### 2019-05-09 main.tex
Initial Overleaf Import

\documentclass{article}
\usepackage[utf8]{inputenc}

\title{scRNA_pipelines_paper}
\author{mark.robinson }
\date{May 2019}

\usepackage{natbib}
\usepackage{graphicx}

\begin{document}

\maketitle

\section{Introduction}
There is a theory which states that if ever anyone discovers exactly what the Universe is for and why it is here, it will instantly disappear and be replaced by something even more bizarre and inexplicable.
There is another theory which states that this has already happened.

\begin{figure}[h!]
\centering
\includegraphics[scale=1.7]{universe}
\caption{The Universe}
\label{fig:universe}
\end{figure}

\section{Conclusion}
``I always thought something was fundamentally wrong with the universe'' \citep{adams1995hitchhiker}

\bibliographystyle{plain}
\bibliography{references}
\end{document}

### 2019-05-09 bmc_article.tex
initial upload

%% BioMed_Central_Tex_Template_v1.06
%%                                      %
%  bmc_article.tex            ver: 1.06 %
%                                       %

%%IMPORTANT: do not delete the first line of this template
%%It must be present to enable the BMC Submission system to
%%recognise this template!!

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                     %%
%%  LaTeX template for BioMed Central  %%
%%     journal article submissions     %%
%%                                     %%
%%          <8 June 2012>              %%
%%                                     %%
%%                                     %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                                                 %%
%% For instructions on how to fill out this Tex template           %%
%% document please refer to Readme.html and the instructions for   %%
%% authors page on the biomed central website                      %%
%% http://www.biomedcentral.com/info/authors/                      %%
%%                                                                 %%
%% Please do not use \input{...} to include other tex files.       %%
%% Submit your LaTeX manuscript as one .tex document.              %%
%%                                                                 %%
%% All additional figures and files should be attached             %%
%% separately and not embedded in the \TeX\ document itself.       %%
%%                                                                 %%
%% BioMed Central currently use the MikTex distribution of         %%
%% TeX for Windows) of TeX and LaTeX.  This is available from      %%
%% http://www.miktex.org                                           %%
%%                                                                 %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% additional documentclass options:
%  [doublespacing]
%  [linenumbers]   - put the line numbers on margins

%%% loading packages, author definitions

%\documentclass[twocolumn]{bmcart}% uncomment this for twocolumn layout and comment line below
\documentclass{bmcart}

%%% Load packages
%\usepackage{amsthm,amsmath}
%\RequirePackage{natbib}
%\RequirePackage[authoryear]{natbib}% uncomment this for author-year bibliography
%\RequirePackage{hyperref}
\usepackage[utf8]{inputenc} %unicode support
%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
\usepackage[pdf]{graphviz}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                             %%
%%  If you wish to display your graphics for   %%
%%  your own use using includegraphic or       %%
%%  includegraphics, then comment out the      %%
%%  following two lines of code.               %%
%%  NB: These line *must* be included when     %%
%%  submitting to BMC.                         %%
%%  All figure files must be submitted as      %%
%%  separate graphics through the BMC          %%
%%  submission process, not included in the    %%
%%  submitted article.                         %%
%%                                             %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\def\includegraphic{}
\def\includegraphics{}

%%% Put your definitions there:
\startlocaldefs
\endlocaldefs

%%% Begin ...
\begin{document}

%%% Start of article front matter
\begin{frontmatter}

\begin{fmbox}
\dochead{Research}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the title of your article here     %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\title{A sample article title}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the authors here                   %%
%%                                          %%
%% Specify information, if available,       %%
%% in the form:                             %%
%%   <key>={<id1>,<id2>}                    %%
%%   <key>=                                 %%
%% Comment or delete the keys which are     %%
%% not used. Repeat \author command as much %%
%% as required.                             %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\author[
   addressref={aff1,aff2},                   % id's of addresses, e.g. {aff1,aff2}
   corref={aff1},                       % id of corresponding address, if any
   email={pierre-luc.germain@hest.ethz.ch}   % email address
]{\inits{PLG}\fnm{Pierre-Luc} \snm{Germain}}
\author[
   addressref={aff1},
   corref={aff1},                       % id of corresponding address, if any
   email={mark.robinson@imls.uzh.ch}
]{\inits{MR}\fnm{Mark} \snm{Robinson}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the authors' addresses here        %%
%%                                          %%
%% Repeat \address commands as much as      %%
%% required.                                %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\address[id=aff1]{
  \orgname{Statistical Bioinformatics, IMLS, University of Z\"{u}rich}, % university, etc
  \street{Winterthurerstrasse 190},
  \postcode{8057}
  \city{Z\"{u}erich},
  \cny{Switzerland}
}
\address[id=aff2]{
  \orgname{D-HEST Institute for Neurosciences, ETH Z\"{u}rich},
  \street{Winterthurerstrasse 190},
  \postcode{8057}
  \city{Z\"{u}erich},
  \cny{Switzerland}
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\end{fmbox}% comment this for two column layout

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% The Abstract begins here                 %%
%%                                          %%
%% Please refer to the Instructions for     %%
%% authors on http://www.biomedcentral.com  %%
%% and include the section headings         %%
%% accordingly for your article type.       %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{abstractbox}

\begin{abstract} % abstract
\parttitle{Background} %if any
Text for this section.
\parttitle{Results} %if any
Text for this section.
\parttitle{Conclusions}
We offer 
\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% The keywords begin here                  %%
%%                                          %%
%% Put each keyword in separate \kwd{}.     %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{keyword}
\kwd{single-cell RNAseq}
\kwd{pipeline}
\kwd{clustering}
\kwd{filtering}
\kwd{benchmark}
\end{keyword}

% MSC classifications codes, if any
%\begin{keyword}[class=AMS]
%\kwd[Primary ]{}
%\kwd{}
%\kwd[; secondary ]{}
%\end{keyword}

\end{abstractbox}
%
%\end{fmbox}% uncomment this for twcolumn layout

\end{frontmatter}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% The Main Body begins here                %%
%%                                          %%
%% Please refer to the instructions for     %%
%% authors on:                              %%
%% http://www.biomedcentral.com/info/authors%%
%% and include the section headings         %%
%% accordingly for your article type.       %%
%%                                          %%
%% See the Results and Discussion section   %%
%% for details on how to create sub-sections%%
%%                                          %%
%% use \cite{...} to cite references        %%
%%  \cite{koon} and                         %%
%%  \cite{oreg,khar,zvai,xjon,schn,pond}    %%
%%  \nocite{smith,marg,hunn,advi,koha,mouse}%%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%% start of article main body
% <put your article body there>

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Background}

Single-cell RNA-sequencing (scRNAseq) and the set of attached analysis methods are evolving fast, and while a number of good comparison/benchmark studies can inform analytical decisions [REFs], they need constant updating and often leave open many details of an analysis. Here, we go in more detail on the various steps of analysis leading from an initial count matrix to a cluster assignment, which are critical steps in wide range of applications. Weharness datasets of known cell composition [TABLE 1] to investigate the impact of various parameters and variations around a core pipeline on various multilevel evaluation metrics.

Although we use some datasets based on other protocols, we focus especially on the 10X datasets and on methods that do not require exogenous control RNA, i.e. spike-ins (see Table 1 and Supplementary Figure 1 for a description of the datasets). As Seurat [REF] was previously shown to outperform other clustering approaches [REFs], we used the Seurat pipeline as the starting framework to perform an integrated investigation of 1) doublet identification, 2) cell filtering, 3) normalization, 4) feature selection, 5) dimensional reduction, 6) clustering. We compare not only competing approaches, but also more fine-grained parameter variations on common methods. Importantly, the success of methods at a certain analytical step might be dependent on choices at other steps. Therefore, instead of evaluating each step in isolation, we developed a general R framework for evaluating nested variations on a pipeline, and suggest a multilevel panel of metrics. Finally, we evaluate several recent methods and provide concrete recommendations.

\section*{Results}

\subsection*{A flexible framework for pipeline evaluation}

We defined a pipeline, in R, as a list of functions executed consecutively on the output of the previous one...  The basic pipeline used through most of this paper is described in Figure XXX. A number of metrics can be considered at each step which can be used to evaluate this and the previous steps.

\subsection*{Doublet detection}

Doublets, defined as two cells sequenced under the same cellular barcode (for instance as a result of being captured in the same droplet), are fairly frequent in scRNAseq datasets, with estimates ranging from X to Y \% [REFs]. While doublets of the same cell type are relatively innocuous for most downstream applications due to their conservation of the relative expression between genes, doublets formed from different cell types or states are likely to be misclassified and could potentially distort downstream analysis. In some cases, doublets can be identify through their unusually high number of reads and detected features, but this is not always the case (Supplementary Figure 2). A number of methods were developed to identify doublets, in most cases by comparing each cell to artificially-created doublets [REFs]. We therefore first evaluated the capacity of these methods to detect doublets using the two 10x datasets produced by Tian et al. (\citey{mixology}) with cells of different genetic identity, and for which therefore SNP information can be used as ground truth. We tested DoubletFinder [REF] and scran's doubletCells [REF], both of which use similarity to artificial doublets, and scds [REF], which relies on a combination of coexpression and binary classification. DoubletFinder integrates a thresholding based on the proportion of expected doublets, while scran and scds return scores which must be manually thresholded. In these cases, we ensured that the right number of cells would be called doublets.

In addition to these methods, we reasoned that an approach such as DoubletFinder could be considerably sped up by being applied directly on counts and by using a pre-clustering to create neotypic doublets more efficiently. We therefore developed a simple and fast package implementing this for doublet detection, with the added perk of accounting for uncertainty in the expected doublet rate and using meta-cells from the clusters to even include triplets.

While most methods accurately identified the doublets in the 3 cell lines dataset (mixology10x3cl), the 5 lines dataset (mixology10x5cl) proved more difficult [Figure]. Of note, our method ran in a fraction of the time required by other methods, and nevertheless proved more accurate [Figure]. We therefore tested whether this method also improved the accuracy of the clustering across all benchmark datasets, and found that it did [Figure].

\subsection*{Filtering: excluded more cells is not necessarily good}

Beyond doublets, a dataset might include low-quality cells whose elimination would reduce noise. This has for instance been demonstrated for cells that have a high content of mitochondrial reads, often as a result of cell degradation [REF]. A common practice is to exclude cells that differ considerably from most other cells on the basis of some such properties, for instance through the `isOutlier` function of `scater` which measures, for a given control property, the number of median absolute deviation of each cell from the mean of all cells. Supplementary Figure 1 shows some of the typical cell properties commonly used. Of note, these properties tend to be correlated, but are not always: for instance, while a high proportion of mitochondrial reads is often correlated with a high proportion of the counts in the top features, there can also be other reasons for an over-representation of highly-expressed features [Supplementary figure?]. In addition, in our experience 10x datasets also exhibit a very tight correlation between the total counts and the total features even across very different cell types (Supplementary Figure 4). We therefore also measure the ratio between the two, and treat cells strongly departing from this trend with suspicion.

Reasoning that the cells we wish to eliminate are cells that would be misclassified, we measured the rate of misclassification of each cell in each dataset across a variety of clustering pipelines, correcting for the median misclassification rate of the subpopulation. We then evaluated what properties could be predictive of misclassification [Supplementary]. The only feature that consistently stood out across multiple datasets was that cells with a very high read counts have a higher chance of being misclassified.

We next investigated the impact of filtering according to various criteria on the accuracy of the clustering. We tested each aforementioned cell property, alone and in combinations, filtering above and/or below a certain threshold defined by different numbers of median absolute deviations. Reasoning that a bad cell would be bad in several distributions, we also included the optional rule that a cell should be excluded if it is an outlier in at least 2 or 3 distributions.

A first observation was that very stringent filtering (removing up to 20\% of the cells) was not generally associated with an increase in accuracy of the clustering [Supplementary], and it somewhat expectedly tended to be biased towards the exclusion of specific subpopulations. We next investigated the criteria [Supplementary]. Interestingly, while filtering out cells with very high total counts and high total features has a significant positive impact on ARI, filtering the lower end of the distribution did not appear to do so, most likely because of prior droplet selection steps.

Surprisingly, filtering out on the basis of the proportion of mitochondrial counts or of counts in the top features did not significantly improve the clustering; however, when testing for the impact of the proportion of mitochondrial counts in the absence of other covariates (i.e. total counts and features), it did have a significant positive impact, suggesting that ‘bad’ cells are anyway excluded on the basis of other distributions. This being said, it must be noted that none of the benchmark datasets has a very high mitochondrial content, and that this filtering might well be important in other datasets. However, filtering solely based on deviations from the distribution might be inadequate when the distribution is tight; in our experience, the proportion of mitochondrial counts in otherwise ‘healthy’ cells can vary up to 8-12\% depending on the tissue. We therefore recommend using a combination of fixed and deviation based threshold, e.g. x $>$ 0.08 and x $>$ 2.5*MADs.

When focusing on the 10x datasets, anova suggests that increasing the number of distribution on which a cell should be an outlier in order to be excluded (`filt.times`) had a significant positive impact on ARI while greatly reducing the of excluded cells; in contrast, increasing the number of MADs leads to a (less robust) decrease in ARI. This suggests that having stringent thresholds but requiring to be an outlier on multiple distributions is preferable.

\subsection*{Normalization and scaling}

\subsection*{Sub-heading for section}
Text for this sub-heading \ldots
\subsubsection*{Sub-sub heading for section}
Text for this sub-sub-heading \ldots
\paragraph*{Sub-sub-sub heading for section}
Text for this sub-sub-sub-heading \ldots
In this section we examine the growth rate of the mean of $Z_0$, $Z_1$ and $Z_2$. In
addition, we examine a common modeling assumption and note the
importance of considering the tails of the extinction time $T_x$ in
studies of escape dynamics.
We will first consider the expected resistant population at $vT_x$ for
some $v>0$, (and temporarily assume $\alpha=0$)

Thus we observe that this expected value is finite for all $v>0$ (also see \cite{koon,khar,zvai,xjon,marg}).
%\nocite{oreg,schn,pond,smith,marg,hunn,advi,koha,mouse}

\section*{Discussion}

For research articles this section should discuss the implications of the findings in context of existing research and highlight limitations of the study. For study protocols and methodology manuscripts this section should include a discussion of any practical or operational issues involved in performing the study and any issues not covered in other sections.

\section*{Conclusions}

This should state clearly the main conclusions and provide an explanation of the importance and relevance of the study to the field.

\section*{Methods}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Backmatter begins here                   %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{backmatter}

\section*{Competing interests}
  The authors declare that they have no competing interests.

\section*{Author's contributions}
    Text for this section \ldots

\section*{Acknowledgements}
  Text for this section \ldots
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                  The Bibliography                       %%
%%                                                         %%
%%  Bmc_mathpys.bst  will be used to                       %%
%%  create a .BBL file for submission.                     %%
%%  After submission of the .TEX file,                     %%
%%  you will be prompted to submit your .BBL file.         %%
%%                                                         %%
%%                                                         %%
%%  Note that the displayed Bibliography will not          %%
%%  necessarily be rendered by Latex exactly as specified  %%
%%  in the online Instructions for Authors.                %%
%%                                                         %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% if your bibliography is in bibtex format, use those commands:
\bibliographystyle{bmc-mathphys} % Style BST file (bmc-mathphys, vancouver, spbasic).
\bibliography{bmc_article}      % Bibliography file (usually '*.bib' )
% for author-year bibliography (bmc-mathphys or spbasic)
% a) write to bib file (bmc-mathphys only)
% @settings{label, options="nameyear"}
% b) uncomment next line
%\nocite{label}

% or include bibliography directly:
% \begin{thebibliography}
% \bibitem{b1}
% \end{thebibliography}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                               %%
%% Figures                       %%
%%                               %%
%% NB: this is for captions and  %%
%% Titles. All graphics must be  %%
%% submitted separately and NOT  %%
%% included in the Tex document  %%
%%                               %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%
%% Do not use \listoffigures as most will included as separate files

\section*{Figures}
  \begin{figure}[h!]
  \caption{\csentence{Sample figure title.}
      A short description of the figure content
      should go here.}
      \end{figure}

\begin{figure}[h!]
  \caption{\csentence{Sample figure title.}
      Figure legend text.}
      \end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                               %%
%% Tables                        %%
%%                               %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% Use of \listoftables is discouraged.
%%
\section*{Tables}
\begin{table}[h!]
\caption{Sample table title. This is where the description of the table should go.}
      \begin{tabular}{cccc}
        \hline
           & B1  &B2   & B3\\ \hline
        A1 & 0.1 & 0.2 & 0.3\\
        A2 & ... & ..  & .\\
        A3 & ..  & .   & .\\ \hline
      \end{tabular}
\end{table}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                               %%
%% Additional Files              %%
%%                               %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section*{Additional Files}
  \subsection*{Additional file 1 --- Sample additional file title}
    Additional file descriptions text (including details of how to
    view the file, if it is in a non-standard format or the file extension).  This might
    refer to a multi-page table or a figure.

  \subsection*{Additional file 2 --- Sample additional file title}
    Additional file descriptions text.

\end{backmatter}
\end{document}

### 2019-05-09 main_figures/figure1_doublets.Rmd
initial upload

---
title: "Figure 1 doublets"
author: "Pierre-Luc Germain"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    keep_md: true

---

```{r}
suppressPackageStartupMessages({
  library(SingleCellExperiment)
  library(ROCit)
  library(cowplot)
})
# we load the dataset
sce <- readRDS("../datasets/mixology10x5cl.SCE.rds")
CD <- as.data.frame(colData(sce))
```

```{r}
truth <- factor(CD$demuxlet_cls, c("SNG","DBL"))
roclist <- lapply( list(plDoubletFinder=CD$pldf.ratio,
                        DoubletFinder.p=CD$doubletFinder.pANN_0.25_0.02_82,
                        scran.score=CD$scran.doubletscore,
                        scds.hybrid=CD$hybrid_score), 
                   class=truth, FUN=rocit)
colors <- c("blue","red","black","pink")
plot(roclist[[1]], lwd=2, col=c(colors[[1]], "grey"), legend=FALSE, YIndex=FALSE, main="mixology10x5cl")
for(i in 2:length(roclist)){
  lines(roclist[[i]]$FPR, roclist[[i]]$TPR, lwd=2, col=colors[[i]])
}
legend("bottomright", fill=colors, legend=names(roclist))
```

### 2019-05-09 main_figures/figure2_filtering.Rmd
initial upload

---
title: "Figure 2 filtering"
author: "Pierre-Luc Germain"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    keep_md: true

---

### 2019-05-09 supplementary_figures/supp1_dataset_description.Rmd
initial upload

---
title: "Supplementary Figures 1-x"
author: "Pierre-Luc Germain"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r, include=FALSE}
if(!exists('FIG_NB')){
  FIG_NB <- 0
  getFigNb <- function(increment=FALSE){
    if(increment) FIG_NB <<- FIG_NB + 1
    FIG_NB
  }
}
```

```{r}
suppressPackageStartupMessages({
  library(ggplot2)
  library(cowplot)
  library(SingleCellExperiment)
  library(scater)
  library(reshape2)
  library(pheatmap)
})
suppressMessages(devtools::load_all("../../pipComp/"))

# we load the datasets
lf <- list.files("../datasets/", full.names=TRUE)
names(lf) <- gsub(".SCE.rds", "", basename(lf), fixed=TRUE)
SCEs <- lapply(lf, readRDS)
```

# Supplementary Figure `r getFigNb(TRUE)`

```{r dist_cell_properties, fig.width=8.5, fig.height=8.5}
vars <- c( "log10(features)"="log10_total_features", "log10(counts)"="log10_total_counts", 
           "% in top 20"="pct_counts_in_top_20_features", "% mito"="pct_counts_Mt", 
           "% ribosomal"="pct_counts_ribosomal", "% coding"="pct_counts_coding" )
plot_grid(plotlist=lapply(1:length(vars), FUN=function(x){ 
  p <- plotFilters(SCEs, vars[x], nmads=c(2,5), ncol=1) + xlab(names(vars)[x])
  p <- p + scale_y_continuous(breaks=scales::pretty_breaks(2))
  p <- p + scale_x_continuous(breaks=scales::pretty_breaks(3))
  p + theme( axis.title.y = element_blank(), 
             strip.text.x = element_text(size=10),
             axis.title=element_text(size=11),
             axis.text=element_text(size=11) )
}), nrow=1)
```

### Supplementary Figure `r getFigNb()`
Distribution across cells of various control properties in the different datasets. The 
lines indicate respectively 2 and 5 median absolute deviations (MADs).

\newpage

# Supplementary Figure `r getFigNb(TRUE)`

```{r mixology_doublet_featcount, fig.height=4}
plotDoublets <- function(sce, title="", mads=c(2,2.5,3), vars=c("log10_total_features", "log10_total_counts")){
  cd <- as.data.frame(colData(sce))
  p <- ggplot(cd, aes_string(x=vars[1], y=vars[2], colour="demuxlet_cls")) + geom_point()
  p <- p + geom_point(data=cd[which(cd$demuxlet_cls=="DBL"),], size=3)
  lt <- c("solid","dashed","dotted","dotdash")
  for(i in 1:length(mads)){
    th <- .tmads(cd[[vars[1]]], mads[i])[2]
    p <- p + geom_vline(aes_string(xintercept=th), linetype=lt[i])
    th <- .tmads(cd[[vars[2]]], mads[i])[2]
    p <- p + geom_hline(aes_string(yintercept=th), linetype=lt[i])
  }
  p  
}
pl <- list( plotDoublets(SCEs$mixology10x3cl) + ggtitle("mixology10x3cl"),
            plotDoublets(SCEs$mixology10x5cl) + ggtitle("mixology10x5cl") )
plot_grid(plotlist=pl, labels=LETTERS[1:2])
```

### Supplementary Figure `r getFigNb()`
The total counts and total features per cell of doublets versus other cells. We used the demuxlet annotation of doublets (based on SNPs) made available through CellBench. While doublets tend to have a higher total count and especially number of detected features, these features alone are not always sufficient for their identification.

\newpage

# Supplementary Figure `r getFigNb(TRUE)`
[to come]
### Supplementary Figure `r getFigNb()`
Doublets that cannot be identified by our method show lower signs of being doublets.

\newpage

# Supplementary Figure `r getFigNb(TRUE)`

```{r featcount_ratio}
d <- data.frame( dataset=rep(names(SCEs),sapply(SCEs,ncol)), 
                 log10_counts=unlist(lapply(SCEs,FUN=function(x) x$log10_total_counts)),
                 log10_features=unlist(lapply(SCEs,FUN=function(x) x$log10_total_features)),
                 featcount_dist=unlist(lapply(SCEs,FUN=function(x) x$featcount_dist)))
ggplot(d, aes(x=log10_features, y=log10_counts, colour=featcount_dist)) + geom_point() + facet_wrap(~dataset, scale="free") + scale_colour_gradient2(limits=c(-0.2,0.2))
```

### Supplementary Figure `r getFigNb()`
There is a tight relationship, in 10x datasets (i.e. not the `Koh` and `Kumar` datasets), between the total counts of a cell and its number of detected features. We therefore include, among control variables, deviation from this ratio.

### 2019-05-09 supplementary_figures/supp_figures.Rmd
initial upload

---
title: "Supplementary Figures"
author: "Pierre-Luc Germain"
date: "`r format(Sys.time(), '%d %B, %Y')`"
geometry: margin=1.5cm
output:
  pdf_document:
    keep_tex: true
    toc: true
    fig_width: 8
    fig_height: 6
    citation_package: natbib
bibliography: ../bmc_article.bib
---

```{r setup, include=FALSE}
# comment this out to print code
knitr::opts_chunk$set(echo=FALSE)
FIG_NB <- 0
getFigNb <- function(increment=FALSE){
  if(increment) FIG_NB <<- FIG_NB + 1
  FIG_NB
}
```

\newpage

```{r child='supp1_dataset_description.Rmd'}
```

### 2019-05-09 supplementary_figures/supp_figures.tex
initial upload

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1.5cm]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Supplementary Figures},
            pdfauthor={Pierre-Luc Germain},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{natbib}
\bibliographystyle{plainnat}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\providecommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}

  \title{Supplementary Figures}
    \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
    \author{Pierre-Luc Germain}
    \preauthor{\centering\large\emph}
  \postauthor{\par}
      \predate{\centering\large\emph}
  \postdate{\par}
    \date{09 Mai, 2019}

\begin{document}
\maketitle

{
\setcounter{tocdepth}{2}
\tableofcontents
}
\newpage

\hypertarget{supplementary-figure-1}{%
\section{Supplementary Figure 1}\label{supplementary-figure-1}}

\includegraphics{supp_figures_files/figure-latex/unnamed-chunk-4-1.pdf}

\hypertarget{supplementary-figure-1-1}{%
\subsubsection{Supplementary Figure 1}\label{supplementary-figure-1-1}}

Distribution across cells of various control properties in the different
datasets. The lines indicate respectively 2 and 5 median absolute
deviations (MADs).

\newpage

\hypertarget{supplementary-figure-2}{%
\section{Supplementary Figure 2}\label{supplementary-figure-2}}

\includegraphics{supp_figures_files/figure-latex/unnamed-chunk-5-1.pdf}

\hypertarget{supplementary-figure-2-1}{%
\subsubsection{Supplementary Figure 2}\label{supplementary-figure-2-1}}

The total counts and total features per cell of doublets versus other
cells. We used the demuxlet annotation of doublets (based on SNPs) made
available through CellBench. While doublets tend to have a higher total
count and especially number of detected features, these features alone
are not always sufficient for their identification.

\newpage

\hypertarget{supplementary-figure-3}{%
\section{Supplementary Figure 3}\label{supplementary-figure-3}}

{[}to come{]} \#\#\# Supplementary Figure 3 Doublets that cannot be
identified by our method show lower signs of being doublets.

\newpage

\hypertarget{supplementary-figure-4}{%
\section{Supplementary Figure 4}\label{supplementary-figure-4}}

\includegraphics{supp_figures_files/figure-latex/unnamed-chunk-6-1.pdf}

\hypertarget{supplementary-figure-4-1}{%
\subsubsection{Supplementary Figure 4}\label{supplementary-figure-4-1}}

There is a tight relationship, in 10x datasets (i.e.~not the
\texttt{Koh} and \texttt{Kumar} datasets), between the total counts of a
cell and its number of detected features. We therefore include, among
control variables, deviation from this ratio.

\bibliography{../bmc\_article.bib}

\end{document}

### 2019-05-13 bmc_article.tex
...

@@
 
-Doublets, defined as two cells sequenced under the same cellular barcode (for instance as a result of being captured in the same droplet), are fairly frequent in scRNAseq datasets, with estimates ranging from X to Y \% [REFs].
+Doublets, defined as two cells sequenced under the same cellular barcode (for instance as a result of being captured in the same droplet), are fairly frequent in scRNAseq datasets, with estimates ranging from 1 to 10\% depending on the platform and cell concentration used \citep{bloom_estimating_2018,kang_multiplexed_2018}.
 While doublets of the same cell type are relatively innocuous for most downstream applications due to their conservation of the relative expression between genes, doublets formed from different cell types or states are likely to be misclassified and could potentially distort downstream analysis.
-In some cases, doublets can be identify through their unusually high number of reads and detected features, but this is not always the case (Supplementary Figure 2).
+In some cases, doublets can be identified through their unusually high number of reads and detected features, but this is not always the case (Supplementary Figure 2).
 A number of methods were developed to identify doublets, in most cases by comparing each cell to artificially-created doublets [REFs].
 We therefore first evaluated the capacity of these methods to detect doublets using the two 10x datasets produced by Tian et al.
-(\citey{mixology}) with cells of different genetic identity, and for which therefore SNP information can be used as ground truth.
+(\citey{tian_mixology_2018}) with cells of different genetic identity, and for which therefore SNP information can be used as ground truth.
 We tested DoubletFinder [REF] and scran's doubletCells [REF], both of which use similarity to artificial doublets, and scds [REF], which relies on a combination of coexpression and binary classification.
@@
 In addition to these methods, we reasoned that an approach such as DoubletFinder could be considerably sped up by being applied directly on counts and by using a pre-clustering to create neotypic doublets more efficiently.
-We therefore developed a simple and fast package implementing this for doublet detection, with the added perk of accounting for uncertainty in the expected doublet rate and using meta-cells from the clusters to even include triplets.
+We therefore developed a simple and fast package implementing this for doublet detection, with the added perk of accounting for uncertainty in the expected doublet rate and using meta-cells from the clusters to even include triplets (see methods).
 
-While most methods accurately identified the doublets in the 3 cell lines dataset (mixology10x3cl), the 5 lines dataset (mixology10x5cl) proved more difficult [Figure].
-Of note, our method ran in a fraction of the time required by other methods, and nevertheless proved more accurate [Figure].
-We therefore tested whether this method also improved the accuracy of the clustering across all benchmark datasets, and found that it did [Figure].
+While most methods accurately identified the doublets in the 3 cell lines dataset (mixology10x3cl), the 5 lines dataset (mixology10x5cl) proved more difficult (Figure 1B).
+Our method achieved the best accuracy while running in a very reasonable time (Figure 1).
+Across datasets, cells called as doublets tended to be classified in the wrong cluster more often than other cells (Figure 1D).
+We therefore tested whether this method also improved the accuracy of the clustering across all benchmark datasets, and found that it consistently did (Figure 2).
 
@@
 A common practice is to exclude cells that differ considerably from most other cells on the basis of some such properties, for instance through the `isOutlier` function of `scater` which measures, for a given control property, the number of median absolute deviation of each cell from the mean of all cells.
-Supplementary Figure 1 shows some of the typical cell properties commonly used.
-Of note, these properties tend to be correlated, but are not always: for instance, while a high proportion of mitochondrial reads is often correlated with a high proportion of the counts in the top features, there can also be other reasons for an over-representation of highly-expressed features [Supplementary figure?].
-In addition, in our experience 10x datasets also exhibit a very tight correlation between the total counts and the total features even across very different cell types (Supplementary Figure 4).
+Supplementary Figure 1 shows the distributions of some of the typical cell properties commonly used.
+Of note, these properties tend to be correlated, but are not always: for instance, while a high proportion of mitochondrial reads is often correlated with a high proportion of the counts in the top features, there can also be other reasons for an over-representation of highly-expressed features (Supplementary Figure 3).
+In addition, in our experience 10X datasets also exhibit a very tight correlation between the total counts and the total features even across very different cell types (Supplementary Figure 4).
 We therefore also measure the ratio between the two, and treat cells strongly departing from this trend with suspicion.
 
-Reasoning that the cells we wish to eliminate are cells that would be misclassified, we measured the rate of misclassification of each cell in each dataset across a variety of clustering pipelines, correcting for the median misclassification rate of the subpopulation.
-We then evaluated what properties could be predictive of misclassification [Supplementary].
-The only feature that consistently stood out across multiple datasets was that cells with a very high read counts have a higher chance of being misclassified.
+Reasoning that the cells we wish to eliminate are cells that would be misclassified, we measured the rate of misclassification of each cell in each dataset across a variety of clustering pipelines, correcting for the median misclassification rate of the subpopulation, and then evaluated what properties could be predictive of misclassification (Supplementary Figures 5-7).
+We could not identify any property or simple combination thereof which would be consistently predictive of misclassification; the only feature that consistently stood out across multiple datasets (the Zheng datasets) was that cells with a very high read counts have a higher chance of being misclassified.
 
@@
 
-A first observation was that very stringent filtering (removing up to 20\% of the cells) was not generally associated with an increase in accuracy of the clustering [Supplementary], and it somewhat expectedly tended to be biased towards the exclusion of specific subpopulations.
-We next investigated the criteria [Supplementary].
-Interestingly, while filtering out cells with very high total counts and high total features has a significant positive impact on ARI, filtering the lower end of the distribution did not appear to do so, most likely because of prior droplet selection steps.
+A first observation was that very stringent filtering (removing up to 20\% of the cells) was not generally associated with an increase in accuracy of the clustering (Supplementary Figure 8A).
+Moreover, stringent filtering tended to be biased towards the exclusion of specific subpopulations (Supplementary Figure 8B).
+Indeed, applying the same filtering criteria on individual clusters of cells (identified through scran's `quickCluster` method) resulted in nearly no cell being filtered out, which suggests that filtering on the global population tends to discard cells of subpopulations with more extreme properties, rather than `bad` cells.
 
-Surprisingly, filtering out on the basis of the proportion of mitochondrial counts or of counts in the top features did not significantly improve the clustering; however, when testing for the impact of the proportion of mitochondrial counts in the absence of other covariates (i.e.
-total counts and features), it did have a significant positive impact, suggesting that ‘bad’ cells are anyway excluded on the basis of other distributions.
-This being said, it must be noted that none of the benchmark datasets has a very high mitochondrial content, and that this filtering might well be important in other datasets.
-However, filtering solely based on deviations from the distribution might be inadequate when the distribution is tight; in our experience, the proportion of mitochondrial counts in otherwise ‘healthy’ cells can vary up to 8-12\% depending on the tissue.
+We next investigated the contribution of various criteria individually (Supplementary Figure 9).
+Interestingly, regression analysis suggests that increasing the number of distributions on which a cell must deviate in order to be excluded tended to be positively associated with ARI, while increasing the number of median absolute deviations required to be an outlier as associated with a decrease in ARI.
+We therefore suggest excluding cells that are outliers on more than 1 distributions.
+The impact of filtering on the basis of total counts or feature counts was not always robust, nor always positive, and we could detect no significant impact of filtering on the basis of the proportion of mitochondrial reads - however it must be said that the benchmark datasets count very few cells with a very high proportion of mitochondrial reads (Supplementary Figure 1).
+We therefore suggest to keep this criterion, but in a relatively lenient fashion, and not solely based on the median absolute deviation of the distribution.
+Filtering solely based on deviations from the distribution might be inadequate when the distribution is tight; in our experience, the proportion of mitochondrial counts in otherwise `healthy` cells can vary up to 8-12\% depending on the tissue.
 We therefore recommend using a combination of fixed and deviation based threshold, e.g.
@@
 
-When focusing on the 10x datasets, anova suggests that increasing the number of distribution on which a cell should be an outlier in order to be excluded (`filt.times`) had a significant positive impact on ARI while greatly reducing the of excluded cells; in contrast, increasing the number of MADs leads to a (less robust) decrease in ARI.
-This suggests that having stringent thresholds but requiring to be an outlier on multiple distributions is preferable.
+On the basis of these observations, we defined three sets of criteria as lenient, default and stringent filtering (see methods), and investigated them further.
 

### 2019-05-13 main_figures/figure1_doublets.Rmd
...

@@
   library(ROCit)
+  library(ggplot2)
   library(cowplot)
 })
+theme_set(theme_cowplot(font_size = 12))
+
 # we load the dataset
-sce <- readRDS("../datasets/mixology10x5cl.SCE.rds")
-CD <- as.data.frame(colData(sce))
+lf <- list.files("../datasets/", pattern="SCE", full.names=TRUE)
+names(lf) <- sapply(strsplit(basename(lf),".",fixed=TRUE), head, 1)
+CDs <- lapply(lf, FUN=function(x){ as.data.frame(colData(readRDS(x))) })
 ```
@@
 ```{r}
+colors <- c("#4477AA", "#117733", "#DDCC77", "#CC6677")
+CD <- CDs$mixology10x3cl
 truth <- factor(CD$demuxlet_cls, c("SNG","DBL"))
-roclist <- lapply( list(plDoubletFinder=CD$pldf.ratio,
-                        DoubletFinder.p=CD$doubletFinder.pANN_0.25_0.02_82,
-                        scran.score=CD$scran.doubletscore,
-                        scds.hybrid=CD$hybrid_score), 
+roclist <- lapply( list(scDblFinder=CD$pldf.ratio,
+                        scds.hybrid=CD$hybrid_score,
+                        scran.score=CD$scran.DblScore),
                    class=truth, FUN=rocit)
-colors <- c("blue","red","black","pink")
-plot(roclist[[1]], lwd=2, col=c(colors[[1]], "grey"), legend=FALSE, YIndex=FALSE, main="mixology10x5cl")
-for(i in 2:length(roclist)){
-  lines(roclist[[i]]$FPR, roclist[[i]]$TPR, lwd=2, col=colors[[i]])
+d <- data.frame( method=rep( c("scDblFinder","scds","scran"),
+                             sapply(roclist, FUN=function(x) length(x$TPR)) ),
+                 FPR=unlist(lapply(roclist,FUN=function(x) x$FPR)),
+                 TPR=unlist(lapply(roclist,FUN=function(x) x$TPR)) )
+p1 <- ggplot(d, aes(FPR, TPR, colour=method)) + geom_line(size=1.5) + ggtitle("mixology10x3cl")
+p1 <- p1 + theme(legend.position = c(0.6, 0.3)) + scale_color_manual(values=colors[-2])
+
+CD <- CDs$mixology10x5cl
+truth <- factor(CD$demuxlet_cls, c("SNG","DBL"))
+roclist <- lapply( list(scDblFinder=CD$pldf.ratio,
+                        DoubletFinder=CD$doubletFinder.pANN_0.25_0.02_82,
+                        scds=CD$hybrid_score,
+                        scran=CD$scran.DblScore), 
+                   class=truth, FUN=rocit)
+d <- data.frame( method=factor( rep(names(roclist), sapply(roclist, FUN=function(x) length(x$TPR))), levels=names(roclist)),
+                 FPR=unlist(lapply(roclist,FUN=function(x) x$FPR)),
+                 TPR=unlist(lapply(roclist,FUN=function(x) x$TPR)) )
+p2 <- ggplot(d, aes(FPR, TPR, colour=method)) + geom_line(size=1.5) + ggtitle("mixology10x5cl")
+p2 <- p2 + theme(legend.position = c(0.6, 0.3)) + scale_color_manual(values=colors)
+```
+
+```{r}
+rt <- data.frame( time=c(12.83, 41.46, 25.08, NA, 56.15, 310.24, 268.93, 722.031),
+                  method=rep(c("scds", "scran", "scDblFinder", "DoubletFinder"),2),
+                  dataset=rep(c("mixology10x3cl", "mixology10x5cl"), each=4))
+p3 <- ggplot(rt, aes(method, time, fill=method)) + geom_col() + scale_color_manual(values=colors) + facet_wrap(~dataset, scale="free", nrow=2) + coord_flip() + ylab("Wall-time, in seconds") + theme(legend.position="none", axis.title.y=element_blank(), plot.margin=unit(c(1,2,1,1),"lines"))
+```
+
+```{r}
+mis <- readRDS("../data/misclassification.rds")
+dbl <- lapply(CDs, FUN=function(x) x[,c("pldf.ratio","pldf.class")])
+dbl <- dbl[intersect(names(dbl),names(mis))]
+for(n in names(dbl)){
+  dbl[[n]]$pct_misclassified_cor <- mis[[n]]$corrected_pct_mis[row.names(dbl[[n]])]
+  dbl[[n]]$pct_misclassified <- mis[[n]]$pct_misclassified[row.names(dbl[[n]])]
 }
-legend("bottomright", fill=colors, legend=names(roclist))
+d <- as.data.frame(data.table::rbindlist(dbl, idcol="dataset"))
+p4 <- ggplot(d, aes(pldf.class, pct_misclassified_cor, fill=pldf.class)) + geom_violin() + facet_wrap(~dataset) + ggtitle("Cluster mis-assignment") + theme(axis.title.x=element_blank(), legend.position="none") + ylab("Corrected % misclassification") + stat_summary(fun.y="mean", geom="point")
 ```
+
+```{r, fig.height=6, fig.width=7}
+plot_grid( plot_grid(p1, p2, nrow=1, labels=LETTERS[1:2]), plot_grid(p3, p4, nrow=1, labels=LETTERS[3:4], rel_widths=c(2,3)), nrow=2)
+```
+
+**A-B** Receiver operating characteristic (ROC) curves of the tested doublet detection methods for the mixology10x3l (**A**) and mixology10x5cl (**B**) datasets.
+DoubletFinder failed due to an error on the mixology10x3cl dataset.
+**C:** Running time of the different methods.
+**D:** Cell called as doublets by scDblFinder tend to be more often misclassified, i.e.
+assigned a wrong cluster.
+The corrected percentage of misclassification is the proportion of times a given cell is misclassified across hundreds of clustering pipelines with various parameters, minus the median misclassification rate of the cell's subpopulation.

### 2019-05-13 supplementary_figures/supp1_dataset_description.Rmd
...

@@
 ```{r mixology_doublet_featcount, fig.height=4}
+theme_set(theme_cowplot(font_size = 13))
 plotDoublets <- function(sce, title="", mads=c(2,2.5,3), vars=c("log10_total_features", "log10_total_counts")){
   cd <- as.data.frame(colData(sce))
-  p <- ggplot(cd, aes_string(x=vars[1], y=vars[2], colour="demuxlet_cls")) + geom_point()
+  p <- ggplot(cd, aes_string(x=vars[1], y=vars[2], colour="demuxlet_cls")) + geom_point() + theme(legend.position="none")
   p <- p + geom_point(data=cd[which(cd$demuxlet_cls=="DBL"),], size=3)
@@
 ### Supplementary Figure `r getFigNb()`
-The total counts and total features per cell of doublets versus other cells.
+The total counts and total features per cell of doublets (red) versus other cells.
 We used the demuxlet annotation of doublets (based on SNPs) made available through CellBench.
+The lines indicate, respectively, 2, 2.5, and 3 median absolute deviations.
 While doublets tend to have a higher total count and especially number of detected features, these features alone are not always sufficient for their identification.
-
-\newpage
-
-# Supplementary Figure `r getFigNb(TRUE)`
-[to come]
-### Supplementary Figure `r getFigNb()`
-Doublets that cannot be identified by our method show lower signs of being doublets.
 
@@
 We therefore include, among control variables, deviation from this ratio.
+
+\newpage
+# Supplementary Figure `r getFigNb(TRUE)`
+
+```{r cellprops_cor, fig.width=7.5, fig.height=7}
+d <- data.frame( dataset=rep(names(SCEs),sapply(SCEs,ncol)), 
+                 pct_counts_Mt=unlist(lapply(SCEs,FUN=function(x) x$pct_counts_Mt)),
+                 pct_top_50=unlist(lapply(SCEs,FUN=function(x) x$pct_counts_in_top_50_features)),
+                 log10_features=unlist(lapply(SCEs,FUN=function(x) x$log10_total_features)))
+plot_grid( ggplot(d, aes(x=pct_top_50, y=pct_counts_Mt)) + geom_point() + facet_wrap(~dataset, scale="free") + geom_density_2d(),
+           ggplot(d, aes(x=log10_features, y=pct_counts_Mt)) + geom_point() + facet_wrap(~dataset, scale="free") + geom_density_2d(),
+           labels=LETTERS[1:2], nrow=2 )
+
+ggplot(d, aes(x=log10_features, y=log10_counts, colour=featcount_dist)) + geom_point() + facet_wrap(~dataset, scale="free") + scale_colour_gradient2(limits=c(-0.2,0.2))
+```
+
+### Supplementary Figure `r getFigNb()`
+There is a tight relationship, in 10x datasets (i.e.
+not the `Koh` and `Kumar` datasets), between the total counts of a cell and its number of detected features.
+We therefore include, among control variables, deviation from this ratio.
+
+\newpage
+
+# Supplementary Figure `r getFigNb(TRUE)`
+
+```{r misclass, fig.height=4}
+theme_set(theme_cowplot(font_size = 11))
+mis <- readRDS("../data/misclassification.rds")
+pdb <- function(x, ...){
+  n <- names(mis)[x]
+  plotBadCells(SCEs[[n]], mis[[n]], name=n, ...)
+}
+pdb(2, labels=c("A",rep("",6)))
+```
+\vfill
+```{r, fig.height=4}
+pdb(3, labels=c("B",rep("",6)))
+```
+
+### Supplementary Figure `r getFigNb()`
+Relationship between various cellular properties and the frequency of cluster mis-assignment for the mixology10x3cl (A) and mixology10x5cl (B) datasets.
+The percentage of misclassification refers to the frequency with which a given cell is assigned the wrong cluster (using the Hungarian algorithm for cluster matching) across several hundred clustering runs with varying parameters.
+Since some subpopulations tend to be more misclassified than others, the adjusted rate of misclassification (`adj_mis`) is substracted for the subpopulation median misclassification rate.
+
+\newpage
+
+# Supplementary Figure `r getFigNb(TRUE)`
+
+```{r, fig.height=4}
+pdb(4, labels=c("A",rep("",6)))
+```
+\vfill
+```{r, fig.height=4}
+pdb(5, labels=c("B",rep("",6)))
+```
+
+### Supplementary Figure `r getFigNb()`
+Relationship between various cellular properties and the frequency of cluster mis-assignment for the Zheng equal (A) or unequal (B) mixtures of four cell types.
+See Supplementary Figure `r getFigNb()-1` for more information.
+The only clear pattern is that cells with a high number of reads or features tend to have a higher misclassification rate.
+
+\newpage
+
+# Supplementary Figure `r getFigNb(TRUE)`
+
+```{r, fig.height=4}
+pdb(6)
+```
+
+### Supplementary Figure `r getFigNb()`
+Relationship between various cellular properties and the frequency of cluster mis-assignment for the Zheng mixture of 8 cell types. See Supplementary Figure `r getFigNb()-2` for more information.
+
+\newpage

### 2019-05-13 supplementary_figures/supp_figures.tex
...

@@
   \postdate{\par}
-    \date{09 Mai, 2019}
+    \date{12 Mai, 2019}
 
@@
 
-\includegraphics{supp_figures_files/figure-latex/unnamed-chunk-4-1.pdf}
+\includegraphics{supp_figures_files/figure-latex/dist_cell_properties-1.pdf}
+
+\vfill
 
@@
 
-\includegraphics{supp_figures_files/figure-latex/unnamed-chunk-5-1.pdf}
+\includegraphics{supp_figures_files/figure-latex/mixology_doublet_featcount-1.pdf}
+
+\vfill
 
@@
 
-The total counts and total features per cell of doublets versus other
-cells. We used the demuxlet annotation of doublets (based on SNPs) made
-available through CellBench. While doublets tend to have a higher total
-count and especially number of detected features, these features alone
-are not always sufficient for their identification.
+The total counts and total features per cell of doublets (red) versus
+other cells. We used the demuxlet annotation of doublets (based on SNPs)
+made available through CellBench. The lines indicate, respectively, 2,
+2.5, and 3 median absolute deviations. While doublets tend to have a
+higher total count and especially number of detected features, these
+features alone are not always sufficient for their identification.
 
@@
 
-{[}to come{]} \#\#\# Supplementary Figure 3 Doublets that cannot be
-identified by our method show lower signs of being doublets.
+\includegraphics{supp_figures_files/figure-latex/featcount_ratio-1.pdf}
+
+\vfill
+
+\hypertarget{supplementary-figure-3-1}{%
+\subsubsection{Supplementary Figure 3}\label{supplementary-figure-3-1}}
+
+There is a tight relationship, in 10x datasets (i.e.~not the
+\texttt{Koh} and \texttt{Kumar} datasets), between the total counts of a
+cell and its number of detected features. We therefore include, among
+control variables, deviation from this ratio.
 
@@
 
-\includegraphics{supp_figures_files/figure-latex/unnamed-chunk-6-1.pdf}
+\includegraphics{supp_figures_files/figure-latex/misclass-1.pdf}
+\vspace{1cm}
+\includegraphics{supp_figures_files/figure-latex/unnamed-chunk-4-1.pdf}
+
+\vfill
 
@@
 
-There is a tight relationship, in 10x datasets (i.e.~not the
-\texttt{Koh} and \texttt{Kumar} datasets), between the total counts of a
-cell and its number of detected features. We therefore include, among
-control variables, deviation from this ratio.
+Relationship between various cellular properties and the frequency of
+cluster mis-assignment for the mixology10x3cl (A) and mixology10x5cl (B)
+datasets. The percentage of misclassification refers to the frequency
+with which a given cell is assigned the wrong cluster (using the
+Hungarian algorithm for cluster matching) across several hundred
+clustering runs with varying parameters. Since some subpopulations tend
+to be more misclassified than others, the adjusted rate of
+misclassification (\texttt{adj\_mis}) is substracted for the
+subpopulation median misclassification rate.
+
+\newpage
+
+\hypertarget{supplementary-figure-5}{%
+\section{Supplementary Figure 5}\label{supplementary-figure-5}}
+
+\includegraphics{supp_figures_files/figure-latex/unnamed-chunk-5-1.pdf}
+\vspace{1cm}
+\includegraphics{supp_figures_files/figure-latex/unnamed-chunk-6-1.pdf}
+
+\vfill
+
+\hypertarget{supplementary-figure-5-1}{%
+\subsubsection{Supplementary Figure 5}\label{supplementary-figure-5-1}}
+
+Relationship between various cellular properties and the frequency of
+cluster mis-assignment for the Zheng equal (A) or unequal (B) mixtures
+of four cell types. See Supplementary Figure 4 for more information. The
+only clear pattern is that cells with a high number of reads or features
+tend to have a higher misclassification rate.
+
+\newpage
+
+\hypertarget{supplementary-figure-6}{%
+\section{Supplementary Figure 6}\label{supplementary-figure-6}}
+
+\includegraphics{supp_figures_files/figure-latex/unnamed-chunk-7-1.pdf}
+
+\vfill
+
+\hypertarget{supplementary-figure-6-1}{%
+\subsubsection{Supplementary Figure 6}\label{supplementary-figure-6-1}}
+
+Relationship between various cellular properties and the frequency of
+cluster mis-assignment for the Zheng mixture of 8 cell types. See
+Supplementary Figure 4 for more information.
+
+\newpage
 

### 2019-05-15 bmc_article.tex
...

@@
 
-\subsection*{Doublet detection}
+\subsection*{Filtering}
+
+\subsubsection*{Doublet detection}
 
@@
 
-\subsection*{Filtering: excluded more cells is not necessarily good}
+\subsubsection*{Excluded more cells is not necessarily good}
 
@@
 
-On the basis of these observations, we defined three sets of criteria as lenient, default and stringent filtering (see methods), and investigated them further.
+On the basis of these observations, we defined three sets of criteria as lenient, default and stringent filtering (see methods), and investigated them further. [...]
+
+\subsubsection*{Filtering features by type}
+
+Mitochondrial reads have been associated with cell degradation, and it has been observed that much of scRNAseq clustering is determined by ribosomal genes [REF].
+We therefore considered whether excluding one category of features or the other, or using only protein-coding genes, had impact on the ability to distinguish subpopulations.
 
 \subsection*{Normalization and scaling}
+
+We next investigated the impact of different normalization strategies.
+In addition to log-normalization, the standard Seurat clustering pipeline scales features so that the PCA is not dominated by a small subset of the features, and we therefore included versions of the different normalization, with or without a subsequent scaling (except sctransform's variance-stabilizing transformation, given that, conceptually, it involves something analogous to scaling).
+Finally, Seurat's `ScaleData` includes the option to regress out the effect of certain covariates, and we therefore tested its use with the proportion of mitochondrial counts and the number of detected features.
+
+An important motivation of sctransform was the observation that, even after normalization, the first principal components of various datasets tended to correlate with library size, suggesting an imperfect correlation \citep{hafemeister_sctransform_2019}.
+We therefore first assessed to what extent the first principal component still retained a correlation with the library size and number of detected features (Supplementary XX).
+In the 10X datasets, most methods successfully decreased the correlation with covariates, including the mere scaling of non-normalized data.
+This was not the case in the smartseq datasets, which required the additional regressing out of covariates.
+
+(NOTE: a correlation of PC1 with library size might be caused by the fact that subpopulations have a different average library size.
+I'll therefore repeat this evaluation with the variance explained by libSize in a PC1~subpopulation+libSize model.)
+
+We next investigated the impact of normalization on the ability to distinguish subpopulations, which was much greater than that of filtering (Supplementary Figure XX).
+For many clusters, scaling tended to reduce both the average silhouette width (XXX) and the proportion of variance in the top principal components explained by subpopulations (Figure 3 and Supplementary Figure XXX), and was associated with a poorer adjusted Rand index (ARI) of the resulting clustering (Figure 3).
+Regressing out covariates systematically gave poorer performance on all metrics.
+sctransform produced the highest cluster average silhouette width (Supplementary Figure XXX) as well as ARI (Figure 3), but not always the highest mean proportion of variance in the top components explained by subpopulations (Figure 3 and Supplementary Figure XXX).
+Interestingly, clusters with the lowest average silhouette width (and which are therefore more difficult to classify) tended to be less well distinguished with sctransform, and instead showed a good difference in non-normalized and/or non-scaled data, suggesting that some procedures emphasize differences more useful to track some subpopulations than others, and hence the possibility that further developments in methods could improve this step.
+
+\subsection*{Feature selection}
+
+A standard clustering pipeline typically involves a step of selecting highly-variable genes, which is complicated by the digital nature and the mean-variance relationship of RNAseq.
+Seurat's earlier approaches involved the use of dispersion estimates standardized for the expression levels, while more recent versions rely of a different measure of variance, again standardized.
+Of note, while adjusting for the mean-variance relationship removes much of the bias towards highly-expressed genes, it is nevertheless plausible that this bias reflects biological relevance and hence is actually helpful in classifying cell types.
+\citep{townes_glmpca_2019}	instead suggested to use deviance, while sctransform provides its own ordering the genes by transformed variance.
+We therefore compared these methods.
+
+Reasoning that a selection method should ideally select genes whose variability is more between subpopulations than within, we first assessed whether each selection method selected genes with a high proportion of variance or deviance explained by (real) subpopulations (the latter estimated through edgeR).
+Estimates of variance explained by subpopulations based on a standard Seurat normalization or on sctransform data were highly correlated (Supplementary Figure XXX A), and there was also a good agreement with deviance explained, although lowly-expressed genes in particular could have a high deviance explained without having much of their variance explained by subpopulations (Supplementary Figure XXX B-D).
+We therefore compared, using both of these estimates, the proportion of the cumulative variance/deviance explained by the top X genes that could be retrieved by each gene ranking method (Supplementary Figures XX and XX).
+We focused on the first 1000 genes to highlight differences between methods, although the differences tended to decrease with a higher number of genes selected.
+A first observation was that standardized measures of variability were systematically worse than their non-standardized  counterparts in selecting genes with a high proportion of variance explained by subpopulations, while instead proving most of the time superior in selecting genes with a high _deviance_ explained by subpopulations (Figure 4 and Supplementary Figure XX).
+Deviance proved the method of choice to maximize the former (with mere expression level proving surprisingly good), but did not perform so well to select genes with a high deviance explained.
+
+We next investigated whether a selection based on these methods impacted the proportion of variance explained by subpopulations in the first principal components (Supplementary Figure XX).
+To validate the previous assay, included a selection based on the genes whose proportion variance or deviance explained was highest.
+Interestingly, while these selections were on average the top-ranking methods, they were not systematically so for all dataset.
+
+These main observations were reflected in the ARI of the resulting clustering (Figure 4B): non-standardized measures of variability, including mere expression level, tended to outperform more complex metrics.
+In general, we found Seurat's unstandardized estimate of variance to provide the best results across all normalization methods, followed by deviance.
+
+We next investigated the impact of the number of features selected.
+
+\subsection*{Dimensionality reduction}
 

### 2019-05-15 supplementary_figures/supp1_dataset_description.Rmd
updated supplementary figures

@@
 \newpage
+
 # Supplementary Figure `r getFigNb(TRUE)`
 
-```{r cellprops_cor, fig.width=7.5, fig.height=7}
+```{r cellprops_cor, fig.width=7, fig.height=8}
 d <- data.frame( dataset=rep(names(SCEs),sapply(SCEs,ncol)), 
+                 featcount_dist=unlist(lapply(SCEs,FUN=function(x) x$featcount_dist)),
+                 log10_counts=unlist(lapply(SCEs,FUN=function(x) x$log10_total_counts)),
                  pct_counts_Mt=unlist(lapply(SCEs,FUN=function(x) x$pct_counts_Mt)),
@@
            labels=LETTERS[1:2], nrow=2 )
-
-ggplot(d, aes(x=log10_features, y=log10_counts, colour=featcount_dist)) + geom_point() + facet_wrap(~dataset, scale="free") + scale_colour_gradient2(limits=c(-0.2,0.2))
 ```

### 2019-05-15 supplementary_figures/supp2_filtering.Rmd
updated supplementary figures

---
title: "Supplementary Figures relative to filtering"
author: "Pierre-Luc Germain"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r, include=FALSE}
if(!exists('FIG_NB')){
  FIG_NB <- 0
  getFigNb <- function(increment=FALSE){
    if(increment) FIG_NB <<- FIG_NB + 1
    FIG_NB
  }
}
```

```{r}
suppressPackageStartupMessages({
  library(ggplot2)
  library(cowplot)
  library(SingleCellExperiment)
  library(reshape2)
  library(ComplexHeatmap)
  library(knitr)
})
suppressMessages(devtools::load_all("../../pipComp/"))
source("../misc_functions.R")
data(datasets)

# we load the datasets
lf <- list.files("../datasets/", full.names=TRUE)
names(lf) <- gsub(".SCE.rds", "", basename(lf), fixed=TRUE)
SCEs <- lapply(lf, readRDS)
```

# Supplementary Figure `r getFigNb(TRUE)`

```{r filt_more_ari, fig.width=8, fig.height=8.5}
res <- readRDS("../data/filtering_old.rds")
res$pcCells <- round(100*res$nbCells/datasets[as.character(res$dataset),"nbCells"],2)
# we restrict to analyses that gave a number of clusters close to reality:
res2 <- res[which(abs(res$nbClusters-datasets[as.character(res$dataset), "subpopulations"])<=1),]
p1 <- ggplot(res2, aes(x=pcCells, y=ARI, colour=norm)) + geom_point() + facet_wrap(~dataset, scale="free")

fres <- readRDS("../../resNew/filtering_intermediateSummary.rds")
pc <- fres$filtering$pcCells
pc <- pc[grep("doublet",row.names(pc),invert=T),]
row.names(pc) <- reduceMethodNames(row.names(pc),2)
ha <- HeatmapAnnotation(dataset=reduceMethodNames(colnames(pc),1))
shortDSNames <- c("mixology10x3cl"="mix3cl", "mixology10x5cl"="mix5cl", "Zhengmix4eq"="Zh4eq", "Zhengmix4uneq"="Zh4uneq", "Zhengmix8eq"="Zh8eq")
colnames(pc) <- paste(shortDSNames[reduceMethodNames(colnames(pc),1)], reduceMethodNames(colnames(pc),2))
p2 <- grid.grabExpr(draw(Heatmap(pc, name="% out", bottom_annotation=ha)))

plot_grid(p1, p2, labels=LETTERS[1:2], nrow=2)
```

### Supplementary Figure `r getFigNb()`
**A:** Adjusted rand index (ARI) of various clustering pipelines using various combinations of filtering criteria. Only clustering analyses with a number of clusters +/-1 from the real number of clusters were considered. In general, filtering more does not result in more accurate clustering. **B:** Percentage of cells filtered out per subpopulation, using various combination of filters (see methods).

\newpage

# Supplementary Figure `r getFigNb(TRUE)`

```{r, fig.height=4}
res$filt.times <- sapply(strsplit(res$filt,"_"), FUN=function(x){ if(length(x)<3) return(1); as.numeric(x[[3]])})
res$filt.MADs <- sapply(strsplit(res$filt,"_"), FUN=function(x){ as.numeric(x[[2]]) })
filt.vars <- strsplit(sapply(strsplit(res$filt,"_"), FUN=function(x) x[[1]]),";")
fv <- unique(unlist(filt.vars))
for(f in fv[grep("both",fv,invert=TRUE)]){
  f2 <- gsub("lower$|higher$","both",f)
  res[[f]] <- sapply(filt.vars, FUN=function(x){ any(c(f,f2) %in% x) })
}
form <- paste0("~nbClusters*dataset+norm+",paste(grep(".",colnames(res),fixed=TRUE,value=TRUE),collapse="+"))
form <- gsub("filt.MADs","factor(filt.MADs)",form,fixed=T)
form <- gsub("filt.times","factor(filt.times)",form,fixed=T)
form <- paste0("ARI",form)
message(strwrap(paste("lm(", form, ")")))
mod <- lm(as.formula(form), data=res)
mod <- coef(summary(mod))[grep(".",row.names(coef(summary(mod))),fixed=T),]
colnames(mod)[4] <- "p value"
kable(mod)
```

### Supplementary Figure `r getFigNb()`
Linear regression analysis of ARI on various filtering criteria, correcting for the effect of the number of clusters. Increasing the number of distributions on which a cell must deviate in order to be excluded (`filt.times`) tended to be positively associated with ARI, while increasing the number of median absolute deviations required to be an outlier (`filt.MADs`) was associated with a decrease in ARI.

\newpage

# Supplementary Figure `r getFigNb(TRUE)`

```{r}
sw <- fres$dimreduction$clust.avg.silwidth[[1]]
sw <- sw[,grep("doublet",colnames(sw))]
w <- grep("seuratvst",colnames(sw))
colnames(sw) <- reduceMethodNames(colnames(sw),2)

chm(list("standard Seurat norm"=sw[,-w], "sctransform"=sw[,w]), scale="row", scaleTitle="z-score", row_ann=data.frame(meanSilWidth=rowMeans(sw), dataset=reduceMethodNames(row.names(sw),1)), show_row_names=FALSE)
```

```{r fig.height=3}
r <- fres$dimreduction$PC.R2[which(fres$dimreduction$PC.R2$dim<=10),]
r <- r[grep("doublet",r$method),]
r <- r[grep("clust",r$method, invert=T),]
r <- aggregate(r$value, by=r[,c(1,3)], FUN=mean)
r <- t(cast2(r, dataset~method, value.var="x", fun.aggregate=mean))
w <- grep("seuratvst", row.names(r))
row.names(r) <- reduceMethodNames(row.names(r), 2)
chm(list("standard Seurat norm"=t(r[-w,]), "sctransform"=t(r[w,])), sameScale=TRUE, 
    scale="row", scaleTitle="z-score", value_format="%.2f")
```

### Supplementary Figure `r getFigNb()`
Impact of the set of filters on the average silhouette width of the subpopulations (top) and on the proportion of variance in the first components explained by clusters (bottom).

\newpage

# Supplementary Figure `r getFigNb(TRUE)`

```{r}
res <- readRDS("../data/selection_old.rds")
res <- res[grep("2000$",res$sel),]
res <- res[grep("deviance|seurat.variance.standardized",res$sel),]
res$filt <- as.factor(res$filt)
levels(res$filt) <- c("coding only", "no mitochondrial", "no ribosomal", "all")
res2 <- res[which(abs(res$nbClusters-datasets[res$dataset,"subpopulations"])<=1),]
ll <- lapply(split(res2, res2$norm), FUN=function(x){
  m2 <- cast2(x, formula=filt~dataset, fun.aggregate=median, value.var="ARI")
})
chm(ll, scale="column", scaleTitle="z-score", sameScale=TRUE, value_format="%.2f", row_sortFn=function(x){ median(x,na.rm=T) }, cluster_columns=FALSE)
```

### Supplementary Figure `r getFigNb()`

Impact of restricting the type of features used on the ARI of the clustering.
\newpage

### 2019-05-15 supplementary_figures/supp3_normalization.Rmd
updated supplementary figures

---
title: "Supplementary Figures relative to normalization"
author: "Pierre-Luc Germain"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r, include=FALSE}
if(!exists('FIG_NB')){
  FIG_NB <- 0
  getFigNb <- function(increment=FALSE){
    if(increment) FIG_NB <<- FIG_NB + 1
    FIG_NB
  }
}
```

```{r}
suppressPackageStartupMessages({
  library(ggplot2)
  library(cowplot)
  library(ComplexHeatmap)
  library(reshape2)
  library(dplyr)
})
suppressMessages(devtools::load_all("../../pipComp/"))
data(datasets)
source("../misc_functions.R")
```

# Supplementary Figure `r getFigNb(TRUE)`

```{r norm_r2, fig.width=8, fig.height=9}
res <- readRDS("../../resNew/norm_intermediateSummary.rds")
r2 <- res$dimreduction$PC.R2
r2$norm <- gsub("norm.","",reduceMethodNames(r2$method,3),fixed=T)
r2 <- r2[which(r2$dim <= 10),]
r2$R.on.Clusters <- sqrt(r2$value)
r3 <- cast2(r2, formula=norm~dataset, value.var="R.on.Clusters")
r2m <- dplyr::summarise(group_by(r2,dataset), meanR=mean(R.on.Clusters))
p1 <- ggplot(r2, aes(norm, R.on.Clusters, fill=norm)) + geom_violin() + facet_wrap(~dataset, nrow=1) + coord_flip() + theme(legend.position="none") + stat_summary(fun.y="mean", geom="point") + geom_hline(data=r2m, aes(yintercept=meanR)) + ggtitle("Proportion of the top dimensions explained by real clusters") + ylab("sqrt(R^2) of dimension ~ real cluster")

pc1.covar <- lapply( res$dimreduction$PC1.covar[2:3], FUN=function(x){
  x <- aggregate(x, by=list(reduceMethodNames(row.names(x), 3)), FUN=median)
  row.names(x) <- x[,1]
  x[,1] <- NULL
  x
})
names(pc1.covar) <- c("Correlation with log10-counts", "Correlation with nb features")
p2 <- grid.grabExpr(draw(chm(pc1.covar, sameScale=TRUE, scaleTitle="cor", value_format="%.2f", row_sortFn=median)))
plot_grid(p2, p1, labels=LETTERS[1:2], nrow=2)
```

### Supplementary Figure `r getFigNb()`
**A:** Proportion of the variance in the first 10 PCA components that is explained by real subpopulations. The square-root of R-squared values is plotted for ease of visualization.
**B:** Correlation of the first component with library size (left) or with the number of detected features (right).

\newpage

# Supplementary Figure `r getFigNb(TRUE)`

```{r norm_silwidth, fig.width=8, fig.height=7}
res <- readRDS("../../resNew/norm_intermediateSummary.rds")
sw <- res$dimreduction$clust.avg.silwidth[[1]]
hma <- HeatmapAnnotation(filtering=reduceMethodNames(colnames(sw),2), col=list(filtering=c(deffilter="#CC6677",lenientfilter="#4477AA",clustFilterStringent="#DDCC77")))
colnames(sw) <- gsub("norm.","",reduceMethodNames(colnames(sw), 3),fixed=T)
chm(list("Cluster mean silhouette width"=sw), scaleTitle="z-score", scale="row", row_ann=data.frame(meanSilWidth=rowMeans(sw), dataset=reduceMethodNames(row.names(sw),1)), show_row_names=FALSE, top_annotation=hma)
```

### Supplementary Figure `r getFigNb()`
Mean silhouette width of true subpopulations across the first 10 components with different normalization and scaling approaches.

\newpage

### 2019-05-15 supplementary_figures/supp4_selection.Rmd
updated supplementary figures

---
title: "Supplementary Figures relative to feature selection"
author: "Pierre-Luc Germain"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r, include=FALSE}
if(!exists('FIG_NB')){
  FIG_NB <- 0
  getFigNb <- function(increment=FALSE){
    if(increment) FIG_NB <<- FIG_NB + 1
    FIG_NB
  }
}
```

```{r}
suppressPackageStartupMessages({
  library(ggplot2)
  library(cowplot)
  library(ComplexHeatmap)
})
suppressMessages(devtools::load_all("../../pipComp/"))
data(datasets)
source("../misc_functions.R")
```

# Supplementary Figure `r getFigNb(TRUE)`

```{r, fig.width=8, fig.height=8}
gi <- readRDS("../data/GI.rds")

layout(matrix(1:4,nrow=2,byrow=TRUE))
LSD::heatscatter(gi$mixology10x3cl$vst.R2, gi$mixology10x3cl$lognorm.R2, xlab="Prop variance explained based, sctransform", ylab="Prop variance explained, standard seurat normalization", main="A. Variance explained by real subpopulations (mixology10x3cl)")

LSD::heatscatter(gi$mixology10x3cl$devianceExplained, gi$mixology10x3cl$vst.R2, xlab="Deviance explained by real subpopulations", ylab="sctransform-based variance explained", , main="B. Variance explained vs deviance explained (mixology10x3cl)")

diffplot <- function(x, title=""){
  LSD::heatscatter(x$log10_mean_counts, x$devianceExplained - x$vst.R2, xlab="log10(mean counts)", ylab="Prop deviance explained - prop variance explained", main=title)
}
diffplot(gi$mixology10x3cl, "C. Difference between the estimates (mixology10x3cl)")
diffplot(gi$Zhengmix4eq, "D. Difference between the estimates (Zhengmix4eq)")
```

### Supplementary Figure `r getFigNb()`
**A:** Comparison of the gene-wise proportion of variance explained by real subpopulations based on Seurat's standard log normalization and on `sctransform` variance-stabilizing transformation. Across 10x datasets, there is a good agreement between the two, the correlation ranging between 0.92 and 0.97.
**B:** There is also a good agreement between _variance_ and _deviance_ explained, with some genes having a higher deviance explained.
**C-D:** Relationship between mean expression and the difference between the proportion of deviance explained and the proportion of variance explained in two datasets. Genes that have a higher proportion of the deviance explained than of the variance explained are generally the lowly-expressed ones.

\newpage

# Supplementary Figure `r getFigNb(TRUE)`

```{r, fig.width=8, fig.height=9, warnings=FALSE, messages=FALSE}
ll <- lapply(names(gi), FUN=function(x){
  cumVE2(gi[[x]], gi[[x]]$vst.R2) + xlim(0,2000) + ggtitle(x) + ylab("Cumulative var explained") + theme(legend.position="none")
})
#ggpubr::ggarrange(plotlist=ll, common.legend = TRUE, ncol=2, nrow=4, labels=LETTERS[1:length(ll)])
ll$legend <- get_legend(ll[[1]] + theme(legend.position="right"))
plot_grid(plotlist=ll, ncol=2)
```

### Supplementary Figure `r getFigNb()`

Proportion of the cumulative _variance_ explained by real subpulations that is retrieved through the selection. For each gene, we compute the proportion of the variance explained by real subpopulations. For each rank X, we sum this proportion for the X genes selected by a given method, and divide it by the sum when selecting the X genes with the highest variance explained. An ideal selection would therefore be a horizontal line at 1.

\newpage

# Supplementary Figure `r getFigNb(TRUE)`

```{r, fig.width=8, fig.height=9, warnings=FALSE, messages=FALSE}
ll <- lapply(names(gi), FUN=function(x){
  cumVE2(gi[[x]], gi[[x]]$devianceExplained) + xlim(0,2000) + ggtitle(x) + ylab("cumul deviance explained") + theme(legend.position="none")
})
ll$legend <- get_legend(ll[[1]] + theme(legend.position="right"))
plot_grid(plotlist=ll, ncol=2)
```

### Supplementary Figure `r getFigNb()`

Proportion of the cumulative _deviance_ explained by real subpulations that is retrieved through the selection. For each gene, we compute the proportion of the variance explained by real subpopulations. As for Supplementary Figure `r getFigNb()-1`, except using deviance explained.

\newpage

# Supplementary Figure `r getFigNb(TRUE)`

```{r, fig.width=8, fig.height=6}
res <- readRDS("../../resNew/sel_intermediateSummary.rds")
r2 <- res$dimreduction$PC.R2
r2 <- r2[grep("1000$",r2$method),]
r2 <- r2[grep("doublet.pl deffilter",r2$method,fixed=T),]
r2$norm <- reduceMethodNames(r2$method,3)
r2$method <- reduceMethodNames(r2$method,4)
r2 <- r2[which(r2$dim <= 5),]
r2$R.on.Clusters <- sqrt(r2$value)
r3 <- lapply(split(r2, r2$norm), FUN=function(x){ 
  cast2(x, formula=method~dataset, value.var="R.on.Clusters")
})
chm(r3[-1], scale="column", scaleTitle="z-score", value_format="%.2f", row_sortFn=median, sameScale=TRUE)
```

### Supplementary Figure `r getFigNb()`

Proportion of the variance in the first 5 principal components that is explained by real subpopulations according to different normalization and selection procedures, selecting 1000 genes in each case.

\newpage

### 2019-05-15 main_figures/figure3_normalization.Rmd
updated main figures

---
title: "Figure 3 normalization"
author: "Pierre-Luc Germain"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    keep_md: true
---

```{r}
suppressPackageStartupMessages({
  library(ggplot2)
  library(cowplot)
  library(ComplexHeatmap)
})
suppressMessages(devtools::load_all("../../pipComp/"))
data(datasets)
source("../misc_functions.R")
theme_set(theme_cowplot(font_size = 11))
```

```{r}
res <- readRDS("../../resNew/norm_intermediateSummary.rds")
```

```{r figure3, fig.width=9, fig.height=5}
r2 <- res$dimreduction$PC.R2
r2$norm <- gsub("norm.","",reduceMethodNames(r2$method,3),fixed=T)
r2 <- r2[which(r2$dim <= 5),]
r2$R.on.Clusters <- sqrt(r2$value)
r3 <- cast2(r2, formula=norm~dataset, value.var="R.on.Clusters")

res.n <- readRDS("../../resNew/norm_endSummary.rds")
res.n <- res.n[which(res.n$norm != "norm.edgeR"),]
res.n$norm <- relevel(factor(res.n$norm), "norm.seurat")
levels(res.n$norm) <- gsub("norm.","",levels(res.n$norm),fixed=T)

m <- cast2(res.n, formula=norm~dataset, fun.aggregate=median, value.var="ARI")
m2 <- cast2(res.n[which(res.n$nbClusters==datasets[res.n$dataset,"subpopulations"]),], formula=norm~dataset, fun.aggregate=median, value.var="ARI")

ll <- list("Var in PCs 1-5 explained"=r3,
           "median ARI"=m,
           "ARI at true number of clusters"=m2)
chm(ll, scale="column", scaleTitle="z-score", value_format="%.2f", 
            sameScale=T, row_sortFn=function(x){ mean(x,na.rm=T) }, cluster_columns=FALSE)
```

### 2019-05-15 main_figures/figure4_selection.Rmd
updated main figures

---
title: "Figure 4 selection"
author: "Pierre-Luc Germain"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    keep_md: true
---

```{r}
suppressPackageStartupMessages({
  library(ggplot2)
  library(cowplot)
  library(ComplexHeatmap)
})
source("../misc_functions.R")
suppressMessages(devtools::load_all("../../pipComp/"))
data(datasets)
theme_set(theme_cowplot(font_size = 12))
```

```{r}
GI <- readRDS("../data/GI.rds")
get_auc <- function(preds,ve.var="vst.R2",n=1000){
    preds2 <- list( deviance=preds$deviance,
                   seurat.disp=preds$seurat.dispersion,
                   seurat.disp.std=preds$seurat.dispersion.scaled,
                   seurat.var=preds$seurat.variance,
                   "seurat.var.std*"=preds$seurat.variance.standardized,
                   sctransform=preds$seurat.res.var,
                   expression=preds$total_counts
                   )
    preds2 <- lapply(preds2, as.numeric)
    vesum <- sum(sort(preds[[ve.var]],decreasing=T)[1:n],na.rm=T)
    sapply(preds2, ve=preds[[ve.var]], FUN=function(x, ve) sum(ve[order(x, decreasing=T)[1:n]],na.rm=T))/vesum
}
# aucs1 <- melt(sapply(GI,FUN=get_auc))
# colnames(aucs1) <- c("method", "dataset", "varExplained")
# p1 <- noXlegWith(ggplot(aucs1, aes(x=method, y=varExplained, fill=method)) + geom_col())
# p1 <- p1 + ylab("Proportion of the explained variance") + facet_wrap(~dataset, scale="free",dir="v")
# plot_grid( p1, p2, labels=LETTERS[1:2], nrow=2)

aucs1 <- sapply(GI,FUN=get_auc)
aucs2 <- sapply(GI,ve.var="devianceExplained", FUN=get_auc)
p1 <- grid.grabExpr(draw(chm(list("Variance explained"=aucs1, "Deviance explained"=aucs2), sameScale=TRUE, scaleTitle="Proportion", cluster_columns=FALSE)))
```

```{r figure4, fig.width=6, fig.height=9}
res <- readRDS("../../resNew/sel_endSummary.rds")
res <- res[grep("devianceExplained|varExp", res$sel, invert=TRUE),]
res$selection <- paste(res$sel, res$selnb)
res2 <- res[which(abs(res$nbClusters-datasets[res$dataset,"subpopulations"])<=0),]
ll <- lapply( split(res2, res2$norm), FUN=function(x){
  cast2(x, formula=selection~dataset, value.var="ARI")
})
names(ll) <- c("scran.noscale","standard norm", "sctransform")
p2 <- grid.grabExpr(draw(chm(ll[-1], sameScale=TRUE, scaleTitle="ARI", cluster_rows=FALSE, cluster_columns=FALSE)))

plot_grid(p1, p2, labels=LETTERS[1:2], nrow=2, rel_heights=1:2)
```

**A:** Estimating different methods of ranking genes based on their ability to capture genes with a high proportion of variance (left) or deviance (right) explained. 
**B: ** Accuracy of clusterings (at the right number of clusters) based on the given clustering methods and number of genes selected.

### 2019-05-15 bmc_article.tex
updated manuscript

@@
 
-On the basis of these observations, we defined three sets of criteria as lenient, default and stringent filtering (see methods), and investigated them further. [...]
+On the basis of these observations, we defined three sets of criteria as lenient, default and stringent filtering (see methods), and investigated them further.
+We found that filtering had a much milder impact than doublet removal (Figure 2), and that more stringent filtering tended to result in higher silhouette width, in a higher proportion of the variance in the first components explained by subpopulations (Supplementary Figure 10), and often in a higher adjusted Rand index (ARI) of the clustering at the right number of clusters (Figure 2).
+However, part of the gain is to be ascribed to the removal of cells from subpopulations (especially in the Zheng datasets) which are difficult to distinguish, and for which more stringent clustering did not improve silhouette width (Supplementary Figure 10A).
+For these reasons, we therefore recommend using a more lenient filter (e.g.
+lenient or default filter described here) or a cluster-based filter to avoids most of the subpopulation bias while excluding strong outliers.
 
@@
 Mitochondrial reads have been associated with cell degradation, and it has been observed that much of scRNAseq clustering is determined by ribosomal genes [REF].
-We therefore considered whether excluding one category of features or the other, or using only protein-coding genes, had impact on the ability to distinguish subpopulations.
+We therefore considered whether excluding one category of features or the other, or using only protein-coding genes, had impact on the ability to distinguish subpopulations (Supplementary Figure 11).
+Removal of ribosomal genes robustly reduced the quality of the clustering, suggesting that they represent real biological differences between subpopulations.
+Removing mitochondrial genes and restricting to protein-coding genes had a very mild impact.
 
@@
 An important motivation of sctransform was the observation that, even after normalization, the first principal components of various datasets tended to correlate with library size, suggesting an imperfect correlation \citep{hafemeister_sctransform_2019}.
-We therefore first assessed to what extent the first principal component still retained a correlation with the library size and number of detected features (Supplementary XX).
+We therefore first assessed to what extent the first principal component still retained a correlation with the library size and number of detected features (Supplementary Figure 12A).
 In the 10X datasets, most methods successfully decreased the correlation with covariates, including the mere scaling of non-normalized data.
@@
 
-We next investigated the impact of normalization on the ability to distinguish subpopulations, which was much greater than that of filtering (Supplementary Figure XX).
-For many clusters, scaling tended to reduce both the average silhouette width (XXX) and the proportion of variance in the top principal components explained by subpopulations (Figure 3 and Supplementary Figure XXX), and was associated with a poorer adjusted Rand index (ARI) of the resulting clustering (Figure 3).
+We next investigated the impact of normalization on the ability to distinguish subpopulations, which was much greater than that of filtering (Supplementary Figure 13).
+For many clusters, scaling tended to reduce both the average silhouette width (Supplementary Figure 12B) and the proportion of variance in the top principal components explained by subpopulations (Figure 3 and Supplementary Figure 13), and was associated with a poorer adjusted Rand index (ARI) of the resulting clustering (Figure 3).
 Regressing out covariates systematically gave poorer performance on all metrics.
-sctransform produced the highest cluster average silhouette width (Supplementary Figure XXX) as well as ARI (Figure 3), but not always the highest mean proportion of variance in the top components explained by subpopulations (Figure 3 and Supplementary Figure XXX).
+sctransform produced the highest cluster average silhouette width (Supplementary Figure 12B) as well as ARI (Figure 3), but not always the highest mean proportion of variance in the top components explained by subpopulations (Figure 3 and Supplementary Figure 13).
 Interestingly, clusters with the lowest average silhouette width (and which are therefore more difficult to classify) tended to be less well distinguished with sctransform, and instead showed a good difference in non-normalized and/or non-scaled data, suggesting that some procedures emphasize differences more useful to track some subpopulations than others, and hence the possibility that further developments in methods could improve this step.
@@
 Reasoning that a selection method should ideally select genes whose variability is more between subpopulations than within, we first assessed whether each selection method selected genes with a high proportion of variance or deviance explained by (real) subpopulations (the latter estimated through edgeR).
-Estimates of variance explained by subpopulations based on a standard Seurat normalization or on sctransform data were highly correlated (Supplementary Figure XXX A), and there was also a good agreement with deviance explained, although lowly-expressed genes in particular could have a high deviance explained without having much of their variance explained by subpopulations (Supplementary Figure XXX B-D).
-We therefore compared, using both of these estimates, the proportion of the cumulative variance/deviance explained by the top X genes that could be retrieved by each gene ranking method (Supplementary Figures XX and XX).
-We focused on the first 1000 genes to highlight differences between methods, although the differences tended to decrease with a higher number of genes selected.
-A first observation was that standardized measures of variability were systematically worse than their non-standardized  counterparts in selecting genes with a high proportion of variance explained by subpopulations, while instead proving most of the time superior in selecting genes with a high _deviance_ explained by subpopulations (Figure 4 and Supplementary Figure XX).
+Estimates of variance explained by subpopulations based on a standard Seurat normalization or on sctransform data were highly correlated (Supplementary Figure 14A), and there was also a good agreement with deviance explained, although lowly-expressed genes in particular could have a high deviance explained without having much of their variance explained by subpopulations (Supplementary Figure 14B-D).
+We therefore compared, using both of these estimates, the proportion of the cumulative variance/deviance explained by the top X genes that could be retrieved by each gene ranking method (Supplementary Figures 15-16).
+We first focused on the first 1000 genes to highlight differences between methods, although the differences tended to decrease with a higher number of genes selected (Figure 4A).
+A first observation was that standardized measures of variability were systematically worse than their non-standardized  counterparts in selecting genes with a high proportion of variance explained by subpopulations, while instead proving most of the time superior in selecting genes with a high _deviance_ explained by subpopulations (Figure 4 and Supplementary Figures 15-16).
 Deviance proved the method of choice to maximize the former (with mere expression level proving surprisingly good), but did not perform so well to select genes with a high deviance explained.
 
-We next investigated whether a selection based on these methods impacted the proportion of variance explained by subpopulations in the first principal components (Supplementary Figure XX).
+We next investigated whether a selection based on these methods impacted the proportion of variance explained by subpopulations in the first principal components (Supplementary Figure 17).
 To validate the previous assay, included a selection based on the genes whose proportion variance or deviance explained was highest.
@@
 
-These main observations were reflected in the ARI of the resulting clustering (Figure 4B): non-standardized measures of variability, including mere expression level, tended to outperform more complex metrics.
-In general, we found Seurat's unstandardized estimate of variance to provide the best results across all normalization methods, followed by deviance.
-
-We next investigated the impact of the number of features selected.
+The main observations were reflected in the ARI of the resulting clustering (Figure 4B): non-standardized measures of variability, including mere expression level, tended to outperform more complex metrics.
+In general, we found deviance and Seurat's unstandardized estimate of variance to provide the best results across datasets and normalization methods.
+Increasing the number of features selected also systematically led to an increase in the accuracy of the clustering.
 
 \subsection*{Dimensionality reduction}
-
-\subsection*{Sub-heading for section}
-Text for this sub-heading \ldots
-\subsubsection*{Sub-sub heading for section}
-Text for this sub-sub-heading \ldots
-\paragraph*{Sub-sub-sub heading for section}
-Text for this sub-sub-sub-heading \ldots
-In this section we examine the growth rate of the mean of $Z_0$, $Z_1$ and $Z_2$. In
-addition, we examine a common modeling assumption and note the
-importance of considering the tails of the extinction time $T_x$ in
-studies of escape dynamics.
-We will first consider the expected resistant population at $vT_x$ for
-some $v>0$, (and temporarily assume $\alpha=0$)
-
-Thus we observe that this expected value is finite for all $v>0$ (also see \cite{koon,khar,zvai,xjon,marg}).
-%\nocite{oreg,schn,pond,smith,marg,hunn,advi,koha,mouse}
 

### 2019-05-15 bmc_article.tex
Updates from Overleaf
updated paper, included figures

@@
 %\RequirePackage{natbib}
-%\RequirePackage[authoryear]{natbib}% uncomment this for author-year bibliography
-%\RequirePackage{hyperref}
+\RequirePackage[authoryear]{natbib}% uncomment this for author-year bibliography
+\usepackage{hyperref}
 \usepackage[utf8]{inputenc} %unicode support
@@
 %\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
-\usepackage[pdf]{graphviz}
+\usepackage{graphicx}
 
@@
 
-\def\includegraphic{}
-\def\includegraphics{}
+%\def\includegraphic{}
+%\def\includegraphics{}
 
@@
 
-Single-cell RNA-sequencing (scRNAseq) and the set of attached analysis methods are evolving fast, and while a number of good comparison/benchmark studies can inform analytical decisions [REFs], they need constant updating and often leave open many details of an analysis.
-Here, we go in more detail on the various steps of analysis leading from an initial count matrix to a cluster assignment, which are critical steps in wide range of applications.
-Weharness datasets of known cell composition [TABLE 1] to investigate the impact of various parameters and variations around a core pipeline on various multilevel evaluation metrics.
+Single-cell RNA-sequencing (scRNAseq) and the set of attached analysis methods are evolving fast, and while a number of good comparison/benchmark studies can inform analytical decisions (\citealp{duoClustering2018}, etc) they need constant updating and often leave open many details of an analysis.
+Here, we go in more detail on the various steps of analysis leading from an initial count matrix to a cluster assignment, which are critical steps in a wide range of applications.
+We harness datasets of known cell composition (Table \ref{tab:table1}) to investigate the impact of various parameters and variations around a core pipeline on various multilevel evaluation metrics.
 
 Although we use some datasets based on other protocols, we focus especially on the 10X datasets and on methods that do not require exogenous control RNA, i.e.
-spike-ins (see Table 1 and Supplementary Figure 1 for a description of the datasets).
-As Seurat [REF] was previously shown to outperform other clustering approaches [REFs], we used the Seurat pipeline as the starting framework to perform an integrated investigation of 1) doublet identification, 2) cell filtering, 3) normalization, 4) feature selection, 5) dimensional reduction, 6) clustering.
+spike-ins (see Table \ref{tab:table1} and Supplementary Figure 1 for a description of the datasets).
+As Seurat [REF] was previously shown to outperform other clustering approaches \cite{duoClustering2018,tianMixology2018}, we used the Seurat pipeline as the starting framework to perform an integrated investigation of 1) doublet identification, 2) cell filtering, 3) normalization, 4) feature selection, 5) dimensional reduction, 6) clustering.
 We compare not only competing approaches, but also more fine-grained parameter variations on common methods.
 Importantly, the success of methods at a certain analytical step might be dependent on choices at other steps.
-Therefore, instead of evaluating each step in isolation, we developed a general R framework for evaluating nested variations on a pipeline, and suggest a multilevel panel of metrics.
+Therefore, instead of evaluating each step in isolation, we developed a general framework for evaluating nested variations on a pipeline, and suggest a multilevel panel of metrics.
 Finally, we evaluate several recent methods and provide concrete recommendations.
@@
 
-We defined a pipeline, in R, as a list of functions executed consecutively on the output of the previous one...
-The basic pipeline used through most of this paper is described in Figure XXX.
-A number of metrics can be considered at each step which can be used to evaluate this and the previous steps.
+We define a pipeline, in R, as a list of functions executed consecutively on the output of the previous one (Figure 0).
+Each step can include various parameters, all of which must be given alternative values.
+The \textit{runPipeline} command will then proceed in a step-wise fashion to computer only the parts of the pipelines that have not already been computed.
+In addition, each step can output evaluation metrics, which together provide a multi-layered evaluation, where variations in a given parameter can be evaluated using all metrics from this point downward in the pipeline.
+This is especially important because end-point metrics, such as the adjusted Rand index (ARI) for clustering, are not perfect.
+For example, by far the most important determinant of the ARI is the number of detected clusters: the farther it is from the actual number of subpopulations, the worse the ARI.
+In this context, one strategy has been to cluster across various resolutions and only consider the results that have the right number of cluster \citep{duoClustering2018}.
+While this has the virtue of making the results comparable, in practice the number of subpopulations is typically unknown, and tools that operate well in this optimal context might not necessarily be best overall.
+Clustering results are also very sensitive, and as we will see they might not always capture improvements in earlier steps of the pipeline.
+
+Another motivation for the present framework is that clustering is relatively fragile to variations in the pipeline.
+We therefore also wanted to capture whether the effect of a given parameter alteration is robust to changes in the rest of the pipeline, or rather specific to a set of other pipeline parameters.
+We therefore proceeded in a step-wise fashion, first testing a large variety of parameters at the early steps of the pipeline along with only a set of mainstream options downstream, then selecting main alternatives and proceeding to a more detailed benchmark of the next step.
+The basic pipeline used through most of this paper is described at the bottom of Figure 0.
+
+\includegraphics[scale=0.8]{figure0}
 
@@
 
-Doublets, defined as two cells sequenced under the same cellular barcode (for instance as a result of being captured in the same droplet), are fairly frequent in scRNAseq datasets, with estimates ranging from 1 to 10\% depending on the platform and cell concentration used \citep{bloom_estimating_2018,kang_multiplexed_2018}.
+Doublets, defined as two cells sequenced under the same cellular barcode (for instance as a result of being captured in the same droplet), are fairly frequent in scRNAseq datasets, with estimates ranging from 1 to 10\% depending on the platform and cell concentration used \citep{bloomEstimating2018,kangMultiplexedDemuxlet2018}.
 While doublets of the same cell type are relatively innocuous for most downstream applications due to their conservation of the relative expression between genes, doublets formed from different cell types or states are likely to be misclassified and could potentially distort downstream analysis.
@@
 A number of methods were developed to identify doublets, in most cases by comparing each cell to artificially-created doublets [REFs].
-We therefore first evaluated the capacity of these methods to detect doublets using the two 10x datasets produced by Tian et al.
-(\citey{tian_mixology_2018}) with cells of different genetic identity, and for which therefore SNP information can be used as ground truth.
+We therefore first evaluated the capacity of these methods to detect doublets using the two 10x datasets produced by \citep{tianMixology2018} with cells of different genetic identity, and for which therefore SNP information can be used as ground truth.
 We tested DoubletFinder [REF] and scran's doubletCells [REF], both of which use similarity to artificial doublets, and scds [REF], which relies on a combination of coexpression and binary classification.
@@
 
-While most methods accurately identified the doublets in the 3 cell lines dataset (mixology10x3cl), the 5 lines dataset (mixology10x5cl) proved more difficult (Figure 1B).
-Our method achieved the best accuracy while running in a very reasonable time (Figure 1).
+While most methods accurately identified the doublets in the 3 cell lines dataset (mixology10x3cl), the 5 lines dataset (mixology10x5cl) proved more difficult (Figure \ref{fig:figure1}B).
+Our method (scDblFinder) achieved the best accuracy while running in a very reasonable time (Figure \ref{fig:figure1}).
 Across datasets, cells called as doublets tended to be classified in the wrong cluster more often than other cells (Figure 1D).
-We therefore tested whether this method also improved the accuracy of the clustering across all benchmark datasets, and found that it consistently did (Figure 2).
+We therefore tested whether this method also improved the accuracy of the clustering across all benchmark datasets, and found that it consistently did (Figure \ref{fig:figure2}).
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\textwidth]{{main_figures/figure1_doublets_files/figure-html/figure1-1.png}}
+    \caption{\textbf{Identification of doublet cells.
+A-B:} Receiver operating characteristic (ROC) curves of the tested doublet detection methods for the mixology10x3l (\textbf{A}) and mixology10x5cl (\textbf{B}) datasets.
+DoubletFinder failed due to an error on the mixology10x3cl dataset.
+\textbf{C:} Running time of the different methods.
+\textbf{D} Cell called as doublets by scDblFinder tend to be more often misclassified, i.e.
+assigned a wrong cluster.
+The corrected percentage of misclassification is the proportion of times a given cell is misclassified across hundreds of clustering pipelines with various parameters, minus the median misclassification rate of the cell's subpopulation.}
+    \label{fig:figure1}
+\end{figure}
 
@@
 
+\begin{figure}
+    \centering
+    \includegraphics[width=\textwidth]{{main_figures/figure2_filtering_files/figure-html/figure2-1.png}}
+    \caption{Proportion of cells filtered out and accuracy of the cluster (under two normalizations) for different sets of filters (see methods).}
+    \label{fig:figure2}
+\end{figure}
+
 On the basis of these observations, we defined three sets of criteria as lenient, default and stringent filtering (see methods), and investigated them further.
-We found that filtering had a much milder impact than doublet removal (Figure 2), and that more stringent filtering tended to result in higher silhouette width, in a higher proportion of the variance in the first components explained by subpopulations (Supplementary Figure 10), and often in a higher adjusted Rand index (ARI) of the clustering at the right number of clusters (Figure 2).
+We found that filtering had a much milder impact than doublet removal (Figure \ref{fig:figure3}), and that more stringent filtering tended to result in higher silhouette width, in a higher proportion of the variance in the first components explained by subpopulations (Supplementary Figure 10), and often in a higher adjusted Rand index (ARI) of the clustering at the right number of clusters (Figure \ref{fig:figure3}).
 However, part of the gain is to be ascribed to the removal of cells from subpopulations (especially in the Zheng datasets) which are difficult to distinguish, and for which more stringent clustering did not improve silhouette width (Supplementary Figure 10A).
@@
 
-An important motivation of sctransform was the observation that, even after normalization, the first principal components of various datasets tended to correlate with library size, suggesting an imperfect correlation \citep{hafemeister_sctransform_2019}.
+An important motivation of sctransform was the observation that, even after normalization, the first principal components of various datasets tended to correlate with library size, suggesting an imperfect correlation \citep{hafemeisterSCtransform2019}.
 We therefore first assessed to what extent the first principal component still retained a correlation with the library size and number of detected features (Supplementary Figure 12A).
@@
 (NOTE: a correlation of PC1 with library size might be caused by the fact that subpopulations have a different average library size.
-I'll therefore repeat this evaluation with the variance explained by libSize in a PC1~subpopulation+libSize model.)
+I'll therefore repeat this evaluation with the variance explained by libSize in a PC1\textasciitilde subpopulation+libSize model.)
 
 We next investigated the impact of normalization on the ability to distinguish subpopulations, which was much greater than that of filtering (Supplementary Figure 13).
-For many clusters, scaling tended to reduce both the average silhouette width (Supplementary Figure 12B) and the proportion of variance in the top principal components explained by subpopulations (Figure 3 and Supplementary Figure 13), and was associated with a poorer adjusted Rand index (ARI) of the resulting clustering (Figure 3).
+For many clusters, scaling tended to reduce both the average silhouette width (Supplementary Figure 12B) and the proportion of variance in the top principal components explained by subpopulations (Figure \ref{fig:figure3} and Supplementary Figure 13), and was associated with a poorer adjusted Rand index (ARI) of the resulting clustering (Figure \ref{fig:figure3}).
 Regressing out covariates systematically gave poorer performance on all metrics.
-sctransform produced the highest cluster average silhouette width (Supplementary Figure 12B) as well as ARI (Figure 3), but not always the highest mean proportion of variance in the top components explained by subpopulations (Figure 3 and Supplementary Figure 13).
+sctransform produced the highest cluster average silhouette width (Supplementary Figure 12B) as well as ARI (Figure \ref{fig:figure3}), but not always the highest mean proportion of variance in the top components explained by subpopulations (Figure \ref{fig:figure3} and Supplementary Figure 13).
 Interestingly, clusters with the lowest average silhouette width (and which are therefore more difficult to classify) tended to be less well distinguished with sctransform, and instead showed a good difference in non-normalized and/or non-scaled data, suggesting that some procedures emphasize differences more useful to track some subpopulations than others, and hence the possibility that further developments in methods could improve this step.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\textwidth]{{main_figures/figure3_normalization_files/figure-html/figure3-1.png}}
+    \caption{\textbf{Evaluation of normalization procedures:} proportion of the variance in the first 5 components explained by the subpopulations, median ARI, and ARI at the true number of clusters.}
+    \label{fig:figure3}
+\end{figure}
 
@@
 Of note, while adjusting for the mean-variance relationship removes much of the bias towards highly-expressed genes, it is nevertheless plausible that this bias reflects biological relevance and hence is actually helpful in classifying cell types.
-\citep{townes_glmpca_2019}	instead suggested to use deviance, while sctransform provides its own ordering the genes by transformed variance.
+\citep{townesGlmpca2019} instead suggested to use deviance, while sctransform provides its own ordering the genes by transformed variance.
 We therefore compared these methods.
@@
 We therefore compared, using both of these estimates, the proportion of the cumulative variance/deviance explained by the top X genes that could be retrieved by each gene ranking method (Supplementary Figures 15-16).
-We first focused on the first 1000 genes to highlight differences between methods, although the differences tended to decrease with a higher number of genes selected (Figure 4A).
-A first observation was that standardized measures of variability were systematically worse than their non-standardized  counterparts in selecting genes with a high proportion of variance explained by subpopulations, while instead proving most of the time superior in selecting genes with a high _deviance_ explained by subpopulations (Figure 4 and Supplementary Figures 15-16).
+We first focused on the first 1000 genes to highlight differences between methods, although the differences tended to decrease with a higher number of genes selected (Figure \ref{fig:figure4}A).
+A first observation was that standardized measures of variability were systematically worse than their non-standardized  counterparts in selecting genes with a high proportion of variance explained by subpopulations, while instead proving most of the time superior in selecting genes with a high deviance explained by subpopulations (Figure \ref{fig:figure4} and Supplementary Figures 15-16).
 Deviance proved the method of choice to maximize the former (with mere expression level proving surprisingly good), but did not perform so well to select genes with a high deviance explained.
+
+\begin{figure}
+    \centering
+    \includegraphics[scale=0.6]{{main_figures/figure4_selection_files/figure-html/figure4-1.png}}
+    \caption{\textbf{Evaluation of feature selection methods.
+A:} Ability of different feature ranking methods to capture genes with a high proportion of variance (left) or deviance (right) explained.
+\texbf{B:} Accuracy of clustering (at the right number of clusters) based on the given clustering methods and number of genes selected.}
+    \label{fig:figure4}
+\end{figure}
 
@@
 \section*{Tables}
+
 \begin{table}[h!]
-\caption{Sample table title. This is where the description of the table should go.}
-      \begin{tabular}{cccc}
+\caption{Description of the benchmark datasets (GSEids to be added)}
+\label{tab:table1}
+\begin{tabular}{rlrrl}
   \hline
-           & B1  &B2   & B3\\ \hline
-        A1 & 0.1 & 0.2 & 0.3\\
-        A2 & ... & ..  & .\\
-        A3 & ..  & .   & .\\ \hline
+dataset & protocol & nbCells & subpops & description \\ 
+  \hline
+Koh & SMARTer & 531 &   9 & FACS purified H7 hESC in different differention stages \\ 
+  Kumar & SMARTer & 246 &   3 & Mouse ESC cultured in different conditions \\ 
+  Zhengmix4eq & 10x & 3994 &   4 & Mixtures of FACS purified PBMCs \\ 
+  Zhengmix4uneq & 10x & 6498 &   4 & Mixtures of FACS purified PBMCs \\ 
+  Zhengmix8eq & 10x & 3994 &   8 & Mixtures of FACS purified PBMCs \\ 
+  mixology10x3cl & 10x & 902 &   3 & Mixture of 3  cancer cell lines from CellBench \\ 
+  mixology10x5cl & 10x & 3918 &   5 & Mixture of 5 cancer cell lines from CellBench \\ 
+   \hline
 \end{tabular}

### 2019-05-15 bmc_article.tex

@@
 
-\title{A sample article title}
+\title{scRNAseq benchmark in want of a title}
 
@@
 \parttitle{Conclusions}
-We offer 
+Text for this section.
 \end{abstract}
@@
 Here, we go in more detail on the various steps of analysis leading from an initial count matrix to a cluster assignment, which are critical steps in a wide range of applications.
-We harness datasets of known cell composition (Table \ref{tab:table1}) to investigate the impact of various parameters and variations around a core pipeline on various multilevel evaluation metrics.
+We harness real datasets of known cell composition (Table \ref{tab:table1}) and develop new evaluation metrics to investigate in a multilevel fashion the impact of various parameters and variations around a core scRNAseq pipeline.
 
-Although we use some datasets based on other protocols, we focus especially on the 10X datasets and on methods that do not require exogenous control RNA, i.e.
+Although we use some datasets based on other protocols, our focus is especially on 10X datasets and on methods that do not include exogenous control RNA, i.e.
 spike-ins (see Table \ref{tab:table1} and Supplementary Figure 1 for a description of the datasets).
-As Seurat [REF] was previously shown to outperform other clustering approaches \cite{duoClustering2018,tianMixology2018}, we used the Seurat pipeline as the starting framework to perform an integrated investigation of 1) doublet identification, 2) cell filtering, 3) normalization, 4) feature selection, 5) dimensional reduction, 6) clustering.
-We compare not only competing approaches, but also more fine-grained parameter variations on common methods.
+As Seurat \cite{satijaSeurat2015} was previously shown to outperform other clustering approaches \cite{duoClustering2018,tianMixology2018}, we used the Seurat pipeline as the starting point to perform an integrated investigation of 1) doublet identification, 2) cell filtering, 3) normalization, 4) feature selection, 5) dimensional reduction, 6) clustering.
+We compare not only competing approaches, but also more fine-grained parameter and variations on common methods.
 Importantly, the success of methods at a certain analytical step might be dependent on choices at other steps.
@@
 
+All analyses were performed through the \url[pipComp]{https://github.com/plger/pipComp} R package, which implements the pipeline framework described here, and all code to reproduce each figure is available in the \url[https://github.com/markrobinsonuzh/scRNA\_pipelines\_paper]{https://github.com/markrobinsonuzh/scRNA_pipelines_paper} repository, which also includes the basic datasets with a standardized annotation.
+
+\subsection*{How the data was obtained}
+(to come)
+
+\subsection*{Software and package versions}
+(to come)
+
+\subsection*{Default pipeline parameters}
+(to come)
+
+\subsection*{Doublet detection method}
+Our new doublet detection method is available at \url{https://github.com/plger/plDoubletFinder}.
+We first cluster the cells using scran's \textit{quickCluster} method, favouring overclustering, and reduce the dataset to the union of 1500 most expressed genes of each cluster.
+We then create 75\% of the artifical doublets by sampling the two cells specifically from different clusters, and 25\% by entirely random sampling in case the clustering was erroneous, and summing their counts.
+We also create meta-cells from each cluster, and use them to create additional doublets and triplets.
+We then combine them with the real cells, convert the counts to scaled column ranks, and build a graph of the \textit{k} (default 10) nearest neighbors using scran.
+We then calculate, for each cell, the proportion of its neighbors that are artificial doublets.
+This ratio serves as a doublet score, which is then thresholded by simultaneously minimizing the error in classifying real vs artificial doublets and the deviation from the distribution of expected doublet rate (accounting for homotypic doublets as done by DoubletFinder).
+For the pipeline comparison, the expected doublet rate was set to 2.1\% (with a 1\% standard deviation) on the basis of the mixology datasets.
+
+\subsection*{Filter sets}
+Lenient, default, and stringent sets of filters...
+
+\subsection*{Variance and deviance explained}
+
+Unless specified otherwise, the variance in gene expression explained by subpopulations was calculated on the data normalized and transfored through sctransform.
+For each gene, we fitted a linear model using the subpopulation as only independent variable (\textasciitilde subpopulation), and used the R-squared as a measure of the variance explained.
+The same method was used for principal components.
+
+The deviance explained by subpopulations was calculated directly on counts using the \textit{pipComp::getDevianceExplained} function.
+The function uses edgeR to fit two models, a full \textasciitilde librarySize+subpopulation model and a reduced \textasciitilde librarySize model.
+For each gene, the deviance explained is then the difference between the deviance of each models, divided by the deviance of the reduced model.
+In the rare cases where this resulted in a negative deviance explained, it was set to zero.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2019-05-17 main_figures/figure3_normalization.Rmd
included norm based on stable genes

@@
 ```{r}
-res <- readRDS("../../resNew/norm_intermediateSummary.rds")
+res <- readRDS("../../resNew/norm2_intermediateSummary.rds")
 ```
@@
 
-res.n <- readRDS("../../resNew/norm_endSummary.rds")
+res.n <- readRDS("../../resNew/norm2_endSummary.rds")
 res.n <- res.n[which(res.n$norm != "norm.edgeR"),]

### 2019-05-18 main_figures/figure2_filtering.Rmd
added missing datasets

@@
 ---
+
+```{r}
+suppressPackageStartupMessages({
+  library(ComplexHeatmap)
+})
+suppressMessages(devtools::load_all("../../pipComp/"))
+source("../misc_functions.R")
+data(datasets)
+```
+
+```{r figure2, fig.width=9, fig.height=4}
+res.f <- readRDS("../../resNew/filtering_endSummary.rds")
+
+res.f$pcCells <- 100-100*res.f$nbCells/datasets[as.character(res.f$dataset),"nbCells"]
+res.f$method <- factor(paste(res.f$doubletmethod, res.f$filt))
+res.f <- res.f[grep("clust",res.f$method, invert=TRUE),]
+levels(res.f$method) <- gsub("nothingFUN ","",levels(res.f$method))
+
+w <- which(abs(res.f$nbClusters-datasets[as.character(res.f$dataset),"subpopulations"])<=1)
+#w <- which(res.f$nbClusters==datasets[as.character(res.f$dataset),"subpopulations"])
+aris.1 <- cast2(res.f[intersect(w, which(res.f$norm=="norm.seurat")),], formula=method~dataset, value.var="ARI")
+aris.2 <- cast2(res.f[intersect(w, which(res.f$norm=="norm.seuratvst")),], formula=method~dataset, value.var="ARI")
+pc <- cast2(res.f, formula=method~dataset, value.var="pcCells")
+chm(list("ARI (standard norm)"=aris.1, "ARI (sctransform)"=aris.2, "% cells out"=pc), scale=c("column","column","none"), value_format=c("%#.2f","%#.2f","%#.1f"), cluster_columns=FALSE, cluster_rows=FALSE, scaleVar=FALSE)
+```

### 2019-05-18 supplementary_figures/supp2_filtering.Rmd
updated supp figures

@@
   library(knitr)
+  library(circlize)
 })
@@
 data(datasets)
-
-# we load the datasets
-lf <- list.files("../datasets/", full.names=TRUE)
-names(lf) <- gsub(".SCE.rds", "", basename(lf), fixed=TRUE)
-SCEs <- lapply(lf, readRDS)
 ```
@@
 
-```{r filt_more_ari, fig.width=8, fig.height=8.5}
+```{r filt_more_ari, fig.width=8, fig.height=6.5}
 res <- readRDS("../data/filtering_old.rds")
@@
 ha <- HeatmapAnnotation(dataset=reduceMethodNames(colnames(pc),1))
-shortDSNames <- c("mixology10x3cl"="mix3cl", "mixology10x5cl"="mix5cl", "Zhengmix4eq"="Zh4eq", "Zhengmix4uneq"="Zh4uneq", "Zhengmix8eq"="Zh8eq")
-colnames(pc) <- paste(shortDSNames[reduceMethodNames(colnames(pc),1)], reduceMethodNames(colnames(pc),2))
-p2 <- grid.grabExpr(draw(Heatmap(pc, name="% out", bottom_annotation=ha)))
+# shortDSNames <- c("mixology10x3cl"="mix3cl", "mixology10x5cl"="mix5cl", "Zhengmix4eq"="Zh4eq", "Zhengmix4uneq"="Zh4uneq", "Zhengmix8eq"="Zh8eq")
+# colnames(pc) <- paste(shortDSNames[reduceMethodNames(colnames(pc),1)], reduceMethodNames(colnames(pc),2))
+# p2 <- grid.grabExpr(draw(Heatmap(pc, name="% out", bottom_annotation=ha)))
 
-plot_grid(p1, p2, labels=LETTERS[1:2], nrow=2)
+pal <- colorRamp2(c(0,1,5,20,max(pc)),c("#0000FFFF", "#B38BFFFF", "#FFFFFFFF", "#FF9E81FF", "#FF0000FF"))
+p2 <- grid.grabExpr(draw(Heatmap(pc, name="% out", bottom_annotation=ha, col=pal, cluster_columns=FALSE, show_column_names=FALSE)))
+
+plot_grid(p1, p2, labels=LETTERS[1:2], nrow=2, rel_heights=c(3,2))
 ```
@@
 
-```{r}
+```{r, fig.height=5}
 sw <- fres$dimreduction$clust.avg.silwidth[[1]]
@@
 colnames(sw) <- reduceMethodNames(colnames(sw),2)
-
-chm(list("standard Seurat norm"=sw[,-w], "sctransform"=sw[,w]), scale="row", scaleTitle="z-score", row_ann=data.frame(meanSilWidth=rowMeans(sw), dataset=reduceMethodNames(row.names(sw),1)), show_row_names=FALSE)
+sw <- sw[order(reduceMethodNames(row.names(sw),1), rowMeans(sw)),]
+ra <- data.frame(meanSilWidth=rowMeans(sw), dataset=reduceMethodNames(row.names(sw),1))
+chm(list("standard Seurat norm"=sw[,-w], "sctransform"=sw[,w]), scale="row", scaleVar=FALSE, scaleTitle="meanSilWidth\ndifference", sameScale=TRUE, row_ann=ra, show_row_names=FALSE, cluster_rows=FALSE)
 ```
 
-```{r fig.height=3}
+```{r fig.height=3.5, fig.width=6}
 r <- fres$dimreduction$PC.R2[which(fres$dimreduction$PC.R2$dim<=10),]
@@
 res <- readRDS("../data/selection_old.rds")
-res <- res[grep("2000$",res$sel),]
 res <- res[grep("deviance|seurat.variance.standardized",res$sel),]

### 2019-05-18 supplementary_figures/supp3_normalization.Rmd
updated supp figures

@@
 ```{r norm_r2, fig.width=8, fig.height=9}
-res <- readRDS("../../resNew/norm_intermediateSummary.rds")
+res <- readRDS("../../resNew/norm2_intermediateSummary.rds")
 r2 <- res$dimreduction$PC.R2
@@
 r2m <- dplyr::summarise(group_by(r2,dataset), meanR=mean(R.on.Clusters))
-p1 <- ggplot(r2, aes(norm, R.on.Clusters, fill=norm)) + geom_violin() + facet_wrap(~dataset, nrow=1) + coord_flip() + theme(legend.position="none") + stat_summary(fun.y="mean", geom="point") + geom_hline(data=r2m, aes(yintercept=meanR)) + ggtitle("Proportion of the top dimensions explained by real clusters") + ylab("sqrt(R^2) of dimension ~ real cluster")
+p1 <- ggplot(r2, aes(norm, R.on.Clusters, fill=norm)) + geom_violin() + facet_wrap(~dataset, nrow=1) + coord_flip() + theme(legend.position="none") + stat_summary(fun.y="mean", geom="point") + geom_hline(data=r2m, aes(yintercept=meanR)) + ggtitle("Proportion of the top dimensions explained by real clusters") + ylab("sqrt(R^2) of dimension ~ real subpopulation")
 
-pc1.covar <- lapply( res$dimreduction$PC1.covar[2:3], FUN=function(x){
+pc1.covar <- lapply( res$dimreduction$PC1.covarR[2:3], FUN=function(x){
   x <- aggregate(x, by=list(reduceMethodNames(row.names(x), 3)), FUN=median)
@@
 ### Supplementary Figure `r getFigNb()`
-**A:** Proportion of the variance in the first 10 PCA components that is explained by real subpopulations. The square-root of R-squared values is plotted for ease of visualization.
-**B:** Correlation of the first component with library size (left) or with the number of detected features (right).
+**A:** Correlation between library size (left) or the number of detected features (right) and the residuals of a linear regression of the first principal component on the subpopulation.
+**B:** Proportion of the variance in the first 10 PCA components that is explained by real subpopulations. The square-root of R-squared values is plotted for ease of visualization.
 
@@
 ```{r norm_silwidth, fig.width=8, fig.height=7}
-res <- readRDS("../../resNew/norm_intermediateSummary.rds")
+res <- readRDS("../../resNew/norm2_intermediateSummary.rds")
 sw <- res$dimreduction$clust.avg.silwidth[[1]]
-hma <- HeatmapAnnotation(filtering=reduceMethodNames(colnames(sw),2), col=list(filtering=c(deffilter="#CC6677",lenientfilter="#4477AA",clustFilterStringent="#DDCC77")))
 colnames(sw) <- gsub("norm.","",reduceMethodNames(colnames(sw), 3),fixed=T)
-chm(list("Cluster mean silhouette width"=sw), scaleTitle="z-score", scale="row", row_ann=data.frame(meanSilWidth=rowMeans(sw), dataset=reduceMethodNames(row.names(sw),1)), show_row_names=FALSE, top_annotation=hma)
+sw <- sw[order(reduceMethodNames(row.names(sw),1), rowMeans(sw)),]
+ra <- data.frame(meanSilWidth=rowMeans(sw), dataset=reduceMethodNames(row.names(sw),1))
+chm(list("Cluster mean silhouette width"=sw), scaleTitle="row z-score", scale="row", row_ann=ra, show_row_names=FALSE, col_sortFn=mean, cluster_rows=FALSE)
 ```

### 2019-05-18 supplementary_figures/supp_figures.tex
updated supp figures

@@
 \bibliographystyle{plainnat}
+\usepackage{longtable,booktabs}
 \usepackage{graphicx,grffile}
@@
   \postdate{\par}
-    \date{12 Mai, 2019}
+    \date{18 Mai, 2019}
 
@@
 
-\vfill
-
 \hypertarget{supplementary-figure-1-1}{%
@@
 \includegraphics{supp_figures_files/figure-latex/mixology_doublet_featcount-1.pdf}
-
-\vfill
 
@@
 
-\vfill
-
 \hypertarget{supplementary-figure-3-1}{%
@@
 
-\includegraphics{supp_figures_files/figure-latex/misclass-1.pdf}
-\vspace{1cm}
-\includegraphics{supp_figures_files/figure-latex/unnamed-chunk-4-1.pdf}
-
-\vfill
+\includegraphics{supp_figures_files/figure-latex/cellprops_cor-1.pdf}
 
@@
 \subsubsection{Supplementary Figure 4}\label{supplementary-figure-4-1}}
+
+There is a tight relationship, in 10x datasets (i.e.~not the
+\texttt{Koh} and \texttt{Kumar} datasets), between the total counts of a
+cell and its number of detected features. We therefore include, among
+control variables, deviation from this ratio.
+
+\newpage
+
+\hypertarget{supplementary-figure-5}{%
+\section{Supplementary Figure 5}\label{supplementary-figure-5}}
+
+\includegraphics{supp_figures_files/figure-latex/misclass-1.pdf} \vfill
+\includegraphics{supp_figures_files/figure-latex/unnamed-chunk-7-1.pdf}
+
+\hypertarget{supplementary-figure-5-1}{%
+\subsubsection{Supplementary Figure 5}\label{supplementary-figure-5-1}}
 
@@
 
-\hypertarget{supplementary-figure-5}{%
-\section{Supplementary Figure 5}\label{supplementary-figure-5}}
+\hypertarget{supplementary-figure-6}{%
+\section{Supplementary Figure 6}\label{supplementary-figure-6}}
 
-\includegraphics{supp_figures_files/figure-latex/unnamed-chunk-5-1.pdf}
-\vspace{1cm}
-\includegraphics{supp_figures_files/figure-latex/unnamed-chunk-6-1.pdf}
+\includegraphics{supp_figures_files/figure-latex/unnamed-chunk-8-1.pdf}
+\vfill
+\includegraphics{supp_figures_files/figure-latex/unnamed-chunk-9-1.pdf}
 
-\vfill
-
-\hypertarget{supplementary-figure-5-1}{%
-\subsubsection{Supplementary Figure 5}\label{supplementary-figure-5-1}}
+\hypertarget{supplementary-figure-6-1}{%
+\subsubsection{Supplementary Figure 6}\label{supplementary-figure-6-1}}
 
@@
 cluster mis-assignment for the Zheng equal (A) or unequal (B) mixtures
-of four cell types. See Supplementary Figure 4 for more information. The
+of four cell types. See Supplementary Figure 5 for more information. The
 only clear pattern is that cells with a high number of reads or features
@@
 
-\hypertarget{supplementary-figure-6}{%
-\section{Supplementary Figure 6}\label{supplementary-figure-6}}
+\hypertarget{supplementary-figure-7}{%
+\section{Supplementary Figure 7}\label{supplementary-figure-7}}
 
-\includegraphics{supp_figures_files/figure-latex/unnamed-chunk-7-1.pdf}
+\includegraphics{supp_figures_files/figure-latex/unnamed-chunk-10-1.pdf}
 
-\vfill
-
-\hypertarget{supplementary-figure-6-1}{%
-\subsubsection{Supplementary Figure 6}\label{supplementary-figure-6-1}}
+\hypertarget{supplementary-figure-7-1}{%
+\subsubsection{Supplementary Figure 7}\label{supplementary-figure-7-1}}
 
@@
 cluster mis-assignment for the Zheng mixture of 8 cell types. See
-Supplementary Figure 4 for more information.
+Supplementary Figure 5 for more information.
+
+\newpage
+
+\hypertarget{supplementary-figure-8}{%
+\section{Supplementary Figure 8}\label{supplementary-figure-8}}
+
+\includegraphics{supp_figures_files/figure-latex/filt_more_ari-1.pdf}
+
+\hypertarget{supplementary-figure-8-1}{%
+\subsubsection{Supplementary Figure 8}\label{supplementary-figure-8-1}}
+
+\textbf{A:} Adjusted rand index (ARI) of various clustering pipelines
+using various combinations of filtering criteria. Only clustering
+analyses with a number of clusters +/-1 from the real number of clusters
+were considered. In general, filtering more does not result in more
+accurate clustering. \textbf{B:} Percentage of cells filtered out per
+subpopulation, using various combination of filters (see methods).
+
+\newpage
+
+\hypertarget{supplementary-figure-9}{%
+\section{Supplementary Figure 9}\label{supplementary-figure-9}}
+
+\begin{verbatim}
+## lm(ARI~nbClusters*dataset+norm+factor(filt.times)+factor(filt.MADs)+counts.lower+counts.higher+mt.higher+feat.lower+feat.higher+top50.lower+top50.higher+ratiodist.lower+ratiodist.higher)
+\end{verbatim}
+
+\begin{longtable}[]{@{}lrrrr@{}}
+\toprule
+& Estimate & Std. Error & t value & p value\tabularnewline
+\midrule
+\endhead
+normnorm.seuratvst & 0.0069338 & 0.0003104 & 22.3415248 &
+0.0000000\tabularnewline
+factor(filt.times)2 & 0.0086024 & 0.0003617 & 23.7847962 &
+0.0000000\tabularnewline
+factor(filt.times)3 & 0.0168470 & 0.0003811 & 44.2112922 &
+0.0000000\tabularnewline
+factor(filt.MADs)2.5 & -0.0040702 & 0.0004199 & -9.6944075 &
+0.0000000\tabularnewline
+factor(filt.MADs)3 & -0.0054701 & 0.0004202 & -13.0177667 &
+0.0000000\tabularnewline
+factor(filt.MADs)5 & -0.0025099 & 0.0004205 & -5.9684908 &
+0.0000000\tabularnewline
+counts.lowerTRUE & -0.0001193 & 0.0003534 & -0.3375833 &
+0.7356780\tabularnewline
+counts.higherTRUE & -0.0073031 & 0.0003552 & -20.5607582 &
+0.0000000\tabularnewline
+mt.higherTRUE & 0.0006318 & 0.0007765 & 0.8136392 &
+0.4158539\tabularnewline
+feat.lowerTRUE & 0.0002074 & 0.0003494 & 0.5934348 &
+0.5528917\tabularnewline
+feat.higherTRUE & -0.0058113 & 0.0003506 & -16.5729653 &
+0.0000000\tabularnewline
+top50.lowerTRUE & -0.0000717 & 0.0003045 & -0.2354817 &
+0.8138354\tabularnewline
+top50.higherTRUE & -0.0000717 & 0.0003045 & -0.2354817 &
+0.8138354\tabularnewline
+ratiodist.lowerTRUE & -0.0009712 & 0.0003045 & -3.1895182 &
+0.0014256\tabularnewline
+ratiodist.higherTRUE & 0.0012337 & 0.0003045 & 4.0515679 &
+0.0000509\tabularnewline
+\bottomrule
+\end{longtable}
+
+\hypertarget{supplementary-figure-9-1}{%
+\subsubsection{Supplementary Figure 9}\label{supplementary-figure-9-1}}
+
+Linear regression analysis of ARI on various filtering criteria,
+correcting for the effect of the number of clusters. Increasing the
+number of distributions on which a cell must deviate in order to be
+excluded (\texttt{filt.times}) tended to be positively associated with
+ARI, while increasing the number of median absolute deviations required
+to be an outlier (\texttt{filt.MADs}) was associated with a decrease in
+ARI.
+
+\newpage
+
+\hypertarget{supplementary-figure-10}{%
+\section{Supplementary Figure 10}\label{supplementary-figure-10}}
+
+\includegraphics{supp_figures_files/figure-latex/unnamed-chunk-14-1.pdf}
+
+\includegraphics{supp_figures_files/figure-latex/unnamed-chunk-15-1.pdf}
+
+\hypertarget{supplementary-figure-10-1}{%
+\subsubsection{Supplementary Figure
+10}\label{supplementary-figure-10-1}}
+
+Impact of the set of filters on the average silhouette width of the
+subpopulations (top) and on the proportion of variance in the first
+components explained by clusters (bottom).
+
+\newpage
+
+\hypertarget{supplementary-figure-11}{%
+\section{Supplementary Figure 11}\label{supplementary-figure-11}}
+
+\includegraphics{supp_figures_files/figure-latex/unnamed-chunk-16-1.pdf}
+
+\hypertarget{supplementary-figure-11-1}{%
+\subsubsection{Supplementary Figure
+11}\label{supplementary-figure-11-1}}
+
+Impact of restricting the type of features used on the ARI of the
+clustering. \newpage
+
+\hypertarget{supplementary-figure-12}{%
+\section{Supplementary Figure 12}\label{supplementary-figure-12}}
+
+\includegraphics{supp_figures_files/figure-latex/norm_r2-1.pdf}
+
+\hypertarget{supplementary-figure-12-1}{%
+\subsubsection{Supplementary Figure
+12}\label{supplementary-figure-12-1}}
+
+\textbf{A:} Correlation between library size (left) or the number of
+detected features (right) and the residuals of a linear regression of
+the first principal component on the subpopulation. \textbf{B:}
+Proportion of the variance in the first 10 PCA components that is
+explained by real subpopulations. The square-root of R-squared values is
+plotted for ease of visualization.
+
+\newpage
+
+\hypertarget{supplementary-figure-13}{%
+\section{Supplementary Figure 13}\label{supplementary-figure-13}}
+
+\includegraphics{supp_figures_files/figure-latex/norm_silwidth-1.pdf}
+
+\hypertarget{supplementary-figure-13-1}{%
+\subsubsection{Supplementary Figure
+13}\label{supplementary-figure-13-1}}
+
+Mean silhouette width of true subpopulations across the first 10
+components with different normalization and scaling approaches.
+
+\newpage
+
+\hypertarget{supplementary-figure-14}{%
+\section{Supplementary Figure 14}\label{supplementary-figure-14}}
+
+\includegraphics{supp_figures_files/figure-latex/unnamed-chunk-21-1.pdf}
+
+\hypertarget{supplementary-figure-14-1}{%
+\subsubsection{Supplementary Figure
+14}\label{supplementary-figure-14-1}}
+
+\textbf{A:} Comparison of the gene-wise proportion of variance explained
+by real subpopulations based on Seurat's standard log normalization and
+on \texttt{sctransform} variance-stabilizing transformation. Across 10x
+datasets, there is a good agreement between the two, the correlation
+ranging between 0.92 and 0.97. \textbf{B:} There is also a good
+agreement between \emph{variance} and \emph{deviance} explained, with
+some genes having a higher deviance explained. \textbf{C-D:}
+Relationship between mean expression and the difference between the
+proportion of deviance explained and the proportion of variance
+explained in two datasets. Genes that have a higher proportion of the
+deviance explained than of the variance explained are generally the
+lowly-expressed ones.
+
+\newpage
+
+\hypertarget{supplementary-figure-15}{%
+\section{Supplementary Figure 15}\label{supplementary-figure-15}}
+
+\begin{verbatim}
+## Warning: Removed 328867 rows containing missing values (geom_path).
+
+## Warning: Removed 328867 rows containing missing values (geom_path).
+\end{verbatim}
+
+\begin{verbatim}
+## Warning: Removed 302113 rows containing missing values (geom_path).
+\end{verbatim}
+
+\begin{verbatim}
+## Warning: Removed 101276 rows containing missing values (geom_path).
+\end{verbatim}
+
+\begin{verbatim}
+## Warning: Removed 68502 rows containing missing values (geom_path).
+\end{verbatim}
+
+\begin{verbatim}
+## Warning: Removed 94976 rows containing missing values (geom_path).
+\end{verbatim}
+
+\begin{verbatim}
+## Warning: Removed 101101 rows containing missing values (geom_path).
+\end{verbatim}
+
+\begin{verbatim}
+## Warning: Removed 96012 rows containing missing values (geom_path).
+\end{verbatim}
+
+\includegraphics{supp_figures_files/figure-latex/unnamed-chunk-22-1.pdf}
+
+\hypertarget{supplementary-figure-15-1}{%
+\subsubsection{Supplementary Figure
+15}\label{supplementary-figure-15-1}}
+
+Proportion of the cumulative \emph{variance} explained by real
+subpulations that is retrieved through the selection. For each gene, we
+compute the proportion of the variance explained by real subpopulations.
+For each rank X, we sum this proportion for the X genes selected by a
+given method, and divide it by the sum when selecting the X genes with
+the highest variance explained. An ideal selection would therefore be a
+horizontal line at 1.
+
+\newpage
+
+\hypertarget{supplementary-figure-16}{%
+\section{Supplementary Figure 16}\label{supplementary-figure-16}}
+
+\begin{verbatim}
+## Warning: Removed 328867 rows containing missing values (geom_path).
+
+## Warning: Removed 328867 rows containing missing values (geom_path).
+\end{verbatim}
+
+\begin{verbatim}
+## Warning: Removed 302113 rows containing missing values (geom_path).
+\end{verbatim}
+
+\begin{verbatim}
+## Warning: Removed 101276 rows containing missing values (geom_path).
+\end{verbatim}
+
+\begin{verbatim}
+## Warning: Removed 68502 rows containing missing values (geom_path).
+\end{verbatim}
+
+\begin{verbatim}
+## Warning: Removed 94976 rows containing missing values (geom_path).
+\end{verbatim}
+
+\begin{verbatim}
+## Warning: Removed 101101 rows containing missing values (geom_path).
+\end{verbatim}
+
+\begin{verbatim}
+## Warning: Removed 96012 rows containing missing values (geom_path).
+\end{verbatim}
+
+\includegraphics{supp_figures_files/figure-latex/unnamed-chunk-23-1.pdf}
+
+\hypertarget{supplementary-figure-16-1}{%
+\subsubsection{Supplementary Figure
+16}\label{supplementary-figure-16-1}}
+
+Proportion of the cumulative \emph{deviance} explained by real
+subpulations that is retrieved through the selection. For each gene, we
+compute the proportion of the variance explained by real subpopulations.
+As for Supplementary Figure 15, except using deviance explained.
+
+\newpage
+
+\hypertarget{supplementary-figure-17}{%
+\section{Supplementary Figure 17}\label{supplementary-figure-17}}
+
+\includegraphics{supp_figures_files/figure-latex/unnamed-chunk-24-1.pdf}
+
+\hypertarget{supplementary-figure-17-1}{%
+\subsubsection{Supplementary Figure
+17}\label{supplementary-figure-17-1}}
+
+Proportion of the variance in the first 5 principal components that is
+explained by real subpopulations according to different normalization
+and selection procedures, selecting 1000 genes in each case.
 

### 2019-05-19 bmc_article.tex

@@
 \usepackage{graphicx}
+\usepackage{url}
 
@@
 
-\subsubsection*{Excluded more cells is not necessarily good}
+\subsubsection*{Excluding more cells is not necessarily good}
 
@@
 Moreover, stringent filtering tended to be biased towards the exclusion of specific subpopulations (Supplementary Figure 8B).
-Indeed, applying the same filtering criteria on individual clusters of cells (identified through scran's `quickCluster` method) resulted in nearly no cell being filtered out, which suggests that filtering on the global population tends to discard cells of subpopulations with more extreme properties, rather than `bad` cells.
+Indeed, applying the same filtering criteria on individual clusters of cells (identified through scran's \textit{quickCluster} method) resulted in nearly no cell being filtered out, which suggests that filtering on the global population tends to discard cells of subpopulations with more extreme properties, rather than `bad` cells.
 
@@
 sctransform produced the highest cluster average silhouette width (Supplementary Figure 12B) as well as ARI (Figure \ref{fig:figure3}), but not always the highest mean proportion of variance in the top components explained by subpopulations (Figure \ref{fig:figure3} and Supplementary Figure 13).
+Of note, even though sctransform is meant to be applied to data with unique molecular identifiers (UMI), it also performed fairly well with smartseq datasets.
 Interestingly, clusters with the lowest average silhouette width (and which are therefore more difficult to classify) tended to be less well distinguished with sctransform, and instead showed a good difference in non-normalized and/or non-scaled data, suggesting that some procedures emphasize differences more useful to track some subpopulations than others, and hence the possibility that further developments in methods could improve this step.
@@
 
-All analyses were performed through the \url[pipComp]{https://github.com/plger/pipComp} R package, which implements the pipeline framework described here, and all code to reproduce each figure is available in the \url[https://github.com/markrobinsonuzh/scRNA\_pipelines\_paper]{https://github.com/markrobinsonuzh/scRNA_pipelines_paper} repository, which also includes the basic datasets with a standardized annotation.
+\subsection*{Code and data availability}
 
-\subsection*{How the data was obtained}
-(to come)
+All analyses were performed through the \href{pipComp}{https://github.com/plger/pipComp} R package, which implements the pipeline framework described here.
+All code to reproduce each figure is available in the \url{https://github.com/markrobinsonuzh/scRNA\_pipelines\_paper} repository, which also includes the basic datasets with a standardized annotation.
+
+The gene counts for the two mixology datasets were downloaded from the  \href[CellBench repository]{https://github.com/LuyiTian/CellBench\_data/data/sincell\_with\_class.RData}, commit 74fe79e.
+For the other datasets, we used the unfiltered counts from \citep{duoClustering2018}, available on the corresponding repository \url{https://github.com/markrobinsonuzh/scRNAseq\_clustering\_comparison}.
+For simplicity, all starting \textit{SingleCellExperiment} objects with standardized metadata are available on \url{https://github.com/markrobinsonuzh/scRNA\_pipelines\_paper}.
 
 \subsection*{Software and package versions}
-(to come)
+Analyses were performed in R 3.6.0 (bioconductor 3.9), and the following packages were installed from github repositories: Seurat (version 3.0.0), sctransform (0.2.0), DoubletFinder (2.0.1), scds (1.0.0), pipComp (0.2).
+The code for the glmPCA was obtained from \url{https://github.com/willtownes/scrna2019} (commit 1ddcc30ebb95d083a685f12fe81d35dd1b0cb1b2).
 
@@
 \subsection*{Doublet detection method}
-Our new doublet detection method is available at \url{https://github.com/plger/plDoubletFinder}.
-We first cluster the cells using scran's \textit{quickCluster} method, favouring overclustering, and reduce the dataset to the union of 1500 most expressed genes of each cluster.
-We then create 75\% of the artifical doublets by sampling the two cells specifically from different clusters, and 25\% by entirely random sampling in case the clustering was erroneous, and summing their counts.
+Our doublet detection method is available at \url{https://github.com/plger/plDoubletFinder}.
+Briefly, we first cluster the cells using scran's \textit{quickCluster} method, favouring overclustering (default minimum size of 20), and reduce the dataset to the union of 1500 most expressed genes of each cluster.
+We then create 75\% of the artifical doublets by sampling the two cells specifically from different clusters, and 25\% by entirely random sampling (in case the clustering was erroneous), and summing the counts of each pair of cells.
 We also create meta-cells from each cluster, and use them to create additional doublets and triplets.
@@
 This ratio serves as a doublet score, which is then thresholded by simultaneously minimizing the error in classifying real vs artificial doublets and the deviation from the distribution of expected doublet rate (accounting for homotypic doublets as done by DoubletFinder).
-For the pipeline comparison, the expected doublet rate was set to 2.1\% (with a 1\% standard deviation) on the basis of the mixology datasets.
+For the pipeline comparison here, the expected doublet rate was set to 2.1\% (with a 1\% standard deviation) on the basis of the mixology datasets.
 
@@
 
-Unless specified otherwise, the variance in gene expression explained by subpopulations was calculated on the data normalized and transfored through sctransform.
+Unless specified otherwise, the variance in gene expression explained by subpopulations was calculated on the data normalized and transformed through sctransform.
 For each gene, we fitted a linear model using the subpopulation as only independent variable (\textasciitilde subpopulation), and used the R-squared as a measure of the variance explained.
@@
 In the rare cases where this resulted in a negative deviance explained, it was set to zero.
+
+To estimate the correlation between the first principal components and covariates such as library size, we first fitted a linear model on the subpopulations, and correlated the residuals of this model with the covariate of interest.
 
@@
 \begin{table}[h!]
-\caption{Description of the benchmark datasets (GSEids to be added)}
+\caption{Overview of the benchmark datasets}
 \label{tab:table1}
@@
   \hline
-dataset & protocol & nbCells & subpops & description \\ 
+dataset & source & protocol & nbCells & subpops & description \\ 
   \hline
-Koh & SMARTer & 531 &   9 & FACS purified H7 hESC in different differention stages \\ 
-  Kumar & SMARTer & 246 &   3 & Mouse ESC cultured in different conditions \\ 
-  Zhengmix4eq & 10x & 3994 &   4 & Mixtures of FACS purified PBMCs \\ 
-  Zhengmix4uneq & 10x & 6498 &   4 & Mixtures of FACS purified PBMCs \\ 
-  Zhengmix8eq & 10x & 3994 &   8 & Mixtures of FACS purified PBMCs \\ 
-  mixology10x3cl & 10x & 902 &   3 & Mixture of 3  cancer cell lines from CellBench \\ 
-  mixology10x5cl & 10x & 3918 &   5 & Mixture of 5 cancer cell lines from CellBench \\ 
+Koh & GSE85066 & SMARTer & 531 &   9 & FACS purified H7 hESC in different differention stages \\ 
+  Kumar & GSE60749 & SMARTer & 246 &   3 & Mouse ESC cultured in different conditions \\ 
+  Zhengmix4eq & \citep{duoClustering2018} & 10x & 3994 &   4 & Mixtures of FACS purified PBMCs \\ 
+  Zhengmix4uneq & \citep{duoClustering2018} & 10x & 6498 &   4 & Mixtures of FACS purified PBMCs \\ 
+  Zhengmix8eq & \citep{duoClustering2018} & 10x & 3994 &   8 & Mixtures of FACS purified PBMCs \\ 
+  mixology10x3cl & \cite{tianMixology2018} & 10x & 902 &   3 & Mixture of 3  cancer cell lines from CellBench \\ 
+  mixology10x5cl & \cite{tianMixology2018} & 10x & 3918 &   5 & Mixture of 5 cancer cell lines from CellBench \\ 
    \hline

### 2019-05-19 main_figures/figure4_selection.Rmd
included combined selection in panel A

@@
   library(ComplexHeatmap)
+  library(matrixStats)
 })
@@
 get_auc <- function(preds,ve.var="vst.R2",n=1000){
+    preds$vst.Gini <- -1*preds$vst.Gini
+    p2 <- as.matrix(preds[,c("seurat.variance", "seurat.variance.standardized", "seurat.res.var", "deviance")])
     preds2 <- list( deviance=preds$deviance,
@@
                    sctransform=preds$seurat.res.var,
+                   rankMean=colMeans(colRanks(p2)),
+                   rankMax=colMaxs(colRanks(p2)),
                    expression=preds$total_counts
@@
 
-```{r figure4, fig.width=6, fig.height=9}
+```{r figure4, fig.width=6, fig.height=10}
 res <- readRDS("../../resNew/sel_endSummary.rds")
@@
 res$selection <- paste(res$sel, res$selnb)
-res2 <- res[which(abs(res$nbClusters-datasets[res$dataset,"subpopulations"])<=0),]
+nbClusDiff <- abs(res$nbClusters-datasets[res$dataset,"subpopulations"])
+#res2 <- res[which(nbClusDiff==0),]
+res2 <- res[which(nbClusDiff==0 | (nbClusDiff<=1 & res$dataset=="mixology10x3cl")),]
 ll <- lapply( split(res2, res2$norm), FUN=function(x){
@@
 
-plot_grid(p1, p2, labels=LETTERS[1:2], nrow=2, rel_heights=1:2)
+plot_grid(p1, p2, labels=LETTERS[1:2], nrow=2, rel_heights=c(6,10))
 ```
@@
 **A:** Estimating different methods of ranking genes based on their ability to capture genes with a high proportion of variance (left) or deviance (right) explained.
+rankMean and rankMax refer to, respectively, the mean rank and the maximum rank in the following metrics: variance, standardized variance, sctransform, deviance.
 **B: ** Accuracy of clusterings (at the right number of clusters) based on the given clustering methods and number of genes selected.

### 2019-05-20 supplementary_figures/supp3_normalization.Rmd
...

@@
 sw <- res$dimreduction$clust.avg.silwidth[[1]]
-colnames(sw) <- gsub("norm.","",reduceMethodNames(colnames(sw), 3),fixed=T)
-sw <- sw[order(reduceMethodNames(row.names(sw),1), rowMeans(sw)),]
+colnames(sw) <- gsub("norm.","",reduceMethodNames(colnames(sw), 2:3),fixed=T)
+filter <- factor(reduceMethodNames(colnames(sw),1), levels=c("lenientfilter", "clustFilterStringent", "deffilter", "stringentfilter"))
+levels(filter) <- c("lenient", "stringent cluster-wise", "default", "stringent")
+o <- order(reduceMethodNames(colnames(sw),2), filter)
+filter <- filter[o]
+sw <- sw[order(reduceMethodNames(row.names(sw),1), rowMeans(sw)), o]
 ra <- data.frame(meanSilWidth=rowMeans(sw), dataset=reduceMethodNames(row.names(sw),1))
-chm(list("Cluster mean silhouette width"=sw), scaleTitle="row z-score", scale="row", row_ann=ra, show_row_names=FALSE, col_sortFn=mean, cluster_rows=FALSE)
+colnames(sw) <- reduceMethodNames(colnames(sw),2)
+w <- 4*1:(ncol(sw)/4)-2
+colnames(sw)[-w] <- ""
+chm(list("Cluster mean silhouette width"=sw), scaleTitle="row z-score", scale="row", row_ann=ra, show_row_names=FALSE, cluster_rows=FALSE, cluster_columns=FALSE, top_annotation=HeatmapAnnotation(Filter=filter))
 ```

### 2019-05-22 main_figures/figure3_normalization.Rmd
updated figure 3

@@
 chm(ll, scale="column", scaleTitle="z-score", value_format="%.2f", 
-            sameScale=T, row_sortFn=function(x){ mean(x,na.rm=T) }, cluster_columns=FALSE)
+            sameScale=TRUE, row_sortFn=function(x){ mean(x,na.rm=T) }, cluster_columns=FALSE)
 ```

### 2019-05-23 bmc_article.tex
minor text updates

@@
 In addition to log-normalization, the standard Seurat clustering pipeline scales features so that the PCA is not dominated by a small subset of the features, and we therefore included versions of the different normalization, with or without a subsequent scaling (except sctransform's variance-stabilizing transformation, given that, conceptually, it involves something analogous to scaling).
-Finally, Seurat's `ScaleData` includes the option to regress out the effect of certain covariates, and we therefore tested its use with the proportion of mitochondrial counts and the number of detected features.
+Seurat's scaling function includes the option to regress out the effect of certain covariates, and we therefore tested its use with the proportion of mitochondrial counts and the number of detected features.
+Finally, it has been proposed that the use of stable genes, and in particular cytosolic ribosomal genes, can be used to normalized scRNAseq; we therefore evaluated a simple linear normalization based on the sum of these genes, as well as nuclear genes.
 
-An important motivation of sctransform was the observation that, even after normalization, the first principal components of various datasets tended to correlate with library size, suggesting an imperfect correlation \citep{hafemeisterSCtransform2019}.
-We therefore first assessed to what extent the first principal component still retained a correlation with the library size and number of detected features (Supplementary Figure 12A).
-In the 10X datasets, most methods successfully decreased the correlation with covariates, including the mere scaling of non-normalized data.
-This was not the case in the smartseq datasets, which required the additional regressing out of covariates.
-
-(NOTE: a correlation of PC1 with library size might be caused by the fact that subpopulations have a different average library size.
-I'll therefore repeat this evaluation with the variance explained by libSize in a PC1\textasciitilde subpopulation+libSize model.)
+An important motivation of sctransform was the observation that, even after normalization, the first principal components of various datasets tended to correlate with library size, suggesting an inadequate normalization \citep{hafemeisterSCtransform2019}.
+We therefore first assessed to what extent the first principal component still retained a correlation with the library size and number of detected features, removing the confounding covariation with the subpopulation (Supplementary Figure 12A).
+The simple step of scaling tended to remove much of the correlation with these features, and in the absence of scaling, standard Seurat normalization resulted in fairly high correlation.
+sctransform had the lowest correlation, but most methods (including normalization based on stable genes) were able to remove most of the correlation when combined with scaling.
+The only exception is Seurat normalization with scaling regressing out the number of features, which surprisingly increased the correlation.
 
@@
 Of note, even though sctransform is meant to be applied to data with unique molecular identifiers (UMI), it also performed fairly well with smartseq datasets.
-Interestingly, clusters with the lowest average silhouette width (and which are therefore more difficult to classify) tended to be less well distinguished with sctransform, and instead showed a good difference in non-normalized and/or non-scaled data, suggesting that some procedures emphasize differences more useful to track some subpopulations than others, and hence the possibility that further developments in methods could improve this step.
+Interestingly, some clusters tended to be less well distinguished with sctransform than with non-scaled (and in some cases non-normalized) methods, suggesting that some procedures emphasize differences more useful to track some subpopulations than others, and hence the possibility that further developments in methods could improve this step.
 
@@
 A standard clustering pipeline typically involves a step of selecting highly-variable genes, which is complicated by the digital nature and the mean-variance relationship of RNAseq.
-Seurat's earlier approaches involved the use of dispersion estimates standardized for the expression levels, while more recent versions rely of a different measure of variance, again standardized.
-Of note, while adjusting for the mean-variance relationship removes much of the bias towards highly-expressed genes, it is nevertheless plausible that this bias reflects biological relevance and hence is actually helpful in classifying cell types.
-\citep{townesGlmpca2019} instead suggested to use deviance, while sctransform provides its own ordering the genes by transformed variance.
+Seurat's earlier approaches involved the use of dispersion estimates standardized for the mean expression levels, while more recent versions rely of a different measure of variance, again standardized.
+Of note, while adjusting for the mean-variance relationship removes much of the bias towards highly-expressed genes, it is plausible that this bias reflects biological relevance and hence is actually helpful in classifying cell types.
+Another common practice for selecting features is to use those with the highest mean expression, and recently \citep{townesGlmpca2019} instead suggested to use deviance, while sctransform provides its own ordering the genes by transformed variance.
 We therefore compared these methods.
@@
 The main observations were reflected in the ARI of the resulting clustering (Figure 4B): non-standardized measures of variability, including mere expression level, tended to outperform more complex metrics.
-In general, we found deviance and Seurat's unstandardized estimate of variance to provide the best results across datasets and normalization methods.
+In general, we found deviance and Seurat's un-standardized estimate of variance to provide the best results across datasets and normalization methods.
 Increasing the number of features selected also systematically led to an increase in the accuracy of the clustering.

### 2019-05-26 supplementary_figures/supp5_dr.Rmd
added DR-related supplementaries

---
title: "Supplementary Figures relative to dimensionality reduction"
author: "Pierre-Luc Germain"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r, include=FALSE}
if(!exists('FIG_NB')){
  FIG_NB <- 0
  getFigNb <- function(increment=FALSE){
    if(increment) FIG_NB <<- FIG_NB + 1
    FIG_NB
  }
}
```

```{r}
suppressPackageStartupMessages({
  library(ggplot2)
  library(cowplot)
  library(ComplexHeatmap)
})
theme_set(theme_cowplot(font_size = 13))
suppressMessages(devtools::load_all("../../pipComp/"))
data(datasets)
source("../misc_functions.R")
```

# Supplementary Figure `r getFigNb(TRUE)`

```{r, fig.width=8, fig.height=8}
res <- readRDS("../data/dr_intermediateSummary.rds")
sw <- res$dimreduction$clust.avg.silwidth[[1]]
colnames(sw) <- reduceMethodNames(colnames(sw),c(3,4,6))
colnames(sw) <- gsub("norm.","",colnames(sw),fixed=T)
sw <- matrix(unlist(sw),ncol=ncol(sw), dimnames=dimnames(sw))
df <- data.frame(Norm=reduceMethodNames(colnames(sw),1), Selection=reduceMethodNames(colnames(sw),2), DimRed=reduceMethodNames(colnames(sw),3), stringsAsFactors=FALSE)
df$Norm[grep("GlmPCA", df$DimRed)] <- ""
ag <- aggregate(t(sw), by=df, FUN=median)
ag <- ag[order(ag[,3],ag[,1],ag[,2]),]
sw2 <- t(ag[,-1:-3])
colnames(sw2) <- apply(ag[,c(1,3)],1,collapse=" ",FUN=paste)

# hma <- HeatmapAnnotation(df=data.frame(meanSilWidth=rowMeans(sw2), dataset=reduceMethodNames(row.names(sw),1)))
# chm(list("Cluster mean silhouette width"=t(sw2)), scaleTitle="meanSilWidth", scale="none", show_row_names=TRUE, show_column_names=FALSE, row_ann=ag[,1:3], top_annotation=hma, cluster_rows=FALSE)

hma <- HeatmapAnnotation(df=ag[,1:3])
chm(list("Cluster mean silhouette width"=sw2), scaleTitle="meanSilWidth", scale="none", row_ann=data.frame(meanSilWidth=rowMeans(sw), dataset=reduceMethodNames(row.names(sw),1)), show_row_names=FALSE, top_annotation=hma)
```

### Supplementary Figure `r getFigNb()`
Average silhouette width of real subpopulations when using the first 10 components of different dimensionality reductions. Weighing the components according to their variance increases the distinctiveness of the subpopulations. Nevertheless, Seurat's standard PCA appears superior to the alternatives.

\newpage

# Supplementary Figure `r getFigNb(TRUE)`

```{r, fig.width=8.5, fig.height=5}
r2 <- res$dimreduction$PC.R2
r2$norm <- gsub("norm.","",reduceMethodNames(r2$method,6),fixed=T)
r2 <- r2[which(r2$dim <= 10),]
r2$R.on.Clusters <- sqrt(r2$value)
r3 <- cast2(r2, formula=norm~dataset, value.var="R.on.Clusters")
r2m <- dplyr::summarise(group_by(r2,dataset), meanR=mean(R.on.Clusters))
ggplot(r2, aes(norm, R.on.Clusters, fill=norm)) + geom_violin() + facet_wrap(~dataset, nrow=1) + coord_flip() + theme(legend.position="none") + stat_summary(fun.y="mean", geom="point") + geom_hline(data=r2m, aes(yintercept=meanR)) + ggtitle("Proportion of the top dimensions explained by subpopulations") + ylab("sqrt(R^2) of dimension ~ subpopulation") + theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

### Supplementary Figure `r getFigNb()`

Proportion of the variance in the first 10 principal components that is explained by real subpopulations according to different dimensionality reduction methods.

\newpage

# Supplementary Figure `r getFigNb(TRUE)`

```{r}
res <- readRDS("../data/dr_endSummary.rds")
res2 <- res[which(abs(res$nbClusters-datasets[res$dataset,"subpopulations"])<=1),]
ll <- lapply(split(res2, res2$norm), FUN=function(x){
  m2 <- cast2(x, formula=dr~dataset, fun.aggregate=median, value.var="ARI")
})

chm(ll, scale="column", scaleTitle="z-score", sameScale=TRUE, value_format="%.2f", row_sortFn=function(x){ median(x,na.rm=T) }, cluster_columns=FALSE)
```

### Supplementary Figure `r getFigNb()`
ARI of the clustering based on different normalization and dimensionality reduction procedures.

\newpage

# Supplementary Figure `r getFigNb(TRUE)`

```{r, fig.width=8, fig.height=8.5}
library(viridis)
res <- readRDS("../data/clust_endSummary.rds")
res2 <- res[which(res$clust=="clust.seurat" & res$k==20 & res$dims>2),]
res2 <- res2[which(res2$dr=="seurat.pca" & res2$dataset!="Kumar"),]

res3 <- aggregate(res2[,"nbClusters"], by=res2[,c("dataset","dims","resolution")], FUN=mean)
colnames(res3)[4] <- "nb"
ll <- lapply(split(res3, res3$dataset), FUN=function(x){
  ggplot(x, aes(factor(dims), factor(resolution), fill=nb)) + stat_bin2d() + 
    scale_fill_viridis() + ylab("Resolution") + xlab("Number of dimensions") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1))
})
for(i in names(ll)) ll[[i]] <- ll[[i]] + ggtitle(i)
plot_grid(plotlist = ll, nrow=3)
```

### Supplementary Figure `r getFigNb()`
Mean number of clusters detected depending on the resolution and number of dimensions used. Increasing the number of dimensions tends to decrease the number of identified clusters, especially at resolutions around the default value.

\newpage

### 2019-05-26 bmc_article.tex
updated manuscript

@@
 
+We next investigated the step of dimensionality reduction, comparing in particular Seurat's PCA, GLM-PCA \citep{townesGlmpca2019}, and a two-step PCA.
+Seurat's default PCA weighs the cell embeddings by the variance of each component,+
+
 \section*{Discussion}
@@
 
-To estimate the correlation between the first principal components and covariates such as library size, we first fitted a linear model on the subpopulations, and correlated the residuals of this model with the covariate of interest.
+To estimate the correlation between the principal components and covariates such as library size, we first fitted a linear model on the subpopulations, and correlated the residuals of this model with the covariate of interest.
 

### 2019-05-27 bmc_article.tex
updated manuscript...

@@
 
+Finally, although there were differences in specific cases, we observed no robust tendency of certain normalization methods towards over- or under-estimating the number of clusters when using with Seurat clustering (Supplementary Figure 14).
+
 \subsection*{Feature selection}
@@
 Reasoning that a selection method should ideally select genes whose variability is more between subpopulations than within, we first assessed whether each selection method selected genes with a high proportion of variance or deviance explained by (real) subpopulations (the latter estimated through edgeR).
-Estimates of variance explained by subpopulations based on a standard Seurat normalization or on sctransform data were highly correlated (Supplementary Figure 14A), and there was also a good agreement with deviance explained, although lowly-expressed genes in particular could have a high deviance explained without having much of their variance explained by subpopulations (Supplementary Figure 14B-D).
-We therefore compared, using both of these estimates, the proportion of the cumulative variance/deviance explained by the top X genes that could be retrieved by each gene ranking method (Supplementary Figures 15-16).
+Estimates of variance explained by subpopulations based on a standard Seurat normalization or on sctransform data were highly correlated (Supplementary Figure 14A), and there was also a good agreement with deviance explained, although lowly-expressed genes in particular could have a high deviance explained without having much of their variance explained by subpopulations (Supplementary Figure 15B-D).
+We therefore compared, using both of these estimates, the proportion of the cumulative variance/deviance explained by the top X genes that could be retrieved by each gene ranking method (Supplementary Figures 16-17).
 We first focused on the first 1000 genes to highlight differences between methods, although the differences tended to decrease with a higher number of genes selected (Figure \ref{fig:figure4}A).
-A first observation was that standardized measures of variability were systematically worse than their non-standardized  counterparts in selecting genes with a high proportion of variance explained by subpopulations, while instead proving most of the time superior in selecting genes with a high deviance explained by subpopulations (Figure \ref{fig:figure4} and Supplementary Figures 15-16).
+A first observation was that standardized measures of variability were systematically worse than their non-standardized  counterparts in selecting genes with a high proportion of variance explained by subpopulations, while instead proving most of the time superior in selecting genes with a high deviance explained by subpopulations (Figure \ref{fig:figure4} and Supplementary Figures 16-17).
 Deviance proved the method of choice to maximize the former (with mere expression level proving surprisingly good), but did not perform so well to select genes with a high deviance explained.
@@
 
-We next investigated whether a selection based on these methods impacted the proportion of variance explained by subpopulations in the first principal components (Supplementary Figure 17).
+We next investigated whether a selection based on these methods impacted the proportion of variance explained by subpopulations in the first principal components (Supplementary Figure 18).
 To validate the previous assay, included a selection based on the genes whose proportion variance or deviance explained was highest.
@@
 
-We next investigated the step of dimensionality reduction, comparing in particular Seurat's PCA, GLM-PCA \citep{townesGlmpca2019}, and a two-step PCA.
-Seurat's default PCA weighs the cell embeddings by the variance of each component,+
+We next investigated the step of dimensionality reduction, comparing in particular Seurat's PCA, GLM-PCA \citep{townesGlmpca2019}, and a two-step PCA procedure in which cells are first clustered in rough clusters according to a standard PCA, a new PCA is performed on each, and the main components of the two are put together (see methods).
+Seurat's default PCA weighs the cell embeddings by the variance of each component, so we also tested the impact of this weighting with each method.
+
+The impact of the dimensionality reduction method was far greater than that of normalization or feature selection (e.g.
+Supplementary Figure 19).
+Although the first components identified by GLM-PCA tended to have a higher proportion of their variance explained by real subpopulations and increase the average silhouette width of already well-defined subpopulations, overall Seurat's PCA procedure proved superior on all metrics (Figure 5 and Supplementary Figures 19-21).
+In particular, weighing the components had a strong positive impact on silhouette widths and resulting ARI.
+
+We next investigated the impact of the number of dimensions used on the clustering.
+Overall, increasing the number of dimensions robustly led to a decrease in the number of clusters, although the effect was milder with data processed through sctransform (Supplementary Figures 14 and 22).
+This tended to be translated to the accuracy of the clustering (Supplementary Figure 23), although in both cases (number of clusters and ARI) the resolution parameter had a much stronger impact.
+
+\subsubsection*{Rules for selecting components}
+
+\subsection*{Clustering}
 
@@
 \subsection*{Doublet detection method}
-Our doublet detection method is available at \url{https://github.com/plger/plDoubletFinder}.
+Our doublet detection method is available at \url{https://github.com/plger/scDblFinder}.
+Version 0.1 was used here.
 Briefly, we first cluster the cells using scran's \textit{quickCluster} method, favouring overclustering (default minimum size of 20), and reduce the dataset to the union of 1500 most expressed genes of each cluster.
@@
 \subsection*{Filter sets}
-Lenient, default, and stringent sets of filters...
+The \textit{default} set of filters excludes cells that are outliers according to at least two of the following thresholds: log10_total_counts $>$2.5 MADs or $<$5 MADs, log10_total_features $>$2.5 MADs or $<$5 MADs, pct_counts_in_top_20_features $>$ or $<$ 5 MADs, featcount_dist (distance to expected ratio of log10 counts and features) $>$ or $<$ 5 MADs, pct_counts_Mt $>$ 2.5 MADs and $>$ 0.08.
+
+The \textit{stringent} set of filters uses the same thresholds, but excludes a cell if it is an outlier on any single distribution.
+
+The \textit{lenient} set of filters excludes cells that are outliers on at least two distributions by at least 5 MADs, except for pct_counts_Mt where the threshold is $>$ 3 MADs and $>$ 0.08.
 

### 2019-05-27 supplementary_figures/supp3_normalization.Rmd
updated supp

--- 2019-05-18 supplementary_figures/supp3_normalization.Rmd
@@
   library(dplyr)
+  library(viridis)
 })
@@
 
-```{r norm_silwidth, fig.width=8, fig.height=7}
+```{r norm_silwidth, fig.width=8, fig.height=4.5}
 res <- readRDS("../../resNew/norm2_intermediateSummary.rds")
 sw <- res$dimreduction$clust.avg.silwidth[[1]]
-colnames(sw) <- gsub("norm.","",reduceMethodNames(colnames(sw), 3),fixed=T)
-sw <- sw[order(reduceMethodNames(row.names(sw),1), rowMeans(sw)),]
+colnames(sw) <- gsub("norm.","",reduceMethodNames(colnames(sw), 2:3),fixed=T)
+filter <- factor(reduceMethodNames(colnames(sw),1), levels=c("lenientfilter", "clustFilterStringent", "deffilter", "stringentfilter"))
+levels(filter) <- c("lenient", "stringent cluster-wise", "default", "stringent")
+filtCols <- c(default="#4477AA", lenient="#117733", "stringent cluster-wise"="#DDCC77", stringent="#CC6677")
+o <- order(reduceMethodNames(colnames(sw),2), filter)
+filter <- filter[o]
+sw <- sw[order(reduceMethodNames(row.names(sw),1), rowMeans(sw)), o]
 ra <- data.frame(meanSilWidth=rowMeans(sw), dataset=reduceMethodNames(row.names(sw),1))
-chm(list("Cluster mean silhouette width"=sw), scaleTitle="row z-score", scale="row", row_ann=ra, show_row_names=FALSE, col_sortFn=mean, cluster_rows=FALSE)
+colnames(sw) <- reduceMethodNames(colnames(sw),2)
+w <- 4*1:(ncol(sw)/4)-2
+colnames(sw)[-w] <- ""
+chm(list("Cluster mean silhouette width"=sw), scaleTitle="row z-score", scale="row", row_ann=ra, show_row_names=FALSE, cluster_rows=FALSE, cluster_columns=FALSE, top_annotation=HeatmapAnnotation(Filter=filter, col=list(Filter=filtCols)))
+```
+
+```{r norm_silwidth2, fig.width=8, fig.height=4}
+w <- order(ra$meanSilWidth)[1:10]
+sw2 <- sw[w,]
+chm(list("Cluster mean silhouette width"=sw2), scaleTitle="Mean\nSilhouette\nWidth", show_row_names=TRUE, cluster_rows=FALSE, cluster_columns=FALSE, top_annotation=HeatmapAnnotation(Filter=filter, col=list(Filter=filtCols)))
 ```
@@
 Mean silhouette width of true subpopulations across the first 10 components with different normalization and scaling approaches.
+Above the silhouette widths are scaled for each subpopulation; below the subpopulations with the lowest mean silhouette width are represented.
 
 \newpage
+
+# Supplementary Figure `r getFigNb(TRUE)`
+
+```{r norm_nbClusters, fig.width=8, fig.height=8.5}
+res <- readRDS("../data/norm2_endSummary.rds")
+res2 <- res[which(res$filt=="deffilter" & res$norm %in% c("norm.scran", "norm.scran.noscale", "norm.seurat", "norm.seuratvst", "norm.stableGsum")),]
+res2$diff.nbClusters <- res2$nbClusters-datasets[res2$dataset,"subpopulations"]
+res3 <- aggregate(res2[,"diff.nbClusters"], by=res2[,c("dataset","dims","resolution","norm")], FUN=mean)
+colnames(res3)[5] <- "nb"
+res3$norm <- gsub("norm.","",res3$norm, fixed=T)
+# ggplot(res3, aes(factor(dims), factor(resolution), fill=nb)) + stat_bin2d() + 
+#   scale_fill_viridis() + ylab("Resolution") + xlab("Number of dimensions") +
+#   theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
+#   coord_flip() + facet_grid(dataset~norm)
+
+ll <- lapply(split(res3, res3$dataset), FUN=function(x){
+  ggplot(x, aes(factor(resolution), factor(dims), fill=nb)) + stat_bin2d() + 
+    scale_fill_viridis() + xlab("Resolution") + ylab("dims") +
+    facet_grid(.~norm)
+})
+for(i in names(ll)){
+  ll[[i]] <- ll[[i]] + ggtitle(i)
+  if(i==names(ll)[length(ll)]){
+    ll[[i]] <- ll[[i]] + theme(axis.text.x = element_text(angle = 90, hjust = 1))
+  }else{
+    ll[[i]] <- ll[[i]] + theme(axis.title.x=element_blank(), axis.text.x=element_blank(), axis.ticks.x=element_blank())
+  }
+}
+ll[[4]] <- ll[[4]] + ylab("Number of dimensions")
+plot_grid(plotlist = ll[-2], nrow=6, rel_heights=c(1,1,1,1,1,1.5))
+```
+
+### Supplementary Figure `r getFigNb()`
+Mean difference between the number of detected clusters and the number of real subpopulations, depending on the normalization method, the resolution and the number of dimensions used.
+The Kumar dataset is not shown here due to a lack of variation in the number of clusters detected.

### 2019-05-27 supplementary_figures/supp4_selection.Rmd
updated supp

@@
 
-```{r, fig.width=8, fig.height=9, warnings=FALSE, messages=FALSE}
+```{r, fig.width=8, fig.height=9, results="hide", warnings=FALSE, messages=FALSE}
 ll <- lapply(names(gi), FUN=function(x){

### 2019-05-27 supplementary_figures/supp6_clust.Rmd
updated supp

---
title: "Supplementary Figures relative to clustering"
author: "Pierre-Luc Germain"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r, include=FALSE}
if(!exists('FIG_NB')){
  FIG_NB <- 0
  getFigNb <- function(increment=FALSE){
    if(increment) FIG_NB <<- FIG_NB + 1
    FIG_NB
  }
}
```

```{r}
suppressPackageStartupMessages({
  library(ggplot2)
  library(cowplot)
  library(ComplexHeatmap)
  library(reshape2)
  library(dplyr)
})
theme_set(theme_cowplot(font_size = 13))
suppressMessages(devtools::load_all("../../pipComp/"))
data(datasets)
source("../misc_functions.R")
```

# Supplementary Figure `r getFigNb(TRUE)`

```{r, fig.width=8, fig.height=8}
res <- readRDS("../data/clust_endSummary.rds")
res <- res[which(res$dims>2 & res$resolution %in% c(0.05, 0.15, 0.3, 0.5, 0.8, 1, 1.2)),]
res$method <- paste0( gsub("clust.","",res$clustmethod, fixed=T), 
                      " k=",res$k, 
                      ifelse( res$clustmethod=="clust.seurat", paste(" res=",res$resolution), paste0(" steps=",res$steps)) )
res$diff.nbClusters <- res$nbClusters-datasets[res$dataset,"subpopulations"]
ll <- lapply( unique(res$dataset), FUN=function(x){
  as.matrix(cast2(res[which(res$dataset==x),], formula=method~dims, fun.aggregate=mean, value.var="diff.nbClusters"))
})
names(ll) <- unique(res$dataset)
mmax <- ceiling(max(abs(sapply(ll,FUN=range))))
cf <- circlize::colorRamp2(c(-mmax, 0, mmax), colors=c("blue", "white", "red"))
chm(ll, sameScale=TRUE, scaleTitle="Difference\nto real nb\nsubpopulations\n", colors=cf, cluster_columns=FALSE, cluster_rows=FALSE)
```

### Supplementary Figure `r getFigNb()`
Difference between the number of detected clusters and the number of real subpopulations according to different numbers of dimensions and clustering paramters.

\newpage

### 2019-05-27 bmc_article.tex
Updates from Overleaf

--- 2019-05-23 bmc_article.tex
@@
 We next investigated the impact of different normalization strategies.
+Beside the standard log-normalization included in Seurat, we tested scran's pooling-based normalization \citep{lunPooling2016}, sctransform's variance-stabilizing transformation \citep{hafemeisterSCtransform2019}, and normalization based on stable genes \citep{linStableGenes2018, deekeStablyExpressed2018}.
 In addition to log-normalization, the standard Seurat clustering pipeline scales features so that the PCA is not dominated by a small subset of the features, and we therefore included versions of the different normalization, with or without a subsequent scaling (except sctransform's variance-stabilizing transformation, given that, conceptually, it involves something analogous to scaling).
-Seurat's scaling function includes the option to regress out the effect of certain covariates, and we therefore tested its use with the proportion of mitochondrial counts and the number of detected features.
-Finally, it has been proposed that the use of stable genes, and in particular cytosolic ribosomal genes, can be used to normalized scRNAseq; we therefore evaluated a simple linear normalization based on the sum of these genes, as well as nuclear genes.
+Finally, Seurat's `ScaleData` includes the option to regress out the effect of certain covariates, and we therefore tested its use with the proportion of mitochondrial counts and the number of detected features.
 
-An important motivation of sctransform was the observation that, even after normalization, the first principal components of various datasets tended to correlate with library size, suggesting an inadequate normalization \citep{hafemeisterSCtransform2019}.
-We therefore first assessed to what extent the first principal component still retained a correlation with the library size and number of detected features, removing the confounding covariation with the subpopulation (Supplementary Figure 12A).
-The simple step of scaling tended to remove much of the correlation with these features, and in the absence of scaling, standard Seurat normalization resulted in fairly high correlation.
-sctransform had the lowest correlation, but most methods (including normalization based on stable genes) were able to remove most of the correlation when combined with scaling.
-The only exception is Seurat normalization with scaling regressing out the number of features, which surprisingly increased the correlation.
+An important motivation of sctransform was the observation that, even after normalization, the first principal components of various datasets tended to correlate with library size, suggesting an imperfect correlation \citep{hafemeisterSCtransform2019}.
+We therefore first assessed to what extent the first principal component still retained a correlation with the library size and number of detected features (Supplementary Figure 12A).
+In the 10X datasets, most methods successfully decreased the correlation with covariates, including the mere scaling of non-normalized data.
+This was not the case in the smartseq datasets, which required the additional regressing out of covariates.
+
+(NOTE: a correlation of PC1 with library size might be caused by the fact that subpopulations have a different average library size.
+I'm repeating this evaluation with the variance explained by libSize in a PC1\textasciitilde subpopulation+libSize model.)
 
@@
 Of note, even though sctransform is meant to be applied to data with unique molecular identifiers (UMI), it also performed fairly well with smartseq datasets.
-Interestingly, some clusters tended to be less well distinguished with sctransform than with non-scaled (and in some cases non-normalized) methods, suggesting that some procedures emphasize differences more useful to track some subpopulations than others, and hence the possibility that further developments in methods could improve this step.
+Interestingly, clusters with the lowest average silhouette width (and which are therefore more difficult to classify) tended to be less well distinguished with sctransform, and instead showed a good difference in non-normalized and/or non-scaled data, suggesting that some procedures emphasize differences more useful to track some subpopulations than others, and hence the possibility that further developments in methods could improve this step.
 
@@
 A standard clustering pipeline typically involves a step of selecting highly-variable genes, which is complicated by the digital nature and the mean-variance relationship of RNAseq.
-Seurat's earlier approaches involved the use of dispersion estimates standardized for the mean expression levels, while more recent versions rely of a different measure of variance, again standardized.
-Of note, while adjusting for the mean-variance relationship removes much of the bias towards highly-expressed genes, it is plausible that this bias reflects biological relevance and hence is actually helpful in classifying cell types.
-Another common practice for selecting features is to use those with the highest mean expression, and recently \citep{townesGlmpca2019} instead suggested to use deviance, while sctransform provides its own ordering the genes by transformed variance.
+Seurat's earlier approaches involved the use of dispersion estimates standardized for the expression levels, while more recent versions rely of a different measure of variance, again standardized.
+Of note, while adjusting for the mean-variance relationship removes much of the bias towards highly-expressed genes, it is nevertheless plausible that this bias reflects biological relevance and hence is actually helpful in classifying cell types.
+\citep{townesGlmpca2019} instead suggested to use deviance, while sctransform provides its own ordering the genes by transformed variance.
 We therefore compared these methods.
@@
     \caption{\textbf{Evaluation of feature selection methods.
-A:} Ability of different feature ranking methods to capture genes with a high proportion of variance (left) or deviance (right) explained.
-\texbf{B:} Accuracy of clustering (at the right number of clusters) based on the given clustering methods and number of genes selected.}
+A:} Ability of different feature ranking methods to capture genes with a high proportion of variance (left) or deviance (right) explained by real subpopulations.
+rankMean and rankMax refer to, respectively, the mean rank and the maximum rank in the following metrics: variance, standardized variance, sctransform, deviance.
+\texbf{B:} Accuracy of clustering (at the right number of clusters) based on the given selection methods and number of genes selected.}
     \label{fig:figure4}
@@
 The main observations were reflected in the ARI of the resulting clustering (Figure 4B): non-standardized measures of variability, including mere expression level, tended to outperform more complex metrics.
-In general, we found deviance and Seurat's un-standardized estimate of variance to provide the best results across datasets and normalization methods.
+In general, we found deviance and Seurat's unstandardized estimate of variance to provide the best results across datasets and normalization methods.
 Increasing the number of features selected also systematically led to an increase in the accuracy of the clustering.

### 2019-05-27 supplementary_figures/supp5_dr.Rmd

@@
   library(ComplexHeatmap)
+  library(reshape2)
+  library(dplyr)
 })
@@
 res2 <- res2[which(res2$dr=="seurat.pca" & res2$dataset!="Kumar"),]
-
-res3 <- aggregate(res2[,"nbClusters"], by=res2[,c("dataset","dims","resolution")], FUN=mean)
+res2$diff.nbClusters <- res2$nbClusters-datasets[res2$dataset,"subpopulations"]
+res3 <- aggregate(res2[,"diff.nbClusters"], by=res2[,c("dataset","dims","resolution")], FUN=mean)
 colnames(res3)[4] <- "nb"
@@
 ### Supplementary Figure `r getFigNb()`
-Mean number of clusters detected depending on the resolution and number of dimensions used.
+Mean difference between the number of detected clusters and the number of real subpopulations, depending on the resolution and number of dimensions used.
+Based on sctransform and seurat PCA.
 Increasing the number of dimensions tends to decrease the number of identified clusters, especially at resolutions around the default value.
@@
 \newpage
+
+# Supplementary Figure `r getFigNb(TRUE)`
+
+```{r, fig.width=8, fig.height=8.5}
+library(viridis)
+res <- readRDS("../data/clust_endSummary.rds")
+res2 <- res[which(res$clust=="clust.seurat" & res$k==20 & res$dims>2 & res$resolution > 0.05 & res$resolution <= 1),]
+res2 <- res2[which(res2$dr=="seurat.pca" & res2$dataset!="Kumar"),]
+res3 <- aggregate(res2[,"ARI"], by=res2[,c("dataset","dims","resolution")], FUN=mean)
+ll <- lapply(split(res3, res3$dataset), FUN=function(x){
+  ggplot(x, aes(factor(dims), factor(resolution), fill=ARI)) + stat_bin2d() + 
+    scale_fill_viridis() + ylab("Resolution") + xlab("Number of dimensions") +
+    theme(axis.text.x = element_text(angle = 90, hjust = 1))
+})
+for(i in names(ll)) ll[[i]] <- ll[[i]] + ggtitle(i)
+plot_grid(plotlist = ll, nrow=3)
+```
+
+### Supplementary Figure `r getFigNb()`
+Adjusted Rand Index of clustering depending on the resolution and number of dimensions used. Based on sctransform and seurat PCA.

### 2019-05-27 bmc_article.tex

--- 2019-05-27 bmc_article.tex
@@
 We next investigated the impact of different normalization strategies.
+Beside the standard log-normalization included in Seurat, we tested scran's pooling-based normalization \citep{lunPooling2016}, sctransform's variance-stabilizing transformation \citep{hafemeisterSCtransform2019}, and normalization based on stable genes \citep{linStableGenes2018, deekeStablyExpressed2018}.
 In addition to log-normalization, the standard Seurat clustering pipeline scales features so that the PCA is not dominated by a small subset of the features, and we therefore included versions of the different normalization, with or without a subsequent scaling (except sctransform's variance-stabilizing transformation, given that, conceptually, it involves something analogous to scaling).

### 2019-05-27 bmc_article.tex
Updates from Overleaf

@@
 Supplementary Figure 19).
-Although the first components identified by GLM-PCA tended to have a higher proportion of their variance explained by real subpopulations and increase the average silhouette width of already well-defined subpopulations, overall Seurat's PCA procedure proved superior on all metrics (Figure 5 and Supplementary Figures 19-21).
+Although the first components identified by GLM-PCA tended to have a higher proportion of their variance explained by real subpopulations and increase the average silhouette width of already well-defined subpopulations, overall Seurat's PCA procedure proved superior on all metrics (Figure \ref{fig:figure5} and Supplementary Figures 19-21).
 In particular, weighing the components had a strong positive impact on silhouette widths and resulting ARI.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\textwidth]{{main_figures/figure5_dr_files/figure-html/unnamed-chunk-2-1.png}}
+    \caption{Evaluation of the dimensionality reduction strategies.}
+    \label{fig:figure5}
+\end{figure}
 
@@
 \subsection*{Filter sets}
-The \textit{default} set of filters excludes cells that are outliers according to at least two of the following thresholds: log10_total_counts $>$2.5 MADs or $<$5 MADs, log10_total_features $>$2.5 MADs or $<$5 MADs, pct_counts_in_top_20_features $>$ or $<$ 5 MADs, featcount_dist (distance to expected ratio of log10 counts and features) $>$ or $<$ 5 MADs, pct_counts_Mt $>$ 2.5 MADs and $>$ 0.08.
+The \textit{default} set of filters excludes cells that are outliers according to at least two of the following thresholds: log10\_total\_counts $>$2.5 MADs or $<$5 MADs, log10\_total\_features $>$2.5 MADs or $<$5 MADs, pct\_counts\_in\_top\_20\_features $>$ or $<$ 5 MADs, featcount\_dist (distance to expected ratio of log10 counts and features) $>$ or $<$ 5 MADs, pct\_counts\_Mt $>$ 2.5 MADs and $>$ 0.08.
 
@@
 
-The \textit{lenient} set of filters excludes cells that are outliers on at least two distributions by at least 5 MADs, except for pct_counts_Mt where the threshold is $>$ 3 MADs and $>$ 0.08.
+The \textit{lenient} set of filters excludes cells that are outliers on at least two distributions by at least 5 MADs, except for pct\_counts\_Mt where the threshold is $>$ 3 MADs and $>$ 0.08.
+
+For cluster-wise filters, clusters were first identified with \textit{scran::quickCluster} and the filters were then applied separately for each cluster.
 

### 2019-06-03 bmc_article.tex
mark edits jun 3

@@
 
-Although we use some datasets based on other protocols, our focus is especially on 10X datasets and on methods that do not include exogenous control RNA, i.e.
+Although we use some datasets based on other protocols, our focus is especially on droplet-based datasets and on methods that do not include exogenous control RNA, i.e.
 spike-ins (see Table \ref{tab:table1} and Supplementary Figure 1 for a description of the datasets).
-As Seurat \cite{satijaSeurat2015} was previously shown to outperform other clustering approaches \cite{duoClustering2018,tianMixology2018}, we used the Seurat pipeline as the starting point to perform an integrated investigation of 1) doublet identification, 2) cell filtering, 3) normalization, 4) feature selection, 5) dimensional reduction, 6) clustering.
+Since graph-based clustering  \cite{satijaSeurat2015} was previously shown to consistently perform well across several datasets \cite{duoClustering2018,tianMixology2018}, we used the Seurat pipeline as the starting point to perform an integrated investigation of: 1) doublet identification, 2) cell filtering, 3) normalization, 4) feature selection, 5) dimension reduction, 6) clustering.
 We compare not only competing approaches, but also more fine-grained parameter and variations on common methods.
@@
 Each step can include various parameters, all of which must be given alternative values.
-The \textit{runPipeline} command will then proceed in a step-wise fashion to computer only the parts of the pipelines that have not already been computed.
+The \textit{runPipeline} command will then proceed in a step-wise fashion to compute only the parts of the pipelines that have not already been computed.
 In addition, each step can output evaluation metrics, which together provide a multi-layered evaluation, where variations in a given parameter can be evaluated using all metrics from this point downward in the pipeline.
@@
 While this has the virtue of making the results comparable, in practice the number of subpopulations is typically unknown, and tools that operate well in this optimal context might not necessarily be best overall.
-Clustering results are also very sensitive, and as we will see they might not always capture improvements in earlier steps of the pipeline.
+Clustering results are also very sensitive, and as shown below, might not always capture improvements in earlier steps of the pipeline.
 
@@
 In addition to these methods, we reasoned that an approach such as DoubletFinder could be considerably sped up by being applied directly on counts and by using a pre-clustering to create neotypic doublets more efficiently.
-We therefore developed a simple and fast package implementing this for doublet detection, with the added perk of accounting for uncertainty in the expected doublet rate and using meta-cells from the clusters to even include triplets (see methods).
+We therefore developed a simple and fast package implementing this for doublet detection, with the added advantage of accounting for uncertainty in the expected doublet rate and using meta-cells from the clusters to even include triplets (see methods).
 
@@
 
-\subsubsection*{Excluding more cells is not necessarily good}
+\subsubsection*{Excluding more cells is not necessarily better}
 
@@
 Supplementary Figure 1 shows the distributions of some of the typical cell properties commonly used.
-Of note, these properties tend to be correlated, but are not always: for instance, while a high proportion of mitochondrial reads is often correlated with a high proportion of the counts in the top features, there can also be other reasons for an over-representation of highly-expressed features (Supplementary Figure 3).
+Of note, these properties tend to be correlated, but not always: for instance, while a high proportion of mitochondrial reads is often correlated with a high proportion of the counts in the top features, there can also be other reasons for an over-representation of highly-expressed features (Supplementary Figure 3).
 In addition, in our experience 10X datasets also exhibit a very tight correlation between the total counts and the total features even across very different cell types (Supplementary Figure 4).
@@
 Reasoning that the cells we wish to eliminate are cells that would be misclassified, we measured the rate of misclassification of each cell in each dataset across a variety of clustering pipelines, correcting for the median misclassification rate of the subpopulation, and then evaluated what properties could be predictive of misclassification (Supplementary Figures 5-7).
-We could not identify any property or simple combination thereof which would be consistently predictive of misclassification; the only feature that consistently stood out across multiple datasets (the Zheng datasets) was that cells with a very high read counts have a higher chance of being misclassified.
+We could not identify any property or simple combination thereof that would be consistently predictive of misclassification; the only feature that consistently stood out across multiple datasets (the Zheng datasets) was that cells with a very high read counts have a higher chance of being misclassified.
 
@@
 Interestingly, regression analysis suggests that increasing the number of distributions on which a cell must deviate in order to be excluded tended to be positively associated with ARI, while increasing the number of median absolute deviations required to be an outlier as associated with a decrease in ARI.
-We therefore suggest excluding cells that are outliers on more than 1 distributions.
+We therefore suggest excluding cells that are outliers in more than 1 distribution.
 The impact of filtering on the basis of total counts or feature counts was not always robust, nor always positive, and we could detect no significant impact of filtering on the basis of the proportion of mitochondrial reads - however it must be said that the benchmark datasets count very few cells with a very high proportion of mitochondrial reads (Supplementary Figure 1).
@@
 We found that filtering had a much milder impact than doublet removal (Figure \ref{fig:figure3}), and that more stringent filtering tended to result in higher silhouette width, in a higher proportion of the variance in the first components explained by subpopulations (Supplementary Figure 10), and often in a higher adjusted Rand index (ARI) of the clustering at the right number of clusters (Figure \ref{fig:figure3}).
-However, part of the gain is to be ascribed to the removal of cells from subpopulations (especially in the Zheng datasets) which are difficult to distinguish, and for which more stringent clustering did not improve silhouette width (Supplementary Figure 10A).
+However, part of the gain is to be ascribed to the removal of cells from subpopulations (especially in the Zheng datasets) that are difficult to distinguish, and for which more stringent clustering did not improve silhouette width (Supplementary Figure 10A).
 For these reasons, we therefore recommend using a more lenient filter (e.g.
-lenient or default filter described here) or a cluster-based filter to avoids most of the subpopulation bias while excluding strong outliers.
+lenient or default filter described here) or a cluster-based filter to avoid most of the subpopulation bias while excluding strong outliers.
 
@@
 Seurat's scaling function includes the option to regress out the effect of certain covariates, and we therefore tested its use with the proportion of mitochondrial counts and the number of detected features.
-Finally, it has been proposed that the use of stable genes, and in particular cytosolic ribosomal genes, can be used to normalized scRNAseq; we therefore evaluated a simple linear normalization based on the sum of these genes, as well as nuclear genes.
+Finally, it has been proposed that the use of stable genes, and in particular cytosolic ribosomal genes, can be used to normalize scRNAseq; we therefore evaluated a simple linear normalization based on the sum of these genes, as well as nuclear genes.
 

### 2019-06-18 main_figures/sample_plots_for_diagram.Rmd
updated diagram

---
title: "Sample plots for diagram"
author: "Pierre-Luc Germain"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
suppressPackageStartupMessages({
  library(ComplexHeatmap)
  library(circlize)
  library(ggplot2)
  library(cowplot)
  devtools::load_all("../../pipComp/")
})
source("../misc_functions.R")
```

# Filtering

```{r, fig.width=2.6, fig.height=1.8}
fres <- readRDS("../../resNew/filtering_intermediateSummary.rds")
pc <- fres$filtering$pcCells
pc <- pc[grep("doublet",row.names(pc),invert=T),]
pc <- pc[2:4,c(-12,-13,-15,-17:-30)]
pc[2,] <- pc[2,]/2
pc[1,] <- pc[1,]*2
pal <- colorRamp2(c(0,2,5,9,max(pc)),c("#0000FFFF", "#B38BFFFF", "#FFFFFFFF", "#FF9E81FF", "#FF0000FF"))
Heatmap(pc[c(2,1,3),], name="% out", col=pal, cluster_columns=FALSE, show_column_names=FALSE, show_row_names=FALSE, cluster_rows=FALSE, row_title="Filter methods", column_title="Subpopulations")
```

# Feature selection

```{r, fig.width=2.6, fig.height=1.8}
noAx <-   theme(axis.text.x=element_blank(),
                axis.ticks.x=element_blank(),
                axis.text.y=element_blank(),
                axis.ticks.y=element_blank() )

gi <- readRDS("../data/GI.rds")
gi <- gi$Zhengmix4uneq
preds <- list( method1=gi$deviance,
               method2=gi$seurat.variance.standardized,
               method3=gi$seurat.variance,
               method4=gi$seurat.dispersion.scaled)
cumVE2(preds, gi$vst.R2) + xlim(0,400) + ylab("Cum.var.expl") + noAx + theme(legend.position="none")
```

# Reduction

```{r, fig.width=5, fig.height=4.2}
rdat <- data.frame( cluster=rep(LETTERS[1:3],c(50,80,50)) )
rdat$x1 <- unlist(mapply(sd=c(2,2,2.5), n=c(50,80,50), mean=c(-5,5,-5), FUN=rnorm))
rdat$y1 <- unlist(mapply(sd=c(2,2,2.5), n=c(50,80,50), mean=c(-5,0,5), FUN=rnorm))
rdat$x2 <- unlist(mapply(sd=c(1, 1, 1.25), n=c(50,80,50), mean=c(-5,5,-5), FUN=rnorm))
rdat$y2 <- unlist(mapply(sd=c(1, 1, 1.25), n=c(50,80,50), mean=c(-5,0,5), FUN=rnorm))
cm.x <- sapply(unique(rdat$cluster), FUN=function(x) mean(rdat$x1[which(rdat$cluster==x)]))
cm.y <- sapply(unique(rdat$cluster), FUN=function(x) mean(rdat$y1[which(rdat$cluster==x)]))
ls <- jitter(rdat$x1)-cm.x[rdat$cluster]
rdat$lib.size <- ls-min(c(0,ls))+2
w <- which(rdat$cluster=="C")
rdat$lib.size[w] <- rdat$lib.size[w]*1.02
rdat$lib.size <- ceiling(rdat$lib.size)
seg <- data.frame( x1=c(cm.x[1], cm.x[1]),
                   y1=rep(cm.y[1],2),
                   x2=cm.x[-1],
                   y2=cm.y[-1]
                    )
p1 <-   ggplot(data=rdat, aes(x1,y1,colour=cluster)) + geom_point() + noAx + 
    labs(x="dim1", y="dim2") + theme(legend.position="none") + ggtitle("Poorly-defined") +
    geom_segment(data=seg, aes(x=x1, y=y1, xend=x2, yend=y2, colour=NULL), linetype="dashed") + 
   geom_point(data=data.frame(x1=cm.x[1],y1=cm.y[1]), size=15, shape=1, colour="black")
p2 <- ggplot(data=rdat, aes(x2,y2,colour=cluster)) + geom_point() + noAx+ labs(x="dim1", y="dim2") + theme(legend.position="none") + ggtitle("Well-defined")

m1 <- rbind( c(120, 90, 100),
             c(100, 80, 100),
             c(90, 70, 90) )
m2 <- rbind( c(90, 70, 80),
             c(90, 75, 75),
             c(70, 60, 75) )
m3 <- as.matrix( c(5,3,7) )
m1 <- jitter(m1,factor=100)
m2 <- jitter(m2,factor=100)

gp <- gpar(fontsize=12)
h <- Heatmap(m1, cluster_rows=FALSE, cluster_columns=FALSE, show_heatmap_legend=FALSE, column_title="silhouette\nwidth", row_title="method", column_title_gp=gp) + 
    Heatmap(cbind(rep(NA,3), m2, rep(NA,3)), cluster_columns=FALSE, show_heatmap_legend=FALSE, column_title="var expl.\nby subpop", na_col="white", column_title_gp=gp) + 
    Heatmap(cbind(rep(NA,3),m3,rep(NA,3)), cluster_columns=FALSE, column_title="libsize\ncorr.", show_heatmap_legend=FALSE, na_col="white", column_title_gp=gp)
h <- grid.grabExpr(draw(h))

plot_grid(
  p1,
  ggplot(data=rdat, aes(x1,y1,colour=lib.size)) + geom_point() + noAx + theme(legend.text=element_blank()) + ggtitle("Correlation with\nlibrary size") + labs(x="dim1", y="dim2", colour="lib\nsize"),
  p2,
  plot_grid( h, NULL, nrow=2, rel_heights=c(9,1)),
  rel_widths=c(2.2,3), scale=c(0.95,1,0.95,0.9)
)
```

# Clustering

```{r, fig.width=4.3, fig.height=1.8}
rdat$pred <- rdat$cluster
rdat$pred[sample(1:nrow(rdat), 40)] <- rdat$pred[sample(1:nrow(rdat), 40)]
m <- sapply(LETTERS[1:3], FUN=function(x){
  x <- which(rdat$cluster==x)
  sapply(LETTERS[1:3], FUN=function(y){
    y <- which(rdat$pred==y)
    length(intersect(x,y))/length(union(x,y))
  })
})
colnames(m) <- LETTERS[24:26]
h <- Heatmap(m+matrix(rnorm(9,sd=0.1),nrow=3), cluster_columns=FALSE, cluster_rows=FALSE, row_title="real\nsubpopulation", column_title="predicted\ncluster", show_heatmap_legend=FALSE, row_title_gp=gp, column_title_gp=gp)
h <- grid.grabExpr(draw(h))

d <- data.frame(method=rep(LETTERS[1:3], 3), type=rep(c("precision","recall","ARI"), each=3), value=c(1,0.9,0.85, 0.77, 0.78, 0.65, 0.95, 0.9, 0.8))
p <- ggplot(d, aes(x=method, y=value, fill=method)) + geom_col() + facet_grid(~type) + theme(axis.text.x=element_blank(), axis.text.y=element_blank(), axis.title.y=element_blank(), axis.ticks = element_blank(), legend.position="none", axis.line = element_blank(), strip.background = element_rect(colour="white", fill="white"))

plot_grid( h, p, rel_widths=c(8,10))
```

### 2019-06-18 bmc_article.tex

@@
 
-In addition to these methods, we reasoned that an approach such as DoubletFinder could be considerably sped up by being applied directly on counts and by using a pre-clustering to create neotypic doublets more efficiently.
+In addition to these methods, we reasoned that an approach such as DoubletFinder could be considerably sped up by being applied directly on counts and by using a pre-clustering to create neotypic doublets (i.e.
+doublets from different cell types) more efficiently.
 We therefore developed a simple and fast package implementing this for doublet detection, with the added advantage of accounting for uncertainty in the expected doublet rate and using meta-cells from the clusters to even include triplets (see methods).

### 2019-11-03 main_figures/figure3_normalization.Rmd
updated comparison of normalization methods

--- 2019-05-15 main_figures/figure3_normalization.Rmd
@@
   library(ComplexHeatmap)
+  library(matrixStats)
 })
-suppressMessages(devtools::load_all("../../pipComp/"))
-data(datasets)
+suppressMessages(devtools::load_all("../../pipeComp/"))
 source("../misc_functions.R")
@@
 ```{r}
-res <- readRDS("../../resNew/norm_intermediateSummary.rds")
+res <- readRDS("../data/results_norm.rds")
 ```
 
-```{r figure3, fig.width=9, fig.height=5}
-r2 <- res$dimreduction$PC.R2
-r2$norm <- gsub("norm.","",reduceMethodNames(r2$method,3),fixed=T)
-r2 <- r2[which(r2$dim <= 5),]
-r2$R.on.Clusters <- sqrt(r2$value)
-r3 <- cast2(r2, formula=norm~dataset, value.var="R.on.Clusters")
+```{r}
+p1 <- scrna_evalPlot_DR(res, "covarRes", agg.by = "norm", anno_legend=FALSE) +
+    scrna_evalPlot_DR(res, "covarRes", covar = "log10_total_features", agg.by = "norm", anno_legend=FALSE)
+p1 <- grid.grabExpr(draw(p1))
 
-res.n <- readRDS("../../resNew/norm_endSummary.rds")
-res.n <- res.n[which(res.n$norm != "norm.edgeR"),]
-res.n$norm <- relevel(factor(res.n$norm), "norm.seurat")
-levels(res.n$norm) <- gsub("norm.","",levels(res.n$norm),fixed=T)
+p2 <- grid.grabExpr(draw(
+  scrna_evalPlot_DR(res, what="silhouette", agg.by = "norm", value_format = "", show_heatmap_legend = TRUE)))
 
-m <- cast2(res.n, formula=norm~dataset, fun.aggregate=median, value.var="ARI")
-m2 <- cast2(res.n[which(res.n$nbClusters==datasets[res.n$dataset,"subpopulations"]),], formula=norm~dataset, fun.aggregate=median, value.var="ARI")
+H <- scrna_evalPlot_clustAtTrueK(res, "ARI", agg.by = "norm", scale = TRUE, reorder_rows = TRUE, anno_legend=FALSE)
+H <- H + scrna_evalPlot_clust(res, "MI", agg.by = "norm", scale = TRUE, reorder_rows=row.names(H@matrix), anno_legend=FALSE)
+p3 <- grid.grabExpr(draw(H))
+```
 
-ll <- list("Var in PCs 1-5 explained"=r3,
-           "median ARI"=m,
-           "ARI at true number of clusters"=m2)
-chm(ll, scale="column", scaleTitle="z-score", value_format="%.2f", 
-            sameScale=T, row_sortFn=function(x){ mean(x,na.rm=T) }, cluster_columns=FALSE)
+```{r figure3, fig.width=9, fig.height=9}
+plot_grid(p1,p2,p3, nrow = 3)
 ```

### 2019-11-04 bmc_article.tex
Updates from Overleaf

@@
 
-We define a pipeline, in R, as a list of functions executed consecutively on the output of the previous one (Figure 0).
+We define a pipeline, in R, as a list of functions executed consecutively on the output of the previous one (Figure 0A).
 Each step can include various parameters, all of which must be given alternative values.
@@
 We therefore also wanted to capture whether the effect of a given parameter alteration is robust to changes in the rest of the pipeline, or rather specific to a set of other pipeline parameters.
-We therefore proceeded in a step-wise fashion, first testing a large variety of parameters at the early steps of the pipeline along with only a set of mainstream options downstream, then selecting main alternatives and proceeding to a more detailed benchmark of the next step.
-The basic pipeline used through most of this paper is described at the bottom of Figure 0.
+We therefore proceeded in a step-wise fashion, first testing a large variety of parameters at the early steps of the pipeline along with only a set of mainstream options downstream, then selecting main alternatives and proceeding to a more detailed benchmark of the next step (Figure 0B).
 

### 2019-11-09 main_figures/figure4_selection.Rmd
updated selection figure

--- 2019-05-15 main_figures/figure4_selection.Rmd
@@
 suppressPackageStartupMessages({
+  library(SingleCellExperiment)
   library(ggplot2)
@@
   library(ComplexHeatmap)
+  library(matrixStats)
 })
 source("../misc_functions.R")
-suppressMessages(devtools::load_all("../../pipComp/"))
+suppressMessages(devtools::load_all("../../pipeComp/"))
 data(datasets)
@@
 ```{r}
-GI <- readRDS("../data/GI.rds")
-get_auc <- function(preds,ve.var="vst.R2",n=1000){
+lf <- list.files("../datasets", "SCE", full.names=TRUE)
+names(lf) <- gsub("\\.SCE\\.rds","",basename(lf))
+GI <- lapply(lf, FUN=function(x) rowData(readRDS(x)))
+get_auc <- function(preds,ve.var="vst.varExp",n=1000){
+    preds$vst.Gini <- -1*preds$vst.Gini
+    p2 <- as.matrix(preds[,c("seurat.variance", "seurat.variance.standardized", "seurat.res.var", "deviance")])
     preds2 <- list( deviance=preds$deviance,
@@
                    sctransform=preds$seurat.res.var,
+                   #rankMean=colMeans(colRanks(p2)),
+                   #rankMax=colMaxs(colRanks(p2)),
                    expression=preds$total_counts
@@
 aucs1 <- sapply(GI,FUN=get_auc)
+h1 <- Heatmap(aucs1 , name="Proportion", cluster_columns=FALSE, 
+           bottom_annotation=.ds_anno(colnames(aucs1),TRUE), show_heatmap_legend = TRUE,
+           show_column_names = FALSE, col=viridisLite::inferno(100), column_title="Variance explained")
 aucs2 <- sapply(GI,ve.var="devianceExplained", FUN=get_auc)
-p1 <- grid.grabExpr(draw(chm(list("Variance explained"=aucs1, "Deviance explained"=aucs2), sameScale=TRUE, scaleTitle="Proportion", cluster_columns=FALSE)))
+h2 <- Heatmap( aucs2[row.names(h1@matrix),], 
+               cluster_rows=FALSE, cluster_columns=FALSE, 
+              bottom_annotation=.ds_anno(colnames(aucs2),FALSE), show_heatmap_legend = FALSE,
+              show_column_names = FALSE, col=viridisLite::inferno(100), column_title="Deviance explained")
+p1 <- grid.grabExpr(draw(h1+h2))
 ```
 
-```{r figure4, fig.width=6, fig.height=9}
-res <- readRDS("../../resNew/sel_endSummary.rds")
-res <- res[grep("devianceExplained|varExp", res$sel, invert=TRUE),]
-res$selection <- paste(res$sel, res$selnb)
-res2 <- res[which(abs(res$nbClusters-datasets[res$dataset,"subpopulations"])<=0),]
-ll <- lapply( split(res2, res2$norm), FUN=function(x){
-  cast2(x, formula=selection~dataset, value.var="ARI")
-})
-names(ll) <- c("scran.noscale","standard norm", "sctransform")
-p2 <- grid.grabExpr(draw(chm(ll[-1], sameScale=TRUE, scaleTitle="ARI", cluster_rows=FALSE, cluster_columns=FALSE)))
-
-plot_grid(p1, p2, labels=LETTERS[1:2], nrow=2, rel_heights=1:2)
+```{r figure4, fig.width=7.5, fig.height=6.5}
+res <- readRDS("../data/results_selection.rds")
+cl <- res$clustering
+vst <- cl[grep("seuratvst",row.names(cl)),]
+snorm <- cl[grep("seuratvst",row.names(cl), invert=TRUE),]
+h1 <- scrna_evalPlot_clustAtTrueK(snorm[grep("=1000;",row.names(snorm)),], "ARI", scale = FALSE, title="ARI at true # clusters", anno_legend = FALSE)
+h2 <- scrna_evalPlot_clustAtTrueK(vst[grep("=1000;",row.names(vst)),], "ARI", scale = FALSE, reorder_rows=row.names(h1@matrix), title="sctransform\nARI at true # clusters", name="ARI2", anno_legend = FALSE)
+p2 <- grid.grabExpr(draw(h1+h2))
+plot_grid(p1, p2, labels=LETTERS[1:2], nrow=2, rel_heights = c(4,5))
 ```
 
-**A:** Estimating different methods of ranking genes based on their ability to capture genes with a high proportion of variance (left) or deviance (right) explained. 
-**B: ** Accuracy of clusterings (at the right number of clusters) based on the given clustering methods and number of genes selected.
+**A:** Estimating different methods of ranking genes based on their ability to capture genes with a high proportion of variance (left) or deviance (right) explained by the subpopulations.
+**B: ** Accuracy of clusterings (at the true number of clusters) when selecting 1000 genes using the given methods.
+Based on standard Seurat normalization (left) or sctransform (right).
+The methods "vst.varExp" and "devianceExplained" are the estimates used in **A** to evaluate the selection methods, and were included here only for validation purpose.

### 2019-11-09 bmc_article.tex

@@
 Although we use some datasets based on other protocols, our focus is especially on droplet-based datasets and on methods that do not include exogenous control RNA, i.e.
-spike-ins (see Table \ref{tab:table1} and Supplementary Figure 1 for a description of the datasets).
+spike-ins (see Table \ref{tab:table1} and \ref{fig:figure1} for a description of the datasets).
 Since graph-based clustering  \cite{satijaSeurat2015} was previously shown to consistently perform well across several datasets \cite{duoClustering2018,tianMixology2018}, we used the Seurat pipeline as the starting point to perform an integrated investigation of: 1) doublet identification, 2) cell filtering, 3) normalization, 4) feature selection, 5) dimension reduction, 6) clustering.
@@
 
-We define a pipeline, in R, as a list of functions executed consecutively on the output of the previous one (Figure 0A).
+We define a pipeline, in R, as a list of functions executed consecutively on the output of the previous one (\ref{fig:figure2}A).
 Each step can include various parameters, all of which must be given alternative values.
@@
 We therefore also wanted to capture whether the effect of a given parameter alteration is robust to changes in the rest of the pipeline, or rather specific to a set of other pipeline parameters.
-We therefore proceeded in a step-wise fashion, first testing a large variety of parameters at the early steps of the pipeline along with only a set of mainstream options downstream, then selecting main alternatives and proceeding to a more detailed benchmark of the next step (Figure 0B).
+We therefore proceeded in a step-wise fashion, first testing a large variety of parameters at the early steps of the pipeline along with only a set of mainstream options downstream, then selecting main alternatives and proceeding to a more detailed benchmark of the next step (\ref{fig:figure2}B).
 
-\includegraphics[scale=0.8]{figure0}
+\begin{figure}
+    \centering
+    \includegraphics[width=\textwidth,keepaspectratio]{dataset_description}
+    \caption{\textbf{Overview of the benchmark datasets used.}}
+    \label{fig:figure1}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\textwidth,keepaspectratio]{figure0}
+    \caption{\textbf{Overview of the pipeComp framework and its application to a scRNAseq clustering pipeline.}}
+    \label{fig:figure2}
+\end{figure}
 
@@
 
-While most methods accurately identified the doublets in the 3 cell lines dataset (mixology10x3cl), the 5 lines dataset (mixology10x5cl) proved more difficult (Figure \ref{fig:figure1}B).
-Our method (scDblFinder) achieved the best accuracy while running in a very reasonable time (Figure \ref{fig:figure1}).
+While most methods accurately identified the doublets in the 3 cell lines dataset (mixology10x3cl), the 5 lines dataset (mixology10x5cl) proved more difficult (Figure \ref{fig:figure3}B).
+Our method (scDblFinder) achieved the best accuracy while running in a very reasonable time (Figure \ref{fig:figure3}).
 Across datasets, cells called as doublets tended to be classified in the wrong cluster more often than other cells (Figure 1D).
-We therefore tested whether this method also improved the accuracy of the clustering across all benchmark datasets, and found that it consistently did (Figure \ref{fig:figure2}).
+We therefore tested whether this method also improved the accuracy of the clustering across all benchmark datasets, and found that it consistently did (Figure \ref{fig:figure4}).
 
@@
 The corrected percentage of misclassification is the proportion of times a given cell is misclassified across hundreds of clustering pipelines with various parameters, minus the median misclassification rate of the cell's subpopulation.}
-    \label{fig:figure1}
+    \label{fig:figure3}
 \end{figure}
@@
     \caption{Proportion of cells filtered out and accuracy of the cluster (under two normalizations) for different sets of filters (see methods).}
-    \label{fig:figure2}
+    \label{fig:figure4}
 \end{figure}
@@
 On the basis of these observations, we defined three sets of criteria as lenient, default and stringent filtering (see methods), and investigated them further.
-We found that filtering had a much milder impact than doublet removal (Figure \ref{fig:figure3}), and that more stringent filtering tended to result in higher silhouette width, in a higher proportion of the variance in the first components explained by subpopulations (Supplementary Figure 10), and often in a higher adjusted Rand index (ARI) of the clustering at the right number of clusters (Figure \ref{fig:figure3}).
+We found that filtering had a much milder impact than doublet removal (Figure \ref{fig:figure4}), and that more stringent filtering tended to result in higher silhouette width, in a higher proportion of the variance in the first components explained by subpopulations (Supplementary Figure 10), and often in a higher adjusted Rand index (ARI) of the clustering at the right number of clusters (Figure \ref{fig:figure4}).
 However, part of the gain is to be ascribed to the removal of cells from subpopulations (especially in the Zheng datasets) that are difficult to distinguish, and for which more stringent clustering did not improve silhouette width (Supplementary Figure 10A).
@@
 We next investigated the impact of normalization on the ability to distinguish subpopulations, which was much greater than that of filtering (Supplementary Figure 13).
-For many clusters, scaling tended to reduce both the average silhouette width (Supplementary Figure 12B) and the proportion of variance in the top principal components explained by subpopulations (Figure \ref{fig:figure3} and Supplementary Figure 13), and was associated with a poorer adjusted Rand index (ARI) of the resulting clustering (Figure \ref{fig:figure3}).
+For many clusters, scaling tended to reduce both the average silhouette width (Supplementary Figure 12B) and the proportion of variance in the top principal components explained by subpopulations (Figure \ref{fig:figure5} and Supplementary Figure 13), and was associated with a poorer adjusted Rand index (ARI) of the resulting clustering (Figure \ref{fig:figure5}).
 Regressing out covariates systematically gave poorer performance on all metrics.
-sctransform produced the highest cluster average silhouette width (Supplementary Figure 12B) as well as ARI (Figure \ref{fig:figure3}), but not always the highest mean proportion of variance in the top components explained by subpopulations (Figure \ref{fig:figure3} and Supplementary Figure 13).
+sctransform produced the highest cluster average silhouette width (Supplementary Figure 12B) as well as ARI (Figure \ref{fig:figure5}), but not always the highest mean proportion of variance in the top components explained by subpopulations (Figure \ref{fig:figure5} and Supplementary Figure 13).
 Of note, even though sctransform is meant to be applied to data with unique molecular identifiers (UMI), it also performed fairly well with smartseq datasets.
@@
     \caption{\textbf{Evaluation of normalization procedures:} proportion of the variance in the first 5 components explained by the subpopulations, median ARI, and ARI at the true number of clusters.}
-    \label{fig:figure3}
+    \label{fig:figure5}
 \end{figure}
@@
 We therefore compared, using both of these estimates, the proportion of the cumulative variance/deviance explained by the top X genes that could be retrieved by each gene ranking method (Supplementary Figures 16-17).
-We first focused on the first 1000 genes to highlight differences between methods, although the differences tended to decrease with a higher number of genes selected (Figure \ref{fig:figure4}A).
-A first observation was that standardized measures of variability were systematically worse than their non-standardized  counterparts in selecting genes with a high proportion of variance explained by subpopulations, while instead proving most of the time superior in selecting genes with a high deviance explained by subpopulations (Figure \ref{fig:figure4} and Supplementary Figures 16-17).
+We first focused on the first 1000 genes to highlight differences between methods, although the differences tended to decrease with a higher number of genes selected (Figure \ref{fig:figure6}A).
+A first observation was that standardized measures of variability were systematically worse than their non-standardized  counterparts in selecting genes with a high proportion of variance explained by subpopulations, while instead proving most of the time superior in selecting genes with a high deviance explained by subpopulations (Figure \ref{fig:figure6} and Supplementary Figures 16-17).
 Deviance proved the method of choice to maximize the former (with mere expression level proving surprisingly good), but did not perform so well to select genes with a high deviance explained.
@@
 \texbf{B:} Accuracy of clustering (at the right number of clusters) based on the given selection methods and number of genes selected.}
-    \label{fig:figure4}
+    \label{fig:figure6}
 \end{figure}
@@
 Supplementary Figure 19).
-Although the first components identified by GLM-PCA tended to have a higher proportion of their variance explained by real subpopulations and increase the average silhouette width of already well-defined subpopulations, overall Seurat's PCA procedure proved superior on all metrics (Figure \ref{fig:figure5} and Supplementary Figures 19-21).
+Although the first components identified by GLM-PCA tended to have a higher proportion of their variance explained by real subpopulations and increase the average silhouette width of already well-defined subpopulations, overall Seurat's PCA procedure proved superior on all metrics (Figure \ref{fig:figure7} and Supplementary Figures 19-21).
 In particular, weighing the components had a strong positive impact on silhouette widths and resulting ARI.
@@
     \caption{Evaluation of the dimensionality reduction strategies.}
-    \label{fig:figure5}
+    \label{fig:figure7}
 \end{figure}

### 2019-11-09 main_figures/figure3_normalization.Rmd
added legend

@@
 ```{r figure3, fig.width=9, fig.height=9}
-plot_grid(p1,p2,p3, nrow = 3)
+plot_grid(p1,p2,p3, nrow = 3, labels=LETTERS[1:3])
 ```
+
+Evaluation of normalization procedures.
+**A:** Residual correlation of the first principal component with library size (left) and detection rate (right), after accounting for biological differences between subpopulations.
+**B:** Average silhouette width per true subpopulation, where higher silhouette width means a higher separability.
+**C:** Clustering accuracy, measured by the ARI at the true number of cluster (left) and by the average mutual information (MI) of the cluster assignment with true subpopulations.

### 2019-11-20 main_figures/figure2_filtering.Rmd
updated figure2

--- 2019-05-09 main_figures/figure2_filtering.Rmd
@@
 ---
+
+```{r}
+suppressPackageStartupMessages({
+  library(ComplexHeatmap)
+  library(circlize)
+  library(ggplot2)
+  library(cowplot)
+})
+theme_set(theme_cowplot())
+suppressMessages(devtools::load_all("../../pipeComp/"))
+source("../misc_functions.R")
+data(datasets)
+```
+
+```{r figure2, fig.width=9, fig.height=7}
+res <- readRDS("../../flt/summary.rds")
+m <- scrna_evalPlot_filtering(res, returnTable=TRUE)
+m$doubletRemoval <- !grepl("doubletmethod=none;", m$method)
+m$filter <- factor(sapply(strsplit(as.character(m$method),"\\.|="), FUN=function(x) rev(x)[1]),
+                     c("nofilter","lenient","default","pca","pca2","stringent"))
+cols <- c("black",plgINS::getQualitativePalette(length(levels(m$filter))-1))
+names(cols) <- levels(m$filter)
+
+# ggplot(m, aes(maxPCout, meanF1, group=method, colour=filter, shape=doubletRemoval)) + geom_point(size=3) + facet_wrap(~dataset, scales="free") + scale_color_manual(values=cols) + xlab("Max proportion of subpopulation excluded")
+
+ggplot(m, aes(maxPCout, F1atK, group=method, colour=filter, shape=doubletRemoval)) + geom_point(size=3) + facet_wrap(~dataset, scales="free") + scale_color_manual(values=cols) + xlab("Max proportion of subpopulation excluded") + ylab("mean F1 at true number of clusters")
+```

### 2019-11-27 main_figures/dimensionality.Rmd
renamed main figures, updated filtering and dimensionality figures

---
title: "Dimensionality"
author: "Pierre-Luc Germain"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    keep_md: true
---

```{r}
suppressPackageStartupMessages({
  library(ggplot2)
  library(cowplot)
  library(ComplexHeatmap)
  library(matrixStats)
})
source("../misc_functions.R")
suppressMessages(devtools::load_all("../../pipeComp/"))
source("../../pipeComp/inst/extdata/scrna_alternatives.R")
theme_set(theme_cowplot(font_size = 12))
```

```{r}
o <- readRDS("../../ndims.RDS")
o <- lapply(o, FUN=function(x){
  x$nselected$elbow <- x$nselected$elbow-1
  x
})
o2 <- data.table::rbindlist( lapply(o, FUN=function(x){
  data.frame( method=names(x$nselected), 
              nsel=unlist(x$nselected), 
              time=x$elapsed )
  }), idcol="dataset")
ve <- sapply(o, FUN=function(x) x$varExplained)
row.names(ve) <- 1:nrow(ve)

cols <- getQualitativePalette(9)[-5:-6]
names(cols) <- colnames(ve)
p1a <- Heatmap(t(ve)[,1:15],col=c("white","black"), cluster_columns=FALSE, cluster_rows=FALSE,name="Prop.\nvar", column_title = "Variance in each PC\nexplained by subpopulations", row_title = "Dataset", right_annotation = rowAnnotation(dataset=colnames(ve), col=list(dataset=cols), show_legend=FALSE), show_row_names = FALSE)
p1a <- grid.grabExpr(draw(p1a))

ve.elbows <- apply(ve,2,FUN=farthestPoint)-1
o2$diff <- o2$nsel-ve.elbows[as.character(o2$dataset)]
tmp <- sapply(split(o2$time, o2$method), mean)
mt2 <- cut(tmp, breaks=c(0,1,5,20,3000), labels=c("<1s","<5s","<20s","minutes \nto hours"))
names(mt2) <- names(tmp)
o2$Time <- mt2[as.character(o2$method)]
tcols <- c("darkblue", "blue", "violet", "red")
names(tcols) <- levels(mt2)
o2$avgTime <- tmp[as.character(o2$method)]
p1b <- ggplot(o2, aes(reorder(method, diff, function(x) mean(abs(x))), diff, group=method, fill=Time)) + geom_violin() + coord_flip() + geom_hline(yintercept=0, linetype="dashed") + ylab("Difference to 'real' number of dimensions") + scale_fill_manual(values=tcols) + theme(axis.title.y=element_blank(), legend.position=c(0.7, 0.3))
p1 <- plot_grid(p1a, p1b, rel_widths = c(0.42,0.58), labels=LETTERS[1:2])
```

```{r, fig.width=8, fig.height=4}
res <- readRDS("../data/results_dimensionality.rds")
sw <- res$dimensionality$clust.avg.silwidth$selected
sw <- sw[grep("=doublet",row.names(sw)),]
n <- sapply(strsplit(row.names(sw),"="),FUN=function(x) rev(x)[1])
w <- grep("seuratvst",row.names(sw))
n[w] <- paste0("sctransform+",n[w])
row.names(sw) <- n
sw2 <- sw[order(grepl("sctransform",row.names(sw))*1000+rowMeans(sw)+apply(sw,1,probs=0.8,FUN=quantile),decreasing=TRUE),order(colMeans(sw),decreasing=TRUE)]
h1 <- Heatmap(sw2, cluster_columns = FALSE, cluster_rows = FALSE, 
        name="silhouette width", col=.silScale(sw2), show_column_names = FALSE,
        column_title = "Subpopulation mean silhouette width",
        bottom_annotation=.ds_anno(colnames(sw2)), heatmap_legend_param = list(direction="horizontal", legend_width=unit(3, "cm")))
p2 <- grid.grabExpr(draw(h1, heatmap_legend_side="bottom"))

res2 <- res$clustering
res2 <- res2[grep("=doublet",row.names(res2)),]
method <- as.character(parsePipNames(row.names(res2))$dims)
w <- grep("seuratvst", row.names(res2))
method[w] <- paste0("sctransform+",method[w])
row.names(res2) <- paste0("method=",method,";",row.names(res2))

h2 <- scrna_evalPlot_clust(res2, what="NMI", agg.by = c("method"), scale=TRUE, reorder_rows = TRUE, anno_legend = FALSE)
h2 <- h2 + 
  scrna_evalPlot_clust(res2, what="min_pr", agg.by = c("method"), reorder_rows=h2, scale=TRUE, anno_legend = FALSE) + 
scrna_evalPlot_clustAtTrueK(res2, agg.by = c("method"), reorder_rows=h2, scale=TRUE, anno_legend = FALSE)
p3 <- grid.grabExpr(draw(h2))
```

```{r figure6, fig.width=9, fig.height=9.5}
plot_grid( p1, p2, p3, nrow=3, rel_heights=c(0.39,0.3,0.31), labels=c(NA,LETTERS[3:4]))
```

**A:** Estimated dimensionality using the proportion of variance in each component explained by subpopulations. **B:** Difference between `real' dimensionality (from **A**) and various dimensionality estimation methods, along with computing time. **C:** Average silhouette width per subpopulation across a selection of dimensionality estimators, using sctransform or standard seurat normalization. **D:** Clustering accuracy of across the normalization/dimensionality methods.

### 2019-11-27 main_figures/doublets.Rmd
renamed main figures, updated filtering and dimensionality figures

--- 2019-05-13 main_figures/figure1_doublets.Rmd
@@
 
-```{r, fig.height=6, fig.width=7}
+```{r figure1, fig.height=6, fig.width=7}
 plot_grid( plot_grid(p1, p2, nrow=1, labels=LETTERS[1:2]), plot_grid(p3, p4, nrow=1, labels=LETTERS[3:4], rel_widths=c(2,3)), nrow=2)

### 2019-11-27 main_figures/dr.Rmd
renamed main figures, updated filtering and dimensionality figures

---
title: "Figure 5 dimensionality reduction"
author: "Pierre-Luc Germain"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    keep_md: true
---

```{r}
suppressPackageStartupMessages({
  library(ggplot2)
  library(cowplot)
  library(ComplexHeatmap)
  library(matrixStats)
})
source("../misc_functions.R")
suppressMessages(devtools::load_all("../../pipComp/"))
data(datasets)
theme_set(theme_cowplot(font_size = 12))
```

```{r, fig.width=8, fig.height=4}
res <- readRDS("../data/dr_intermediateSummary.rds")
sw <- res$dimreduction$clust.avg.silwidth[[1]]
sw <- sw[,grep("seuratvst",colnames(sw))]
sw <- sw[,grep("deviance",colnames(sw))]
colnames(sw) <- reduceMethodNames(colnames(sw),6)
sw <- matrix(unlist(sw),ncol=ncol(sw), dimnames=dimnames(sw))
ag <- aggregate(sw, by=list(reduceMethodNames(row.names(sw),1)), FUN=mean)
row.names(ag) <- ag[,1]; ag[,1] <- NULL
sw <- t(ag)

r2 <- res$dimreduction$PC.R2
r2 <- r2[grep("seuratvst",r2$method),]
r2 <- r2[grep("deviance",r2$method),]
r2$dr <- reduceMethodNames(r2$method, 6)
r2 <- r2[which(r2$dim <= 5),]
r3 <- cast2(r2, formula=dr~dataset, value.var="value")

res <- readRDS("../data/dr_endSummary.rds")
res2 <- res[which(abs(res$nbClusters-datasets[res$dataset,"subpopulations"])<=1 & 
                    res$norm=="norm.seuratvst"),]
m.ari <- cast2(res2, formula=dr~dataset, fun.aggregate=median, value.var="ARI")

ll <- list( "mean silhouette width"=sw,
            "PC1-5 var explained"=r3,
            "median ARI"=m.ari)
ll <- lapply(ll, FUN=function(x){
  row.names(x) <- gsub(".noW weighted", "", paste(row.names(x), "weighted"), fixed=T)
  x
})
chm(ll, scale="column", scaleTitle="z-score", sameScale=TRUE, value_format="%.2f", row_sortFn=function(x){ median(x,na.rm=T) }, cluster_columns=FALSE)
```

Mean of subpopulation average silhouette width (left), mean proportion of the variance in the first 5 components explained by real subpopulations (center), and median ARI of the resulting clustering (right).

### 2019-11-27 main_figures/filtering.Rmd
renamed main figures, updated filtering and dimensionality figures

---
title: "Figure 2 filtering"
author: "Pierre-Luc Germain"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    keep_md: true

---

```{r}
suppressPackageStartupMessages({
  library(ComplexHeatmap)
  library(circlize)
  library(ggplot2)
  library(cowplot)
})
theme_set(theme_cowplot())
suppressMessages(devtools::load_all("../../pipeComp/"))
source("../misc_functions.R")
data(datasets)
```

```{r figure2, fig.width=9, fig.height=7}
res <- readRDS("../data/results_filtering.rds")
m <- scrna_evalPlot_filtering(res, returnTable=TRUE)
m$doubletRemoval <- !grepl("doubletmethod=none;", m$method)
m$filter <- factor(sapply(strsplit(as.character(m$method),"\\.|="), FUN=function(x) rev(x)[1]),
                     c("nofilter","lenient","default","pca","pca2","stringent","veryStringent"))
cols <- c("black",plgINS::getQualitativePalette(length(levels(m$filter))-1))
names(cols) <- levels(m$filter)

# ggplot(m, aes(maxPCout, meanF1, group=method, colour=filter, shape=doubletRemoval)) + geom_point(size=3) + facet_wrap(~dataset, scales="free") + scale_color_manual(values=cols) + xlab("Max proportion of subpopulation excluded")

ggplot(m, aes(maxPCout, F1atK, group=method, colour=filter, shape=doubletRemoval)) + geom_point(size=3) +
  # geom_line(aes(group=doubletRemoval), colour="lightgrey") + 
facet_wrap(~dataset, scales="free") + scale_color_manual(values=cols) + xlab("Max proportion of subpopulation excluded") + ylab("mean F1 at true number of clusters")
```

Relationship between the maximum subpopulation exclusion rate and the average clustering accuracy per subpopulation across various filtering strategies. Of note, doublet removal appears to be desirable even when, due to the design, there are no heterotypic doublets in the data.

### 2019-11-27 main_figures/filtering_old.Rmd
renamed main figures, updated filtering and dimensionality figures

---
title: "Figure 2 filtering"
author: "Pierre-Luc Germain"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    keep_md: true

---

```{r}
suppressPackageStartupMessages({
  library(ComplexHeatmap)
  library(circlize)
})
suppressMessages(devtools::load_all("../../pipeComp/"))
source("../misc_functions.R")
data(datasets)
```

```{r figure2, fig.width=9, fig.height=4}
res <- readRDS("../data/results_filtering.rds")
co <- res$filtering
co <- co + res$doublet[sapply(strsplit(row.names(res$filtering),";"), FUN=function(x) x[1]),]
co <- abs(as.matrix(co[,grep("^pcOut",colnames(co))]))
ds <- sapply(strsplit(colnames(co),"\\."), FUN=function(x) x[2])
ds <- split(seq_len(ncol(co)), ds)
df <- data.frame(
  method=rep(row.names(co), length(ds)),
  dataset=rep(names(ds), each=nrow(co)),
  maxPCout=unlist(lapply(ds, FUN=function(x) rowMaxs(co[,x]))),
  medianPCout=unlist(lapply(ds, FUN=function(x) rowMedians(co[,x]))),
  meanF1=NA_real_,
  F1atK=NA_real_
)
ds <- sapply(strsplit(colnames(res$clustering)," "), FUN=function(x) x[1])
ds <- split(seq_len(ncol(res$clustering)), ds)
ds$stepElapsed.Koh <- NULL
ds <- lapply(ds, FUN=function(x){
  cl <- res$clustering[,x]
  cl <- cl[grep("seuratvst",row.names(cl),invert=TRUE),]
  by <- sapply(strsplit(row.names(cl),";"), FUN=function(x) paste(x[1:2],collapse=";"))
  meanF1 <- aggregate(cl[[grep("mean_F1",colnames(cl))]], by=list(by), na.rm=TRUE, FUN=mean)
  topF1 <- aggregate(cl[[grep("mean_F1",colnames(cl))]], by=list(by), na.rm=TRUE, FUN=max)
  w <- which(cl[[grep("n_clus",colnames(cl))]]==cl[[grep("nbClusts",colnames(cl))]])
  cl <- cl[w,]
  tmp <- aggregate(cl[[grep("mean_F1",colnames(cl))]], by=list(by[w]), na.rm=TRUE, FUN=mean)
  F1atK <- tmp[,2]
  names(F1atK) <- tmp[,1]
  d <- data.frame(method=meanF1[,1], meanF1=as.numeric(meanF1[,2]), 
                  topF1=as.numeric(topF1[,2]), F1atK=NA_real_)
  d[as.character(tmp[,1]),"F1atK"] <- tmp[,2]
  d
})
ds2 <- bind_rows(ds,.id="dataset")
m <- merge(df, ds2, by=c("dataset","method"))
m$doubletRemoval <- grepl("scDblFinder", m$method)
m$filter <- sapply(strsplit(as.character(m$method),"\\."), FUN=function(x) rev(x)[1])
ag <- aggregate(res$clustering)
df$F1 <- apply(df[,1:2],1,FUN=function(x){
  res$clustering[,grep("")]
  mean()
})

res.f <- readRDS("../../resNew/filtering_endSummary.rds")

res.f$pcCells <- 100-100*res.f$nbCells/datasets[as.character(res.f$dataset),"nbCells"]
res.f$method <- factor(paste(res.f$doubletmethod, res.f$filt))
res.f <- res.f[grep("clust",res.f$method, invert=TRUE),]
levels(res.f$method) <- gsub("nothingFUN ","",levels(res.f$method))

w <- which(abs(res.f$nbClusters-datasets[as.character(res.f$dataset),"subpopulations"])<=1)
#w <- which(res.f$nbClusters==datasets[as.character(res.f$dataset),"subpopulations"])
aris.1 <- cast2(res.f[intersect(w, which(res.f$norm=="norm.seurat")),], formula=method~dataset, value.var="ARI")
aris.2 <- cast2(res.f[intersect(w, which(res.f$norm=="norm.seuratvst")),], formula=method~dataset, value.var="ARI")
pc <- cast2(res.f, formula=method~dataset, value.var="pcCells")
chm(list("ARI (standard norm)"=aris.1, "ARI (sctransform)"=aris.2, "% cells out"=pc), scale=c("column","column","column"), value_format=c("%#.2f","%#.2f","%#.1f"), cluster_columns=FALSE, cluster_rows=FALSE, scaleVar=FALSE)
```

```{r figure2b, fig.width=11, fig.height=4}
HM <- chm(list("ARI"=aris.1, "Total % out"=pc), scale=c("column","column","none"), value_format=c("%#.2f","%#.1f","%#.1f"), cluster_columns=FALSE, cluster_rows=FALSE, scaleVar=FALSE, width=1)

fres <- readRDS("../../resNew/filtering_intermediateSummary.rds")
pc2 <- fres$filtering$pcCell
row.names(pc2) <- reduceMethodNames(row.names(pc2),1:2)
ha <- HeatmapAnnotation(dataset=reduceMethodNames(colnames(pc2),1))
pc2 <- pc2[row.names(aris.1),]
pal <- colorRamp2(c(0,1,5,20,max(pc2)),c("#0000FFFF", "#B38BFFFF", "#FFFFFFFF", "#FF9E81FF", "#FF0000FF"))
HM + Heatmap(pc2, name="% out", bottom_annotation=ha, col=pal, cluster_columns=FALSE, show_column_names=FALSE, width=1, column_title="% per subpopulation")
```

```{r figure2b, fig.width=8.5, fig.height=4}
nb <- fres$filtering$nbCells
pc <- fres$filtering$pcCells
tot <- tot <- 100*pc/nb
row.names(nb) <- row.names(tot) <- row.names(pc) <- reduceMethodNames(row.names(nb),1:2)
subpops <- split(colnames(nb), sapply(strsplit(colnames(nb)," "),FUN=function(x) x[[1]]))
maxOut <- sapply(subpops, FUN=function(x) rowMax(pc[,x]))
maxDeltaPC <- sapply(subpops, FUN=function(x){
  apply(pc[,x],1,FUN=function(x) max(x)-median(x))
})
row.names(maxOut) <- row.names(pc)
# HM + Heatmap(maxDeltaPC[row.names(aris.1),], name="Max % delta", col=pal, cluster_columns=FALSE, show_column_names=FALSE, width=0.5, column_title="Max % difference\nto median")
HM + Heatmap(maxOut[row.names(aris.1),], name="Max %", col=pal, cluster_columns=FALSE, width=1, column_title="Max cluster % out"
```

```{r}
chm(list("ARI"=aris.1, "Total % out"=pc[row.names(aris.1),], "Max cluster % out"=maxOut[row.names(aris.1),]), scale=c("column","column","none"), value_format=c("%#.2f","%#.1f","%#.1f"), cluster_columns=FALSE, cluster_rows=FALSE, scaleVar=FALSE, width=1)
```

```{r, fig.height=3, fig.width=7}
d <- data.frame( celltype=rep(LETTERS[1:3],c(1000,1500,200)), 
                 value=c( rnorm(1000,90,sd=50),
                          rnorm(1500,130,sd=40),
                          rnorm(200,200,sd=40) ) )
library(ggridges)
vl <- function(x) geom_vline(xintercept=x, linetype="dashed")
plot_grid(
  ggplot(d, aes(x=value)) + geom_density(fill="grey") + xlab("value (e.g. log-total-counts)") + vl(35) + vl(205),
  ggplot(d, aes(y=celltype, x=value)) + geom_density_ridges(panel_scaling = FALSE) + xlab("value (e.g. log-total-counts)") + vl(40) + vl(200)
)
```

```{r}
res.f
rf2 <- aggregate(res.f[,c(14:20)], by=res.f[,c(1,2,3,4,13,21)], FUN=mean)
rf2$doublet <- rf2$doubletmethod=="doublet.pl"
ggplot(rf2[which(rf2$norm=="norm.seurat"),], aes(pcCells, ARI, shape=doublet)) + geom_line(mapping=aes(group=resolution), colour="grey") + geom_point(mapping=aes(colour=filt, size=resolution)) + facet_wrap(~dataset, scales="free") + scale_size(range=c(1.5,4)) + scale_x_sqrt() + xlab("% cells excluded")
```

```{r}
load("../../pipComp/data/datasets.RData", verbose=TRUE)
o <- readRDS("../data/filt_allSummaries.rds")
doag <- function(n, prefix="F1"){
  x <- o[[n]]
  x <- x[which(x$norm=="norm.seurat"),]
  x <- x[grep("norm\\.seuratvst|clustFilter", row.names(x), invert=TRUE),]
  x <- x[which(abs(x$nbClusters-datasets[n,"subpopulations"]) <=1),]
  x$pcCells <- round(100*x$nbCells/datasets[n,"nbCells"],2)
  x$method <- ifelse(x$doubletmethod=="doublet.pl", paste("doublet",x$filt), x$filt)
  x <- aggregate(x[,grep(paste0("^",prefix,"\\."),colnames(x))], list(method=x[,"method"]), FUN=mean)
  row.names(x) <- x[,1]
  x[,-1]
}
pr <- do.call(cbind,lapply(names(o), prefix="pr", FUN=doag))
re <- do.call(cbind,lapply(names(o), prefix="re", FUN=doag))
f1 <- do.call(cbind,lapply(names(o), FUN=doag))
ha <- HeatmapAnnotation(dataset=rep(names(o), sapply(o,FUN=function(x) length(grep("^F1\\.",colnames(x))))))
oi <- order(rowSums(f1), decreasing=TRUE)
Heatmap(pr[oi,], cluster_rows=FALSE, bottom_annotation=ha, show_column_names=FALSE, show_heatmap_legend=FALSE, column_title="Precision", show_column_dend=FALSE) +
    Heatmap(re[oi,], cluster_rows = FALSE, bottom_annotation=ha, show_column_names=FALSE, show_heatmap_legend=FALSE, column_title="Recall", show_column_dend=FALSE) + 
    Heatmap(f1[oi,], cluster_rows = FALSE, bottom_annotation=ha, show_column_names=FALSE, name="Score", column_title="F1 score", show_column_dend=FALSE)
```

```{r, fig.width=10, fig.height=14}
cd <- readRDS("../data/cellDists.rds")
lf <- list.files("../datasets", pattern="\\.SCE\\.rds", full.names = TRUE)
names(lf) <- gsub("\\.SCE\\.rds","",basename(lf))
meta <- lapply(lf, FUN=function(x) as.data.frame(colData(readRDS(x))))
cn <- table(unlist(lapply(meta, colnames)))
cn <- names(cn)[which(cn==length(meta))]
meta2 <- lapply(names(meta), FUN=function(x){
  y <- as.data.frame(cd[[x]])
  colnames(y) <- paste("dist",colnames(y))
  x <- meta[[x]]
  n2 <- paste(x$phenoid,row.names(x),sep=".")
  x <- cbind(x[,cn], y[n2,])
  x <- split(x, x$phenoid)
  x <- lapply(x, FUN=function(x){
    x$dist.scaled <- scale(x$`dist seurat`)
    x
  })
  as.data.frame(data.table::rbindlist(x))
})
names(meta2) <- names(meta)
meta2 <- data.table::rbindlist(meta2,idcol="dataset")
# plot_grid(
#   ggplot(meta2, aes(total_counts, total_features_by_counts, colour=dist.scaled)) + geom_point(size=1) + facet_wrap(~dataset, scales = "free") + scale_x_log10() + scale_y_log10(),
#   ggplot(meta2, aes(pct_counts_in_top_50_features, pct_counts_Mt, colour=dist.scaled)) + geom_point() + facet_wrap(~dataset, scales = "free"),
#   nrow = 2, labels=c("A","B")
# )
plot_grid(plotlist=lapply(c("total_counts","total_features_by_counts"), FUN=function(x){
  ggplot(meta2, aes_string(x, 'dist.scaled')) + geom_point() + geom_density_2d() + facet_wrap(~dataset, scales="free") + scale_x_log10()
}), labels=c("A","B"), nrow=2)
plot_grid(plotlist=lapply(c("pct_counts_in_top_50_features","pct_counts_Mt"), FUN=function(x){
  ggplot(meta2, aes_string(x, 'dist.scaled')) + geom_point() + geom_density_2d() + facet_wrap(~dataset, scales="free")
}), labels=c("A","B"), nrow=2)
```

### 2019-11-27 main_figures/normalization.Rmd
renamed main figures, updated filtering and dimensionality figures

--- 2019-11-09 main_figures/figure3_normalization.Rmd
@@
 theme_set(theme_cowplot(font_size = 11))
+renameNorm <- function(x){
+  .renameHrows(x, f=function(x){
+    x <- gsub("^norm\\.", "", x)
+    x <- gsub("seuratvst","sctransform",x)
+    x <- gsub(".noscale", " (no scaling)", x)
+  })
+}
 ```
@@
 ```{r}
-p1 <- scrna_evalPlot_DR(res, "covarRes", agg.by = "norm", anno_legend=FALSE) +
-    scrna_evalPlot_DR(res, "covarRes", covar = "log10_total_features", agg.by = "norm", anno_legend=FALSE)
-p1 <- grid.grabExpr(draw(p1))
+p1 <- scrna_evalPlot_DR(res, "covarRes", agg.by = "norm", anno_legend=FALSE, reorder_rows = TRUE)
+p1 <- p1 + scrna_evalPlot_DR(res, "covarRes", covar = "log10_total_features", agg.by = "norm", anno_legend=FALSE, reorder_rows=p1)
+p1 <- grid.grabExpr(draw(renameNorm(p1)))
 
 p2 <- grid.grabExpr(draw(
-  scrna_evalPlot_DR(res, what="silhouette", agg.by = "norm", value_format = "", show_heatmap_legend = TRUE)))
+  renameNorm(scrna_evalPlot_DR(res, what="silhouette", agg.by = "norm", value_format = "", show_heatmap_legend = TRUE))))
 
@@
 H <- H + scrna_evalPlot_clust(res, "MI", agg.by = "norm", scale = TRUE, reorder_rows=row.names(H@matrix), anno_legend=FALSE)
-p3 <- grid.grabExpr(draw(H))
+p3 <- grid.grabExpr(draw(renameNorm(H)))
 ```

### 2019-11-27 bmc_article.tex
Updates from Overleaf

@@
 spike-ins (see Table \ref{tab:table1} and \ref{fig:figure1} for a description of the datasets).
+In addition to previously-used benchmark datasets with true cell labels \cite{duoClustering2018,tianMixology2018}, we simulated two datasets with a hierarchical subpopulation structure based on real 10x data using \textit{muscat} [REF].
 Since graph-based clustering  \cite{satijaSeurat2015} was previously shown to consistently perform well across several datasets \cite{duoClustering2018,tianMixology2018}, we used the Seurat pipeline as the starting point to perform an integrated investigation of: 1) doublet identification, 2) cell filtering, 3) normalization, 4) feature selection, 5) dimension reduction, 6) clustering.
@@
 
-We define a pipeline, in R, as a list of functions executed consecutively on the output of the previous one (\ref{fig:figure2}A).
-Each step can include various parameters, all of which must be given alternative values.
-The \textit{runPipeline} command will then proceed in a step-wise fashion to compute only the parts of the pipelines that have not already been computed.
-In addition, each step can output evaluation metrics, which together provide a multi-layered evaluation, where variations in a given parameter can be evaluated using all metrics from this point downward in the pipeline.
+The \textit{pipeComp} package defines a pipeline as, minimally, a list of functions executed consecutively on the output of the previous one (\ref{fig:figure2}A).
+In addition, optional benchmark functions can be set for each step to provide standardized, multi-layered evaluation metrics.
+Given such a \textit{PipelineDefinition} object, a set of alternative parameters (which might include different subroutines) and benchmark datasets, the \textit{runPipeline} function then proceeds through all combinations arguments, avoiding recomputing the same step twice and compiling evaluations on the fly to avoid storing potentially large intermediate data.
+Variations in a given parameter can be evaluated using all metrics from this point downward in the pipeline.
 This is especially important because end-point metrics, such as the adjusted Rand index (ARI) for clustering, are not perfect.
@@
     \centering
-    \includegraphics[width=\textwidth]{{main_figures/figure1_doublets_files/figure-html/figure1-1.png}}
+    \includegraphics[width=\textwidth]{{main_figures/doublets_files/figure-html/figure1-1.png}}
     \caption{\textbf{Identification of doublet cells.
@@
     \centering
-    \includegraphics[width=\textwidth]{{main_figures/figure2_filtering_files/figure-html/figure2-1.png}}
+    \includegraphics[width=\textwidth]{{main_figures/filtering_files/figure-html/figure2-1.png}}
     \caption{Proportion of cells filtered out and accuracy of the cluster (under two normalizations) for different sets of filters (see methods).}
@@
     \centering
-    \includegraphics[width=\textwidth]{{main_figures/figure3_normalization_files/figure-html/figure3-1.png}}
-    \caption{\textbf{Evaluation of normalization procedures:} proportion of the variance in the first 5 components explained by the subpopulations, median ARI, and ARI at the true number of clusters.}
+    \includegraphics[width=\textwidth]{{main_figures/normalization_files/figure-html/figure3-1.png}}
+    \caption{\textbf{Evaluation of normalization procedures.
+A:} Residual correlation of the first principal component with library size (left) and detection rate (right), after accounting for biological differences between subpopulations.
+\textbf{B:} Average silhouette width per true subpopulation, where higher silhouette width means a higher separability.
+\textbf{C:} Clustering accuracy, measured by the ARI at the true number of cluster (left) and by the average mutual information (MI) of the cluster assignment with true subpopulations.}
     \label{fig:figure5}
@@
     \centering
-    \includegraphics[scale=0.6]{{main_figures/figure4_selection_files/figure-html/figure4-1.png}}
+    \includegraphics[scale=0.6]{{main_figures/selection_files/figure-html/figure4-1.png}}
     \caption{\textbf{Evaluation of feature selection methods.
 A:} Ability of different feature ranking methods to capture genes with a high proportion of variance (left) or deviance (right) explained by real subpopulations.
-rankMean and rankMax refer to, respectively, the mean rank and the maximum rank in the following metrics: variance, standardized variance, sctransform, deviance.
-\texbf{B:} Accuracy of clustering (at the right number of clusters) based on the given selection methods and number of genes selected.}
+\textbf{B:} Accuracy of clusterings (at the true number of clusters) when selecting 1000 genes using the given methods.
+Based on standard Seurat normalization (left) or sctransform (right).
+The methods `vst.varExp' and `devianceExplained' are the estimates used in \textbf{A} to evaluate the selection methods, and were included here only for validation purpose.}
     \label{fig:figure6}
@@
 
-We next investigated the impact of the number of dimensions used on the clustering.
+\subsubsection*{Estimating dimension}
+
+We next investigated the impact of the number of dimensions used on the separability of the subpopulations.
+Euclidean distance decreases as the number of non-discriminating dimensions increases, suggesting a tradeoff between the amount of information and the proportion of relevant dimensions.
 Overall, increasing the number of dimensions robustly led to a decrease in the number of clusters, although the effect was milder with data processed through sctransform (Supplementary Figures 14 and 22).
 This tended to be translated to the accuracy of the clustering (Supplementary Figure 23), although in both cases (number of clusters and ARI) the resolution parameter had a much stronger impact.
+We then compared various methods to estimate the intrinsic dimension based on the reduced space (from Seurat weighted PCA).
+As a first approximation of the real dimension, we computed the variance in each principal component which was explained by the subpopulations; this dropped dramatically after the first few components, suggesting an overall low dimensionality (Figure \ref{fig:figure8}A) at least for the purpose of cell type definition.
+We next estimated dimensions using various methods (see methods), some of which are already commonly used for scRNAseq, while others were borrowed from other fields.
+Figure \ref{fig:figure8}B shows the difference between the dimension estimates of these methods and that based on the subpopulations (i.e.
+from Figure \ref{fig:figure8}A).
+Of note, the estimates differ widely in terms of their computing time (Figure \ref{fig:figure8}B), and we saw no relationship between the accuracy of the estimate and the complexity of the method.
+Reasoning that over-estimating dimension is less problematic than under-estimating it, we eliminated methods below the estimate and selected candidate methods for a full analysis of their impact on clustering (Figure \ref{fig:figure8}C-D) in combination with both sctransform or normal Seurat normalization.
+Although most methods performed well, the global maximum likelihood based on translated Poisson Mixture Model \citep{haroTranslated2008}, i.e.
+maxLikGlobal, provided the dimensionality estimate that best separated the subpopulations (Figure \ref{fig:figure8}C) and resulted in the best clustering accuracy (Figure \ref{fig:figure8}D).
+Of note, this method systematically estimated dimensions above those for which a large part of the variance is explained by subpopulations.
 
-\subsubsection*{Rules for selecting components}
+\begin{figure}
+    \centering
+    \includegraphics[width=\textwidth]{{main_figures/dimensionality_files/figure-html/figure6-1.png}}
+    \caption{\textbf{Estimating dimensionality.
+A:} Estimated dimensionality using the proportion of variance in each component explained by subpopulations.
+\textbf{B:} Difference between `real' dimensionality (from \textbf{A}) and various dimensionality estimation methods, along with computing time.
+\textbf{C:} Average silhouette width per subpopulation across a selection of dimensionality estimators, using sctransform or standard seurat normalization.
+\textbf{D:} Clustering accuracy of across the normalization/dimensionality methods.}
+    \label{fig:figure8}
+\end{figure}
 
@@
 
-All analyses were performed through the \href{pipComp}{https://github.com/plger/pipComp} R package, which implements the pipeline framework described here.
-All code to reproduce each figure is available in the \url{https://github.com/markrobinsonuzh/scRNA\_pipelines\_paper} repository, which also includes the basic datasets with a standardized annotation.
+All analyses were performed through the \href{pipeComp}{https://github.com/plger/pipeComp} R package, which implements the pipeline framework described here.
+All code to reproduce the simulations and figures is available in the \url{https://github.com/markrobinsonuzh/scRNA\_pipelines\_paper} repository, which also includes the basic datasets with a standardized annotation.
 
@@
 The code for the glmPCA was obtained from \url{https://github.com/willtownes/scrna2019} (commit 1ddcc30ebb95d083a685f12fe81d35dd1b0cb1b2).
+
+\subsection*{Simulated datasets}
+The `simMix1` dataset was based on the human PBMC CITE-seq data deposited under `GSE100866`.
+Both RNA and ADT count data were downloaded from GEO and processed independently using Seurat.
+We then considered cells that were in the same cluster both in the RNA-based and ADT-based analyses to be real subpopulations, and focused on the 4 most abundant such subpopulations.
+We then performed 3 sampling-based simulations with various degrees of separation using \textit{muscat}, and merged the three simulations.
+The `simMix2` dataset was generated from the mouse brain data published with \textit{muscat} [REF] in a similar fashion.
+The specific code is available on \url{https://github.com/markrobinsonuzh/scRNA\_pipelines\_paper}.
 
@@
   \hline
-dataset & source & protocol & nbCells & subpops & description \\ 
+dataset & source & protocol & description \\ 
   \hline
-Koh & GSE85066 & SMARTer & 531 &   9 & FACS purified H7 hESC in different differention stages \\ 
-  Kumar & GSE60749 & SMARTer & 246 &   3 & Mouse ESC cultured in different conditions \\ 
-  Zhengmix4eq & \citep{duoClustering2018} & 10x & 3994 &   4 & Mixtures of FACS purified PBMCs \\ 
-  Zhengmix4uneq & \citep{duoClustering2018} & 10x & 6498 &   4 & Mixtures of FACS purified PBMCs \\ 
-  Zhengmix8eq & \citep{duoClustering2018} & 10x & 3994 &   8 & Mixtures of FACS purified PBMCs \\ 
-  mixology10x3cl & \cite{tianMixology2018} & 10x & 902 &   3 & Mixture of 3  cancer cell lines from CellBench \\ 
-  mixology10x5cl & \cite{tianMixology2018} & 10x & 3918 &   5 & Mixture of 5 cancer cell lines from CellBench \\ 
+Koh & GSE85066 & SMARTer & FACS purified H7 hESC in different differention stages \\ 
+  Kumar & GSE60749 & SMARTer & Mouse ESC cultured in different conditions \\ 
+  Zhengmix4eq & \citep{duoClustering2018} & 10x & Mixtures of FACS purified PBMCs \\ 
+  Zhengmix4uneq & \citep{duoClustering2018} & 10x & Mixtures of FACS purified PBMCs \\ 
+  Zhengmix8eq & \citep{duoClustering2018} & 10x & Mixtures of FACS purified PBMCs \\ 
+  mixology10x3cl & \cite{tianMixology2018} & 10x & Mixture of 3  cancer cell lines from CellBench \\ 
+  mixology10x5cl & \cite{tianMixology2018} & 10x & Mixture of 5 cancer cell lines from CellBench \\ 
    \hline

### 2019-11-27 main_figures/filtering.Rmd
added subpop exclusion example

@@
   library(ggplot2)
+  library(ggridges)
   library(cowplot)
+  library(SingleCellExperiment)
 })
-theme_set(theme_cowplot())
+theme_set(theme_cowplot(font_size = 12))
 suppressMessages(devtools::load_all("../../pipeComp/"))
@@
 
-```{r figure2, fig.width=9, fig.height=7}
+```{r}
+thresplot <- function(ds, var="log10_total_counts", title="", nmads=2.5, relh=c(1,2), xlims=range(ds[[var]])){
+  noy <- theme( axis.line.y=element_blank(), axis.text.y=element_blank(),
+                  axis.ticks.y=element_blank(), axis.title = element_text(size=11), 
+                  strip.background = element_blank(), legend.position = "none",
+                  axis.text=element_text(size=10), plot.title = element_text(size=12))
+  x <- ds[[var]]
+  xm <- median(x)
+  md <- median(abs(x-xm))
+  mads <- c(xm-nmads*md, xm+nmads*md)
+  p1 <- ggplot(ds, aes_string(x=var, y=1)) + geom_density_ridges() + noy + 
+    ylab("All") + ggtitle(title) +  
+    theme(axis.line.x=element_blank(), axis.text.x=element_blank(), 
+          axis.ticks.x=element_blank(), axis.title.x = element_blank() ) +
+    geom_vline(xintercept=mads, linetype="dashed") + xlim(xlims)
+  p2 <- ggplot(ds, aes_string(x=var, y="phenoid", fill="phenoid", colour="phenoid")) + 
+    noy + geom_density_ridges() + ylab("Subpopulations") +
+    geom_vline(xintercept=mads, linetype="dashed") + xlim(xlims)
+  plot_grid( p1, p2, nrow=2, rel_heights=relh)
+}
+
+koh <- as.data.frame(colData(readRDS("../datasets/Koh.SCE.rds")))
+zhen <- as.data.frame(colData(readRDS("../datasets/Zhengmix8eq.SCE.rds")))
+sim <- as.data.frame(colData(readRDS("../datasets/simMix1.SCE.rds")))
+
+p1 <- plot_grid( thresplot(koh, title="Koh"),
+                 thresplot(zhen, title="Zhengmix8eq", xlims=c(2.65,3.8)),
+                 thresplot(sim, title="simMix1", xlims=c(2.4, 3.82)), nrow=1)
+```
+
+```{r figure2, fig.width=8, fig.height=8.5}
 res <- readRDS("../data/results_filtering.rds")
@@
 
-ggplot(m, aes(maxPCout, F1atK, group=method, colour=filter, shape=doubletRemoval)) + geom_point(size=3) +
+p2 <- ggplot(m, aes(maxPCout, F1atK, group=method, colour=filter, shape=doubletRemoval)) + geom_point(size=3) +
   # geom_line(aes(group=doubletRemoval), colour="lightgrey") + 
-facet_wrap(~dataset, scales="free") + scale_color_manual(values=cols) + xlab("Max proportion of subpopulation excluded") + ylab("mean F1 at true number of clusters")
+  facet_wrap(~dataset, scales="free") + scale_color_manual(values=cols) + 
+  xlab("Max proportion of subpopulation excluded") + ylab("mean F1 at true number of clusters")
+
+plot_grid(p1, p2, nrow=2, rel_heights = c(3,8), labels=LETTERS[1:2])
 ```
 
-Relationship between the maximum subpopulation exclusion rate and the average clustering accuracy per subpopulation across various filtering strategies.
+**A:** Filtering on the basis of distance to the whole distribution can lead to strong bias against certain subpopulations.
+**B:** Relationship between the maximum subpopulation exclusion rate and the average clustering accuracy per subpopulation across various filtering strategies.
 Of note, doublet removal appears to be desirable even when, due to the design, there are no heterotypic doublets in the data.

### 2019-11-27 bmc_article.tex

@@
 Across datasets, cells called as doublets tended to be classified in the wrong cluster more often than other cells (Figure 1D).
-We therefore tested whether this method also improved the accuracy of the clustering across all benchmark datasets, and found that it consistently did (Figure \ref{fig:figure4}).
+We therefore tested whether this method also improved the accuracy of the clustering across all benchmark datasets, and found that it did even when, by design, the data contained no heterotypic doublet (Figure \ref{fig:figure4}).
 
@@
 
-We next investigated the impact of filtering according to various criteria on the accuracy of the clustering.
-We tested each aforementioned cell property, alone and in combinations, filtering above and/or below a certain threshold defined by different numbers of median absolute deviations.
-Reasoning that a bad cell would be bad in several distributions, we also included the optional rule that a cell should be excluded if it is an outlier in at least 2 or 3 distributions.
-
-A first observation was that very stringent filtering (removing up to 20\% of the cells) was not generally associated with an increase in accuracy of the clustering (Supplementary Figure 8A).
-Moreover, stringent filtering tended to be biased towards the exclusion of specific subpopulations (Supplementary Figure 8B).
-Indeed, applying the same filtering criteria on individual clusters of cells (identified through scran's \textit{quickCluster} method) resulted in nearly no cell being filtered out, which suggests that filtering on the global population tends to discard cells of subpopulations with more extreme properties, rather than `bad` cells.
-
-We next investigated the contribution of various criteria individually (Supplementary Figure 9).
-Interestingly, regression analysis suggests that increasing the number of distributions on which a cell must deviate in order to be excluded tended to be positively associated with ARI, while increasing the number of median absolute deviations required to be an outlier as associated with a decrease in ARI.
-We therefore suggest excluding cells that are outliers in more than 1 distribution.
-The impact of filtering on the basis of total counts or feature counts was not always robust, nor always positive, and we could detect no significant impact of filtering on the basis of the proportion of mitochondrial reads - however it must be said that the benchmark datasets count very few cells with a very high proportion of mitochondrial reads (Supplementary Figure 1).
-We therefore suggest to keep this criterion, but in a relatively lenient fashion, and not solely based on the median absolute deviation of the distribution.
-Filtering solely based on deviations from the distribution might be inadequate when the distribution is tight; in our experience, the proportion of mitochondrial counts in otherwise `healthy` cells can vary up to 8-12\% depending on the tissue.
-We therefore recommend using a combination of fixed and deviation based threshold, e.g.
-x $>$ 0.08 and x $>$ 2.5*MADs.
+We next investigated the impact of filtering according to various criteria (see methods).
+An important risk of excluding cells on the basis of their distance from the whole distribution of cells on some properties (e.g.
+library size) is that these properties tend to have different distributions across subpopulations.
+As a result, thresholds in terms of number of median absolute deviations (MADs) from the whole distribution can lead to strong biases against certain subpopulations (Figure \label{fig:figure4}A).
+We therefore examined the tradeoff between the increased accuracy of filtering and the maximum proportion of cells excluded per subpopulation (Figure \label{fig:figure4}B and Supplementary Figure 8).
+Since filtering changes the relative abundance of the different subpopulations, global clustering accuracy metrics such as ARI are not appropriate here, and we therefore calculated the per-subpopulation precision and recall using the Hungarian algorithm, and monitored their mean F1 score.
+A first observation was that, although the stringent filtering tended to be associated with an increase in accuracy, it could also become deleterious, and most of the benefits could be achieved without very stringent filtering and minimizing subpopulation bias.
+Applying the same filtering criteria on individual clusters of cells (identified through scran's \textit{quickCluster} method) resulted in nearly no cell being filtered out, which suggests that filtering on the global population tends to discard cells of subpopulations with more extreme properties, rather than `bad` cells.
+Finally, by changing the distributions, the doublet removal step in conjunction with filtering sometimes resulted in a net decrease in the proportion of excluded cells while retaining or improving accuracy.
+We therefore recommend the use of doublet removal followed by relatively mild filtering, such as is implemented in our `default' filtering (see methods).
 
@@
     \includegraphics[width=\textwidth]{{main_figures/filtering_files/figure-html/figure2-1.png}}
-    \caption{Proportion of cells filtered out and accuracy of the cluster (under two normalizations) for different sets of filters (see methods).}
+    \caption{\textbf{A:} Filtering on the basis of distance to the whole distribution can lead to strong bias against certain subpopulations.
+The dashed line indicates a threshold of 2.5 median absolute deviations (MADs) from the median of the overall population.
+\textbf{B:} Relationship between the maximum subpopulation exclusion rate and the average clustering accuracy per subpopulation across various filtering strategies.
+Of note, doublet removal appears to be desirable even when, due to the design, there are no heterotypic doublets in the data.
+The PCA methods refer to multivariate outlier detected as implemented in `scater` (see methods for details).}
     \label{fig:figure4}
 \end{figure}
-
-On the basis of these observations, we defined three sets of criteria as lenient, default and stringent filtering (see methods), and investigated them further.
-We found that filtering had a much milder impact than doublet removal (Figure \ref{fig:figure4}), and that more stringent filtering tended to result in higher silhouette width, in a higher proportion of the variance in the first components explained by subpopulations (Supplementary Figure 10), and often in a higher adjusted Rand index (ARI) of the clustering at the right number of clusters (Figure \ref{fig:figure4}).
-However, part of the gain is to be ascribed to the removal of cells from subpopulations (especially in the Zheng datasets) that are difficult to distinguish, and for which more stringent clustering did not improve silhouette width (Supplementary Figure 10A).
-For these reasons, we therefore recommend using a more lenient filter (e.g.
-lenient or default filter described here) or a cluster-based filter to avoid most of the subpopulation bias while excluding strong outliers.
 
@@
 Mitochondrial reads have been associated with cell degradation, and it has been observed that much of scRNAseq clustering is determined by ribosomal genes [REF].
-We therefore considered whether excluding one category of features or the other, or using only protein-coding genes, had impact on the ability to distinguish subpopulations (Supplementary Figure 11).
+We therefore considered whether excluding one category of features or the other, or using only protein-coding genes, had impact on the ability to distinguish subpopulations (Supplementary Figure 9).
 Removal of ribosomal genes robustly reduced the quality of the clustering, suggesting that they represent real biological differences between subpopulations.
@@
 Beside the standard log-normalization included in Seurat, we tested scran's pooling-based normalization \citep{lunPooling2016}, sctransform's variance-stabilizing transformation \citep{hafemeisterSCtransform2019}, and normalization based on stable genes \citep{linStableGenes2018, deekeStablyExpressed2018}.
-In addition to log-normalization, the standard Seurat clustering pipeline scales features so that the PCA is not dominated by a small subset of the features, and we therefore included versions of the different normalization, with or without a subsequent scaling (except sctransform's variance-stabilizing transformation, given that, conceptually, it involves something analogous to scaling).
-Seurat's scaling function includes the option to regress out the effect of certain covariates, and we therefore tested its use with the proportion of mitochondrial counts and the number of detected features.
-Finally, it has been proposed that the use of stable genes, and in particular cytosolic ribosomal genes, can be used to normalize scRNAseq; we therefore evaluated a simple linear normalization based on the sum of these genes, as well as nuclear genes.
+In addition to log-normalization, the standard Seurat clustering pipeline performs per-feature unit-variance scaling so that the PCA is not too dominated by highly-expressed features, and we therefore included versions of the different normalization, with or without a subsequent scaling (sctransform's variance-stabilizing transformation involves something analogous to scaling).
+Seurat's scaling function also includes the option to regress out the effect of certain covariates, and we therefore tested its use with the proportion of mitochondrial counts and the number of detected features.
+Finally, it has been proposed that the use of stable genes, in particular cytosolic ribosomal genes, can be used to normalize scRNAseq; we therefore evaluated a simple linear normalization based on the sum of these genes, as well as nuclear genes.
 
 An important motivation of sctransform was the observation that, even after normalization, the first principal components of various datasets tended to correlate with library size, suggesting an inadequate normalization \citep{hafemeisterSCtransform2019}.
-We therefore first assessed to what extent the first principal component still retained a correlation with the library size and number of detected features, removing the confounding covariation with the subpopulation (Supplementary Figure 12A).
-The simple step of scaling tended to remove much of the correlation with these features, and in the absence of scaling, standard Seurat normalization resulted in fairly high correlation.
+We therefore first assessed to what extent the first principal component still retained a correlation with the library size and number of detected features, removing the confounding covariation with the subpopulation (Figure \label{fig:figure5}A).
+The simple step of scaling tended to remove much of the correlation with these features, and in the absence of scaling, standard Seurat normalization resulted in fairly high correlation with technical covariates.
 sctransform had the lowest correlation, but most methods (including normalization based on stable genes) were able to remove most of the correlation when combined with scaling.
@@
 
-We next investigated the impact of normalization on the ability to distinguish subpopulations, which was much greater than that of filtering (Supplementary Figure 13).
-For many clusters, scaling tended to reduce both the average silhouette width (Supplementary Figure 12B) and the proportion of variance in the top principal components explained by subpopulations (Figure \ref{fig:figure5} and Supplementary Figure 13), and was associated with a poorer adjusted Rand index (ARI) of the resulting clustering (Figure \ref{fig:figure5}).
+We next investigated the impact of normalization on the separability of the subpopulations (Figure \label{fig:figure5}B-C), and found most methods (including no normalization at all) to perform fairly well.
+Since clustering accuracy metrics such as the ARI are very strongly influenced by the number of clusters, we complemented it with silhouette width and mutual information.
+Scaling tended to reduce the average silhouette width of some subpopulations and to increase that of some less distinguishable ones, and was generally but not always beneficial on the accuracy of the final clustering.
 Regressing out covariates systematically gave poorer performance on all metrics.
-sctransform produced the highest cluster average silhouette width (Supplementary Figure 12B) as well as ARI (Figure \ref{fig:figure5}), but not always the highest mean proportion of variance in the top components explained by subpopulations (Figure \ref{fig:figure5} and Supplementary Figure 13).
-Of note, even though sctransform is meant to be applied to data with unique molecular identifiers (UMI), it also performed fairly well with smartseq datasets.
-Interestingly, some clusters tended to be less well distinguished with sctransform than with non-scaled (and in some cases non-normalized) methods, suggesting that some procedures emphasize differences more useful to track some subpopulations than others, and hence the possibility that further developments in methods could improve this step.
+sctransform systematically outperformed other methods, and even though it was developed to be applied to data with unique molecular identifiers (UMI), it also performed fairly well with smartseq datasets.
 
@@
 
-Finally, although there were differences in specific cases, we observed no robust tendency of certain normalization methods towards over- or under-estimating the number of clusters when using with Seurat clustering (Supplementary Figure 14).
+Finally, although some normalization methods had a tendency to lead to a higher (e.g.
+sctransform) or lower (e.g.
+stable genes) number of clusters, the effect was very mild and not entirely systematic, and we observed no major tendency towards over- or under-estimating the number of clusters (Supplementary Figure 10).
 
@@
 
-Reasoning that a selection method should ideally select genes whose variability is more between subpopulations than within, we first assessed whether each selection method selected genes with a high proportion of variance or deviance explained by (real) subpopulations (the latter estimated through edgeR).
-Estimates of variance explained by subpopulations based on a standard Seurat normalization or on sctransform data were highly correlated (Supplementary Figure 14A), and there was also a good agreement with deviance explained, although lowly-expressed genes in particular could have a high deviance explained without having much of their variance explained by subpopulations (Supplementary Figure 15B-D).
-We therefore compared, using both of these estimates, the proportion of the cumulative variance/deviance explained by the top X genes that could be retrieved by each gene ranking method (Supplementary Figures 16-17).
-We first focused on the first 1000 genes to highlight differences between methods, although the differences tended to decrease with a higher number of genes selected (Figure \ref{fig:figure6}A).
-A first observation was that standardized measures of variability were systematically worse than their non-standardized  counterparts in selecting genes with a high proportion of variance explained by subpopulations, while instead proving most of the time superior in selecting genes with a high deviance explained by subpopulations (Figure \ref{fig:figure6} and Supplementary Figures 16-17).
-Deviance proved the method of choice to maximize the former (with mere expression level proving surprisingly good), but did not perform so well to select genes with a high deviance explained.
+Reasoning that a selection method should ideally select genes whose variability is more between subpopulations than within, we first assessed whether each selection method selected genes with a high proportion of variance or deviance explained by (real) subpopulations.
+Estimates of variance explained by subpopulations based on a standard Seurat normalization or on sctransform data were highly correlated (Supplementary Figure 11A), and there was also a good agreement with deviance explained, although lowly-expressed genes in particular could have a high deviance explained without having much of their variance explained by subpopulations (Supplementary Figure 11B-D).
+We therefore compared, using both of these estimates, the proportion of the cumulative variance/deviance explained by the top X genes that could be retrieved by each gene ranking method (Supplementary Figures 12-13).
+We first focused on the first 1000 genes to highlight differences between methods, although the differences tended to decrease with a higher number of genes selected (Supplementary Figures 12-14).
+A first observation was that standardized measures of variability were systematically worse than their non-standardized  counterparts in selecting genes with a high proportion of variance explained by subpopulations, while instead proving most of the time superior in selecting genes with a high deviance explained by subpopulations (Figure \ref{fig:figure6}A and Supplementary Figures 12-13).
+Deviance proved the method of choice to maximize the former (with mere expression level proving surprisingly good), but surprisingly did not perform so well to select genes with a high deviance explained.
 
@@
 
-We next investigated whether a selection based on these methods impacted the proportion of variance explained by subpopulations in the first principal components (Supplementary Figure 18).
-To validate the previous assay, included a selection based on the genes whose proportion variance or deviance explained was highest.
-Interestingly, while these selections were on average the top-ranking methods, they were not systematically so for all dataset.
-
-The main observations were reflected in the ARI of the resulting clustering (Figure 4B): non-standardized measures of variability, including mere expression level, tended to outperform more complex metrics.
-In general, we found deviance and Seurat's un-standardized estimate of variance to provide the best results across datasets and normalization methods.
-Increasing the number of features selected also systematically led to an increase in the accuracy of the clustering.
+We next investigated whether a selection based on these methods impacted the clustering accuracy (Figure \ref{fig:figure6}A).
+To validate the previous assay, we included a selection based on the genes whose proportion variance or deviance explained was highest.
+Interestingly, while these selections were on average the top-ranking methods, they were not systematically best for all dataset.
+The previous observations were reflected in the ARI of the resulting clustering (Figure 4B): non-standardized measures of variability, including mere expression level, tended to outperform more complex metrics.
+In general, we found deviance and un-standardized estimate of variance to provide the best results across datasets and normalization methods.
+Increasing the number of features selected also systematically led to an increase in the accuracy of the clustering, typically plateauing after 4000 features (Supplementary Figure 14).
 
@@
 The impact of the dimensionality reduction method was far greater than that of normalization or feature selection (e.g.
-Supplementary Figure 19).
-Although the first components identified by GLM-PCA tended to have a higher proportion of their variance explained by real subpopulations and increase the average silhouette width of already well-defined subpopulations, overall Seurat's PCA procedure proved superior on all metrics (Figure \ref{fig:figure7} and Supplementary Figures 19-21).
+Supplementary Figure 15).
+Although the first components identified by GLM-PCA tended to have a higher proportion of their variance explained by real subpopulations and increase the average silhouette width of already well-defined subpopulations, overall Seurat's PCA procedure proved superior on all metrics (Figure \ref{fig:figure7} and Supplementary Figures 15-17).
 In particular, weighing the components had a strong positive impact on silhouette widths and resulting ARI.
@@
     \centering
-    \includegraphics[width=\textwidth]{{main_figures/figure5_dr_files/figure-html/unnamed-chunk-2-1.png}}
+    \includegraphics[width=\textwidth]{{main_figures/dr_files/figure-html/unnamed-chunk-2-1.png}}
     \caption{Evaluation of the dimensionality reduction strategies.}

### 2019-11-29 main_figures/dr.Rmd
updated dimensionality reduction comparison

@@
 source("../misc_functions.R")
-suppressMessages(devtools::load_all("../../pipComp/"))
-data(datasets)
-theme_set(theme_cowplot(font_size = 12))
+suppressMessages(devtools::load_all("../../pipeComp/"))
 ```
 
-```{r, fig.width=8, fig.height=4}
-res <- readRDS("../data/dr_intermediateSummary.rds")
-sw <- res$dimreduction$clust.avg.silwidth[[1]]
-sw <- sw[,grep("seuratvst",colnames(sw))]
-sw <- sw[,grep("deviance",colnames(sw))]
-colnames(sw) <- reduceMethodNames(colnames(sw),6)
-sw <- matrix(unlist(sw),ncol=ncol(sw), dimnames=dimnames(sw))
-ag <- aggregate(sw, by=list(reduceMethodNames(row.names(sw),1)), FUN=mean)
-row.names(ag) <- ag[,1]; ag[,1] <- NULL
-sw <- t(ag)
+```{r dimred, fig.width=12, fig.height=6}
+res <- readRDS("../data/results_dimred.rds")
+x <- res$dimreduction$clust.avg.silwidth
+x <- x[grep("norm\\.seurat\\.noscale.*denoise", row.names(x), invert=TRUE),]
+x <- x[order(colSums(apply(x,1,prob=c(0.05,0.5),FUN=quantile))+
+                         rowMeans(x), decreasing=TRUE),order(colMeans(x), decreasing=TRUE)]
+row.names(x) <- gsub("norm\\.","",apply(parsePipNames(row.names(x))[,c("norm","dr")],1,collapse=" + ",FUN=paste))
+row.names(x) <- gsub("seuratvst","sctransform",row.names(x))
+h1 <- Heatmap(x, cluster_columns = FALSE, cluster_rows = FALSE, 
+        name="silhouette width", col=.silScale(x), show_column_names = FALSE,
+        column_title = "Subpopulation mean silhouette width", row_names_max_width = unit(9,"cm"),
+        bottom_annotation=.ds_anno(colnames(x)), heatmap_legend_param = list(direction="horizontal", legend_width=unit(3, "cm")))
+p1 <- grid.grabExpr(draw(h1, heatmap_legend_side="bottom"))
 
-r2 <- res$dimreduction$PC.R2
-r2 <- r2[grep("seuratvst",r2$method),]
-r2 <- r2[grep("deviance",r2$method),]
-r2$dr <- reduceMethodNames(r2$method, 6)
-r2 <- r2[which(r2$dim <= 5),]
-r3 <- cast2(r2, formula=dr~dataset, value.var="value")
+res$clustering <- res$clustering[grep("norm\\.seurat\\.noscale.*denoise", row.names(res$clustering), invert=TRUE),]
+h <- scrna_evalPlot_clust(res, what="MI", agg.by=c("norm","dr"), scale=TRUE, anno_legend = FALSE)
+h <- h + scrna_evalPlot_clust(res, what="mean_F1", agg.by=c("norm","dr"), scale=TRUE, reorder_rows = h, anno_legend = FALSE) +
+  scrna_evalPlot_clust(res, what="min_pr", agg.by=c("norm","dr"), scale=TRUE, reorder_rows = h, anno_legend = FALSE) +
+  scrna_evalPlot_clustAtTrueK(res, agg.by=c("norm","dr"), reorder_rows = h, anno_legend = FALSE)
+h <- .renameHrows(h, function(x) gsub("; dr=", " + ", gsub("seuratvst", "sctransform", x)))
 
-res <- readRDS("../data/dr_endSummary.rds")
-res2 <- res[which(abs(res$nbClusters-datasets[res$dataset,"subpopulations"])<=1 & 
-                    res$norm=="norm.seuratvst"),]
-m.ari <- cast2(res2, formula=dr~dataset, fun.aggregate=median, value.var="ARI")
+p2 <- grid.grabExpr(draw(h))
 
-ll <- list( "mean silhouette width"=sw,
-            "PC1-5 var explained"=r3,
-            "median ARI"=m.ari)
-ll <- lapply(ll, FUN=function(x){
-  row.names(x) <- gsub(".noW weighted", "", paste(row.names(x), "weighted"), fixed=T)
-  x
-})
-chm(ll, scale="column", scaleTitle="z-score", sameScale=TRUE, value_format="%.2f", row_sortFn=function(x){ median(x,na.rm=T) }, cluster_columns=FALSE)
+plot_grid(p1,p2, nrow=2, labels=LETTERS[1:2])
 ```
 
-Mean of subpopulation average silhouette width (left), mean proportion of the variance in the first 5 components explained by real subpopulations (center), and median ARI of the resulting clustering (right).
+**A:** Average silhouette width per subpopulation according to various normalization and dimension reductions.
+**B:** Clustering accuracy , mean proportion of the variance in the first 5 components explained by real subpopulations (center), and median ARI of the resulting clustering (right).

### 2019-11-29 bmc_article.tex
Updates from Overleaf

@@
 
-We next investigated the step of dimensionality reduction, comparing in particular Seurat's PCA, GLM-PCA \citep{townesGlmpca2019}, and a two-step PCA procedure in which cells are first clustered in rough clusters according to a standard PCA, a new PCA is performed on each, and the main components of the two are put together (see methods).
+We next investigated the step of dimensionality reduction, comparing in particular Seurat's PCA, scran's denoise PCA, and GLM-PCA \citep{townesGlmpca2019}, combined (where relevant) with sctransform normalization.
 Seurat's default PCA weighs the cell embeddings by the variance of each component, so we also tested the impact of this weighting with each method.
@@
     \centering
-    \includegraphics[width=\textwidth]{{main_figures/dr_files/figure-html/unnamed-chunk-2-1.png}}
-    \caption{Evaluation of the dimensionality reduction strategies.}
+    \includegraphics[width=\textwidth]{{main_figures/dr_files/figure-html/dimred-1.png}}
+    \caption{\textbf{Evaluation of the dimensionality reduction strategies.
+A:} Average silhouette width per subpopulation according to various normalization and dimension reductions.
+\textbf{B:} Clustering accuracy , mean proportion of the variance in the first 5 components explained by real subpopulations (center), and median ARI of the resulting clustering (right).}
     \label{fig:figure7}
@@
 
-\subsubsection*{Estimating dimension}
+\subsubsection*{Estimating the number of dimensions}
 

### 2019-12-02 bmc_article.tex
Updates from Overleaf

@@
 
-Although we use some datasets based on other protocols, our focus is especially on droplet-based datasets and on methods that do not include exogenous control RNA, i.e.
-spike-ins (see Table \ref{tab:table1} and \ref{fig:figure1} for a description of the datasets).
+Although we use some datasets based on other protocols, our focus is especially on droplet-based datasets that do not include exogenous control RNA (i.e.
+spike-ins); see Table \ref{tab:table1} and \ref{fig:figure1} for a description of the datasets.
 In addition to previously-used benchmark datasets with true cell labels \cite{duoClustering2018,tianMixology2018}, we simulated two datasets with a hierarchical subpopulation structure based on real 10x data using \textit{muscat} [REF].
-Since graph-based clustering  \cite{satijaSeurat2015} was previously shown to consistently perform well across several datasets \cite{duoClustering2018,tianMixology2018}, we used the Seurat pipeline as the starting point to perform an integrated investigation of: 1) doublet identification, 2) cell filtering, 3) normalization, 4) feature selection, 5) dimension reduction, 6) clustering.
+Since graph-based clustering  \cite{satijaSeurat2015} was previously shown to consistently perform well across several datasets \cite{duoClustering2018,tianMixology2018}, we used the \texttt{Seurat} pipeline as the starting point to perform an integrated investigation of: 1) doublet identification, 2) cell filtering, 3) normalization, 4) feature selection, 5) dimension reduction, 6) clustering.
 We compare not only competing approaches, but also more fine-grained parameter and variations on common methods.
@@
 
-The \textit{pipeComp} package defines a pipeline as, minimally, a list of functions executed consecutively on the output of the previous one (\ref{fig:figure2}A).
+The \texttt{pipeComp} package defines a pipeline as, minimally, a list of functions executed consecutively on the output of the previous one (\ref{fig:figure2}A).
 In addition, optional benchmark functions can be set for each step to provide standardized, multi-layered evaluation metrics.
-Given such a \textit{PipelineDefinition} object, a set of alternative parameters (which might include different subroutines) and benchmark datasets, the \textit{runPipeline} function then proceeds through all combinations arguments, avoiding recomputing the same step twice and compiling evaluations on the fly to avoid storing potentially large intermediate data.
+Given such a \texttt{PipelineDefinition} object, a set of alternative parameters (which might include different subroutines) and benchmark datasets, the \texttt{runPipeline} function then proceeds through all combinations arguments, avoiding recomputing the same step twice and compiling evaluations on the fly to avoid storing potentially large intermediate data.
 Variations in a given parameter can be evaluated using all metrics from this point downward in the pipeline.
@@
 We therefore first evaluated the capacity of these methods to detect doublets using the two 10x datasets produced by \citep{tianMixology2018} with cells of different genetic identity, and for which therefore SNP information can be used as ground truth.
-We tested DoubletFinder [REF] and scran's doubletCells [REF], both of which use similarity to artificial doublets, and scds [REF], which relies on a combination of coexpression and binary classification.
-DoubletFinder integrates a thresholding based on the proportion of expected doublets, while scran and scds return scores which must be manually thresholded.
+We tested \texttt{DoubletFinder} [REF] and \texttt{scran}'s \texttt{doubletCells} [REF], both of which use similarity to artificial doublets, and \texttt{scds} [REF], which relies on a combination of coexpression and binary classification.
+\texttt{DoubletFinder} integrates a thresholding based on the proportion of expected doublets, while \texttt{scran} and \texttt{scds} return scores which must be manually thresholded.
 In these cases, we ensured that the right number of cells would be called doublets.
 
-In addition to these methods, we reasoned that an approach such as DoubletFinder could be considerably sped up by being applied directly on counts and by using a pre-clustering to create neotypic doublets (i.e.
+In addition to these methods, we reasoned that an approach such as \texttt{DoubletFinder} could be simplified by being applied directly on counts and by using a pre-clustering to create neotypic doublets (i.e.
 doublets from different cell types) more efficiently.
@@
 While most methods accurately identified the doublets in the 3 cell lines dataset (mixology10x3cl), the 5 lines dataset (mixology10x5cl) proved more difficult (Figure \ref{fig:figure3}B).
-Our method (scDblFinder) achieved the best accuracy while running in a very reasonable time (Figure \ref{fig:figure3}).
+Our method (\texttt{scDblFinder}) achieved the best accuracy while running in a very reasonable time (Figure \ref{fig:figure3}).
 Across datasets, cells called as doublets tended to be classified in the wrong cluster more often than other cells (Figure 1D).
@@
 This has for instance been demonstrated for cells that have a high content of mitochondrial reads, often as a result of cell degradation [REF].
-A common practice is to exclude cells that differ considerably from most other cells on the basis of some such properties, for instance through the `isOutlier` function of `scater` which measures, for a given control property, the number of median absolute deviation of each cell from the mean of all cells.
+A common practice is to exclude cells that differ considerably from most other cells on the basis of some such properties, for instance through the \texttt{isOutlier} function of \texttt{scater} which measures, for a given control property, the number of median absolute deviation of each cell from the mean of all cells.
 Supplementary Figure 1 shows the distributions of some of the typical cell properties commonly used.
@@
 A first observation was that, although the stringent filtering tended to be associated with an increase in accuracy, it could also become deleterious, and most of the benefits could be achieved without very stringent filtering and minimizing subpopulation bias.
-Applying the same filtering criteria on individual clusters of cells (identified through scran's \textit{quickCluster} method) resulted in nearly no cell being filtered out, which suggests that filtering on the global population tends to discard cells of subpopulations with more extreme properties, rather than `bad` cells.
+Applying the same filtering criteria on individual clusters of cells (identified through scran's \texttt{quickCluster} method) resulted in nearly no cell being filtered out, which suggests that filtering on the global population tends to discard cells of subpopulations with more extreme properties, rather than `bad` cells.
 Finally, by changing the distributions, the doublet removal step in conjunction with filtering sometimes resulted in a net decrease in the proportion of excluded cells while retaining or improving accuracy.
@@
 Of note, doublet removal appears to be desirable even when, due to the design, there are no heterotypic doublets in the data.
-The PCA methods refer to multivariate outlier detected as implemented in `scater` (see methods for details).}
+The PCA methods refer to multivariate outlier detected as implemented in \texttt{scater} (see methods for details).}
     \label{fig:figure4}
@@
 We next investigated the impact of different normalization strategies.
-Beside the standard log-normalization included in Seurat, we tested scran's pooling-based normalization \citep{lunPooling2016}, sctransform's variance-stabilizing transformation \citep{hafemeisterSCtransform2019}, and normalization based on stable genes \citep{linStableGenes2018, deekeStablyExpressed2018}.
-In addition to log-normalization, the standard Seurat clustering pipeline performs per-feature unit-variance scaling so that the PCA is not too dominated by highly-expressed features, and we therefore included versions of the different normalization, with or without a subsequent scaling (sctransform's variance-stabilizing transformation involves something analogous to scaling).
-Seurat's scaling function also includes the option to regress out the effect of certain covariates, and we therefore tested its use with the proportion of mitochondrial counts and the number of detected features.
+Beside the standard log-normalization included in \texttt{Seurat}, we tested \texttt{scran}'s pooling-based normalization \citep{lunPooling2016}, \texttt{sctransform}'s variance-stabilizing transformation \citep{hafemeisterSCtransform2019}, and normalization based on stable genes \citep{linStableGenes2018, deekeStablyExpressed2018}.
+In addition to log-normalization, the standard \texttt{Seurat} clustering pipeline performs per-feature unit-variance scaling so that the PCA is not too dominated by highly-expressed features, and we therefore included versions of the different normalization, with or without a subsequent scaling (\texttt{sctransform}'s variance-stabilizing transformation involves something analogous to scaling).
+\texttt{Seurat}'s scaling function also includes the option to regress out the effect of certain covariates, and we therefore tested its use with the proportion of mitochondrial counts and the number of detected features.
 Finally, it has been proposed that the use of stable genes, in particular cytosolic ribosomal genes, can be used to normalize scRNAseq; we therefore evaluated a simple linear normalization based on the sum of these genes, as well as nuclear genes.
@@
 We therefore first assessed to what extent the first principal component still retained a correlation with the library size and number of detected features, removing the confounding covariation with the subpopulation (Figure \label{fig:figure5}A).
-The simple step of scaling tended to remove much of the correlation with these features, and in the absence of scaling, standard Seurat normalization resulted in fairly high correlation with technical covariates.
-sctransform had the lowest correlation, but most methods (including normalization based on stable genes) were able to remove most of the correlation when combined with scaling.
-The only exception is Seurat normalization with scaling regressing out the number of features, which surprisingly increased the correlation.
+The simple step of scaling tended to remove much of the correlation with these features, and in the absence of scaling, standard \texttt{Seurat} normalization resulted in fairly high correlation with technical covariates.
+\texttt{sctransform} had the lowest correlation, but most methods (including normalization based on stable genes) were able to remove most of the correlation when combined with scaling.
+The only exception is \texttt{Seurat} normalization with scaling regressing out the number of features, which surprisingly increased the correlation.
 
@@
 Finally, although some normalization methods had a tendency to lead to a higher (e.g.
-sctransform) or lower (e.g.
+\texttt{sctransform}) or lower (e.g.
 stable genes) number of clusters, the effect was very mild and not entirely systematic, and we observed no major tendency towards over- or under-estimating the number of clusters (Supplementary Figure 10).
@@
 A standard clustering pipeline typically involves a step of selecting highly-variable genes, which is complicated by the digital nature and the mean-variance relationship of RNAseq.
-Seurat's earlier approaches involved the use of dispersion estimates standardized for the mean expression levels, while more recent versions rely of a different measure of variance, again standardized.
+\texttt{Seurat}'s earlier approaches involved the use of dispersion estimates standardized for the mean expression levels, while more recent versions rely of a different measure of variance, again standardized.
 Of note, while adjusting for the mean-variance relationship removes much of the bias towards highly-expressed genes, it is plausible that this bias reflects biological relevance and hence is actually helpful in classifying cell types.
-Another common practice for selecting features is to use those with the highest mean expression, and recently \citep{townesGlmpca2019} instead suggested to use deviance, while sctransform provides its own ordering the genes by transformed variance.
+Another common practice for selecting features is to use those with the highest mean expression, and recently \citep{townesGlmpca2019} instead suggested to use deviance, while \texttt{sctransform} provides its own ordering the genes by transformed variance.
 We therefore compared these methods.
@@
 Reasoning that a selection method should ideally select genes whose variability is more between subpopulations than within, we first assessed whether each selection method selected genes with a high proportion of variance or deviance explained by (real) subpopulations.
-Estimates of variance explained by subpopulations based on a standard Seurat normalization or on sctransform data were highly correlated (Supplementary Figure 11A), and there was also a good agreement with deviance explained, although lowly-expressed genes in particular could have a high deviance explained without having much of their variance explained by subpopulations (Supplementary Figure 11B-D).
+Estimates of variance explained by subpopulations based on a standard \texttt{Seurat} normalization or on sctransform data were highly correlated (Supplementary Figure 11A), and there was also a good agreement with deviance explained, although lowly-expressed genes in particular could have a high deviance explained without having much of their variance explained by subpopulations (Supplementary Figure 11B-D).
 We therefore compared, using both of these estimates, the proportion of the cumulative variance/deviance explained by the top X genes that could be retrieved by each gene ranking method (Supplementary Figures 12-13).
@@
 \textbf{B:} Accuracy of clusterings (at the true number of clusters) when selecting 1000 genes using the given methods.
-Based on standard Seurat normalization (left) or sctransform (right).
+Based on standard \texttt{Seurat} normalization (left) or \texttt{sctransform} (right).
 The methods `vst.varExp' and `devianceExplained' are the estimates used in \textbf{A} to evaluate the selection methods, and were included here only for validation purpose.}
@@
 
-We next investigated the step of dimensionality reduction, comparing in particular Seurat's PCA, scran's denoise PCA, and GLM-PCA \citep{townesGlmpca2019}, combined (where relevant) with sctransform normalization.
-Seurat's default PCA weighs the cell embeddings by the variance of each component, so we also tested the impact of this weighting with each method.
+We next investigated the step of dimensionality reduction, comparing in particular \texttt{Seurat}'s PCA, \texttt{scran}'s \texttt{denoise PCA}, and GLM-PCA \citep{townesGlmpca2019}, combined (where relevant) with sctransform normalization.
+\texttt{Seurat}'s default PCA weighs the cell embeddings by the variance of each component, so we also tested the impact of this weighting with each method.
 
@@
 Euclidean distance decreases as the number of non-discriminating dimensions increases, suggesting a tradeoff between the amount of information and the proportion of relevant dimensions.
-Overall, increasing the number of dimensions robustly led to a decrease in the number of clusters, although the effect was milder with data processed through sctransform (Supplementary Figures 14 and 22).
+Overall, increasing the number of dimensions robustly led to a decrease in the number of clusters, although the effect was milder with data processed through \texttt{sctransform} (Supplementary Figures 14 and 22).
 This tended to be translated to the accuracy of the clustering (Supplementary Figure 23), although in both cases (number of clusters and ARI) the resolution parameter had a much stronger impact.
@@
 maxLikGlobal, provided the dimensionality estimate that best separated the subpopulations (Figure \ref{fig:figure8}C) and resulted in the best clustering accuracy (Figure \ref{fig:figure8}D).
-Of note, this method systematically estimated dimensions above those for which a large part of the variance is explained by subpopulations.
+Of note, this method systematically estimated dimensions above those for which a large part of the variance is explained by subpopulations, suggesting that these components help classification despite being individually uninformative.
 
@@
 
+We finally moved to the clustering step.
+Given previous work on the topic\citep{duoClustering2018} and the success of graph-based clustering methods for scRNA-seq, we simply compared Seurat to two \texttt{scran} SNN-based clustering approaches respectively using random walks (`walktrap') or optimizing the modularity score (`fast greedy'), with standard normalization or after \texttt{sctransform}.
+
+(Interpretation of the results... walktrap top?)
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\textwidth]{{main_figures/clustering_files/figure-html/clustering-1.png}}
+    \caption{\textbf{Evaluation of clustering methods.}}
+    \label{fig:figure9}
+\end{figure}
+
+\subsection*{Further extensions to the pipeline: imputation/denoising}
+
+The basic pipeline used here can be extended by adding steps and using the same evaluation metrics.
+To demonstrate this, we evaluated various imputation or denoising techniques on their impact on classification.
+Observing that all methods performed equally well or better on normalized data, we applied them after filtering and normalization, but before scaling and reduction.
+While all methods improved the separability of some subpopulations, no method had a positive impact on the silhouette width of all populations (Figure \ref{fig:figure10}A).
+Similarly, not all methods performed as well across all datasets in terms of clustering accuracy (Figure \ref{fig:figure10}A).
+As expected, 10X datasets which have more cells at a lower coverage benefited more from these methods, which were instead rather deleterious in smart-seq datasets.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\textwidth]{{main_figures/imputation_files/figure-html/imputation-1.png}}
+    \caption{\textbf{Imputation/denoising methods.
+A:} Average silhouette width per subpopulation across various denoising/imputation methods.
+\textbf{B:} Clustering accuracy across denoising/imputation methods.}
+    \label{fig:figure10}
+\end{figure}
+
 \section*{Discussion}
@@
 For study protocols and methodology manuscripts this section should include a discussion of any practical or operational issues involved in performing the study and any issues not covered in other sections.
+
+\subsection*{Concrete recommendations}
+
+\subsection*{Limitations and open questions}
 
@@
 \subsection*{Software and package versions}
-Analyses were performed in R 3.6.0 (bioconductor 3.9), and the following packages were installed from github repositories: Seurat (version 3.0.0), sctransform (0.2.0), DoubletFinder (2.0.1), scds (1.0.0), pipComp (0.2).
+Analyses were performed in R 3.6.0 (bioconductor 3.9), and the following packages were installed from github repositories: \texttt{Seurat} (version 3.0.0), \texttt{sctransform} (0.2.0), \texttt{DoubletFinder} (2.0.1), \texttt{scds} (1.0.0).
 The code for the glmPCA was obtained from \url{https://github.com/willtownes/scrna2019} (commit 1ddcc30ebb95d083a685f12fe81d35dd1b0cb1b2).
@@
 \subsection*{Simulated datasets}
-The `simMix1` dataset was based on the human PBMC CITE-seq data deposited under `GSE100866`.
+The `simMix1' dataset was based on the human PBMC CITE-seq data deposited under `GSE100866'.
 Both RNA and ADT count data were downloaded from GEO and processed independently using Seurat.
@@
 We then performed 3 sampling-based simulations with various degrees of separation using \textit{muscat}, and merged the three simulations.
-The `simMix2` dataset was generated from the mouse brain data published with \textit{muscat} [REF] in a similar fashion.
+The `simMix2' dataset was generated from the mouse brain data published with \textit{muscat} [REF] in a similar fashion.
 The specific code is available on \url{https://github.com/markrobinsonuzh/scRNA\_pipelines\_paper}.
@@
 Version 0.1 was used here.
-Briefly, we first cluster the cells using scran's \textit{quickCluster} method, favouring overclustering (default minimum size of 20), and reduce the dataset to the union of 1500 most expressed genes of each cluster.
-We then create 75\% of the artifical doublets by sampling the two cells specifically from different clusters, and 25\% by entirely random sampling (in case the clustering was erroneous), and summing the counts of each pair of cells.
+Briefly, we first cluster the cells using \texttt{scran}'s \texttt{quickCluster} method, favouring overclustering (default minimum size of 20), and reduce the dataset to the union of 1500 most expressed genes of each cluster.
+We then create 75\% of the artificial doublets by sampling the two cells specifically from different clusters, and 25\% by entirely random sampling (in case the clustering was erroneous), and summing the counts of each pair of cells.
 We also create meta-cells from each cluster, and use them to create additional doublets and triplets.
@@
 
-The deviance explained by subpopulations was calculated directly on counts using the \textit{pipComp::getDevianceExplained} function.
+The deviance explained by subpopulations was calculated directly on counts using the \texttt{pipeComp::getDevianceExplained} function.
 The function uses edgeR to fit two models, a full \textasciitilde librarySize+subpopulation model and a reduced \textasciitilde librarySize model.

### 2019-12-02 main_figures/clustering.Rmd
added imputation and clustering results, not final

---
title: "Clustering"
author: "Pierre-Luc Germain"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    keep_md: true
---

```{r}
suppressPackageStartupMessages({
  library(ggplot2)
  library(cowplot)
  library(ComplexHeatmap)
  library(matrixStats)
})
source("../misc_functions.R")
suppressMessages(devtools::load_all("../../pipeComp/"))
theme_set(theme_cowplot(font_size = 12))
```

```{r clustering, fig.width=12, fig.height=3}
res <- readRDS("../data/results_clust.rds")
h <- scrna_evalPlot_clust(res, what="MI", agg.by=c("norm","clustmethod"), scale=TRUE, anno_legend = FALSE)
h <- h + scrna_evalPlot_clust(res, what="mean_F1", agg.by=c("norm","clustmethod"), scale=TRUE, reorder_rows = h, anno_legend = FALSE) +
  scrna_evalPlot_clust(res, what="min_pr", agg.by=c("norm","clustmethod"), scale=TRUE, reorder_rows = h, anno_legend = FALSE) +
  scrna_evalPlot_clustAtTrueK(res, agg.by=c("norm","clustmethod"), reorder_rows = h, anno_legend = FALSE)
h <- .renameHrows(h, function(x) gsub("seurat; clustmethod=|; clustmethod=", "", gsub("seuratvst", "sctransform + ", x)))
h <- .renameHrows(h, function(x) gsub("clust.scran","scran (walktrap)",gsub("clust.scran.fg", "scran (fast_greedy)", x)))
h
```

### 2019-12-02 main_figures/imputation.Rmd
added imputation and clustering results, not final

---
title: "Imputation/denoising"
author: "Pierre-Luc Germain"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    keep_md: true
---

```{r}
suppressPackageStartupMessages({
  library(ggplot2)
  library(cowplot)
  library(ComplexHeatmap)
  library(matrixStats)
})
source("../misc_functions.R")
suppressMessages(devtools::load_all("../../pipeComp/"))
theme_set(theme_cowplot(font_size = 10))
```

```{r}
res <- readRDS("../data/results_imputation.rds")
dr <- res$dr
dr <- dr[order(rowMeans(dr>0,na.rm=TRUE)+rowMedians(dr, na.rm=TRUE), decreasing=TRUE),order(colMeans(dr,na.rm=TRUE),decreasing=TRUE)]
ds <- factor(sapply(strsplit(colnames(dr),"\\."),FUN=function(x) x[[1]]))
cols <- getQualitativePalette(length(unique(ds)))
names(cols) <- sort(unique(ds))
dsa <- ComplexHeatmap::HeatmapAnnotation(dataset=ds, col=list(dataset=cols), show_annotation_name = FALSE, show_legend=TRUE)

h1 <- Heatmap(dr, cluster_columns = FALSE, col=.silScale(dr), cluster_rows = FALSE, 
        bottom_annotation=dsa,
        show_column_names = FALSE, show_column_dend = FALSE, 
        name="silh.\nwidth", column_title="Mean silhouette width", 
        right_annotation=rowAnnotation( mean=rowMeans(dr, na.rm=TRUE), 
                                        median=rowMedians(dr, na.rm=TRUE), 
                                        min=rowMins(dr, na.rm=TRUE), 
                                        show_legend=FALSE)
        )
p1 <- grid.grabExpr(draw(h1))

cl <- res$cl
mycast <- function(x,y){
  x <- reshape2::dcast(x, method~dataset, value.var=y, fun.aggregate = mean, na.rm=TRUE)
  row.names(x) <- x[,1]
  as.matrix(x[,-1])
}
ag <- aggregate(cl[,c("ARI",grep("_",colnames(cl),value=TRUE))], by=cl[,1:2], FUN=function(x){
  mean(as.numeric(unlist(x)),na.rm=TRUE)
  })
clw <- cl[which(cl$nbClustDeviation==0),]
ag2 <- aggregate(clw[,c("ARI",grep("_",colnames(clw),value=TRUE))], by=clw[,1:2], FUN=function(x){
  mean(as.numeric(unlist(x)),na.rm=TRUE)
})

ll2 <- list( "mean ARI"=mycast(ag, "ARI"),
            "ARI at true\nnb clusters"=mycast(ag2, "ARI"),
            "Mean precision"=mycast(ag, "mean_pr"),
            "Mean recall"=mycast(ag, "mean_re")
            )

e <- do.call(cbind, ll2)
o <- names(sort(rowMeans(e, na.rm=TRUE)+rowMedians(e, na.rm=TRUE), decreasing=TRUE))
ll2 <- lapply(ll2, FUN=function(x) x[o,])
ll2s <- lapply(ll2, scale)

myh <- function(k,title=k,...){
  rmLead0 <- sum(abs(ll2[[k]])>1,na.rm=TRUE)==0
  sp <- paste0('sprintf("%.2f", ll2[["',k,'"]][i, j])')
  resmid <- range(ll2s[[k]], na.rm=TRUE)
  resmid <- resmid[1]+(resmid[2]-resmid[1])/2
  cellfn <- eval(parse(text=paste0('function(j, i, x, y, width, height, fill){',
      'lab <- ', ifelse(rmLead0, paste('gsub("^0\\\\.",".",', sp,')'), sp),
      '; grid.text(lab, x, y, gp = gpar(fontsize = 10, 
      col=ifelse(ll2s[["',k,'"]][i,j]>',resmid,',"black","white"))) }')))
  Heatmap(ll2s[[k]], col = viridis::inferno(100), show_column_names = FALSE, bottom_annotation = .ds_anno(colnames(ll2[[k]]), FALSE), cluster_columns=FALSE, cluster_rows = FALSE, show_heatmap_legend = FALSE, column_title=title, cell_fun=cellfn, ...)
}

p2 <- grid.grabExpr(draw( myh(names(ll2)[1]) + myh(names(ll2)[2]) + 
                            myh(names(ll2)[3]) + myh(names(ll2)[4]) ))
```

```{r imputation, fig.width=10, fig.height=7}
plot_grid( p1, p2, nrow=2, labels=LETTERS[1:2], rel_heights = c(4,6))
```

**A:** Average silhouette width per subpopulation across various denoising/imputation methods. **B:** Clustering accuracy across denoising/imputation methods.

### 2020-01-22 main_figures/dimensionality.Rmd
updated dimensionality figure

@@
 o2$avgTime <- tmp[as.character(o2$method)]
-p1b <- ggplot(o2, aes(reorder(method, diff, function(x) mean(abs(x))), diff, group=method, fill=Time)) + geom_violin() + coord_flip() + geom_hline(yintercept=0, linetype="dashed") + ylab("Difference to 'real' number of dimensions") + scale_fill_manual(values=tcols) + theme(axis.title.y=element_blank(), legend.position=c(0.7, 0.3))
+p1b <- ggplot(o2, aes(reorder(method, diff, function(x) mean(abs(x))), diff, group=method, fill=Time)) + geom_violin() + coord_flip() + geom_hline(yintercept=0, linetype="dashed") + ylab("Difference to expected dimensionality") + scale_fill_manual(values=tcols) + theme(axis.title.y=element_blank(), legend.position=c(0.7, 0.3))
 p1 <- plot_grid(p1a, p1b, rel_widths = c(0.42,0.58), labels=LETTERS[1:2])
@@
 ```{r, fig.width=8, fig.height=4}
-res <- readRDS("../data/results_dimensionality.rds")
-sw <- res$dimensionality$clust.avg.silwidth$selected
+res1 <- readRDS("../data/results_dimensionality.rds")
+res2 <- readRDS("../data/results_dimensionality2.rds")
+sw1 <- res1$dimensionality$clust.avg.silwidth$selected
+sw2 <- res2$dimensionality$clust.avg.silwidth$selected
+sw2 <- sw2[grep("Local\\.b",row.names(sw2),invert = TRUE),] # redundant
+sw <- rbind(sw1,sw2[setdiff(row.names(sw2),row.names(sw1)),])
 sw <- sw[grep("=doublet",row.names(sw)),]
@@
 row.names(sw) <- n
-sw2 <- sw[order(grepl("sctransform",row.names(sw))*1000+rowMeans(sw)+apply(sw,1,probs=0.8,FUN=quantile),decreasing=TRUE),order(colMeans(sw),decreasing=TRUE)]
+sw2 <- sw[order(grepl("sctransform",row.names(sw))*1000+rowMeans(sw)+apply(sw,1,probs=0.2,FUN=quantile),decreasing=TRUE),order(colMeans(sw),decreasing=TRUE)]
 h1 <- Heatmap(sw2, cluster_columns = FALSE, cluster_rows = FALSE, 
@@
 
-res2 <- res$clustering
+cl1 <- res1$clustering
+cl2 <- res2$clustering
+res2 <- rbind(cl1, cl2[setdiff(row.names(cl2),row.names(cl1)), colnames(cl1)])
 res2 <- res2[grep("=doublet",row.names(res2)),]
@@
 ```{r figure6, fig.width=9, fig.height=9.5}
-plot_grid( p1, p2, p3, nrow=3, rel_heights=c(0.39,0.3,0.31), labels=c(NA,LETTERS[3:4]))
+plot_grid( p1, p2, p3, nrow=3, rel_heights=c(0.36,0.32,0.32), labels=c(NA,LETTERS[3:4]))
 ```

### 2020-01-22 bmc_article.tex

@@
 For example, by far the most important determinant of the ARI is the number of detected clusters: the farther it is from the actual number of subpopulations, the worse the ARI.
-In this context, one strategy has been to cluster across various resolutions and only consider the results that have the right number of cluster \citep{duoClustering2018}.
+In this context, one strategy has been to cluster across various resolutions and only consider the results that have the right number of clusters \citep{duoClustering2018}.
 While this has the virtue of making the results comparable, in practice the number of subpopulations is typically unknown, and tools that operate well in this optimal context might not necessarily be best overall.
-Clustering results are also very sensitive, and as shown below, might not always capture improvements in earlier steps of the pipeline.
+Clustering results are also very sensitive, and might not always capture improvements in earlier steps of the pipeline.
 
@@
 
-Doublets, defined as two cells sequenced under the same cellular barcode (for instance as a result of being captured in the same droplet), are fairly frequent in scRNAseq datasets, with estimates ranging from 1 to 10\% depending on the platform and cell concentration used \citep{bloomEstimating2018,kangMultiplexedDemuxlet2018}.
+Doublets, defined as two cells sequenced under the same cellular barcode (e.g., being captured in the same droplet), are fairly frequent in scRNAseq datasets, with estimates ranging from 1 to 10\% depending on the platform and cell concentration used \citep{bloomEstimating2018,kangMultiplexedDemuxlet2018}.
 While doublets of the same cell type are relatively innocuous for most downstream applications due to their conservation of the relative expression between genes, doublets formed from different cell types or states are likely to be misclassified and could potentially distort downstream analysis.
@@
 We therefore first evaluated the capacity of these methods to detect doublets using the two 10x datasets produced by \citep{tianMixology2018} with cells of different genetic identity, and for which therefore SNP information can be used as ground truth.
-We tested \texttt{DoubletFinder} [REF] and \texttt{scran}'s \texttt{doubletCells} [REF], both of which use similarity to artificial doublets, and \texttt{scds} [REF], which relies on a combination of coexpression and binary classification.
-\texttt{DoubletFinder} integrates a thresholding based on the proportion of expected doublets, while \texttt{scran} and \texttt{scds} return scores which must be manually thresholded.
+We tested \texttt{DoubletFinder} [REF] and \texttt{scran}'s \texttt{doubletCells} [REF], both of which use similarity to artificial doublets, and \texttt{scds} [REF], which relies on a combination of co-expression and binary classification.
+\texttt{DoubletFinder} integrates a thresholding based on the proportion of expected doublets, while \texttt{scran} and \texttt{scds} return scores that must be manually thresholded.
 In these cases, we ensured that the right number of cells would be called doublets.
@@
 This has for instance been demonstrated for cells that have a high content of mitochondrial reads, often as a result of cell degradation [REF].
-A common practice is to exclude cells that differ considerably from most other cells on the basis of some such properties, for instance through the \texttt{isOutlier} function of \texttt{scater} which measures, for a given control property, the number of median absolute deviation of each cell from the mean of all cells.
+A common practice is to exclude cells that differ considerably from most other cells on the basis of such properties, for instance through the \texttt{isOutlier} function of \texttt{scater} that measures, for a given control property, the number of median absolute deviations of each cell from the mean of all cells.
 Supplementary Figure 1 shows the distributions of some of the typical cell properties commonly used.
@@
 Reasoning that the cells we wish to eliminate are cells that would be misclassified, we measured the rate of misclassification of each cell in each dataset across a variety of clustering pipelines, correcting for the median misclassification rate of the subpopulation, and then evaluated what properties could be predictive of misclassification (Supplementary Figures 5-7).
-We could not identify any property or simple combination thereof that would be consistently predictive of misclassification; the only feature that consistently stood out across multiple datasets (the Zheng datasets) was that cells with a very high read counts have a higher chance of being misclassified.
+We could not identify any property or simple combination thereof that would be consistently predictive of misclassification; the only feature that consistently stood out across multiple datasets (the Zheng datasets) was that cells with very high read counts have a higher chance of being misclassified.
 
@@
 
-Reasoning that a selection method should ideally select genes whose variability is more between subpopulations than within, we first assessed whether each selection method selected genes with a high proportion of variance or deviance explained by (real) subpopulations.
-Estimates of variance explained by subpopulations based on a standard \texttt{Seurat} normalization or on sctransform data were highly correlated (Supplementary Figure 11A), and there was also a good agreement with deviance explained, although lowly-expressed genes in particular could have a high deviance explained without having much of their variance explained by subpopulations (Supplementary Figure 11B-D).
+Reasoning that a selection method should ideally select genes whose variability is more between subpopulations than within, we first assessed whether each selection method selected genes with a high proportion of variance or deviance explained by (real) subpopulation.
+Estimates of variance explained by subpopulation based on a standard \texttt{Seurat} normalization or on sctransform data were highly correlated (Supplementary Figure 11A), and there was also a good agreement with deviance explained, although lowly-expressed genes in particular could have a high deviance explained without having much of their variance explained by subpopulation (Supplementary Figure 11B-D).
 We therefore compared, using both of these estimates, the proportion of the cumulative variance/deviance explained by the top X genes that could be retrieved by each gene ranking method (Supplementary Figures 12-13).
 We first focused on the first 1000 genes to highlight differences between methods, although the differences tended to decrease with a higher number of genes selected (Supplementary Figures 12-14).
-A first observation was that standardized measures of variability were systematically worse than their non-standardized  counterparts in selecting genes with a high proportion of variance explained by subpopulations, while instead proving most of the time superior in selecting genes with a high deviance explained by subpopulations (Figure \ref{fig:figure6}A and Supplementary Figures 12-13).
+A first observation was that standardized measures of variability were systematically worse than their non-standardized  counterparts in selecting genes with a high proportion of variance explained by subpopulation, while instead proving most of the time superior in selecting genes with a high deviance explained by subpopulation (Figure \ref{fig:figure6}A and Supplementary Figures 12-13).
 Deviance proved the method of choice to maximize the former (with mere expression level proving surprisingly good), but surprisingly did not perform so well to select genes with a high deviance explained.
@@
 To validate the previous assay, we included a selection based on the genes whose proportion variance or deviance explained was highest.
-Interestingly, while these selections were on average the top-ranking methods, they were not systematically best for all dataset.
+Interestingly, while these selections were on average the top-ranking methods, they were not systematically best for all datasets.
 The previous observations were reflected in the ARI of the resulting clustering (Figure 4B): non-standardized measures of variability, including mere expression level, tended to outperform more complex metrics.
-In general, we found deviance and un-standardized estimate of variance to provide the best results across datasets and normalization methods.
+In general, we found deviance and unstandardized estimates of variance to provide the best results across datasets and normalization methods.
 Increasing the number of features selected also systematically led to an increase in the accuracy of the clustering, typically plateauing after 4000 features (Supplementary Figure 14).
@@
 We then compared various methods to estimate the intrinsic dimension based on the reduced space (from Seurat weighted PCA).
-As a first approximation of the real dimension, we computed the variance in each principal component which was explained by the subpopulations; this dropped dramatically after the first few components, suggesting an overall low dimensionality (Figure \ref{fig:figure8}A) at least for the purpose of cell type definition.
+As a first approximation of the real dimension, we computed the variance in each principal component that was explained by the subpopulations; this dropped dramatically after the first few components, suggesting an overall low dimensionality (Figure \ref{fig:figure8}A) at least for the purpose of cell type definition.
 We next estimated dimensions using various methods (see methods), some of which are already commonly used for scRNAseq, while others were borrowed from other fields.
@@
 \subsection*{Software and package versions}
-Analyses were performed in R 3.6.0 (bioconductor 3.9), and the following packages were installed from github repositories: \texttt{Seurat} (version 3.0.0), \texttt{sctransform} (0.2.0), \texttt{DoubletFinder} (2.0.1), \texttt{scds} (1.0.0).
+Analyses were performed in R 3.6.0 (Bioconductor 3.9), and the following packages were installed from github repositories: \texttt{Seurat} (version 3.0.0), \texttt{sctransform} (0.2.0), \texttt{DoubletFinder} (2.0.1), \texttt{scds} (1.0.0).
 The code for the glmPCA was obtained from \url{https://github.com/willtownes/scrna2019} (commit 1ddcc30ebb95d083a685f12fe81d35dd1b0cb1b2).

### 2020-01-23 main_figures/doublets.Rmd
updated doublets figure

--- 2019-05-09 main_figures/figure1_doublets.Rmd
@@
   library(ROCit)
+  library(ggplot2)
   library(cowplot)
 })
+devtools::load_all("~/bioinfo/robinson/scDblFinder/")
+theme_set(theme_cowplot(font_size = 12))
+
 # we load the dataset
-sce <- readRDS("../datasets/mixology10x5cl.SCE.rds")
-CD <- as.data.frame(colData(sce))
+lf <- list.files("../datasets/", pattern="SCE", full.names=TRUE)
+names(lf) <- sapply(strsplit(basename(lf),".",fixed=TRUE), head, 1)
+CDs <- lapply(lf, FUN=function(x){ as.data.frame(colData(readRDS(x))) })
 ```
@@
 ```{r}
-truth <- factor(CD$demuxlet_cls, c("SNG","DBL"))
-roclist <- lapply( list(plDoubletFinder=CD$pldf.ratio,
-                        DoubletFinder.p=CD$doubletFinder.pANN_0.25_0.02_82,
-                        scran.score=CD$scran.doubletscore,
-                        scds.hybrid=CD$hybrid_score), 
-                   class=truth, FUN=rocit)
-colors <- c("blue","red","black","pink")
-plot(roclist[[1]], lwd=2, col=c(colors[[1]], "grey"), legend=FALSE, YIndex=FALSE, main="mixology10x5cl")
-for(i in 2:length(roclist)){
-  lines(roclist[[i]]$FPR, roclist[[i]]$TPR, lwd=2, col=colors[[i]])
+res <- dblres <- readRDS("../data/doublets.rds")
+scoresNames <- c("scran.score"="scran", "scDblFinder"="scDblFinder", "DF.score"="DoubletFinder", "hybrid_score"="scds")
+cols <- c("#4477AA", "#DDCC77", "#117733", "#CC6677")
+names(cols) <- as.character(scoresNames)
+
+pw <- function(i){
+  scores <- res[[i]][,intersect(names(scoresNames),colnames(res[[i]]))]
+  colnames(scores) <- scoresNames[colnames(scores)]
+  plotROCs(scores, factor(res[[i]]$demuxlet_cls, c("SNG","DBL"))) +
+    ylim(0,1) + scale_color_manual(values=cols)
 }
-legend("bottomright", fill=colors, legend=names(roclist))
+
+p1 <- pw(1) + theme(legend.position="none")
+p2 <- pw(2) + theme(legend.position="none")
+p3 <- pw(3) + theme(legend.position=c(0.25,0.25))
 ```
+
+```{r}
+rt <- data.frame(
+  method = structure(c(3L, 4L, 2L, 1L, 3L, 4L, 2L, 1L, 
+3L, 4L, 2L, 1L), .Label = c("DoubletFinder", "scDblFinder", "scds", 
+"scran"), class = "factor"),
+  dataset = structure(c(1L, 1L, 1L, 
+1L, 2L, 2L, 2L, 2L, 3L, 3L, 3L, 3L), .Label = c("mixology10x3cl", 
+"mixology10x5cl", "demuxlet"), class = "factor"),
+  time = c(14.52, 26.36, 7.82400000000007, NA, 42.74, 
+250.14, 21.8199999999999, 678.81, 100.05, 823.09, 54.0140000000001, 
+956.63)
+)
+
+pt <- ggplot(rt, aes(method, time, fill=method)) + geom_col() + facet_wrap(~dataset, scale="free_x") + coord_flip() + xlab("") + ylab("Running time (s)") + scale_fill_manual(values=cols) + theme(legend.position="none")
+
+# le <- get_legend(plotROCs(scores, factor(CDs[[i]]$demuxlet_cls, c("SNG","DBL"))))
+```
+
+```{r}
+mis <- readRDS("../data/misclassification.rds")
+lf <- list.files("../datasets/", full.names = TRUE)
+lf <- lf[grep("simMix",lf,invert=TRUE)]
+names(lf) <- sapply(strsplit(basename(lf),"\\."), FUN=function(x) x[1])
+CDs <- lapply(lf, FUN=function(x) colData(readRDS(x)))
+dbl <- lapply(CDs, FUN=function(x) x[,c("pldf.ratio","pldf.class")])
+dbl <- dbl[intersect(names(dbl),names(mis))]
+for(n in names(dbl)){
+  dbl[[n]]$pct_misclassified_cor <- mis[[n]]$corrected_pct_mis[row.names(dbl[[n]])]
+  dbl[[n]]$pct_misclassified <- mis[[n]]$pct_misclassified[row.names(dbl[[n]])]
+}
+d <- as.data.frame(data.table::rbindlist(lapply(dbl,as.data.frame), idcol="dataset"))
+d$pldf.class <- factor(d$pldf.class, c("doublet","singlet"))
+levels(d$pldf.class) <- c("DBL","SNG")
+p4 <- ggplot(d, aes(pldf.class, pct_misclassified, fill=pldf.class)) + geom_violin() + facet_wrap(~dataset, nrow = 1) + theme(axis.title.x=element_blank(), legend.position="none") + ylab("% misclassified") + stat_summary(fun.y="mean", geom="point")
+```
+
+```{r figure1, fig.height=6, fig.width=7}
+p1m <- plot_grid(p1, p2, p3, nrow=1)
+plot_grid(p1m, pt, p4, rel_heights=c(3.5,2,2), labels = LETTERS[1:3], ncol = 1, hjust=0)
+```
+
+**A:** Receiver operating characteristic (ROC) curves of the tested doublet detection methods for three datasets with SNP-identified doublets.
+**B:** Running time of the different methods (DoubletFinder failed on one of the datasets).
+**C:** Rate of misclassification of the cells identified by scDblFinder as doublets (DBL), across a large range of clustering analysis.
+Even in datasets which should not have neotypic doublets, the cells identified as such tend to be misclassified.

### 2020-01-23 bmc_article.tex
Updates from Overleaf

@@
 
-Single-cell RNA-sequencing (scRNAseq) and the set of attached analysis methods are evolving fast, and while a number of good comparison/benchmark studies can inform analytical decisions (\citealp{duoClustering2018}, etc) they need constant updating and often leave open many details of an analysis.
+Single-cell RNA-sequencing (scRNAseq) and the set of attached analysis methods are evolving fast, and while a number of good comparison/benchmark studies can inform analytical decisions (\citealp{duoClustering2018, SonesonDE2018, SunDimRed2019}) they need constant updating and often leave open many details of an analysis.
 Here, we go in more detail on the various steps of analysis leading from an initial count matrix to a cluster assignment, which are critical steps in a wide range of applications.
@@
     \centering
-    \includegraphics[width=\textwidth,keepaspectratio]{figure0}
+    \includegraphics[width=\textwidth,keepaspectratio]{main_figures/pipeline_explanation.png}
     \caption{\textbf{Overview of the pipeComp framework and its application to a scRNAseq clustering pipeline.}}
@@
 
-In addition to these methods, we reasoned that an approach such as \texttt{DoubletFinder} could be simplified by being applied directly on counts and by using a pre-clustering to create neotypic doublets (i.e.
+In addition to these methods, we reasoned that an approach such as \texttt{DoubletFinder} could be simplified by being applied directly on counts and by using a pre-clustering to create neotypic/heterotypic doublets (i.e.
 doublets from different cell types) more efficiently.

### 2020-01-29 supplementary_figures/supp1_dataset_description.Rmd
updated supplementary figures

@@
 suppressMessages(devtools::load_all("../../pipComp/"))
-
+theme_set(theme_cowplot(font_size = 12))
 # we load the datasets
@@
 names(lf) <- gsub(".SCE.rds", "", basename(lf), fixed=TRUE)
+lf <- lf[grep("simMix", lf, invert=TRUE)]
 SCEs <- lapply(lf, readRDS)
@@
 ```{r misclass, fig.height=4}
-theme_set(theme_cowplot(font_size = 11))
+theme_set(theme_cowplot(font_size = 10))
 mis <- readRDS("../data/misclassification.rds")

### 2020-01-29 supplementary_figures/supp2_filtering.Rmd
updated supplementary figures

--- 2019-05-15 supplementary_figures/supp2_filtering.Rmd
@@
 
-```{r}
+```{r, include=FALSE}
 suppressPackageStartupMessages({
@@
   library(knitr)
+  library(circlize)
 })
-suppressMessages(devtools::load_all("../../pipComp/"))
+suppressMessages(devtools::load_all("../../pipeComp/"))
 source("../misc_functions.R")
-data(datasets)
-
-# we load the datasets
-lf <- list.files("../datasets/", full.names=TRUE)
-names(lf) <- gsub(".SCE.rds", "", basename(lf), fixed=TRUE)
-SCEs <- lapply(lf, readRDS)
+load("../../pipComp/data/datasets.RData")
 ```
@@
 
-```{r filt_more_ari, fig.width=8, fig.height=8.5}
-res <- readRDS("../data/filtering_old.rds")
-res$pcCells <- round(100*res$nbCells/datasets[as.character(res$dataset),"nbCells"],2)
-# we restrict to analyses that gave a number of clusters close to reality:
-res2 <- res[which(abs(res$nbClusters-datasets[as.character(res$dataset), "subpopulations"])<=1),]
-p1 <- ggplot(res2, aes(x=pcCells, y=ARI, colour=norm)) + geom_point() + facet_wrap(~dataset, scale="free")
+```{r filt_more, fig.width=8, fig.height=9}
+res <- readRDS("../data/results_filtering.rds")
+m <- scrna_evalPlot_filtering(res, returnTable=TRUE)
+m <- m[grep("veryStringent", m$method, invert=TRUE),]
 
-fres <- readRDS("../../resNew/filtering_intermediateSummary.rds")
-pc <- fres$filtering$pcCells
-pc <- pc[grep("doublet",row.names(pc),invert=T),]
-row.names(pc) <- reduceMethodNames(row.names(pc),2)
-ha <- HeatmapAnnotation(dataset=reduceMethodNames(colnames(pc),1))
-shortDSNames <- c("mixology10x3cl"="mix3cl", "mixology10x5cl"="mix5cl", "Zhengmix4eq"="Zh4eq", "Zhengmix4uneq"="Zh4uneq", "Zhengmix8eq"="Zh8eq")
-colnames(pc) <- paste(shortDSNames[reduceMethodNames(colnames(pc),1)], reduceMethodNames(colnames(pc),2))
-p2 <- grid.grabExpr(draw(Heatmap(pc, name="% out", bottom_annotation=ha)))
+# # horizontal version
+# m <- m[grep("veryStringent", m$method, invert=TRUE),]
+# m.f1 <- reshape2::dcast(m, method~dataset, value.var="F1atK")
+# n <- sapply(strsplit(gsub("=",".",m.f1[,1]),"\\."),FUN=function(x) rev(x)[1])
+# n[grep("=doublet", m.f1[,1])] <- paste0("doubletRemoval+",n[grep("=doublet", m.f1[,1])])
+# m.pc <- reshape2::dcast(m, method~dataset, value.var="maxPCout")[,-1]
+# m.f1m <- reshape2::dcast(m, method~dataset, value.var="meanF1")[,-1]
+# m.pcm <- reshape2::dcast(m, method~dataset, value.var="medianPCout")[,-1]
+# m.f1 <- m.f1[,-1]
+# row.names(m.f1m) <- row.names(m.pcm) <- row.names(m.f1) <- row.names(m.pc) <- n
+# 
+# chm(list("mean F1"=m.f1m, "F1 at true \n# clusters"=m.f1, "max % lost per\nsubpopulation"=m.pc, "median % lost per\nsubpopulation"=m.pcm), scale = "column", colors = viridis::inferno(3), value_format = c("%.2f","%.2f","%.0f","%.0f"), scaleFn=list(scale,scale, scale2,scale2), cluster_columns = FALSE,  row_sortFn = function(x, na.rm){ sum(x[1:18],na.rm=TRUE)-median(x[19:36], na.rm=TRUE)/100})
 
-plot_grid(p1, p2, labels=LETTERS[1:2], nrow=2)
+mycast <- function(x, val, form=dataset~method){ 
+  x <- reshape2::dcast(x, form, value.var=val)
+  row.names(x) <- x[,1]
+  as.matrix(x[,-1])
+}
+m.f1 <- mycast(m, "F1atK")
+n <- sapply(strsplit(gsub("=",".",colnames(m.f1)),"\\."),FUN=function(x) rev(x)[1])
+n[grep("=doublet", colnames(m.f1))] <- paste0("doubletRemoval+",n[grep("=doublet", colnames(m.f1))])
+m.pc <- mycast(m, "maxPCout")
+colnames(m.f1) <- colnames(m.pc) <- n
+o <- order(colMeans(t(scale(t(m.f1))), na.rm=TRUE)*100-colMeans(t(scale(t(m.pc))), na.rm=TRUE))
+m.f1 <- m.f1[,o]
+m.pc <- m.pc[,o]
+m.f1m <- mycast(m, "meanF1")[,o]
+m.pcm <- mycast(m, "medianPCout")[,o]
+colnames(m.f1m) <- colnames(m.pcm) <- colnames(m.f1)
+
+names(nn) <- nn <- c("m.f1m","m.f1","m.pc","m.pcm")
+cellfns <- lapply(nn, FUN=function(mname){
+  if(grepl("f1",mname)){
+    sp <- paste0('sprintf("%.1f", 100*',mname,'[i, j])')
+  }else{
+    sp <- paste0('sprintf("%.1f", ',mname,'[i, j])')
+  }
+  eval(parse(text=paste0('function(j, i, x, y, width, height, fill){',
+      'lab <- ', sp, '; grid.text(lab, x, y, gp = gpar(fontsize = 10)) }')))
+})
+
+names(nn) <- c("mean F1", "F1 at true \n# clusters", "max % lost per\nsubpopulation", "median % lost per\nsubpopulation")
+hs <- lapply(seq_along(nn), FUN=function(i){
+  mat <- get(nn[[i]])
+  if(grepl("f1",nn[[i]])){
+    mat <- t(scale(t(mat)))
+  }else{
+    mat <- t(scale(t(sqrt(mat))))
+  }
+  Heatmap(mat, row_title=names(nn)[[i]], col=viridis::inferno(100), cell_fun = cellfns[[i]], show_heatmap_legend = FALSE, cluster_columns = FALSE, cluster_rows = FALSE)
+})
+
+hs[[1]] %v% hs[[2]] %v% hs[[3]] %v% hs[[4]]
 ```
@@
 ### Supplementary Figure `r getFigNb()`
-**A:** Adjusted rand index (ARI) of various clustering pipelines using various combinations of filtering criteria.
-Only clustering analyses with a number of clusters +/-1 from the real number of clusters were considered.
-In general, filtering more does not result in more accurate clustering.
-**B:** Percentage of cells filtered out per subpopulation, using various combination of filters (see methods).
-
-\newpage
-
-# Supplementary Figure `r getFigNb(TRUE)`
-
-```{r, fig.height=4}
-res$filt.times <- sapply(strsplit(res$filt,"_"), FUN=function(x){ if(length(x)<3) return(1); as.numeric(x[[3]])})
-res$filt.MADs <- sapply(strsplit(res$filt,"_"), FUN=function(x){ as.numeric(x[[2]]) })
-filt.vars <- strsplit(sapply(strsplit(res$filt,"_"), FUN=function(x) x[[1]]),";")
-fv <- unique(unlist(filt.vars))
-for(f in fv[grep("both",fv,invert=TRUE)]){
-  f2 <- gsub("lower$|higher$","both",f)
-  res[[f]] <- sapply(filt.vars, FUN=function(x){ any(c(f,f2) %in% x) })
-}
-form <- paste0("~nbClusters*dataset+norm+",paste(grep(".",colnames(res),fixed=TRUE,value=TRUE),collapse="+"))
-form <- gsub("filt.MADs","factor(filt.MADs)",form,fixed=T)
-form <- gsub("filt.times","factor(filt.times)",form,fixed=T)
-form <- paste0("ARI",form)
-message(strwrap(paste("lm(", form, ")")))
-mod <- lm(as.formula(form), data=res)
-mod <- coef(summary(mod))[grep(".",row.names(coef(summary(mod))),fixed=T),]
-colnames(mod)[4] <- "p value"
-kable(mod)
-```
-
-### Supplementary Figure `r getFigNb()`
-Linear regression analysis of ARI on various filtering criteria, correcting for the effect of the number of clusters.
-Increasing the number of distributions on which a cell must deviate in order to be excluded (`filt.times`) tended to be positively associated with ARI, while increasing the number of median absolute deviations required to be an outlier (`filt.MADs`) was associated with a decrease in ARI.
-
-\newpage
-
-# Supplementary Figure `r getFigNb(TRUE)`
-
-```{r}
-sw <- fres$dimreduction$clust.avg.silwidth[[1]]
-sw <- sw[,grep("doublet",colnames(sw))]
-w <- grep("seuratvst",colnames(sw))
-colnames(sw) <- reduceMethodNames(colnames(sw),2)
-
-chm(list("standard Seurat norm"=sw[,-w], "sctransform"=sw[,w]), scale="row", scaleTitle="z-score", row_ann=data.frame(meanSilWidth=rowMeans(sw), dataset=reduceMethodNames(row.names(sw),1)), show_row_names=FALSE)
-```
-
-```{r fig.height=3}
-r <- fres$dimreduction$PC.R2[which(fres$dimreduction$PC.R2$dim<=10),]
-r <- r[grep("doublet",r$method),]
-r <- r[grep("clust",r$method, invert=T),]
-r <- aggregate(r$value, by=r[,c(1,3)], FUN=mean)
-r <- t(cast2(r, dataset~method, value.var="x", fun.aggregate=mean))
-w <- grep("seuratvst", row.names(r))
-row.names(r) <- reduceMethodNames(row.names(r), 2)
-chm(list("standard Seurat norm"=t(r[-w,]), "sctransform"=t(r[w,])), sameScale=TRUE, 
-    scale="row", scaleTitle="z-score", value_format="%.2f")
-```
-
-### Supplementary Figure `r getFigNb()`
-Impact of the set of filters on the average silhouette width of the subpopulations (top) and on the proportion of variance in the first components explained by clusters (bottom).
+Mean clustering F1 score per subpopulation, mean F1 at true number of clusters, as well as maximum and median proportion of excluded cells per subpopulation across various filtering strategies.
+Doublet removal generally improves clustering accuracy with very mild exclusion rates, even in datasets that do not have heterotypic doublets.
+Stringent distribution-based filtering creates large cell type biases.
 
@@
 })
-chm(ll, scale="column", scaleTitle="z-score", sameScale=TRUE, value_format="%.2f", row_sortFn=function(x){ median(x,na.rm=T) }, cluster_columns=FALSE)
+chm(ll, scale="column", scaleTitle="z-score", sameScale=TRUE, value_format="%.2f", row_sortFn=function(x,...){ median(x,...) }, cluster_columns=FALSE)
 ```

### 2020-01-29 supplementary_figures/supp3_normalization.Rmd
updated supplementary figures

--- 2019-05-18 supplementary_figures/supp3_normalization.Rmd
@@
 data(datasets)
+theme_set(theme_cowplot(font_size = 11))
 source("../misc_functions.R")
@@
 
-```{r norm_r2, fig.width=8, fig.height=9}
-res <- readRDS("../../resNew/norm2_intermediateSummary.rds")
-r2 <- res$dimreduction$PC.R2
-r2$norm <- gsub("norm.","",reduceMethodNames(r2$method,3),fixed=T)
-r2 <- r2[which(r2$dim <= 10),]
-r2$R.on.Clusters <- sqrt(r2$value)
-r3 <- cast2(r2, formula=norm~dataset, value.var="R.on.Clusters")
-r2m <- dplyr::summarise(group_by(r2,dataset), meanR=mean(R.on.Clusters))
-p1 <- ggplot(r2, aes(norm, R.on.Clusters, fill=norm)) + geom_violin() + facet_wrap(~dataset, nrow=1) + coord_flip() + theme(legend.position="none") + stat_summary(fun.y="mean", geom="point") + geom_hline(data=r2m, aes(yintercept=meanR)) + ggtitle("Proportion of the top dimensions explained by real clusters") + ylab("sqrt(R^2) of dimension ~ real subpopulation")
+```{r norm_nbClusters, fig.width=8, fig.height=8.5}
+res <- readRDS("../data/norm2_endSummary.rds")
+res2 <- res[which(res$filt=="deffilter" & res$norm %in% c("norm.scran", "norm.scran.noscale", "norm.seurat", "norm.seuratvst", "norm.stableGsum")),]
+res2$diff.nbClusters <- res2$nbClusters-datasets[res2$dataset,"subpopulations"]
+res3 <- aggregate(res2[,"diff.nbClusters"], by=res2[,c("dataset","dims","resolution","norm")], FUN=mean)
+colnames(res3)[5] <- "nb"
+res3$norm <- gsub("norm.","",res3$norm, fixed=T)
+# ggplot(res3, aes(factor(dims), factor(resolution), fill=nb)) + stat_bin2d() + 
+#   scale_fill_viridis() + ylab("Resolution") + xlab("Number of dimensions") +
+#   theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
+#   coord_flip() + facet_grid(dataset~norm)
 
-pc1.covar <- lapply( res$dimreduction$PC1.covarR[2:3], FUN=function(x){
-  x <- aggregate(x, by=list(reduceMethodNames(row.names(x), 3)), FUN=median)
-  row.names(x) <- x[,1]
-  x[,1] <- NULL
-  x
+ll <- lapply(split(res3, res3$dataset), FUN=function(x){
+  ggplot(x, aes(factor(resolution), factor(dims), fill=nb)) + stat_bin2d() + 
+    scale_fill_viridis() + xlab("Resolution") + ylab("dims") +
+    facet_grid(.~norm)
 })
-names(pc1.covar) <- c("Correlation with log10-counts", "Correlation with nb features")
-p2 <- grid.grabExpr(draw(chm(pc1.covar, sameScale=TRUE, scaleTitle="cor", value_format="%.2f", row_sortFn=median)))
-plot_grid(p2, p1, labels=LETTERS[1:2], nrow=2)
+for(i in names(ll)){
+  ll[[i]] <- ll[[i]] + ggtitle(i)
+  if(i==names(ll)[length(ll)]){
+    ll[[i]] <- ll[[i]] + theme(axis.text.x = element_text(angle = 90, hjust = 1))
+  }else{
+    ll[[i]] <- ll[[i]] + theme(axis.title.x=element_blank(), axis.text.x=element_blank(), axis.ticks.x=element_blank())
+  }
+}
+ll[[4]] <- ll[[4]] + ylab("Number of dimensions")
+
+mod <- summary(lm(nbClusters~dataset+norm+dims+resolution, data=res2))
+plot_grid(plotlist = ll[-2], nrow=6, rel_heights=c(1,1,1,1,1,1.5))
 ```
@@
 ### Supplementary Figure `r getFigNb()`
-**A:** Correlation between library size (left) or the number of detected features (right) and the residuals of a linear regression of the first principal component on the subpopulation.
-**B:** Proportion of the variance in the first 10 PCA components that is explained by real subpopulations. The square-root of R-squared values is plotted for ease of visualization.
-
-\newpage
-
-# Supplementary Figure `r getFigNb(TRUE)`
-
-```{r norm_silwidth, fig.width=8, fig.height=7}
-res <- readRDS("../../resNew/norm2_intermediateSummary.rds")
-sw <- res$dimreduction$clust.avg.silwidth[[1]]
-colnames(sw) <- gsub("norm.","",reduceMethodNames(colnames(sw), 3),fixed=T)
-sw <- sw[order(reduceMethodNames(row.names(sw),1), rowMeans(sw)),]
-ra <- data.frame(meanSilWidth=rowMeans(sw), dataset=reduceMethodNames(row.names(sw),1))
-chm(list("Cluster mean silhouette width"=sw), scaleTitle="row z-score", scale="row", row_ann=ra, show_row_names=FALSE, col_sortFn=mean, cluster_rows=FALSE)
-```
-
-### Supplementary Figure `r getFigNb()`
-Mean silhouette width of true subpopulations across the first 10 components with different normalization and scaling approaches.
-
-\newpage
+Mean difference between the number of detected clusters and the number of real subpopulations, depending on the normalization method, the resolution and the number of dimensions used.
+The Kumar dataset is not shown here due to a lack of variation in the number of clusters detected.
+A rough ANOVA on `nbClusters~dataset+norm+dims+resolution` suggests that seuratvst (sctransform) is associated with a higher number of clusters (p~`r round(mod$coefficients["normnorm.seuratvst",4],3)`).

### 2020-01-29 supplementary_figures/supp4_selection.Rmd
updated supplementary figures

--- 2019-05-15 supplementary_figures/supp4_selection.Rmd
@@
 }
-```
 
-```{r}
 suppressPackageStartupMessages({
@@
 layout(matrix(1:4,nrow=2,byrow=TRUE))
-LSD::heatscatter(gi$mixology10x3cl$vst.R2, gi$mixology10x3cl$lognorm.R2, xlab="Prop variance explained based, sctransform", ylab="Prop variance explained, standard seurat normalization", main="A.
-Variance explained by real subpopulations (mixology10x3cl)")
+par(mar= c(7, 4, 5, 2) + 0.1)
+LSD::heatscatter(gi$mixology10x3cl$vst.R2, gi$mixology10x3cl$lognorm.R2, xlab="Prop var explained based, sctransform", ylab="Prop var explained, lognorm", main="A.
+Var explained by subpopulations\n(mixology10x3cl)", cex.main=0.8)
 
-LSD::heatscatter(gi$mixology10x3cl$devianceExplained, gi$mixology10x3cl$vst.R2, xlab="Deviance explained by real subpopulations", ylab="sctransform-based variance explained", , main="B.
-Variance explained vs deviance explained (mixology10x3cl)")
+LSD::heatscatter(gi$mixology10x3cl$devianceExplained, gi$mixology10x3cl$vst.R2, xlab="Deviance explained by subpopulations", ylab="sctransform-based variance explained", , main="B.
+Var explained vs deviance explained\n(mixology10x3cl)", cex.main=0.8)
 
 diffplot <- function(x, title=""){
-  LSD::heatscatter(x$log10_mean_counts, x$devianceExplained - x$vst.R2, xlab="log10(mean counts)", ylab="Prop deviance explained - prop variance explained", main=title)
+  LSD::heatscatter(x$log10_mean_counts, x$devianceExplained - x$vst.R2, xlab="log10(mean counts)", ylab="Prop deviance explained - prop variance explained", main=title, cex.main=0.8)
 }
-diffplot(gi$mixology10x3cl, "C. Difference between the estimates (mixology10x3cl)")
-diffplot(gi$Zhengmix4eq, "D. Difference between the estimates (Zhengmix4eq)")
+diffplot(gi$mixology10x3cl, "C. Difference between the estimates\n(mixology10x3cl)")
+diffplot(gi$Zhengmix4eq, "D. Difference between the estimates\n(Zhengmix4eq)")
 ```
@@
 
-```{r, fig.width=8, fig.height=9, warnings=FALSE, messages=FALSE}
+```{r, include=FALSE}
 ll <- lapply(names(gi), FUN=function(x){
@@
 ll$legend <- get_legend(ll[[1]] + theme(legend.position="right"))
+```
+
+```{r, fig.width=8, fig.height=9, results="hide", warnings=FALSE, messages=FALSE}
 plot_grid(plotlist=ll, ncol=2)
@@
 ll$legend <- get_legend(ll[[1]] + theme(legend.position="right"))
+plot_grid(plotlist=ll, ncol=2)
+```
+
+```{r, fig.width=8, fig.height=9, results="hide", warnings=FALSE, messages=FALSE}
 plot_grid(plotlist=ll, ncol=2)
@@
 
-```{r, fig.width=8, fig.height=6}
-res <- readRDS("../../resNew/sel_intermediateSummary.rds")
-r2 <- res$dimreduction$PC.R2
-r2 <- r2[grep("1000$",r2$method),]
-r2 <- r2[grep("doublet.pl deffilter",r2$method,fixed=T),]
-r2$norm <- reduceMethodNames(r2$method,3)
-r2$method <- reduceMethodNames(r2$method,4)
-r2 <- r2[which(r2$dim <= 5),]
-r2$R.on.Clusters <- sqrt(r2$value)
-r3 <- lapply(split(r2, r2$norm), FUN=function(x){ 
-  cast2(x, formula=method~dataset, value.var="R.on.Clusters")
+```{r, fig.width=8, fig.height=9}
+res <- readRDS("../data/results_selection.rds")
+cl <- res$clustering
+cl <- cl[grep("devianceExplained|varExp", row.names(cl), invert=TRUE),]
+vst <- cl[grep("seuratvst",row.names(cl)),]
+snorm <- cl[grep("seuratvst",row.names(cl), invert=TRUE),]
+
+ds <- sapply(strsplit(colnames(cl)," "), FUN=function(x) x[1])
+ds <- split(seq_len(ncol(cl)), ds)
+ds <- ds[grep("stepElapsed",names(ds),invert=TRUE)]
+myag <- function(cl, out="ARI", trueK=FALSE){
+  y <- lapply(ds, FUN=function(x){
+    x <- cl[,x]
+    colnames(x) <- sapply(strsplit(colnames(x)," "), FUN=function(x) x[2])
+    x <- parsePipNames(x)
+    if(!trueK) return( aggregate( x[,out], 
+                                  by=list(method=x$sel, nb=x$selnb), 
+                                  na.rm=TRUE, FUN=mean) )
+    w <- which(x$n_clus==x$true.nbClusts)
+    aggregate(x[w,out], by=list(method=x$sel[w], nb=x$selnb[w]), na.rm=TRUE, FUN=mean)
   })
-chm(r3[-1], scale="column", scaleTitle="z-score", value_format="%.2f", row_sortFn=median, sameScale=TRUE)
+  bind_rows(y, .id="dataset")
+}
+vst.ARIatK <- myag(vst, trueK=TRUE)
+snorm.ARIatK <- myag(snorm, trueK=TRUE)
+
+cols <- getQualitativePalette(9)
+names(cols) <- unique(vst.ARIatK$method)
+
+plot_grid( 
+  ggplot(vst.ARIatK, aes(nb, x, colour=method)) + geom_line(size=1.3) +facet_wrap(~dataset, scales="free") + scale_color_manual(values = cols) + xlab("Number of genes selected") + ylab("ARI at true # of clusters") + theme(axis.text=element_text(size=10)),
+  ggplot(snorm.ARIatK, aes(nb, x, colour=method)) + geom_line(size=1.3) +facet_wrap(~dataset, scales="free") + scale_color_manual(values = cols) + xlab("Number of genes selected") + ylab("ARI at true # of clusters") + theme(axis.text=element_text(size=10)),
+  nrow=2, labels = LETTERS[1:2]
+)
 ```
@@
 
-Proportion of the variance in the first 5 principal components that is explained by real subpopulations according to different normalization and selection procedures, selecting 1000 genes in each case.
+Clustering accuracy according to the number of genes selected using various ranking/selection methods. **A:** Based on sctransform, **B:** Based on standard Seurat normalization.
 

### 2020-01-29 supplementary_figures/supp5_dr.Rmd
updated supplementary figures

@@
 ```
-
-# Supplementary Figure `r getFigNb(TRUE)`
-
-```{r, fig.width=8, fig.height=8}
-res <- readRDS("../data/dr_intermediateSummary.rds")
-sw <- res$dimreduction$clust.avg.silwidth[[1]]
-colnames(sw) <- reduceMethodNames(colnames(sw),c(3,4,6))
-colnames(sw) <- gsub("norm.","",colnames(sw),fixed=T)
-sw <- matrix(unlist(sw),ncol=ncol(sw), dimnames=dimnames(sw))
-df <- data.frame(Norm=reduceMethodNames(colnames(sw),1), Selection=reduceMethodNames(colnames(sw),2), DimRed=reduceMethodNames(colnames(sw),3), stringsAsFactors=FALSE)
-df$Norm[grep("GlmPCA", df$DimRed)] <- ""
-ag <- aggregate(t(sw), by=df, FUN=median)
-ag <- ag[order(ag[,3],ag[,1],ag[,2]),]
-sw2 <- t(ag[,-1:-3])
-colnames(sw2) <- apply(ag[,c(1,3)],1,collapse=" ",FUN=paste)
-
-# hma <- HeatmapAnnotation(df=data.frame(meanSilWidth=rowMeans(sw2), dataset=reduceMethodNames(row.names(sw),1)))
-# chm(list("Cluster mean silhouette width"=t(sw2)), scaleTitle="meanSilWidth", scale="none", show_row_names=TRUE, show_column_names=FALSE, row_ann=ag[,1:3], top_annotation=hma, cluster_rows=FALSE)
-
-hma <- HeatmapAnnotation(df=ag[,1:3])
-chm(list("Cluster mean silhouette width"=sw2), scaleTitle="meanSilWidth", scale="none", row_ann=data.frame(meanSilWidth=rowMeans(sw), dataset=reduceMethodNames(row.names(sw),1)), show_row_names=FALSE, top_annotation=hma)
-```
-
-### Supplementary Figure `r getFigNb()`
-Average silhouette width of real subpopulations when using the first 10 components of different dimensionality reductions.
-Weighing the components according to their variance increases the distinctiveness of the subpopulations.
-Nevertheless, Seurat's standard PCA appears superior to the alternatives.
-
-\newpage
-
-# Supplementary Figure `r getFigNb(TRUE)`
-
-```{r, fig.width=8.5, fig.height=5}
-r2 <- res$dimreduction$PC.R2
-r2$norm <- gsub("norm.","",reduceMethodNames(r2$method,6),fixed=T)
-r2 <- r2[which(r2$dim <= 10),]
-r2$R.on.Clusters <- sqrt(r2$value)
-r3 <- cast2(r2, formula=norm~dataset, value.var="R.on.Clusters")
-r2m <- dplyr::summarise(group_by(r2,dataset), meanR=mean(R.on.Clusters))
-ggplot(r2, aes(norm, R.on.Clusters, fill=norm)) + geom_violin() + facet_wrap(~dataset, nrow=1) + coord_flip() + theme(legend.position="none") + stat_summary(fun.y="mean", geom="point") + geom_hline(data=r2m, aes(yintercept=meanR)) + ggtitle("Proportion of the top dimensions explained by subpopulations") + ylab("sqrt(R^2) of dimension ~ subpopulation") + theme(axis.text.x = element_text(angle = 90, hjust = 1))
-```
-
-### Supplementary Figure `r getFigNb()`
-
-Proportion of the variance in the first 10 principal components that is explained by real subpopulations according to different dimensionality reduction methods.
-
-\newpage
-
-# Supplementary Figure `r getFigNb(TRUE)`
-
-```{r}
-res <- readRDS("../data/dr_endSummary.rds")
-res2 <- res[which(abs(res$nbClusters-datasets[res$dataset,"subpopulations"])<=1),]
-ll <- lapply(split(res2, res2$norm), FUN=function(x){
-  m2 <- cast2(x, formula=dr~dataset, fun.aggregate=median, value.var="ARI")
-})
-
-chm(ll, scale="column", scaleTitle="z-score", sameScale=TRUE, value_format="%.2f", row_sortFn=function(x){ median(x,na.rm=T) }, cluster_columns=FALSE)
-```
-
-### Supplementary Figure `r getFigNb()`
-ARI of the clustering based on different normalization and dimensionality reduction procedures.
-
-\newpage
 
@@
 res2 <- res2[which(res2$dr=="seurat.pca" & res2$dataset!="Kumar"),]
-res3 <- aggregate(res2[,"ARI"], by=res2[,c("dataset","dims","resolution")], FUN=mean)
+res3 <- aggregate(res2[,"ARI",drop=FALSE], by=res2[,c("dataset","dims","resolution")], FUN=mean)
 ll <- lapply(split(res3, res3$dataset), FUN=function(x){

### 2020-01-29 supplementary_figures/supp6_clust.Rmd
updated supplementary figures

@@
   library(dplyr)
+  library(patchwork)
 })
-theme_set(theme_cowplot(font_size = 13))
-suppressMessages(devtools::load_all("../../pipComp/"))
-data(datasets)
+theme_set(theme_cowplot(font_size = 12))
+devtools::load_all("../../pipeComp/")
 source("../misc_functions.R")
@@
 
+```{r, include=FALSE}
+#res <- readRDS("../data/clust_endSummary.rds")
+res <- readRDS("../data/results_clust_2.rds")
+res <- parsePipNames(res$clustering)
+res <- res[which(res$norm=="norm.seurat" & res$resolution <5),]
+x <- grep("true\\.nbClusts",colnames(res),value=TRUE)
+names(x) <- x <- sapply(strsplit(x," "),FUN=function(x) x[[1]])
+y <- cbind(res[,c("k","steps","resolution","clustmethod","dims")],sapply(x, FUN=function(x){
+  res[[paste(x,"n_clus")]] - res[[paste(x,"true.nbClusts")]]
+}))
+y$param <- ifelse( res$clustmethod=="clust.seurat", paste0(" res=",res$resolution), paste0(" steps=",res$steps))
+ag <- aggregate(y[,names(x)], by=y[,c("clustmethod", "k","param"),drop=FALSE], FUN=mean)
+agl <- cbind(ag[rep(1:nrow(ag), 9),1:3], dataset=rep(colnames(ag)[4:ncol(ag)], each=nrow(ag)), delta=as.numeric(as.matrix(ag[,4:ncol(ag)])))
+agl$k <- factor(agl$k, as.character(sort(unique(agl$k))))
+x <- sapply(unique(strsplit(ag$param, "=")), FUN=function(x) as.numeric(x[[2]]) )
+names(x) <- unique(ag$param)
+agl$param <- factor(agl$param, names(sort(x)))
+
+ssq <- trans_new("ssq", function(x) sign(x)*sqrt(abs(x)), function(x) sign(x)*x^2)
+#slg <- trans_new("slg", function(x) sign(x)*log1p(abs(x)), function(x) sign(x)*(exp(x)-1))
+pf <- function(x) ggplot(x, aes(k, param, fill=delta)) + geom_tile() + 
+  facet_wrap(~dataset, nrow = 1) + labs(x="", fill="Difference to real number of clusters") + 
+  theme(axis.text.x=element_blank(), axis.ticks.x = element_blank(), axis.line.x = element_blank()) +
+  scale_fill_gradient2(low="darkblue", mid="white", high="red", na.value = "red", trans=ssq, limit=c(-7,20), breaks=c(-7,-2,0,2,7,20)) + 
+  theme(legend.position = "none")
+```
+
 ```{r, fig.width=8, fig.height=8}
-res <- readRDS("../data/clust_endSummary.rds")
-res <- res[which(res$dims>2 & res$resolution %in% c(0.05, 0.15, 0.3, 0.5, 0.8, 1, 1.2)),]
-res$method <- paste0( gsub("clust.","",res$clustmethod, fixed=T), 
-                      " k=",res$k, 
-                      ifelse( res$clustmethod=="clust.seurat", paste(" res=",res$resolution), paste0(" steps=",res$steps)) )
-res$diff.nbClusters <- res$nbClusters-datasets[res$dataset,"subpopulations"]
-ll <- lapply( unique(res$dataset), FUN=function(x){
-  as.matrix(cast2(res[which(res$dataset==x),], formula=method~dims, fun.aggregate=mean, value.var="diff.nbClusters"))
-})
-names(ll) <- unique(res$dataset)
-mmax <- ceiling(max(abs(sapply(ll,FUN=range))))
-cf <- circlize::colorRamp2(c(-mmax, 0, mmax), colors=c("blue", "white", "red"))
-chm(ll, sameScale=TRUE, scaleTitle="Difference\nto real nb\nsubpopulations\n", colors=cf, cluster_columns=FALSE, cluster_rows=FALSE)
+(pf(agl[which(agl$clustmethod=="clust.seurat"),]) + ylab("Seurat") + theme(legend.position = "top")) / 
+    (pf(agl[which(agl$clustmethod=="clust.scran"),]) + ylab("scran walktrap")) /
+    (pf(agl[which(agl$clustmethod=="clust.scran.fg"),]) + ylab("scran fast_greedy")) /
+    (ggplot(agl, aes(k)) + theme(axis.line.y=element_blank(), plot.margin = unit(rep(0,4),"pt"))) + 
+    plot_layout(heights = c(10,8,8,1))
 ```
@@
 ### Supplementary Figure `r getFigNb()`
-Difference between the number of detected clusters and the number of real subpopulations according to different numbers of dimensions and clustering paramters.
+Difference between the number of detected clusters and the number of real subpopulations according to different clustering paramters.
 

### 2020-01-29 supplementary_figures/supp_figures.tex
updated supplementary figures

@@
 \bibliographystyle{plainnat}
-\usepackage{longtable,booktabs}
 \usepackage{graphicx,grffile}
@@
   \postdate{\par}
-    \date{18 Mai, 2019}
+    \date{29 Januar, 2020}
 
@@
 \includegraphics{supp_figures_files/figure-latex/misclass-1.pdf} \vfill
-\includegraphics{supp_figures_files/figure-latex/unnamed-chunk-7-1.pdf}
+\includegraphics{supp_figures_files/figure-latex/unnamed-chunk-9-1.pdf}
 
@@
 
-\includegraphics{supp_figures_files/figure-latex/unnamed-chunk-8-1.pdf}
+\includegraphics{supp_figures_files/figure-latex/unnamed-chunk-10-1.pdf}
 \vfill
-\includegraphics{supp_figures_files/figure-latex/unnamed-chunk-9-1.pdf}
+\includegraphics{supp_figures_files/figure-latex/unnamed-chunk-11-1.pdf}
 
@@
 
-\includegraphics{supp_figures_files/figure-latex/unnamed-chunk-10-1.pdf}
+\includegraphics{supp_figures_files/figure-latex/unnamed-chunk-12-1.pdf}
 
@@
 
-\includegraphics{supp_figures_files/figure-latex/filt_more_ari-1.pdf}
+\includegraphics{supp_figures_files/figure-latex/filt_more-1.pdf}
 
@@
 
-\textbf{A:} Adjusted rand index (ARI) of various clustering pipelines
-using various combinations of filtering criteria. Only clustering
-analyses with a number of clusters +/-1 from the real number of clusters
-were considered. In general, filtering more does not result in more
-accurate clustering. \textbf{B:} Percentage of cells filtered out per
-subpopulation, using various combination of filters (see methods).
+Mean clustering F1 score per subpopulation, mean F1 at true number of
+clusters, as well as maximum and median proportion of excluded cells per
+subpopulation across various filtering strategies. Doublet removal
+generally improves clustering accuracy with very mild exclusion rates,
+even in datasets that do not have heterotypic doublets. Stringent
+distribution-based filtering creates large cell type biases.
 
@@
 
-\begin{verbatim}
-## lm(ARI~nbClusters*dataset+norm+factor(filt.times)+factor(filt.MADs)+counts.lower+counts.higher+mt.higher+feat.lower+feat.higher+top50.lower+top50.higher+ratiodist.lower+ratiodist.higher)
-\end{verbatim}
-
-\begin{longtable}[]{@{}lrrrr@{}}
-\toprule
-& Estimate & Std. Error & t value & p value\tabularnewline
-\midrule
-\endhead
-normnorm.seuratvst & 0.0069338 & 0.0003104 & 22.3415248 &
-0.0000000\tabularnewline
-factor(filt.times)2 & 0.0086024 & 0.0003617 & 23.7847962 &
-0.0000000\tabularnewline
-factor(filt.times)3 & 0.0168470 & 0.0003811 & 44.2112922 &
-0.0000000\tabularnewline
-factor(filt.MADs)2.5 & -0.0040702 & 0.0004199 & -9.6944075 &
-0.0000000\tabularnewline
-factor(filt.MADs)3 & -0.0054701 & 0.0004202 & -13.0177667 &
-0.0000000\tabularnewline
-factor(filt.MADs)5 & -0.0025099 & 0.0004205 & -5.9684908 &
-0.0000000\tabularnewline
-counts.lowerTRUE & -0.0001193 & 0.0003534 & -0.3375833 &
-0.7356780\tabularnewline
-counts.higherTRUE & -0.0073031 & 0.0003552 & -20.5607582 &
-0.0000000\tabularnewline
-mt.higherTRUE & 0.0006318 & 0.0007765 & 0.8136392 &
-0.4158539\tabularnewline
-feat.lowerTRUE & 0.0002074 & 0.0003494 & 0.5934348 &
-0.5528917\tabularnewline
-feat.higherTRUE & -0.0058113 & 0.0003506 & -16.5729653 &
-0.0000000\tabularnewline
-top50.lowerTRUE & -0.0000717 & 0.0003045 & -0.2354817 &
-0.8138354\tabularnewline
-top50.higherTRUE & -0.0000717 & 0.0003045 & -0.2354817 &
-0.8138354\tabularnewline
-ratiodist.lowerTRUE & -0.0009712 & 0.0003045 & -3.1895182 &
-0.0014256\tabularnewline
-ratiodist.higherTRUE & 0.0012337 & 0.0003045 & 4.0515679 &
-0.0000509\tabularnewline
-\bottomrule
-\end{longtable}
+\includegraphics{supp_figures_files/figure-latex/unnamed-chunk-15-1.pdf}
 
@@
 
-Linear regression analysis of ARI on various filtering criteria,
-correcting for the effect of the number of clusters. Increasing the
-number of distributions on which a cell must deviate in order to be
-excluded (\texttt{filt.times}) tended to be positively associated with
-ARI, while increasing the number of median absolute deviations required
-to be an outlier (\texttt{filt.MADs}) was associated with a decrease in
-ARI.
-
-\newpage
+Impact of restricting the type of features used on the ARI of the
+clustering. \newpage
 
@@
 
-\includegraphics{supp_figures_files/figure-latex/unnamed-chunk-14-1.pdf}
-
-\includegraphics{supp_figures_files/figure-latex/unnamed-chunk-15-1.pdf}
+\includegraphics{supp_figures_files/figure-latex/norm_nbClusters-1.pdf}
 
@@
 
-Impact of the set of filters on the average silhouette width of the
-subpopulations (top) and on the proportion of variance in the first
-components explained by clusters (bottom).
-
-\newpage
+Mean difference between the number of detected clusters and the number
+of real subpopulations, depending on the normalization method, the
+resolution and the number of dimensions used. The Kumar dataset is not
+shown here due to a lack of variation in the number of clusters
+detected. A rough ANOVA on
+\texttt{nbClusters\textasciitilde{}dataset+norm+dims+resolution}
+suggests that seuratvst (sctransform) is associated with a higher number
+of clusters (p\textasciitilde{}0.002).
 
@@
 
-\includegraphics{supp_figures_files/figure-latex/unnamed-chunk-16-1.pdf}
+\includegraphics{supp_figures_files/figure-latex/unnamed-chunk-19-1.pdf}
 
@@
 11}\label{supplementary-figure-11-1}}
-
-Impact of restricting the type of features used on the ARI of the
-clustering. \newpage
-
-\hypertarget{supplementary-figure-12}{%
-\section{Supplementary Figure 12}\label{supplementary-figure-12}}
-
-\includegraphics{supp_figures_files/figure-latex/norm_r2-1.pdf}
-
-\hypertarget{supplementary-figure-12-1}{%
-\subsubsection{Supplementary Figure
-12}\label{supplementary-figure-12-1}}
-
-\textbf{A:} Correlation between library size (left) or the number of
-detected features (right) and the residuals of a linear regression of
-the first principal component on the subpopulation. \textbf{B:}
-Proportion of the variance in the first 10 PCA components that is
-explained by real subpopulations. The square-root of R-squared values is
-plotted for ease of visualization.
-
-\newpage
-
-\hypertarget{supplementary-figure-13}{%
-\section{Supplementary Figure 13}\label{supplementary-figure-13}}
-
-\includegraphics{supp_figures_files/figure-latex/norm_silwidth-1.pdf}
-
-\hypertarget{supplementary-figure-13-1}{%
-\subsubsection{Supplementary Figure
-13}\label{supplementary-figure-13-1}}
-
-Mean silhouette width of true subpopulations across the first 10
-components with different normalization and scaling approaches.
-
-\newpage
-
-\hypertarget{supplementary-figure-14}{%
-\section{Supplementary Figure 14}\label{supplementary-figure-14}}
-
-\includegraphics{supp_figures_files/figure-latex/unnamed-chunk-21-1.pdf}
-
-\hypertarget{supplementary-figure-14-1}{%
-\subsubsection{Supplementary Figure
-14}\label{supplementary-figure-14-1}}
 
@@
 
-\hypertarget{supplementary-figure-15}{%
-\section{Supplementary Figure 15}\label{supplementary-figure-15}}
+\hypertarget{supplementary-figure-12}{%
+\section{Supplementary Figure 12}\label{supplementary-figure-12}}
 
 \begin{verbatim}
-## Warning: Removed 328867 rows containing missing values (geom_path).
-
 ## Warning: Removed 328867 rows containing missing values (geom_path).
@@
 
-\includegraphics{supp_figures_files/figure-latex/unnamed-chunk-22-1.pdf}
+\includegraphics{supp_figures_files/figure-latex/unnamed-chunk-21-1.pdf}
 
-\hypertarget{supplementary-figure-15-1}{%
+\hypertarget{supplementary-figure-12-1}{%
 \subsubsection{Supplementary Figure
-15}\label{supplementary-figure-15-1}}
+12}\label{supplementary-figure-12-1}}
 
@@
 
-\hypertarget{supplementary-figure-16}{%
-\section{Supplementary Figure 16}\label{supplementary-figure-16}}
+\hypertarget{supplementary-figure-13}{%
+\section{Supplementary Figure 13}\label{supplementary-figure-13}}
 
 \begin{verbatim}
-## Warning: Removed 328867 rows containing missing values (geom_path).
-
 ## Warning: Removed 328867 rows containing missing values (geom_path).
@@
 
+\hypertarget{supplementary-figure-13-1}{%
+\subsubsection{Supplementary Figure
+13}\label{supplementary-figure-13-1}}
+
+Proportion of the cumulative \emph{deviance} explained by real
+subpulations that is retrieved through the selection. For each gene, we
+compute the proportion of the variance explained by real subpopulations.
+As for Supplementary Figure 12, except using deviance explained.
+
+\newpage
+
+\hypertarget{supplementary-figure-14}{%
+\section{Supplementary Figure 14}\label{supplementary-figure-14}}
+
+\includegraphics{supp_figures_files/figure-latex/unnamed-chunk-24-1.pdf}
+
+\hypertarget{supplementary-figure-14-1}{%
+\subsubsection{Supplementary Figure
+14}\label{supplementary-figure-14-1}}
+
+Clustering accuracy according to the number of genes selected using
+various ranking/selection methods. \textbf{A:} Based on sctransform,
+\textbf{B:} Based on standard Seurat normalization.
+
+\newpage
+
+\hypertarget{supplementary-figure-15}{%
+\section{Supplementary Figure 15}\label{supplementary-figure-15}}
+
+\includegraphics{supp_figures_files/figure-latex/unnamed-chunk-27-1.pdf}
+
+\hypertarget{supplementary-figure-15-1}{%
+\subsubsection{Supplementary Figure
+15}\label{supplementary-figure-15-1}}
+
+Mean difference between the number of detected clusters and the number
+of real subpopulations, depending on the resolution and number of
+dimensions used. Based on sctransform and seurat PCA. Increasing the
+number of dimensions tends to decrease the number of identified
+clusters, especially at resolutions around the default value.
+
+\newpage
+
+\hypertarget{supplementary-figure-16}{%
+\section{Supplementary Figure 16}\label{supplementary-figure-16}}
+
+\includegraphics{supp_figures_files/figure-latex/unnamed-chunk-28-1.pdf}
+
 \hypertarget{supplementary-figure-16-1}{%
@@
 
-Proportion of the cumulative \emph{deviance} explained by real
-subpulations that is retrieved through the selection. For each gene, we
-compute the proportion of the variance explained by real subpopulations.
-As for Supplementary Figure 15, except using deviance explained.
-
-\newpage
+Adjusted Rand Index of clustering depending on the resolution and number
+of dimensions used. Based on sctransform and seurat PCA.
 
@@
 
-\includegraphics{supp_figures_files/figure-latex/unnamed-chunk-24-1.pdf}
+\includegraphics{supp_figures_files/figure-latex/unnamed-chunk-31-1.pdf}
 
@@
 
-Proportion of the variance in the first 5 principal components that is
-explained by real subpopulations according to different normalization
-and selection procedures, selecting 1000 genes in each case.
+Difference between the number of detected clusters and the number of
+real subpopulations according to different clustering paramters.
 

### 2020-01-29 bmc_article.tex

@@
 
-\title{scRNAseq benchmark in want of a title}
+\title{A framework for pipeline benchmarking and its application to single-cell RNAseq clustering}
 
@@
    addressref={aff1,aff2},          % id's of addresses, e.g. {aff1,aff2}
-   corref={aff1},                       % id of corresponding address, if any
    email={pierre-luc.germain@hest.ethz.ch}   % email address
 ]{\inits{PLG}\fnm{Pierre-Luc} \snm{Germain}}
+\author[
+   addressref={aff1},
+    email={anthony.sonrel@uzh.ch}
+]{\fnm{Anthony} \snm{Sonrel}}
 \author[
@@
 \begin{abstract} % abstract
-\parttitle{Background} %if any
-Text for this section.
-\parttitle{Results} %if any
-Text for this section.
-\parttitle{Conclusions}
-Text for this section.
+%\parttitle{Background} %if any
+We present a flexible R framework for pipeline comparison with multi-level evaluation metrics, and apply it to the benchmark of single-cell RNAseq analysis pipelines using datasets with known cell identities.
+We discuss common steps of such analyses, including filtering and doublet detection (offering a new package), normalization, feature selection, denoising, dimensionality reduction and clustering.
+On the basis of these analyses, we make a number of concrete key recommendations about analysis choices.
 \end{abstract}
@@
 
-Single-cell RNA-sequencing (scRNAseq) and the set of attached analysis methods are evolving fast, and while a number of good comparison/benchmark studies can inform analytical decisions (\citealp{duoClustering2018, SonesonDE2018, SunDimRed2019}) they need constant updating and often leave open many details of an analysis.
-Here, we go in more detail on the various steps of analysis leading from an initial count matrix to a cluster assignment, which are critical steps in a wide range of applications.
+Single-cell RNA-sequencing (scRNAseq) and the set of attached analysis methods are evolving fast, with more than 560 software tools available to the community \cite{ZappiaDB2018}, roughly half of which are dedicated to tasks related to data processing such as clustering, ordering, dimension reduction or normalization.
+This increase in the number of available tools follows the development of new sequencing technologies and the growing number of reported cells, genes and cell populations \cite{SvenssonDB2019}.
+Data processing is a critical step in any scRNAseq analysis as these data are affected by various technical effects (batch effect, dropout, ...) that have to be corrected to distinguish the biological variation form the technical variation.
+The choice of the scRNAseq protocol is already of prime importance as different protocols differ greatly in view of parameters such as gene detection or mappability \cite{MereuCellAtlas2019}, which can in turn affect downstream analysis.
+In view of the intrinsic challenges of scRNAseq technology and the growing number of tools and studies complexity, it is critical to evaluate the available tools to leverage future studies as it can ultimately increase the power of further differential expression analysis \cite{viethSystematic2019}.
+
+While a number of good comparison and benchmark studies have already been performed on various steps related to scRNAseq analysis and can inform analytical decisions (\citealp{duoClustering2018, SonesonDE2018, SunDimRed2019}), they need constant updating and often leave open many details of an analysis.
+Another missing aspect of current benchmarking studies is their limitation to capture all aspects of scRNAseq processing workflow.
+Although previous benchmarks already brought valuable recommendations for data processing, some only focused on one aspect of data processing (e.g., \citealp{SunDimRed2019}), did not evaluate the effect of the choice of a tool on downstream analysis (e.g., \citealp{TsuyuzakiPCA2020}) or did not tackle all aspects of data processing, such as doublet identification or cell filtering (e.g., \citealp{viethSystematic2019}).
+A thorough evaluation of the tools covering all major processing steps is however urgently needed as previous benchmarking studies highlighted that a combination of tools can have a drastic impact on downstream analysis, such as differential expression analysis and cell-type deconvolution (\citealp{viethSystematic2019, CobosDeconvolution2020}).
+It is then likely that processing algorithms also interact in a positive or negative fashion and it is then critical to evaluate not only the single effect of a method but also its interaction with all parts of a workflow.
+
+Here, we develop a flexible R framework for pipeline comparison and evaluate in details the various steps of analysis leading from an initial count matrix to a cluster assignment, which are critical in a wide range of applications.
 We harness real datasets of known cell composition (Table \ref{tab:table1}) and develop new evaluation metrics to investigate in a multilevel fashion the impact of various parameters and variations around a core scRNAseq pipeline.
-
 Although we use some datasets based on other protocols, our focus is especially on droplet-based datasets that do not include exogenous control RNA (i.e.
 spike-ins); see Table \ref{tab:table1} and \ref{fig:figure1} for a description of the datasets.
-In addition to previously-used benchmark datasets with true cell labels \cite{duoClustering2018,tianMixology2018}, we simulated two datasets with a hierarchical subpopulation structure based on real 10x data using \textit{muscat} [REF].
+In addition to previously-used benchmark datasets with true cell labels \cite{duoClustering2018,tianMixology2018}, we simulated two datasets with a hierarchical subpopulation structure based on real 10x human and mouse data using \textit{muscat} (\cite{CrowellMuscat2019})).
 Since graph-based clustering \cite{satijaSeurat2015} was previously shown to consistently perform well across several datasets \cite{duoClustering2018,tianMixology2018}, we used the \texttt{Seurat} pipeline as the starting point to perform an integrated investigation of: 1) doublet identification, 2) cell filtering, 3) normalization, 4) feature selection, 5) dimension reduction, 6) clustering.
-We compare not only competing approaches, but also more fine-grained parameter and variations on common methods.
+We compare not only competing approaches, but also more fine-grained parameters and variations on common methods.
 Importantly, the success of methods at a certain analytical step might be dependent on choices at other steps.
@@
 In addition, optional benchmark functions can be set for each step to provide standardized, multi-layered evaluation metrics.
-Given such a \texttt{PipelineDefinition} object, a set of alternative parameters (which might include different subroutines) and benchmark datasets, the \texttt{runPipeline} function then proceeds through all combinations arguments, avoiding recomputing the same step twice and compiling evaluations on the fly to avoid storing potentially large intermediate data.
+Given such a \texttt{PipelineDefinition} object, a set of alternative parameters (which might include different subroutines) and benchmark datasets, the \texttt{runPipeline} function then proceeds through all combinations of arguments, avoiding recomputing the same step twice and compiling evaluations on the fly to avoid storing potentially large intermediate data.
 Variations in a given parameter can be evaluated using all metrics from this point downward in the pipeline.
@@
 In some cases, doublets can be identified through their unusually high number of reads and detected features, but this is not always the case (Supplementary Figure 2).
-A number of methods were developed to identify doublets, in most cases by comparing each cell to artificially-created doublets [REFs].
+A number of methods were developed to identify doublets, in most cases by comparing each cell to artificially-created doublets \citep{mcginnisDoubletfinder2019, LunScran2016, BaisScds2019}.
 We therefore first evaluated the capacity of these methods to detect doublets using the two 10x datasets produced by \citep{tianMixology2018} with cells of different genetic identity, and for which therefore SNP information can be used as ground truth.
-We tested \texttt{DoubletFinder} [REF] and \texttt{scran}'s \texttt{doubletCells} [REF], both of which use similarity to artificial doublets, and \texttt{scds} [REF], which relies on a combination of co-expression and binary classification.
+We tested \texttt{DoubletFinder} \citep{mcginnisDoubletfinder2019} and \texttt{scran}'s \texttt{doubletCells} \citep{LunScran2016}, both of which use similarity to artificial doublets, and \texttt{scds} \citep{BaisScds2019}, which relies on a combination of co-expression and binary classification.
 \texttt{DoubletFinder} integrates a thresholding based on the proportion of expected doublets, while \texttt{scran} and \texttt{scds} return scores that must be manually thresholded.
@@
 doublets from different cell types) more efficiently.
-We therefore developed a simple and fast package implementing this for doublet detection, with the added advantage of accounting for uncertainty in the expected doublet rate and using meta-cells from the clusters to even include triplets (see methods).
+We therefore developed a simple and fast Biocondcutor package implementing this for doublet detection, \texttt{scDblFinder}, with the added advantage of accounting for uncertainty in the expected doublet rate and using meta-cells from the clusters to even include triplets (see methods).
 
-While most methods accurately identified the doublets in the 3 cell lines dataset (mixology10x3cl), the 5 lines dataset (mixology10x5cl) proved more difficult (Figure \ref{fig:figure3}B).
-Our method (\texttt{scDblFinder}) achieved the best accuracy while running in a very reasonable time (Figure \ref{fig:figure3}).
-Across datasets, cells called as doublets tended to be classified in the wrong cluster more often than other cells (Figure 1D).
+While most methods accurately identified the doublets in the 3 cell lines dataset (mixology10x3cl), the other two datasets proved more difficult (Figure \ref{fig:figure3}A).
+\texttt{scDblFinder} achieved comparable or better accuracy than top alternatives while running in a very reasonable time (Figure \ref{fig:figure3}B).
+Across datasets, cells called as doublets tended to be classified in the wrong cluster more often than other cells (Figure \ref{fig:figure3}C.
 We therefore tested whether this method also improved the accuracy of the clustering across all benchmark datasets, and found that it did even when, by design, the data contained no heterotypic doublet (Figure \ref{fig:figure4}).
@@
     \caption{\textbf{Identification of doublet cells.
-A-B:} Receiver operating characteristic (ROC) curves of the tested doublet detection methods for the mixology10x3l (\textbf{A}) and mixology10x5cl (\textbf{B}) datasets.
-DoubletFinder failed due to an error on the mixology10x3cl dataset.
-\textbf{C:} Running time of the different methods.
-\textbf{D} Cell called as doublets by scDblFinder tend to be more often misclassified, i.e.
-assigned a wrong cluster.
-The corrected percentage of misclassification is the proportion of times a given cell is misclassified across hundreds of clustering pipelines with various parameters, minus the median misclassification rate of the cell's subpopulation.}
-    \label{fig:figure3}
+A:} Receiver operating characteristic (ROC) curves of the tested doublet detection methods for three datasets with SNP-identified doublets.
+Dots indicate threshold determined by the true number of doublets.
+\textbf{B:} Running time of the different methods (DoubletFinder failed on one of the datasets).
+\textbf{C:} Rate of misclassification of the cells identified by scDblFinder as doublets (DBL) or singlets (SNG), across a large range of clustering analysis.
+Even in datasets which should not have neotypic doublets, the cells identified as such tend to be misclassified.
+\label{fig:figure3}}
 \end{figure}
@@
 Beyond doublets, a dataset might include low-quality cells whose elimination would reduce noise.
-This has for instance been demonstrated for cells that have a high content of mitochondrial reads, often as a result of cell degradation [REF].
-A common practice is to exclude cells that differ considerably from most other cells on the basis of such properties, for instance through the \texttt{isOutlier} function of \texttt{scater} that measures, for a given control property, the number of median absolute deviations of each cell from the mean of all cells.
+This has for instance been demonstrated for droplets containing a high content of mitochondrial reads, often as a result of cell degradation and resulting loss of cytoplasmic mRNAs \citep{IlicicLowQual2016}.
+A common practice is to exclude cells that differ considerably from most other cells on the basis of such properties, for instance through the \texttt{isOutlier} function of \texttt{scater} that measures, for a given control property, the number of median absolute deviations of each cell from the median of all cells.
 Supplementary Figure 1 shows the distributions of some of the typical cell properties commonly used.
-Of note, these properties tend to be correlated, but not always: for instance, while a high proportion of mitochondrial reads is often correlated with a high proportion of the counts in the top features, there can also be other reasons for an over-representation of highly-expressed features (Supplementary Figure 3).
+Of note, these properties tend to be correlated, but not always: for instance, while a high proportion of mitochondrial reads is often correlated with a high proportion of the counts in the top features, there can also be other reasons for an over-representation of highly-expressed features (Supplementary Figure 3), such as an over-amplification in non-UMI datasets.
 In addition, in our experience 10X datasets also exhibit a very tight correlation between the total counts and the total features even across very different cell types (Supplementary Figure 4).
@@
 library size) is that these properties tend to have different distributions across subpopulations.
-As a result, thresholds in terms of number of median absolute deviations (MADs) from the whole distribution can lead to strong biases against certain subpopulations (Figure \label{fig:figure4}A).
-We therefore examined the tradeoff between the increased accuracy of filtering and the maximum proportion of cells excluded per subpopulation (Figure \label{fig:figure4}B and Supplementary Figure 8).
-Since filtering changes the relative abundance of the different subpopulations, global clustering accuracy metrics such as ARI are not appropriate here, and we therefore calculated the per-subpopulation precision and recall using the Hungarian algorithm, and monitored their mean F1 score.
+As a result, thresholds in terms of number of median absolute deviations (MADs) from the whole distribution can lead to strong biases against certain subpopulations (Figure \ref{fig:figure4}A).
+We therefore examined the tradeoff between the increased accuracy of filtering and the maximum proportion of cells excluded per subpopulation (Figure \ref{fig:figure4}B and Supplementary Figure 8).
+Since filtering changes the relative abundance of the different subpopulations, global clustering accuracy metrics such as ARI are not appropriate here, and we therefore calculated the per-subpopulation precision and recall using the Hungarian algorithm \citep{PapadimitriouHu1998}, and monitored their mean F1 score.
 A first observation was that, although the stringent filtering tended to be associated with an increase in accuracy, it could also become deleterious, and most of the benefits could be achieved without very stringent filtering and minimizing subpopulation bias.
-Applying the same filtering criteria on individual clusters of cells (identified through scran's \texttt{quickCluster} method) resulted in nearly no cell being filtered out, which suggests that filtering on the global population tends to discard cells of subpopulations with more extreme properties, rather than `bad` cells.
+Applying the same filtering criteria on individual clusters of cells (identified through scran's \texttt{quickCluster} method) resulted in nearly no cell being filtered out.
+This suggests that filtering on the global population tends to discard cells of subpopulations with more extreme properties, such as high library size, rather than low-quality cells.
 Finally, by changing the distributions, the doublet removal step in conjunction with filtering sometimes resulted in a net decrease in the proportion of excluded cells while retaining or improving accuracy.
@@
 
-Mitochondrial reads have been associated with cell degradation, and it has been observed that much of scRNAseq clustering is determined by ribosomal genes [REF].
-We therefore considered whether excluding one category of features or the other, or using only protein-coding genes, had impact on the ability to distinguish subpopulations (Supplementary Figure 9).
+Mitochondrial reads have been associated with cell degradation and there is evidence that ribosomal genes can influence the clustering output, hiding other biological structure in the analysis \citep{freytagComparison2018}.
+We therefore investigated whether excluding one category of features or the other, or using only protein-coding genes, had an impact on the ability to distinguish subpopulations (Supplementary Figure 9).
 Removal of ribosomal genes robustly reduced the quality of the clustering, suggesting that they represent real biological differences between subpopulations.
@@
 Beside the standard log-normalization included in \texttt{Seurat}, we tested \texttt{scran}'s pooling-based normalization \citep{lunPooling2016}, \texttt{sctransform}'s variance-stabilizing transformation \citep{hafemeisterSCtransform2019}, and normalization based on stable genes \citep{linStableGenes2018, deekeStablyExpressed2018}.
-In addition to log-normalization, the standard \texttt{Seurat} clustering pipeline performs per-feature unit-variance scaling so that the PCA is not too dominated by highly-expressed features, and we therefore included versions of the different normalization, with or without a subsequent scaling (\texttt{sctransform}'s variance-stabilizing transformation involves something analogous to scaling).
-\texttt{Seurat}'s scaling function also includes the option to regress out the effect of certain covariates, and we therefore tested its use with the proportion of mitochondrial counts and the number of detected features.
-Finally, it has been proposed that the use of stable genes, in particular cytosolic ribosomal genes, can be used to normalize scRNAseq; we therefore evaluated a simple linear normalization based on the sum of these genes, as well as nuclear genes.
+In addition to log-normalization, the standard \texttt{Seurat} clustering pipeline performs per-feature unit-variance scaling so that the PCA is not too strongly dominated by highly-expressed features.
+We therefore included versions of the different normalization procedures, with or without a subsequent scaling (\texttt{sctransform}'s variance-stabilizing transformation involves something analogous to scaling).
+\texttt{Seurat}'s scaling function also includes the option to regress out the effect of certain covariates.
+We tested its performance by using the proportion of mitochondrial counts and the number of detected features as covariates.
+Finally, it has been proposed that the use of stable genes, in particular cytosolic ribosomal genes, can be used to normalize scRNAseq\citep{deekeStablyExpressed2018}.
+We therefore evaluated a simple linear normalization based on the sum of these genes, as well as nuclear genes.
 
-An important motivation of sctransform was the observation that, even after normalization, the first principal components of various datasets tended to correlate with library size, suggesting an inadequate normalization \citep{hafemeisterSCtransform2019}.
-We therefore first assessed to what extent the first principal component still retained a correlation with the library size and number of detected features, removing the confounding covariation with the subpopulation (Figure \label{fig:figure5}A).
-The simple step of scaling tended to remove much of the correlation with these features, and in the absence of scaling, standard \texttt{Seurat} normalization resulted in fairly high correlation with technical covariates.
-\texttt{sctransform} had the lowest correlation, but most methods (including normalization based on stable genes) were able to remove most of the correlation when combined with scaling.
-The only exception is \texttt{Seurat} normalization with scaling regressing out the number of features, which surprisingly increased the correlation.
+An important motivation for the development \texttt{sctransform} was the observation that, even after normalization, the first principal components of various datasets tended to correlate with library size, suggesting an inadequate normalization \citep{hafemeisterSCtransform2019}.
+We therefore first assessed to what extent the first principal component still retained a correlation with the library size and the number of detected features, removing the confounding covariation with the subpopulations (Figure \ref{fig:figure5}A).
+The simple step of scaling tended to remove much of the correlation with these features and, in the absence of scaling, standard \texttt{Seurat} normalization resulted in fairly high correlation with technical covariates.
+\texttt{sctransform} led to the lowest correlation but most methods, including normalization based on stable genes, were able to remove most of the correlation when combined with scaling.
+The only exception is \texttt{Seurat} normalization with scaling regressing out covariates, which surprisingly increased the correlation in 8 of the 9 datasets.
 
-We next investigated the impact of normalization on the separability of the subpopulations (Figure \label{fig:figure5}B-C), and found most methods (including no normalization at all) to perform fairly well.
+We next investigated the impact of normalization on the separability of the subpopulations (Figure \ref{fig:figure5}B-C).
 Since clustering accuracy metrics such as the ARI are very strongly influenced by the number of clusters, we complemented it with silhouette width and mutual information.
+We found most methods (including no normalization at all) to perform fairly well in most of the subpopulations.
 Scaling tended to reduce the average silhouette width of some subpopulations and to increase that of some less distinguishable ones, and was generally but not always beneficial on the accuracy of the final clustering.
 Regressing out covariates systematically gave poorer performance on all metrics.
-sctransform systematically outperformed other methods, and even though it was developed to be applied to data with unique molecular identifiers (UMI), it also performed fairly well with smartseq datasets.
+\texttt{sctransform} systematically outperformed other methods, and even though it was developed to be applied to data with unique molecular identifiers (UMI), it also performed fairly well with Smart-seq protocol (Koh and Kumar datasets).
 
@@
     \caption{\textbf{Evaluation of normalization procedures.
+`Seurat.feat_mt_regress' indicates the regressing out of the number of features and proportion of mitochondrial reads during scaling, represents, and `Seurat.feat_regress` of the number of features only.
 A:} Residual correlation of the first principal component with library size (left) and detection rate (right), after accounting for biological differences between subpopulations.
@@
 
-Finally, although some normalization methods had a tendency to lead to a higher (e.g.
+Finally, we monitored whether, under the same downstream clustering analysis, different normalization methods tended to lead to an over- or under-estimation of the number of clusters.
+Although some methods had a tendency to lead to a higher (e.g.
 \texttt{sctransform}) or lower (e.g.
-stable genes) number of clusters, the effect was very mild and not entirely systematic, and we observed no major tendency towards over- or under-estimating the number of clusters (Supplementary Figure 10).
+stable genes) number of clusters, the effect was very mild and not entirely systematic (Supplementary Figure 10).
 
@@
 
-A standard clustering pipeline typically involves a step of selecting highly-variable genes, which is complicated by the digital nature and the mean-variance relationship of RNAseq.
-\texttt{Seurat}'s earlier approaches involved the use of dispersion estimates standardized for the mean expression levels, while more recent versions rely of a different measure of variance, again standardized.
-Of note, while adjusting for the mean-variance relationship removes much of the bias towards highly-expressed genes, it is plausible that this bias reflects biological relevance and hence is actually helpful in classifying cell types.
-Another common practice for selecting features is to use those with the highest mean expression, and recently \citep{townesGlmpca2019} instead suggested to use deviance, while \texttt{sctransform} provides its own ordering the genes by transformed variance.
-We therefore compared these methods.
+A standard clustering pipeline typically involves a step of highly-variable genes selection, which is complicated by the digital nature and the mean-variance relationship of (sc)RNAseq.
+\texttt{Seurat}'s earlier approaches involved the use of dispersion estimates standardized for the mean expression levels, while more recent versions (>3) rely on a different measure of variance, again standardized.
+Of note, while adjusting for the mean-variance relationship removes much of the bias towards highly-expressed genes, it is plausible that this relationship may in fact sometimes reflects biological relevance and would be helpful in classifying cell types.
+Another common practice in features selection is to use those with the highest mean expression, and recently, \citep{townesGlmpca2019} instead suggested to use deviance, while \texttt{sctransform} provides its own ordering of genes based on transformed variance.
 
-Reasoning that a selection method should ideally select genes whose variability is more between subpopulations than within, we first assessed whether each selection method selected genes with a high proportion of variance or deviance explained by (real) subpopulation.
-Estimates of variance explained by subpopulation based on a standard \texttt{Seurat} normalization or on sctransform data were highly correlated (Supplementary Figure 11A), and there was also a good agreement with deviance explained, although lowly-expressed genes in particular could have a high deviance explained without having much of their variance explained by subpopulation (Supplementary Figure 11B-D).
-We therefore compared, using both of these estimates, the proportion of the cumulative variance/deviance explained by the top X genes that could be retrieved by each gene ranking method (Supplementary Figures 12-13).
-We first focused on the first 1000 genes to highlight differences between methods, although the differences tended to decrease with a higher number of genes selected (Supplementary Figures 12-14).
-A first observation was that standardized measures of variability were systematically worse than their non-standardized  counterparts in selecting genes with a high proportion of variance explained by subpopulation, while instead proving most of the time superior in selecting genes with a high deviance explained by subpopulation (Figure \ref{fig:figure6}A and Supplementary Figures 12-13).
-Deviance proved the method of choice to maximize the former (with mere expression level proving surprisingly good), but surprisingly did not perform so well to select genes with a high deviance explained.
+Reasoning that a selection method should ideally select genes whose variability is more between subpopulations than within, we first assessed whether a method selected genes with a high proportion of variance or deviance explained by (real) subpopulation.
+As the proportion of variability in a gene attributable to subpopulations can be measured in various ways, we first compared three such alternatives.
+Analysis of variance performed on a standard \texttt{Seurat} normalization or on \texttt{sctransform} data were highly correlated (Supplementary Figure 11A).
+These estimates were also in good agreement with the deviance explained, although lowly-expressed genes could have a high deviance explained without having much of their variance explained by subpopulation (Supplementary Figure 11B-D).
+We therefore compared the proportion of the cumulative variance/deviance explained by the top X genes that could be retrieved by each gene ranking method (Supplementary Figures 12-13).
+We first focused on the first 1000 genes to highlight the differences between the methods, although a higher number of selected genes decreased the differences between the methods (Supplementary Figures 12-14).
+The standardized measures of variability were systematically worse than their non-standardized counterparts in selecting genes with a high proportion of variance explained by subpopulation.
+Regarding to the percentage of deviance explained however, the standardized measures were often superior(Figure \ref{fig:figure6}A and Supplementary Figures 12-13).
+Deviance proved the method of choice to prioritize genes with a high variance explained by subpopulations (with mere expression level proving surprisingly good), but surprisingly did not perform so well to select genes with a high deviance explained.
 
@@
 A:} Ability of different feature ranking methods to capture genes with a high proportion of variance (left) or deviance (right) explained by real subpopulations.
+The asterix denotes the default Seurat method.
 \textbf{B:} Accuracy of clusterings (at the true number of clusters) when selecting 1000 genes using the given methods.
@@
 
-We next investigated whether a selection based on these methods impacted the clustering accuracy (Figure \ref{fig:figure6}A).
-To validate the previous assay, we included a selection based on the genes whose proportion variance or deviance explained was highest.
+We next evaluated how the use of the different feature selection methods affected the clustering accuracy (Figure \ref{fig:figure6}B).
+To validate the previous assay on the proportion of variance/ deviance explained by real populations, we included genes that maximized these two latter measures.
 Interestingly, while these selections were on average the top-ranking methods, they were not systematically best for all datasets.
-The previous observations were reflected in the ARI of the resulting clustering (Figure 4B): non-standardized measures of variability, including mere expression level, tended to outperform more complex metrics.
+The previous observations were reflected in the ARI of the resulting clustering (Figure \ref{fig:figure6}B): non-standardized measures of variability, including mere expression level, tended to outperform more complex metrics.
 In general, we found deviance and unstandardized estimates of variance to provide the best results across datasets and normalization methods.
@@
 
-We next investigated the step of dimensionality reduction, comparing in particular \texttt{Seurat}'s PCA, \texttt{scran}'s \texttt{denoise PCA}, and GLM-PCA \citep{townesGlmpca2019}, combined (where relevant) with sctransform normalization.
-\texttt{Seurat}'s default PCA weighs the cell embeddings by the variance of each component, so we also tested the impact of this weighting with each method.
+We investigated the methods of dimensionality reduction.
+Since the various PCA approaches and implementations were recently benchmarked in a similar context \citep{TsuyuzakiPCA2020}, we focused on widely used approaches which had not yet been compared \texttt{Seurat}'s PCA, \texttt{scran}'s \texttt{denoise PCA}, and GLM-PCA \citep{townesGlmpca2019} and combined (when relevant) with \texttt{sctransform} normalization.
+Given that \texttt{Seurat}'s default PCA weights the cell embeddings by the variance of each component, we also evaluated the impact of this weighting with each method.
 
-The impact of the dimensionality reduction method was far greater than that of normalization or feature selection (e.g.
+The impact of the choice of dimensionality reduction method was far greater than that of normalization or feature selection (e.g.
 Supplementary Figure 15).
-Although the first components identified by GLM-PCA tended to have a higher proportion of their variance explained by real subpopulations and increase the average silhouette width of already well-defined subpopulations, overall Seurat's PCA procedure proved superior on all metrics (Figure \ref{fig:figure7} and Supplementary Figures 15-17).
-In particular, weighing the components had a strong positive impact on silhouette widths and resulting ARI.
+The first components identified by GLM-PCA tended to have a higher proportion of their variance explained by real subpopulations and increase the average silhouette width of already well-defined subpopulations.
+Seurat's PCA procedure however proved superior on all metrics (Figure \ref{fig:figure7} and Supplementary Figures 15-17).
+Overall, weighting the principal components by their variance in the same fashion as Seurat's method does had a strong positive impact on silhouette widths and ARI scores.
 
@@
     \includegraphics[width=\textwidth]{{main_figures/dr_files/figure-html/dimred-1.png}}
-    \caption{\textbf{Evaluation of the dimensionality reduction strategies.
-A:} Average silhouette width per subpopulation according to various normalization and dimension reductions.
+    \caption{\textbf{Evaluation of common dimensionality reduction methods.
+A:} Average silhouette width per subpopulation resulting from combinations of normalization and dimension reductions.
 \textbf{B:} Clustering accuracy , mean proportion of the variance in the first 5 components explained by real subpopulations (center), and median ARI of the resulting clustering (right).}
@@
 
-We next investigated the impact of the number of dimensions used on the separability of the subpopulations.
-Euclidean distance decreases as the number of non-discriminating dimensions increases, suggesting a tradeoff between the amount of information and the proportion of relevant dimensions.
+A common step following dimension reduction is the selection of an appropriate number of dimensions to use for downstream analysis.
+Since euclidean distance decreases as the number of non-discriminating dimensions increases, there is usually a trade-off between selecting enough dimensions to keep most information and excluding smaller dimensions that may represent technical noise or other unwanted sources of variation.
 Overall, increasing the number of dimensions robustly led to a decrease in the number of clusters, although the effect was milder with data processed through \texttt{sctransform} (Supplementary Figures 14 and 22).
 This tended to be translated to the accuracy of the clustering (Supplementary Figure 23), although in both cases (number of clusters and ARI) the resolution parameter had a much stronger impact.
-We then compared various methods to estimate the intrinsic dimension based on the reduced space (from Seurat weighted PCA).
-As a first approximation of the real dimension, we computed the variance in each principal component that was explained by the subpopulations; this dropped dramatically after the first few components, suggesting an overall low dimensionality (Figure \ref{fig:figure8}A) at least for the purpose of cell type definition.
-We next estimated dimensions using various methods (see methods), some of which are already commonly used for scRNAseq, while others were borrowed from other fields.
-Figure \ref{fig:figure8}B shows the difference between the dimension estimates of these methods and that based on the subpopulations (i.e.
+
+Different approaches have been proposed to select the appropriate number of dimensions, from the visual identification of an 'Elbow' (inflexion point) of the variance explained, to more complex algorithms.
+We evaluated the performance of common tools, namely [TBD] (\texttt{pcaLocal.FO} label, [REF needed?]), global maximum likelihood based on translated Poisson Mixture Model (\texttt{maxLikGlobal}, \citep{haroTranslated2008}), [TBD] (\texttt{essLocal.b}, \texttt{essLocal.b},[REF needed?]), JackStraw procedure (\texttt{jackstraw.elbow}, \citep{ChungJackstraw2015}), [TBD](\texttt{pcaLocal.maxgap}, [REF needed?]), [TBD] (\texttt{pcaLocal.fan} [REF needed?]), Fisher Separability analysis (\texttt{fisherSeparability}, \citep{AlberganteSepar2019}), \texttt{scran}'s \texttt{denoisePCA} function (\texttt{scran.denoisePCA}, \citep{LunScran2016}) and [TBD] (\texttt{pcaOptmPointwise.max}, [REF needed?]).
+
+We compared the various estimates of dimensionality in their ability to retrieve the intrinsic number of dimensions in a dataset, based on Seurat's weighted PCA space.
+As a first approximation of the true dimensionality, we computed the variance in each principal component that was explained by the subpopulations, which sharply decreased after the first few components in most datasets (Figure \ref{fig:figure8}A).
+Figure \ref{fig:figure8}B shows the difference between the dimension estimates of the above methods and that based on the subpopulations (i.e.
 from Figure \ref{fig:figure8}A).
-Of note, the estimates differ widely in terms of their computing time (Figure \ref{fig:figure8}B), and we saw no relationship between the accuracy of the estimate and the complexity of the method.
-Reasoning that over-estimating dimension is less problematic than under-estimating it, we eliminated methods below the estimate and selected candidate methods for a full analysis of their impact on clustering (Figure \ref{fig:figure8}C-D) in combination with both sctransform or normal Seurat normalization.
-Although most methods performed well, the global maximum likelihood based on translated Poisson Mixture Model \citep{haroTranslated2008}, i.e.
-maxLikGlobal, provided the dimensionality estimate that best separated the subpopulations (Figure \ref{fig:figure8}C) and resulted in the best clustering accuracy (Figure \ref{fig:figure8}D).
-Of note, this method systematically estimated dimensions above those for which a large part of the variance is explained by subpopulations, suggesting that these components help classification despite being individually uninformative.
+Of note, the methods differ widely in view of their computing time (Figure \ref{fig:figure8}B) and we saw no relationship between the accuracy of the estimates and the complexity of the method.
+Reasoning that over-estimating the number of dimensions is less problematic than under-estimating it, we kept the former methods for a full analysis of their impact on clustering (Figure \ref{fig:figure8}C-D), when combined with sctransform or the standard Seurat normalization.
+Although most methods performed well on the various clustering measures, the global maximum likelihood based on translated Poisson Mixture Model (\texttt{maxLikGlobal}) provided the dimensionality estimate that best separated the subpopulations (Figure \ref{fig:figure8}C) and resulted in the best clustering accuracy (Figure \ref{fig:figure8}D).
+This method systematically selected many more components than were associated with the subpopulations, suggesting that although these additional components appear individually uninformative, in combination they nevertheless contribute to classification.
 
@@
     \caption{\textbf{Estimating dimensionality.
-A:} Estimated dimensionality using the proportion of variance in each component explained by subpopulations.
-\textbf{B:} Difference between `real' dimensionality (from \textbf{A}) and various dimensionality estimation methods, along with computing time.
-\textbf{C:} Average silhouette width per subpopulation across a selection of dimensionality estimators, using sctransform or standard seurat normalization.
-\textbf{D:} Clustering accuracy of across the normalization/dimensionality methods.}
+A:} Estimated dimensionality using the proportion of variance in each component explained by the true subpopulations.
+\textbf{B:} Difference between `real' dimensionality (from \textbf{A}) and dimensionality estimation methods, along with computing time.
+\textbf{C:} Average silhouette width per subpopulation across methods that do not systematically under-estimate the dimensionality, combined with sctransform or standard seurat normalization.
+\textbf{D:} Clustering accuracy across the normalization/dimensionality estimation methods.}
     \label{fig:figure8}
@@
 
-We finally moved to the clustering step.
-Given previous work on the topic\citep{duoClustering2018} and the success of graph-based clustering methods for scRNA-seq, we simply compared Seurat to two \texttt{scran} SNN-based clustering approaches respectively using random walks (`walktrap') or optimizing the modularity score (`fast greedy'), with standard normalization or after \texttt{sctransform}.
+The last analysis evaluated by the \citep{pipeComp} workflow was the clustering step, commonly used to aggregate cells with a common biological (e.g.
+cell-type) or experimental background (e.g., drug treatment).
+Given previous works on the topic \citep{duoClustering2018,freytagComparison2018} and the success of graph-based clustering methods for scRNA-seq, we restricted our evaluation to Seurat's \texttt{FindClusters} method and the two \texttt{scran} SNN-based clustering approaches, based on random walks (\texttt{walktrap} method) or on the optimization of the modularity score (\texttt{fast\_ greedy}).
+Again, the tested methods were combined with Seurat's standard normalization and \texttt{sctransform}.
 
-(Interpretation of the results... walktrap top?)
+The clustering method varied greatly in their ability to retrieve the true subpopulations but the results also varied across the datasets and across the measure used to compare the methods \ref{fig:figure9}.
+Overall, the walktrap method lead to the best identification of the subpopulations.
+The method performed especially good with regard to the MI index, where the clustering was consistently better or similar to other methods in all datasets.
+The precision at the true number of clusters was also near-perfect in most datasets, except in the Zhengmix8eq dataset which showed to be particularly difficult to cluster on, no matter the measure considered.
+In the scenarios where the number of clusters is free to vary (related to the mean F1 and minimum precision scores), the \texttt{scran}'s greedy algorithm results in a good clustering in some datasets.
+The performance of all methods is however systematically poor in the datasets with complex simulated/ real cell-type composition (simMix1 and Zhengmix8eq) in comparison with datasets containing more distinctive or less subpopulations (e.g., Zhengmix4eq).
 
@@
     \includegraphics[width=\textwidth]{{main_figures/clustering_files/figure-html/clustering-1.png}}
-    \caption{\textbf{Evaluation of clustering methods.}}
+    \caption{\textbf{Performance of clustering methods.} Clustering accuracy of common clustering tools in combination with \textttt{sctransform} and standard Seurat's normalization.}
     \label{fig:figure9}
@@
 
-The basic pipeline used here can be extended by adding steps and using the same evaluation metrics.
-To demonstrate this, we evaluated various imputation or denoising techniques on their impact on classification.
-Observing that all methods performed equally well or better on normalized data, we applied them after filtering and normalization, but before scaling and reduction.
-While all methods improved the separability of some subpopulations, no method had a positive impact on the silhouette width of all populations (Figure \ref{fig:figure10}A).
+The basic pipeline presented here can be extended with additional analysis steps to evaluate while keeping the same evaluation metrics.
+To demonstrate this, we evaluated various imputation or denoising techniques based on their impact on classification.
+As preliminary analysis showed that all methods performed equally well or better on normalized data, we applied these methods after filtering and normalization, but before scaling and reduction.
+While all methods improved the separability of some subpopulations compared to a scenario with no imputation step (e.g., simMix1 populations), no method had a systematic positive impact on the silhouette width across all populations (Figure \ref{fig:figure10}A).
 Similarly, not all methods performed as well across all datasets in terms of clustering accuracy (Figure \ref{fig:figure10}A).
@@
 
+On the basis of our findings, we can make a number of concrete analysis recommendations:
+
+\begin{enumerate}
+   \item Filtering
+   \begin{itemize}
+     \item Doublet detection and removal is advised, and can be performed at little computing cost with software such as \texttt{scDblFinder} or \texttt{scds}.
+     \item Distribution-based cell filtering fails to capture doublets, and should use relatively lenient cutoffs (e.g.
+5 MADs, or 3 MADs in at least 2 distributions) to exclude bad cells while avoiding bias.
+   \end{itemize}
+   \item Normalization and scaling
+   \begin{itemize}
+     \item Most normalization methods tested yielded a fair performance, especially when combined with scaling, which tended to have a positive impact.
+     \item \texttt{sctransform} offered the best overall performance in terms of the separability of the subpopulations, as well as removing the effect of library size and detection rate.
+     \item The common practice of regressing out cell covariates such as the detection rate or proportion of mitochondrial reads nearly always had a negative impact, leading to increased correlation with covariates and decreased clustering accuracy.
+Where therefore advise against this practice.
+   \end{itemize}
+   \item Feature selection
+   \begin{itemize}
+     \item Deviance \citep{townesGlmpca2019} offered the best ranking of genes for feature selection.
+     \item Increasing the number of features included tended to lead to better classifications, plateauing after 4000.
+   \end{itemize}
+   \item PCA
+   \begin{itemize}
+     \item As recently reported \citep{SunDimRed2019}, we recommend the irlba-based PCA using weighting of the components, as implemented in \texttt{Seurat}. 
+     \item Instead of the common elbow or jackstraw methods for deciding on the components to include, we recommend the `maxLikGlobal` method.
+   \end{itemize}
+   \item Clustering
+   \begin{itemize}
+     \item TO DO
+   \end{itemize}
+\end{enumerate}
+
 \subsection*{Limitations and open questions}
@@
 \subsection*{Software and package versions}
-Analyses were performed in R 3.6.0 (Bioconductor 3.9), and the following packages were installed from github repositories: \texttt{Seurat} (version 3.0.0), \texttt{sctransform} (0.2.0), \texttt{DoubletFinder} (2.0.1), \texttt{scds} (1.0.0).
+Analyses were performed in R 3.6.0 (Bioconductor 3.9), and the following packages were installed from github repositories: \texttt{Seurat} (version 3.0.0), \texttt{sctransform} (0.2.0), \texttt{DoubletFinder} (2.0.1), \texttt{scds} (1.0.0), \texttt{SAVERX} (1.0.0), \texttt{scImpute} (0.0.9), \texttt{ALRA} (commit 7636de8), \texttt{DCA} (0.2.2), \texttt{DrImpute} (1.2), \texttt{ENHANCE} (R version, commit 1571696), \textttt{SAVERX} (1.0.0).
 The code for the glmPCA was obtained from \url{https://github.com/willtownes/scrna2019} (commit 1ddcc30ebb95d083a685f12fe81d35dd1b0cb1b2).
@@
 \subsection*{Default pipeline parameters}
-(to come)
+
+Most imputations/ denoising methods were run with the default parameters with the following exceptions; \texttt{DrImpute} documentation advises to process the data prior to imputation by removing lowly expressed genes and cells expressing less than 2 genes.
+As it is not clear if this step is a hard requirement for the method to perform well, \texttt{DrImpute} was run with and without prior processing (\texttt{DrImpute\_process} and \texttt{DrImpute\_noprocess} labels, respectively).
+\texttt{dca} method does not accept non-integers counts but two of the datasets being used contain such table of counts due to multimapping.
+For these datasets, we rounded up the counts prior to imputation.
+\texttt{alra} is designed for normalized data but as we are evaluating normalization downstream to imputation, we used the method on both non-normalized (\texttt{alra} label) and normalized counts (\texttt{alra\_norm} label).
+\texttt{scImpute} requires an estimation of the expected number of clusters with the input data.
+As the estimation of the true number of cluster may not be known by the user, we evaluated the tool using the true number of clusters (\texttt{scImpute} label) and using an over/under-estimation of this number (\texttt{scImpute\_plus5} and \texttt{scImpute\_min5} labels, respectively).
+\texttt{ENHANCE} uses a k-nearest neighbor aggregation method and automatically estimates the number of neighbors to merge prior to the imputation.
+With the smallest datasets that we used, this lead to an early stop of the function as this parameters was estimated to 1.
+In such cases, we manually set this parameter to 2 for the method to work.
 
@@
 
-\section*{Figures}
-  \begin{figure}[h!]
-  \caption{\csentence{Sample figure title.}
-      A short description of the figure content
-      should go here.}
-      \end{figure}
-
-\begin{figure}[h!]
-  \caption{\csentence{Sample figure title.}
-      Figure legend text.}
-      \end{figure}
+% \section*{Figures}
+%   \begin{figure}[h!]
+%   \caption{\csentence{Sample figure title.}
+%       A short description of the figure content
+%       should go here.}
+%       \end{figure}
+% 
+% \begin{figure}[h!]
+%   \caption{\csentence{Sample figure title.}
+%       Figure legend text.}
+%       \end{figure}
 
@@
   mixology10x5cl & \cite{tianMixology2018} & 10x & Mixture of 5 cancer cell lines from CellBench \\ 
+  simMix1 & \cite{CrowellMuscat2019} & - & Simulation of 10 human cell subpopulations \\
+  simMix2 & \cite{CrowellMuscat2019} & - & Simulation of 9 mouse cell subpopulations \\
    \hline

### 2020-01-29 bmc_article.tex
Updates from Overleaf

@@
 %\parttitle{Background} %if any
-We present a flexible R framework for pipeline comparison with multi-level evaluation metrics, and apply it to the benchmark of single-cell RNAseq analysis pipelines using datasets with known cell identities.
-We discuss common steps of such analyses, including filtering and doublet detection (offering a new package), normalization, feature selection, denoising, dimensionality reduction and clustering.
-On the basis of these analyses, we make a number of concrete key recommendations about analysis choices.
+The massive growth of single-cell RNA sequencing (scRNAseq) and methods for its analysis is currently only insufficiently accompanied by proper and up-to-date benchmarks that would guide analytical choices.
+Moreover, current studies are often focused on isolated steps of the process.
+Here, we present a flexible R framework for pipeline comparison with multi-level evaluation metrics, and apply it to the benchmark of scRNAseq analysis pipelines using datasets with known cell identities.
+We evaluate common steps of such analyses, including filtering, doublet detection (suggesting a new R package, \textit{scDblFinder}), normalization, feature selection, denoising, dimensionality reduction and clustering.
+On the basis of these analyses, we make a number of key concrete recommendations about analysis choices.
+The evaluation framework, \textit{pipeComp}, has been implemented so as to easily integrate any other step or tool and permit further extending benchmarks (\url{https://github.com/plger/pipeComp}).
+
 \end{abstract}
@@
 This increase in the number of available tools follows the development of new sequencing technologies and the growing number of reported cells, genes and cell populations \cite{SvenssonDB2019}.
-Data processing is a critical step in any scRNAseq analysis as these data are affected by various technical effects (batch effect, dropout, ...) that have to be corrected to distinguish the biological variation form the technical variation.
-The choice of the scRNAseq protocol is already of prime importance as different protocols differ greatly in view of parameters such as gene detection or mappability \cite{MereuCellAtlas2019}, which can in turn affect downstream analysis.
-In view of the intrinsic challenges of scRNAseq technology and the growing number of tools and studies complexity, it is critical to evaluate the available tools to leverage future studies as it can ultimately increase the power of further differential expression analysis \cite{viethSystematic2019}.
+Data processing is a critical step in any scRNAseq analysis, affecting downstream analysis and interpretation.
+It is therefore critical to evaluate the available tools to leverage future studies as it can ultimately increase the power of further differential expression analysis \cite{viethSystematic2019}.
 
@@
 Here, we develop a flexible R framework for pipeline comparison and evaluate in details the various steps of analysis leading from an initial count matrix to a cluster assignment, which are critical in a wide range of applications.
-We harness real datasets of known cell composition (Table \ref{tab:table1}) and develop new evaluation metrics to investigate in a multilevel fashion the impact of various parameters and variations around a core scRNAseq pipeline.
+We harnessed real datasets of known cell composition (Table \ref{tab:table1}) and develop new evaluation metrics to investigate in a multilevel fashion the impact of various parameters and variations around a core scRNAseq pipeline.
 Although we use some datasets based on other protocols, our focus is especially on droplet-based datasets that do not include exogenous control RNA (i.e.
 spike-ins); see Table \ref{tab:table1} and \ref{fig:figure1} for a description of the datasets.
-In addition to previously-used benchmark datasets with true cell labels \cite{duoClustering2018,tianMixology2018}, we simulated two datasets with a hierarchical subpopulation structure based on real 10x human and mouse data using \textit{muscat} (\cite{CrowellMuscat2019})).
-Since graph-based clustering \cite{satijaSeurat2015} was previously shown to consistently perform well across several datasets \cite{duoClustering2018,tianMixology2018}, we used the \texttt{Seurat} pipeline as the starting point to perform an integrated investigation of: 1) doublet identification, 2) cell filtering, 3) normalization, 4) feature selection, 5) dimension reduction, 6) clustering.
+In addition to previously-used benchmark datasets with true cell labels \cite{duoClustering2018,tianMixology2018}, we simulated two datasets with a hierarchical subpopulation structure based on real 10x human and mouse data using \textit{muscat} \cite{CrowellMuscat2019}).
+Since graph-based clustering \cite{satijaSeurat2015} was previously shown to consistently perform well across several datasets \cite{duoClustering2018,tianMixology2018}, we used the \textit{Seurat} pipeline as the starting point to perform an integrated investigation of: 1) doublet identification, 2) cell filtering, 3) normalization, 4) feature selection, 5) dimension reduction, 6) clustering.
 We compare not only competing approaches, but also more fine-grained parameters and variations on common methods.
@@
 
-The \texttt{pipeComp} package defines a pipeline as, minimally, a list of functions executed consecutively on the output of the previous one (\ref{fig:figure2}A).
+The \textit{pipeComp} package defines a pipeline as, minimally, a list of functions executed consecutively on the output of the previous one (\ref{fig:figure2}A).
 In addition, optional benchmark functions can be set for each step to provide standardized, multi-layered evaluation metrics.
-Given such a \texttt{PipelineDefinition} object, a set of alternative parameters (which might include different subroutines) and benchmark datasets, the \texttt{runPipeline} function then proceeds through all combinations of arguments, avoiding recomputing the same step twice and compiling evaluations on the fly to avoid storing potentially large intermediate data.
+Given such a \texttt{PipelineDefinition} object, a set of alternative parameters (which might include different subroutines) and benchmark datasets, the \texttt{runPipeline} function then proceeds through all combinations of arguments, avoiding recomputing the same step twice and compiling evaluations (including running time) on the fly.
+The \textit{CellBench} package was recently proposed to address a similar need\citep{su_cellbench}, offering an elegant piping syntax to combine alternative methods at successive steps.
+In addition to offering the same capabilities, a key additional feature of \textit{pipeComp} is that, since the evaluation metrics are not treated as a step but parallel to it, they can be performed at various points in the pipeline without needing to store the potentially large intermediate data, and thereby enabling combinatorial benchmarks.
 Variations in a given parameter can be evaluated using all metrics from this point downward in the pipeline.
-This is especially important because end-point metrics, such as the adjusted Rand index (ARI) for clustering, are not perfect.
+This is especially important because end-point metrics, such as the adjusted Rand index (ARI, \citep{HubertARI1985}) for clustering, are not perfect.
 For example, by far the most important determinant of the ARI is the number of detected clusters: the farther it is from the actual number of subpopulations, the worse the ARI.
@@
 A number of methods were developed to identify doublets, in most cases by comparing each cell to artificially-created doublets \citep{mcginnisDoubletfinder2019, LunScran2016, BaisScds2019}.
-We therefore first evaluated the capacity of these methods to detect doublets using the two 10x datasets produced by \citep{tianMixology2018} with cells of different genetic identity, and for which therefore SNP information can be used as ground truth.
-We tested \texttt{DoubletFinder} \citep{mcginnisDoubletfinder2019} and \texttt{scran}'s \texttt{doubletCells} \citep{LunScran2016}, both of which use similarity to artificial doublets, and \texttt{scds} \citep{BaisScds2019}, which relies on a combination of co-expression and binary classification.
-\texttt{DoubletFinder} integrates a thresholding based on the proportion of expected doublets, while \texttt{scran} and \texttt{scds} return scores that must be manually thresholded.
+We therefore first evaluated the capacity of these methods to detect doublets using the two 10x datasets with cells of different genetic identity \citep{tianMixology2018}, and for which therefore SNP information can be used as ground truth.
+We included a further such dataset\citep{kangMultiplexedDemuxlet2018} not included for the rest of our benchmark due to the absence of true cell labels beside SNPs.
+Of note, SNP-based analyses also call doublets created by cells of the same cell type (but from different individuals), whose identification is generally not the primary aim of doublet detection methods, since these can be considered homotypic (as opposed to neotypic or heterotypic doublets, i.e.
+doublets from different cell types) and can be identified through other means.
+
+We tested \textit{DoubletFinder} \citep{mcginnisDoubletfinder2019} and \textit{scran}'s \texttt{doubletCells} \citep{LunScran2016}, both of which use similarity to artificial doublets, and \textit{scds} \citep{BaisScds2019}, which relies on a combination of co-expression and binary classification.
+\textit{DoubletFinder} integrates a thresholding based on the proportion of expected doublets, while \textit{scran} and \textit{scds} return scores that must be manually thresholded.
 In these cases, we ensured that the right number of cells would be called doublets.
-
-In addition to these methods, we reasoned that an approach such as \texttt{DoubletFinder} could be simplified by being applied directly on counts and by using a pre-clustering to create neotypic/heterotypic doublets (i.e.
-doublets from different cell types) more efficiently.
-We therefore developed a simple and fast Biocondcutor package implementing this for doublet detection, \texttt{scDblFinder}, with the added advantage of accounting for uncertainty in the expected doublet rate and using meta-cells from the clusters to even include triplets (see methods).
+In addition to these methods, we reasoned that an approach such as \textit{DoubletFinder} could be simplified by being applied directly on counts and by using a pre-clustering to create neotypic/heterotypic doublets more efficiently.
+We therefore developed a simple and fast Bioconductor package implementing this method for doublet detection, \textit{scDblFinder}, with the added advantage of accounting for uncertainty in the expected doublet rate and using meta-cells from the clusters to even include triplets (see methods).
 
 While most methods accurately identified the doublets in the 3 cell lines dataset (mixology10x3cl), the other two datasets proved more difficult (Figure \ref{fig:figure3}A).
-\texttt{scDblFinder} achieved comparable or better accuracy than top alternatives while running in a very reasonable time (Figure \ref{fig:figure3}B).
+\textit{scDblFinder} achieved comparable or better accuracy than top alternatives while running in a very reasonable time (Figure \ref{fig:figure3}B).
 Across datasets, cells called as doublets tended to be classified in the wrong cluster more often than other cells (Figure \ref{fig:figure3}C.
@@
 \textbf{B:} Running time of the different methods (DoubletFinder failed on one of the datasets).
-\textbf{C:} Rate of misclassification of the cells identified by scDblFinder as doublets (DBL) or singlets (SNG), across a large range of clustering analysis.
+\textbf{C:} Rate of misclassification of the cells identified by \textit{scDblFinder} as doublets (DBL) or singlets (SNG), across a large range of clustering analysis.
 Even in datasets which should not have neotypic doublets, the cells identified as such tend to be misclassified.
@@
 This has for instance been demonstrated for droplets containing a high content of mitochondrial reads, often as a result of cell degradation and resulting loss of cytoplasmic mRNAs \citep{IlicicLowQual2016}.
-A common practice is to exclude cells that differ considerably from most other cells on the basis of such properties, for instance through the \texttt{isOutlier} function of \texttt{scater} that measures, for a given control property, the number of median absolute deviations of each cell from the median of all cells.
+A common practice is to exclude cells that differ considerably from most other cells on the basis of such properties, for instance through the \texttt{isOutlier} function of \textit{scater} that measures, for a given control property, the number of median absolute deviations of each cell from the median of all cells.
 Supplementary Figure 1 shows the distributions of some of the typical cell properties commonly used.
-Of note, these properties tend to be correlated, but not always: for instance, while a high proportion of mitochondrial reads is often correlated with a high proportion of the counts in the top features, there can also be other reasons for an over-representation of highly-expressed features (Supplementary Figure 3), such as an over-amplification in non-UMI datasets.
-In addition, in our experience 10X datasets also exhibit a very tight correlation between the total counts and the total features even across very different cell types (Supplementary Figure 4).
+Of note, these properties tend to be correlated, with some exceptions.
+For instance, while a high proportion of mitochondrial reads is often correlated with a high proportion of the counts in the top features, there can also be other reasons for an over-representation of highly-expressed features (Supplementary Figure 3), such as an over-amplification in non-UMI datasets.
+In our experience, 10X datasets also exhibit a very strong correlation between the total counts and the total features even across very different cell types (Supplementary Figure 4).
 We therefore also measure the ratio between the two, and treat cells strongly departing from this trend with suspicion.
@@
 Applying the same filtering criteria on individual clusters of cells (identified through scran's \texttt{quickCluster} method) resulted in nearly no cell being filtered out.
-This suggests that filtering on the global population tends to discard cells of subpopulations with more extreme properties, such as high library size, rather than low-quality cells.
+This suggests that filtering on the global population tends to discard cells of subpopulations with more extreme properties (e.g.
+high library size), rather than low-quality cells.
 Finally, by changing the distributions, the doublet removal step in conjunction with filtering sometimes resulted in a net decrease in the proportion of excluded cells while retaining or improving accuracy.
@@
 Of note, doublet removal appears to be desirable even when, due to the design, there are no heterotypic doublets in the data.
-The PCA methods refer to multivariate outlier detected as implemented in \texttt{scater} (see methods for details).}
+The PCA methods refer to multivariate outlier detected as implemented in \textit{scater} (see methods for details).}
     \label{fig:figure4}
@@
 We next investigated the impact of different normalization strategies.
-Beside the standard log-normalization included in \texttt{Seurat}, we tested \texttt{scran}'s pooling-based normalization \citep{lunPooling2016}, \texttt{sctransform}'s variance-stabilizing transformation \citep{hafemeisterSCtransform2019}, and normalization based on stable genes \citep{linStableGenes2018, deekeStablyExpressed2018}.
-In addition to log-normalization, the standard \texttt{Seurat} clustering pipeline performs per-feature unit-variance scaling so that the PCA is not too strongly dominated by highly-expressed features.
-We therefore included versions of the different normalization procedures, with or without a subsequent scaling (\texttt{sctransform}'s variance-stabilizing transformation involves something analogous to scaling).
-\texttt{Seurat}'s scaling function also includes the option to regress out the effect of certain covariates.
+Beside the standard log-normalization included in \textit{Seurat}, we tested \textit{scran}'s pooling-based normalization \citep{lunPooling2016}, \textit{sctransform}'s variance-stabilizing transformation \citep{hafemeisterSCtransform2019}, and normalization based on stable genes \citep{linStableGenes2018, deekeStablyExpressed2018}.
+In addition to log-normalization, the standard \textit{Seurat} clustering pipeline performs per-feature unit-variance scaling so that the PCA is not too strongly dominated by highly-expressed features.
+We therefore included versions of the different normalization procedures, with or without a subsequent scaling (\textit{sctransform}'s variance-stabilizing transformation involves something analogous to scaling).
+\textit{Seurat}'s scaling function also includes the option to regress out the effect of certain covariates.
 We tested its performance by using the proportion of mitochondrial counts and the number of detected features as covariates.
@@
 
-An important motivation for the development \texttt{sctransform} was the observation that, even after normalization, the first principal components of various datasets tended to correlate with library size, suggesting an inadequate normalization \citep{hafemeisterSCtransform2019}.
+An important motivation for the development \textit{sctransform} was the observation that, even after normalization, the first principal components of various datasets tended to correlate with library size, suggesting an inadequate normalization \citep{hafemeisterSCtransform2019}.
 We therefore first assessed to what extent the first principal component still retained a correlation with the library size and the number of detected features, removing the confounding covariation with the subpopulations (Figure \ref{fig:figure5}A).
-The simple step of scaling tended to remove much of the correlation with these features and, in the absence of scaling, standard \texttt{Seurat} normalization resulted in fairly high correlation with technical covariates.
-\texttt{sctransform} led to the lowest correlation but most methods, including normalization based on stable genes, were able to remove most of the correlation when combined with scaling.
-The only exception is \texttt{Seurat} normalization with scaling regressing out covariates, which surprisingly increased the correlation in 8 of the 9 datasets.
+The simple step of scaling tended to remove much of the correlation with these features and, in the absence of scaling, standard \textit{Seurat} normalization resulted in fairly high correlation with technical covariates.
+\textit{sctransform} led to the lowest correlation but most methods, including normalization based on stable genes, were able to remove most of the correlation when combined with scaling.
+The only exception is \textit{Seurat} normalization with scaling regressing out covariates, which surprisingly increased the correlation in 8 of the 9 datasets.
 
 We next investigated the impact of normalization on the separability of the subpopulations (Figure \ref{fig:figure5}B-C).
-Since clustering accuracy metrics such as the ARI are very strongly influenced by the number of clusters, we complemented it with silhouette width and mutual information.
+Since clustering accuracy metrics such as the ARI are very strongly influenced by the number of clusters, we complemented it with silhouette width\citep{RousseeuwSil1987} and mutual information.
 We found most methods (including no normalization at all) to perform fairly well in most of the subpopulations.
@@
 Regressing out covariates systematically gave poorer performance on all metrics.
-\texttt{sctransform} systematically outperformed other methods, and even though it was developed to be applied to data with unique molecular identifiers (UMI), it also performed fairly well with Smart-seq protocol (Koh and Kumar datasets).
+\textit{sctransform} systematically outperformed other methods, and even though it was developed to be applied to data with unique molecular identifiers (UMI), it also performed fairly well with Smart-seq protocol (Koh and Kumar datasets).
 
@@
     \includegraphics[width=\textwidth]{{main_figures/normalization_files/figure-html/figure3-1.png}}
-    \caption{\textbf{Evaluation of normalization procedures.
-`Seurat.feat_mt_regress' indicates the regressing out of the number of features and proportion of mitochondrial reads during scaling, represents, and `Seurat.feat_regress` of the number of features only.
-A:} Residual correlation of the first principal component with library size (left) and detection rate (right), after accounting for biological differences between subpopulations.
+    \caption{\textbf{Evaluation of normalization procedures.} `Seurat.feat\_mt\_regress' indicates the regressing out of the number of features and proportion of mitochondrial reads during scaling, represents, and `Seurat.feat\_regress` of the number of features only.
+Residual correlation of the first principal component with library size (left) and detection rate (right), after accounting for biological differences between subpopulations.
 \textbf{B:} Average silhouette width per true subpopulation, where higher silhouette width means a higher separability.
@@
 Although some methods had a tendency to lead to a higher (e.g.
-\texttt{sctransform}) or lower (e.g.
+\textit{sctransform}) or lower (e.g.
 stable genes) number of clusters, the effect was very mild and not entirely systematic (Supplementary Figure 10).
@@
 A standard clustering pipeline typically involves a step of highly-variable genes selection, which is complicated by the digital nature and the mean-variance relationship of (sc)RNAseq.
-\texttt{Seurat}'s earlier approaches involved the use of dispersion estimates standardized for the mean expression levels, while more recent versions (>3) rely on a different measure of variance, again standardized.
+\textit{Seurat}'s earlier approaches involved the use of dispersion estimates standardized for the mean expression levels, while more recent versions (>3) rely on a different measure of variance, again standardized.
 Of note, while adjusting for the mean-variance relationship removes much of the bias towards highly-expressed genes, it is plausible that this relationship may in fact sometimes reflects biological relevance and would be helpful in classifying cell types.
-Another common practice in features selection is to use those with the highest mean expression, and recently, \citep{townesGlmpca2019} instead suggested to use deviance, while \texttt{sctransform} provides its own ordering of genes based on transformed variance.
+Another common practice in features selection is to use those with the highest mean expression, and recently, \citep{townesGlmpca2019} instead suggested to use deviance, while \textit{sctransform} provides its own ordering of genes based on transformed variance.
 
-Reasoning that a selection method should ideally select genes whose variability is more between subpopulations than within, we first assessed whether a method selected genes with a high proportion of variance or deviance explained by (real) subpopulation.
+Reasoning that a selection method should ideally select genes whose variability is more between subpopulations than within, we first assessed to what extent each method selected genes with a high proportion of variance or deviance explained by (real) subpopulation.
 As the proportion of variability in a gene attributable to subpopulations can be measured in various ways, we first compared three such alternatives.
-Analysis of variance performed on a standard \texttt{Seurat} normalization or on \texttt{sctransform} data were highly correlated (Supplementary Figure 11A).
+Analysis of variance performed on a standard \textit{Seurat} normalization or on \textit{sctransform} data were highly correlated (Supplementary Figure 11A).
 These estimates were also in good agreement with the deviance explained, although lowly-expressed genes could have a high deviance explained without having much of their variance explained by subpopulation (Supplementary Figure 11B-D).
 We therefore compared the proportion of the cumulative variance/deviance explained by the top X genes that could be retrieved by each gene ranking method (Supplementary Figures 12-13).
-We first focused on the first 1000 genes to highlight the differences between the methods, although a higher number of selected genes decreased the differences between the methods (Supplementary Figures 12-14).
+We first focused on the first 1000 genes to highlight the differences between methods, although a higher number of selected genes decreased the differences between methods (Supplementary Figures 12-14).
 The standardized measures of variability were systematically worse than their non-standardized counterparts in selecting genes with a high proportion of variance explained by subpopulation.
 Regarding to the percentage of deviance explained however, the standardized measures were often superior(Figure \ref{fig:figure6}A and Supplementary Figures 12-13).
-Deviance proved the method of choice to prioritize genes with a high variance explained by subpopulations (with mere expression level proving surprisingly good), but surprisingly did not perform so well to select genes with a high deviance explained.
+Deviance proved the method of choice to prioritize genes with a high variance explained by subpopulations (with mere expression level proving surprisingly good), but did not perform so well to select genes with a high deviance explained.
 
@@
 \textbf{B:} Accuracy of clusterings (at the true number of clusters) when selecting 1000 genes using the given methods.
-Based on standard \texttt{Seurat} normalization (left) or \texttt{sctransform} (right).
+Based on standard \textit{Seurat} normalization (left) or \textit{sctransform} (right).
 The methods `vst.varExp' and `devianceExplained' are the estimates used in \textbf{A} to evaluate the selection methods, and were included here only for validation purpose.}
@@
 
-We next evaluated how the use of the different feature selection methods affected the clustering accuracy (Figure \ref{fig:figure6}B).
+We next evaluated how the use of different feature selection methods affected the clustering accuracy (Figure \ref{fig:figure6}B).
 To validate the previous assay on the proportion of variance/ deviance explained by real populations, we included genes that maximized these two latter measures.
@@
 
-We investigated the methods of dimensionality reduction.
-Since the various PCA approaches and implementations were recently benchmarked in a similar context \citep{TsuyuzakiPCA2020}, we focused on widely used approaches which had not yet been compared \texttt{Seurat}'s PCA, \texttt{scran}'s \texttt{denoise PCA}, and GLM-PCA \citep{townesGlmpca2019} and combined (when relevant) with \texttt{sctransform} normalization.
-Given that \texttt{Seurat}'s default PCA weights the cell embeddings by the variance of each component, we also evaluated the impact of this weighting with each method.
+We next investigated the methods of dimensionality reduction.
+Since the various PCA approaches and implementations were recently benchmarked in a similar context \citep{TsuyuzakiPCA2020}, we focused on widely used approaches which had not yet been compared: \textit{Seurat}'s PCA, \textit{scran}'s \texttt{denoise PCA}, and GLM-PCA \citep{townesGlmpca2019}, and combined them (when relevant) with \textit{sctransform} normalization.
+Given that \textit{Seurat}'s default PCA weights the cell embeddings by the variance of each component, we also evaluated the impact of this weighting with each method.
 
@@
 The first components identified by GLM-PCA tended to have a higher proportion of their variance explained by real subpopulations and increase the average silhouette width of already well-defined subpopulations.
-Seurat's PCA procedure however proved superior on all metrics (Figure \ref{fig:figure7} and Supplementary Figures 15-17).
+Seurat's PCA procedure however proved superior on all metrics (Figure \ref{fig:figure7}).
 Overall, weighting the principal components by their variance in the same fashion as Seurat's method does had a strong positive impact on silhouette widths and ARI scores.
@@
 Since euclidean distance decreases as the number of non-discriminating dimensions increases, there is usually a trade-off between selecting enough dimensions to keep most information and excluding smaller dimensions that may represent technical noise or other unwanted sources of variation.
-Overall, increasing the number of dimensions robustly led to a decrease in the number of clusters, although the effect was milder with data processed through \texttt{sctransform} (Supplementary Figures 14 and 22).
-This tended to be translated to the accuracy of the clustering (Supplementary Figure 23), although in both cases (number of clusters and ARI) the resolution parameter had a much stronger impact.
+Overall, increasing the number of dimensions robustly led to a decrease in the number of clusters (Supplementary Figures 15).
+This tended to be translated to the accuracy of the clustering (Supplementary Figure 16), although in both cases (number of clusters and ARI) \textit{Seurat}'s resolution parameter had a much stronger impact.
 
 Different approaches have been proposed to select the appropriate number of dimensions, from the visual identification of an 'Elbow' (inflexion point) of the variance explained, to more complex algorithms.
-We evaluated the performance of common tools, namely [TBD] (\texttt{pcaLocal.FO} label, [REF needed?]), global maximum likelihood based on translated Poisson Mixture Model (\texttt{maxLikGlobal}, \citep{haroTranslated2008}), [TBD] (\texttt{essLocal.b}, \texttt{essLocal.b},[REF needed?]), JackStraw procedure (\texttt{jackstraw.elbow}, \citep{ChungJackstraw2015}), [TBD](\texttt{pcaLocal.maxgap}, [REF needed?]), [TBD] (\texttt{pcaLocal.fan} [REF needed?]), Fisher Separability analysis (\texttt{fisherSeparability}, \citep{AlberganteSepar2019}), \texttt{scran}'s \texttt{denoisePCA} function (\texttt{scran.denoisePCA}, \citep{LunScran2016}) and [TBD] (\texttt{pcaOptmPointwise.max}, [REF needed?]).
+We evaluated the performance of common tools, namely [TBD] (\textit{pcaLocal.FO} label, [REF needed?]), global maximum likelihood based on translated Poisson Mixture Model (\textit{maxLikGlobal}, \citep{haroTranslated2008}), [TBD] (\textit{essLocal.b}, \textit{essLocal.b},[REF needed?]), JackStraw procedure (\textit{jackstraw.elbow}, \citep{ChungJackstraw2015}), [TBD](\textit{pcaLocal.maxgap}, [REF needed?]), [TBD] (\textit{pcaLocal.fan} [REF needed?]), Fisher Separability analysis (\textit{fisherSeparability}, \citep{AlberganteSepar2019}), \textit{scran}'s \texttt{denoisePCA} function (\textit{scran.denoisePCA}, \citep{LunScran2016}) and [TBD] (\textit{pcaOptmPointwise.max}, [REF needed?]).
 
@@
 Reasoning that over-estimating the number of dimensions is less problematic than under-estimating it, we kept the former methods for a full analysis of their impact on clustering (Figure \ref{fig:figure8}C-D), when combined with sctransform or the standard Seurat normalization.
-Although most methods performed well on the various clustering measures, the global maximum likelihood based on translated Poisson Mixture Model (\texttt{maxLikGlobal}) provided the dimensionality estimate that best separated the subpopulations (Figure \ref{fig:figure8}C) and resulted in the best clustering accuracy (Figure \ref{fig:figure8}D).
+Although most methods performed well on the various clustering measures, the global maximum likelihood based on translated Poisson Mixture Model (\textit{maxLikGlobal}) provided the dimensionality estimate that best separated the subpopulations (Figure \ref{fig:figure8}C) and resulted in the best clustering accuracy (Figure \ref{fig:figure8}D).
 This method systematically selected many more components than were associated with the subpopulations, suggesting that although these additional components appear individually uninformative, in combination they nevertheless contribute to classification.
@@
 \textbf{B:} Difference between `real' dimensionality (from \textbf{A}) and dimensionality estimation methods, along with computing time.
-\textbf{C:} Average silhouette width per subpopulation across methods that do not systematically under-estimate the dimensionality, combined with sctransform or standard seurat normalization.
+\textbf{C:} Average silhouette width per subpopulation across a selection of methods, combined with \textit{sctransform} or standard \textit{Seurat} normalization.
 \textbf{D:} Clustering accuracy across the normalization/dimensionality estimation methods.}
@@
 
-The last analysis evaluated by the \citep{pipeComp} workflow was the clustering step, commonly used to aggregate cells with a common biological (e.g.
-cell-type) or experimental background (e.g., drug treatment).
-Given previous works on the topic \citep{duoClustering2018,freytagComparison2018} and the success of graph-based clustering methods for scRNA-seq, we restricted our evaluation to Seurat's \texttt{FindClusters} method and the two \texttt{scran} SNN-based clustering approaches, based on random walks (\texttt{walktrap} method) or on the optimization of the modularity score (\texttt{fast\_ greedy}).
-Again, the tested methods were combined with Seurat's standard normalization and \texttt{sctransform}.
+The last step evaluated in our pipeline was clustering.
+Given previous works on the topic \citep{duoClustering2018,freytagComparison2018} and the success of graph-based clustering methods for scRNA-seq, we restricted our evaluation to Seurat's method and two \textit{scran} SNN-based clustering approaches, based on random walks (\textit{walktrap} method) or on the optimization of the modularity score (\textit{fast\_greedy}).
+Again, the tested methods were combined with Seurat's standard normalization and \textit{sctransform}, otherwise using the parameters found optimal in the previous steps.
 
-The clustering method varied greatly in their ability to retrieve the true subpopulations but the results also varied across the datasets and across the measure used to compare the methods \ref{fig:figure9}.
-Overall, the walktrap method lead to the best identification of the subpopulations.
-The method performed especially good with regard to the MI index, where the clustering was consistently better or similar to other methods in all datasets.
-The precision at the true number of clusters was also near-perfect in most datasets, except in the Zhengmix8eq dataset which showed to be particularly difficult to cluster on, no matter the measure considered.
-In the scenarios where the number of clusters is free to vary (related to the mean F1 and minimum precision scores), the \texttt{scran}'s greedy algorithm results in a good clustering in some datasets.
-The performance of all methods is however systematically poor in the datasets with complex simulated/ real cell-type composition (simMix1 and Zhengmix8eq) in comparison with datasets containing more distinctive or less subpopulations (e.g., Zhengmix4eq).
+Since ARI is dominated by differences in the number of clusters, and no single metric is perfect, we diversified them (Figure \ref{fig:figure9}).
+Mutual information (MI) has the virtue of not decreasing when a true subpopulation is split into two clusters, which is arguably less problematic (and might well reflect unknown biological subgroups), but as a consequence it can be biased towards methods producing higher resolution clustering.
+Similarly, precision per true subpopulation is considerably more robust to differences in the number of clusters.
+We also tracked the mean F1 score and the ARI at the true number of clusters.
+
+The MI score and minimum precision, which are largely independent of the estimated number of clusters, were overall higher for the walktrap method (Figure \ref{fig:figure9}), while the mean F1 score favored both \textitt{scran} methods (walktrap and fast greedy) over \textit{Seurat}.
+Finally, the ARI score at the true number of clusters, when available, showed similar performances.
+However, while \textit{Seurat}'s resolution parameter had a large impact on the number of clusters identified, and therefore could be used to find a parameter combination producing the right number of clusters, the number of clusters found by \textit{scran}-based clustering was considerably less influenced by the available parameters (number of nearest neighbors or steps in the random walk - see Supplementary Figure 17).
+The occasional lack of scran-based clustering with the right number of clusters, along with a higher mutual information, suggest that \textit{scran} sometimes simply divides a real subpopulation into two (possibly tracking some unknown biological differences) rather than committing misclassification errors.
+Overall, the walktrap method appeared superior to the fast greedy algorithm and was generally less prone to misclassification than \textit{Seurat} clustering, although the latter offered more control over the resolution.
+Of note, very difficult subpopulations, both from the Zhengmix8eq and simMix1 datasets, remained very inaccurately classified by all methods and in regards to all metrics.
 
@@
     \includegraphics[width=\textwidth]{{main_figures/clustering_files/figure-html/clustering-1.png}}
-    \caption{\textbf{Performance of clustering methods.} Clustering accuracy of common clustering tools in combination with \textttt{sctransform} and standard Seurat's normalization.}
+    \caption{\textbf{Evaluation of clustering methods.} Clustering accuracy of common clustering tools in combination with \textitt{sctransform} and standard Seurat's normalization.}
     \label{fig:figure9}
@@
 
-The basic pipeline presented here can be extended with additional analysis steps to evaluate while keeping the same evaluation metrics.
+The basic pipeline presented here can be extended with additional analysis steps while keeping the same evaluation metrics.
 To demonstrate this, we evaluated various imputation or denoising techniques based on their impact on classification.
-As preliminary analysis showed that all methods performed equally well or better on normalized data, we applied these methods after filtering and normalization, but before scaling and reduction.
-While all methods improved the separability of some subpopulations compared to a scenario with no imputation step (e.g., simMix1 populations), no method had a systematic positive impact on the silhouette width across all populations (Figure \ref{fig:figure10}A).
-Similarly, not all methods performed as well across all datasets in terms of clustering accuracy (Figure \ref{fig:figure10}A).
-As expected, 10X datasets which have more cells at a lower coverage benefited more from these methods, which were instead rather deleterious in smart-seq datasets.
+As preliminary analysis showed that all methods performed equally well or better on normalized data, we applied them after filtering and normalization, but before scaling and reduction.
+Although some of the methods (e.g.
+\textit{DRImpute\_process} and \textit{alra\_norm}) did improve the separability of some more elusive subpopulations, no method had a systematically positive impact on the average silhouette width across all subpopulations (Figure \ref{fig:figure10}A).
+When restricting ourselves to clustering analyses that yielded the `right' number of clusters, all tested methods improved classification  compared to a scenario with no imputation step (`none' label, Figure \ref{fig:figure10}B).
+However, the situation was not so straightforward with alternative metrics, where some methods (e.g.
+\textit{enhance}) consistently underperformed.
+10X datasets, which are typically characterized by a lower per-cell coverage and feature detection rate, expectedly benefited more from imputation, and in this context \textit{DRImpute} and \textit{dca} tended to show the best performance.
+On the contrary, imputing on normalized counts originating from Smart-seq technology was instead rather deleterious to the clustering accuracy.
+Our recommendation follows the previous benchmarking of imputation strategies performed by \citep{ZhangImput2018}.
+Although this previous study focused on trajectory inference and preservation of data structure as main metrics, \textit{DrImpute} also revealed to be among the top performers and was recommended for clustering analysis (\textit{alra} was however not included in this previous evaluation).
 
@@
     \includegraphics[width=\textwidth]{{main_figures/imputation_files/figure-html/imputation-1.png}}
-    \caption{\textbf{Imputation/denoising methods.
-A:} Average silhouette width per subpopulation across various denoising/imputation methods.
-\textbf{B:} Clustering accuracy across denoising/imputation methods.}
+    \caption{\textbf{Imputation/denoising methods} Average silhouette width per subpopulation (\textbf{A}) and clustering accuracy (\textbf{B}) with or without (\textit{none}) application of a denoising/imputation method}
     \label{fig:figure10}
@@
 \section*{Discussion}
-
-For research articles this section should discuss the implications of the findings in context of existing research and highlight limitations of the study.
-For study protocols and methodology manuscripts this section should include a discussion of any practical or operational issues involved in performing the study and any issues not covered in other sections.
 
@@
    \begin{itemize}
-     \item Doublet detection and removal is advised, and can be performed at little computing cost with software such as \texttt{scDblFinder} or \texttt{scds}.
+     \item Doublet detection and removal is advised, and can be performed at little computing cost with software such as \textit{scDblFinder} or \textit{scds}.
      \item Distribution-based cell filtering fails to capture doublets, and should use relatively lenient cutoffs (e.g.
-5 MADs, or 3 MADs in at least 2 distributions) to exclude bad cells while avoiding bias.
+5 MADs, or 3 MADs in at least 2 distributions) to exclude poor-quality cells while avoiding bias.
+     \item Features filtering based on feature type did not appear beneficial.
    \end{itemize}
@@
    \begin{itemize}
-     \item Most normalization methods tested yielded a fair performance, especially when combined with scaling, which tended to have a positive impact.
-     \item \texttt{sctransform} offered the best overall performance in terms of the separability of the subpopulations, as well as removing the effect of library size and detection rate.
+     \item Most normalization methods tested yielded a fair performance, especially when combined with scaling, which tended to have a positive impact on clustering.
+     \item \textit{sctransform} offered the best overall performance in terms of the separability of the subpopulations, as well as removing the effect of library size and detection rate.
      \item The common practice of regressing out cell covariates such as the detection rate or proportion of mitochondrial reads nearly always had a negative impact, leading to increased correlation with covariates and decreased clustering accuracy.
-Where therefore advise against this practice.
+We therefore advise against this practice.
    \end{itemize}
@@
      \item Deviance \citep{townesGlmpca2019} offered the best ranking of genes for feature selection.
-     \item Increasing the number of features included tended to lead to better classifications, plateauing after 4000.
+     \item Increasing the number of features included tended to lead to better classifications, plateauing from 4000 features.
+   \end{itemize}
+   \item Denoising/imputation
+   \begin{itemize}
+       \item Denoising appeared beneficial to the identification of subpopulations in 10x datasets, but not in Smart-seq datasets.
+       \item We found especially \textit{alra} (with prior normalization), \textit{DrImpute} (with prior processing) and \textit{dca} to offer the best performances.
    \end{itemize}
@@
    \begin{itemize}
-     \item As recently reported \citep{SunDimRed2019}, we recommend the irlba-based PCA using weighting of the components, as implemented in \texttt{Seurat}. 
-     \item Instead of the common elbow or jackstraw methods for deciding on the components to include, we recommend the `maxLikGlobal` method.
+     \item Similarly to previous reports \citep{SunDimRed2019}, we recommend the irlba-based PCA using weighting of the components, as implemented in \textit{Seurat}. 
+     \item Instead of the common elbow or jackstraw methods for deciding on the components to include, we recommend the \textit{global maximum likelihood based on translated Poisson Mixture Model} method (e.g.
+implemented in \texttt{intrinsicDimension::maxLikGlobalDimEst()}).
    \end{itemize}
@@
    \begin{itemize}
-     \item TO DO
+    \item We found \textit{scran}-based walktrap clustering to show good performances.
+    \item In cases where prior knowledge can guide the choice of a resolution, \textit{Seurat} can be useful in affording manual control of it while, in the absence of such knowledge, \textit{scran}-based walktrap clustering provided reasonable estimates.
    \end{itemize}
@@
 
+In this study, we evaluated tools commonly used for the processing of scRNAseq with a focus on droplet-based datasets, namely from the 10x technology.
+Although this platform has been used in almost half of the scRNAseq studies in 2019 \citep{SvenssonDB2019}, other popular technologies such as Drop-seq, InDrops or Smart-seq2/3 were not represented in the present benchmarking.
+Differences between such protocols have a very large impact on clustering \citep{MereuCellAtlas2019}.
+Although most top-ranking methods in our comparison performed well on both Smart-seq and 10x datasets that we tested, future benchmarking efforts should strive to include less represented technologies.
+In addition, we did not compare any of the alignment and/or quantification methods used to obtain the count matrix \citep{MereuCellAtlas2019}.
+Some steps, such as the implementation of the PCA step, were also not explored in detail here as they have already been the object of recent and thorough study elsewhere \citep{SunDimRed2019}.
+We also considered only methods relying on Euclidean distance, while correlation was recently reported to be superior\citep{kim_impact_2019} and would require further investigation.
+
+Here, we chose to concentrate on what could be learned from datasets with known cell labels (as opposed to labels inferred from the data, as in \citep{MereuCellAtlas2019}).
+In contrast to \citep{tianMixology2018}, who relied on RNA mixtures of known proportions, we chose to rely chiefly on real cells and their representative form of variability.
+Given the limited availability of such datasets however, several aspects of single-cell analysis could not be compared, such as batch effect correction or multi-dataset integration.
+In addition, a focus on the identification of the subpopulations might fail to reveal methods which are instead best performing for tasks other than clustering, such as differential expression analysis or trajectory inference.
+Some informative benchmarks have already been performed on some of these topics \citealp{SaelensTraject2019, SonesonDE2018}.
+Yes, such evaluations could benefit from considering methods not in isolation, but as parts of a connected workflow, as we have done here.
+We believe that the \textit{pipeComp} framework is modular and flexible enough to integrate new steps in the pipeline, as it was shown as an example with the evaluation of imputation/ denoising methods.
+
+We believe there is much space for improvement in methods.
+Concerning cell filtering, we noticed that the current approach based on whole-population characteristic (e.g.
+MAD cut-off) can be biased against certain subpopulations, suggesting that more refined methods expect multimodal distributions could be used rather than relying on whole-population characteristics.
+In addition, most common filtering approaches do not harness the relationship between cell QC properties.
+Finally, imputation had a varied impact on the clustering analysis and seemed to be linked to the technology that was used to generate the data.
+We support the hypothesis of \citep{ZhangImput2018} on the respective strengths of linear models and non-linear models and their use to different types of data, such as the performance of non-linear methods in developmental studies.
+
 \section*{Conclusions}
 
-This should state clearly the main conclusions and provide an explanation of the importance and relevance of the study to the field.
+\textit{pipeComp} is a flexible R framework for evaluating methods and parameter combinations in a complex pipeline, computing on-the-fly multilevel evaluation metrics.
+Application of this framework to scRNAseq clustering enabled us to make concrete recommendations on the steps of filtering, normalization, feature selection, denoising, dimensionality reduction and clustering.
+We show a diversity of multilevel metrics to be more robust, more sensitive, and more nuanced than simply evaluating the final clustering.
+In addition, we provide a new, efficient Bioconductor package for doublet detection.
+Finally, we hope that the \textit{pipeComp} framework can be applied to extend the current benchmark, as well as to apply it to other contexts.
 
@@
 
-All analyses were performed through the \href{pipeComp}{https://github.com/plger/pipeComp} R package, which implements the pipeline framework described here.
+All analyses were performed through the \href{https://github.com/plger/pipeComp}{pipeComp} R package, which implements the pipeline framework described here.
 All code to reproduce the simulations and figures is available in the \url{https://github.com/markrobinsonuzh/scRNA\_pipelines\_paper} repository, which also includes the basic datasets with a standardized annotation.
@@
 \subsection*{Software and package versions}
-Analyses were performed in R 3.6.0 (Bioconductor 3.9), and the following packages were installed from github repositories: \texttt{Seurat} (version 3.0.0), \texttt{sctransform} (0.2.0), \texttt{DoubletFinder} (2.0.1), \texttt{scds} (1.0.0), \texttt{SAVERX} (1.0.0), \texttt{scImpute} (0.0.9), \texttt{ALRA} (commit 7636de8), \texttt{DCA} (0.2.2), \texttt{DrImpute} (1.2), \texttt{ENHANCE} (R version, commit 1571696), \textttt{SAVERX} (1.0.0).
+Analyses were performed in R 3.6.0 (Bioconductor 3.9), and the following packages were installed from github repositories: \textit{Seurat} (version 3.0.0), \textit{sctransform} (0.2.0), \textit{DoubletFinder} (2.0.1), \textit{scDblFinder} (1.1.1), \textit{scds} (1.0.0), \textit{SAVERX} (1.0.0), \textit{scImpute} (0.0.9), \textit{ALRA} (commit 7636de8), \textit{DCA} (0.2.2), \textit{DrImpute} (1.2), \textit{ENHANCE} (R version, commit 1571696), \textitt{SAVERX} (1.0.0).
 The code for the glmPCA was obtained from \url{https://github.com/willtownes/scrna2019} (commit 1ddcc30ebb95d083a685f12fe81d35dd1b0cb1b2).
@@
 \subsection*{Simulated datasets}
-The `simMix1' dataset was based on the human PBMC CITE-seq data deposited under `GSE100866'.
-Both RNA and ADT count data were downloaded from GEO and processed independently using Seurat.
-We then considered cells that were in the same cluster both in the RNA-based and ADT-based analyses to be real subpopulations, and focused on the 4 most abundant such subpopulations.
-We then performed 3 sampling-based simulations with various degrees of separation using \textit{muscat}, and merged the three simulations.
-The `simMix2' dataset was generated from the mouse brain data published with \textit{muscat} [REF] in a similar fashion.
+The simMix1 dataset was based on the human PBMC CITE-seq data deposited under accession code GSE100866 of the Gene Expression Omnibus (GEO) website.
+Both RNA and ADT count data were downloaded from GEO and processed independently using \textit{Seurat}.
+We then considered cells that were in the same cluster both in the RNA-based and ADT-based analyses to be real subpopulations, and focused on the 4 most abundant ones.
+We then performed 3 sampling-based simulations with various degrees of separation using \textit{muscat} \citep{CrowellMuscat2019}, and merged the three simulations.
+The simMix2 dataset was generated from the mouse brain data published in \cite{CrowellMuscat2019} in a similar fashion.
 The specific code is available on \url{https://github.com/markrobinsonuzh/scRNA\_pipelines\_paper}.
@@
 
-Most imputations/ denoising methods were run with the default parameters with the following exceptions; \texttt{DrImpute} documentation advises to process the data prior to imputation by removing lowly expressed genes and cells expressing less than 2 genes.
-As it is not clear if this step is a hard requirement for the method to perform well, \texttt{DrImpute} was run with and without prior processing (\texttt{DrImpute\_process} and \texttt{DrImpute\_noprocess} labels, respectively).
-\texttt{dca} method does not accept non-integers counts but two of the datasets being used contain such table of counts due to multimapping.
+Where unspecified, the following default parameters or parameter sets were used:
+\begin{itemize}
+    \item \textit{scDblFinder} was used for doublet identification
+    \item the default filter sets (see below) were applied
+    \item standard \textit{Seurat} normalization was employed
+    \item \textit{Seurat} (v3) variable feature selection was employed, selecting 2000 genes
+    \item standard \textit{Seurat} scaling and PCA were employed
+    \item To study the impact of upstream steps on clustering, various \textit{Seurat} clustering analyses were performed, selecting different numbers of dimensions (5, 10, 15, 20, 30, and 50) and using various resolution parameters (0.005, 0.01, 0.02, 0.05, 0.1, 0.15, 0.2, 0.3, 0.4, 0.5, 0.8, 1, 1.2, 1.5, 2, 4).
+The range of resolution parameters was selected to ensure that the right number of clusters could be obtained in all datasets.
+\end{itemize}
+
+\subsubsection{Denoising/imputation}
+Most imputations/ denoising methods were run with the default parameters with the following exceptions; \textit{DrImpute} documentation advises to process the data prior to imputation by removing lowly expressed genes and cells expressing less than 2 genes.
+As it is not clear if this step is a hard requirement for the method to perform well, \textit{DrImpute} was run with and without prior processing (\textit{DrImpute\_process} and \textit{DrImpute\_noprocess} labels, respectively).
+\textit{dca} method does not accept non-integers counts but two of the datasets being used contain such table of counts due to multimapping.
 For these datasets, we rounded up the counts prior to imputation.
-\texttt{alra} is designed for normalized data but as we are evaluating normalization downstream to imputation, we used the method on both non-normalized (\texttt{alra} label) and normalized counts (\texttt{alra\_norm} label).
-\texttt{scImpute} requires an estimation of the expected number of clusters with the input data.
-As the estimation of the true number of cluster may not be known by the user, we evaluated the tool using the true number of clusters (\texttt{scImpute} label) and using an over/under-estimation of this number (\texttt{scImpute\_plus5} and \texttt{scImpute\_min5} labels, respectively).
-\texttt{ENHANCE} uses a k-nearest neighbor aggregation method and automatically estimates the number of neighbors to merge prior to the imputation.
+\textit{alra} is designed for normalized data but as we are evaluating normalization downstream to imputation, we used the method on both non-normalized (\textit{alra} label) and normalized counts (\textit{alra\_norm} label).
+\textit{scImpute} requires an estimation of the expected number of clusters with the input data.
+As the estimation of the true number of cluster may not be known by the user, we evaluated the tool using the true number of clusters (\textit{scImpute} label) and using an over/under-estimation of this number (\textit{scImpute\_plus5} and \textit{scImpute\_min5} labels, respectively).
+\textit{ENHANCE} uses a k-nearest neighbor aggregation method and automatically estimates the number of neighbors to merge prior to the imputation.
 With the smallest datasets that we used, this lead to an early stop of the function as this parameters was estimated to 1.
@@
 Our doublet detection method is available at \url{https://github.com/plger/scDblFinder}.
-Version 0.1 was used here.
-Briefly, we first cluster the cells using \texttt{scran}'s \texttt{quickCluster} method, favouring overclustering (default minimum size of 20), and reduce the dataset to the union of 1500 most expressed genes of each cluster.
-We then create 75\% of the artificial doublets by sampling the two cells specifically from different clusters, and 25\% by entirely random sampling (in case the clustering was erroneous), and summing the counts of each pair of cells.
+Briefly, after reducing the data to the most expressed genes, we cluster the cells using the fast-greedy algorithm, favouring overclustering.
+We then create artificial doublets by sampling the two cells specifically from different clusters, and sum the counts of each pair of cells.
 We also create meta-cells from each cluster, and use them to create additional doublets and triplets.
-We then combine them with the real cells, convert the counts to scaled column ranks, and build a graph of the \textit{k} (default 10) nearest neighbors using scran.
-We then calculate, for each cell, the proportion of its neighbors that are artificial doublets.
-This ratio serves as a doublet score, which is then thresholded by simultaneously minimizing the error in classifying real vs artificial doublets and the deviation from the distribution of expected doublet rate (accounting for homotypic doublets as done by DoubletFinder).
-For the pipeline comparison here, the expected doublet rate was set to 2.1\% (with a 1\% standard deviation) on the basis of the mixology datasets.
+We combine them with the real cells, perform PCA and build a KNN graph using \textit{BiocNeighbors}.
+We then calculate, for each cell, the proportion of its neighbors that are artificial doublets, weighted by the distance.
+This ratio serves as a doublet score, which is then thresholded by simultaneously minimizing the error in classifying real vs artificial doublets and the deviation from the distribution of expected doublet rate (accounting for homotypic doublets as done by \textit{DoubletFinder}).
 
@@
 
+The `pca' and `pca2' clusters refer to the multivariate outlier detection methods implemented in \textit{scater}, running \textit{runPCA} with \texttt{use\_coldata=TRUE, detect\_outliers=TRUE}.
+`pca` uses all covariates, while `pca2` uses only the log10(counts), log10(features), proportion mitochondrial and proportion in the top 50 features.
+
 \subsection*{Variance and deviance explained}
 
-Unless specified otherwise, the variance in gene expression explained by subpopulations was calculated on the data normalized and transformed through sctransform.
+Unless specified otherwise, the variance in gene expression explained by subpopulations was calculated on the data normalized and transformed through \textit{sctransform}.
 For each gene, we fitted a linear model using the subpopulation as only independent variable (\textasciitilde subpopulation), and used the R-squared as a measure of the variance explained.
@@
 The deviance explained by subpopulations was calculated directly on counts using the \texttt{pipeComp::getDevianceExplained} function.
-The function uses edgeR to fit two models, a full \textasciitilde librarySize+subpopulation model and a reduced \textasciitilde librarySize model.
+The function uses \textit{edgeR} to fit two models, a full \textasciitilde librarySize+subpopulation model and a reduced \textasciitilde librarySize model.
 For each gene, the deviance explained is then the difference between the deviance of each models, divided by the deviance of the reduced model.
@@
   mixology10x5cl & \cite{tianMixology2018} & 10x & Mixture of 5 cancer cell lines from CellBench \\ 
-  simMix1 & \cite{CrowellMuscat2019} & - & Simulation of 10 human cell subpopulations \\
-  simMix2 & \cite{CrowellMuscat2019} & - & Simulation of 9 mouse cell subpopulations \\
+  simMix1 & - & 10x-based & Simulation of 10 human cell subpopulations \\
+  simMix2 & - & 10x-based & Simulation of 9 mouse cell subpopulations \\
    \hline
@@
 \section*{Additional Files}
-  \subsection*{Additional file 1 --- Sample additional file title}
-    Additional file descriptions text (including details of how to
-    view the file, if it is in a non-standard format or the file extension).  This might
-    refer to a multi-page table or a figure.
-
-  \subsection*{Additional file 2 --- Sample additional file title}
-    Additional file descriptions text.
+  \subsection*{Supplementary figures}
+    % Additional file descriptions text
 

### 2020-01-30 main_figures/clustering.Rmd
updated colours in doublet figure, x label in filtering figure, sticker in pipeline

@@
 ```{r clustering, fig.width=12, fig.height=3}
-res <- readRDS("../data/results_clust.rds")
+res <- readRDS("../data/results_clust_2.rds")
 h <- scrna_evalPlot_clust(res, what="MI", agg.by=c("norm","clustmethod"), scale=TRUE, anno_legend = FALSE)

### 2020-01-30 main_figures/doublets.Rmd
updated colours in doublet figure, x label in filtering figure, sticker in pipeline

@@
 
-pw <- function(i){
+pw <- function(i, dot.size=4, ...){
   scores <- res[[i]][,intersect(names(scoresNames),colnames(res[[i]]))]
   colnames(scores) <- scoresNames[colnames(scores)]
-  plotROCs(scores, factor(res[[i]]$demuxlet_cls, c("SNG","DBL"))) +
+  plotROCs(scores, factor(res[[i]]$demuxlet_cls, c("SNG","DBL")), dot.size=dot.size) +
     ylim(0,1) + scale_color_manual(values=cols)

### 2020-01-30 main_figures/filtering.Rmd
added imputation and clustering results, not final
updated colours in doublet figure, x label in filtering figure, sticker in pipeline

@@
   facet_wrap(~dataset, scales="free") + scale_color_manual(values=cols) + 
-  xlab("Max proportion of subpopulation excluded") + ylab("mean F1 at true number of clusters")
+  xlab("Maximum proportion of cells excluded per subpopulation") + ylab("mean F1 at true number of clusters")
 
@@
 **A:** Filtering on the basis of distance to the whole distribution can lead to strong bias against certain subpopulations.
+The dashed line indicates a threshold of 2.5 median absolute deviations (MADs) from the median of the overall population.
 **B:** Relationship between the maximum subpopulation exclusion rate and the average clustering accuracy per subpopulation across various filtering strategies.

### 2020-01-30 main_figures/normalization.Rmd
updated colours in doublet figure, x label in filtering figure, sticker in pipeline

@@
 ```{r}
-p1 <- scrna_evalPlot_DR(res, "covarRes", agg.by = "norm", anno_legend=FALSE, reorder_rows = TRUE)
+p1 <- scrna_evalPlot_DR(res, "covarRes", agg.by = "norm", anno_legend=FALSE, reorder_rows = )
 p1 <- p1 + scrna_evalPlot_DR(res, "covarRes", covar = "log10_total_features", agg.by = "norm", anno_legend=FALSE, reorder_rows=p1)

### 2020-01-30 bmc_article.tex
Updates from Overleaf

@@
    email={mark.robinson@imls.uzh.ch}
-]{\inits{MR}\fnm{Mark} \snm{Robinson}}
+]{\inits{MR}\fnm{Mark D.} \snm{Robinson}}
 
@@
 \address[id=aff1]{
-  \orgname{Statistical Bioinformatics, IMLS, University of Z\"{u}rich}, % university, etc
+  \orgname{Department of Molecular Life Sciences and SIB Swiss Institute of Bioinformatics, University of Zürich}, % university, etc
   \street{Winterthurerstrasse 190},
   \postcode{8057}
-  \city{Z\"{u}erich},
+  \city{Zürich},
   \cny{Switzerland}
@@
 \address[id=aff2]{
-  \orgname{D-HEST Institute for Neurosciences, ETH Z\"{u}rich},
+  \orgname{D-HEST Institute for Neurosciences, ETH Zürich},
   \street{Winterthurerstrasse 190},
   \postcode{8057}
-  \city{Z\"{u}erich},
+  \city{Zürich},
   \cny{Switzerland}
@@
 %\parttitle{Background} %if any
-The massive growth of single-cell RNA sequencing (scRNAseq) and methods for its analysis is currently only insufficiently accompanied by proper and up-to-date benchmarks that would guide analytical choices.
+The massive growth of single-cell RNA sequencing (scRNAseq) and methods for its analysis still lacks sufficient and up-to-date benchmarks that would guide analytical choices.
 Moreover, current studies are often focused on isolated steps of the process.
@@
 We evaluate common steps of such analyses, including filtering, doublet detection (suggesting a new R package, \textit{scDblFinder}), normalization, feature selection, denoising, dimensionality reduction and clustering.
-On the basis of these analyses, we make a number of key concrete recommendations about analysis choices.
-The evaluation framework, \textit{pipeComp}, has been implemented so as to easily integrate any other step or tool and permit further extending benchmarks (\url{https://github.com/plger/pipeComp}).
+On the basis of these analyses, we make a number of concrete recommendations about analysis choices.
+The evaluation framework, \textit{pipeComp}, has been implemented so as to easily integrate any other step or tool and permit extensible benchmarks (\url{https://github.com/plger/pipeComp}).
 
@@
 This increase in the number of available tools follows the development of new sequencing technologies and the growing number of reported cells, genes and cell populations \cite{SvenssonDB2019}.
-Data processing is a critical step in any scRNAseq analysis, affecting downstream analysis and interpretation.
-It is therefore critical to evaluate the available tools to leverage future studies as it can ultimately increase the power of further differential expression analysis \cite{viethSystematic2019}.
+As data processing is a critical step in any scRNAseq analysis, affecting downstream analysis and interpretation, it is critical to evaluate the available tools.
 
@@
 
-Here, we develop a flexible R framework for pipeline comparison and evaluate in details the various steps of analysis leading from an initial count matrix to a cluster assignment, which are critical in a wide range of applications.
+Here, we develop a flexible R framework for pipeline comparison and evaluate the various steps of analysis leading from an initial count matrix to a cluster assignment, which are critical in a wide range of applications.
 We harnessed real datasets of known cell composition (Table \ref{tab:table1}) and develop new evaluation metrics to investigate in a multilevel fashion the impact of various parameters and variations around a core scRNAseq pipeline.
 Although we use some datasets based on other protocols, our focus is especially on droplet-based datasets that do not include exogenous control RNA (i.e.
-spike-ins); see Table \ref{tab:table1} and \ref{fig:figure1} for a description of the datasets.
+spike-ins); see Table \ref{tab:table1} and Figure \ref{fig:figure1} for a description of the datasets.
 In addition to previously-used benchmark datasets with true cell labels \cite{duoClustering2018,tianMixology2018}, we simulated two datasets with a hierarchical subpopulation structure based on real 10x human and mouse data using \textit{muscat} \cite{CrowellMuscat2019}).
@@
 
-The \textit{pipeComp} package defines a pipeline as, minimally, a list of functions executed consecutively on the output of the previous one (\ref{fig:figure2}A).
+The \textit{pipeComp} package defines a pipeline as, minimally, a list of functions executed consecutively on the output of the previous one (Figure \ref{fig:figure2}A).
 In addition, optional benchmark functions can be set for each step to provide standardized, multi-layered evaluation metrics.
@@
 We therefore also wanted to capture whether the effect of a given parameter alteration is robust to changes in the rest of the pipeline, or rather specific to a set of other pipeline parameters.
-We therefore proceeded in a step-wise fashion, first testing a large variety of parameters at the early steps of the pipeline along with only a set of mainstream options downstream, then selecting main alternatives and proceeding to a more detailed benchmark of the next step (\ref{fig:figure2}B).
+We therefore proceeded in a step-wise fashion, first testing a large variety of parameters at the early steps of the pipeline along with only a set of mainstream options downstream, then selecting main alternatives and proceeding to a more detailed benchmark of the next step (Figure \ref{fig:figure2}B).
 
@@
 A number of methods were developed to identify doublets, in most cases by comparing each cell to artificially-created doublets \citep{mcginnisDoubletfinder2019, LunScran2016, BaisScds2019}.
-We therefore first evaluated the capacity of these methods to detect doublets using the two 10x datasets with cells of different genetic identity \citep{tianMixology2018}, and for which therefore SNP information can be used as ground truth.
-We included a further such dataset\citep{kangMultiplexedDemuxlet2018} not included for the rest of our benchmark due to the absence of true cell labels beside SNPs.
+We therefore first evaluated the capacity of these methods to detect doublets using the two 10x datasets with cells of different genetic identity \citep{tianMixology2018}, using SNP genotypes as the ground truth.
+For the sole purpose of this section, we included an additional dataset with SNP information but lacking true cell labels \citep{kangMultiplexedDemuxlet2018}.
 Of note, SNP-based analyses also call doublets created by cells of the same cell type (but from different individuals), whose identification is generally not the primary aim of doublet detection methods, since these can be considered homotypic (as opposed to neotypic or heterotypic doublets, i.e.
@@
 In addition to these methods, we reasoned that an approach such as \textit{DoubletFinder} could be simplified by being applied directly on counts and by using a pre-clustering to create neotypic/heterotypic doublets more efficiently.
-We therefore developed a simple and fast Bioconductor package implementing this method for doublet detection, \textit{scDblFinder}, with the added advantage of accounting for uncertainty in the expected doublet rate and using meta-cells from the clusters to even include triplets (see methods).
+We therefore developed a simple and fast Bioconductor package implementing this method for doublet detection, \textit{scDblFinder}, with the added advantage of accounting for uncertainty in the expected doublet rate and using meta-cells from the clusters to even include triplets (see Methods).
 
@@
 This has for instance been demonstrated for droplets containing a high content of mitochondrial reads, often as a result of cell degradation and resulting loss of cytoplasmic mRNAs \citep{IlicicLowQual2016}.
-A common practice is to exclude cells that differ considerably from most other cells on the basis of such properties, for instance through the \texttt{isOutlier} function of \textit{scater} that measures, for a given control property, the number of median absolute deviations of each cell from the median of all cells.
+A common practice is to exclude cells that differ considerably from most other cells on the basis of such properties, for instance through the \texttt{isOutlier} function from \textit{scater} that measures, for a given control property, the number of median absolute deviations (MADs) of each cell from the median of all cells.
 Supplementary Figure 1 shows the distributions of some of the typical cell properties commonly used.
@@
 
-We next investigated the impact of filtering according to various criteria (see methods).
+We next investigated the impact of filtering according to various criteria (see Methods).
 An important risk of excluding cells on the basis of their distance from the whole distribution of cells on some properties (e.g.
 library size) is that these properties tend to have different distributions across subpopulations.
-As a result, thresholds in terms of number of median absolute deviations (MADs) from the whole distribution can lead to strong biases against certain subpopulations (Figure \ref{fig:figure4}A).
+As a result, thresholds in terms of number of MADs from the whole distribution can lead to strong biases against certain subpopulations (Figure \ref{fig:figure4}A).
 We therefore examined the tradeoff between the increased accuracy of filtering and the maximum proportion of cells excluded per subpopulation (Figure \ref{fig:figure4}B and Supplementary Figure 8).
 Since filtering changes the relative abundance of the different subpopulations, global clustering accuracy metrics such as ARI are not appropriate here, and we therefore calculated the per-subpopulation precision and recall using the Hungarian algorithm \citep{PapadimitriouHu1998}, and monitored their mean F1 score.
-A first observation was that, although the stringent filtering tended to be associated with an increase in accuracy, it could also become deleterious, and most of the benefits could be achieved without very stringent filtering and minimizing subpopulation bias.
+A first observation was that, although the stringent filtering tended to be associated with an increase in accuracy, it could also become deleterious, and most of the benefits could be achieved without very stringent filtering and minimizing subpopulation bias (Figure \ref{fig:figure4}B).
 Applying the same filtering criteria on individual clusters of cells (identified through scran's \texttt{quickCluster} method) resulted in nearly no cell being filtered out.
@@
 Finally, by changing the distributions, the doublet removal step in conjunction with filtering sometimes resulted in a net decrease in the proportion of excluded cells while retaining or improving accuracy.
-We therefore recommend the use of doublet removal followed by relatively mild filtering, such as is implemented in our `default' filtering (see methods).
+We therefore recommend the use of doublet removal followed by relatively mild filtering, such as is implemented in our `default' filtering (see Methods).
 
@@
 Of note, doublet removal appears to be desirable even when, due to the design, there are no heterotypic doublets in the data.
-The PCA methods refer to multivariate outlier detected as implemented in \textit{scater} (see methods for details).}
+The PCA methods refer to multivariate outlier detected as implemented in \textit{scater} (see Methods for details).}
     \label{fig:figure4}
@@
 
-An important motivation for the development \textit{sctransform} was the observation that, even after normalization, the first principal components of various datasets tended to correlate with library size, suggesting an inadequate normalization \citep{hafemeisterSCtransform2019}.
+An important motivation for the development of \textit{sctransform} was the observation that, even after normalization, the first principal components of various datasets tended to correlate with library size, suggesting an inadequate normalization \citep{hafemeisterSCtransform2019}.
 We therefore first assessed to what extent the first principal component still retained a correlation with the library size and the number of detected features, removing the confounding covariation with the subpopulations (Figure \ref{fig:figure5}A).
@@
 We next investigated the impact of normalization on the separability of the subpopulations (Figure \ref{fig:figure5}B-C).
-Since clustering accuracy metrics such as the ARI are very strongly influenced by the number of clusters, we complemented it with silhouette width\citep{RousseeuwSil1987} and mutual information.
+Since clustering accuracy metrics such as the ARI are very strongly influenced by the number of clusters, we complemented it with silhouette width \citep{RousseeuwSil1987} and mutual information.
 We found most methods (including no normalization at all) to perform fairly well in most of the subpopulations.
@@
 Regressing out covariates systematically gave poorer performance on all metrics.
-\textit{sctransform} systematically outperformed other methods, and even though it was developed to be applied to data with unique molecular identifiers (UMI), it also performed fairly well with Smart-seq protocol (Koh and Kumar datasets).
+\textit{sctransform} systematically outperformed other methods, and even though it was developed to be applied to data with unique molecular identifiers (UMI), it also performed fairly well with the Smart-seq protocol (Koh and Kumar datasets).
 
@@
 A standard clustering pipeline typically involves a step of highly-variable genes selection, which is complicated by the digital nature and the mean-variance relationship of (sc)RNAseq.
-\textit{Seurat}'s earlier approaches involved the use of dispersion estimates standardized for the mean expression levels, while more recent versions (>3) rely on a different measure of variance, again standardized.
+\textit{Seurat}'s earlier approaches involved the use of dispersion estimates standardized for the mean expression levels, while more recent versions ($\geq$3.0) rely on a different measure of variance, again standardized.
 Of note, while adjusting for the mean-variance relationship removes much of the bias towards highly-expressed genes, it is plausible that this relationship may in fact sometimes reflects biological relevance and would be helpful in classifying cell types.
-Another common practice in features selection is to use those with the highest mean expression, and recently, \citep{townesGlmpca2019} instead suggested to use deviance, while \textit{sctransform} provides its own ordering of genes based on transformed variance.
+Another common practice in feature selection is to use those with the highest mean expression, and recently, \citep{townesGlmpca2019} instead suggested to use deviance, while \textit{sctransform} provides its own ordering of genes based on transformed variance.
 
-Reasoning that a selection method should ideally select genes whose variability is more between subpopulations than within, we first assessed to what extent each method selected genes with a high proportion of variance or deviance explained by (real) subpopulation.
+Reasoning that a selection method should ideally select genes whose variability is higher between subpopulations than within, we first assessed to what extent each method selected genes with a high proportion of variance or deviance explained by (real) subpopulation.
 As the proportion of variability in a gene attributable to subpopulations can be measured in various ways, we first compared three such alternatives.
@@
 The standardized measures of variability were systematically worse than their non-standardized counterparts in selecting genes with a high proportion of variance explained by subpopulation.
-Regarding to the percentage of deviance explained however, the standardized measures were often superior(Figure \ref{fig:figure6}A and Supplementary Figures 12-13).
+Regarding the percentage of deviance explained however, the standardized measures were often superior(Figure \ref{fig:figure6}A and Supplementary Figures 12-13).
 Deviance proved the method of choice to prioritize genes with a high variance explained by subpopulations (with mere expression level proving surprisingly good), but did not perform so well to select genes with a high deviance explained.
@@
 We next evaluated how the use of different feature selection methods affected the clustering accuracy (Figure \ref{fig:figure6}B).
-To validate the previous assay on the proportion of variance/ deviance explained by real populations, we included genes that maximized these two latter measures.
+To validate the previous assay on the proportion of variance/deviance explained by real populations, we included genes that maximized these two latter measures.
 Interestingly, while these selections were on average the top-ranking methods, they were not systematically best for all datasets.
@@
 Seurat's PCA procedure however proved superior on all metrics (Figure \ref{fig:figure7}).
-Overall, weighting the principal components by their variance in the same fashion as Seurat's method does had a strong positive impact on silhouette widths and ARI scores.
+Overall, weighting the principal components by their variance, in the same fashion as Seurat's method, had a strong positive impact on silhouette widths and ARI scores.
 
@@
 A common step following dimension reduction is the selection of an appropriate number of dimensions to use for downstream analysis.
-Since euclidean distance decreases as the number of non-discriminating dimensions increases, there is usually a trade-off between selecting enough dimensions to keep most information and excluding smaller dimensions that may represent technical noise or other unwanted sources of variation.
+Since Euclidean distance decreases as the number of non-discriminating dimensions increases, there is usually a trade-off between selecting enough dimensions to keep most information and excluding smaller dimensions that may represent technical noise or other unwanted sources of variation.
 Overall, increasing the number of dimensions robustly led to a decrease in the number of clusters (Supplementary Figures 15).
@@
 
-The MI score and minimum precision, which are largely independent of the estimated number of clusters, were overall higher for the walktrap method (Figure \ref{fig:figure9}), while the mean F1 score favored both \textitt{scran} methods (walktrap and fast greedy) over \textit{Seurat}.
+The MI score and minimum precision, which are largely independent of the estimated number of clusters, were overall higher for the \textit{walktrap} method (Figure \ref{fig:figure9}), while the mean F1 score favored both \textitt{scran} methods (\textit{walktrap} and \textit{fast greedy}) over \textit{Seurat}.
 Finally, the ARI score at the true number of clusters, when available, showed similar performances.
 However, while \textit{Seurat}'s resolution parameter had a large impact on the number of clusters identified, and therefore could be used to find a parameter combination producing the right number of clusters, the number of clusters found by \textit{scran}-based clustering was considerably less influenced by the available parameters (number of nearest neighbors or steps in the random walk - see Supplementary Figure 17).
-The occasional lack of scran-based clustering with the right number of clusters, along with a higher mutual information, suggest that \textit{scran} sometimes simply divides a real subpopulation into two (possibly tracking some unknown biological differences) rather than committing misclassification errors.
-Overall, the walktrap method appeared superior to the fast greedy algorithm and was generally less prone to misclassification than \textit{Seurat} clustering, although the latter offered more control over the resolution.
+The occasional lack of scran-based clustering with the right number of clusters, along with a higher mutual information, suggest that \textit{scran} sometimes simply divides a real subpopulation into two clusters (possibly tracking some unknown biological differences) rather than committing misclassification errors.
+Overall, the \textit{walktrap} method appeared superior to the \textit{fast greedy} algorithm and was generally less prone to misclassification than \textit{Seurat} clustering, although the latter offered more control over the resolution.
 Of note, very difficult subpopulations, both from the Zhengmix8eq and simMix1 datasets, remained very inaccurately classified by all methods and in regards to all metrics.
@@
 To demonstrate this, we evaluated various imputation or denoising techniques based on their impact on classification.
-As preliminary analysis showed that all methods performed equally well or better on normalized data, we applied them after filtering and normalization, but before scaling and reduction.
+Since preliminary analysis showed that all methods performed equally well or better on normalized data, we applied them after filtering and normalization, but before scaling and reduction.
 Although some of the methods (e.g.
-\textit{DRImpute\_process} and \textit{alra\_norm}) did improve the separability of some more elusive subpopulations, no method had a systematically positive impact on the average silhouette width across all subpopulations (Figure \ref{fig:figure10}A).
+\textit{DRImpute\_process} and \textit{alra\_norm}) did improve the separability of the more elusive subpopulations, no method had a systematically positive impact on the average silhouette width across all subpopulations (Figure \ref{fig:figure10}A).
 When restricting ourselves to clustering analyses that yielded the `right' number of clusters, all tested methods improved classification  compared to a scenario with no imputation step (`none' label, Figure \ref{fig:figure10}B).
@@
 \textit{enhance}) consistently underperformed.
-10X datasets, which are typically characterized by a lower per-cell coverage and feature detection rate, expectedly benefited more from imputation, and in this context \textit{DRImpute} and \textit{dca} tended to show the best performance.
+10X datasets, which are typically characterized by a lower per-cell coverage and feature detection rate, benefited more from imputation, and in this context, \textit{DRImpute} and \textit{dca} tended to show the best performance.
 On the contrary, imputing on normalized counts originating from Smart-seq technology was instead rather deleterious to the clustering accuracy.
@@
      \item \textit{sctransform} offered the best overall performance in terms of the separability of the subpopulations, as well as removing the effect of library size and detection rate.
-     \item The common practice of regressing out cell covariates such as the detection rate or proportion of mitochondrial reads nearly always had a negative impact, leading to increased correlation with covariates and decreased clustering accuracy.
+     \item The common practice of regressing out cell covariates, such as the detection rate or proportion of mitochondrial reads nearly always had a negative impact, leading to increased correlation with covariates and decreased clustering accuracy.
 We therefore advise against this practice.
@@
    \begin{itemize}
-    \item We found \textit{scran}-based walktrap clustering to show good performances.
-    \item In cases where prior knowledge can guide the choice of a resolution, \textit{Seurat} can be useful in affording manual control of it while, in the absence of such knowledge, \textit{scran}-based walktrap clustering provided reasonable estimates.
+    \item We found \textit{scran}-based \textit{walktrap} clustering to show good performances.
+    \item In cases where prior knowledge can guide the choice of a resolution, \textit{Seurat} can be useful in affording manual control of it while, in the absence of such knowledge, \textit{scran}-based \textit{walktrap} clustering provided reasonable estimates.
    \end{itemize}
@@
 Given the limited availability of such datasets however, several aspects of single-cell analysis could not be compared, such as batch effect correction or multi-dataset integration.
-In addition, a focus on the identification of the subpopulations might fail to reveal methods which are instead best performing for tasks other than clustering, such as differential expression analysis or trajectory inference.
-Some informative benchmarks have already been performed on some of these topics \citealp{SaelensTraject2019, SonesonDE2018}.
-Yes, such evaluations could benefit from considering methods not in isolation, but as parts of a connected workflow, as we have done here.
-We believe that the \textit{pipeComp} framework is modular and flexible enough to integrate new steps in the pipeline, as it was shown as an example with the evaluation of imputation/ denoising methods.
+For these aspects of scRNAseq processing that are critical in some experimental designs, we refer the reader to previous evaluations \citealp{TranBatch2020}.
+In addition, a focus on the identification of the subpopulations might fail to reveal methods that are instead best performing for tasks other than clustering, such as differential expression analysis or trajectory inference.
+Several informative benchmarks have already been performed on some of these topics \citealp{DalMolinDE2017, JaakkolaDE2017, SaelensTraject2019, SonesonDE2018}.
+Yet, such evaluations could benefit from considering methods not in isolation, but as parts of a connected workflow, as we have done here.
+We believe that the \textit{pipeComp} framework is modular and flexible enough to integrate new steps in the pipeline, as shown by an example with imputation/denoising methods.
 
@@
 Concerning cell filtering, we noticed that the current approach based on whole-population characteristic (e.g.
-MAD cut-off) can be biased against certain subpopulations, suggesting that more refined methods expect multimodal distributions could be used rather than relying on whole-population characteristics.
+MAD cut-off) can be biased against certain subpopulations, suggesting that more refined methods expecting multimodal distributions should be used rather than relying on whole-population characteristics.
 In addition, most common filtering approaches do not harness the relationship between cell QC properties.
 Finally, imputation had a varied impact on the clustering analysis and seemed to be linked to the technology that was used to generate the data.
-We support the hypothesis of \citep{ZhangImput2018} on the respective strengths of linear models and non-linear models and their use to different types of data, such as the performance of non-linear methods in developmental studies.
+We support the hypothesis of \citep{ZhangImput2018} on the respective strengths of linear and non-linear models and their use to different types of data, such as the highest performance of non-linear methods in developmental studies.
 
-\section*{Conclusions}
+\section*{Conclusion}
 
 \textit{pipeComp} is a flexible R framework for evaluating methods and parameter combinations in a complex pipeline, computing on-the-fly multilevel evaluation metrics.
-Application of this framework to scRNAseq clustering enabled us to make concrete recommendations on the steps of filtering, normalization, feature selection, denoising, dimensionality reduction and clustering.
-We show a diversity of multilevel metrics to be more robust, more sensitive, and more nuanced than simply evaluating the final clustering.
-In addition, we provide a new, efficient Bioconductor package for doublet detection.
-Finally, we hope that the \textit{pipeComp} framework can be applied to extend the current benchmark, as well as to apply it to other contexts.
+Applying this framework to scRNAseq clustering enabled us to make concrete recommendations on the steps of filtering, normalization, feature selection, denoising, dimensionality reduction and clustering.
+We demonstrate how a diversity of multilevel metrics can be more robust, more sensitive, and more nuanced than simply evaluating the final clustering performance.
+In addition, we provide a new and efficient Bioconductor package for doublet detection, \texttt{scDblFinder}.
+We hope that the \textit{pipeComp} framework can be applied to extend the current field of benchmarking, as well as to apply it to other range of methods in scRNAseq analysis.
 
@@
 
-\subsubsection{Denoising/imputation}
-Most imputations/ denoising methods were run with the default parameters with the following exceptions; \textit{DrImpute} documentation advises to process the data prior to imputation by removing lowly expressed genes and cells expressing less than 2 genes.
+\subsubsection*{Denoising/imputation}
+Most imputation/denoising methods were run with the default parameters with the following exceptions; \textit{DrImpute} documentation advises to process the data prior to imputation by removing lowly expressed genes and cells expressing less than 2 genes.
 As it is not clear if this step is a hard requirement for the method to perform well, \textit{DrImpute} was run with and without prior processing (\textit{DrImpute\_process} and \textit{DrImpute\_noprocess} labels, respectively).
-\textit{dca} method does not accept non-integers counts but two of the datasets being used contain such table of counts due to multimapping.
+\textit{dca} method does not accept non-integer counts, but two of the datasets used contained such table of counts due to multimapping.
 For these datasets, we rounded up the counts prior to imputation.
@@
 \textit{ENHANCE} uses a k-nearest neighbor aggregation method and automatically estimates the number of neighbors to merge prior to the imputation.
-With the smallest datasets that we used, this lead to an early stop of the function as this parameters was estimated to 1.
+With the smallest datasets, this parameters was estimated to be 1, which lead to an early stop of the function.
 In such cases, we manually set this parameter to 2 for the method to work.
@@
 \section*{Acknowledgements}
-  Text for this section \ldots
+This work was supported by the Swiss National Science Foundation (grants 310030_175841, CRSII5_177208) as well as the Chan
+Zuckerberg Initiative DAF (grant number 2018-182828), an advised fund of Silicon Valley Community Foundation. MDR
+acknowledges support from the University Research Priority Program Evolution in Action at the University of Zurich.
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2020-01-30 supplementary_figures/supp4_selection.Rmd
added figure on nb of clusters, updated suppl

@@
 ```{r, fig.width=8, fig.height=9, results="hide", warnings=FALSE, messages=FALSE}
-plot_grid(plotlist=ll, ncol=2)
+suppressWarnings(plot_grid(plotlist=ll, ncol=2))
 ```
@@
 ll$legend <- get_legend(ll[[1]] + theme(legend.position="right"))
-plot_grid(plotlist=ll, ncol=2)
 ```

### 2020-01-30 supplementary_figures/supp6_clust.Rmd
added figure on nb of clusters, updated suppl

@@
 res <- parsePipNames(res$clustering)
+x <- grep("true\\.nbClusts",colnames(res),value=TRUE)
+names(x) <- x <- sapply(strsplit(x," "),FUN=function(x) x[[1]])
+aris <- as.data.frame(rbindlist(lapply(x, FUN=function(x){
+  x <- res[,paste(x,"ARI"),drop=FALSE]
+  colnames(x) <- "ARI"
+  x
+}), idcol = "dataset"))
+aris$nbClusters <- unlist(lapply(x, FUN=function(x) res[,paste(x,"n_clus")]))
+aris$clust.method <- res$clustmethod
+aris$norm.method <- res$norm
+levels(aris$norm.method) <- c("norm.seurat","sctransform")
+aris <- aris[which(aris$nbClusters <= 40),]
+dd <- data.frame( dataset=names(x), k=sapply(x, FUN=function(x) res[,paste(x,"true.nbClusts")][1]))
+```
+
+```{r, fig.width=7.5, fig.height=7}
+ggplot(aris, aes(nbClusters, ARI, colour=clust.method, shape=norm.method)) + 
+  geom_point() + xlim(c(1,40)) + facet_wrap(~dataset) + xlab("Number of clusters") +
+  geom_vline(aes(xintercept=k), data=dd, linetype="dashed") + theme(legend.position="bottom") + 
+  guides(colour=guide_legend(ncol=1, title = ""), shape=guide_legend(ncol=1, title=""))
+```
+
+### Supplementary Figure `r getFigNb()`
+The number of clusters called has a much bigger impact on the Adjusted Rand Index (ARI) than differences between methods. The dashed line indicates the true number of clusters.
+
+# Supplementary Figure `r getFigNb(TRUE)`
+
+```{r, include=FALSE}
 res <- res[which(res$norm=="norm.seurat" & res$resolution <5),]

### 2020-01-30 supplementary_figures/supp_figures.tex
added figure on nb of clusters, updated suppl

@@
   \postdate{\par}
-    \date{29 Januar, 2020}
+    \date{30 Januar, 2020}
 
@@
 
-\begin{verbatim}
-## Warning: Removed 328867 rows containing missing values (geom_path).
-\end{verbatim}
-
-\begin{verbatim}
-## Warning: Removed 302113 rows containing missing values (geom_path).
-\end{verbatim}
-
-\begin{verbatim}
-## Warning: Removed 101276 rows containing missing values (geom_path).
-\end{verbatim}
-
-\begin{verbatim}
-## Warning: Removed 68502 rows containing missing values (geom_path).
-\end{verbatim}
-
-\begin{verbatim}
-## Warning: Removed 94976 rows containing missing values (geom_path).
-\end{verbatim}
-
-\begin{verbatim}
-## Warning: Removed 101101 rows containing missing values (geom_path).
-\end{verbatim}
-
-\begin{verbatim}
-## Warning: Removed 96012 rows containing missing values (geom_path).
-\end{verbatim}
-
 \includegraphics{supp_figures_files/figure-latex/unnamed-chunk-21-1.pdf}
@@
 \section{Supplementary Figure 13}\label{supplementary-figure-13}}
-
-\begin{verbatim}
-## Warning: Removed 328867 rows containing missing values (geom_path).
-\end{verbatim}
-
-\begin{verbatim}
-## Warning: Removed 302113 rows containing missing values (geom_path).
-\end{verbatim}
-
-\begin{verbatim}
-## Warning: Removed 101276 rows containing missing values (geom_path).
-\end{verbatim}
-
-\begin{verbatim}
-## Warning: Removed 68502 rows containing missing values (geom_path).
-\end{verbatim}
-
-\begin{verbatim}
-## Warning: Removed 94976 rows containing missing values (geom_path).
-\end{verbatim}
-
-\begin{verbatim}
-## Warning: Removed 101101 rows containing missing values (geom_path).
-\end{verbatim}
-
-\begin{verbatim}
-## Warning: Removed 96012 rows containing missing values (geom_path).
-\end{verbatim}
 
@@
 
+The number of clusters called has a much bigger impact on the Adjusted
+Rand Index (ARI) than differences between methods. The dashed line
+indicates the true number of clusters.
+
+\hypertarget{supplementary-figure-18}{%
+\section{Supplementary Figure 18}\label{supplementary-figure-18}}
+
+\includegraphics{supp_figures_files/figure-latex/unnamed-chunk-33-1.pdf}
+
+\hypertarget{supplementary-figure-18-1}{%
+\subsubsection{Supplementary Figure
+18}\label{supplementary-figure-18-1}}
+
 Difference between the number of detected clusters and the number of

### 2020-01-30 bmc_article.tex
Updates from Overleaf

@@
 
-While a number of good comparison and benchmark studies have already been performed on various steps related to scRNAseq analysis and can inform analytical decisions (\citealp{duoClustering2018, SonesonDE2018, SunDimRed2019}), they need constant updating and often leave open many details of an analysis.
+While a number of good comparison and benchmark studies have already been performed on various steps related to scRNAseq analysis and can inform analytical decisions (\citealp{CobosDeconvolution2020, colePerformance2018, DalMolinDE2017, duoClustering2018, freytagComparison2018, HeiserDimred2019, JaakkolaDE2017, SonesonDE2018, SunDimRed2019, tianMixology2018, TranBatch2020, TsuyuzakiPCA2020, viethSystematic2019, WangDE2019, YipHVGs2018, ZhangImput2018}), they need constant updating and often leave open many details of an analysis.
 Another missing aspect of current benchmarking studies is their limitation to capture all aspects of scRNAseq processing workflow.
@@
 Here, we develop a flexible R framework for pipeline comparison and evaluate the various steps of analysis leading from an initial count matrix to a cluster assignment, which are critical in a wide range of applications.
-We harnessed real datasets of known cell composition (Table \ref{tab:table1}) and develop new evaluation metrics to investigate in a multilevel fashion the impact of various parameters and variations around a core scRNAseq pipeline.
+We harnessed real datasets of known cell composition (Table \ref{tab:table1}) and developed new evaluation metrics to investigate in a multilevel fashion the impact of various parameters and variations around a core scRNAseq pipeline.
 Although we use some datasets based on other protocols, our focus is especially on droplet-based datasets that do not include exogenous control RNA (i.e.
@@
 This is especially important because end-point metrics, such as the adjusted Rand index (ARI, \citep{HubertARI1985}) for clustering, are not perfect.
-For example, by far the most important determinant of the ARI is the number of detected clusters: the farther it is from the actual number of subpopulations, the worse the ARI.
+For example, although the meaning of an ARI score is independent of the number of true subpopulations \citep{steinley_properties_2004}, the number of clusters called is by far the most important determinant of the score: the farther it is from the actual number of subpopulations, the worse the ARI (Supplementary Figure 17).
 In this context, one strategy has been to cluster across various resolutions and only consider the results that have the right number of clusters \citep{duoClustering2018}.
@@
 
-Mitochondrial reads have been associated with cell degradation and there is evidence that ribosomal genes can influence the clustering output, hiding other biological structure in the analysis \citep{freytagComparison2018}.
+Mitochondrial reads have been associated with cell degradation, and there is evidence that ribosomal genes can influence the clustering output, hiding other biological structure in the analysis \citep{freytagComparison2018}.
 We therefore investigated whether excluding one category of features or the other, or using only protein-coding genes, had an impact on the ability to distinguish subpopulations (Supplementary Figure 9).
@@
 \textbf{B:} Average silhouette width per true subpopulation, where higher silhouette width means a higher separability.
-\textbf{C:} Clustering accuracy, measured by the ARI at the true number of cluster (left) and by the average mutual information (MI) of the cluster assignment with true subpopulations.}
+\textbf{C:} Clustering accuracy (\textit{Seurat} clustering), measured by the ARI at the true number of cluster (left) and by the average mutual information (MI) of the cluster assignment with true subpopulations.}
     \label{fig:figure5}
@@
 We next investigated the methods of dimensionality reduction.
-Since the various PCA approaches and implementations were recently benchmarked in a similar context \citep{TsuyuzakiPCA2020}, we focused on widely used approaches which had not yet been compared: \textit{Seurat}'s PCA, \textit{scran}'s \texttt{denoise PCA}, and GLM-PCA \citep{townesGlmpca2019}, and combined them (when relevant) with \textit{sctransform} normalization.
+Since the various PCA approaches and implementations were recently benchmarked in a similar context \citep{TsuyuzakiPCA2020}, we focused on widely used approaches which had not yet been compared: \textit{Seurat}'s PCA, \textit{scran}'s \texttt{denoisePCA}, and GLM-PCA \citep{townesGlmpca2019}, and combined them (when relevant) with \textit{sctransform} normalization.
 Given that \textit{Seurat}'s default PCA weights the cell embeddings by the variance of each component, we also evaluated the impact of this weighting with each method.
@@
 Supplementary Figure 15).
-The first components identified by GLM-PCA tended to have a higher proportion of their variance explained by real subpopulations and increase the average silhouette width of already well-defined subpopulations.
-Seurat's PCA procedure however proved superior on all metrics (Figure \ref{fig:figure7}).
-Overall, weighting the principal components by their variance, in the same fashion as Seurat's method, had a strong positive impact on silhouette widths and ARI scores.
+GLM-PCA tended to increase the average silhouette width of already well-defined subpopulations, but \textit{Seurat}'s PCA procedure however proved superior on all metrics (Figure \ref{fig:figure7}).
+Overall, weighting the principal components by their variance (as \textit{Seurat} does) had a positive impact on silhouette widths and ARI scores.
 
@@
 A:} Average silhouette width per subpopulation resulting from combinations of normalization and dimension reductions.
-\textbf{B:} Clustering accuracy , mean proportion of the variance in the first 5 components explained by real subpopulations (center), and median ARI of the resulting clustering (right).}
+\textbf{B:} Clustering accuracy, mean proportion of the variance in the first 5 components explained by real subpopulations (center), and median ARI of the resulting clustering (right).}
     \label{fig:figure7}
@@
 Since Euclidean distance decreases as the number of non-discriminating dimensions increases, there is usually a trade-off between selecting enough dimensions to keep most information and excluding smaller dimensions that may represent technical noise or other unwanted sources of variation.
-Overall, increasing the number of dimensions robustly led to a decrease in the number of clusters (Supplementary Figures 15).
+Overall, increasing the number of dimensions robustly led to a decrease in the number of clusters (Supplementary Figures 10 and 15).
 This tended to be translated to the accuracy of the clustering (Supplementary Figure 16), although in both cases (number of clusters and ARI) \textit{Seurat}'s resolution parameter had a much stronger impact.
@@
 
-Since ARI is dominated by differences in the number of clusters, and no single metric is perfect, we diversified them (Figure \ref{fig:figure9}).
+Since ARI is dominated by differences in the number of clusters (Supplementary Figure 17), and no single metric is perfect, we diversified them (Figure \ref{fig:figure9}).
 Mutual information (MI) has the virtue of not decreasing when a true subpopulation is split into two clusters, which is arguably less problematic (and might well reflect unknown biological subgroups), but as a consequence it can be biased towards methods producing higher resolution clustering.
@@
 Finally, the ARI score at the true number of clusters, when available, showed similar performances.
-However, while \textit{Seurat}'s resolution parameter had a large impact on the number of clusters identified, and therefore could be used to find a parameter combination producing the right number of clusters, the number of clusters found by \textit{scran}-based clustering was considerably less influenced by the available parameters (number of nearest neighbors or steps in the random walk - see Supplementary Figure 17).
-The occasional lack of scran-based clustering with the right number of clusters, along with a higher mutual information, suggest that \textit{scran} sometimes simply divides a real subpopulation into two clusters (possibly tracking some unknown biological differences) rather than committing misclassification errors.
+However, while \textit{Seurat}'s resolution parameter had a large impact on the number of clusters identified, and therefore could be used to find a parameter combination producing the right number of clusters, the number of clusters found by \textit{scran}-based clustering was considerably less influenced by the available parameters (number of nearest neighbors or steps in the random walk - see Supplementary Figure 18).
+The occasional lack of \textit{scran}-based clustering with the right number of clusters, along with a higher mutual information, suggest that \textit{scran} sometimes simply divides a real subpopulation into two clusters (possibly tracking some unknown biological differences) rather than committing misclassification errors.
 Overall, the \textit{walktrap} method appeared superior to the \textit{fast greedy} algorithm and was generally less prone to misclassification than \textit{Seurat} clustering, although the latter offered more control over the resolution.
@@
 Although some of the methods (e.g.
-\textit{DRImpute\_process} and \textit{alra\_norm}) did improve the separability of the more elusive subpopulations, no method had a systematically positive impact on the average silhouette width across all subpopulations (Figure \ref{fig:figure10}A).
+\textit{DRImpute\_process} and \textit{alra\_norm}) did improve the separability of some more elusive subpopulations, no method had a systematically positive impact on the average silhouette width across all subpopulations (Figure \ref{fig:figure10}A).
 When restricting ourselves to clustering analyses that yielded the `right' number of clusters, all tested methods improved classification  compared to a scenario with no imputation step (`none' label, Figure \ref{fig:figure10}B).
@@
 \textit{enhance}) consistently underperformed.
-10X datasets, which are typically characterized by a lower per-cell coverage and feature detection rate, benefited more from imputation, and in this context, \textit{DRImpute} and \textit{dca} tended to show the best performance.
+10X datasets, which are typically characterized by a lower per-cell coverage and feature detection rate, benefited more from imputation, and in this context, \textit{DrImpute} and \textit{dca} tended to show the best performance.
 On the contrary, imputing on normalized counts originating from Smart-seq technology was instead rather deleterious to the clustering accuracy.
-Our recommendation follows the previous benchmarking of imputation strategies performed by \citep{ZhangImput2018}.
-Although this previous study focused on trajectory inference and preservation of data structure as main metrics, \textit{DrImpute} also revealed to be among the top performers and was recommended for clustering analysis (\textit{alra} was however not included in this previous evaluation).
+The good performance of \textit{DrImpute} is in line with a previous study focused on the preservation of data structure in trajectory inference from scRNAseq \citep{ZhangImput2018} (\textit{alra} was however not included in this previous benchmark).
 
@@
      \item Distribution-based cell filtering fails to capture doublets, and should use relatively lenient cutoffs (e.g.
-5 MADs, or 3 MADs in at least 2 distributions) to exclude poor-quality cells while avoiding bias.
+5 MADs, or 3 MADs in at least 2 distributions) to exclude poor-quality cells while avoiding bias against some subpopulations.
      \item Features filtering based on feature type did not appear beneficial.
@@
 Although this platform has been used in almost half of the scRNAseq studies in 2019 \citep{SvenssonDB2019}, other popular technologies such as Drop-seq, InDrops or Smart-seq2/3 were not represented in the present benchmarking.
-Differences between such protocols have a very large impact on clustering \citep{MereuCellAtlas2019}.
+Differences between such protocols can have a very large impact on clustering \citep{MereuCellAtlas2019}.
 Although most top-ranking methods in our comparison performed well on both Smart-seq and 10x datasets that we tested, future benchmarking efforts should strive to include less represented technologies.
-In addition, we did not compare any of the alignment and/or quantification methods used to obtain the count matrix \citep{MereuCellAtlas2019}.
-Some steps, such as the implementation of the PCA step, were also not explored in detail here as they have already been the object of recent and thorough study elsewhere \citep{SunDimRed2019}.
+In addition, we did not compare any of the alignment and/or quantification methods used to obtain the count matrix, which was for instance discussed in \citep{viethSystematic2019}.
+Some steps, such as the implementation of the PCA, were also not explored in detail here as they have already been the object of recent and thorough study elsewhere \citep{SunDimRed2019, TsuyuzakiPCA2020}.
 We also considered only methods relying on Euclidean distance, while correlation was recently reported to be superior\citep{kim_impact_2019} and would require further investigation.
@@
 Here, we chose to concentrate on what could be learned from datasets with known cell labels (as opposed to labels inferred from the data, as in \citep{MereuCellAtlas2019}).
-In contrast to \citep{tianMixology2018}, who relied on RNA mixtures of known proportions, we chose to rely chiefly on real cells and their representative form of variability.
+In contrast to Tian et al.
+\citep{tianMixology2018}, who used on RNA mixtures of known proportions, we chose to rely chiefly on real cells and their representative form of variability.
 Given the limited availability of such datasets however, several aspects of single-cell analysis could not be compared, such as batch effect correction or multi-dataset integration.
-For these aspects of scRNAseq processing that are critical in some experimental designs, we refer the reader to previous evaluations \citealp{TranBatch2020}.
+For these aspects of scRNAseq processing that are critical in some experimental designs, we refer the reader to previous evaluations \citealp{TranBatch2020, StuartIntegration2019}.
 In addition, a focus on the identification of the subpopulations might fail to reveal methods that are instead best performing for tasks other than clustering, such as differential expression analysis or trajectory inference.
-Several informative benchmarks have already been performed on some of these topics \citealp{DalMolinDE2017, JaakkolaDE2017, SaelensTraject2019, SonesonDE2018}.
+Several informative benchmarks have already been performed on some of these topics \citealp{CrowellMuscat2019, DalMolinDE2017, JaakkolaDE2017, SaelensTraject2019, SonesonDE2018, WangDE2019}.
 Yet, such evaluations could benefit from considering methods not in isolation, but as parts of a connected workflow, as we have done here.
@@
 In addition, we provide a new and efficient Bioconductor package for doublet detection, \texttt{scDblFinder}.
-We hope that the \textit{pipeComp} framework can be applied to extend the current field of benchmarking, as well as to apply it to other range of methods in scRNAseq analysis.
+We hope that the \textit{pipeComp} framework can be applied to extend the current field of benchmarking, as well as to apply it to other range of methods.
 
@@
 As it is not clear if this step is a hard requirement for the method to perform well, \textit{DrImpute} was run with and without prior processing (\textit{DrImpute\_process} and \textit{DrImpute\_noprocess} labels, respectively).
-\textit{dca} method does not accept non-integer counts, but two of the datasets used contained such table of counts due to multimapping.
+The \textit{dca} method accepts only integers, while two of the datasets had non-integer quantification of expected counts.
 For these datasets, we rounded up the counts prior to imputation.
@@
 \section*{Acknowledgements}
-This work was supported by the Swiss National Science Foundation (grants 310030_175841, CRSII5_177208) as well as the Chan
-Zuckerberg Initiative DAF (grant number 2018-182828), an advised fund of Silicon Valley Community Foundation. MDR
-acknowledges support from the University Research Priority Program Evolution in Action at the University of Zurich.
+This work was supported by the Swiss National Science Foundation (grants 310030_175841, CRSII5_177208) as well as the Chan Zuckerberg Initiative DAF (grant number 2018-182828), an advised fund of Silicon Valley Community Foundation.
+MDR acknowledges support from the University Research Priority Program Evolution in Action at the University of Zurich.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2020-01-31 draft.tex

\documentclass[11pt]{article}

\renewcommand\familydefault{\sfdefault}
\usepackage[left=3cm,right=3cm,top=2cm,bottom=3cm]{geometry}

% graphics, colors & math
\usepackage{graphicx}
\usepackage[dvipsnames,usenames]{xcolor}
\usepackage{amsmath,amsfonts,nicefrac,textcomp,icomma,bm}

% links & figures
\usepackage{hyperref}
\usepackage[labelfont=bf,font=small]{caption}
\captionsetup[figure]{font={stretch=1}}
\newcommand{\bcaption}[2]{\caption[#1]{\textbf{#1} #2}}

% cross-referencing
\usepackage{xr}
\makeatletter
\newcommand*{\addFileDependency}[1]{%
  \typeout{(#1)}
  \@addtofilelist{#1}
  \IfFileExists{#1}{}{\typeout{No file #1.}}
}
\makeatother

\usepackage{pdfpages}

% line spacing
\usepackage{setspace}
\renewcommand{\baselinestretch}{1.5}
\setlength{\parindent}{2em}

% line numbering
\usepackage{etoolbox,lineno}
\newcommand*{\Section}[1]{\par\nolinenumbers\section*{#1}\linenumbers}

% title layout
\usepackage[affil-it]{authblk}
\renewcommand\Authands{ \& }

\usepackage{lipsum}
\makeatletter
\renewcommand{\maketitle}{\bgroup
\setlength{\parindent}{0pt}\begin{flushleft}
\textbf{\Large{\@title}}\\\vspace{1em}\@author
\end{flushleft}\egroup}
\makeatother

\author[1,2,3]{Pierre-Luc Germain}
\author[1,2]{Anthony Sonrel}
\author[1,2]{Mark D. Robinson}

\affil[1]{Department of Molecular Life Sciences, University of Z{\"u}rich, Switzerland}
\affil[2]{SIB Swiss Institute of Bioinformatics, Z{\"u}rich, Switzerland}
\affil[3]{D-HEST Institute for Neuroscience, Swiss Federal Institute of Technology (ETH), Z{\"u}rich, Switzerland}

\date{\today}

% code font-style
\newsavebox{\foobox}
\newcommand{\slantbox}[2][0]{\mbox{%
        \sbox{\foobox}{#2}%
        \hskip\wd\foobox
        \pdfsave
        \pdfsetmatrix{1 0 #1 1}%
        \llap{\usebox{\foobox}}%
        \pdfrestore
}}
\newcommand\unslant[2][-.25]{\slantbox[#1]{$#2$}}
\newcommand{\code}[1]{\unslant{\mathit{#1}}}

% citations
\usepackage[style=nature,citestyle=numeric,sorting=none]{biblatex}
\renewcommand*{\bibfont}{\footnotesize}
\addbibresource{refs.bib}

\renewcommand{\cite}[1]{\textcolor{Blue}{$^[$\supercite{#1}$^]$}}
\newcommand{\citet}[1]{\citeauthor{#1}\textcolor{Blue}{$^[$\supercite{#1}$^]$}}

% check marks
\usepackage{pifont}
\newcommand{\cmark}{\textcolor{Green}{\ding{51}}}
\newcommand{\xmark}{\textcolor{Red}{\ding{55}}}

% tables
\usepackage{colortbl}

\title{\textit{pipeComp}: A framework for pipeline benchmarking and its application to single-cell RNAseq clustering}

\begin{document}

\thispagestyle{empty}
\maketitle

\linenumbers
%\modulolinenumbers[5]

\Section{Abstract}

The massive growth of single-cell RNA sequencing (scRNAseq) and methods for its analysis still lacks sufficient and up-to-date benchmarks that would guide analytical choices. Moreover, current studies are often focused on isolated steps of the process. Here, we present a flexible R framework for pipeline comparison with multi-level evaluation metrics, and apply it to the benchmark of scRNAseq analysis pipelines using datasets with known cell identities. We evaluate common steps of such analyses, including filtering, doublet detection (suggesting a new R package, \textit{scDblFinder}), normalization, feature selection, denoising, dimensionality reduction and clustering. On the basis of these analyses, we make a number of concrete recommendations about analysis choices. The evaluation framework, \textit{pipeComp}, has been implemented so as to easily integrate any other step or tool and permit extensible benchmarks (\url{https://github.com/plger/pipeComp}).

\newpage
\pagenumbering{arabic}

\clearpage
\singlespacing
\nolinenumbers
\printbibliography

\includepdf[pages=-]{supplementary_figures/supp_figures.pdf}

\end{document}

### 2020-01-31 draft.tex
Updates from Overleaf

@@
 \usepackage[dvipsnames,usenames]{xcolor}
-\usepackage{amsmath,amsfonts,nicefrac,textcomp,icomma,bm}
+\usepackage{amsmath,amssymb, amsfonts,nicefrac,textcomp,icomma,bm}
 
@@
 
-\title{\textit{pipeComp}: A framework for pipeline benchmarking and its application to single-cell RNAseq clustering}
+\title{pipeComp, a general framework for the evaluation of computational pipelines, reveals performant single-cell RNA-seq preprocessing tools}
 
@@
 
-The massive growth of single-cell RNA sequencing (scRNAseq) and methods for its analysis still lacks sufficient and up-to-date benchmarks that would guide analytical choices.
+The massive growth of single-cell RNA-sequencing (scRNAseq) and methods for its analysis still lacks sufficient and up-to-date benchmarks that would guide analytical choices.
 Moreover, current studies are often focused on isolated steps of the process.
-Here, we present a flexible R framework for pipeline comparison with multi-level evaluation metrics, and apply it to the benchmark of scRNAseq analysis pipelines using datasets with known cell identities.
+Here, we present a flexible R framework for pipeline comparison with multi-level evaluation metrics and apply it to the benchmark of scRNAseq analysis pipelines using datasets with known cell identities.
 We evaluate common steps of such analyses, including filtering, doublet detection (suggesting a new R package, \textit{scDblFinder}), normalization, feature selection, denoising, dimensionality reduction and clustering.
 On the basis of these analyses, we make a number of concrete recommendations about analysis choices.
-The evaluation framework, \textit{pipeComp}, has been implemented so as to easily integrate any other step or tool and permit extensible benchmarks (\url{https://github.com/plger/pipeComp}).
+The evaluation framework, \textit{pipeComp}, has been implemented so as to easily integrate any other step or tool, allowing extensible benchmarks and easy application to other fields (\url{https://github.com/plger/pipeComp}).
 
@@
 \pagenumbering{arabic}
+
+\section*{Background}
+
+Single-cell RNA-sequencing (scRNAseq) and the set of attached analysis methods are evolving fast, with more than 560 software tools available to the community \cite{ZappiaDB2018}, roughly half of which are dedicated to tasks related to data processing such as clustering, ordering, dimension reduction or normalization.
+This increase in the number of available tools follows the development of new sequencing technologies and the growing number of reported cells, genes and cell populations \cite{SvenssonDB2019}.
+As data processing is a critical step in any scRNAseq analysis, affecting downstream analysis and interpretation, it is critical to evaluate the available tools.
+
+A number of good comparison and benchmark studies have already been performed on various steps related to scRNAseq analysis and can guide the choice of methodology \cite{CobosDeconvolution2020, colePerformance2018, DalMolinDE2017, duoClustering2018, freytagComparison2018, HeiserDimred2019, JaakkolaDE2017, SonesonDE2018, SunDimRed2019, tianMixology2018, TranBatch2020, TsuyuzakiPCA2020, viethSystematic2019, WangDE2019, YipHVGs2018, ZhangImput2018}.
+However these recommendations need constant updating and often leave open many details of an analysis.
+Another missing aspect of current benchmarking studies is their limitation to capture all aspects of scRNAseq processing workflow.
+Although previous benchmarks already brought valuable recommendations for data processing, some only focused on one aspect of data processing (e.g., \cite{SunDimRed2019}), did not evaluate how the tool selection affects downstream analysis (e.g., \cite{TsuyuzakiPCA2020}) or did not tackle all aspects of data processing, such as doublet identification or cell filtering (e.g., \cite{viethSystematic2019}).
+A thorough evaluation of the tools covering all major processing steps is however urgently needed as previous benchmarking studies highlighted that a combination of tools can have a drastic impact on downstream analysis, such as differential expression analysis and cell-type deconvolution\cite{viethSystematic2019,CobosDeconvolution2020}.
+It is then critical to evaluate not only the single effect of a preprocessing method but also its positive or negative interaction with all parts of a workflow.
+
+Here, we develop a flexible R framework for pipeline comparison and evaluate the various steps of analysis leading from an initial count matrix to a cluster assignment, which are critical in a wide range of applications.
+We collected real datasets of known cell composition (Table \ref{tab:table1}) and used a variety of evaluation metrics to investigate in a multilevel fashion the impact of various parameters and variations around a core scRNAseq pipeline.
+Although we use some datasets based on other protocols, our focus is especially on droplet-based datasets that do not include exogenous control RNA (i.e.
+spike-ins); see Table \ref{tab:table1} and Figure \ref{fig:figure1} for more details.
+In addition to previously-used benchmark datasets with true cell labels \cite{duoClustering2018,tianMixology2018}, we simulated two datasets with a hierarchical subpopulation structure based on real 10x human and mouse data using \textit{muscat} \cite{CrowellMuscat2019}.
+Since graph-based clustering \cite{satijaSeurat2015} was previously shown to consistently perform well across several datasets \cite{duoClustering2018,tianMixology2018}, we used the \textit{Seurat} pipeline as the starting point to perform an integrated investigation of: 1) doublet identification, 2) cell filtering, 3) normalization, 4) feature selection, 5) dimension reduction, 6) clustering.
+We compared competing approaches and also explored more fine-grained parameters and variations on common methods.
+Importantly, the success of methods at a certain analytical step might be dependent on choices at other steps.
+Therefore, instead of evaluating each step in isolation, we developed a general framework for evaluating nested variations on a pipeline and suggest a multilevel panel of metrics.
+Finally, we evaluate several recent methods and provide concrete recommendations.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\textwidth,keepaspectratio]{dataset_description}
+    \caption{\textbf{Overview of the benchmark datasets used.}}
+    \label{fig:figure1}
+\end{figure}
+
+\section*{Results}
+
+\subsection*{A flexible framework for pipeline evaluation}
+
+The \textit{pipeComp} package defines a pipeline as, minimally, a list of functions executed consecutively on the output of the previous one (Figure \ref{fig:figure2}A).
+In addition, optional benchmark functions can be set for each step to provide standardized, multi-layered evaluation metrics.
+Given such a \texttt{PipelineDefinition} object, a set of alternative parameters (which might include different subroutines) and benchmark datasets, the \texttt{runPipeline} function then proceeds through all combinations of arguments, avoiding recomputing the same step twice and compiling evaluations (including running time) on the fly.
+Variations in a given parameter can be evaluated using all metrics from this point downward in the pipeline.
+This is especially important because end-point metrics, such as the adjusted Rand index (ARI) \cite{HubertARI1985} for clustering, are not perfect.
+For example, although the meaning of an ARI score is independent of the number of true subpopulations \cite{steinleyProperties2004}, the number of clusters called is by far the most important determinant of the score: the farther it is from the actual number of subpopulations, the worse the ARI (Supplementary Figure 17).
+In this context, one strategy has been to cluster across various resolutions and only consider the results that have the right number of clusters \cite{duoClustering2018}.
+While this has the virtue of making the results comparable, in practice the number of subpopulations is typically unknown and tools that operate well in this optimal context might not necessarily be best overall.
+Clustering results are also very sensitive and might not always capture improvements in earlier steps of the pipeline.
+We therefore favour monitoring several complementary metrics across multiple steps of the process.
+
+A clustering output is relatively fragile to variations in the pipeline.
+We therefore wanted to capture whether the effect of a given parameter alteration is robust to changes in the rest of the pipeline, or rather specific to a set of other pipeline parameters.
+We proceeded in a step-wise fashion, first testing a large variety of parameters at the early steps of the pipeline along with only a set of mainstream options downstream, then selecting main alternatives and proceeding to a more detailed benchmark of the next step (Figure \ref{fig:figure2}B).
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\textwidth,keepaspectratio]{main_figures/pipeline_explanation.png}
+    \caption{\textbf{Overview of the \textit{pipeComp} framework and its application to a scRNAseq clustering pipeline.
+A:} The package is built around a \texttt{PipelineDefinition} S4 class which defines a set of functions to be executed consecutively, as well as optional evaluation functions for each step.
+Each step can accept of number of parameters whose alternative values are provided as a list.
+All (or subsets of) combinations of parameters can then be simultaneously ran and evaluated using the \texttt{runPipeline} function.
+\textbf{B:} Scheme representing the application of \textit{pipeComp} to evaluate a range of methods that are commonly used in scRNAseq studies and some of the metrics monitored at various steps.}
+    \label{fig:figure2}
+\end{figure}
+
+\subsection*{Filtering}
+
+\subsubsection*{Doublet detection}
+
+Doublets, defined as two cells sequenced under the same cellular barcode (e.g., being captured in the same droplet), are fairly frequent in scRNAseq datasets, with estimates ranging from 1 to 10\% depending on the platform and cell concentration used \cite{bloomEstimating2018,kangMultiplexedDemuxlet2018}.
+While doublets of the same cell type are relatively innocuous for most downstream applications due to their conservation of the relative expression between genes, doublets formed from different cell types or states are likely to be misclassified and could potentially distort downstream analysis.
+In some cases, doublets can be identified through their unusually high number of reads and detected features, but this is not always the case (Supplementary Figure 2).
+A number of methods were developed to identify doublets, in most cases by comparing each cell to artificially-created doublets \cite{mcginnisDoubletfinder2019, LunScran2016, BaisScds2019}.
+We first evaluated the capacity of these methods to detect doublets using the two 10x datasets with cells of different genetic identity \cite{tianMixology2018}, using SNP genotypes as the ground truth.
+For the sole purpose of this section, we included an additional dataset with SNP information but lacking true cell labels \cite{kangMultiplexedDemuxlet2018}.
+Of note, SNP-based analyses also call doublets created by cells of the same cell type (but from different individuals) and which are generally described as homotypic (as opposed to neotypic or heterotypic doublets, i.e.
+doublets from different cell types).
+These homotypic doublets might not be identifiable from the mere gene counts, and their identification is not generally the primary aim of doublet callers since they are often considered innocuous and, when across individuals, can be identified through other means (e.g.
+SNPs).
+We therefore do not expect a perfect accuracy in datasets involving cells of the same type across individuals (as in the demuxlet dataset).
+
+We tested \textit{DoubletFinder} \cite{mcginnisDoubletfinder2019} and \textit{scran}'s \texttt{doubletCells} \cite{LunScran2016}, both of which use similarity to artificial doublets, and \textit{scds} \cite{BaisScds2019}, which relies on a combination of co-expression and binary classification.
+\textit{DoubletFinder} integrates a thresholding based on the proportion of expected doublets, while \textit{scran} and \textit{scds} return scores that must be manually thresholded.
+In these cases, we ensured that the right number of cells would be called doublets.
+In addition to these methods, we reasoned that an approach such as \textit{DoubletFinder} could be simplified by being applied directly on counts and by using a pre-clustering to create neotypic/heterotypic doublets more efficiently.
+We therefore developed a simple and fast Bioconductor package implementing this method for doublet detection, \textit{scDblFinder}, with the added advantage of accounting for uncertainty in the expected doublet rate and using meta-cells from the clusters to even include triplets (see Methods).
+
+While most methods accurately identified the doublets in the 3 cell lines dataset (mixology10x3cl), the other two datasets proved more difficult (Figure \ref{fig:figure3}A).
+\textit{scDblFinder} achieved comparable or better accuracy than top alternatives while being the fastest method (Figure \ref{fig:figure3}B).
+Across datasets, cells called as doublets tended to be classified in the wrong cluster more often than other cells (Figure \ref{fig:figure3}C).
+We also found that \textit{scDblFinder} improved the accuracy of the clustering across all benchmark datasets even when, by design, the data contained no heterotypic doublet (Figure \ref{fig:figure4}).
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\textwidth]{{main_figures/doublets_files/figure-html/figure1-1.png}}
+    \caption{\textbf{Identification of doublet cells.
+A:} Receiver operating characteristic (ROC) curves of the tested doublet detection methods for three datasets with SNP-identified doublets.
+Dots indicate threshold determined by the true number of doublets.
+\textbf{B:} Running time of the different methods (DoubletFinder failed on one of the datasets).
+\textbf{C:} Rate of misclassification of the cells identified by \textit{scDblFinder} as doublets (DBL) or singlets (SNG), across a large range of clustering analysis.
+Even in datasets which should not have neotypic doublets, the cells identified as such tend to be misclassified.
+\label{fig:figure3}}
+\end{figure}
+
+\subsubsection*{Excluding more cells is not necessarily better}
+
+Beyond doublets, a dataset might include low-quality cells whose elimination would reduce noise.
+This has for instance been demonstrated for droplets containing a high content of mitochondrial reads, often as a result of cell degradation and resulting loss of cytoplasmic mRNAs \cite{IlicicLowQual2016}.
+A common practice is to exclude cells that differ considerably from most other cells on the basis of such properties.
+This can for instance be performed through the \texttt{isOutlier} function from \textit{scater} that measures, for a given control property, the number of median absolute deviations (MADs) of each cell from the median of all cells.
+Supplementary Figure 1 shows the distributions of some of the typical cell properties commonly used.
+Of note, these properties tend to be correlated, with some exceptions.
+For example, while a high proportion of mitochondrial reads is often correlated with a high proportion of the counts in the top features, there can also be other reasons for an over-representation of highly-expressed features (Supplementary Figure 3), such as an over-amplification in non-UMI datasets.
+In our experience, 10X datasets also exhibit a very strong correlation between the total counts and the total features even across very different cell types (Supplementary Figure 4).
+We therefore also measure the ratio between the two and treat cells strongly departing from this trend with suspicion.
+
+Reasoning that the cells we wish to exclude are the cells that would be misclassified, we measured the rate of misclassification of each cell in each dataset across a variety of clustering pipelines, correcting for the median misclassification rate of the subpopulation and then evaluated what properties could be predictive of misclassification (Supplementary Figures 5-7).
+We could not identify any property or simple combination thereof that would be consistently predictive of misclassification; the only feature that consistently stood out across multiple datasets (the Zheng datasets) was that cells with very high read counts have a higher chance of being misclassified.
+
+We next investigated the impact of filtering according to various criteria (see Methods).
+An important risk of excluding cells on the basis of their distance from the whole distribution of cells on some properties (e.g.
+library size) is that these properties tend to have different distributions across subpopulations.
+As a result, thresholds in terms of number of MADs from the whole distribution can lead to strong biases against certain subpopulations (Figure \ref{fig:figure4}A).
+We therefore examined the tradeoff between the increased accuracy of filtering and the maximum proportion of cells excluded per subpopulation (Figure \ref{fig:figure4}B and Supplementary Figure 8).
+Since filtering changes the relative abundance of the different subpopulations, global clustering accuracy metrics such as ARI are not appropriate here.
+We therefore calculated the per-subpopulation precision and recall using the Hungarian algorithm \cite{PapadimitriouHu1998} and monitored their mean F1 score.
+A first observation was that, although the stringent filtering tended to be associated with an increase in accuracy, it could also become deleterious and most of the benefits could be achieved without very stringent filtering and minimizing subpopulation bias (Figure \ref{fig:figure4}B).
+Applying the same filtering criteria on individual clusters of cells (identified through scran's \texttt{quickCluster} method) resulted in nearly no cell being filtered out.
+This suggests that filtering on the global population tends to discard cells of subpopulations with more extreme properties (e.g.
+high library size), rather than low-quality cells.
+Finally, by changing the distributions, the doublet removal step in conjunction with filtering sometimes resulted in a net decrease in the proportion of excluded cells while retaining or improving accuracy.
+We therefore recommend the use of doublet removal followed by relatively mild filtering, such as is implemented in our `default' filtering (see Methods).
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\textwidth]{{main_figures/filtering_files/figure-html/figure2-1.png}}
+    \caption{\textbf{A:} Filtering on the basis of distance to the whole distribution can lead to strong bias against certain subpopulations.
+The dashed line indicates a threshold of 2.5 median absolute deviations (MADs) from the median of the overall population.
+\textbf{B:} Relationship between the maximum subpopulation exclusion rate and the average clustering accuracy per subpopulation across various filtering strategies.
+Of note, doublet removal appears to be desirable even when, due to the design, there are no heterotypic doublets in the data.
+The PCA methods refer to multivariate outlier detected as implemented in \textit{scater} (see Methods for details).}
+    \label{fig:figure4}
+\end{figure}
+
+\subsubsection*{Filtering features by type}
+
+Mitochondrial reads have been associated with cell degradation and there is evidence that ribosomal genes can influence the clustering output, hiding other biological structure in the analysis \cite{freytagComparison2018}.
+We therefore investigated whether excluding one category of features or the other, or using only protein-coding genes, had an impact on the ability to distinguish subpopulations (Supplementary Figure 9).
+Removal of ribosomal genes robustly reduced the quality of the clustering, suggesting that they represent real biological differences between subpopulations.
+Removing mitochondrial genes and restricting to protein-coding genes had a very mild impact.
+
+\subsection*{Normalization and scaling}
+
+We next investigated the impact of different normalization strategies.
+Beside the standard log-normalization included in \textit{Seurat}, we tested \textit{scran}'s pooling-based normalization \cite{lunPooling2016}, \textit{sctransform}'s variance-stabilizing transformation \cite{hafemeisterSCtransform2019}, and normalization based on stable genes \cite{linStableGenes2018, deekeStablyExpressed2018}.
+In addition to log-normalization, the standard \textit{Seurat} clustering pipeline performs per-feature unit-variance scaling so that the PCA is not too strongly dominated by highly-expressed features.
+We therefore included versions of the different normalization procedures, with or without a subsequent scaling (\textit{sctransform}'s variance-stabilizing transformation involves a transformation analogous to scaling).
+\textit{Seurat}'s scaling function also includes the option to regress out the effect of certain covariates.
+We tested its performance by using the proportion of mitochondrial counts and the number of detected features as covariates.
+Finally, it has been proposed that the use of stable genes, in particular cytosolic ribosomal genes, can be used to normalize scRNAseq\cite{deekeStablyExpressed2018}.
+We therefore evaluated a simple linear normalization based on the sum of these genes, as well as nuclear genes.
+
+An important motivation for the development of \textit{sctransform} was the observation that, even after normalization, the first principal components of various datasets tended to correlate with library size, suggesting an inadequate normalization \cite{hafemeisterSCtransform2019}.
+However, as library size tends to vary across subpopulations, part of this effect can simply reflect biological differences.
+We therefore assessed to what extent the first principal component still retained a correlation with the library size and the number of detected features, removing the confounding covariation with the subpopulations (Figure \ref{fig:figure5}A).
+The simple step of scaling tended to remove much of the correlation with these features and, in the absence of scaling, standard \textit{Seurat} normalization resulted in fairly high correlation with technical covariates.
+\textit{sctransform} led to the lowest correlation but most methods, including normalization based on stable genes, were able to remove most of the correlation when combined with scaling.
+The only exception is \textit{Seurat} normalization with scaling regressing out covariates, which surprisingly increased the correlation in 8 of the 9 datasets.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\textwidth]{{main_figures/normalization_files/figure-html/figure3-1.png}}
+    \caption{\textbf{Evaluation of normalization procedures.} `Seurat.feat\_mt\_regress' regresses out the number of features and proportion of mitochondrial reads during scaling; `Seurat.feat\_regress` regresses out the number of features only.
+Residual correlation of the first principal component with library size (left) and detection rate (right), after accounting for biological differences between subpopulations.
+\textbf{B:} Average silhouette width per true subpopulation, where higher silhouette width means a higher separability.
+\textbf{C:} Clustering accuracy (\textit{Seurat} clustering), measured by the ARI at the true number of cluster (left) and by the average mutual information (MI) of the cluster assignment with true subpopulations.}
+    \label{fig:figure5}
+\end{figure}
+
+We further evaluated normalization methods by investigating their impact on the separability of the subpopulations (Figure \ref{fig:figure5}B-C).
+Since clustering accuracy metrics such as the ARI are very strongly influenced by the number of clusters, we complemented it with silhouette width \cite{RousseeuwSil1987} and mutual information.
+We found most methods (including no normalization at all) to perform fairly well in most of the subpopulations.
+Scaling tended to reduce the average silhouette width of some subpopulations and to increase that of some less distinguishable ones and was generally, but not always, beneficial on the accuracy of the final clustering.
+Regressing out covariates systematically gave poorer performance on all metrics.
+\textit{sctransform} systematically outperformed other methods and, even though it was developed to be applied to data with unique molecular identifiers (UMI), it also performed fairly well with the Smart-seq protocol (Koh and Kumar datasets).
+
+Finally, we monitored whether, under the same downstream clustering analysis, different normalization methods tended to lead to an over- or under-estimation of the number of clusters.
+Although some methods had a tendency to lead to a higher (e.g.
+\textit{sctransform}) or lower (e.g.
+stable genes) number of clusters, the effect was very mild and not entirely systematic (Supplementary Figure 10).
+
+\subsection*{Feature selection}
+
+A standard clustering pipeline typically involves a step of highly-variable genes selection, which is complicated by the digital nature and the mean-variance relationship of (sc)RNAseq.
+\textit{Seurat}'s earlier approaches involved the use of dispersion estimates standardized for the mean expression levels, while more recent versions ($\geq$3.0) rely on a different measure of variance, again standardized.
+While adjusting for the mean-variance relationship removes much of the bias towards highly-expressed genes, it is plausible that this relationship may in fact sometimes reflects biological relevance and would be helpful in classifying cell types.
+Another common practice in feature selection is to use those with the highest mean expression.
+Recently, \cite{townesGlmpca2019} instead suggested to use deviance, while \textit{sctransform} provides its own ordering of genes based on transformed variance.
+
+Reasoning that a selection method should ideally select genes whose variability is higher between subpopulations than within, we first assessed to what extent each method selected genes with a high proportion of variance or deviance explained by (real) subpopulation.
+As the proportion of variability in a gene attributable to subpopulations can be measured in various ways, we first compared three approaches: ANOVA on log-normalized count, ANOVA on \textit{sctransform} normalization, and deviance explained.
+The ANOVAs performed on a standard \textit{Seurat} normalization and on \textit{sctransform} data were highly correlated (Supplementary Figure 11A).
+These estimates were also in good agreement with the deviance explained, although lowly-expressed genes could have a high deviance explained without having much of their variance explained by subpopulation (Supplementary Figure 11B-D).
+We therefore compared the proportion of the cumulative variance/deviance explained by the top X genes that could be retrieved by each gene ranking method (Supplementary Figures 12-13).
+We first focused on the first 1000 genes to highlight the differences between methods, although a higher number of selected genes decreased the differences between methods (Supplementary Figures 12-14).
+The standardized measures of variability were systematically worse than their non-standardized counterparts in selecting genes with a high proportion of variance explained by subpopulation.
+Regarding the percentage of deviance explained however, the standardized measures were often superior (Figure \ref{fig:figure6}A and Supplementary Figures 12-13).
+Deviance proved the method of choice to prioritize genes with a high variance explained by subpopulations (with mere expression level proving surprisingly good) but did not perform so well to select genes with a high deviance explained.
+
+\begin{figure}
+    \centering
+    \includegraphics[scale=0.6]{{main_figures/selection_files/figure-html/figure4-1.png}}
+    \caption{\textbf{Evaluation of feature selection methods.
+A:} Ability of different feature ranking methods to capture genes with a high proportion of variance (left) or deviance (right) explained by real subpopulations.
+The asterix denotes the default Seurat method.
+\textbf{B:} Accuracy of clusterings (at the true number of clusters) when selecting 1000 genes using the given methods.
+Based on standard \textit{Seurat} normalization (left) or \textit{sctransform} (right).
+The  \textit{vst.varExp} and \textit{devianceExplained} methods correspond to the estimates used in \textbf{A} to evaluate the selection methods and were included here only for validation purpose.}
+    \label{fig:figure6}
+\end{figure}
+
+We next evaluated how the use of different feature selection methods affected the clustering accuracy (Figure \ref{fig:figure6}B).
+To validate the previous assay on the proportion of variance/deviance explained by real populations, we included genes that maximized these two latter measures.
+Interestingly, while these selections were on average the top-ranking methods, they were not systematically best for all datasets.
+The previous observations were reflected in the ARI of the resulting clustering (Figure \ref{fig:figure6}B): non-standardized measures of variability, including mere expression level, tended to outperform more complex metrics.
+In general, we found deviance and unstandardized estimates of variance to provide the best results across datasets and normalization methods.
+Increasing the number of features selected also systematically led to an increase in the accuracy of the clustering, typically plateauing after 4000 features (Supplementary Figure 14).
+
+\subsection*{Dimensionality reduction}
+
+Since the various PCA approaches and implementations were recently benchmarked in a similar context \cite{TsuyuzakiPCA2020}, we focused on widely used approaches which had not yet been compared: \textit{Seurat}'s PCA, \textit{scran}'s \texttt{denoisePCA}, and GLM-PCA \cite{townesGlmpca2019}.
+When relevant, we combined them with \textit{sctransform} normalization.
+Given that \textit{Seurat}'s default PCA weights the cell embeddings by the variance of each component, we also evaluated the impact of this weighting with each method.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\textwidth]{{main_figures/dr_files/figure-html/dimred-1.png}}
+    \caption{\textbf{Evaluation of common dimensionality reduction methods.
+A:} Average silhouette width per subpopulation resulting from combinations of normalization and dimension reductions.
+\textbf{B:} Clustering accuracy, mean proportion of the variance in the first 5 components explained by real subpopulations (center), and median ARI of the resulting clustering (right).}
+    \label{fig:figure7}
+\end{figure}
+
+The impact of the choice of dimensionality reduction method was far greater than that of normalization or feature selection (e.g.
+Supplementary Figure 15).
+GLM-PCA tended to increase the average silhouette width of already well-defined subpopulations, but \textit{Seurat}'s PCA procedure however proved superior on all metrics (Figure \ref{fig:figure7}).
+Overall, weighting the principal components by their variance (as \textit{Seurat} does) had a positive impact on silhouette widths and ARI scores.
+
+\subsubsection*{Estimating the number of dimensions}
+
+A common step following dimension reduction is the selection of an appropriate number of dimensions to use for downstream analysis.
+Since Euclidean distance decreases as the number of non-discriminating dimensions increases, there is usually a trade-off between selecting enough dimensions to keep most information and excluding smaller dimensions that may represent technical noise or other unwanted sources of variation.
+Overall, increasing the number of dimensions robustly led to a decrease in the number of clusters (Supplementary Figures 10 and 15).
+This tended to affect the accuracy of the clustering (Supplementary Figure 16), although in both cases (number of clusters and ARI) \textit{Seurat}'s resolution parameter had a much stronger impact.
+
+Different approaches have been proposed to select the appropriate number of dimensions, from the visual identification of an 'Elbow' (inflexion point) of the variance explained, to more complex algorithms.
+We evaluated the performance of dimensionality estimators implemented in the \textit{intrinsicDimension} package \cite{johnssonDimensionality2015}, as well as common procedures such as the `elbow` method (inflexion point in the variance explained by each component), some scRNAseq-specific methods such as the JackStraw procedure \cite{ChungJackstraw2015} or \textit{scran}'s \texttt{denoisePCA} \cite{LunScran2016}, and the recent application of Fisher Separability analysis \cite{AlberganteSepar2019}.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\textwidth]{{main_figures/dimensionality_files/figure-html/figure6-1.png}}
+    \caption{\textbf{Estimating dimensionality.
+A:} Estimated dimensionality using the proportion of variance in each component explained by the true subpopulations.
+\textbf{B:} Difference between `real' dimensionality (from \textbf{A}) and dimensionality estimation methods, along with computing time.
+\textbf{C:} Average silhouette width per subpopulation across a selection of methods, combined with \textit{sctransform} or standard \textit{Seurat} normalization.
+\textbf{D:} Clustering accuracy across the normalization/dimensionality estimation methods.}
+    \label{fig:figure8}
+\end{figure}
+
+We compared the various estimates of dimensionality in their ability to retrieve the intrinsic number of dimensions in a dataset, based on Seurat's weighted PCA space.
+As a first approximation of the true dimensionality, we computed the variance in each principal component that was explained by the subpopulations, which sharply decreased after the first few components in most datasets (Figure \ref{fig:figure8}A).
+Figure \ref{fig:figure8}B shows the difference between the dimension estimates of the above methods and that based on the subpopulations (i.e.
+from Figure \ref{fig:figure8}A).
+Of note, the methods differ widely in view of their computing time (Figure \ref{fig:figure8}B) and we saw no relationship between the accuracy of the estimates and the complexity of the method.
+Reasoning that over-estimating the number of dimensions is less problematic than under-estimating it, we kept the former methods for a full analysis of their impact on clustering (Figure \ref{fig:figure8}C-D), when combined with sctransform or the standard Seurat normalization.
+Although most methods performed well on the various clustering measures, the global maximum likelihood based on translated Poisson Mixture Model (\textit{maxLikGlobal}) provided the dimensionality estimate that best separated the subpopulations (Figure \ref{fig:figure8}C) and resulted in the best clustering accuracy (Figure \ref{fig:figure8}D).
+This method systematically selected many more components than were associated with the subpopulations, suggesting that although these additional components appear individually uninformative, in combination they nevertheless contribute to classification.
+
+\subsection*{Clustering}
+
+The last step evaluated in our pipeline was clustering.
+Given previous works on the topic \cite{duoClustering2018,freytagComparison2018} and the success of graph-based clustering methods for scRNAseq, we restricted our evaluation to Seurat's method and two \textit{scran} SNN-based clustering approaches, based on random walks (\textit{walktrap} method) or on the optimization of the modularity score (\textit{fast\_greedy}).
+Again, the tested methods were combined with \textit{Seurat}'s standard normalization and \textit{sctransform}, otherwise using the parameters found optimal in the previous steps.
+
+Since ARI is dominated by differences in the number of clusters (Supplementary Figure 17) and no single metric is perfect, we diversified them (Figure \ref{fig:figure9}).
+Mutual information (MI) has the virtue of not decreasing when a true subpopulation is split into two clusters, which is arguably less problematic (and might well reflect unknown biological subgroups), but as a consequence it can be biased towards methods producing higher resolution clustering.
+Similarly, precision per true subpopulation is considerably more robust to differences in the number of clusters.
+We also tracked the mean F1 score and the ARI at the true number of clusters.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\textwidth]{{main_figures/clustering_files/figure-html/clustering-1.png}}
+    \caption{\textbf{Evaluation of clustering methods.} Clustering accuracy of common clustering tools in combination with \textit{sctransform} and standard Seurat's normalization.}
+    \label{fig:figure9}
+\end{figure}
+
+The MI score and minimum precision, which are largely independent of the estimated number of clusters, were overall higher for the \textit{walktrap} method (Figure \ref{fig:figure9}), while the mean F1 score favored both \textit{scran} methods (\textit{walktrap} and \textit{fast greedy}) over \textit{Seurat}.
+Finally, the ARI score at the true number of clusters, when available, showed similar performances.
+However, because \textit{Seurat}'s resolution parameter had a large impact on the number of clusters identified, \textit{Seurat} could always be tweaked into producing the right number of clusters.
+Instead, the number of clusters found by \textit{scran} was considerably less influenced by the available parameters (number of nearest neighbors or steps in the random walk - see Supplementary Figure 18), and as a result \textit{scran}-based clustering sometimes never produced a partitioning with the right number of clusters.
+This observation, along with \textit{scran}'s higher mutual information, suggest that \textit{scran} sometimes simply divides a real subpopulation into two clusters (possibly tracking some unknown biological differences) rather than committing misclassification errors.
+Overall, the \textit{walktrap} method appeared superior to the \textit{fast greedy} algorithm and was generally less prone to misclassification than \textit{Seurat} clustering, although the latter offered more control over the resolution.
+Finally, some poorly distinguishable subpopulations from both the Zhengmix8eq and simMix1 datasets remained very inaccurately classified by all methods and in regards to all metrics.
+
+\subsection*{Further extensions to the pipeline: imputation/denoising}
+
+The basic pipeline presented here can be extended with additional analysis steps while keeping the same evaluation metrics.
+To demonstrate this, we evaluated various imputation or denoising techniques based on their impact on classification.
+Since preliminary analysis showed that all methods performed equally well or better on normalized data, we applied them after filtering and normalization, but before scaling and reduction.
+Although some of the methods (e.g.
+\textit{DRImpute\_process} and \textit{alra\_norm}) did improve the separability of some more elusive subpopulations, no method had a systematically positive impact on the average silhouette width across all subpopulations (Figure \ref{fig:figure10}A).
+When restricting ourselves to clustering analyses that yielded the `right' number of clusters, all tested methods improved classification  compared to a scenario with no imputation step (`none' label, Figure \ref{fig:figure10}B).
+However, the situation was not so straightforward with alternative metrics, where some methods (e.g.
+\textit{enhance}) consistently underperformed.
+10X datasets, which are typically characterized by a lower per-cell coverage and feature detection rate, benefited more from imputation and, in this context, \textit{DrImpute} and \textit{dca} tended to show the best performance.
+On the contrary, imputing on normalized counts originating from Smart-seq technology was instead rather deleterious to the clustering accuracy.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\textwidth]{{main_figures/imputation_files/figure-html/imputation-1.png}}
+    \caption{\textbf{Imputation/denoising methods} Average silhouette width per subpopulation (\textbf{A}) and clustering accuracy (\textbf{B}) with or without (\textit{none}) application of a denoising/imputation method}
+    \label{fig:figure10}
+\end{figure}
+
+\section*{Discussion}
+
+\subsection*{Concrete recommendations}
+
+On the basis of our findings, we can make a number of concrete analysis recommendations:
+
+\begin{enumerate}
+   \item Filtering
+   \begin{itemize}
+     \item Doublet detection and removal is advised and can be performed at little computing cost with software such as \textit{scDblFinder} or \textit{scds}.
+     \item Distribution-based cell filtering fails to capture doublets and should use relatively lenient cutoffs (e.g.
+5 MADs, or 3 MADs in at least 2 distributions) to exclude poor-quality cells while avoiding bias against some subpopulations.
+     \item Features filtering based on feature type did not appear beneficial.
+   \end{itemize}
+   \item Normalization and scaling
+   \begin{itemize}
+     \item Most normalization methods tested yielded a fair performance, especially when combined with scaling, which tended to have a positive impact on clustering.
+     \item \textit{sctransform} offered the best overall performance in terms of the separability of the subpopulations, as well as removing the effect of library size and detection rate.
+     \item The common practice of regressing out cell covariates, such as the detection rate or proportion of mitochondrial reads nearly always had a negative impact, leading to increased correlation with covariates and decreased clustering accuracy.
+We therefore advise against this practice.
+   \end{itemize}
+   \item Feature selection
+   \begin{itemize}
+     \item Deviance \cite{townesGlmpca2019} offered the best ranking of genes for feature selection.
+     \item Increasing the number of features included tended to lead to better classifications, plateauing from 4000 features in our datasets.
+   \end{itemize}
+   \item Denoising/imputation
+   \begin{itemize}
+       \item Denoising appeared beneficial to the identification of subpopulations in 10x datasets, but not in Smart-seq datasets.
+       \item We found especially \textit{alra} (with prior normalization), \textit{DrImpute} (with prior processing) and \textit{dca} to offer the best performances.
+   \end{itemize}
+   \item PCA
+   \begin{itemize}
+     \item Similarly to previous reports \cite{SunDimRed2019}, we recommend the irlba-based PCA using weighting of the components, as implemented in \textit{Seurat}. 
+     \item Instead of the common elbow or jackstraw methods for deciding on the components to include, we recommend the \textit{global maximum likelihood based on translated Poisson Mixture Model} method (e.g.
+implemented in the  \texttt{maxLikGlobalDimEst} function of \texttt{intrinsicDimension}).
+   \end{itemize}
+   \item Clustering
+   \begin{itemize}
+    \item We found \textit{scran}-based \textit{walktrap} clustering to show good performances.
+    \item In cases where prior knowledge can guide the choice of a resolution, \textit{Seurat} can be useful in affording manual control of it while, in the absence of such knowledge, \textit{scran}-based \textit{walktrap} clustering provided reasonable estimates.
+   \end{itemize}
+\end{enumerate}
+
+\subsection*{Limitations and open questions}
+
+In this study, we evaluated tools commonly used for the processing of scRNAseq with a focus on droplet-based datasets, namely from the 10x technology.
+Although this platform has been used in almost half of the scRNAseq studies in 2019 \cite{SvenssonDB2019}, other popular technologies such as Drop-seq, InDrops or Smart-seq2/3 were not represented in the present benchmarking.
+Differences between such protocols, even among droplet-based technologies, can have a very large impact on cell capture efficiency, cell numbers and clustering \cite{MereuCellAtlas2019, ZhangDroplet2019, SalomonDroplet2019}.
+Although most top-ranking methods in our comparison performed well on both Smart-seq and 10x datasets that we tested, future benchmarking efforts should strive to include less represented technologies.
+In addition, we did not compare any of the alignment and/or quantification methods used to obtain the count matrix, which was for instance discussed in \cite{viethSystematic2019}.
+Some steps, such as the implementation of the PCA, were also not explored in detail here as they have already been the object of recent and thorough study elsewhere \cite{SunDimRed2019, TsuyuzakiPCA2020}.
+We also considered only methods relying on Euclidean distance, while correlation was recently reported to be superior\cite{kim_impact_2019} and would require further investigation.
+
+Here, we chose to concentrate on what could be learned from datasets with known cell labels (as opposed to labels inferred from the data, as in \cite{MereuCellAtlas2019}).
+In contrast to Tian et al.
+\cite{tianMixology2018}, who used RNA mixtures of known proportions, we chose to rely chiefly on real cells and their representative form of variability.
+Given the limited availability of such well-described datasets however, several aspects of single-cell analysis could not be compared, such as batch effect correction or multi-dataset integration.
+For these aspects of scRNAseq processing that are critical in some experimental designs, we refer the reader to previous evaluations \cite{TranBatch2020, StuartIntegration2019}.
+In addition, a focus on the identification of the subpopulations might fail to reveal methods that are instead best performing for tasks other than clustering, such as differential expression analysis or trajectory inference.
+Several informative benchmarks have already been performed on some of these topics \cite{CrowellMuscat2019, DalMolinDE2017, JaakkolaDE2017, SaelensTraject2019, SonesonDE2018, WangDE2019}.
+Yet, such evaluations could benefit from considering methods not in isolation, but as parts of a connected workflow, as we have done here.
+We believe that the \textit{pipeComp} framework is modular and flexible enough to integrate new steps in the pipeline, as shown by an example with imputation/denoising methods.
+
+We developed \texttt{pipeComp} to address the need of a framework that can simultaneously evaluate the interaction of multiple tools.
+The work of Vieth and colleagues\cite{viethSystematic2019} already offered an important precedent in this respect, evaluating the interaction of various steps in the context of scRNAseq differential expression analysis, but did not offer a platform for doing so.
+Instead, the \textit{CellBench} package was recently proposed to address a similar need \cite{suCellbench}, offering an elegant piping syntax to combine alternative methods at successive steps.
+A key additional feature of \textit{pipeComp} is its ability to perform evaluation in parallel to the pipeline and offering on-the-fly evaluation at any step.
+This avoid the need to store potentially large intermediate data for all possible permutations and thus allowing a combinatorial benchmark.
+In addition, the fact that benchmark functions are stored in the \texttt{PipelineDefinition} makes the benchmark more smoothly portable, making it easy to modify, extend with further methods, or apply to other datasets.
+
+With respect to the methods themselves, we believe there is still space for improvement in some of the steps.
+Concerning cell filtering, we noticed that the current approach based on whole-population characteristic (e.g.
+MAD cut-off) can be biased against certain subpopulations, suggesting that more refined methods expecting multimodal distributions should be used rather than relying on whole-population characteristics.
+In addition, most common filtering approaches do not harness the relationship between cell QC properties.
+Finally, imputation had a varied impact on the clustering analysis and seemed to be linked to the technology that was used to generate the data.
+Also, the good performance of \textit{DrImpute} is in line with a previous study focused on the preservation of data structure in trajectory inference from scRNAseq \cite{ZhangImput2018} (\textit{alra} was however not included in this previous benchmark).
+Our results also align with the hypothesis of this study on the respective strengths of linear and non-linear models and their use to different types of data, such as the highest performance of non-linear methods in developmental studies.
+Nevertheless, the other steps of preprocessing could already reach good performances when using the right tools, which is especially the case for normalization.
+
+\section*{Conclusion}
+
+\textit{pipeComp} is a flexible R framework for evaluating methods and parameter combinations in a complex pipeline, computing on-the-fly multilevel evaluation metrics.
+Applying this framework to scRNAseq clustering enabled us to make concrete recommendations on the steps of filtering, normalization, feature selection, denoising, dimensionality reduction and clustering.
+We demonstrate how a diversity of multilevel metrics can be more robust, more sensitive, and more nuanced than simply evaluating the final clustering performance.
+In addition, we provide a new and efficient Bioconductor package for doublet detection, \texttt{scDblFinder}.
+We hope that the \textit{pipeComp} framework can be applied to extend the current field of benchmarking, as well as to apply it to other range of methods.
+
+\newpage
+
+\section*{Methods}
+
+\subsection*{Code and data availability}
+
+All analyses were performed through the \href{https://github.com/plger/pipeComp}{pipeComp} R package, which implements the pipeline framework described here.
+All code to reproduce the simulations and figures is available in the \url{https://github.com/markrobinsonuzh/scRNA\_pipelines\_paper} repository, which also includes the basic datasets with a standardized annotation.
+
+The gene counts for the two mixology datasets were downloaded from the  \href{https://github.com/LuyiTian/CellBench_data/data/sincell_with_class.RData}{CellBench repository}, commit 74fe79e.
+For the other datasets, we used the unfiltered counts from \cite{duoClustering2018}, available on the corresponding repository \url{https://github.com/markrobinsonuzh/scRNAseq\_clustering\_comparison}.
+For simplicity, all starting \textit{SingleCellExperiment} objects with standardized metadata are available on \url{https://github.com/markrobinsonuzh/scRNA\_pipelines\_paper}.
+
+\subsection*{Software and package versions}
+Analyses were performed in R 3.6.0 (Bioconductor 3.9) and the following packages were installed from GitHub repositories: \textit{Seurat} (version 3.0.0), \textit{sctransform} (0.2.0), \textit{DoubletFinder} (2.0.1), \textit{scDblFinder} (1.1.1), \textit{scds} (1.0.0), \textit{SAVERX} (1.0.0), \textit{scImpute} (0.0.9), \textit{ALRA} (commit 7636de8), \textit{DCA} (0.2.2), \textit{DrImpute} (1.2), \textit{ENHANCE} (R version, commit 1571696), \textit{SAVERX} (1.0.0).
+The code for the glmPCA was obtained from \url{https://github.com/willtownes/scrna2019} (commit 1ddcc30ebb95d083a685f12fe81d35dd1b0cb1b2).
+
+\subsection*{Simulated datasets}
+The simMix1 dataset was based on the human PBMC CITE-seq data deposited under accession code GSE100866 of the Gene Expression Omnibus (GEO) website.
+Both RNA and ADT count data were downloaded from GEO and processed independently using \textit{Seurat}.
+We then considered cells that were in the same cluster both in the RNA-based and ADT-based analyses to be real subpopulations and focused on the 4 most abundant ones.
+We then performed 3 sampling-based simulations with various degrees of separation using \textit{muscat} \cite{CrowellMuscat2019} and merged the three simulations.
+The simMix2 dataset was generated from the mouse brain data published in \cite{CrowellMuscat2019} in a similar fashion.
+The specific code is available on \url{https://github.com/markrobinsonuzh/scRNA\_pipelines\_paper}.
+
+\subsection*{Default pipeline parameters}
+
+Where unspecified, the following default parameters or parameter sets were used:
+\begin{itemize}
+    \item \textit{scDblFinder} was used for doublet identification
+    \item the default filter sets (see below) were applied
+    \item standard \textit{Seurat} normalization was employed
+    \item \textit{Seurat} ($\geq$3.0) variable feature selection was employed, selecting 2000 genes
+    \item standard \textit{Seurat} scaling and PCA were employed
+    \item To study the impact of upstream steps on clustering, various \textit{Seurat} clustering analyses were performed, selecting different numbers of dimensions (5, 10, 15, 20, 30, and 50) and using various resolution parameters (0.005, 0.01, 0.02, 0.05, 0.1, 0.15, 0.2, 0.3, 0.4, 0.5, 0.8, 1, 1.2, 1.5, 2, 4).
+The range of resolution parameters was selected to ensure that the right number of clusters could be obtained in all datasets.
+\end{itemize}
+
+\subsection*{Denoising/imputation}
+Most imputation/denoising methods were run with the default parameters with the following exceptions; \textit{DrImpute} documentation advises to process the data prior to imputation by removing lowly expressed genes and cells expressing less than 2 genes.
+As it is not clear if this step is a hard requirement for the method to perform well, \textit{DrImpute} was run with and without prior processing (\textit{DrImpute\_process} and \textit{DrImpute\_noprocess} labels, respectively).
+The \textit{dca} method only accepts integers counts, while two of the datasets had non-integer quantification of expected counts.
+For these datasets, we rounded up the counts prior to imputation.
+\textit{alra} is designed for normalized data but as we are evaluating normalization downstream to imputation, we used the method on both non-normalized (\textit{alra} label) and normalized counts (\textit{alra\_norm} label).
+\textit{scImpute} requires an estimation of the expected number of clusters with the input data.
+As the estimation of the true number of cluster may not be known by the user, we evaluated the tool using the true number of clusters (\textit{scImpute} label) and using an over/under-estimation of this number (\textit{scImpute\_plus5} and \textit{scImpute\_min5} labels, respectively).
+\textit{ENHANCE} uses a k-nearest neighbor aggregation method and automatically estimates the number of neighbors to merge prior to the imputation.
+With the smallest datasets, this parameters was estimated to be 1, which lead to an early stop of the function.
+In such cases, we manually set this parameter to 2 for the method to work.
+
+\subsection*{Dimensionality estimates}
+
+For the methods that produce local dimensionality estimates, we used the maximum.
+For the elbow method, we implemented an automatic procedure by taking the farthest point from a line drawn between the variance explained by the first and last (i.e.
+50th) components calculated.
+For the \texttt{JackStraw} method, since the \textit{Seurat} documentation advises not to use a p-value threshold (which would typically yield a very large number of dimensions) but rather look for a drop in significance, we applied the same farthest point algorithm on the log10(p-values), which in our hands reproduced manual threshold selection.
+
+\subsection*{Doublet detection method}
+Our doublet detection method is available at \url{https://github.com/plger/scDblFinder}.
+Briefly, after reducing the data to the most expressed genes, we cluster the cells using the fast-greedy algorithm, favouring overclustering.
+We then create artificial doublets by sampling the two cells specifically from different clusters and sum the counts of each pair of cells.
+We also create meta-cells from each cluster and use them to create additional doublets and triplets.
+We combine them with the real cells, perform PCA and build a KNN graph using \textit{BiocNeighbors}.
+We then calculate, for each cell, the proportion of its neighbors that are artificial doublets, weighted by the distance.
+This ratio serves as a doublet score, which is then thresholded by simultaneously minimizing the error in classifying real vs artificial doublets and the deviation from the distribution of expected doublet rate (accounting for homotypic doublets as done by \textit{DoubletFinder}).
+
+\subsection*{Filter sets}
+The \textit{default} set of filters excludes cells that are outliers according to at least two of the following thresholds: log10\_total\_counts $>$2.5 MADs or $<$5 MADs, log10\_total\_features $>$2.5 MADs or $<$5 MADs, pct\_counts\_in\_top\_20\_features $>$ or $<$ 5 MADs, featcount\_dist (distance to expected ratio of log10 counts and features) $>$ or $<$ 5 MADs, pct\_counts\_Mt $>$ 2.5 MADs and $>$ 0.08.
+
+The \textit{stringent} set of filters uses the same thresholds, but excludes a cell if it is an outlier on any single distribution.
+
+The \textit{lenient} set of filters excludes cells that are outliers on at least two distributions by at least 5 MADs, except for pct\_counts\_Mt where the threshold is $>$ 3 MADs and $>$ 0.08.
+
+For cluster-wise filters, clusters were first identified with \textit{scran::quickCluster} and the filters were then applied separately for each cluster.
+
+The `pca' and `pca2' clusters refer to the multivariate outlier detection methods implemented in \textit{scater}, running \textit{runPCA} with \texttt{use\_coldata=TRUE, detect\_outliers=TRUE}.
+`pca` uses all covariates, while `pca2` uses only the log10(counts), log10(features), proportion mitochondrial and proportion in the top 50 features.
+
+\subsection*{Variance and deviance explained}
+
+Unless specified otherwise, the variance in gene expression explained by subpopulations was calculated on the data normalized and transformed through \textit{sctransform}.
+For each gene, we fitted a linear model using the subpopulation as only independent variable ($\sim subpopulation$) and used the R-squared as a measure of the variance explained.
+The same method was used for principal components.
+
+The deviance explained by subpopulations was calculated directly on counts using the \linebreak \texttt{getDevianceExplained} function of \texttt{pipeComp}.
+The function uses \textit{edgeR} to fit two models, a full ($\sim librarySize + population$) and a reduced one ($\sim librarySize$).
+For each gene, the deviance explained is then the difference between the deviance of each models, divided by the deviance of the reduced model.
+In the rare cases where this resulted in a negative deviance explained, it was set to zero.
+
+To estimate the correlation between the principal components and covariates such as library size, we first fitted a linear model on the subpopulations and correlated the residuals of this model with the covariate of interest.
+
+\section*{Tables}
+
+\begin{table}[h!]
+\caption{Overview of the benchmark datasets.}
+\label{tab:table1}
+\begin{tabular}{rlrrl}
+  \hline
+dataset & source & protocol & description \\ 
+  \hline
+Koh & GSE85066 & SMARTer & FACS purified H7 hESC in different differention stages \\ 
+  Kumar & GSE60749 & SMARTer & Mouse ESC cultured in different conditions \\ 
+  Zhengmix4eq & \cite{duoClustering2018} & 10x & Mixtures of FACS purified PBMCs \\ 
+  Zhengmix4uneq & \cite{duoClustering2018} & 10x & Mixtures of FACS purified PBMCs \\ 
+  Zhengmix8eq & \cite{duoClustering2018} & 10x & Mixtures of FACS purified PBMCs \\ 
+  mixology10x3cl & \cite{tianMixology2018} & 10x & Mixture of 3  cancer cell lines from CellBench \\ 
+  mixology10x5cl & \cite{tianMixology2018} & 10x & Mixture of 5 cancer cell lines from CellBench \\ 
+  simMix1 & - & 10x-based & Simulation of 10 human cell subpopulations \\
+  simMix2 & - & 10x-based & Simulation of 9 mouse cell subpopulations \\
+   \hline
+\end{tabular}
+\end{table}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                          %%
+%% Backmatter begins here                   %%
+%%                                          %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{backmatter}
+
+\section*{Competing interests}
+As developers of the \textit{scDblFinder} package, the authors have an interest in it performing well. The authors declare no further competing interests.
+
+\section*{Author's contributions}
+    Text for this section \ldots
+
+\section*{Acknowledgements}
+This work was supported by the Swiss National Science Foundation (grants 310030\_175841, CRSII5\_177208) as well as the Chan Zuckerberg Initiative DAF (grant number 2018-182828), an advised fund of Silicon Valley Community Foundation.
+MDR acknowledges support from the University Research Priority Program Evolution in Action at the University of Zurich.
+
+\end{backmatter}
 
@@
 
-\includepdf[pages=-]{supplementary_figures/supp_figures.pdf}
+\newpage
+\pagenumbering{arabic}
+
+% \includepdf[pages=-]{supplementary_figures/supp_figures.pdf}
 

### 2020-02-01 supplementary_figures/supp_figures.tex
added title and authors to supp

@@
 \hypersetup{unicode=true,
-            pdftitle={Supplementary Figures},
-            pdfauthor={Pierre-Luc Germain},
+            pdfauthor={Pierre-Luc Germain; Anthony Sonrel; Mark D. Robinson},
             pdfborder={0 0 0},
@@
 
-  \title{Supplementary Figures}
+  \title{\Large{pipeComp, a general framework for the evaluation of computational pipelines, reveals performant single-cell RNA-seq preprocessing tools}}
     \pretitle{\vspace{\droptitle}\centering\huge}
   \posttitle{\par}
-    \author{Pierre-Luc Germain}
+  \subtitle{\vspace{0.5cm}\LARGE{Supplementary Figures}\vspace{0.5cm}}
+  \author{Pierre-Luc Germain \\ Anthony Sonrel \\ Mark D. Robinson}
     \preauthor{\centering\large\emph}
@@
   \postdate{\par}
-    \date{30 Januar, 2020}
+    \date{01 Februar, 2020}
 
@@
 }
+\renewcommand{\thepage}{S\arabic{page}}
+
 \newpage

### 2020-02-01 bmc_article.tex

@@
 
-\title{A framework for pipeline benchmarking and its application to single-cell RNAseq clustering}
+\title{pipeComp, a general framework for the evaluation of computational pipelines, reveals performant single cell RNA-seq preprocessing tools}
 
@@
 On the basis of these analyses, we make a number of concrete recommendations about analysis choices.
-The evaluation framework, \textit{pipeComp}, has been implemented so as to easily integrate any other step or tool and permit extensible benchmarks (\url{https://github.com/plger/pipeComp}).
+The evaluation framework, \textit{pipeComp}, has been implemented so as to easily integrate any other step or tool, permit extensible benchmarks and easy application to other fields (\url{https://github.com/plger/pipeComp}).
 
@@
 
-While a number of good comparison and benchmark studies have already been performed on various steps related to scRNAseq analysis and can inform analytical decisions (\citealp{CobosDeconvolution2020, colePerformance2018, DalMolinDE2017, duoClustering2018, freytagComparison2018, HeiserDimred2019, JaakkolaDE2017, SonesonDE2018, SunDimRed2019, tianMixology2018, TranBatch2020, TsuyuzakiPCA2020, viethSystematic2019, WangDE2019, YipHVGs2018, ZhangImput2018}), they need constant updating and often leave open many details of an analysis.
+A number of good comparison and benchmark studies have already been performed on various steps related to scRNAseq analysis and can inform analytical decisions (\citealp{CobosDeconvolution2020, colePerformance2018, DalMolinDE2017, duoClustering2018, freytagComparison2018, HeiserDimred2019, JaakkolaDE2017, SonesonDE2018, SunDimRed2019, tianMixology2018, TranBatch2020, TsuyuzakiPCA2020, viethSystematic2019, WangDE2019, YipHVGs2018, ZhangImput2018}).
+However these recommendations need constant updating and often leave open many details of an analysis.
 Another missing aspect of current benchmarking studies is their limitation to capture all aspects of scRNAseq processing workflow.
-Although previous benchmarks already brought valuable recommendations for data processing, some only focused on one aspect of data processing (e.g., \citealp{SunDimRed2019}), did not evaluate the effect of the choice of a tool on downstream analysis (e.g., \citealp{TsuyuzakiPCA2020}) or did not tackle all aspects of data processing, such as doublet identification or cell filtering (e.g., \citealp{viethSystematic2019}).
+Although previous benchmarks already brought valuable recommendations for data processing, some only focused on one aspect of data processing (e.g., \citealp{SunDimRed2019}), did not evaluate how the tool selection affects downstream analysis (e.g., \citealp{TsuyuzakiPCA2020}) or did not tackle all aspects of data processing, such as doublet identification or cell filtering (e.g., \citealp{viethSystematic2019}).
 A thorough evaluation of the tools covering all major processing steps is however urgently needed as previous benchmarking studies highlighted that a combination of tools can have a drastic impact on downstream analysis, such as differential expression analysis and cell-type deconvolution (\citealp{viethSystematic2019, CobosDeconvolution2020}).
-It is then likely that processing algorithms also interact in a positive or negative fashion and it is then critical to evaluate not only the single effect of a method but also its interaction with all parts of a workflow.
+It is then critical to evaluate not only the single effect of a preprocessing method but also its positive or negative interaction with all parts of a workflow.
 
 Here, we develop a flexible R framework for pipeline comparison and evaluate the various steps of analysis leading from an initial count matrix to a cluster assignment, which are critical in a wide range of applications.
-We harnessed real datasets of known cell composition (Table \ref{tab:table1}) and developed new evaluation metrics to investigate in a multilevel fashion the impact of various parameters and variations around a core scRNAseq pipeline.
+We collected real datasets of known cell composition (Table \ref{tab:table1}) and used a variety of evaluation metrics to investigate in a multilevel fashion the impact of various parameters and variations around a core scRNAseq pipeline.
 Although we use some datasets based on other protocols, our focus is especially on droplet-based datasets that do not include exogenous control RNA (i.e.
-spike-ins); see Table \ref{tab:table1} and Figure \ref{fig:figure1} for a description of the datasets.
-In addition to previously-used benchmark datasets with true cell labels \cite{duoClustering2018,tianMixology2018}, we simulated two datasets with a hierarchical subpopulation structure based on real 10x human and mouse data using \textit{muscat} \cite{CrowellMuscat2019}).
+spike-ins); see Table \ref{tab:table1} and Figure \ref{fig:figure1} for more details.
+In addition to previously-used benchmark datasets with true cell labels \cite{duoClustering2018,tianMixology2018}, we simulated two datasets with a hierarchical subpopulation structure based on real 10x human and mouse data using \textit{muscat} \cite{CrowellMuscat2019}.
 Since graph-based clustering \cite{satijaSeurat2015} was previously shown to consistently perform well across several datasets \cite{duoClustering2018,tianMixology2018}, we used the \textit{Seurat} pipeline as the starting point to perform an integrated investigation of: 1) doublet identification, 2) cell filtering, 3) normalization, 4) feature selection, 5) dimension reduction, 6) clustering.
-We compare not only competing approaches, but also more fine-grained parameters and variations on common methods.
+We compared competing approaches and also explored more fine-grained parameters and variations on common methods.
 Importantly, the success of methods at a certain analytical step might be dependent on choices at other steps.
@@
 Given such a \texttt{PipelineDefinition} object, a set of alternative parameters (which might include different subroutines) and benchmark datasets, the \texttt{runPipeline} function then proceeds through all combinations of arguments, avoiding recomputing the same step twice and compiling evaluations (including running time) on the fly.
-The \textit{CellBench} package was recently proposed to address a similar need\citep{su_cellbench}, offering an elegant piping syntax to combine alternative methods at successive steps.
-In addition to offering the same capabilities, a key additional feature of \textit{pipeComp} is that, since the evaluation metrics are not treated as a step but parallel to it, they can be performed at various points in the pipeline without needing to store the potentially large intermediate data, and thereby enabling combinatorial benchmarks.
 Variations in a given parameter can be evaluated using all metrics from this point downward in the pipeline.
@@
 Clustering results are also very sensitive, and might not always capture improvements in earlier steps of the pipeline.
+We therefore favour monitoring several complementary metrics across multiple steps of the process.
 
-Another motivation for the present framework is that clustering is relatively fragile to variations in the pipeline.
-We therefore also wanted to capture whether the effect of a given parameter alteration is robust to changes in the rest of the pipeline, or rather specific to a set of other pipeline parameters.
-We therefore proceeded in a step-wise fashion, first testing a large variety of parameters at the early steps of the pipeline along with only a set of mainstream options downstream, then selecting main alternatives and proceeding to a more detailed benchmark of the next step (Figure \ref{fig:figure2}B).
+A clustering output is relatively fragile to variations in the pipeline.
+We therefore wanted to capture whether the effect of a given parameter alteration is robust to changes in the rest of the pipeline, or rather specific to a set of other pipeline parameters.
+We proceeded in a step-wise fashion, first testing a large variety of parameters at the early steps of the pipeline along with only a set of mainstream options downstream, then selecting main alternatives and proceeding to a more detailed benchmark of the next step (Figure \ref{fig:figure2}B).
 
@@
     \includegraphics[width=\textwidth,keepaspectratio]{main_figures/pipeline_explanation.png}
-    \caption{\textbf{Overview of the pipeComp framework and its application to a scRNAseq clustering pipeline.}}
+    \caption{\textbf{Overview of the \textit{pipeComp} framework and its application to a scRNAseq clustering pipeline.
+A:} The package is built around a \texttt{PipelineDefinition} S4 class which defines a set of functions to be executed consecutively, as well as optional evaluation functions for each step.
+Each step can accept of number of parameters whose alternative values are provided as a list.
+All (or subsets of) combinations of parameters can then be simultaneously ran and evaluated using the \texttt{runPipeline} function.
+\textbf{B:} Scheme representing the application of \textit{pipeComp} to evaluate a range of methods that are commonly used in scRNAseq studies, and some of the metrics monitored at various steps.}
     \label{fig:figure2}
@@
 A number of methods were developed to identify doublets, in most cases by comparing each cell to artificially-created doublets \citep{mcginnisDoubletfinder2019, LunScran2016, BaisScds2019}.
-We therefore first evaluated the capacity of these methods to detect doublets using the two 10x datasets with cells of different genetic identity \citep{tianMixology2018}, using SNP genotypes as the ground truth.
+We first evaluated the capacity of these methods to detect doublets using the two 10x datasets with cells of different genetic identity \citep{tianMixology2018}, using SNP genotypes as the ground truth.
 For the sole purpose of this section, we included an additional dataset with SNP information but lacking true cell labels \citep{kangMultiplexedDemuxlet2018}.
@@
 While most methods accurately identified the doublets in the 3 cell lines dataset (mixology10x3cl), the other two datasets proved more difficult (Figure \ref{fig:figure3}A).
-\textit{scDblFinder} achieved comparable or better accuracy than top alternatives while running in a very reasonable time (Figure \ref{fig:figure3}B).
-Across datasets, cells called as doublets tended to be classified in the wrong cluster more often than other cells (Figure \ref{fig:figure3}C.
+\textit{scDblFinder} achieved comparable or better accuracy than top alternatives while being the fastest method (Figure \ref{fig:figure3}B).
+Across datasets, cells called as doublets tended to be classified in the wrong cluster more often than other cells (Figure \ref{fig:figure3}C).
 We therefore tested whether this method also improved the accuracy of the clustering across all benchmark datasets, and found that it did even when, by design, the data contained no heterotypic doublet (Figure \ref{fig:figure4}).
@@
 Of note, these properties tend to be correlated, with some exceptions.
-For instance, while a high proportion of mitochondrial reads is often correlated with a high proportion of the counts in the top features, there can also be other reasons for an over-representation of highly-expressed features (Supplementary Figure 3), such as an over-amplification in non-UMI datasets.
+For example, while a high proportion of mitochondrial reads is often correlated with a high proportion of the counts in the top features, there can also be other reasons for an over-representation of highly-expressed features (Supplementary Figure 3), such as an over-amplification in non-UMI datasets.
 In our experience, 10X datasets also exhibit a very strong correlation between the total counts and the total features even across very different cell types (Supplementary Figure 4).
@@
 \textit{Seurat}'s earlier approaches involved the use of dispersion estimates standardized for the mean expression levels, while more recent versions ($\geq$3.0) rely on a different measure of variance, again standardized.
-Of note, while adjusting for the mean-variance relationship removes much of the bias towards highly-expressed genes, it is plausible that this relationship may in fact sometimes reflects biological relevance and would be helpful in classifying cell types.
-Another common practice in feature selection is to use those with the highest mean expression, and recently, \citep{townesGlmpca2019} instead suggested to use deviance, while \textit{sctransform} provides its own ordering of genes based on transformed variance.
+While adjusting for the mean-variance relationship removes much of the bias towards highly-expressed genes, it is plausible that this relationship may in fact sometimes reflects biological relevance and would be helpful in classifying cell types.
+Another common practice in feature selection is to use those with the highest mean expression.
+Recently, \citep{townesGlmpca2019} instead suggested to use deviance, while \textit{sctransform} provides its own ordering of genes based on transformed variance.
 
 Reasoning that a selection method should ideally select genes whose variability is higher between subpopulations than within, we first assessed to what extent each method selected genes with a high proportion of variance or deviance explained by (real) subpopulation.
-As the proportion of variability in a gene attributable to subpopulations can be measured in various ways, we first compared three such alternatives.
+As the proportion of variability in a gene attributable to subpopulations can be measured in various ways, we first compared three of them.
 Analysis of variance performed on a standard \textit{Seurat} normalization or on \textit{sctransform} data were highly correlated (Supplementary Figure 11A).
@@
 
-We next investigated the methods of dimensionality reduction.
 Since the various PCA approaches and implementations were recently benchmarked in a similar context \citep{TsuyuzakiPCA2020}, we focused on widely used approaches which had not yet been compared: \textit{Seurat}'s PCA, \textit{scran}'s \texttt{denoisePCA}, and GLM-PCA \citep{townesGlmpca2019}, and combined them (when relevant) with \textit{sctransform} normalization.
@@
 Although this platform has been used in almost half of the scRNAseq studies in 2019 \citep{SvenssonDB2019}, other popular technologies such as Drop-seq, InDrops or Smart-seq2/3 were not represented in the present benchmarking.
-Differences between such protocols can have a very large impact on clustering \citep{MereuCellAtlas2019}.
+Differences between such protocols, even among droplet-based technologies, can have a very large impact on cell capture efficiency, cell numbers and clustering \citealp{MereuCellAtlas2019, ZhangDroplet2019, SalomonDroplet2019}.
 Although most top-ranking methods in our comparison performed well on both Smart-seq and 10x datasets that we tested, future benchmarking efforts should strive to include less represented technologies.
@@
 In contrast to Tian et al.
-\citep{tianMixology2018}, who used on RNA mixtures of known proportions, we chose to rely chiefly on real cells and their representative form of variability.
-Given the limited availability of such datasets however, several aspects of single-cell analysis could not be compared, such as batch effect correction or multi-dataset integration.
+\citep{tianMixology2018}, who used RNA mixtures of known proportions, we chose to rely chiefly on real cells and their representative form of variability.
+Given the limited availability of such well-described datasets however, several aspects of single-cell analysis could not be compared, such as batch effect correction or multi-dataset integration.
 For these aspects of scRNAseq processing that are critical in some experimental designs, we refer the reader to previous evaluations \citealp{TranBatch2020, StuartIntegration2019}.
@@
 We believe that the \textit{pipeComp} framework is modular and flexible enough to integrate new steps in the pipeline, as shown by an example with imputation/denoising methods.
+
+We developed \ŧexttt{pipeComp} to address the need of a framework that can evaluate multiple tools simultaneously.
+The \textit{CellBench} package was recently proposed to address a similar need\citep{su_cellbench}, offering an elegant piping syntax to combine alternative methods at successive steps.
+A key additional feature of \textit{pipeComp} is its ability to perform evaluation in parallel to the pipeline and offering on-the-fly evaluation at any step.
+This avoid the need to store potentially large intermediate data for all possible permutations and thus allowing a combinatorial benchmark.
+In addition, the fact that benchmark functions are stored in the \texttt{PipelineDefinition} makes the benchmark more smoothly portable, enabling it transportable easy to apply comparable contains all information regarding the benchmark, including
 
@@
 \section*{Competing interests}
-  The authors declare that they have no competing interests.
+As developers of the \textit{scDblFinder} package, the authors have an interest in it performing well. The authors declare no further competing interests.
 
@@
 \begin{table}[h!]
-\caption{Overview of the benchmark datasets}
+\caption{Overview of the benchmark datasets.}
 \label{tab:table1}

### 2020-02-01 draft.tex

@@
 
-\author[1,2,3]{Pierre-Luc Germain}
+\author[1,2,3,*]{Pierre-Luc Germain}
 \author[1,2]{Anthony Sonrel}
-\author[1,2]{Mark D. Robinson}
+\author[1,2,*]{Mark D. Robinson}
 
@@
 \affil[3]{D-HEST Institute for Neuroscience, Swiss Federal Institute of Technology (ETH), Z{\"u}rich, Switzerland}
+\affil[*]{Correspondence to Pierre-Luc Germain (pierre-luc.germain@hest.ethz.ch) and Mark D. Robinson (mark.robinson@imls.uzh.ch)}
 
@@
 
-A clustering output is relatively fragile to variations in the pipeline.
-We therefore wanted to capture whether the effect of a given parameter alteration is robust to changes in the rest of the pipeline, or rather specific to a set of other pipeline parameters.
-We proceeded in a step-wise fashion, first testing a large variety of parameters at the early steps of the pipeline along with only a set of mainstream options downstream, then selecting main alternatives and proceeding to a more detailed benchmark of the next step (Figure \ref{fig:figure2}B).
-
 \begin{figure}
@@
 \end{figure}
+
+A clustering output is relatively fragile to variations in the pipeline.
+We therefore wanted to capture whether the effect of a given parameter alteration is robust to changes in the rest of the pipeline, or rather specific to a set of other pipeline parameters.
+We proceeded in a step-wise fashion, first testing a large variety of parameters at the early steps of the pipeline along with only a set of mainstream options downstream, then selecting main alternatives and proceeding to a more detailed benchmark of the next step (Figure \ref{fig:figure2}B).
 
@@
 doublets from different cell types).
-These homotypic doublets might not be identifiable from the mere gene counts, and their identification is not generally the primary aim of doublet callers since they are often considered innocuous and, when across individuals, can be identified through other means (e.g.
-SNPs).
+These homotypic doublets might not be identifiable from the mere gene counts, and their identification is not generally the primary aim of doublet callers since they are often considered innocuous and, when across individuals, can be identified through other means (e.g., SNPs).
 We therefore do not expect a perfect accuracy in datasets involving cells of the same type across individuals (as in the demuxlet dataset).
@@
 
-While most methods accurately identified the doublets in the 3 cell lines dataset (mixology10x3cl), the other two datasets proved more difficult (Figure \ref{fig:figure3}A).
-\textit{scDblFinder} achieved comparable or better accuracy than top alternatives while being the fastest method (Figure \ref{fig:figure3}B).
-Across datasets, cells called as doublets tended to be classified in the wrong cluster more often than other cells (Figure \ref{fig:figure3}C).
-We also found that \textit{scDblFinder} improved the accuracy of the clustering across all benchmark datasets even when, by design, the data contained no heterotypic doublet (Figure \ref{fig:figure4}).
-
-\begin{figure}
+\begin{figure}[h]
     \centering
-    \includegraphics[width=\textwidth]{{main_figures/doublets_files/figure-html/figure1-1.png}}
+    \includegraphics[width=0.9\textwidth]{{main_figures/doublets_files/figure-html/figure1-1.png}}
     \caption{\textbf{Identification of doublet cells.
@@
 \end{figure}
+
+While most methods accurately identified the doublets in the 3 cell lines dataset (mixology10x3cl), the other two datasets proved more difficult (Figure \ref{fig:figure3}A).
+\textit{scDblFinder} achieved comparable or better accuracy than top alternatives while being the fastest method (Figure \ref{fig:figure3}B).
+Across datasets, cells called as doublets tended to be classified in the wrong cluster more often than other cells (Figure \ref{fig:figure3}C).
+We also found that \textit{scDblFinder} improved the accuracy of the clustering across all benchmark datasets even when, by design, the data contained no heterotypic doublet (Figure \ref{fig:figure4}).
 
@@
 
+\begin{figure}
+    \centering
+    \includegraphics[width=\textwidth]{{main_figures/filtering_files/figure-html/figure2-1.png}}
+    \caption{\textbf{A:} Filtering on the basis of distance to the whole distribution can lead to strong bias against certain subpopulations.
+The dashed line indicates a threshold of 2.5 median absolute deviations (MADs) from the median of the overall population.
+\textbf{B:} Relationship between the maximum subpopulation exclusion rate and the average clustering accuracy per subpopulation across various filtering strategies.
+Of note, doublet removal appears to be desirable even when, due to the design, there are no heterotypic doublets in the data.
+The PCA methods refer to multivariate outlier detected as implemented in \textit{scater} (see Methods for details).}
+    \label{fig:figure4}
+\end{figure}
+
 We next investigated the impact of filtering according to various criteria (see Methods).
-An important risk of excluding cells on the basis of their distance from the whole distribution of cells on some properties (e.g.
-library size) is that these properties tend to have different distributions across subpopulations.
+An important risk of excluding cells on the basis of their distance from the whole distribution of cells on some properties (e.g., library size) is that these properties tend to have different distributions across subpopulations.
 As a result, thresholds in terms of number of MADs from the whole distribution can lead to strong biases against certain subpopulations (Figure \ref{fig:figure4}A).
@@
 We therefore recommend the use of doublet removal followed by relatively mild filtering, such as is implemented in our `default' filtering (see Methods).
-
-\begin{figure}
-    \centering
-    \includegraphics[width=\textwidth]{{main_figures/filtering_files/figure-html/figure2-1.png}}
-    \caption{\textbf{A:} Filtering on the basis of distance to the whole distribution can lead to strong bias against certain subpopulations.
-The dashed line indicates a threshold of 2.5 median absolute deviations (MADs) from the median of the overall population.
-\textbf{B:} Relationship between the maximum subpopulation exclusion rate and the average clustering accuracy per subpopulation across various filtering strategies.
-Of note, doublet removal appears to be desirable even when, due to the design, there are no heterotypic doublets in the data.
-The PCA methods refer to multivariate outlier detected as implemented in \textit{scater} (see Methods for details).}
-    \label{fig:figure4}
-\end{figure}
 
@@
 In addition to log-normalization, the standard \textit{Seurat} clustering pipeline performs per-feature unit-variance scaling so that the PCA is not too strongly dominated by highly-expressed features.
-We therefore included versions of the different normalization procedures, with or without a subsequent scaling (\textit{sctransform}'s variance-stabilizing transformation involves a transformation analogous to scaling).
+We therefore included versions of the different normalization procedures, with or without a subsequent scaling (\textit{sctransform}'s variance-stabilizing transformation involves an approach analogous to scaling).
 \textit{Seurat}'s scaling function also includes the option to regress out the effect of certain covariates.
@@
 Finally, we monitored whether, under the same downstream clustering analysis, different normalization methods tended to lead to an over- or under-estimation of the number of clusters.
-Although some methods had a tendency to lead to a higher (e.g.
-\textit{sctransform}) or lower (e.g.
-stable genes) number of clusters, the effect was very mild and not entirely systematic (Supplementary Figure 10).
+Although some methods had a tendency to lead to a higher (e.g., \textit{sctransform}) or lower (e.g., stable genes) number of clusters, the effect was very mild and not entirely systematic (Supplementary Figure 10).
 
@@
     \centering
-    \includegraphics[scale=0.6]{{main_figures/selection_files/figure-html/figure4-1.png}}
+    \includegraphics[scale=0.75]{{main_figures/selection_files/figure-html/figure4-1.png}}
     \caption{\textbf{Evaluation of feature selection methods.
@@
 
-\begin{figure}
+\begin{figure}[h]
     \centering
@@
 
-The impact of the choice of dimensionality reduction method was far greater than that of normalization or feature selection (e.g.
-Supplementary Figure 15).
+The impact of the choice of dimensionality reduction method was far greater than that of normalization or feature selection (e.g., Supplementary Figure 15).
 GLM-PCA tended to increase the average silhouette width of already well-defined subpopulations, but \textit{Seurat}'s PCA procedure however proved superior on all metrics (Figure \ref{fig:figure7}).
@@
 Finally, the ARI score at the true number of clusters, when available, showed similar performances.
-However, because \textit{Seurat}'s resolution parameter had a large impact on the number of clusters identified, \textit{Seurat} could always be tweaked into producing the right number of clusters.
+However, because \textit{Seurat}'s resolution parameter had a large impact on the number of clusters identified (Supplementary Figure 18), \textit{Seurat} could always be coerced into producing the right number of clusters.
 Instead, the number of clusters found by \textit{scran} was considerably less influenced by the available parameters (number of nearest neighbors or steps in the random walk - see Supplementary Figure 18), and as a result \textit{scran}-based clustering sometimes never produced a partitioning with the right number of clusters.
-This observation, along with \textit{scran}'s higher mutual information, suggest that \textit{scran} sometimes simply divides a real subpopulation into two clusters (possibly tracking some unknown biological differences) rather than committing misclassification errors.
+This observation, along with \textit{scran}'s higher MI score, suggest that \textit{scran} sometimes simply divides a real subpopulation into two clusters (possibly tracking some unknown biological differences) rather than committing misclassification errors.
 Overall, the \textit{walktrap} method appeared superior to the \textit{fast greedy} algorithm and was generally less prone to misclassification than \textit{Seurat} clustering, although the latter offered more control over the resolution.
@@
 
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=\textwidth]{{main_figures/imputation_files/figure-html/imputation-1.png}}
+    \caption{\textbf{Imputation/denoising methods} Average silhouette width per subpopulation (\textbf{A}) and clustering accuracy (\textbf{B}) with or without (indicated as \textit{none}) application of a denoising/imputation method.}
+    \label{fig:figure10}
+\end{figure}
+
 The basic pipeline presented here can be extended with additional analysis steps while keeping the same evaluation metrics.
@@
 Since preliminary analysis showed that all methods performed equally well or better on normalized data, we applied them after filtering and normalization, but before scaling and reduction.
-Although some of the methods (e.g.
-\textit{DRImpute\_process} and \textit{alra\_norm}) did improve the separability of some more elusive subpopulations, no method had a systematically positive impact on the average silhouette width across all subpopulations (Figure \ref{fig:figure10}A).
+Although some of the methods (e.g., \textit{DRImpute\_process} and \textit{alra\_norm}) did improve the separability of some more elusive subpopulations, no method had a systematically positive impact on the average silhouette width across all subpopulations (Figure \ref{fig:figure10}A).
 When restricting ourselves to clustering analyses that yielded the `right' number of clusters, all tested methods improved classification  compared to a scenario with no imputation step (`none' label, Figure \ref{fig:figure10}B).
-However, the situation was not so straightforward with alternative metrics, where some methods (e.g.
-\textit{enhance}) consistently underperformed.
+However, the situation was not so straightforward with alternative metrics, where some methods (e.g., \textit{enhance}) consistently underperformed.
 10X datasets, which are typically characterized by a lower per-cell coverage and feature detection rate, benefited more from imputation and, in this context, \textit{DrImpute} and \textit{dca} tended to show the best performance.
 On the contrary, imputing on normalized counts originating from Smart-seq technology was instead rather deleterious to the clustering accuracy.
-
-\begin{figure}
-    \centering
-    \includegraphics[width=\textwidth]{{main_figures/imputation_files/figure-html/imputation-1.png}}
-    \caption{\textbf{Imputation/denoising methods} Average silhouette width per subpopulation (\textbf{A}) and clustering accuracy (\textbf{B}) with or without (\textit{none}) application of a denoising/imputation method}
-    \label{fig:figure10}
-\end{figure}
 
@@
 
-On the basis of our findings, we can make a number of concrete analysis recommendations:
+On the basis of our findings, we can make a number of concrete analysis recommendations, also summarized in Figure \ref{fig:figure11}:
 
@@
      \item Doublet detection and removal is advised and can be performed at little computing cost with software such as \textit{scDblFinder} or \textit{scds}.
-     \item Distribution-based cell filtering fails to capture doublets and should use relatively lenient cutoffs (e.g.
-5 MADs, or 3 MADs in at least 2 distributions) to exclude poor-quality cells while avoiding bias against some subpopulations.
+     \item Distribution-based cell filtering fails to capture doublets and should use relatively lenient cutoffs (e.g., 5 MADs, or 3 MADs in at least 2 distributions) to exclude poor-quality cells while avoiding bias against some subpopulations.
      \item Features filtering based on feature type did not appear beneficial.
@@
      \item Similarly to previous reports \cite{SunDimRed2019}, we recommend the irlba-based PCA using weighting of the components, as implemented in \textit{Seurat}. 
-     \item Instead of the common elbow or jackstraw methods for deciding on the components to include, we recommend the \textit{global maximum likelihood based on translated Poisson Mixture Model} method (e.g.
-implemented in the  \texttt{maxLikGlobalDimEst} function of \texttt{intrinsicDimension}).
+     \item Instead of the common elbow or jackstraw methods for deciding on the components to include, we recommend the \textit{global maximum likelihood based on translated Poisson Mixture Model} method (e.g., implemented in the  \texttt{maxLikGlobalDimEst} function of \texttt{intrinsicDimension}).
    \end{itemize}
@@
 \end{enumerate}
+
+\begin{figure}[h]
+    \centering
+    \includegraphics[scale=0.75,keepaspectratio]{main_figures/pipeComp_summary.png}
+    \caption{\textbf{Recommendations of tools from filtering, to clustering.}}
+    \label{fig:figure11}
+\end{figure}
 
@@
 With respect to the methods themselves, we believe there is still space for improvement in some of the steps.
-Concerning cell filtering, we noticed that the current approach based on whole-population characteristic (e.g.
-MAD cut-off) can be biased against certain subpopulations, suggesting that more refined methods expecting multimodal distributions should be used rather than relying on whole-population characteristics.
+Concerning cell filtering, we noticed that the current approach based on whole-population characteristic (e.g., MAD cut-off) can be biased against certain subpopulations, suggesting that more refined methods expecting multimodal distributions should be used rather than relying on whole-population characteristics.
 In addition, most common filtering approaches do not harness the relationship between cell QC properties.
@@
 Our results also align with the hypothesis of this study on the respective strengths of linear and non-linear models and their use to different types of data, such as the highest performance of non-linear methods in developmental studies.
-Nevertheless, the other steps of preprocessing could already reach good performances when using the right tools, which is especially the case for normalization.
 
@@
 
-\section*{Author's contributions}
-    Text for this section \ldots
+% \section*{Author's contributions}
+%    Text for this section \ldots
 

### 2020-02-02 simulations/prepare_citeseq.Rmd
added code for simulations

---
title: "Prepare citeseq data for simulation"
author: "Pierre-Luc Germain"
date: "10/27/2019"
output: html_document
---

```{r}
suppressPackageStartupMessages({
  library(Seurat)
  library(SingleCellExperiment)
  library(scater)
  library(muscat)
})
source("pipeComp/inst/extdata/scrna_alternatives.R")
```

```{r}
download.file("ftp://ftp.ncbi.nlm.nih.gov/geo/series/GSE100nnn/GSE100866/suppl/GSE100866_CBMC_8K_13AB_10X-RNA_umi.csv.gz","RNA.csv.gz")
cbmc.rna <- as.sparse(read.csv(file = "RNA.csv.gz", sep = ",", header = TRUE, row.names = 1))
cbmc.rna <- cbmc.rna[-grep("ERCC_|MOUSE_",row.names(cbmc.rna)),]
sce <- SingleCellExperiment(list(counts=cbmc.rna))
sce <- addQCPerCell(sce)
out <- isOutlier(sce$detected, log=TRUE) | isOutlier(sce$sum, log=TRUE) | isOutlier(sce$percent_top_50)
sce <- sce[,!out]
cbmc.rna <- cbmc.rna[,colnames(sce)]

download.file("ftp://ftp.ncbi.nlm.nih.gov/geo/series/GSE100nnn/GSE100866/suppl/GSE100866_CBMC_8K_13AB_10X-ADT_umi.csv.gz", "ADT.csv.gz")
cbmc.adt <- as.sparse(read.csv(file = "ADT.csv.gz", sep = ",", 
    header = TRUE, row.names = 1))
cbmc.adt <- cbmc.adt[,colnames(sce)]
```

Then we more or less follow the Seurat procedure:

```{r}
cbmc <- CreateSeuratObject(counts = cbmc.rna)
cbmc <- NormalizeData(cbmc)
cbmc <- FindVariableFeatures(cbmc, nfeatures=2000)
cbmc <- ScaleData(cbmc)
cbmc <- RunPCA(cbmc, verbose = FALSE)
ElbowPlot(cbmc, ndims = 50)
cbmc <- FindNeighbors(cbmc, dims = 1:9)
cbmc <- FindClusters(cbmc, resolution = 0.8)
cbmc[["rnaClusterID"]] <- Idents(cbmc)
```

```{r}
cbmc[["ADT"]] <- CreateAssayObject(counts = cbmc.adt)
cbmc <- NormalizeData(cbmc, assay = "ADT", normalization.method = "CLR")
cbmc <- ScaleData(cbmc, assay = "ADT")
DefaultAssay(cbmc) <- "ADT"
adt.dist <- dist(t(GetAssayData(cbmc, slot = "data")))
cbmc[["adt_snn"]] <- FindNeighbors(adt.dist)$snn
cbmc <- FindClusters(cbmc, resolution = 0.2, graph.name = "adt_snn")
(tt <- table(Idents(cbmc), cbmc$rnaClusterID))
```

We then select cells of the table (i.e. cells clustered together on both types of data) for which there are enough cells:
```{r}
t2 <- t(apply(tt, 1, FUN=function(x) c(which.max(x)-1, max(x))))
t2 <- t2[order(t2[,2],decreasing=TRUE),]
t2 <- t2[!duplicated(t2[,1]) & t2[,2] > 200,]
to_keep <- paste(row.names(t2),t2[,1])
sce$cluster_id <- factor(paste(Idents(cbmc),cbmc$rnaClusterID))
sce2 <- sce[,sce$cluster_id %in% to_keep]
sce2$cluster_id <- droplevels(sce2$cluster_id)
levels(sce2$cluster_id) <- LETTERS[seq_along(levels(sce2$cluster_id))]
```

We prepare for simulation:
```{r}
sce2$sample_id <- factor(sample(LETTERS[1:2],ncol(sce2),replace=TRUE))
sce2 <- prepSim(sce2)
saveRDS(sce2, file="citeseq_prep4sim.SCE.rds")
```

### 2020-02-02 simulations/simulation.Rmd
added code for simulations

---
title: "simulation"
author: "Pierre-Luc Germain"
date: "9/30/2019"
output: html_document
---

```{r}
suppressPackageStartupMessages({
  library(SingleCellExperiment)
  library(muscat)
  library(purrr)
  library(scater)
  library(cowplot)
})
source("../../pipComp/R/misc.R")
source("../../pipComp/R/filtering.R")
source("../../pipComp/R/selection.R")
theme_set(theme_cowplot())
```

# Simulating dataset 1

```{r}
# see citeseq.Rmd
sce <- readRDS("citeseq_prep4sim.SCE.rds")
sce$sample_id <- factor("A")
```

```{r}
sims <- list()
set.seed(1234)
probs <- list( set_names(c(0.8,0.1,0.1,0,0),levels(sce$cluster_id)), 1, c(1,0))
lfcs <- set_names(c(2,3,2,1,1),levels(sce$cluster_id))
sims$s1 <- simData(sce, nrow(sce), 1000, probs=probs, p_type=0.2, rel_lfc=lfcs)
probs <- list( set_names(c(0.1,0.6,0,0.3,0), levels(sce$cluster_id)), 1, c(1,0))
lfcs <- set_names(c(2,1,1,1,1), levels(sce$cluster_id))
sims$s2 <- simData(sce, nrow(sce), 1000, probs=probs, p_type=0.05, rel_lfc=lfcs)
probs <- list( set_names(c(0.3,0.3,0.1,0,0.3), levels(sce$cluster_id)), 1, c(1,0))
lfcs <- set_names(rep(1,5), levels(sce$cluster_id))
sims$s3 <- simData(sce, nrow(sce), 500, probs=probs, p_type=0.01, rel_lfc=lfcs)

sim <- SingleCellExperiment(list(counts=do.call(cbind, lapply(sims,counts))))
sim$phenoid <- paste0( rep(LETTERS[1:3],sapply(sims,ncol)),
                       unlist(lapply(sims,FUN=function(x){
  levels(x$cluster_id) <- as.character(seq_along(levels(x$cluster_id)))
  x$cluster_id
})))
colnames(sim) <- paste0(sim$phenoid, ".c", 1:ncol(sim))
sim <- add_meta(sim)
sim <- compute_all_gene_info(sim)
saveRDS(sim, file="../datasets/simMix1.SCE.rds")
```

# Simulating dataset 2

```{r, eval=FALSE}
sce <- readRDS("~/bioinfo/robinson/muscat-comparison/MAGL/output/MAGL-SCE.rds")
sce$oldclusters <- sce$cluster_id
sce$cluster_id <- as.factor(gsub("[0-9]$","",sce$cluster_id))
tt <- table(sce$sample_id)
sce <- sce[,which(sce$sample_id %in% names(tt)[which(tt>3000)])]
sce$sample_id <- droplevels(sce$sample_id)
sce <- prepSim(sce,min_size=50)
saveRDS(sce, file="magl_sce_prepSim.rds")
```
```{r, include=FALSE}
sce <- readRDS("magl_sce_prepSim.rds")
```

```{r}
sce$offset <- sce$offset+log(2)
```

```{r}
sims <- list()
set.seed(1234)
probs1 <- list( set_names(c(0.2,0.4,0.35,0.05),levels(sce$cluster_id)),
               c(1,0), c(1,0) )
lfcs1 <- set_names(c(2,4,1,2),levels(sce$cluster_id))
sims$s1 <- simData(sce, nrow(sce), 1500, probs=probs1, p_type=0.1, rel_lfc=lfcs1)
probs2 <- list( set_names(c(0.2,0.6,0.2,0),levels(sce$cluster_id)),
               c(1,0), c(1,0) )
lfcs2 <- set_names(c(3,2,1,1),levels(sce$cluster_id))
sims$s2 <- simData(sce, nrow(sce), 1000, probs=probs2, p_type=0.03, rel_lfc=lfcs2)
probs3 <- list( set_names(c(0.5,0.5,0,0),levels(sce$cluster_id)),
               c(1,0), c(1,0) )
lfcs3 <- set_names(c(1,1,1,1),levels(sce$cluster_id))
sims$s3 <- simData(sce, nrow(sce), 500, probs=probs3, p_type=0.01, rel_lfc=lfcs3)

sim <- SingleCellExperiment(list(counts=do.call(cbind, lapply(sims,counts))))
sim$phenoid <- paste0( rep(LETTERS[1:3],c(1500,1000,500)),
                       unlist(lapply(sims,FUN=function(x){
  levels(x$cluster_id) <- as.character(seq_along(levels(x$cluster_id)))
  x$cluster_id
})))
colnames(sim) <- paste0(sim$phenoid, ".c", 1:ncol(sim))
sim <- add_meta(sim)
sim <- compute_all_gene_info(sim)
saveRDS(sim, file="../datasets/simMix2.SCE.rds")
```

# Overview

```{r, fig.width=8.5, fig.height=3}
ll <- c( simMix1="../datasets/simMix1.SCE.rds",
         simMix2="../datasets/simMix2.SCE.rds" )
devtools::load_all("../../pipeComp")
describeDatasets(ll)
```

### 2020-02-03 main_figures/doublets.Rmd
replace load_all with normal lib

@@
   library(cowplot)
+  library(scDblFinder)
 })
-devtools::load_all("~/bioinfo/robinson/scDblFinder/")
 theme_set(theme_cowplot(font_size = 12))

### 2020-02-03 misc_functions.R
some helper functions for plotting

@@ -1,0 +1,187 @@
+truek <- c(Koh = 9, Kumar = 3, mixology10x3cl = 3, mixology10x5cl = 5, 
+           Zhengmix4eq = 4, Zhengmix4uneq = 4, Zhengmix8eq = 8)
+
+
+# value_format="%.3f"
+chm <- function(x, scale="none", sameScale=FALSE, scaleTitle=NULL, value_format="", rmLead0=TRUE,
+                row_ann=NULL, colors=c("blue", "white", "red"), col_sortFn=NULL, row_sortFn=NULL, 
+                cluster_rows=is.null(row_sortFn), cluster_columns=is.null(col_sortFn), 
+                scaleVar=TRUE, scaleFn=base::scale, ...){
+  x <- lapply(x, as.matrix)
+  xo <- x
+  if(!is.list(scaleFn)) scaleFn <- lapply(x, FUN=function(x) scaleFn)
+  if(length(scale)==1) scale <- rep(scale, length(x))
+  if(length(value_format)==1) value_format <- rep(value_format, length(x))
+  for(i in 1:length(x)){
+    rn <- row.names(x[[i]])
+    if(scaleVar){
+      if(scale[[i]]=="row") x[[i]] <- t(apply(x[[i]], 1, scaleFn[[i]]))
+      if(scale[[i]]=="column") x[[i]] <- apply(x[[i]], 2, scaleFn[[i]])
+    }else{
+      if(scale[[i]]=="row") x[[i]] <- x[[i]]-rowMeans(x[[i]],na.rm=T)
+      if(scale[[i]]=="column") x[[i]] <- t(t(x[[i]])-colMeans(x[[i]], na.rm=T))
+    }
+    row.names(x[[i]]) <- rn
+  }
+  if(!is.null(col_sortFn)){
+    o <- colnames(x[[1]])[order(apply(do.call(cbind, x), 2, FUN=col_sortFn),decreasing=T)]
+    x <- lapply(x, FUN=function(x) x[,o])
+    xo <- lapply(xo, FUN=function(x) x[,o])
+  }
+  if(!is.null(row_sortFn)){
+    o <- row.names(x[[1]])[order(apply(do.call(cbind, x), 1, na.rm=TRUE, FUN=row_sortFn),decreasing=T)]
+    x <- lapply(x, FUN=function(x) x[o,])
+    xo <- lapply(xo, FUN=function(x) x[o,])
+  }
+  if(is.function(colors)){
[…]

### 2020-02-03 main_figures/clustering.Rmd
changed load_all to pipeComp, fixed calls to internal functions

@@
   library(ComplexHeatmap)
-  library(matrixStats)
+  library
+  library(pipeComp)
 })
 source("../misc_functions.R")
-suppressMessages(devtools::load_all("../../pipeComp/"))
 theme_set(theme_cowplot(font_size = 12))

### 2020-02-03 main_figures/dimensionality.Rmd
changed load_all to pipeComp, fixed calls to internal functions

@@
   library(matrixStats)
+  library(pipeComp)
 })
 source("../misc_functions.R")
-suppressMessages(devtools::load_all("../../pipeComp/"))
-source("../../pipeComp/inst/extdata/scrna_alternatives.R")
+source(system.file("extdata/scrna_alternatives.R", package="pipeComp"))
 theme_set(theme_cowplot(font_size = 12))
@@
 ```{r}
-o <- readRDS("../../ndims.RDS")
+o <- readRDS("../data/ndims.RDS")
 o <- lapply(o, FUN=function(x){
@@
 h1 <- Heatmap(sw2, cluster_columns = FALSE, cluster_rows = FALSE, 
-        name="silhouette width", col=.silScale(sw2), show_column_names = FALSE,
+        name="silhouette width", col=pipeComp:::.silScale(sw2), show_column_names = FALSE,
         column_title = "Subpopulation mean silhouette width",
-        bottom_annotation=.ds_anno(colnames(sw2)), heatmap_legend_param = list(direction="horizontal", legend_width=unit(3, "cm")))
+        bottom_annotation=pipeComp:::.ds_anno(colnames(sw2)), heatmap_legend_param = list(direction="horizontal", legend_width=unit(3, "cm")))
 p2 <- grid.grabExpr(draw(h1, heatmap_legend_side="bottom"))

### 2020-02-03 main_figures/dr.Rmd
changed load_all to pipeComp, fixed calls to internal functions

@@
   library(matrixStats)
+  library(pipeComp)
 })
 source("../misc_functions.R")
-suppressMessages(devtools::load_all("../../pipeComp/"))
 ```
@@
 h1 <- Heatmap(x, cluster_columns = FALSE, cluster_rows = FALSE, 
-        name="silhouette width", col=.silScale(x), show_column_names = FALSE,
+        name="silhouette width", col=pipeComp:::.silScale(x), show_column_names = FALSE,
         column_title = "Subpopulation mean silhouette width", row_names_max_width = unit(9,"cm"),
-        bottom_annotation=.ds_anno(colnames(x)), heatmap_legend_param = list(direction="horizontal", legend_width=unit(3, "cm")))
+        bottom_annotation=pipeComp:::.ds_anno(colnames(x)), heatmap_legend_param = list(direction="horizontal", legend_width=unit(3, "cm")))
 p1 <- grid.grabExpr(draw(h1, heatmap_legend_side="bottom"))
@@
   scrna_evalPlot_clustAtTrueK(res, agg.by=c("norm","dr"), reorder_rows = h, anno_legend = FALSE)
-h <- .renameHrows(h, function(x) gsub("; dr=", " + ", gsub("seuratvst", "sctransform", x)))
+h <- pipeComp:::.renameHrows(h, function(x) gsub("; dr=", " + ", gsub("seuratvst", "sctransform", x)))
 

### 2020-02-03 main_figures/filtering.Rmd
changed load_all to pipeComp, fixed calls to internal functions

@@
   library(SingleCellExperiment)
+  library(pipeComp)
 })
 theme_set(theme_cowplot(font_size = 12))
-suppressMessages(devtools::load_all("../../pipeComp/"))
 source("../misc_functions.R")

### 2020-02-03 main_figures/imputation.Rmd
changed load_all to pipeComp, fixed calls to internal functions

@@
   library(matrixStats)
+  library(pipeComp)
 })
 source("../misc_functions.R")
-suppressMessages(devtools::load_all("../../pipeComp/"))
 theme_set(theme_cowplot(font_size = 10))
@@
 
-h1 <- Heatmap(dr, cluster_columns = FALSE, col=.silScale(dr), cluster_rows = FALSE, 
+h1 <- Heatmap(dr, cluster_columns = FALSE, col=pipeComp:::.silScale(dr), cluster_rows = FALSE, 
         bottom_annotation=dsa,
@@
       col=ifelse(ll2s[["',k,'"]][i,j]>',resmid,',"black","white"))) }')))
-  Heatmap(ll2s[[k]], col = viridis::inferno(100), show_column_names = FALSE, bottom_annotation = .ds_anno(colnames(ll2[[k]]), FALSE), cluster_columns=FALSE, cluster_rows = FALSE, show_heatmap_legend = FALSE, column_title=title, cell_fun=cellfn, ...)
+  Heatmap(ll2s[[k]], col = viridis::inferno(100), show_column_names = FALSE, bottom_annotation = pipeComp:::.ds_anno(colnames(ll2[[k]]), FALSE), cluster_columns=FALSE, cluster_rows = FALSE, show_heatmap_legend = FALSE, column_title=title, cell_fun=cellfn, ...)
 }

### 2020-02-03 main_figures/normalization.Rmd
changed load_all to pipeComp, fixed calls to internal functions

@@
   library(matrixStats)
+  library(pipeComp)
 })
-suppressMessages(devtools::load_all("../../pipeComp/"))
 source("../misc_functions.R")

### 2020-02-03 main_figures/selection.Rmd
changed load_all to pipeComp, fixed calls to internal functions

--- 2019-11-09 main_figures/figure4_selection.Rmd
@@
   library(matrixStats)
+  library(pipeComp)
 })
 source("../misc_functions.R")
-suppressMessages(devtools::load_all("../../pipeComp/"))
 data(datasets)

### 2020-02-03 supplementary_figures/supp1_dataset_description.Rmd
changed load_all to pipeComp, fixed calls to internal functions

@@
   library(pheatmap)
+  library(pipeComp)
 })
-suppressMessages(devtools::load_all("../../pipComp/"))
+source("../misc_functions.R")
 theme_set(theme_cowplot(font_size = 12))

### 2020-02-03 supplementary_figures/supp2_filtering.Rmd
changed load_all to pipeComp, fixed calls to internal functions

@@
   library(circlize)
+  library(pipeComp)
 })
-suppressMessages(devtools::load_all("../../pipeComp/"))
 source("../misc_functions.R")
-load("../../pipComp/data/datasets.RData")
 ```
@@
 levels(res$filt) <- c("coding only", "no mitochondrial", "no ribosomal", "all")
-res2 <- res[which(abs(res$nbClusters-datasets[res$dataset,"subpopulations"])<=1),]
+res2 <- res[which(abs(res$nbClusters-truek[res$dataset])<=1),]
 ll <- lapply(split(res2, res2$norm), FUN=function(x){

### 2020-02-03 supplementary_figures/supp3_normalization.Rmd
changed load_all to pipeComp, fixed calls to internal functions

@@
   library(viridis)
+  library(pipeComp)
 })
 suppressMessages(devtools::load_all("../../pipComp/"))
-data(datasets)
 theme_set(theme_cowplot(font_size = 11))
@@
 res2 <- res[which(res$filt=="deffilter" & res$norm %in% c("norm.scran", "norm.scran.noscale", "norm.seurat", "norm.seuratvst", "norm.stableGsum")),]
-res2$diff.nbClusters <- res2$nbClusters-datasets[res2$dataset,"subpopulations"]
+res2$diff.nbClusters <- res2$nbClusters-truek[res2$dataset]
 res3 <- aggregate(res2[,"diff.nbClusters"], by=res2[,c("dataset","dims","resolution","norm")], FUN=mean)

### 2020-02-03 supplementary_figures/supp4_selection.Rmd
changed load_all to pipeComp, fixed calls to internal functions

@@
   library(ComplexHeatmap)
+  library(pipeComp)
 })
-suppressMessages(devtools::load_all("../../pipComp/"))
-data(datasets)
 source("../misc_functions.R")

### 2020-02-03 supplementary_figures/supp5_dr.Rmd
changed load_all to pipeComp, fixed calls to internal functions

@@
   library(dplyr)
+  library(pipeComp)
+  library(viridis)
 })
 theme_set(theme_cowplot(font_size = 13))
-suppressMessages(devtools::load_all("../../pipComp/"))
-data(datasets)
 source("../misc_functions.R")
@@
 ```{r, fig.width=8, fig.height=8.5}
-library(viridis)
 res <- readRDS("../data/clust_endSummary.rds")
@@
 res2 <- res2[which(res2$dr=="seurat.pca" & res2$dataset!="Kumar"),]
-res2$diff.nbClusters <- res2$nbClusters-datasets[res2$dataset,"subpopulations"]
+res2$diff.nbClusters <- res2$nbClusters-truek[res2$dataset]
 res3 <- aggregate(res2[,"diff.nbClusters"], by=res2[,c("dataset","dims","resolution")], FUN=mean)

### 2020-02-03 supplementary_figures/supp6_clust.Rmd
changed load_all to pipeComp, fixed calls to internal functions

@@
   library(patchwork)
+  library(pipeComp)
 })
 theme_set(theme_cowplot(font_size = 12))
-devtools::load_all("../../pipeComp/")
 source("../misc_functions.R")

### 2020-02-03 supplementary_figures/supp_figures.Rmd
updated supplementary figures
added DR-related supplementaries
updated supp
changed load_all to pipeComp, fixed calls to internal functions

@@
 ---
-title: "Supplementary Figures"
-author: "Pierre-Luc Germain"
+title: \Large{pipeComp, a general framework for the evaluation of computational pipelines, reveals performant single-cell RNA-seq preprocessing tools}
+subtitle: \vspace{0.5cm}\LARGE{Supplementary Figures}\vspace{0.5cm}
+author:
+- "Pierre-Luc Germain"
+- "Anthony Sonrel"
+- "Mark D. Robinson"
 date: "`r format(Sys.time(), '%d %B, %Y')`"
@@
 ---
+
+\renewcommand{\thepage}{S\arabic{page}}
 
@@
 ```
+
+```{r child='supp2_filtering.Rmd'}
+```
+
+```{r child='supp3_normalization.Rmd'}
+```
+
+```{r child='supp4_selection.Rmd'}
+```
+
+```{r child='supp5_dr.Rmd'}
+```
+
+```{r child='supp6_clust.Rmd'}
+```

### 2020-02-27 bmc_article.tex

@@
 \author[
-   addressref={aff1,aff2},          % id's of addresses, e.g. {aff1,aff2}
+   addressref={aff1,aff2,aff3},          % id's of addresses, e.g. {aff1,aff2}
    email={pierre-luc.germain@hest.ethz.ch}   % email address
@@
 \author[
-   addressref={aff1},
+   addressref={aff1,aff2},
     email={anthony.sonrel@uzh.ch}
@@
 \author[
-   addressref={aff1},
+   addressref={aff1,aff2},
    corref={aff1},  % id of corresponding address, if any
@@
 \address[id=aff1]{
-  \orgname{Department of Molecular Life Sciences and SIB Swiss Institute of Bioinformatics, University of Zürich}, % university, etc
+  \orgname{Department of Molecular Life Sciences, University of Zürich}, % university, etc
   \street{Winterthurerstrasse 190},
@@
 \address[id=aff2]{
+  \orgname{SIB Swiss Institute of Bioinformatics}, % university, etc
+  \city{Zürich},
+  \cny{Switzerland}
+}
+\address[id=aff3]{
   \orgname{D-HEST Institute for Neurosciences, ETH Zürich},
@@
 %\parttitle{Background} %if any
-The massive growth of single-cell RNA sequencing (scRNAseq) and methods for its analysis still lacks sufficient and up-to-date benchmarks that would guide analytical choices.
+The massive growth of single-cell RNA-sequencing (scRNAseq) and methods for its analysis still lacks sufficient and up-to-date benchmarks that would guide analytical choices.
 Moreover, current studies are often focused on isolated steps of the process.
-Here, we present a flexible R framework for pipeline comparison with multi-level evaluation metrics, and apply it to the benchmark of scRNAseq analysis pipelines using datasets with known cell identities.
-We evaluate common steps of such analyses, including filtering, doublet detection (suggesting a new R package, \textit{scDblFinder}), normalization, feature selection, denoising, dimensionality reduction and clustering.
+Here, we present a flexible R framework for pipeline comparison with multi-level evaluation metrics and apply it to the benchmark of scRNAseq analysis pipelines using datasets with known cell identities.
+We evaluate common steps of such analyses, including filtering, doublet detection, normalization, feature selection, denoising, dimensionality reduction and clustering.
 On the basis of these analyses, we make a number of concrete recommendations about analysis choices.
-The evaluation framework, \textit{pipeComp}, has been implemented so as to easily integrate any other step or tool, permit extensible benchmarks and easy application to other fields (\url{https://github.com/plger/pipeComp}).
+The evaluation framework, \textit{pipeComp}, has been implemented so as to easily integrate any other step or tool, allowing extensible benchmarks and easy application to other fields (\url{https://github.com/plger/pipeComp}).
 
@@
 
-A number of good comparison and benchmark studies have already been performed on various steps related to scRNAseq analysis and can inform analytical decisions (\citealp{CobosDeconvolution2020, colePerformance2018, DalMolinDE2017, duoClustering2018, freytagComparison2018, HeiserDimred2019, JaakkolaDE2017, SonesonDE2018, SunDimRed2019, tianMixology2018, TranBatch2020, TsuyuzakiPCA2020, viethSystematic2019, WangDE2019, YipHVGs2018, ZhangImput2018}).
+A number of good comparison and benchmark studies have already been performed on various steps related to scRNAseq analysis and can guide the choice of methodology (\citealp{CobosDeconvolution2020, colePerformance2018, DalMolinDE2017, duoClustering2018, freytagComparison2018, GaoProcessing2020, HeiserDimred2019, HouImput2020, JaakkolaDE2017, SonesonDE2018, SunDimRed2019, tianMixology2018, TranBatch2020, TsuyuzakiPCA2020, viethSystematic2019, WangDE2019, YipHVGs2018, ZhangImput2018}).
 However these recommendations need constant updating and often leave open many details of an analysis.
@@
 Although previous benchmarks already brought valuable recommendations for data processing, some only focused on one aspect of data processing (e.g., \citealp{SunDimRed2019}), did not evaluate how the tool selection affects downstream analysis (e.g., \citealp{TsuyuzakiPCA2020}) or did not tackle all aspects of data processing, such as doublet identification or cell filtering (e.g., \citealp{viethSystematic2019}).
-A thorough evaluation of the tools covering all major processing steps is however urgently needed as previous benchmarking studies highlighted that a combination of tools can have a drastic impact on downstream analysis, such as differential expression analysis and cell-type deconvolution (\citealp{viethSystematic2019, CobosDeconvolution2020}).
+A thorough evaluation of the tools covering all major processing steps is however urgently needed as previous benchmarking studies highlighted that a combination of tools can have a drastic impact on downstream analysis, such as differential expression analysis and cell-type deconvolution\cite{viethSystematic2019,CobosDeconvolution2020}.
 It is then critical to evaluate not only the single effect of a preprocessing method but also its positive or negative interaction with all parts of a workflow.
@@
 Importantly, the success of methods at a certain analytical step might be dependent on choices at other steps.
-Therefore, instead of evaluating each step in isolation, we developed a general framework for evaluating nested variations on a pipeline, and suggest a multilevel panel of metrics.
+Therefore, instead of evaluating each step in isolation, we developed a general framework for evaluating nested variations on a pipeline and suggest a multilevel panel of metrics.
 Finally, we evaluate several recent methods and provide concrete recommendations.
@@
 Variations in a given parameter can be evaluated using all metrics from this point downward in the pipeline.
-This is especially important because end-point metrics, such as the adjusted Rand index (ARI, \citep{HubertARI1985}) for clustering, are not perfect.
-For example, although the meaning of an ARI score is independent of the number of true subpopulations \citep{steinley_properties_2004}, the number of clusters called is by far the most important determinant of the score: the farther it is from the actual number of subpopulations, the worse the ARI (Supplementary Figure 17).
-In this context, one strategy has been to cluster across various resolutions and only consider the results that have the right number of clusters \citep{duoClustering2018}.
-While this has the virtue of making the results comparable, in practice the number of subpopulations is typically unknown, and tools that operate well in this optimal context might not necessarily be best overall.
-Clustering results are also very sensitive, and might not always capture improvements in earlier steps of the pipeline.
+This is especially important because end-point metrics, such as the adjusted Rand index (ARI) \cite{HubertARI1985} for clustering, are not perfect.
+For example, although the meaning of an ARI score is independent of the number of true subpopulations \cite{steinleyProperties2004}, the number of clusters called is by far the most important determinant of the score: the farther it is from the actual number of subpopulations, the worse the ARI (Supplementary Figure 17).
+In this context, one strategy has been to cluster across various resolutions and only consider the results that have the right number of clusters \cite{duoClustering2018}.
+While this has the virtue of making the results comparable, in practice the number of subpopulations is typically unknown and tools that operate well in this optimal context might not necessarily be best overall.
+Clustering results are also very sensitive and might not always capture improvements in earlier steps of the pipeline.
 We therefore favour monitoring several complementary metrics across multiple steps of the process.
@@
 All (or subsets of) combinations of parameters can then be simultaneously ran and evaluated using the \texttt{runPipeline} function.
-\textbf{B:} Scheme representing the application of \textit{pipeComp} to evaluate a range of methods that are commonly used in scRNAseq studies, and some of the metrics monitored at various steps.}
+\textbf{B:} Scheme representing the application of \textit{pipeComp} to evaluate a range of methods that are commonly used in scRNAseq studies and some of the metrics monitored at various steps.}
     \label{fig:figure2}
@@
 
-Doublets, defined as two cells sequenced under the same cellular barcode (e.g., being captured in the same droplet), are fairly frequent in scRNAseq datasets, with estimates ranging from 1 to 10\% depending on the platform and cell concentration used \citep{bloomEstimating2018,kangMultiplexedDemuxlet2018}.
+Doublets, defined as two cells sequenced under the same cellular barcode (e.g., being captured in the same droplet), are fairly frequent in scRNAseq datasets, with estimates ranging from 1 to 10\% depending on the platform and cell concentration used \cite{bloomEstimating2018,kangMultiplexedDemuxlet2018}.
 While doublets of the same cell type are relatively innocuous for most downstream applications due to their conservation of the relative expression between genes, doublets formed from different cell types or states are likely to be misclassified and could potentially distort downstream analysis.
 In some cases, doublets can be identified through their unusually high number of reads and detected features, but this is not always the case (Supplementary Figure 2).
-A number of methods were developed to identify doublets, in most cases by comparing each cell to artificially-created doublets \citep{mcginnisDoubletfinder2019, LunScran2016, BaisScds2019}.
-We first evaluated the capacity of these methods to detect doublets using the two 10x datasets with cells of different genetic identity \citep{tianMixology2018}, using SNP genotypes as the ground truth.
-For the sole purpose of this section, we included an additional dataset with SNP information but lacking true cell labels \citep{kangMultiplexedDemuxlet2018}.
-Of note, SNP-based analyses also call doublets created by cells of the same cell type (but from different individuals), whose identification is generally not the primary aim of doublet detection methods, since these can be considered homotypic (as opposed to neotypic or heterotypic doublets, i.e.
-doublets from different cell types) and can be identified through other means.
+A number of methods were developed to identify doublets, in most cases by comparing each cell to artificially-created doublets \cite{mcginnisDoubletfinder2019, LunScran2016, BaisScds2019}.
+We first evaluated the capacity of these methods to detect doublets using the two 10x datasets with cells of different genetic identity \cite{tianMixology2018}, using SNP genotypes as the ground truth.
+For the sole purpose of this section, we included an additional dataset with SNP information but lacking true cell labels \cite{kangMultiplexedDemuxlet2018}.
+Of note, SNP-based analyses also call doublets created by cells of the same cell type (but from different individuals) and which are generally described as homotypic (as opposed to neotypic or heterotypic doublets, i.e.
+doublets from different cell types).
+These homotypic doublets might not be identifiable from the mere gene counts, and their identification is not generally the primary aim of doublet callers since they are often considered innocuous and, when across individuals, can be identified through other means (e.g., SNPs).
+We therefore do not expect a perfect accuracy in datasets involving cells of the same type across individuals (as in the demuxlet dataset).
 
-We tested \textit{DoubletFinder} \citep{mcginnisDoubletfinder2019} and \textit{scran}'s \texttt{doubletCells} \citep{LunScran2016}, both of which use similarity to artificial doublets, and \textit{scds} \citep{BaisScds2019}, which relies on a combination of co-expression and binary classification.
+We tested \textit{DoubletFinder} \cite{mcginnisDoubletfinder2019} and \textit{scran}'s \texttt{doubletCells} \cite{LunScran2016}, both of which use similarity to artificial doublets, and \textit{scds} \cite{BaisScds2019}, which relies on a combination of co-expression and binary classification.
 \textit{DoubletFinder} integrates a thresholding based on the proportion of expected doublets, while \textit{scran} and \textit{scds} return scores that must be manually thresholded.
@@
 Across datasets, cells called as doublets tended to be classified in the wrong cluster more often than other cells (Figure \ref{fig:figure3}C).
-We therefore tested whether this method also improved the accuracy of the clustering across all benchmark datasets, and found that it did even when, by design, the data contained no heterotypic doublet (Figure \ref{fig:figure4}).
+We also found that \textit{scDblFinder} improved the accuracy of the clustering across all benchmark datasets even when, by design, the data contained no heterotypic doublet (Figure \ref{fig:figure4}).
 
@@
     \centering
-    \includegraphics[width=\textwidth]{{main_figures/doublets_files/figure-html/figure1-1.png}}
+    \includegraphics[scale=0.65, keepaspectratio]{{main_figures/doublets_files/figure-html/figure1-1.png}}
     \caption{\textbf{Identification of doublet cells.
@@
 Beyond doublets, a dataset might include low-quality cells whose elimination would reduce noise.
-This has for instance been demonstrated for droplets containing a high content of mitochondrial reads, often as a result of cell degradation and resulting loss of cytoplasmic mRNAs \citep{IlicicLowQual2016}.
-A common practice is to exclude cells that differ considerably from most other cells on the basis of such properties, for instance through the \texttt{isOutlier} function from \textit{scater} that measures, for a given control property, the number of median absolute deviations (MADs) of each cell from the median of all cells.
+This has for instance been demonstrated for droplets containing a high content of mitochondrial reads, often as a result of cell degradation and resulting loss of cytoplasmic mRNAs \cite{IlicicLowQual2016}.
+A common practice is to exclude cells that differ considerably from most other cells on the basis of such properties.
+This can for instance be performed through the \texttt{isOutlier} function from \textit{scater} that measures, for a given control property, the number of median absolute deviations (MADs) of each cell from the median of all cells.
 Supplementary Figure 1 shows the distributions of some of the typical cell properties commonly used.
@@
 In our experience, 10X datasets also exhibit a very strong correlation between the total counts and the total features even across very different cell types (Supplementary Figure 4).
-We therefore also measure the ratio between the two, and treat cells strongly departing from this trend with suspicion.
+We therefore also measure the ratio between the two and treat cells strongly departing from this trend with suspicion.
 
-Reasoning that the cells we wish to eliminate are cells that would be misclassified, we measured the rate of misclassification of each cell in each dataset across a variety of clustering pipelines, correcting for the median misclassification rate of the subpopulation, and then evaluated what properties could be predictive of misclassification (Supplementary Figures 5-7).
+Reasoning that the cells we wish to exclude are the cells that would be misclassified, we measured the rate of misclassification of each cell in each dataset across a variety of clustering pipelines, correcting for the median misclassification rate of the subpopulation and then evaluated what properties could be predictive of misclassification (Supplementary Figures 5-7).
 We could not identify any property or simple combination thereof that would be consistently predictive of misclassification; the only feature that consistently stood out across multiple datasets (the Zheng datasets) was that cells with very high read counts have a higher chance of being misclassified.
@@
 We next investigated the impact of filtering according to various criteria (see Methods).
-An important risk of excluding cells on the basis of their distance from the whole distribution of cells on some properties (e.g.
-library size) is that these properties tend to have different distributions across subpopulations.
+An important risk of excluding cells on the basis of their distance from the whole distribution of cells on some properties (e.g., library size) is that these properties tend to have different distributions across subpopulations.
 As a result, thresholds in terms of number of MADs from the whole distribution can lead to strong biases against certain subpopulations (Figure \ref{fig:figure4}A).
 We therefore examined the tradeoff between the increased accuracy of filtering and the maximum proportion of cells excluded per subpopulation (Figure \ref{fig:figure4}B and Supplementary Figure 8).
-Since filtering changes the relative abundance of the different subpopulations, global clustering accuracy metrics such as ARI are not appropriate here, and we therefore calculated the per-subpopulation precision and recall using the Hungarian algorithm \citep{PapadimitriouHu1998}, and monitored their mean F1 score.
-A first observation was that, although the stringent filtering tended to be associated with an increase in accuracy, it could also become deleterious, and most of the benefits could be achieved without very stringent filtering and minimizing subpopulation bias (Figure \ref{fig:figure4}B).
+Since filtering changes the relative abundance of the different subpopulations, global clustering accuracy metrics such as ARI are not appropriate here.
+We therefore calculated the per-subpopulation precision and recall using the Hungarian algorithm \cite{PapadimitriouHu1998} and monitored their mean F1 score.
+A first observation was that, although the stringent filtering tended to be associated with an increase in accuracy, it could also become deleterious and most of the benefits could be achieved without very stringent filtering and minimizing subpopulation bias (Figure \ref{fig:figure4}B).
 Applying the same filtering criteria on individual clusters of cells (identified through scran's \texttt{quickCluster} method) resulted in nearly no cell being filtered out.
@@
     \centering
-    \includegraphics[width=\textwidth]{{main_figures/filtering_files/figure-html/figure2-1.png}}
+    \includegraphics[scale=0.60, keepaspectratio]{{main_figures/filtering_files/figure-html/figure2-1.png}}
     \caption{\textbf{A:} Filtering on the basis of distance to the whole distribution can lead to strong bias against certain subpopulations.
@@
 
-Mitochondrial reads have been associated with cell degradation, and there is evidence that ribosomal genes can influence the clustering output, hiding other biological structure in the analysis \citep{freytagComparison2018}.
+Mitochondrial reads have been associated with cell degradation and there is evidence that ribosomal genes can influence the clustering output, hiding other biological structure in the analysis \cite{freytagComparison2018}.
 We therefore investigated whether excluding one category of features or the other, or using only protein-coding genes, had an impact on the ability to distinguish subpopulations (Supplementary Figure 9).
@@
 We next investigated the impact of different normalization strategies.
-Beside the standard log-normalization included in \textit{Seurat}, we tested \textit{scran}'s pooling-based normalization \citep{lunPooling2016}, \textit{sctransform}'s variance-stabilizing transformation \citep{hafemeisterSCtransform2019}, and normalization based on stable genes \citep{linStableGenes2018, deekeStablyExpressed2018}.
+Beside the standard log-normalization included in \textit{Seurat}, we tested \textit{scran}'s pooling-based normalization \cite{lunPooling2016}, \textit{sctransform}'s variance-stabilizing transformation \cite{hafemeisterSCtransform2019}, and normalization based on stable genes \cite{linStableGenes2018, deekeStablyExpressed2018}.
 In addition to log-normalization, the standard \textit{Seurat} clustering pipeline performs per-feature unit-variance scaling so that the PCA is not too strongly dominated by highly-expressed features.
-We therefore included versions of the different normalization procedures, with or without a subsequent scaling (\textit{sctransform}'s variance-stabilizing transformation involves something analogous to scaling).
+We therefore included versions of the different normalization procedures, with or without a subsequent scaling (\textit{sctransform}'s variance-stabilizing transformation involves an approach analogous to scaling).
 \textit{Seurat}'s scaling function also includes the option to regress out the effect of certain covariates.
 We tested its performance by using the proportion of mitochondrial counts and the number of detected features as covariates.
-Finally, it has been proposed that the use of stable genes, in particular cytosolic ribosomal genes, can be used to normalize scRNAseq\citep{deekeStablyExpressed2018}.
+Finally, it has been proposed that the use of stable genes, in particular cytosolic ribosomal genes, can be used to normalize scRNAseq \cite{deekeStablyExpressed2018}.
 We therefore evaluated a simple linear normalization based on the sum of these genes, as well as nuclear genes.
 
-An important motivation for the development of \textit{sctransform} was the observation that, even after normalization, the first principal components of various datasets tended to correlate with library size, suggesting an inadequate normalization \citep{hafemeisterSCtransform2019}.
-We therefore first assessed to what extent the first principal component still retained a correlation with the library size and the number of detected features, removing the confounding covariation with the subpopulations (Figure \ref{fig:figure5}A).
+An important motivation for the development of \textit{sctransform} was the observation that, even after normalization, the first principal components of various datasets tended to correlate with library size, suggesting an inadequate normalization \cite{hafemeisterSCtransform2019}.
+However, as library size tends to vary across subpopulations, part of this effect can simply reflect biological differences.
+We therefore assessed to what extent the first principal component still retained a correlation with the library size and the number of detected features, removing the confounding covariation with the subpopulations (Figure \ref{fig:figure5}A).
 The simple step of scaling tended to remove much of the correlation with these features and, in the absence of scaling, standard \textit{Seurat} normalization resulted in fairly high correlation with technical covariates.
@@
 
-We next investigated the impact of normalization on the separability of the subpopulations (Figure \ref{fig:figure5}B-C).
-Since clustering accuracy metrics such as the ARI are very strongly influenced by the number of clusters, we complemented it with silhouette width \citep{RousseeuwSil1987} and mutual information.
+We further evaluated normalization methods by investigating their impact on the separability of the subpopulations (Figure \ref{fig:figure5}B-C).
+Since clustering accuracy metrics such as the ARI are very strongly influenced by the number of clusters, we complemented it with silhouette width \cite{RousseeuwSil1987} and mutual information (MI).
 We found most methods (including no normalization at all) to perform fairly well in most of the subpopulations.
-Scaling tended to reduce the average silhouette width of some subpopulations and to increase that of some less distinguishable ones, and was generally but not always beneficial on the accuracy of the final clustering.
+An exception was \textit{scVI} (Supplementary Figure TBD) that performed poorly on Zhengmix and simMix1 datasets in view of the subpopulation silhouette and to a relative poor ARI.
+Scaling tended to reduce the average silhouette width of some subpopulations and to increase that of some less distinguishable ones and was generally, but not always, beneficial on the accuracy of the final clustering.
 Regressing out covariates systematically gave poorer performance on all metrics.
-\textit{sctransform} systematically outperformed other methods, and even though it was developed to be applied to data with unique molecular identifiers (UMI), it also performed fairly well with the Smart-seq protocol (Koh and Kumar datasets).
+\textit{sctransform} systematically outperformed other methods and, even though it was developed to be applied to data with unique molecular identifiers (UMI), it also performed fairly well with the Smart-seq protocol (Koh and Kumar datasets).
 
@@
     \includegraphics[width=\textwidth]{{main_figures/normalization_files/figure-html/figure3-1.png}}
-    \caption{\textbf{Evaluation of normalization procedures.} `Seurat.feat\_mt\_regress' indicates the regressing out of the number of features and proportion of mitochondrial reads during scaling, represents, and `Seurat.feat\_regress` of the number of features only.
+    \caption{\textbf{Evaluation of normalization procedures.} `Seurat.feat\_mt\_regress' regresses out the number of features and proportion of mitochondrial reads during scaling; `Seurat.feat\_regress` regresses out the number of features only.
 Residual correlation of the first principal component with library size (left) and detection rate (right), after accounting for biological differences between subpopulations.
@@
 Another common practice in feature selection is to use those with the highest mean expression.
-Recently, \citep{townesGlmpca2019} instead suggested to use deviance, while \textit{sctransform} provides its own ordering of genes based on transformed variance.
+Recently, \cite{townesGlmpca2019} instead suggested to use deviance, while \textit{sctransform} provides its own ordering of genes based on transformed variance.
 
 Reasoning that a selection method should ideally select genes whose variability is higher between subpopulations than within, we first assessed to what extent each method selected genes with a high proportion of variance or deviance explained by (real) subpopulation.
-As the proportion of variability in a gene attributable to subpopulations can be measured in various ways, we first compared three of them.
-Analysis of variance performed on a standard \textit{Seurat} normalization or on \textit{sctransform} data were highly correlated (Supplementary Figure 11A).
+As the proportion of variability in a gene attributable to subpopulations can be measured in various ways, we first compared three approaches: ANOVA on log-normalized count, ANOVA on \textit{sctransform} normalization, and deviance explained.
+The ANOVAs performed on a standard \textit{Seurat} normalization and on \textit{sctransform} data were highly correlated (Supplementary Figure 11A).
 These estimates were also in good agreement with the deviance explained, although lowly-expressed genes could have a high deviance explained without having much of their variance explained by subpopulation (Supplementary Figure 11B-D).
@@
 The standardized measures of variability were systematically worse than their non-standardized counterparts in selecting genes with a high proportion of variance explained by subpopulation.
-Regarding the percentage of deviance explained however, the standardized measures were often superior(Figure \ref{fig:figure6}A and Supplementary Figures 12-13).
-Deviance proved the method of choice to prioritize genes with a high variance explained by subpopulations (with mere expression level proving surprisingly good), but did not perform so well to select genes with a high deviance explained.
+Regarding the percentage of deviance explained however, the standardized measures were often superior (Figure \ref{fig:figure6}A and Supplementary Figures 12-13).
+Deviance proved the method of choice to prioritize genes with a high variance explained by subpopulations (with mere expression level proving surprisingly good) but did not perform so well to select genes with a high deviance explained.
 
@@
 Based on standard \textit{Seurat} normalization (left) or \textit{sctransform} (right).
-The methods `vst.varExp' and `devianceExplained' are the estimates used in \textbf{A} to evaluate the selection methods, and were included here only for validation purpose.}
+The  \textit{vst.varExp} and \textit{devianceExplained} methods correspond to the estimates used in \textbf{A} to evaluate the selection methods and were included here only for validation purpose.}
     \label{fig:figure6}
@@
 
-Since the various PCA approaches and implementations were recently benchmarked in a similar context \citep{TsuyuzakiPCA2020}, we focused on widely used approaches which had not yet been compared: \textit{Seurat}'s PCA, \textit{scran}'s \texttt{denoisePCA}, and GLM-PCA \citep{townesGlmpca2019}, and combined them (when relevant) with \textit{sctransform} normalization.
+Since the various PCA approaches and implementations were recently benchmarked in a similar context \cite{TsuyuzakiPCA2020}, we focused on widely used approaches which had not yet been compared: \textit{Seurat}'s PCA, \textit{scran}'s \texttt{denoisePCA}, and GLM-PCA \cite{townesGlmpca2019}.
+When relevant, we combined them with \textit{sctransform} normalization.
 Given that \textit{Seurat}'s default PCA weights the cell embeddings by the variance of each component, we also evaluated the impact of this weighting with each method.
 
-The impact of the choice of dimensionality reduction method was far greater than that of normalization or feature selection (e.g.
-Supplementary Figure 15).
+The impact of the choice of dimensionality reduction method was far greater than that of normalization or feature selection (e.g., Supplementary Figure 15).
 GLM-PCA tended to increase the average silhouette width of already well-defined subpopulations, but \textit{Seurat}'s PCA procedure however proved superior on all metrics (Figure \ref{fig:figure7}).
@@
     \centering
-    \includegraphics[width=\textwidth]{{main_figures/dr_files/figure-html/dimred-1.png}}
+    \includegraphics[scale=0.40, keepaspectratio]{{main_figures/dr_files/figure-html/dimred-1.png}}
     \caption{\textbf{Evaluation of common dimensionality reduction methods.
@@
 Overall, increasing the number of dimensions robustly led to a decrease in the number of clusters (Supplementary Figures 10 and 15).
-This tended to be translated to the accuracy of the clustering (Supplementary Figure 16), although in both cases (number of clusters and ARI) \textit{Seurat}'s resolution parameter had a much stronger impact.
+This tended to affect the accuracy of the clustering (Supplementary Figure 16), although in both cases (number of clusters and ARI) \textit{Seurat}'s resolution parameter had a much stronger impact.
 
 Different approaches have been proposed to select the appropriate number of dimensions, from the visual identification of an 'Elbow' (inflexion point) of the variance explained, to more complex algorithms.
-We evaluated the performance of common tools, namely [TBD] (\textit{pcaLocal.FO} label, [REF needed?]), global maximum likelihood based on translated Poisson Mixture Model (\textit{maxLikGlobal}, \citep{haroTranslated2008}), [TBD] (\textit{essLocal.b}, \textit{essLocal.b},[REF needed?]), JackStraw procedure (\textit{jackstraw.elbow}, \citep{ChungJackstraw2015}), [TBD](\textit{pcaLocal.maxgap}, [REF needed?]), [TBD] (\textit{pcaLocal.fan} [REF needed?]), Fisher Separability analysis (\textit{fisherSeparability}, \citep{AlberganteSepar2019}), \textit{scran}'s \texttt{denoisePCA} function (\textit{scran.denoisePCA}, \citep{LunScran2016}) and [TBD] (\textit{pcaOptmPointwise.max}, [REF needed?]).
+We evaluated the performance of dimensionality estimators implemented in the \textit{intrinsicDimension} package \cite{johnssonDimensionality2015}, as well as common procedures such as the `elbow` method (inflexion point in the variance explained by each component), some scRNAseq-specific methods such as the JackStraw procedure \cite{ChungJackstraw2015} or \textit{scran}'s \texttt{denoisePCA} \cite{LunScran2016}, and the recent application of Fisher Separability analysis \cite{AlberganteSepar2019}.
 
@@
     \centering
-    \includegraphics[width=\textwidth]{{main_figures/dimensionality_files/figure-html/figure6-1.png}}
+    \includegraphics[scale=0.53, keepaspectratio]{{main_figures/dimensionality_files/figure-html/figure6-1.png}}
     \caption{\textbf{Estimating dimensionality.
@@
 The last step evaluated in our pipeline was clustering.
-Given previous works on the topic \citep{duoClustering2018,freytagComparison2018} and the success of graph-based clustering methods for scRNA-seq, we restricted our evaluation to Seurat's method and two \textit{scran} SNN-based clustering approaches, based on random walks (\textit{walktrap} method) or on the optimization of the modularity score (\textit{fast\_greedy}).
-Again, the tested methods were combined with Seurat's standard normalization and \textit{sctransform}, otherwise using the parameters found optimal in the previous steps.
+Given previous works on the topic \cite{duoClustering2018,freytagComparison2018} and the success of graph-based clustering methods for scRNAseq, we restricted our evaluation to Seurat's method and two \textit{scran} SNN-based clustering approaches, based on random walks (\textit{walktrap} method) or on the optimization of the modularity score (\textit{fast\_greedy}).
+Again, the tested methods were combined with \textit{Seurat}'s standard normalization and \textit{sctransform}, otherwise using the parameters found optimal in the previous steps.
 
-Since ARI is dominated by differences in the number of clusters (Supplementary Figure 17), and no single metric is perfect, we diversified them (Figure \ref{fig:figure9}).
-Mutual information (MI) has the virtue of not decreasing when a true subpopulation is split into two clusters, which is arguably less problematic (and might well reflect unknown biological subgroups), but as a consequence it can be biased towards methods producing higher resolution clustering.
+Since ARI is dominated by differences in the number of clusters (Supplementary Figure 17) and no single metric is perfect, we diversified them (Figure \ref{fig:figure9}).
+MI has the virtue of not decreasing when a true subpopulation is split into two clusters, which is arguably less problematic (and might well reflect unknown biological subgroups), but as a consequence it can be biased towards methods producing higher resolution clustering.
 Similarly, precision per true subpopulation is considerably more robust to differences in the number of clusters.
@@
 
-The MI score and minimum precision, which are largely independent of the estimated number of clusters, were overall higher for the \textit{walktrap} method (Figure \ref{fig:figure9}), while the mean F1 score favored both \textitt{scran} methods (\textit{walktrap} and \textit{fast greedy}) over \textit{Seurat}.
+The MI score and minimum precision, which are largely independent of the estimated number of clusters, were overall higher for the \textit{walktrap} method (Figure \ref{fig:figure9}), while the mean F1 score favored both \textit{scran} methods (\textit{walktrap} and \textit{fast greedy}) over \textit{Seurat}.
 Finally, the ARI score at the true number of clusters, when available, showed similar performances.
-However, while \textit{Seurat}'s resolution parameter had a large impact on the number of clusters identified, and therefore could be used to find a parameter combination producing the right number of clusters, the number of clusters found by \textit{scran}-based clustering was considerably less influenced by the available parameters (number of nearest neighbors or steps in the random walk - see Supplementary Figure 18).
-The occasional lack of \textit{scran}-based clustering with the right number of clusters, along with a higher mutual information, suggest that \textit{scran} sometimes simply divides a real subpopulation into two clusters (possibly tracking some unknown biological differences) rather than committing misclassification errors.
+However, because \textit{Seurat}'s resolution parameter had a large impact on the number of clusters identified (Supplementary Figure 18), \textit{Seurat} could always be coerced into producing the right number of clusters.
+Instead, the number of clusters found by \textit{scran} was considerably less influenced by the available parameters (number of nearest neighbors or steps in the random walk - see Supplementary Figure 18), and as a result \textit{scran}-based clustering sometimes never produced a partitioning with the right number of clusters.
+This observation, along with \textit{scran}'s higher MI score, suggest that \textit{scran} sometimes simply divides a real subpopulation into two clusters (possibly tracking some unknown biological differences) rather than committing misclassification errors.
 Overall, the \textit{walktrap} method appeared superior to the \textit{fast greedy} algorithm and was generally less prone to misclassification than \textit{Seurat} clustering, although the latter offered more control over the resolution.
-Of note, very difficult subpopulations, both from the Zhengmix8eq and simMix1 datasets, remained very inaccurately classified by all methods and in regards to all metrics.
+Finally, some poorly distinguishable subpopulations from both the Zhengmix8eq and simMix1 datasets remained very inaccurately classified by all methods and in regards to all metrics.
 
@@
     \includegraphics[width=\textwidth]{{main_figures/clustering_files/figure-html/clustering-1.png}}
-    \caption{\textbf{Evaluation of clustering methods.} Clustering accuracy of common clustering tools in combination with \textitt{sctransform} and standard Seurat's normalization.}
+    \caption{\textbf{Evaluation of clustering methods.} Clustering accuracy of common clustering tools in combination with \textit{sctransform} and standard Seurat's normalization.}
     \label{fig:figure9}
@@
 Since preliminary analysis showed that all methods performed equally well or better on normalized data, we applied them after filtering and normalization, but before scaling and reduction.
-Although some of the methods (e.g.
-\textit{DRImpute\_process} and \textit{alra\_norm}) did improve the separability of some more elusive subpopulations, no method had a systematically positive impact on the average silhouette width across all subpopulations (Figure \ref{fig:figure10}A).
+Although some of the methods (e.g., \textit{DRImpute\_process} and \textit{alra\_norm}) did improve the separability of some more elusive subpopulations, no method had a systematically positive impact on the average silhouette width across all subpopulations (Figure \ref{fig:figure10}A).
 When restricting ourselves to clustering analyses that yielded the `right' number of clusters, all tested methods improved classification  compared to a scenario with no imputation step (`none' label, Figure \ref{fig:figure10}B).
-However, the situation was not so straightforward with alternative metrics, where some methods (e.g.
-\textit{enhance}) consistently underperformed.
-10X datasets, which are typically characterized by a lower per-cell coverage and feature detection rate, benefited more from imputation, and in this context, \textit{DrImpute} and \textit{dca} tended to show the best performance.
+However, the situation was not so straightforward with alternative metrics, where some methods (e.g., \textit{enhance}) consistently underperformed.
+10X datasets, which are typically characterized by a lower per-cell coverage and feature detection rate, benefited more from imputation and, in this context, \textit{DrImpute} and \textit{dca} tended to show the best performance.
 On the contrary, imputing on normalized counts originating from Smart-seq technology was instead rather deleterious to the clustering accuracy.
-The good performance of \textit{DrImpute} is in line with a previous study focused on the preservation of data structure in trajectory inference from scRNAseq \citep{ZhangImput2018} (\textit{alra} was however not included in this previous benchmark).
 
@@
     \includegraphics[width=\textwidth]{{main_figures/imputation_files/figure-html/imputation-1.png}}
-    \caption{\textbf{Imputation/denoising methods} Average silhouette width per subpopulation (\textbf{A}) and clustering accuracy (\textbf{B}) with or without (\textit{none}) application of a denoising/imputation method}
+     \caption{\textbf{Imputation/denoising methods} Average silhouette width per subpopulation (\textbf{A}) and clustering accuracy (\textbf{B}) with or without (indicated as \textit{none}) application of a denoising/imputation method.}
     \label{fig:figure10}
@@
    \begin{itemize}
-     \item Doublet detection and removal is advised, and can be performed at little computing cost with software such as \textit{scDblFinder} or \textit{scds}.
-     \item Distribution-based cell filtering fails to capture doublets, and should use relatively lenient cutoffs (e.g.
-5 MADs, or 3 MADs in at least 2 distributions) to exclude poor-quality cells while avoiding bias against some subpopulations.
+     \item Doublet detection and removal is advised and can be performed at little computing cost with software such as \textit{scDblFinder} or \textit{scds}.
+     \item Distribution-based cell filtering fails to capture doublets and should use relatively lenient cutoffs (e.g., 5 MADs, or 3 MADs in at least 2 distributions) to exclude poor-quality cells while avoiding bias against some subpopulations.
      \item Features filtering based on feature type did not appear beneficial.
@@
    \begin{itemize}
-     \item Deviance \citep{townesGlmpca2019} offered the best ranking of genes for feature selection.
-     \item Increasing the number of features included tended to lead to better classifications, plateauing from 4000 features.
+     \item Deviance \cite{townesGlmpca2019} offered the best ranking of genes for feature selection.
+     \item Increasing the number of features included tended to lead to better classifications, plateauing from 4000 features in our datasets.
    \end{itemize}
@@
    \begin{itemize}
-     \item Similarly to previous reports \citep{SunDimRed2019}, we recommend the irlba-based PCA using weighting of the components, as implemented in \textit{Seurat}. 
-     \item Instead of the common elbow or jackstraw methods for deciding on the components to include, we recommend the \textit{global maximum likelihood based on translated Poisson Mixture Model} method (e.g.
-implemented in \texttt{intrinsicDimension::maxLikGlobalDimEst()}).
+     \item Similarly to previous reports \cite{SunDimRed2019}, we recommend the irlba-based PCA using weighting of the components, as implemented in \textit{Seurat}. 
+     \item Instead of the common elbow or jackstraw methods for deciding on the components to include, we recommend the \textit{global maximum likelihood based on translated Poisson Mixture Model} method (e.g., implemented in the  \texttt{maxLikGlobalDimEst} function of \texttt{intrinsicDimension}).
    \end{itemize}
@@
 
+\begin{figure}[h]
+    \centering
+    \includegraphics[scale=0.08,keepaspectratio]{main_figures/pipeComp_summary.png}
+    \caption{\textbf{Recommendations of tools from filtering to clustering.}}
+    \label{fig:figure11}
+\end{figure}
+
 \subsection*{Limitations and open questions}
@@
 In this study, we evaluated tools commonly used for the processing of scRNAseq with a focus on droplet-based datasets, namely from the 10x technology.
-Although this platform has been used in almost half of the scRNAseq studies in 2019 \citep{SvenssonDB2019}, other popular technologies such as Drop-seq, InDrops or Smart-seq2/3 were not represented in the present benchmarking.
-Differences between such protocols, even among droplet-based technologies, can have a very large impact on cell capture efficiency, cell numbers and clustering \citealp{MereuCellAtlas2019, ZhangDroplet2019, SalomonDroplet2019}.
+Although this platform has been used in almost half of the scRNAseq studies in 2019 \cite{SvenssonDB2019}, other popular technologies such as Drop-seq, InDrops or Smart-seq2/3 were not represented in the present benchmarking.
+Differences between such protocols, even among droplet-based technologies, can have a very large impact on cell capture efficiency, cell numbers and clustering \cite{MereuCellAtlas2019, ZhangDroplet2019, SalomonDroplet2019}.
 Although most top-ranking methods in our comparison performed well on both Smart-seq and 10x datasets that we tested, future benchmarking efforts should strive to include less represented technologies.
-In addition, we did not compare any of the alignment and/or quantification methods used to obtain the count matrix, which was for instance discussed in \citep{viethSystematic2019}.
-Some steps, such as the implementation of the PCA, were also not explored in detail here as they have already been the object of recent and thorough study elsewhere \citep{SunDimRed2019, TsuyuzakiPCA2020}.
-We also considered only methods relying on Euclidean distance, while correlation was recently reported to be superior\citep{kim_impact_2019} and would require further investigation.
+In addition, we did not compare any of the alignment and/or quantification methods used to obtain the count matrix, which was for instance discussed in \cite{viethSystematic2019}.
+Some steps, such as the implementation of the PCA, were also not explored in detail here as they have already been the object of recent and thorough study elsewhere \cite{SunDimRed2019, TsuyuzakiPCA2020}.
+We also considered only methods relying on Euclidean distance, while correlation was recently reported to be superior\cite{kim_impact_2019} and would require further investigation.
 
-Here, we chose to concentrate on what could be learned from datasets with known cell labels (as opposed to labels inferred from the data, as in \citep{MereuCellAtlas2019}).
+Our recommendations are in line with the results of several other benchmarking studies.
+For instance, \textit{Seurat} and \textit{scran} methods were already shown to yield good clustering performance \cite{freytagComparison2018} and \textit{DrImpute} to improve the quality of downstream analyses \cite{ZhangImput2018}.
+However, a previous systematic comparison study \cite{viethSystematic2019} revealed that the use of \textit{SAVER} for denoising data and \textit{Scran} for clustering yielded better performance.
+It should be noted that the criteria of evaluations of the latter study differed to ours as the authors focused on the ability of the tools to recover DE genes, whereas we focused on the ability to recover cell populations.
+Another disagreement with the present recommendations come from the study of Hou et al.
+\cite{HouImput2020} that recommends \textit{SAVERX} and discourage the use of \textit{dca} for better clustering, whereas our study found opposite results for these tools.
+Several differences in the design of our study could explain this contrast, such as the use of different metrics (ARI at true number of clusters, mean precision/ recall), of different datasets (Koh, Kumar, simMix simulations) and the use of different filtering and normalization strategies.
+\textit{alra} however yielded good results in both studies and may be the most conservative option.
+The study from Hou et al.
+also highlighted the usefulness of \textit{alra} when followed by differential and trajectory analysis both in UMI and Fluidigm technologies.
+
+Here, we chose to concentrate on what could be learned from datasets with known cell labels (as opposed to labels inferred from the data, as in \cite{MereuCellAtlas2019}).
 In contrast to Tian et al.
-\citep{tianMixology2018}, who used RNA mixtures of known proportions, we chose to rely chiefly on real cells and their representative form of variability.
+\cite{tianMixology2018}, who used RNA mixtures of known proportions, we chose to rely chiefly on real cells and their representative form of variability.
 Given the limited availability of such well-described datasets however, several aspects of single-cell analysis could not be compared, such as batch effect correction or multi-dataset integration.
-For these aspects of scRNAseq processing that are critical in some experimental designs, we refer the reader to previous evaluations \citealp{TranBatch2020, StuartIntegration2019}.
+For these aspects of scRNAseq processing that are critical in some experimental designs, we refer the reader to previous evaluations \cite{TranBatch2020, StuartIntegration2019}.
 In addition, a focus on the identification of the subpopulations might fail to reveal methods that are instead best performing for tasks other than clustering, such as differential expression analysis or trajectory inference.
-Several informative benchmarks have already been performed on some of these topics \citealp{CrowellMuscat2019, DalMolinDE2017, JaakkolaDE2017, SaelensTraject2019, SonesonDE2018, WangDE2019}.
+Several informative benchmarks have already been performed on some of these topics \cite{CrowellMuscat2019, DalMolinDE2017, JaakkolaDE2017, SaelensTraject2019, SonesonDE2018, WangDE2019}.
 Yet, such evaluations could benefit from considering methods not in isolation, but as parts of a connected workflow, as we have done here.
@@
 
-We developed \ŧexttt{pipeComp} to address the need of a framework that can evaluate multiple tools simultaneously.
-The \textit{CellBench} package was recently proposed to address a similar need\citep{su_cellbench}, offering an elegant piping syntax to combine alternative methods at successive steps.
+We developed \texttt{pipeComp} to address the need of a framework that can simultaneously evaluate the interaction of multiple tools.
+The work of Vieth and colleagues\cite{viethSystematic2019} already offered an important precedent in this respect, evaluating the interaction of various steps in the context of scRNAseq differential expression analysis, but did not offer a platform for doing so.
+Instead, the \textit{CellBench} package was recently proposed to address a similar need \cite{suCellbench}, offering an elegant piping syntax to combine alternative methods at successive steps.
 A key additional feature of \textit{pipeComp} is its ability to perform evaluation in parallel to the pipeline and offering on-the-fly evaluation at any step.
 This avoid the need to store potentially large intermediate data for all possible permutations and thus allowing a combinatorial benchmark.
-In addition, the fact that benchmark functions are stored in the \texttt{PipelineDefinition} makes the benchmark more smoothly portable, enabling it transportable easy to apply comparable contains all information regarding the benchmark, including
+In addition, the fact that benchmark functions are stored in the \texttt{PipelineDefinition} makes the benchmark more smoothly portable, making it easy to modify, extend with further methods, or apply to other datasets.
 
-We believe there is much space for improvement in methods.
-Concerning cell filtering, we noticed that the current approach based on whole-population characteristic (e.g.
-MAD cut-off) can be biased against certain subpopulations, suggesting that more refined methods expecting multimodal distributions should be used rather than relying on whole-population characteristics.
+With respect to the methods themselves, we believe there is still space for improvement in some of the steps.
+Concerning cell filtering, we noticed that the current approach based on whole-population characteristic (e.g., MAD cut-off) can be biased against certain subpopulations, suggesting that more refined methods expecting multimodal distributions should be used rather than relying on whole-population characteristics.
 In addition, most common filtering approaches do not harness the relationship between cell QC properties.
 Finally, imputation had a varied impact on the clustering analysis and seemed to be linked to the technology that was used to generate the data.
-We support the hypothesis of \citep{ZhangImput2018} on the respective strengths of linear and non-linear models and their use to different types of data, such as the highest performance of non-linear methods in developmental studies.
+Also, the good performance of \textit{DrImpute} is in line with a previous study focused on the preservation of data structure in trajectory inference from scRNAseq \cite{ZhangImput2018} (\textit{alra} was however not included in this previous benchmark).
+Our results also align with the hypothesis of this study on the respective strengths of linear and non-linear models and their use to different types of data, such as the highest performance of non-linear methods in developmental studies.
 
@@
 
-The gene counts for the two mixology datasets were downloaded from the  \href[CellBench repository]{https://github.com/LuyiTian/CellBench\_data/data/sincell\_with\_class.RData}, commit 74fe79e.
-For the other datasets, we used the unfiltered counts from \citep{duoClustering2018}, available on the corresponding repository \url{https://github.com/markrobinsonuzh/scRNAseq\_clustering\_comparison}.
+The gene counts for the two mixology datasets were downloaded from the  \href{https://github.com/LuyiTian/CellBench_data/data/sincell_with_class.RData}{CellBench repository}, commit 74fe79e.
+For the other datasets, we used the unfiltered counts from \cite{duoClustering2018}, available on the corresponding repository \url{https://github.com/markrobinsonuzh/scRNAseq\_clustering\_comparison}.
 For simplicity, all starting \textit{SingleCellExperiment} objects with standardized metadata are available on \url{https://github.com/markrobinsonuzh/scRNA\_pipelines\_paper}.
@@
 \subsection*{Software and package versions}
-Analyses were performed in R 3.6.0 (Bioconductor 3.9), and the following packages were installed from github repositories: \textit{Seurat} (version 3.0.0), \textit{sctransform} (0.2.0), \textit{DoubletFinder} (2.0.1), \textit{scDblFinder} (1.1.1), \textit{scds} (1.0.0), \textit{SAVERX} (1.0.0), \textit{scImpute} (0.0.9), \textit{ALRA} (commit 7636de8), \textit{DCA} (0.2.2), \textit{DrImpute} (1.2), \textit{ENHANCE} (R version, commit 1571696), \textitt{SAVERX} (1.0.0).
+Analyses were performed in R 3.6.0 (Bioconductor 3.9) and the following packages were installed from GitHub repositories: \textit{Seurat} (version 3.0.0), \textit{sctransform} (0.2.0), \textit{DoubletFinder} (2.0.1), \textit{scDblFinder} (1.1.1), \textit{scds} (1.0.0), \textit{SAVERX} (1.0.0), \textit{scImpute} (0.0.9), \textit{ALRA} (commit 7636de8), \textit{DCA} (0.2.2), \textit{DrImpute} (1.2), \textit{ENHANCE} (R version, commit 1571696), \textit{SAVERX} (1.0.0).
 The code for the glmPCA was obtained from \url{https://github.com/willtownes/scrna2019} (commit 1ddcc30ebb95d083a685f12fe81d35dd1b0cb1b2).
@@
 Both RNA and ADT count data were downloaded from GEO and processed independently using \textit{Seurat}.
-We then considered cells that were in the same cluster both in the RNA-based and ADT-based analyses to be real subpopulations, and focused on the 4 most abundant ones.
-We then performed 3 sampling-based simulations with various degrees of separation using \textit{muscat} \citep{CrowellMuscat2019}, and merged the three simulations.
+We then considered cells that were in the same cluster both in the RNA-based and ADT-based analyses to be real subpopulations and focused on the 4 most abundant ones.
+We then performed 3 sampling-based simulations with various degrees of separation using \textit{muscat} \cite{CrowellMuscat2019} and merged the three simulations.
 The simMix2 dataset was generated from the mouse brain data published in \cite{CrowellMuscat2019} in a similar fashion.
@@
     \item standard \textit{Seurat} normalization was employed
-    \item \textit{Seurat} (v3) variable feature selection was employed, selecting 2000 genes
+    \item \textit{Seurat} ($\geq$3.0) variable feature selection was employed, selecting 2000 genes
     \item standard \textit{Seurat} scaling and PCA were employed
@@
 
-\subsubsection*{Denoising/imputation}
+\subsection*{Denoising/imputation}
 Most imputation/denoising methods were run with the default parameters with the following exceptions; \textit{DrImpute} documentation advises to process the data prior to imputation by removing lowly expressed genes and cells expressing less than 2 genes.
 As it is not clear if this step is a hard requirement for the method to perform well, \textit{DrImpute} was run with and without prior processing (\textit{DrImpute\_process} and \textit{DrImpute\_noprocess} labels, respectively).
-The \textit{dca} method accepts only integers, while two of the datasets had non-integer quantification of expected counts.
+The \textit{dca} method only accepts integers counts, while two of the datasets had non-integer quantification of expected counts.
 For these datasets, we rounded up the counts prior to imputation.
@@
 
+\subsection*{Dimensionality estimates}
+
+For the methods that produce local dimensionality estimates, we used the maximum.
+For the elbow method, we implemented an automatic procedure by taking the farthest point from a line drawn between the variance explained by the first and last (i.e.
+50th) components calculated.
+For the \texttt{JackStraw} method, since the \textit{Seurat} documentation advises not to use a p-value threshold (which would typically yield a very large number of dimensions) but rather look for a drop in significance, we applied the same farthest point algorithm on the log10(p-values), which in our hands reproduced manual threshold selection.
+
 \subsection*{Doublet detection method}
@@
 Briefly, after reducing the data to the most expressed genes, we cluster the cells using the fast-greedy algorithm, favouring overclustering.
-We then create artificial doublets by sampling the two cells specifically from different clusters, and sum the counts of each pair of cells.
-We also create meta-cells from each cluster, and use them to create additional doublets and triplets.
+We then create artificial doublets by sampling the two cells specifically from different clusters and sum the counts of each pair of cells.
+We also create meta-cells from each cluster and use them to create additional doublets and triplets.
 We combine them with the real cells, perform PCA and build a KNN graph using \textit{BiocNeighbors}.
@@
 Unless specified otherwise, the variance in gene expression explained by subpopulations was calculated on the data normalized and transformed through \textit{sctransform}.
-For each gene, we fitted a linear model using the subpopulation as only independent variable (\textasciitilde subpopulation), and used the R-squared as a measure of the variance explained.
+For each gene, we fitted a linear model using the subpopulation as only independent variable ($\sim subpopulation$) and used the R-squared as a measure of the variance explained.
 The same method was used for principal components.
 
-The deviance explained by subpopulations was calculated directly on counts using the \texttt{pipeComp::getDevianceExplained} function.
-The function uses \textit{edgeR} to fit two models, a full \textasciitilde librarySize+subpopulation model and a reduced \textasciitilde librarySize model.
+The deviance explained by subpopulations was calculated directly on counts using the \linebreak \texttt{getDevianceExplained} function of \texttt{pipeComp}.
+The function uses \textit{edgeR} to fit two models, a full ($\sim librarySize + population$) and a reduced one ($\sim librarySize$).
 For each gene, the deviance explained is then the difference between the deviance of each models, divided by the deviance of the reduced model.
@@
 
-To estimate the correlation between the principal components and covariates such as library size, we first fitted a linear model on the subpopulations, and correlated the residuals of this model with the covariate of interest.
+To estimate the correlation between the principal components and covariates such as library size, we first fitted a linear model on the subpopulations and correlated the residuals of this model with the covariate of interest.
+
+\subsection*{\textit{scVI} normalization and LD}
+
+The evaluation of \textit{scVI} algorithm was performed in separate to the other pipeline combinations as the model training on our reference datasets was especially resource-consuming.
+As it was not practically feasible to perform the training step on many combination of parameters as it was done with other tools, we evaluated it on a representative subset of pipelines using \textit{sctransform} normalization and \textit{Seurat}'s PCA dimension reduction method.
+The resulting evaluation is available in the Supplementary Figure [TBD].
 
@@
 \section*{Acknowledgements}
-This work was supported by the Swiss National Science Foundation (grants 310030_175841, CRSII5_177208) as well as the Chan Zuckerberg Initiative DAF (grant number 2018-182828), an advised fund of Silicon Valley Community Foundation.
+This work was supported by the Swiss National Science Foundation (grants 310030\_175841, CRSII5\_177208) as well as the Chan Zuckerberg Initiative DAF (grant number 2018-182828), an advised fund of Silicon Valley Community Foundation.
 MDR acknowledges support from the University Research Priority Program Evolution in Action at the University of Zurich.
@@
   Kumar & GSE60749 & SMARTer & Mouse ESC cultured in different conditions \\ 
-  Zhengmix4eq & \citep{duoClustering2018} & 10x & Mixtures of FACS purified PBMCs \\ 
-  Zhengmix4uneq & \citep{duoClustering2018} & 10x & Mixtures of FACS purified PBMCs \\ 
-  Zhengmix8eq & \citep{duoClustering2018} & 10x & Mixtures of FACS purified PBMCs \\ 
+  Zhengmix4eq & \cite{duoClustering2018} & 10x & Mixtures of FACS purified PBMCs \\ 
+  Zhengmix4uneq & \cite{duoClustering2018} & 10x & Mixtures of FACS purified PBMCs \\ 
+  Zhengmix8eq & \cite{duoClustering2018} & 10x & Mixtures of FACS purified PBMCs \\ 
   mixology10x3cl & \cite{tianMixology2018} & 10x & Mixture of 3  cancer cell lines from CellBench \\ 

### 2020-02-27 draft.tex
Updates from Overleaf

@@
     \centering
-    \includegraphics[scale=0.75,keepaspectratio]{main_figures/pipeComp_summary.png}
-    \caption{\textbf{Recommendations of tools from filtering, to clustering.}}
+    \includegraphics[scale=0.08,keepaspectratio]{main_figures/pipeComp_summary.png}
+    \caption{\textbf{Recommendations of tools from filtering to clustering.}}
     \label{fig:figure11}

### 2020-02-27 supplementary_figures/supp1_dataset_description.Rmd
updated supplementary figures

@@
 source("../misc_functions.R")
-theme_set(theme_cowplot(font_size = 12))
 # we load the datasets
@@
 ```
-
-# Supplementary Figure `r getFigNb(TRUE)`
-
-```{r dist_cell_properties, fig.width=8.5, fig.height=8.5}
-vars <- c( "log10(features)"="log10_total_features", "log10(counts)"="log10_total_counts", 
-           "% in top 20"="pct_counts_in_top_20_features", "% mito"="pct_counts_Mt", 
-           "% ribosomal"="pct_counts_ribosomal", "% coding"="pct_counts_coding" )
-plot_grid(plotlist=lapply(1:length(vars), FUN=function(x){ 
-  p <- plotFilters(SCEs, vars[x], nmads=c(2,5), ncol=1) + xlab(names(vars)[x])
-  p <- p + scale_y_continuous(breaks=scales::pretty_breaks(2))
-  p <- p + scale_x_continuous(breaks=scales::pretty_breaks(3))
-  p + theme( axis.title.y = element_blank(), 
-             strip.text.x = element_text(size=10),
-             axis.title=element_text(size=11),
-             axis.text=element_text(size=11) )
-}), nrow=1)
-```
-
-### Supplementary Figure `r getFigNb()`
-Distribution across cells of various control properties in the different datasets. The 
-lines indicate respectively 2 and 5 median absolute deviations (MADs).
-
-\newpage
 
@@
 
-```{r featcount_ratio}
-d <- data.frame( dataset=rep(names(SCEs),sapply(SCEs,ncol)), 
-                 log10_counts=unlist(lapply(SCEs,FUN=function(x) x$log10_total_counts)),
-                 log10_features=unlist(lapply(SCEs,FUN=function(x) x$log10_total_features)),
-                 featcount_dist=unlist(lapply(SCEs,FUN=function(x) x$featcount_dist)))
-ggplot(d, aes(x=log10_features, y=log10_counts, colour=featcount_dist)) + geom_point() + facet_wrap(~dataset, scale="free") + scale_colour_gradient2(limits=c(-0.2,0.2))
+```{r dist_cell_properties, fig.width=8.5, fig.height=8.5}
+theme_set(theme_cowplot(font_size = 12))
+vars <- c( "log10(features)"="log10_total_features", "log10(counts)"="log10_total_counts", 
+           "% in top 20"="pct_counts_in_top_20_features", "% mito"="pct_counts_Mt", 
+           "% ribosomal"="pct_counts_ribosomal", "% coding"="pct_counts_coding" )
+plot_grid(plotlist=lapply(1:length(vars), FUN=function(x){ 
+  p <- plotFilters(SCEs, vars[x], nmads=c(2,5), ncol=1) + xlab(names(vars)[x])
+  p <- p + scale_y_continuous(breaks=scales::pretty_breaks(2))
+  p <- p + scale_x_continuous(breaks=scales::pretty_breaks(3))
+  p + theme( axis.title.y = element_blank(), 
+             strip.text.x = element_text(size=10),
+             axis.title=element_text(size=11),
+             axis.text=element_text(size=11) )
+}), nrow=1)
 ```
@@
 ### Supplementary Figure `r getFigNb()`
-There is a tight relationship, in 10x datasets (i.e.
-not the `Koh` and `Kumar` datasets), between the total counts of a cell and its number of detected features.
-We therefore include, among control variables, deviation from this ratio.
+Distribution across cells of various control properties in the different datasets. The 
+lines indicate respectively 2 and 5 median absolute deviations (MADs).
 
@@
            labels=LETTERS[1:2], nrow=2 )
+```
+
+### Supplementary Figure `r getFigNb()`
+Relationship between selected cell-level QC metrics.
+
+\newpage
+
+# Supplementary Figure `r getFigNb(TRUE)`
+
+```{r featcount_ratio}
+d <- data.frame( dataset=rep(names(SCEs),sapply(SCEs,ncol)), 
+                 log10_counts=unlist(lapply(SCEs,FUN=function(x) x$log10_total_counts)),
+                 log10_features=unlist(lapply(SCEs,FUN=function(x) x$log10_total_features)),
+                 featcount_dist=unlist(lapply(SCEs,FUN=function(x) x$featcount_dist)))
+ggplot(d, aes(x=log10_features, y=log10_counts, colour=featcount_dist)) + geom_point() + facet_wrap(~dataset, scale="free") + scale_colour_gradient2(limits=c(-0.2,0.2))
 ```

### 2020-02-27 supplementary_figures/supp1b_metrics.Rmd
updated supplementary figures

---
title: "Supplementary Figures relative to clustering"
author: "Pierre-Luc Germain"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r, include=FALSE}
if(!exists('FIG_NB')){
  FIG_NB <- 0
  getFigNb <- function(increment=FALSE){
    if(increment) FIG_NB <<- FIG_NB + 1
    FIG_NB
  }
}

suppressPackageStartupMessages({
  library(ggplot2)
  library(cowplot)
  library(ComplexHeatmap)
  library(reshape2)
  library(dplyr)
  library(patchwork)
  library(pipeComp) # (requires `pipeComp` version >=0.99.3)
  library(scales)
})
theme_set(theme_cowplot(font_size = 12))
source("../misc_functions.R")
```

# Supplementary Figure `r getFigNb(TRUE)`

```{r, include=FALSE}
res <- readRDS("../data/results_clustering.rds")
res <- res$evaluation$clustering
res <- res[which(res$clustmethod %in% c("clust.seurat","clust.scran")),]
res$norm <- gsub("norm\\.seuratvst","sctransform",res$norm)
dd <- res[!duplicated(res[,c("dataset","true.nbClusts")]),]
```

```{r, fig.width=7.5, fig.height=7}
ggplot(res, aes(n_clus, ARI, colour=clustmethod, shape=norm)) + 
    geom_point() + xlim(c(1,40)) + facet_wrap(~dataset) + xlab("Number of clusters") +
    geom_vline(aes(xintercept=k), data=dd, linetype="dashed") + theme(legend.position="bottom") + 
    guides(colour=guide_legend(ncol=1, title = ""), shape=guide_legend(ncol=1, title=""))
```

### Supplementary Figure `r getFigNb()`
The number of clusters called has a much bigger impact on the Adjusted Rand Index (ARI) than differences between methods. The dashed line indicates the true number of clusters.

# Supplementary Figure `r getFigNb(TRUE)`

```{r, fig.width=8, fig.height=5}
res <- readRDS("../data/results_clustering.rds")
cl <- res <- res$evaluation$clustering
cl <- cl[cl$dataset!="mixology10x3cl",]
cl$nbClust.diff <- cl$n_clus-cl$true.nbClusts
cl$nbClust.absDiff <- abs(cl$n_clus-cl$true.nbClusts)
ff <- c( "mean_pr", "mean_re", "mean_F1", "RI", "ARI", "MI", "AMI", "VI", "NVI", 
         "ID", "NID", "NMI", "min_pr", "min_re", "min_F1", "nbClust.diff",
         "nbClust.absDiff" )
e <- split(cl[,ff], cl$dataset, drop=TRUE)
sp <- lapply(e, FUN=function(x) cor(x,method="spearman"))
sp <- Reduce('+',sp)/length(sp)
pe <- lapply(e, FUN=function(x) cor(x))
pe <- Reduce('+',pe)/length(pe)
h <- hclust(as.dist(1-(sp+pe)/2))
sp <- sp[h$order,h$order]
pe <- pe[h$order,h$order]
cs <- circlize::colorRamp2(c(-1,0,1), c("darkblue","white","darkred"))
Heatmap( sp, column_title = "Spearman correlation", col=cs, name="cor", 
         cluster_columns=FALSE, cluster_rows=FALSE) + 
  Heatmap(pe, column_title="Spearman correlation", name="pearson", col=cs,
          cluster_columns=FALSE, cluster_rows=FALSE, show_heatmap_legend=FALSE)
```

### Supplementary Figure `r getFigNb()`
Relationship of various metrics of clustering accuracy between each other and with variations in the number of clusters called (`nbClust.diff` and `nbClust.absDiff`). Correlations were calculated for each dataset separately across various clustering runs and averaged (the `mixology10x3cl` dataset was excluded due to insufficient variation among the results). Information distance metrics (ID, NID, VI, NVI) are highly correlated with the absolute difference between the true and called number of clusters, while the Adjusted Rand Index (ARI) and similar metrics were strongly anticorrelated to it. Precision (mean_pr) and recall (mean_re) were slightly less correlated with discrepancies in the number of clusters. Mutual information (MI) was not at all correlated with the absolute difference in number of clusters (`nbClust.absDiff`), but positively correlated with the difference (`nbClust.diff`), i.e. favouring clusterings calling a higher number of clusters. We therefore recommend using complementary metrics such as ARI and MI, and potentially mean F1 per subpopulation.

\newpage

### 2020-02-27 supplementary_figures/supp2_filtering.Rmd
updated supplementary figures

@@
 res <- readRDS("../data/results_filtering.rds")
-m <- scrna_evalPlot_filtering(res, returnTable=TRUE)
+m <- old_evalPlot_filtering(res, returnTable=TRUE)
 m <- m[grep("veryStringent", m$method, invert=TRUE),]

### 2020-02-27 supplementary_figures/supp3_normalization.Rmd
updated supplementary figures

@@
 })
-suppressMessages(devtools::load_all("../../pipComp/"))
 theme_set(theme_cowplot(font_size = 11))
@@
 A rough ANOVA on `nbClusters~dataset+norm+dims+resolution` suggests that seuratvst (sctransform) is associated with a higher number of clusters (p~`r round(mod$coefficients["normnorm.seuratvst",4],3)`).
+
+\newpage
+
+# Supplementary Figure `r getFigNb(TRUE)`
+
+```{r VST, echo=FALSE, out.width = '100%'}
+knitr::include_graphics("vst.png")
+```
+
+### Supplementary Figure `r getFigNb()`
+Relationship of the variance with mean count after `sctranform`'s variance stabilizing transformation.
+
+\newpage

### 2020-02-27 supplementary_figures/supp3b_scVI.Rmd
updated supplementary figures

---
title: "Supplementary Figures relative to scVI"
author: "Pierre-Luc Germain"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r, include=FALSE}
if(!exists('FIG_NB')){
  FIG_NB <- 0
  getFigNb <- function(increment=FALSE){
    if(increment) FIG_NB <<- FIG_NB + 1
    FIG_NB
  }
}

suppressPackageStartupMessages({
  library(ComplexHeatmap)
  library(matrixStats)
  library(pipeComp)
  library(cowplot)
})
source("../misc_functions.R")
```

# Supplementary Figure `r getFigNb(TRUE)`

```{r}
res <- readRDS("../data/scVI.rds")
sil <- res$evaluation$dimreduction$silhouette$top_10_dims
sil$method <- factor(paste(sil$norm, sil$dr))
levels(sil$method) <- c("scVI.LD","scVI+seurat.pca","scVI.LD","sctransform+\nseurat.pca")
levels(sil$dataset) <- c("Koh","Kumar","mixology10x3cl","mixology10x5cl","simMix1","simMix2","Zhengmix4eq","Zhengmix4uneq","Zhengmix8eq")
res$evaluation$dimreduction$silhouette$top_10_dims <- sil
x <- range(res$evaluation$dimreduction$silhouette$top_10_dims$meanSilWidth)
silColBreaks <- unique(round(c(x[1],0,x[2]/2,x[2]),1))
H <- scrna_evalPlot_DR(res, "meanSilWidth", scale=FALSE, agg.by = "method", 
                       reorder_rows=TRUE, value_format="", 
                       show_heatmap_legend=TRUE, anno_legend = TRUE, 
                       heatmap_legend_param=list(at=silColBreaks))

res$evaluation$clustering$method <- factor(paste(res$evaluation$clustering$norm, res$evaluation$clustering$dr))
res$evaluation$clustering$delta.nbClust <- res$evaluation$clustering$n_clus-res$evaluation$clustering$true.nbClusts
levels(res$evaluation$clustering$method) <- c("scVI.LD","scVI+seurat.pca","scVI.LD","sctransform+\nseurat.pca")
H2 <- scrna_evalPlot_clust(res, what="MI", agg.by="method", anno_legend=FALSE) + 
  scrna_evalPlot_clust(res, what="ARI", agg.by="method", anno_legend=FALSE) + 
  scrna_evalPlot_clust(res, what="delta.nbClust", agg.by="method", col=circlize::colorRamp2(c(-6, 0, 6), c("red", "white", "Darkblue")), show_heatmap_legend = FALSE, value_cols = c("black","black"), value_format = "%.0f", anno_legend=FALSE)
```

```{r, fig.width=9, fig.height=3.5}
plot_grid(grid.grabExpr(draw(H)), grid.grabExpr(draw(H2)), ncol=1, labels = LETTERS[1:2])
```

### Supplementary Figure `r getFigNb()`

**scVI evaluation. A:** Average silhouette width per subpopulation using sctransform and standard Seurat PCA, scVI normalization followed by Seurat PCA, or scVI linear decoder (LD). **B:** Clustering accuracy across the same methods followed by Seurat clustering.

\newpage

### 2020-02-27 supplementary_figures/supp4_selection.Rmd
updated supplementary figures

@@
   })
-  bind_rows(y, .id="dataset")
+  dplyr::bind_rows(y, .id="dataset")
 }

### 2020-02-27 supplementary_figures/supp6_clust.Rmd
updated supplementary figures

--- 2020-01-29 supplementary_figures/supp6_clust.Rmd
@@
   library(patchwork)
+  library(pipeComp) # (requires `pipeComp` version >=0.99.3)
+  library(scales)
 })
 theme_set(theme_cowplot(font_size = 12))
-devtools::load_all("../../pipeComp/")
 source("../misc_functions.R")
@@
 ```{r, include=FALSE}
-#res <- readRDS("../data/clust_endSummary.rds")
-res <- readRDS("../data/results_clust_2.rds")
-res <- parsePipNames(res$clustering)
+res <- readRDS("../data/results_clustering2.rds")
+res <- res$evaluation$clustering
 res <- res[which(res$norm=="norm.seurat" & res$resolution <5),]
-x <- grep("true\\.nbClusts",colnames(res),value=TRUE)
-names(x) <- x <- sapply(strsplit(x," "),FUN=function(x) x[[1]])
-y <- cbind(res[,c("k","steps","resolution","clustmethod","dims")],sapply(x, FUN=function(x){
-  res[[paste(x,"n_clus")]] - res[[paste(x,"true.nbClusts")]]
-}))
-y$param <- ifelse( res$clustmethod=="clust.seurat", paste0(" res=",res$resolution), paste0(" steps=",res$steps))
-ag <- aggregate(y[,names(x)], by=y[,c("clustmethod", "k","param"),drop=FALSE], FUN=mean)
-agl <- cbind(ag[rep(1:nrow(ag), 9),1:3], dataset=rep(colnames(ag)[4:ncol(ag)], each=nrow(ag)), delta=as.numeric(as.matrix(ag[,4:ncol(ag)])))
+res$delta <- res$n_clus-res$true.nbClusts
+res$param <- ifelse( res$clustmethod=="clust.seurat", paste0(" res=",res$resolution), paste0(" steps=",res$steps))
+res$param[which(res$clustmethod=="clust.scran.fg")] <- "NA"
+agl <- aggregate(res[,"delta",drop=FALSE], by=res[,c("dataset","clustmethod", "k","param"),drop=FALSE], FUN=mean)
 agl$k <- factor(agl$k, as.character(sort(unique(agl$k))))
-x <- sapply(unique(strsplit(ag$param, "=")), FUN=function(x) as.numeric(x[[2]]) )
-names(x) <- unique(ag$param)
+x <- sapply(unique(strsplit(agl$param, "=")), FUN=function(x){
+  tryCatch(as.numeric(x[[2]]), error=function(e) NA_real_)
+})
+names(x) <- unique(agl$param)
 agl$param <- factor(agl$param, names(sort(x)))
@@
 ```{r, fig.width=8, fig.height=8}
-(pf(agl[which(agl$clustmethod=="clust.seurat"),]) + ylab("Seurat") + theme(legend.position = "top")) / 
-    (pf(agl[which(agl$clustmethod=="clust.scran"),]) + ylab("scran walktrap")) /
-    (pf(agl[which(agl$clustmethod=="clust.scran.fg"),]) + ylab("scran fast_greedy")) /
+tt <- theme(strip.background = element_blank(), strip.text.x = element_blank())
+p1 <- pf(agl[which(agl$clustmethod=="clust.seurat"),]) + ylab("Seurat") + 
+  theme(legend.position = "top", strip.text.x = element_text(size = 10, angle=60), strip.background = element_blank())
+p1 / 
+    (pf(agl[which(agl$clustmethod=="clust.scran"),]) + ylab("scran walktrap") + tt) /
+    (pf(agl[which(agl$clustmethod=="clust.scran.fg"),]) + ylab("scran fast_greedy") + tt) /
     (ggplot(agl, aes(k)) + theme(axis.line.y=element_blank(), plot.margin = unit(rep(0,4),"pt"))) + 
-    plot_layout(heights = c(10,8,8,1))
+    plot_layout(heights = c(10,6,2,1))
 ```

### 2020-02-27 main_figures/clustering.Rmd
updated clustering results

@@
   library(ComplexHeatmap)
-  library
   library(pipeComp)
@@
 
-```{r clustering, fig.width=12, fig.height=3}
-res <- readRDS("../data/results_clust_2.rds")
-h <- scrna_evalPlot_clust(res, what="MI", agg.by=c("norm","clustmethod"), scale=TRUE, anno_legend = FALSE)
-h <- h + scrna_evalPlot_clust(res, what="mean_F1", agg.by=c("norm","clustmethod"), scale=TRUE, reorder_rows = h, anno_legend = FALSE) +
-  scrna_evalPlot_clust(res, what="min_pr", agg.by=c("norm","clustmethod"), scale=TRUE, reorder_rows = h, anno_legend = FALSE) +
-  scrna_evalPlot_clustAtTrueK(res, agg.by=c("norm","clustmethod"), reorder_rows = h, anno_legend = FALSE)
-h <- .renameHrows(h, function(x) gsub("seurat; clustmethod=|; clustmethod=", "", gsub("seuratvst", "sctransform + ", x)))
-h <- .renameHrows(h, function(x) gsub("clust.scran","scran (walktrap)",gsub("clust.scran.fg", "scran (fast_greedy)", x)))
-h
+```{r clustering_accuracy, fig.width=12, fig.height=3}
+# biorxiv data & figure
+# res <- readRDS("../data/results_clust_2.rds")
+# h <- scrna_evalPlot_clust(res, what="MI", agg.by=c("norm","clustmethod"), scale=TRUE, anno_legend = FALSE)
+# h <- h + scrna_evalPlot_clust(res, what="mean_F1", agg.by=c("norm","clustmethod"), scale=TRUE, reorder_rows = h, anno_legend = FALSE) +
+#   scrna_evalPlot_clust(res, what="min_pr", agg.by=c("norm","clustmethod"), scale=TRUE, reorder_rows = h, anno_legend = FALSE) +
+#   scrna_evalPlot_clustAtTrueK(res, agg.by=c("norm","clustmethod"), reorder_rows = h, anno_legend = FALSE)
+# h <- pipeComp:::.renameHrows(h, function(x) gsub("seurat; clustmethod=|; clustmethod=", "", gsub("seuratvst", "sctransform + ", x)))
+# h <- pipeComp:::.renameHrows(h, function(x) gsub("clust.scran","scran (walktrap)",gsub("clust.scran.fg", "scran (fast_greedy)", x)))
+
+res <- readRDS("../data/results_clustering2.rds")
+cl <- res$evaluation$clustering
+cl$graph <- cl$clustmethod
+levels(cl$graph) <- c(paste0("scran (",c("exact rank SNN", "Annoy rank SNN", "exact rank SNN", "exact KNN", "Annoy KNN", "exact SNN", "Annoy SNN"),")"), "Seurat SNN")
+levels(cl$clustmethod) <- c("walktrap","walktrap","fast-greedy","walktrap","walktrap","walktrap","walktrap","Seurat")
+cl$elapsed <- res$elapsed$stepwise$clustering$elapsed
+cl$nbClust.diff <- cl$n_clus-cl$true.nbClusts
+clo <- cl
+cl <- cl[cl$k<=20,]
+res$evaluation$clustering <- cl
+
+renameFn <- function(x){
+  x <- gsub("seurat; graph=|; graph=", "", gsub("seuratvst", "sctrans.+", x))
+  x <- gsub("; clustmethod=", "+", x)
+  x <- gsub("^graph=","",x)
+  x <- gsub("exact ","",x)
+  x <- gsub(" + walktrap","",x,fixed=TRUE)
+}
+
+tmp <- scale(pipeComp:::.prepRes(cl, "ARI", agg.by=c("norm","graph","clustmethod"))) + 
+  scale(pipeComp:::.prepRes(cl, "MI", agg.by=c("norm","graph","clustmethod")))
+o <- names(sort(rowMeans(tmp, na.rm=TRUE)+apply(tmp,1,na.rm=TRUE,FUN=median), decreasing = TRUE))
+o <- gsub("norm=norm\\.","",o)
+o <- c(o[grep("seuratvst",o)],o[grep("seuratvst",o,invert=TRUE)])
+h <- scrna_evalPlot_clust(cl, what="MI", agg.by=c("norm","graph","clustmethod"), agg.fn = median, scale=TRUE, anno_legend = FALSE, reorder_rows=o, row_names_max_width = unit(8, "cm"), row_names_gp = gpar(fontsize = 11)) +
+  scrna_evalPlot_clust(res, what="ARI", agg.by=c("norm","graph","clustmethod"), agg.fn = median, scale=TRUE, reorder_rows = o, anno_legend = FALSE) +
+  scrna_evalPlot_clust(res, what="min_pr", agg.by=c("norm","graph","clustmethod"), agg.fn = median, scale=TRUE, reorder_rows = o, anno_legend = FALSE) +
+  scrna_evalPlot_clust(res, what="ARI", atTrueK = TRUE, agg.by=c("norm","graph","clustmethod"), agg.fn = median, reorder_rows = o, anno_legend = FALSE, row_names_max_width = unit(8, "cm"), row_names_gp = gpar(fontsize = 11))
+h <- pipeComp:::.renameHrows(h, renameFn)
+h2 <- grid.grabExpr(draw(h))
 ```
+
+```{r clustering_time}
+clo$elapsed <- res$elapsed$stepwise$clustering$elapsed
+# keep reasonable parameters for a realistic time
+res$evaluation$clustering <- clo[which(clo$steps>2 & clo$k <=20),]
+h3 <- scrna_evalPlot_clust(res, what="elapsed", agg.by=c("graph","clustmethod"),
+                           anno_legend = TRUE, scale=TRUE, value_format = "%.1f" )
+h3 <- h3 + scrna_evalPlot_clust(res, what="nbClust.diff", 
+                                agg.by=c("graph","clustmethod"), agg.fn = median,
+                                reorder_rows=h3, scale = FALSE, value_format = "%.1f", 
+                                col = circlize::colorRamp2(c(-5, 0, 5), c("red", "white", "Darkblue")),
+                                value_cols = c("black","black"), 
+                                row_names_max_width = unit(8, "cm"),
+                                row_names_gp = gpar(fontsize = 12) )
+h3 <- pipeComp:::.renameHrows(h3, renameFn)
+h3 <- grid.grabExpr(draw(h3))
+```
+
+```{r clustering, fig.width=12, fig.height=8}
+plot_grid(h2, h3, labels=LETTERS[1:2], ncol=1, rel_heights = c(7,4))
+```
+Clustering accuracy (A), computing time (B left) and mean deviation from the true number of clusters (B right).
+The (exact) rank-based SNN graph represents the default setting in `scran`-based clustering; however we find that the Annoy approximation to the KNN diminishes the computing costs without negatively affecting the accuracy.

### 2020-02-27 misc_functions.R
added old functions to plot old filtering results

@@ -77,6 +77,13 @@ reduceMethods <- function(x, pos=1:2, FUN=mean, sep=" "){
   row.names(x) <- x[,1]
   x[,-1]
 }
+renameNorm <- function(x){
+  pipeComp:::.renameHrows(x, f=function(x){
+    x <- gsub("^norm\\.", "", x)
+    x <- gsub("seuratvst","sctransform",x)
+    x <- gsub(".noscale", " (no scaling)", x)
+  })
+}
 
 cast2 <- function(x, fun.aggregate=median, ...){
   x <- reshape2::dcast(x, fun.aggregate=fun.aggregate, na.rm=T, ...)
@@ -184,4 +191,142 @@ plotBadCells <- function(sce, mis, return_list=FALSE, aspect.ratio=NULL, name=""
   )
   if(return_list) return(pl)
   plot_grid(plotlist = pl, labels=c("",labels[3:7]))
+}
+
+
+
+#' cumVE2
+#'
+#' Plots the proportion of the cumulative variance/deviance explained by clusters, according 
+#' to various gene rankings.
+#'
+#' @param preds A list of names variability measurements, or a data.frame of gene metadata.
+#' @param ve The variance explained by each gene. If NULL, will attempt to read "vst.varExp"
+#' from `preds`.
+#' @param ve.name The name of the measure, default "variance"
+#'
+#' @return a ggplot
+#' @export
+cumVE2 <- function(preds=NULL, ve=NULL, ve.name="variance"){
+  if(is.data.frame(preds) | is(preds,"DataFrame")){
+    ## assume output of `varExplainedByClusters`
+    if(is.null(ve)) ve <- preds[["vst.varExp"]]
+    preds <- list( deviance=preds$deviance,
+                   #    ineq=preds$mean.ineq,
+                   seurat.disp=preds$seurat.dispersion,
+                   seurat.disp.std=preds$seurat.dispersion.scaled,
+                   seurat.vst=preds$seurat.variance,
+                   seurat.vst.std=preds$seurat.variance.standardized,
+                   seurat.trans=preds$seurat.res.var,
+                   expr=preds$total_counts
+    )
+    preds <- lapply(preds, as.numeric)
+  }else{
[…]

### 2020-02-27 bmc_article.tex

@@
 \kwd{clustering}
+\kwd{normalization}
+\kwd{dimensionality}
 \kwd{filtering}
@@
 
-A number of good comparison and benchmark studies have already been performed on various steps related to scRNAseq analysis and can guide the choice of methodology (\citealp{CobosDeconvolution2020, colePerformance2018, DalMolinDE2017, duoClustering2018, freytagComparison2018, GaoProcessing2020, HeiserDimred2019, HouImput2020, JaakkolaDE2017, SonesonDE2018, SunDimRed2019, tianMixology2018, TranBatch2020, TsuyuzakiPCA2020, viethSystematic2019, WangDE2019, YipHVGs2018, ZhangImput2018}).
+A number of good comparison and benchmark studies have already been performed on various steps related to scRNAseq processing and analysis and can guide the choice of methodology (\citealp{CobosDeconvolution2020, colePerformance2018, DalMolinDE2017, duoClustering2018, freytagComparison2018, GaoProcessing2020, HeiserDimred2019, HouImput2020, JaakkolaDE2017, KrzakClust2019, SonesonDE2018, SunDimRed2019, tianMixology2018, TranBatch2020, TsuyuzakiPCA2020, viethSystematic2019, WangDE2019, YipHVGs2018, ZhangImput2018}).
 However these recommendations need constant updating and often leave open many details of an analysis.
@@
 This is especially important because end-point metrics, such as the adjusted Rand index (ARI) \cite{HubertARI1985} for clustering, are not perfect.
-For example, although the meaning of an ARI score is independent of the number of true subpopulations \cite{steinleyProperties2004}, the number of clusters called is by far the most important determinant of the score: the farther it is from the actual number of subpopulations, the worse the ARI (Supplementary Figure 17).
+For example, although the meaning of an ARI score is independent of the number of true subpopulations \cite{steinleyProperties2004}, the number of clusters called is by far the most important determinant of the score: the farther it is from the actual number of subpopulations, the worse the ARI (Supplementary Figures 1-2).
 In this context, one strategy has been to cluster across various resolutions and only consider the results that have the right number of clusters \cite{duoClustering2018}.
@@
     \centering
-    \includegraphics[width=\textwidth,keepaspectratio]{main_figures/pipeline_explanation.png}
+    \includegraphics[scale=0.7, keepaspectratio]{main_figures/pipeline_explanation.png}
     \caption{\textbf{Overview of the \textit{pipeComp} framework and its application to a scRNAseq clustering pipeline.
 A:} The package is built around a \texttt{PipelineDefinition} S4 class which defines a set of functions to be executed consecutively, as well as optional evaluation functions for each step.
-Each step can accept of number of parameters whose alternative values are provided as a list.
+Each step can accept a number of parameters whose alternative values are provided as a list.
 All (or subsets of) combinations of parameters can then be simultaneously ran and evaluated using the \texttt{runPipeline} function.
@@
 While doublets of the same cell type are relatively innocuous for most downstream applications due to their conservation of the relative expression between genes, doublets formed from different cell types or states are likely to be misclassified and could potentially distort downstream analysis.
-In some cases, doublets can be identified through their unusually high number of reads and detected features, but this is not always the case (Supplementary Figure 2).
+In some cases, doublets can be identified through their unusually high number of reads and detected features, but this is not always the case (Supplementary Figure 3).
 A number of methods were developed to identify doublets, in most cases by comparing each cell to artificially-created doublets \cite{mcginnisDoubletfinder2019, LunScran2016, BaisScds2019}.
@@
 In addition to these methods, we reasoned that an approach such as \textit{DoubletFinder} could be simplified by being applied directly on counts and by using a pre-clustering to create neotypic/heterotypic doublets more efficiently.
-We therefore developed a simple and fast Bioconductor package implementing this method for doublet detection, \textit{scDblFinder}, with the added advantage of accounting for uncertainty in the expected doublet rate and using meta-cells from the clusters to even include triplets (see Methods).
+We therefore also evaluated a simple and fast Bioconductor package implementing this method for doublet detection, \textit{scDblFinder}, with the added advantage of accounting for uncertainty in the expected doublet rate and using meta-cells from the clusters to even include triplets (see Methods).
 
@@
 Dots indicate threshold determined by the true number of doublets.
-\textbf{B:} Running time of the different methods (DoubletFinder failed on one of the datasets).
+\textbf{B:} Running time of the different methods (\textit{DoubletFinder} failed on one of the datasets).
 \textbf{C:} Rate of misclassification of the cells identified by \textit{scDblFinder} as doublets (DBL) or singlets (SNG), across a large range of clustering analysis.
@@
 This can for instance be performed through the \texttt{isOutlier} function from \textit{scater} that measures, for a given control property, the number of median absolute deviations (MADs) of each cell from the median of all cells.
-Supplementary Figure 1 shows the distributions of some of the typical cell properties commonly used.
+Supplementary Figure 4 shows the distributions of some of the typical cell properties commonly used.
 Of note, these properties tend to be correlated, with some exceptions.
-For example, while a high proportion of mitochondrial reads is often correlated with a high proportion of the counts in the top features, there can also be other reasons for an over-representation of highly-expressed features (Supplementary Figure 3), such as an over-amplification in non-UMI datasets.
-In our experience, 10X datasets also exhibit a very strong correlation between the total counts and the total features even across very different cell types (Supplementary Figure 4).
+For example, while a high proportion of mitochondrial reads is often correlated with a high proportion of the counts in the top features, there can also be other reasons for an over-representation of highly-expressed features (Supplementary Figure 5), such as an over-amplification in non-UMI datasets.
+In our experience, 10X datasets also exhibit a very strong correlation between the total counts and the total features even across very different cell types (Supplementary Figure 6).
 We therefore also measure the ratio between the two and treat cells strongly departing from this trend with suspicion.
 
-Reasoning that the cells we wish to exclude are the cells that would be misclassified, we measured the rate of misclassification of each cell in each dataset across a variety of clustering pipelines, correcting for the median misclassification rate of the subpopulation and then evaluated what properties could be predictive of misclassification (Supplementary Figures 5-7).
+Reasoning that the cells we wish to exclude are the cells that would be misclassified, we measured the rate of misclassification of each cell in each dataset across a variety of clustering pipelines, correcting for the median misclassification rate of the subpopulation and then evaluated what properties could be predictive of misclassification (Supplementary Figures 7-9).
 We could not identify any property or simple combination thereof that would be consistently predictive of misclassification; the only feature that consistently stood out across multiple datasets (the Zheng datasets) was that cells with very high read counts have a higher chance of being misclassified.
@@
 As a result, thresholds in terms of number of MADs from the whole distribution can lead to strong biases against certain subpopulations (Figure \ref{fig:figure4}A).
-We therefore examined the tradeoff between the increased accuracy of filtering and the maximum proportion of cells excluded per subpopulation (Figure \ref{fig:figure4}B and Supplementary Figure 8).
+We therefore examined the tradeoff between the increased accuracy of filtering and the maximum proportion of cells excluded per subpopulation (Figure \ref{fig:figure4}B and Supplementary Figure 10).
 Since filtering changes the relative abundance of the different subpopulations, global clustering accuracy metrics such as ARI are not appropriate here.
@@
     \includegraphics[scale=0.60, keepaspectratio]{{main_figures/filtering_files/figure-html/figure2-1.png}}
-    \caption{\textbf{A:} Filtering on the basis of distance to the whole distribution can lead to strong bias against certain subpopulations.
+    \caption{\textbf{Effect of filtering on cell subpopulation structure and clustering.
+A:} Filtering on the basis of distance to the whole distribution can lead to strong bias against certain subpopulations.
 The dashed line indicates a threshold of 2.5 median absolute deviations (MADs) from the median of the overall population.
@@
 Mitochondrial reads have been associated with cell degradation and there is evidence that ribosomal genes can influence the clustering output, hiding other biological structure in the analysis \cite{freytagComparison2018}.
-We therefore investigated whether excluding one category of features or the other, or using only protein-coding genes, had an impact on the ability to distinguish subpopulations (Supplementary Figure 9).
+We therefore investigated whether excluding one category of features or the other, or using only protein-coding genes, had an impact on the ability to distinguish subpopulations (Supplementary Figure 11).
 Removal of ribosomal genes robustly reduced the quality of the clustering, suggesting that they represent real biological differences between subpopulations.
@@
 We found most methods (including no normalization at all) to perform fairly well in most of the subpopulations.
-An exception was \textit{scVI} (Supplementary Figure TBD) that performed poorly on Zhengmix and simMix1 datasets in view of the subpopulation silhouette and to a relative poor ARI.
+An exception was \textit{scVI} (Supplementary Figure 12) which performed poorly on most datasets in terms of both the subpopulation silhouette and clustering accuracy.
 Scaling tended to reduce the average silhouette width of some subpopulations and to increase that of some less distinguishable ones and was generally, but not always, beneficial on the accuracy of the final clustering.
@@
     \centering
-    \includegraphics[width=\textwidth]{{main_figures/normalization_files/figure-html/figure3-1.png}}
+    \includegraphics[scale=0.55, keepaspectratio]{{main_figures/normalization_files/figure-html/figure3-1.png}}
     \caption{\textbf{Evaluation of normalization procedures.} `Seurat.feat\_mt\_regress' regresses out the number of features and proportion of mitochondrial reads during scaling; `Seurat.feat\_regress` regresses out the number of features only.
@@
 \textbf{B:} Average silhouette width per true subpopulation, where higher silhouette width means a higher separability.
-\textbf{C:} Clustering accuracy (\textit{Seurat} clustering), measured by the ARI at the true number of cluster (left) and by the average mutual information (MI) of the cluster assignment with true subpopulations.}
+\textbf{C:} Clustering accuracy (\textit{Seurat} clustering), measured by the ARI at the true number of cluster (left) and by the average MI of the cluster assignment with true subpopulations.}
     \label{fig:figure5}
@@
 Finally, we monitored whether, under the same downstream clustering analysis, different normalization methods tended to lead to an over- or under-estimation of the number of clusters.
-Although some methods had a tendency to lead to a higher (e.g., \textit{sctransform}) or lower (e.g., stable genes) number of clusters, the effect was very mild and not entirely systematic (Supplementary Figure 10).
+Although some methods had a tendency to lead to a higher (e.g., \textit{sctransform}) or lower (e.g., stable genes) number of clusters, the effect was very mild and not entirely systematic (Supplementary Figure 13).
+We also confirmed that, especially in the UMI-based datasets, \textit{sctransform} did in fact successfully stabilize variance across mean counts (Supplementary Figure 14).
 
@@
 As the proportion of variability in a gene attributable to subpopulations can be measured in various ways, we first compared three approaches: ANOVA on log-normalized count, ANOVA on \textit{sctransform} normalization, and deviance explained.
-The ANOVAs performed on a standard \textit{Seurat} normalization and on \textit{sctransform} data were highly correlated (Supplementary Figure 11A).
-These estimates were also in good agreement with the deviance explained, although lowly-expressed genes could have a high deviance explained without having much of their variance explained by subpopulation (Supplementary Figure 11B-D).
-We therefore compared the proportion of the cumulative variance/deviance explained by the top X genes that could be retrieved by each gene ranking method (Supplementary Figures 12-13).
-We first focused on the first 1000 genes to highlight the differences between methods, although a higher number of selected genes decreased the differences between methods (Supplementary Figures 12-14).
+The ANOVAs performed on a standard \textit{Seurat} normalization and on \textit{sctransform} data were highly correlated (Supplementary Figure 15A).
+These estimates were also in good agreement with the deviance explained, although lowly-expressed genes could have a high deviance explained without having much of their variance explained by subpopulation (Supplementary Figure 15B-D).
+We therefore compared the proportion of the cumulative variance/deviance explained by the top X genes that could be retrieved by each gene ranking method (Supplementary Figures 16-17).
+We first focused on the first 1000 genes to highlight the differences between methods, although a higher number of selected genes decreased the differences between methods (Supplementary Figures 16-18).
 The standardized measures of variability were systematically worse than their non-standardized counterparts in selecting genes with a high proportion of variance explained by subpopulation.
-Regarding the percentage of deviance explained however, the standardized measures were often superior (Figure \ref{fig:figure6}A and Supplementary Figures 12-13).
+Regarding the percentage of deviance explained however, the standardized measures were often superior (Figure \ref{fig:figure6}A and Supplementary Figures 16-17).
 Deviance proved the method of choice to prioritize genes with a high variance explained by subpopulations (with mere expression level proving surprisingly good) but did not perform so well to select genes with a high deviance explained.
@@
     \centering
-    \includegraphics[scale=0.6]{{main_figures/selection_files/figure-html/figure4-1.png}}
+    \includegraphics[scale=0.65]{{main_figures/selection_files/figure-html/figure4-1.png}}
     \caption{\textbf{Evaluation of feature selection methods.
 A:} Ability of different feature ranking methods to capture genes with a high proportion of variance (left) or deviance (right) explained by real subpopulations.
-The asterix denotes the default Seurat method.
+The asterisk denotes the default Seurat method.
 \textbf{B:} Accuracy of clusterings (at the true number of clusters) when selecting 1000 genes using the given methods.
@@
 In general, we found deviance and unstandardized estimates of variance to provide the best results across datasets and normalization methods.
-Increasing the number of features selected also systematically led to an increase in the accuracy of the clustering, typically plateauing after 4000 features (Supplementary Figure 14).
+Increasing the number of features selected also systematically led to an increase in the accuracy of the clustering, typically plateauing after 4000 features (Supplementary Figure 18).
 
@@
 
-The impact of the choice of dimensionality reduction method was far greater than that of normalization or feature selection (e.g., Supplementary Figure 15).
-GLM-PCA tended to increase the average silhouette width of already well-defined subpopulations, but \textit{Seurat}'s PCA procedure however proved superior on all metrics (Figure \ref{fig:figure7}).
+The impact of the choice of dimensionality reduction method was far greater than that of normalization or feature selection (Figure \ref{fig:figure7}).
+GLM-PCA tended to increase the average silhouette width of already well-defined subpopulations, but \textit{Seurat}'s PCA procedure however proved superior on all metrics.
+Like GLM-PCA, \textit{scVI}'s Linear Decoder (LD) does not explicitly rely on normalized counts data.
+Its performance was lower than a pipeline with non-weighted \textit{Seurat} PCA on all measures for datasets with more subpopulations (Supplementary Figure 12).
 Overall, weighting the principal components by their variance (as \textit{Seurat} does) had a positive impact on silhouette widths and ARI scores.
@@
 Since Euclidean distance decreases as the number of non-discriminating dimensions increases, there is usually a trade-off between selecting enough dimensions to keep most information and excluding smaller dimensions that may represent technical noise or other unwanted sources of variation.
-Overall, increasing the number of dimensions robustly led to a decrease in the number of clusters (Supplementary Figures 10 and 15).
-This tended to affect the accuracy of the clustering (Supplementary Figure 16), although in both cases (number of clusters and ARI) \textit{Seurat}'s resolution parameter had a much stronger impact.
+Overall, increasing the number of dimensions robustly led to a decrease in the number of clusters (Supplementary Figures 13 and 19).
+This tended to affect the accuracy of the clustering (Supplementary Figure 20), although in both cases (number of clusters and ARI) \textit{Seurat}'s resolution parameter had a much stronger impact.
 
@@
 Of note, the methods differ widely in view of their computing time (Figure \ref{fig:figure8}B) and we saw no relationship between the accuracy of the estimates and the complexity of the method.
-Reasoning that over-estimating the number of dimensions is less problematic than under-estimating it, we kept the former methods for a full analysis of their impact on clustering (Figure \ref{fig:figure8}C-D), when combined with sctransform or the standard Seurat normalization.
-Although most methods performed well on the various clustering measures, the global maximum likelihood based on translated Poisson Mixture Model (\textit{maxLikGlobal}) provided the dimensionality estimate that best separated the subpopulations (Figure \ref{fig:figure8}C) and resulted in the best clustering accuracy (Figure \ref{fig:figure8}D).
+Reasoning that over-estimating the number of dimensions is less problematic than under-estimating it, we kept the former methods for a full analysis of their impact on clustering (Figure \ref{fig:figure8}C-D), when combined with \textit{sctransform} or the standard \textit{Seurat} normalization.
+Although most methods performed well on the various clustering measures, the global maximum likelihood based on translated Poisson Mixture Model (\textit{maxLikGlobal}, using the 20 nearest neighbors) provided the dimensionality estimate that best separated the subpopulations (Figure \ref{fig:figure8}C) and resulted in the best clustering accuracy (Figure \ref{fig:figure8}D).
 This method systematically selected many more components than were associated with the subpopulations, suggesting that although these additional components appear individually uninformative, in combination they nevertheless contribute to classification.
+An implication is that the commonly-used elbow method most likely underestimates dimensionality.
 
@@
 The last step evaluated in our pipeline was clustering.
-Given previous works on the topic \cite{duoClustering2018,freytagComparison2018} and the success of graph-based clustering methods for scRNAseq, we restricted our evaluation to Seurat's method and two \textit{scran} SNN-based clustering approaches, based on random walks (\textit{walktrap} method) or on the optimization of the modularity score (\textit{fast\_greedy}).
+Given previous works on the topic \cite{duoClustering2018,freytagComparison2018} and the success of graph-based clustering methods for scRNAseq, we restricted our evaluation to Seurat's method and variations of \textit{scran}'s graph-based clustering using random walks (\textit{walktrap} method) or the optimization of the modularity score (\textit{fast\_greedy}) on various underlying graphs.
+Specifically, we compared rank-based graphs (\textit{scran}'s default) with graphs using the components' values directly, SNN vs KNN graphs, and the exact nearest neighbors to the Annoy approximation.
 Again, the tested methods were combined with \textit{Seurat}'s standard normalization and \textit{sctransform}, otherwise using the parameters found optimal in the previous steps.
 
-Since ARI is dominated by differences in the number of clusters (Supplementary Figure 17) and no single metric is perfect, we diversified them (Figure \ref{fig:figure9}).
-MI has the virtue of not decreasing when a true subpopulation is split into two clusters, which is arguably less problematic (and might well reflect unknown biological subgroups), but as a consequence it can be biased towards methods producing higher resolution clustering.
+Since ARI is dominated by differences in the number of clusters (Supplementary Figures 1-2) and no single metric is perfect, we diversified them (Figure \ref{fig:figure9}).
+MI has the virtue of not decreasing when a true subpopulation is split into two clusters, which is arguably less problematic (and might well reflect unknown biological subgroups), but as a consequence it can be biased towards methods producing higher resolution clustering (Supplementary Figure 2).
 Similarly, precision per true subpopulation is considerably more robust to differences in the number of clusters.
@@
 Finally, the ARI score at the true number of clusters, when available, showed similar performances.
-However, because \textit{Seurat}'s resolution parameter had a large impact on the number of clusters identified (Supplementary Figure 18), \textit{Seurat} could always be coerced into producing the right number of clusters.
-Instead, the number of clusters found by \textit{scran} was considerably less influenced by the available parameters (number of nearest neighbors or steps in the random walk - see Supplementary Figure 18), and as a result \textit{scran}-based clustering sometimes never produced a partitioning with the right number of clusters.
+However, because \textit{Seurat}'s resolution parameter had a large impact on the number of clusters identified (Supplementary Figures 1 and 21), \textit{Seurat} could always be coerced into producing the right number of clusters.
+Instead, the number of clusters found by \textit{scran} was considerably less influenced by the available parameters (number of nearest neighbors or steps in the random walk - see Supplementary Figure 21), and as a result \textit{scran}-based clustering sometimes never produced a partitioning with the right number of clusters.
 This observation, along with \textit{scran}'s higher MI score, suggest that \textit{scran} sometimes simply divides a real subpopulation into two clusters (possibly tracking some unknown biological differences) rather than committing misclassification errors.
 Overall, the \textit{walktrap} method appeared superior to the \textit{fast greedy} algorithm and was generally less prone to misclassification than \textit{Seurat} clustering, although the latter offered more control over the resolution.
+A major difference between \textit{walktrap}-based clustering and \textit{Seurat} is the computing time.
+We found that using the Annoy approximation to the nearest neighbors somewhat reduced computing time at no apparent cost to accuracy; however, it nevertheless remained considerably longer than \textit{Seurat}.
 Finally, some poorly distinguishable subpopulations from both the Zhengmix8eq and simMix1 datasets remained very inaccurately classified by all methods and in regards to all metrics.
@@
     \centering
-    \includegraphics[width=\textwidth]{{main_figures/clustering_files/figure-html/clustering-1.png}}
-    \caption{\textbf{Evaluation of clustering methods.} Clustering accuracy of common clustering tools in combination with \textit{sctransform} and standard Seurat's normalization.}
+    \includegraphics[scale=0.41, keepaspectratio]{{main_figures/clustering_files/figure-html/clustering-1.png}}
+    \caption{\textbf{Evaluation of clustering methods.} Clustering accuracy (\textbf{A}) and computing time (\textbf{B}) of \textit{scran}- and \textit{Seurat}-based clustering tools in combination with \textit{sctransform} and standard \textit{Seurat}'s normalization.}
     \label{fig:figure9}
@@
     \includegraphics[width=\textwidth]{{main_figures/imputation_files/figure-html/imputation-1.png}}
-     \caption{\textbf{Imputation/denoising methods} Average silhouette width per subpopulation (\textbf{A}) and clustering accuracy (\textbf{B}) with or without (indicated as \textit{none}) application of a denoising/imputation method.}
+     \caption{\textbf{Evaluation of imputation/denoising methods} Average silhouette width per subpopulation (\textbf{A}) and clustering accuracy (\textbf{B}) with or without (indicated as \textit{none}) application of a denoising/imputation method.}
     \label{fig:figure10}
@@
      \item Similarly to previous reports \cite{SunDimRed2019}, we recommend the irlba-based PCA using weighting of the components, as implemented in \textit{Seurat}. 
-     \item Instead of the common elbow or jackstraw methods for deciding on the components to include, we recommend the \textit{global maximum likelihood based on translated Poisson Mixture Model} method (e.g., implemented in the  \texttt{maxLikGlobalDimEst} function of \texttt{intrinsicDimension}).
+     \item We advise against the commonly-used elbow method (because it is too conservative) and the jackstraw method (low performance at great computational costs) for deciding on the components to include, and recommend a method like the \textit{global maximum likelihood based on translated Poisson Mixture Model} method (e.g., implemented in the  \texttt{maxLikGlobalDimEst} function of \texttt{intrinsicDimension}, in our case using the 20 nearest neighbors).
    \end{itemize}
@@
    \begin{itemize}
-    \item We found \textit{scran}-based \textit{walktrap} clustering to show good performances.
+    \item We found \textit{scran}-based \textit{walktrap} clustering to show good performances, at however increased computational costs which might be prohibitive with large datasets.
     \item In cases where prior knowledge can guide the choice of a resolution, \textit{Seurat} can be useful in affording manual control of it while, in the absence of such knowledge, \textit{scran}-based \textit{walktrap} clustering provided reasonable estimates.
@@
 For instance, \textit{Seurat} and \textit{scran} methods were already shown to yield good clustering performance \cite{freytagComparison2018} and \textit{DrImpute} to improve the quality of downstream analyses \cite{ZhangImput2018}.
-However, a previous systematic comparison study \cite{viethSystematic2019} revealed that the use of \textit{SAVER} for denoising data and \textit{Scran} for clustering yielded better performance.
-It should be noted that the criteria of evaluations of the latter study differed to ours as the authors focused on the ability of the tools to recover DE genes, whereas we focused on the ability to recover cell populations.
+A previous systematic comparison \cite{viethSystematic2019} instead found that the use of \textit{SAVER} for denoising data and \textit{scran} for clustering yielded better performance on the ability to recover differentially-expressed genes, whereas we focused on the ability to recover cell populations.
 Another disagreement with the present recommendations come from the study of Hou et al.
@@
 We demonstrate how a diversity of multilevel metrics can be more robust, more sensitive, and more nuanced than simply evaluating the final clustering performance.
-In addition, we provide a new and efficient Bioconductor package for doublet detection, \texttt{scDblFinder}.
 We hope that the \textit{pipeComp} framework can be applied to extend the current field of benchmarking, as well as to apply it to other range of methods.
@@
 All analyses were performed through the \href{https://github.com/plger/pipeComp}{pipeComp} R package, which implements the pipeline framework described here.
+Clustering and scVI results were produced with pipeComp 0.99.4, while all other results were produced with pipeComp 0.99.2 (`biorxiv` branch of the repository).
 All code to reproduce the simulations and figures is available in the \url{https://github.com/markrobinsonuzh/scRNA\_pipelines\_paper} repository, which also includes the basic datasets with a standardized annotation.
@@
 For the other datasets, we used the unfiltered counts from \cite{duoClustering2018}, available on the corresponding repository \url{https://github.com/markrobinsonuzh/scRNAseq\_clustering\_comparison}.
-For simplicity, all starting \textit{SingleCellExperiment} objects with standardized metadata are available on \url{https://github.com/markrobinsonuzh/scRNA\_pipelines\_paper}.
+For simplicity, all starting \textit{SingleCellExperiment} objects with standardized metadata are available on \url{https://doi.org/10.6084/m9.figshare.11787210.v1}.
 
 \subsection*{Software and package versions}
-Analyses were performed in R 3.6.0 (Bioconductor 3.9) and the following packages were installed from GitHub repositories: \textit{Seurat} (version 3.0.0), \textit{sctransform} (0.2.0), \textit{DoubletFinder} (2.0.1), \textit{scDblFinder} (1.1.1), \textit{scds} (1.0.0), \textit{SAVERX} (1.0.0), \textit{scImpute} (0.0.9), \textit{ALRA} (commit 7636de8), \textit{DCA} (0.2.2), \textit{DrImpute} (1.2), \textit{ENHANCE} (R version, commit 1571696), \textit{SAVERX} (1.0.0).
+Analyses were performed in R 3.6.0 (Bioconductor 3.9) and the following packages were installed from GitHub repositories: \textit{Seurat} (version 3.0.0), \textit{sctransform} (0.2.0), \textit{DoubletFinder} (2.0.1), \textit{scDblFinder} (1.1.4), \textit{scds} (1.0.0), \textit{SAVERX} (1.0.0), \textit{scImpute} (0.0.9), \textit{ALRA} (commit 7636de8), \textit{DCA} (0.2.2), \textit{DrImpute} (1.2), \textit{ENHANCE} (R version, commit 1571696), \textit{SAVERX} (1.0.0).
 The code for the glmPCA was obtained from \url{https://github.com/willtownes/scrna2019} (commit 1ddcc30ebb95d083a685f12fe81d35dd1b0cb1b2).
@@
 
-\subsection*{Doublet detection method}
-Our doublet detection method is available at \url{https://github.com/plger/scDblFinder}.
-Briefly, after reducing the data to the most expressed genes, we cluster the cells using the fast-greedy algorithm, favouring overclustering.
-We then create artificial doublets by sampling the two cells specifically from different clusters and sum the counts of each pair of cells.
-We also create meta-cells from each cluster and use them to create additional doublets and triplets.
-We combine them with the real cells, perform PCA and build a KNN graph using \textit{BiocNeighbors}.
-We then calculate, for each cell, the proportion of its neighbors that are artificial doublets, weighted by the distance.
-This ratio serves as a doublet score, which is then thresholded by simultaneously minimizing the error in classifying real vs artificial doublets and the deviation from the distribution of expected doublet rate (accounting for homotypic doublets as done by \textit{DoubletFinder}).
-
 \subsection*{Filter sets}
@@
 
-The evaluation of \textit{scVI} algorithm was performed in separate to the other pipeline combinations as the model training on our reference datasets was especially resource-consuming.
+The evaluation of \textit{scVI} algorithm was performed separate to the other pipeline combinations as the model training on our reference datasets was especially resource-consuming.
 As it was not practically feasible to perform the training step on many combination of parameters as it was done with other tools, we evaluated it on a representative subset of pipelines using \textit{sctransform} normalization and \textit{Seurat}'s PCA dimension reduction method.
-The resulting evaluation is available in the Supplementary Figure [TBD].
+The resulting evaluation is available in the Supplementary Figure 12.
 
@@
 \section*{Author's contributions}
-    Text for this section \ldots
+PLG developed the \textit{pipeComp} framework and implemented the evaluation metrics.
+PLG and AS prepared pipelines, methods wrappers and performed the analyses.
+AS performed all denoising/imputation analyses.
+PLG, AS and MDR wrote the manuscript.
 
@@
 \section*{Additional Files}
-  \subsection*{Supplementary figures}
+  \subsection*{Supplementary figures 1-21}
     % Additional file descriptions text

### 2020-02-28 supplementary_figures/supp5_dr.Rmd
added figure on k in maxLikGlobal

@@
 Adjusted Rand Index of clustering depending on the resolution and number of dimensions used. Based on sctransform and seurat PCA.
+
+# Supplementary Figure `r getFigNb(TRUE)`
+
+```{r, include=FALSE, eval=FALSE}
+source(system.file("extdata", "scrna_alternatives.R", package="pipeComp"))
+suppressPackageStartupMessages({
+  library(SingleCellExperiment)
+  library(Seurat)
+  library(scater)
+  library(intrinsicDimension)
+})
+datasets <- list.files("../datasets", full.names = TRUE)
+names(datasets) <- gsub("\\..*", "", basename(datasets))
+datasets <- lapply(datasets, readRDS)
+datasets <- lapply(datasets, filt.mad)
+datasets <- lapply(datasets, norm.seurat)
+datasets <- lapply(datasets, sel.deviance)
+
+met <- c("seurat.pca", "scran.denoisePCA", "scran.runPCA")
+res <- list()
+for(i in met) {
+  res[[i]] <- list()
+  print(i)
+  a <- lapply(datasets, function(x) {
+    switch(i,
+           seurat.pca = seurat.pca(x),
+           scran.denoisePCA = scran.denoisePCA(x),
+           scran.runPCA = scran.runPCA(x))
+  })
+  x <- lapply(a, function(x) reducedDim(x, "PCA"))
+  rm(a); gc()
+  ks <- seq(5, 40, 5)
+  for (d in names(datasets)) {
+    res[[i]][[d]]$ndims <- c()
+    res[[i]][[d]]$k <- c()
+  }
+  for (k in ks) {
+    print(paste("k:", k))
+    for (d in names(datasets)) {
+      res[[i]][[d]]$ndims <- c(res[[i]][[d]]$ndims,
+                               as.numeric(maxLikGlobalDimEst(x[[d]], k=k, unbiased=TRUE)))
+      res[[i]][[d]]$k <- c(res[[i]][[d]]$k, k)
+    }
+  }
+}
+saveRDS(res, "../data/maxLikGlobal.rds")
+```
+
+```{r, fig.width=7, fig.height=5}
+res <- readRDS("../data/maxLikGlobal.rds")
+res <- dplyr::bind_rows(lapply(res$seurat.pca, as.data.frame), .id="dataset")
+cols <- pipeComp::getQualitativePalette(9)
+names(cols) <- unique(res$dataset)
+ggplot(res, aes(k, ndims, colour=dataset)) + geom_point() + geom_line() + 
+  scale_colour_manual(values=cols) + xlab("Number of nearest neighbors (k)") +
+  ylab("Estimated dimensionality") + 
+  ggtitle("intrinsicDimension::maxLikGlobalDimEst")
+```
+
+### Supplementary Figure `r getFigNb()`
+Estimates of dimensionality by the `intrinsicDimension::maxLikGlobalDimEst` method using various 'reasonable' numbers of nearest neighbors (`k` parameter).

### 2020-03-01 supplementary_figures/supp3b_scVI.Rmd
updated scVI results

@@
 sil$method <- factor(paste(sil$norm, sil$dr))
-levels(sil$method) <- c("scVI.LD","scVI+seurat.pca","scVI.LD","sctransform+\nseurat.pca")
+levels(sil$method) <- meths <- c("scVI.LD", "scran+seurat.pca", "scVI.LD", "scVI+seurat.pca","scVI.LD","sctransform")
 levels(sil$dataset) <- c("Koh","Kumar","mixology10x3cl","mixology10x5cl","simMix1","simMix2","Zhengmix4eq","Zhengmix4uneq","Zhengmix8eq")
@@
 res$evaluation$clustering$delta.nbClust <- res$evaluation$clustering$n_clus-res$evaluation$clustering$true.nbClusts
-levels(res$evaluation$clustering$method) <- c("scVI.LD","scVI+seurat.pca","scVI.LD","sctransform+\nseurat.pca")
-H2 <- scrna_evalPlot_clust(res, what="MI", agg.by="method", anno_legend=FALSE) + 
-  scrna_evalPlot_clust(res, what="ARI", agg.by="method", anno_legend=FALSE) + 
-  scrna_evalPlot_clust(res, what="delta.nbClust", agg.by="method", col=circlize::colorRamp2(c(-6, 0, 6), c("red", "white", "Darkblue")), show_heatmap_legend = FALSE, value_cols = c("black","black"), value_format = "%.0f", anno_legend=FALSE)
+levels(res$evaluation$clustering$method) <- meths
+meths <- levels(sil$method)[c(4,2,3,1)]
+H2 <- scrna_evalPlot_clust(res, what="MI", agg.by="method", anno_legend=FALSE, reorder_rows=meths ) + 
+  scrna_evalPlot_clust(res, what="ARI", agg.by="method", anno_legend=FALSE, reorder_rows=meths) + 
+  scrna_evalPlot_clust(res, what="ARI", atTrueK = TRUE, agg.by="method", anno_legend=FALSE, reorder_rows=meths)
 ```
 
-```{r, fig.width=9, fig.height=3.5}
+```{r, fig.width=9, fig.height=4}
 plot_grid(grid.grabExpr(draw(H)), grid.grabExpr(draw(H2)), ncol=1, labels = LETTERS[1:2])
@@
 **scVI evaluation.
-A:** Average silhouette width per subpopulation using sctransform and standard Seurat PCA, scVI normalization followed by Seurat PCA, or scVI linear decoder (LD).
+A:** Average silhouette width per subpopulation using either sctransform, scran or scVI normalization followed by Seurat PCA, or the scVI linear decoder (LD).
 **B:** Clustering accuracy across the same methods followed by Seurat clustering.

### 2020-03-01 draft.tex
Updates from Overleaf

@@
 We collected real datasets of known cell composition (Table \ref{tab:table1}) and used a variety of evaluation metrics to investigate in a multilevel fashion the impact of various parameters and variations around a core scRNAseq pipeline.
-Although we use some datasets based on other protocols, our focus is especially on droplet-based datasets that do not include exogenous control RNA (i.e.
-spike-ins); see Table \ref{tab:table1} and Figure \ref{fig:figure1} for more details.
+Although we use some datasets based on other protocols, our focus is especially on droplet-based datasets that do not include exogenous control RNA (i.e., spike-ins); see Table \ref{tab:table1} and Figure \ref{fig:figure1} for more details.
 In addition to previously-used benchmark datasets with true cell labels \cite{duoClustering2018,tianMixology2018}, we simulated two datasets with a hierarchical subpopulation structure based on real 10x human and mouse data using \textit{muscat} \cite{CrowellMuscat2019}.
@@
 For the sole purpose of this section, we included an additional dataset with SNP information but lacking true cell labels \cite{kangMultiplexedDemuxlet2018}.
-Of note, SNP-based analyses also call doublets created by cells of the same cell type (but from different individuals) and which are generally described as homotypic (as opposed to neotypic or heterotypic doublets, i.e.
-doublets from different cell types).
+Of note, SNP-based analyses also call doublets created by cells of the same cell type (but from different individuals) and which are generally described as homotypic (as opposed to neotypic or heterotypic doublets, i.e., doublets from different cell types).
 These homotypic doublets might not be identifiable from the mere gene counts, and their identification is not generally the primary aim of doublet callers since they are often considered innocuous and, when across individuals, can be identified through other means (e.g., SNPs).
@@
 Applying the same filtering criteria on individual clusters of cells (identified through scran's \texttt{quickCluster} method) resulted in nearly no cell being filtered out.
-This suggests that filtering on the global population tends to discard cells of subpopulations with more extreme properties (e.g.
-high library size), rather than low-quality cells.
+This suggests that filtering on the global population tends to discard cells of subpopulations with more extreme properties (e.g., high library size), rather than low-quality cells.
 Finally, by changing the distributions, the doublet removal step in conjunction with filtering sometimes resulted in a net decrease in the proportion of excluded cells while retaining or improving accuracy.
@@
 As a first approximation of the true dimensionality, we computed the variance in each principal component that was explained by the subpopulations, which sharply decreased after the first few components in most datasets (Figure \ref{fig:figure8}A).
-Figure \ref{fig:figure8}B shows the difference between the dimension estimates of the above methods and that based on the subpopulations (i.e.
-from Figure \ref{fig:figure8}A).
+Figure \ref{fig:figure8}B shows the difference between the dimension estimates of the above methods and that based on the subpopulations (i.e., from Figure \ref{fig:figure8}A).
 Of note, the methods differ widely in view of their computing time (Figure \ref{fig:figure8}B) and we saw no relationship between the accuracy of the estimates and the complexity of the method.
@@
 For the methods that produce local dimensionality estimates, we used the maximum.
-For the elbow method, we implemented an automatic procedure by taking the farthest point from a line drawn between the variance explained by the first and last (i.e.
-50th) components calculated.
+For the elbow method, we implemented an automatic procedure by taking the farthest point from a line drawn between the variance explained by the first and last (i.e., 50th) components calculated.
 For the \texttt{JackStraw} method, since the \textit{Seurat} documentation advises not to use a p-value threshold (which would typically yield a very large number of dimensions) but rather look for a drop in significance, we applied the same farthest point algorithm on the log10(p-values), which in our hands reproduced manual threshold selection.

### 2020-03-01 main_figures/normalization.Rmd
updated normalization figure

@@
 ---
-title: "Figure 3 normalization"
+title: "Normalization"
 author: "Pierre-Luc Germain"
 date: "`r format(Sys.time(), '%d %B, %Y')`"
-output:
-  html_document:
-    keep_md: true
+output: html_document
 ---
@@
 source("../misc_functions.R")
-theme_set(theme_cowplot(font_size = 11))
-renameNorm <- function(x){
-  .renameHrows(x, f=function(x){
-    x <- gsub("^norm\\.", "", x)
-    x <- gsub("seuratvst","sctransform",x)
-    x <- gsub(".noscale", " (no scaling)", x)
-  })
-}
+theme_set(theme_cowplot(font_size = 10))
 ```
+
+(Requires `pipeComp` version >=0.99.3)
 
@@
 ```{r}
-p1 <- scrna_evalPlot_DR(res, "covarRes", agg.by = "norm", anno_legend=FALSE, reorder_rows = )
-p1 <- p1 + scrna_evalPlot_DR(res, "covarRes", covar = "log10_total_features", agg.by = "norm", anno_legend=FALSE, reorder_rows=p1)
+p1 <- scrna_evalPlot_DR(res, "log10_total_counts", agg.by = "norm", anno_legend=FALSE)
+p1 <- p1 + scrna_evalPlot_DR(res, "log10_total_features", agg.by = "norm", anno_legend=FALSE, reorder_rows=p1)
 p1 <- grid.grabExpr(draw(renameNorm(p1)))
@@
 p2 <- grid.grabExpr(draw(
-  renameNorm(scrna_evalPlot_DR(res, what="silhouette", agg.by = "norm", value_format = "", show_heatmap_legend = TRUE))))
+  renameNorm(scrna_evalPlot_DR(res, what="meanSilWidth", agg.by = "norm", value_format = "", show_heatmap_legend = TRUE))))
 
-H <- scrna_evalPlot_clustAtTrueK(res, "ARI", agg.by = "norm", scale = TRUE, reorder_rows = TRUE, anno_legend=FALSE)
-H <- H + scrna_evalPlot_clust(res, "MI", agg.by = "norm", scale = TRUE, reorder_rows=row.names(H@matrix), anno_legend=FALSE)
+H <- scrna_evalPlot_clust(res, "MI", agg.by = "norm", scale = TRUE, reorder_rows=TRUE, anno_legend=FALSE)
+H <- H + scrna_evalPlot_clust(res, "ARI", agg.by = "norm", scale = TRUE, reorder_rows = row.names(H@matrix), anno_legend=FALSE) +
+  scrna_evalPlot_clust(res, "ARI", atTrueK = TRUE, agg.by = "norm", scale = TRUE, reorder_rows = row.names(H@matrix), anno_legend=FALSE)
 p3 <- grid.grabExpr(draw(renameNorm(H)))
@@
 
-```{r figure3, fig.width=9, fig.height=9}
-plot_grid(p1,p2,p3, nrow = 3, labels=LETTERS[1:3])
+```{r figure3, fig.width=9, fig.height=11}
+pdf("normalization.pdf", width=10, height=11)
+plot_grid(p1,p2,p3, nrow = 3, labels=LETTERS[1:3], rel_heights = c(10,9,10))
+dev.off()
+plot_grid(p1,p2,p3, nrow = 3, labels=LETTERS[1:3], rel_heights = c(10,9,10))
 ```
@@
 Evaluation of normalization procedures.
-**A:** Residual correlation of the first principal component with library size (left) and detection rate (right), after accounting for biological differences between subpopulations.
+**A:** Average of the absolute within-subpopulation correlation of the first 5 principal components with library size (left) and detection rate (right).
 **B:** Average silhouette width per true subpopulation, where higher silhouette width means a higher separability.
-**C:** Clustering accuracy, measured by the ARI at the true number of cluster (left) and by the average mutual information (MI) of the cluster assignment with true subpopulations.
+**C:** Clustering accuracy, measured by the mutual information (MI), the adjusted Rand Index (ARI), and the ARI at the true number of cluster.

### 2020-03-01 bmc_article.tex

@@
 %\parttitle{Background} %if any
-The massive growth of single-cell RNA-sequencing (scRNAseq) and methods for its analysis still lacks sufficient and up-to-date benchmarks that would guide analytical choices.
+%The massive growth of single-cell RNA-sequencing (scRNAseq) and methods for its analysis still lacks sufficient and up-to-date benchmarks that would guide analytical choices.
 Moreover, current studies are often focused on isolated steps of the process.
@@
 We evaluate common steps of such analyses, including filtering, doublet detection, normalization, feature selection, denoising, dimensionality reduction and clustering.
-On the basis of these analyses, we make a number of concrete recommendations about analysis choices.
+On the basis of these analyses, we make a number of concrete recommendations about current analysis choices.
 The evaluation framework, \textit{pipeComp}, has been implemented so as to easily integrate any other step or tool, allowing extensible benchmarks and easy application to other fields (\url{https://github.com/plger/pipeComp}).
+
+\parttitle{Background} The massive growth of single-cell RNA-sequencing (scRNAseq) and the methods for its analysis still lack sufficient and up-to-date benchmarks that could guide analytical choices.
+Numerous benchmark studies already exist and cover most of scRNAseq processing and analytical methods but only a few give advice on a comprehensive pipeline.
+Moreover, current studies often focused on isolated steps of the process and do not address the impact of a tool on both the intermediate and the final steps of the analysis.
+
+\parttitle{Results} Here, we present a flexible R framework for pipeline comparison with multi-level evaluation metrics.
+We apply it to the benchmark of scRNAseq analysis pipelines using simulated and real datasets with known cell identities, covering common methods of filtering, doublet detection, normalization, feature selection, denoising, dimensionality reduction and clustering.
+We evaluate the choice of these tools with multi-purpose metrics to assess their ability to reveal cell population structure and lead to efficient clustering.
+
+\parttitle{Conclusion} On the basis of our systematic evaluations of analysis pipelines, we make a number of concrete recommendations about current analysis choices and for a comprehensive pipeline.
+The evaluation framework that we developed, \textit{pipeComp}, has been implemented so as to easily integrate any other step, tool, or evaluation metric allowing extensible benchmarks and easy applications to other fields of reseach in scRNAseq or Bioinformatics (\url{https://github.com/plger/pipeComp}).
 
@@
 \begin{keyword}
-\kwd{single-cell RNAseq}
+\kwd{single-cell RNA sequencing (scRNAseq)}
 \kwd{pipeline}
@@
 \kwd{normalization}
-\kwd{dimensionality}
+\kwd{dimensionality reduction}
 \kwd{filtering}
@@
 In addition to previously-used benchmark datasets with true cell labels \cite{duoClustering2018,tianMixology2018}, we simulated two datasets with a hierarchical subpopulation structure based on real 10x human and mouse data using \textit{muscat} \cite{CrowellMuscat2019}.
-Since graph-based clustering \cite{satijaSeurat2015} was previously shown to consistently perform well across several datasets \cite{duoClustering2018,tianMixology2018}, we used the \textit{Seurat} pipeline as the starting point to perform an integrated investigation of: 1) doublet identification, 2) cell filtering, 3) normalization, 4) feature selection, 5) dimension reduction, 6) clustering.
+Since graph-based clustering \cite{satijaSeurat2015} was previously shown to consistently perform well across several datasets \cite{duoClustering2018,tianMixology2018}, we used the \textit{Seurat} pipeline as the starting point to perform an integrated investigation of: 1) doublet identification; 2) cell filtering; 3) normalization; 4) feature selection; 5) dimension reduction; and, 6) clustering.
 We compared competing approaches and also explored more fine-grained parameters and variations on common methods.
@@
 In addition, optional benchmark functions can be set for each step to provide standardized, multi-layered evaluation metrics.
-Given such a \texttt{PipelineDefinition} object, a set of alternative parameters (which might include different subroutines) and benchmark datasets, the \texttt{runPipeline} function then proceeds through all combinations of arguments, avoiding recomputing the same step twice and compiling evaluations (including running time) on the fly.
+Given such a \texttt{PipelineDefinition} object, a set of alternative parameters (which might include different subroutines) and benchmark datasets, the \texttt{runPipeline} function then proceeds through all combinations of arguments, avoiding recomputing the same step twice and compiling evaluations (including run time) on the fly.
 Variations in a given parameter can be evaluated using all metrics from this point downward in the pipeline.
@@
 Clustering results are also very sensitive and might not always capture improvements in earlier steps of the pipeline.
-We therefore favour monitoring several complementary metrics across multiple steps of the process.
-
-A clustering output is relatively fragile to variations in the pipeline.
-We therefore wanted to capture whether the effect of a given parameter alteration is robust to changes in the rest of the pipeline, or rather specific to a set of other pipeline parameters.
+In addition, we wanted to assess whether the effect of a given parameter alteration is robust to changes in the rest of the pipeline.
+We therefore monitored several complementary metrics across multiple steps of the process.
 We proceeded in a step-wise fashion, first testing a large variety of parameters at the early steps of the pipeline along with only a set of mainstream options downstream, then selecting main alternatives and proceeding to a more detailed benchmark of the next step (Figure \ref{fig:figure2}B).
@@
 We found most methods (including no normalization at all) to perform fairly well in most of the subpopulations.
-An exception was \textit{scVI} (Supplementary Figure 12) which performed poorly on most datasets in terms of both the subpopulation silhouette and clustering accuracy.
+An exception was \textit{scVI} (Supplementary Figure 12), which performed poorly on most datasets in terms of both the subpopulation silhouette and clustering accuracy.
+For time and resource cost reasons, \textit{scVI} was run on a reduced set of pipelines and compared to \textit{sctransform} (more details in the Methods section).
 Scaling tended to reduce the average silhouette width of some subpopulations and to increase that of some less distinguishable ones and was generally, but not always, beneficial on the accuracy of the final clustering.
@@
 Another common practice in feature selection is to use those with the highest mean expression.
-Recently, \cite{townesGlmpca2019} instead suggested to use deviance, while \textit{sctransform} provides its own ordering of genes based on transformed variance.
+Recently, it was instead suggested to use deviance \cite{townesGlmpca2019}, while \textit{sctransform} provides its own ordering of genes based on transformed variance.
 
-Reasoning that a selection method should ideally select genes whose variability is higher between subpopulations than within, we first assessed to what extent each method selected genes with a high proportion of variance or deviance explained by (real) subpopulation.
+Reasoning that a selection method should ideally select genes whose variability is higher between subpopulations than within, we first assessed to what extent each method selected genes with a high proportion of variance or deviance explained by the (real) subpopulations.
 As the proportion of variability in a gene attributable to subpopulations can be measured in various ways, we first compared three approaches: ANOVA on log-normalized count, ANOVA on \textit{sctransform} normalization, and deviance explained.
@@
 Regarding the percentage of deviance explained however, the standardized measures were often superior (Figure \ref{fig:figure6}A and Supplementary Figures 16-17).
-Deviance proved the method of choice to prioritize genes with a high variance explained by subpopulations (with mere expression level proving surprisingly good) but did not perform so well to select genes with a high deviance explained.
+Deviance proved the method of choice to prioritize genes with a high variance explained by subpopulations (with mere expression level proving surprisingly performant) but did not perform so well to select genes with a high deviance explained.
 
@@
 
-Since the various PCA approaches and implementations were recently benchmarked in a similar context \cite{TsuyuzakiPCA2020}, we focused on widely used approaches which had not yet been compared: \textit{Seurat}'s PCA, \textit{scran}'s \texttt{denoisePCA}, and GLM-PCA \cite{townesGlmpca2019}.
+Since the various PCA approaches and implementations were recently benchmarked in a similar context \cite{TsuyuzakiPCA2020}, we focused on widely used approaches that had not yet been compared: \textit{Seurat}'s PCA, \textit{scran}'s \texttt{denoisePCA}, and GLM-PCA \cite{townesGlmpca2019}.
 When relevant, we combined them with \textit{sctransform} normalization.
@@
 GLM-PCA tended to increase the average silhouette width of already well-defined subpopulations, but \textit{Seurat}'s PCA procedure however proved superior on all metrics.
-Like GLM-PCA, \textit{scVI}'s Linear Decoder (LD) does not explicitly rely on normalized counts data.
+Like GLM-PCA, \textit{scVI}'s Linear Decoder (LD) does not explicitly rely on normalized counts.
 Its performance was lower than a pipeline with non-weighted \textit{Seurat} PCA on all measures for datasets with more subpopulations (Supplementary Figure 12).
@@
 Overall, increasing the number of dimensions robustly led to a decrease in the number of clusters (Supplementary Figures 13 and 19).
-This tended to affect the accuracy of the clustering (Supplementary Figure 20), although in both cases (number of clusters and ARI) \textit{Seurat}'s resolution parameter had a much stronger impact.
+This tended to affect the accuracy of the clustering (Supplementary Figure 20), although in both cases (number of clusters and ARI), \textit{Seurat}'s resolution parameter had a much stronger impact.
 
@@
 Reasoning that over-estimating the number of dimensions is less problematic than under-estimating it, we kept the former methods for a full analysis of their impact on clustering (Figure \ref{fig:figure8}C-D), when combined with \textit{sctransform} or the standard \textit{Seurat} normalization.
-Although most methods performed well on the various clustering measures, the global maximum likelihood based on translated Poisson Mixture Model (\textit{maxLikGlobal}, using the 20 nearest neighbors) provided the dimensionality estimate that best separated the subpopulations (Figure \ref{fig:figure8}C) and resulted in the best clustering accuracy (Figure \ref{fig:figure8}D).
+Although most methods performed well on the various clustering measures, the global maximum likelihood based on translated Poisson Mixture Model (\textit{maxLikGlobal}, using the 20 nearest neighbors, although estimates were relatively stable across decent values of \textit{k}; see Supplementary Figure 21) provided the dimensionality estimate that best separated the subpopulations (Figure \ref{fig:figure8}C) and resulted in the best clustering accuracy (Figure \ref{fig:figure8}D).
 This method systematically selected many more components than were associated with the subpopulations, suggesting that although these additional components appear individually uninformative, in combination they nevertheless contribute to classification.
@@
 The last step evaluated in our pipeline was clustering.
-Given previous works on the topic \cite{duoClustering2018,freytagComparison2018} and the success of graph-based clustering methods for scRNAseq, we restricted our evaluation to Seurat's method and variations of \textit{scran}'s graph-based clustering using random walks (\textit{walktrap} method) or the optimization of the modularity score (\textit{fast\_greedy}) on various underlying graphs.
+Given previous work on the topic \cite{duoClustering2018,freytagComparison2018} and the success of graph-based clustering methods for scRNAseq, we restricted our evaluation to Seurat's method and variations of \textit{scran}'s graph-based clustering using random walks (\textit{walktrap} method) or the optimization of the modularity score (\textit{fast\_greedy}) on various underlying graphs.
 Specifically, we compared rank-based graphs (\textit{scran}'s default) with graphs using the components' values directly, SNN vs KNN graphs, and the exact nearest neighbors to the Annoy approximation.
@@
 However, because \textit{Seurat}'s resolution parameter had a large impact on the number of clusters identified (Supplementary Figures 1 and 21), \textit{Seurat} could always be coerced into producing the right number of clusters.
-Instead, the number of clusters found by \textit{scran} was considerably less influenced by the available parameters (number of nearest neighbors or steps in the random walk - see Supplementary Figure 21), and as a result \textit{scran}-based clustering sometimes never produced a partitioning with the right number of clusters.
+Instead, the number of clusters found by \textit{scran} was considerably less influenced by the available parameters (number of nearest neighbors or steps in the random walk - see Supplementary Figure 22), and as a result \textit{scran}-based clustering sometimes never produced a partitioning with the right number of clusters.
 This observation, along with \textit{scran}'s higher MI score, suggest that \textit{scran} sometimes simply divides a real subpopulation into two clusters (possibly tracking some unknown biological differences) rather than committing misclassification errors.
 Overall, the \textit{walktrap} method appeared superior to the \textit{fast greedy} algorithm and was generally less prone to misclassification than \textit{Seurat} clustering, although the latter offered more control over the resolution.
-A major difference between \textit{walktrap}-based clustering and \textit{Seurat} is the computing time.
-We found that using the Annoy approximation to the nearest neighbors somewhat reduced computing time at no apparent cost to accuracy; however, it nevertheless remained considerably longer than \textit{Seurat}.
-Finally, some poorly distinguishable subpopulations from both the Zhengmix8eq and simMix1 datasets remained very inaccurately classified by all methods and in regards to all metrics.
+A major difference between \textit{walktrap}-based clustering and \textit{Seurat} is the computing time (Figure \ref{fig:figure9}B).
+We found that using the Annoy approximation to the nearest neighbors somewhat reduced computing time at no apparent cost to accuracy.
+It nevertheless remained considerably slower than \textit{Seurat}.
+For all methods, some poorly distinguishable subpopulations from both the Zhengmix8eq and simMix1 datasets remained very inaccurately classified in regard to all metrics.
 
@@
     \includegraphics[scale=0.41, keepaspectratio]{{main_figures/clustering_files/figure-html/clustering-1.png}}
-    \caption{\textbf{Evaluation of clustering methods.} Clustering accuracy (\textbf{A}) and computing time (\textbf{B}) of \textit{scran}- and \textit{Seurat}-based clustering tools in combination with \textit{sctransform} and standard \textit{Seurat}'s normalization.}
+    \caption{\textbf{Evaluation of clustering methods.} Clustering accuracy (\textbf{A}) and run time (\textbf{B}) of \textit{scran}- and \textit{Seurat}-based clustering tools in combination with \textit{sctransform} and standard \textit{Seurat}'s normalization.}
     \label{fig:figure9}
@@
 When restricting ourselves to clustering analyses that yielded the `right' number of clusters, all tested methods improved classification  compared to a scenario with no imputation step (`none' label, Figure \ref{fig:figure10}B).
-However, the situation was not so straightforward with alternative metrics, where some methods (e.g., \textit{enhance}) consistently underperformed.
-10X datasets, which are typically characterized by a lower per-cell coverage and feature detection rate, benefited more from imputation and, in this context, \textit{DrImpute} and \textit{dca} tended to show the best performance.
+However, the situation was not so straightforward with alternative metrics, where some methods (e.g., \textit{ENHANCE}) consistently underperformed.
+10X datasets, which are typically characterized by a lower per-cell coverage and feature detection rate, benefited more from imputation and, in this context, \textit{DrImpute} and \textit{DCA} tended to show the best performance.
 On the contrary, imputing on normalized counts originating from Smart-seq technology was instead rather deleterious to the clustering accuracy.
@@
     \centering
-    \includegraphics[width=\textwidth]{{main_figures/imputation_files/figure-html/imputation-1.png}}
+    \includegraphics[scale=0.49, keepaspectratio]{{main_figures/imputation_files/figure-html/imputation-1.png}}
      \caption{\textbf{Evaluation of imputation/denoising methods} Average silhouette width per subpopulation (\textbf{A}) and clustering accuracy (\textbf{B}) with or without (indicated as \textit{none}) application of a denoising/imputation method.}
@@
        \item Denoising appeared beneficial to the identification of subpopulations in 10x datasets, but not in Smart-seq datasets.
-       \item We found especially \textit{alra} (with prior normalization), \textit{DrImpute} (with prior processing) and \textit{dca} to offer the best performances.
+       \item We found especially \textit{ALRA} (with prior normalization), \textit{DrImpute} (with prior processing) and \textit{DCA} to offer the best performances.
    \end{itemize}
@@
 Another disagreement with the present recommendations come from the study of Hou et al.
-\cite{HouImput2020} that recommends \textit{SAVERX} and discourage the use of \textit{dca} for better clustering, whereas our study found opposite results for these tools.
+\cite{HouImput2020} that recommends \textit{SAVERX} and discourage the use of \textit{DCA} for better clustering, whereas our study found opposite results for these tools.
 Several differences in the design of our study could explain this contrast, such as the use of different metrics (ARI at true number of clusters, mean precision/ recall), of different datasets (Koh, Kumar, simMix simulations) and the use of different filtering and normalization strategies.
-\textit{alra} however yielded good results in both studies and may be the most conservative option.
+\textit{ALRA} however yielded good results in both studies and may be the most conservative option.
 The study from Hou et al.
-also highlighted the usefulness of \textit{alra} when followed by differential and trajectory analysis both in UMI and Fluidigm technologies.
+also highlighted the usefulness of \textit{ALRA} when followed by differential and trajectory analysis both in UMI and Fluidigm technologies.
 
@@
 A key additional feature of \textit{pipeComp} is its ability to perform evaluation in parallel to the pipeline and offering on-the-fly evaluation at any step.
-This avoid the need to store potentially large intermediate data for all possible permutations and thus allowing a combinatorial benchmark.
+This avoids the need to store potentially large intermediate data for all possible permutations and thus allowing a combinatorial benchmark.
 In addition, the fact that benchmark functions are stored in the \texttt{PipelineDefinition} makes the benchmark more smoothly portable, making it easy to modify, extend with further methods, or apply to other datasets.
@@
 Finally, imputation had a varied impact on the clustering analysis and seemed to be linked to the technology that was used to generate the data.
-Also, the good performance of \textit{DrImpute} is in line with a previous study focused on the preservation of data structure in trajectory inference from scRNAseq \cite{ZhangImput2018} (\textit{alra} was however not included in this previous benchmark).
+Also, the good performance of \textit{DrImpute} is in line with a previous study focused on the preservation of data structure in trajectory inference from scRNAseq \cite{ZhangImput2018} (\textit{ALRA} was however not included in this previous benchmark).
 Our results also align with the hypothesis of this study on the respective strengths of linear and non-linear models and their use to different types of data, such as the highest performance of non-linear methods in developmental studies.
 
-\section*{Conclusion}
+\section*{Conclusions}
 
 \textit{pipeComp} is a flexible R framework for evaluating methods and parameter combinations in a complex pipeline, computing on-the-fly multilevel evaluation metrics.
-Applying this framework to scRNAseq clustering enabled us to make concrete recommendations on the steps of filtering, normalization, feature selection, denoising, dimensionality reduction and clustering.
+Applying this framework to scRNAseq clustering enabled us to make concrete recommendations on the steps of filtering, normalization, feature selection, denoising, dimensionality reduction and clustering (Figure \ref{fig:figure11}).
 We demonstrate how a diversity of multilevel metrics can be more robust, more sensitive, and more nuanced than simply evaluating the final clustering performance.
-We hope that the \textit{pipeComp} framework can be applied to extend the current field of benchmarking, as well as to apply it to other range of methods.
+We hope that the \textit{pipeComp} framework can be applied to extend the current field of benchmarking, as well as to investigate pipelines other domains with multi-step computational analyses.
 
@@
 All analyses were performed through the \href{https://github.com/plger/pipeComp}{pipeComp} R package, which implements the pipeline framework described here.
-Clustering and scVI results were produced with pipeComp 0.99.4, while all other results were produced with pipeComp 0.99.2 (`biorxiv` branch of the repository).
+scVI results and results of the clustering section were produced with pipeComp 0.99.4, while all other results were produced with pipeComp 0.99.2 (\textit{biorxiv}) branch of the repository).
 All code to reproduce the simulations and figures is available in the \url{https://github.com/markrobinsonuzh/scRNA\_pipelines\_paper} repository, which also includes the basic datasets with a standardized annotation.
@@
 As it is not clear if this step is a hard requirement for the method to perform well, \textit{DrImpute} was run with and without prior processing (\textit{DrImpute\_process} and \textit{DrImpute\_noprocess} labels, respectively).
-The \textit{dca} method only accepts integers counts, while two of the datasets had non-integer quantification of expected counts.
+The \textit{DCA} method only accepts integers counts, while two of the datasets had non-integer quantification of expected counts.
 For these datasets, we rounded up the counts prior to imputation.
-\textit{alra} is designed for normalized data but as we are evaluating normalization downstream to imputation, we used the method on both non-normalized (\textit{alra} label) and normalized counts (\textit{alra\_norm} label).
+\textit{ALRA} is designed for normalized data but as we are evaluating normalization downstream to imputation, we used the method on both non-normalized (\textit{alra} label) and normalized counts (\textit{alra\_norm} label).
 \textit{scImpute} requires an estimation of the expected number of clusters with the input data.
@@
 \section*{Author's contributions}
-PLG developed the \textit{pipeComp} framework and implemented the evaluation metrics.
-PLG and AS prepared pipelines, methods wrappers and performed the analyses.
+PLG developed the \textit{pipeComp} framework, and PLG and MDR designed the benchmark.
+PLG and AS prepared the pipelines, methods wrappers and performed the analyses.
 AS performed all denoising/imputation analyses.
@@
 \section*{Additional Files}
-  \subsection*{Supplementary figures 1-21}
-    % Additional file descriptions text
+  \subsection*{\textbf{Additional file 1:Supplementary figures 1-22}}
+  Additional results of evaluations across datasets among various metrics. (PDF 15 MB)
 

### 2020-03-02 supplementary_figures/supp3_normalization.Rmd
updated supp figures

@@
 
+```{r covar, fig.width=8, fig.height=4.5}
+res <- readRDS("../data/results_norm.rds")
+H <- scrna_evalPlot_DR(res, "log10_total_counts", agg.by = "norm", anno_legend=FALSE, covar.type="mean")
+H <- H + scrna_evalPlot_DR(res, "log10_total_features", agg.by = "norm", anno_legend=FALSE, covar.type="mean", reorder_rows = H)
+H
+```
+
+### Supplementary Figure `r getFigNb()`
+Mean per-subpopulation absolute correlation of the first 5 components with library size (left) and the number of detected features (right) across normalization procedures.
+
+\newpage
+
+# Supplementary Figure `r getFigNb(TRUE)`
+
+```{r}
+res <- readRDS("../data/scVI.rds")
+sil <- res$evaluation$dimreduction$silhouette$top_10_dims
+sil$method <- factor(paste(sil$norm, sil$dr))
+levels(sil$method) <- meths <- c("scVI.LD", "scran+seurat.pca", "scVI.LD", "scVI+seurat.pca","scVI.LD","sctransform")
+levels(sil$dataset) <- c("Koh","Kumar","mixology10x3cl","mixology10x5cl","simMix1","simMix2","Zhengmix4eq","Zhengmix4uneq","Zhengmix8eq")
+res$evaluation$dimreduction$silhouette$top_10_dims <- sil
+x <- range(res$evaluation$dimreduction$silhouette$top_10_dims$meanSilWidth)
+silColBreaks <- unique(round(c(x[1],0,x[2]/2,x[2]),1))
+H <- scrna_evalPlot_DR(res, "meanSilWidth", scale=FALSE, agg.by = "method", 
+                       reorder_rows=TRUE, value_format="", 
+                       show_heatmap_legend=TRUE, anno_legend = TRUE, 
+                       heatmap_legend_param=list(at=silColBreaks))
+
+res$evaluation$clustering$method <- factor(paste(res$evaluation$clustering$norm, res$evaluation$clustering$dr))
+res$evaluation$clustering$delta.nbClust <- res$evaluation$clustering$n_clus-res$evaluation$clustering$true.nbClusts
+levels(res$evaluation$clustering$method) <- meths
+meths <- levels(sil$method)[c(4,2,3,1)]
+H2 <- scrna_evalPlot_clust(res, what="MI", agg.by="method", anno_legend=FALSE, reorder_rows=meths ) + 
+  scrna_evalPlot_clust(res, what="ARI", agg.by="method", anno_legend=FALSE, reorder_rows=meths) + 
+  scrna_evalPlot_clust(res, what="ARI", atTrueK = TRUE, agg.by="method", anno_legend=FALSE, reorder_rows=meths)
+```
+
+```{r, fig.width=9, fig.height=4}
+plot_grid(grid.grabExpr(draw(H)), grid.grabExpr(draw(H2)), ncol=1, labels = LETTERS[1:2])
+```
+
+### Supplementary Figure `r getFigNb()`
+
+**scVI evaluation.
+A:** Average silhouette width per subpopulation using either sctransform, scran or scVI normalization followed by Seurat PCA, or the scVI linear decoder (LD).
+**B:** Clustering accuracy across the same methods followed by Seurat clustering.
+
+\newpage
+
+# Supplementary Figure `r getFigNb(TRUE)`
+
 ```{r norm_nbClusters, fig.width=8, fig.height=8.5}
-res <- readRDS("../data/norm2_endSummary.rds")
-res2 <- res[which(res$filt=="deffilter" & res$norm %in% c("norm.scran", "norm.scran.noscale", "norm.seurat", "norm.seuratvst", "norm.stableGsum")),]
-res2$diff.nbClusters <- res2$nbClusters-truek[res2$dataset]
-res3 <- aggregate(res2[,"diff.nbClusters"], by=res2[,c("dataset","dims","resolution","norm")], FUN=mean)
+res <- readRDS("../data/results_norm.rds")
+res <- res$evaluation$clustering
+res2 <- res[which(res$norm %in% c("norm.scran.scaled", "norm.scran", "norm.seurat", "norm.sctransform", "norm.stableGsum") & res$resolution<4),]
+res2 <- res2[grep("simMix",res2$dataset,invert=TRUE),]
+res2$dataset <- droplevels(res2$dataset)
+res2$delta.nbClusters <- res2$n_clus-res2$true.nbClusts
+res3 <- aggregate(res2[,"delta.nbClusters"], by=res2[,c("dataset","dims","resolution","norm")], FUN=mean)
 colnames(res3)[5] <- "nb"
 res3$norm <- gsub("norm.","",res3$norm, fixed=T)
-# ggplot(res3, aes(factor(dims), factor(resolution), fill=nb)) + stat_bin2d() + 
-#   scale_fill_viridis() + ylab("Resolution") + xlab("Number of dimensions") +
-#   theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
-#   coord_flip() + facet_grid(dataset~norm)
 
@@
 
-mod <- summary(lm(nbClusters~dataset+norm+dims+resolution, data=res2))
+mod <- summary(lm(n_clus~dataset+norm+dims+resolution, data=res2))
 plot_grid(plotlist = ll[-2], nrow=6, rel_heights=c(1,1,1,1,1,1.5))
@@
 The Kumar dataset is not shown here due to a lack of variation in the number of clusters detected.
-A rough ANOVA on `nbClusters~dataset+norm+dims+resolution` suggests that seuratvst (sctransform) is associated with a higher number of clusters (p~`r round(mod$coefficients["normnorm.seuratvst",4],3)`).
+A rough ANOVA on `nbClusters~dataset+norm+dims+resolution` suggests that seuratvst (sctransform) is associated with a higher number of clusters (p~`r round(mod$coefficients["normnorm.sctransform",4],3)`).
 
@@
 \newpage
+
+# Supplementary Figure `r getFigNb(TRUE)`
+
+```{r, fig.width=7.5, fig.height=6}
+res <- readRDS("../data/results_norm.rds")
+res <- res$elapsed$stepwise$normalization
+ag <- aggregate(res[,"elapsed",drop=FALSE], by=res[,c("dataset","norm")], FUN=mean)
+cols <- pipeComp::getQualitativePalette(9)
+names(cols) <- levels(ag$dataset)
+ggplot(ag, aes(norm, elapsed)) + geom_boxplot() + geom_point(aes(colour=dataset)) + 
+  coord_flip() + scale_colour_manual(values=cols) + ylab("") + xlab("Time (s)")
+```
+
+### Supplementary Figure `r getFigNb()`
+Running time of the normalization methods.

### 2020-03-02 main_figures/normalization.Rmd
updated norm figure

@@
 
-Evaluation of normalization procedures.
-**A:** Average of the absolute within-subpopulation correlation of the first 5 principal components with library size (left) and detection rate (right).
+Evaluation of normalization procedures. **A:** Variance in the first principal component explained by 
+library size and detection rate after accounting for subpopulation differences.
+The values are difference in adjusted _R^2_ between a `PC1~subpopulation+covariate` model and `PC1~subpopulation`.
 **B:** Average silhouette width per true subpopulation, where higher silhouette width means a higher separability.

### 2020-03-02 main_figures.tex
Updates from Overleaf

--- 2020-02-27 bmc_article.tex
@@
-%% BioMed_Central_Tex_Template_v1.06
-%%                                      %
-%  bmc_article.tex            ver: 1.06 %
-%                                       %
+\documentclass[12pt]{article}
+\pagestyle{empty}
+\usepackage{graphicx}
 
-%%IMPORTANT: do not delete the first line of this template
-%%It must be present to enable the BMC Submission system to
-%%recognise this template!!
+\title{pipeComp - Main figures}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                     %%
-%%  LaTeX template for BioMed Central  %%
-%%     journal article submissions     %%
-%%                                     %%
-%%          <8 June 2012>              %%
-%%                                     %%
-%%                                     %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\date{\today}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                                                 %%
-%% For instructions on how to fill out this Tex template           %%
-%% document please refer to Readme.html and the instructions for   %%
-%% authors page on the biomed central website                      %%
-%% http://www.biomedcentral.com/info/authors/                      %%
-%%                                                                 %%
-%% Please do not use \input{...} to include other tex files.       %%
-%% Submit your LaTeX manuscript as one .tex document.              %%
-%%                                                                 %%
-%% All additional figures and files should be attached             %%
-%% separately and not embedded in the \TeX\ document itself.       %%
-%%                                                                 %%
-%% BioMed Central currently use the MikTex distribution of         %%
-%% TeX for Windows) of TeX and LaTeX.  This is available from      %%
-%% http://www.miktex.org                                           %%
-%%                                                                 %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%% additional documentclass options:
-%  [doublespacing]
-%  [linenumbers]   - put the line numbers on margins
-
-%%% loading packages, author definitions
-
-%\documentclass[twocolumn]{bmcart}% uncomment this for twocolumn layout and comment line below
-\documentclass{bmcart}
-
-%%% Load packages
-%\usepackage{amsthm,amsmath}
-%\RequirePackage{natbib}
-\RequirePackage[authoryear]{natbib}% uncomment this for author-year bibliography
-\usepackage{hyperref}
-\usepackage[utf8]{inputenc} %unicode support
-%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
-%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
-\usepackage{graphicx}
-\usepackage{url}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                             %%
-%%  If you wish to display your graphics for   %%
-%%  your own use using includegraphic or       %%
-%%  includegraphics, then comment out the      %%
-%%  following two lines of code.               %%
-%%  NB: These line *must* be included when     %%
-%%  submitting to BMC.                         %%
-%%  All figure files must be submitted as      %%
-%%  separate graphics through the BMC          %%
-%%  submission process, not included in the    %%
-%%  submitted article.                         %%
-%%                                             %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%\def\includegraphic{}
-%\def\includegraphics{}
-
-%%% Put your definitions there:
-\startlocaldefs
-\endlocaldefs
-
-%%% Begin ...
 \begin{document}
-
-%%% Start of article front matter
-\begin{frontmatter}
-
-\begin{fmbox}
-\dochead{Research}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% Enter the title of your article here     %%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\title{pipeComp, a general framework for the evaluation of computational pipelines, reveals performant single cell RNA-seq preprocessing tools}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% Enter the authors here                   %%
-%%                                          %%
-%% Specify information, if available,       %%
-%% in the form:                             %%
-%%   <key>={<id1>,<id2>}                    %%
-%%   <key>=                                 %%
-%% Comment or delete the keys which are     %%
-%% not used. Repeat \author command as much %%
-%% as required.                             %%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\author[
-   addressref={aff1,aff2,aff3},          % id's of addresses, e.g. {aff1,aff2}
-   email={pierre-luc.germain@hest.ethz.ch}   % email address
-]{\inits{PLG}\fnm{Pierre-Luc} \snm{Germain}}
-\author[
-   addressref={aff1,aff2},
-    email={anthony.sonrel@uzh.ch}
-]{\fnm{Anthony} \snm{Sonrel}}
-\author[
-   addressref={aff1,aff2},
-   corref={aff1},  % id of corresponding address, if any
-   email={mark.robinson@imls.uzh.ch}
-]{\inits{MR}\fnm{Mark D.} \snm{Robinson}}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% Enter the authors' addresses here        %%
-%%                                          %%
-%% Repeat \address commands as much as      %%
-%% required.                                %%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\address[id=aff1]{
-  \orgname{Department of Molecular Life Sciences, University of Zürich}, % university, etc
-  \street{Winterthurerstrasse 190},
-  \postcode{8057}
-  \city{Zürich},
-  \cny{Switzerland}
-}
-\address[id=aff2]{
-  \orgname{SIB Swiss Institute of Bioinformatics}, % university, etc
-  \city{Zürich},
-  \cny{Switzerland}
-}
-\address[id=aff3]{
-  \orgname{D-HEST Institute for Neurosciences, ETH Zürich},
-  \street{Winterthurerstrasse 190},
-  \postcode{8057}
-  \city{Zürich},
-  \cny{Switzerland}
-}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\end{fmbox}% comment this for two column layout
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% The Abstract begins here                 %%
-%%                                          %%
-%% Please refer to the Instructions for     %%
-%% authors on http://www.biomedcentral.com  %%
-%% and include the section headings         %%
-%% accordingly for your article type.       %%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{abstractbox}
-
-\begin{abstract} % abstract
-%\parttitle{Background} %if any
-The massive growth of single-cell RNA-sequencing (scRNAseq) and methods for its analysis still lacks sufficient and up-to-date benchmarks that would guide analytical choices.
-Moreover, current studies are often focused on isolated steps of the process.
-Here, we present a flexible R framework for pipeline comparison with multi-level evaluation metrics and apply it to the benchmark of scRNAseq analysis pipelines using datasets with known cell identities.
-We evaluate common steps of such analyses, including filtering, doublet detection, normalization, feature selection, denoising, dimensionality reduction and clustering.
-On the basis of these analyses, we make a number of concrete recommendations about analysis choices.
-The evaluation framework, \textit{pipeComp}, has been implemented so as to easily integrate any other step or tool, allowing extensible benchmarks and easy application to other fields (\url{https://github.com/plger/pipeComp}).
-
-\end{abstract}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% The keywords begin here                  %%
-%%                                          %%
-%% Put each keyword in separate \kwd{}.     %%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{keyword}
-\kwd{single-cell RNAseq}
-\kwd{pipeline}
-\kwd{clustering}
-\kwd{normalization}
-\kwd{dimensionality}
-\kwd{filtering}
-\kwd{benchmark}
-\end{keyword}
-
-% MSC classifications codes, if any
-%\begin{keyword}[class=AMS]
-%\kwd[Primary ]{}
-%\kwd{}
-%\kwd[; secondary ]{}
-%\end{keyword}
-
-\end{abstractbox}
-%
-%\end{fmbox}% uncomment this for twcolumn layout
-
-\end{frontmatter}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% The Main Body begins here                %%
-%%                                          %%
-%% Please refer to the instructions for     %%
-%% authors on:                              %%
-%% http://www.biomedcentral.com/info/authors%%
-%% and include the section headings         %%
-%% accordingly for your article type.       %%
-%%                                          %%
-%% See the Results and Discussion section   %%
-%% for details on how to create sub-sections%%
-%%                                          %%
-%% use \cite{...} to cite references        %%
-%%  \cite{koon} and                         %%
-%%  \cite{oreg,khar,zvai,xjon,schn,pond}    %%
-%%  \nocite{smith,marg,hunn,advi,koha,mouse}%%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%% start of article main body
-% <put your article body there>
-
-%%%%%%%%%%%%%%%%
-%% Background %%
-%%
-\section*{Background}
-
-Single-cell RNA-sequencing (scRNAseq) and the set of attached analysis methods are evolving fast, with more than 560 software tools available to the community \cite{ZappiaDB2018}, roughly half of which are dedicated to tasks related to data processing such as clustering, ordering, dimension reduction or normalization.
-This increase in the number of available tools follows the development of new sequencing technologies and the growing number of reported cells, genes and cell populations \cite{SvenssonDB2019}.
-As data processing is a critical step in any scRNAseq analysis, affecting downstream analysis and interpretation, it is critical to evaluate the available tools.
-
-A number of good comparison and benchmark studies have already been performed on various steps related to scRNAseq processing and analysis and can guide the choice of methodology (\citealp{CobosDeconvolution2020, colePerformance2018, DalMolinDE2017, duoClustering2018, freytagComparison2018, GaoProcessing2020, HeiserDimred2019, HouImput2020, JaakkolaDE2017, KrzakClust2019, SonesonDE2018, SunDimRed2019, tianMixology2018, TranBatch2020, TsuyuzakiPCA2020, viethSystematic2019, WangDE2019, YipHVGs2018, ZhangImput2018}).
-However these recommendations need constant updating and often leave open many details of an analysis.
-Another missing aspect of current benchmarking studies is their limitation to capture all aspects of scRNAseq processing workflow.
-Although previous benchmarks already brought valuable recommendations for data processing, some only focused on one aspect of data processing (e.g., \citealp{SunDimRed2019}), did not evaluate how the tool selection affects downstream analysis (e.g., \citealp{TsuyuzakiPCA2020}) or did not tackle all aspects of data processing, such as doublet identification or cell filtering (e.g., \citealp{viethSystematic2019}).
-A thorough evaluation of the tools covering all major processing steps is however urgently needed as previous benchmarking studies highlighted that a combination of tools can have a drastic impact on downstream analysis, such as differential expression analysis and cell-type deconvolution\cite{viethSystematic2019,CobosDeconvolution2020}.
-It is then critical to evaluate not only the single effect of a preprocessing method but also its positive or negative interaction with all parts of a workflow.
-
-Here, we develop a flexible R framework for pipeline comparison and evaluate the various steps of analysis leading from an initial count matrix to a cluster assignment, which are critical in a wide range of applications.
-We collected real datasets of known cell composition (Table \ref{tab:table1}) and used a variety of evaluation metrics to investigate in a multilevel fashion the impact of various parameters and variations around a core scRNAseq pipeline.
-Although we use some datasets based on other protocols, our focus is especially on droplet-based datasets that do not include exogenous control RNA (i.e.
-spike-ins); see Table \ref{tab:table1} and Figure \ref{fig:figure1} for more details.
-In addition to previously-used benchmark datasets with true cell labels \cite{duoClustering2018,tianMixology2018}, we simulated two datasets with a hierarchical subpopulation structure based on real 10x human and mouse data using \textit{muscat} \cite{CrowellMuscat2019}.
-Since graph-based clustering \cite{satijaSeurat2015} was previously shown to consistently perform well across several datasets \cite{duoClustering2018,tianMixology2018}, we used the \textit{Seurat} pipeline as the starting point to perform an integrated investigation of: 1) doublet identification, 2) cell filtering, 3) normalization, 4) feature selection, 5) dimension reduction, 6) clustering.
-We compared competing approaches and also explored more fine-grained parameters and variations on common methods.
-Importantly, the success of methods at a certain analytical step might be dependent on choices at other steps.
-Therefore, instead of evaluating each step in isolation, we developed a general framework for evaluating nested variations on a pipeline and suggest a multilevel panel of metrics.
-Finally, we evaluate several recent methods and provide concrete recommendations.
-
-\section*{Results}
-
-\subsection*{A flexible framework for pipeline evaluation}
-
-The \textit{pipeComp} package defines a pipeline as, minimally, a list of functions executed consecutively on the output of the previous one (Figure \ref{fig:figure2}A).
-In addition, optional benchmark functions can be set for each step to provide standardized, multi-layered evaluation metrics.
-Given such a \texttt{PipelineDefinition} object, a set of alternative parameters (which might include different subroutines) and benchmark datasets, the \texttt{runPipeline} function then proceeds through all combinations of arguments, avoiding recomputing the same step twice and compiling evaluations (including running time) on the fly.
-Variations in a given parameter can be evaluated using all metrics from this point downward in the pipeline.
-This is especially important because end-point metrics, such as the adjusted Rand index (ARI) \cite{HubertARI1985} for clustering, are not perfect.
-For example, although the meaning of an ARI score is independent of the number of true subpopulations \cite{steinleyProperties2004}, the number of clusters called is by far the most important determinant of the score: the farther it is from the actual number of subpopulations, the worse the ARI (Supplementary Figures 1-2).
-In this context, one strategy has been to cluster across various resolutions and only consider the results that have the right number of clusters \cite{duoClustering2018}.
-While this has the virtue of making the results comparable, in practice the number of subpopulations is typically unknown and tools that operate well in this optimal context might not necessarily be best overall.
-Clustering results are also very sensitive and might not always capture improvements in earlier steps of the pipeline.
-We therefore favour monitoring several complementary metrics across multiple steps of the process.
-
-A clustering output is relatively fragile to variations in the pipeline.
-We therefore wanted to capture whether the effect of a given parameter alteration is robust to changes in the rest of the pipeline, or rather specific to a set of other pipeline parameters.
-We proceeded in a step-wise fashion, first testing a large variety of parameters at the early steps of the pipeline along with only a set of mainstream options downstream, then selecting main alternatives and proceeding to a more detailed benchmark of the next step (Figure \ref{fig:figure2}B).
+\maketitle
 
@@
     \centering
-    \includegraphics[scale=0.7, keepaspectratio]{main_figures/pipeline_explanation.png}
+    \includegraphics[scale=0.75, keepaspectratio]{main_figures/pipeline_explanation.pdf}
     \caption{\textbf{Overview of the \textit{pipeComp} framework and its application to a scRNAseq clustering pipeline.
@@
 
-\subsection*{Filtering}
-
-\subsubsection*{Doublet detection}
-
-Doublets, defined as two cells sequenced under the same cellular barcode (e.g., being captured in the same droplet), are fairly frequent in scRNAseq datasets, with estimates ranging from 1 to 10\% depending on the platform and cell concentration used \cite{bloomEstimating2018,kangMultiplexedDemuxlet2018}.
-While doublets of the same cell type are relatively innocuous for most downstream applications due to their conservation of the relative expression between genes, doublets formed from different cell types or states are likely to be misclassified and could potentially distort downstream analysis.
-In some cases, doublets can be identified through their unusually high number of reads and detected features, but this is not always the case (Supplementary Figure 3).
-A number of methods were developed to identify doublets, in most cases by comparing each cell to artificially-created doublets \cite{mcginnisDoubletfinder2019, LunScran2016, BaisScds2019}.
-We first evaluated the capacity of these methods to detect doublets using the two 10x datasets with cells of different genetic identity \cite{tianMixology2018}, using SNP genotypes as the ground truth.
-For the sole purpose of this section, we included an additional dataset with SNP information but lacking true cell labels \cite{kangMultiplexedDemuxlet2018}.
-Of note, SNP-based analyses also call doublets created by cells of the same cell type (but from different individuals) and which are generally described as homotypic (as opposed to neotypic or heterotypic doublets, i.e.
-doublets from different cell types).
-These homotypic doublets might not be identifiable from the mere gene counts, and their identification is not generally the primary aim of doublet callers since they are often considered innocuous and, when across individuals, can be identified through other means (e.g., SNPs).
-We therefore do not expect a perfect accuracy in datasets involving cells of the same type across individuals (as in the demuxlet dataset).
-
-We tested \textit{DoubletFinder} \cite{mcginnisDoubletfinder2019} and \textit{scran}'s \texttt{doubletCells} \cite{LunScran2016}, both of which use similarity to artificial doublets, and \textit{scds} \cite{BaisScds2019}, which relies on a combination of co-expression and binary classification.
-\textit{DoubletFinder} integrates a thresholding based on the proportion of expected doublets, while \textit{scran} and \textit{scds} return scores that must be manually thresholded.
-In these cases, we ensured that the right number of cells would be called doublets.
-In addition to these methods, we reasoned that an approach such as \textit{DoubletFinder} could be simplified by being applied directly on counts and by using a pre-clustering to create neotypic/heterotypic doublets more efficiently.
-We therefore also evaluated a simple and fast Bioconductor package implementing this method for doublet detection, \textit{scDblFinder}, with the added advantage of accounting for uncertainty in the expected doublet rate and using meta-cells from the clusters to even include triplets (see Methods).
-
-While most methods accurately identified the doublets in the 3 cell lines dataset (mixology10x3cl), the other two datasets proved more difficult (Figure \ref{fig:figure3}A).
-\textit{scDblFinder} achieved comparable or better accuracy than top alternatives while being the fastest method (Figure \ref{fig:figure3}B).
-Across datasets, cells called as doublets tended to be classified in the wrong cluster more often than other cells (Figure \ref{fig:figure3}C).
-We also found that \textit{scDblFinder} improved the accuracy of the clustering across all benchmark datasets even when, by design, the data contained no heterotypic doublet (Figure \ref{fig:figure4}).
-
 \begin{figure}
     \centering
-    \includegraphics[scale=0.65, keepaspectratio]{{main_figures/doublets_files/figure-html/figure1-1.png}}
+    \includegraphics[scale=0.85, keepaspectratio]{{main_figures/doublets.pdf}}
     \caption{\textbf{Identification of doublet cells.
@@
 
-\subsubsection*{Excluding more cells is not necessarily better}
-
-Beyond doublets, a dataset might include low-quality cells whose elimination would reduce noise.
-This has for instance been demonstrated for droplets containing a high content of mitochondrial reads, often as a result of cell degradation and resulting loss of cytoplasmic mRNAs \cite{IlicicLowQual2016}.
-A common practice is to exclude cells that differ considerably from most other cells on the basis of such properties.
-This can for instance be performed through the \texttt{isOutlier} function from \textit{scater} that measures, for a given control property, the number of median absolute deviations (MADs) of each cell from the median of all cells.
-Supplementary Figure 4 shows the distributions of some of the typical cell properties commonly used.
-Of note, these properties tend to be correlated, with some exceptions.
-For example, while a high proportion of mitochondrial reads is often correlated with a high proportion of the counts in the top features, there can also be other reasons for an over-representation of highly-expressed features (Supplementary Figure 5), such as an over-amplification in non-UMI datasets.
-In our experience, 10X datasets also exhibit a very strong correlation between the total counts and the total features even across very different cell types (Supplementary Figure 6).
-We therefore also measure the ratio between the two and treat cells strongly departing from this trend with suspicion.
-
-Reasoning that the cells we wish to exclude are the cells that would be misclassified, we measured the rate of misclassification of each cell in each dataset across a variety of clustering pipelines, correcting for the median misclassification rate of the subpopulation and then evaluated what properties could be predictive of misclassification (Supplementary Figures 7-9).
-We could not identify any property or simple combination thereof that would be consistently predictive of misclassification; the only feature that consistently stood out across multiple datasets (the Zheng datasets) was that cells with very high read counts have a higher chance of being misclassified.
-
-We next investigated the impact of filtering according to various criteria (see Methods).
-An important risk of excluding cells on the basis of their distance from the whole distribution of cells on some properties (e.g., library size) is that these properties tend to have different distributions across subpopulations.
-As a result, thresholds in terms of number of MADs from the whole distribution can lead to strong biases against certain subpopulations (Figure \ref{fig:figure4}A).
-We therefore examined the tradeoff between the increased accuracy of filtering and the maximum proportion of cells excluded per subpopulation (Figure \ref{fig:figure4}B and Supplementary Figure 10).
-Since filtering changes the relative abundance of the different subpopulations, global clustering accuracy metrics such as ARI are not appropriate here.
-We therefore calculated the per-subpopulation precision and recall using the Hungarian algorithm \cite{PapadimitriouHu1998} and monitored their mean F1 score.
-A first observation was that, although the stringent filtering tended to be associated with an increase in accuracy, it could also become deleterious and most of the benefits could be achieved without very stringent filtering and minimizing subpopulation bias (Figure \ref{fig:figure4}B).
-Applying the same filtering criteria on individual clusters of cells (identified through scran's \texttt{quickCluster} method) resulted in nearly no cell being filtered out.
-This suggests that filtering on the global population tends to discard cells of subpopulations with more extreme properties (e.g.
-high library size), rather than low-quality cells.
-Finally, by changing the distributions, the doublet removal step in conjunction with filtering sometimes resulted in a net decrease in the proportion of excluded cells while retaining or improving accuracy.
-We therefore recommend the use of doublet removal followed by relatively mild filtering, such as is implemented in our `default' filtering (see Methods).
-
 \begin{figure}
     \centering
-    \includegraphics[scale=0.60, keepaspectratio]{{main_figures/filtering_files/figure-html/figure2-1.png}}
+    \includegraphics[scale=0.72, keepaspectratio]{{main_figures/filtering.pdf}}
     \caption{\textbf{Effect of filtering on cell subpopulation structure and clustering.
@@
 
-\subsubsection*{Filtering features by type}
-
-Mitochondrial reads have been associated with cell degradation and there is evidence that ribosomal genes can influence the clustering output, hiding other biological structure in the analysis \cite{freytagComparison2018}.
-We therefore investigated whether excluding one category of features or the other, or using only protein-coding genes, had an impact on the ability to distinguish subpopulations (Supplementary Figure 11).
-Removal of ribosomal genes robustly reduced the quality of the clustering, suggesting that they represent real biological differences between subpopulations.
-Removing mitochondrial genes and restricting to protein-coding genes had a very mild impact.
-
-\subsection*{Normalization and scaling}
-
-We next investigated the impact of different normalization strategies.
-Beside the standard log-normalization included in \textit{Seurat}, we tested \textit{scran}'s pooling-based normalization \cite{lunPooling2016}, \textit{sctransform}'s variance-stabilizing transformation \cite{hafemeisterSCtransform2019}, and normalization based on stable genes \cite{linStableGenes2018, deekeStablyExpressed2018}.
-In addition to log-normalization, the standard \textit{Seurat} clustering pipeline performs per-feature unit-variance scaling so that the PCA is not too strongly dominated by highly-expressed features.
-We therefore included versions of the different normalization procedures, with or without a subsequent scaling (\textit{sctransform}'s variance-stabilizing transformation involves an approach analogous to scaling).
-\textit{Seurat}'s scaling function also includes the option to regress out the effect of certain covariates.
-We tested its performance by using the proportion of mitochondrial counts and the number of detected features as covariates.
-Finally, it has been proposed that the use of stable genes, in particular cytosolic ribosomal genes, can be used to normalize scRNAseq \cite{deekeStablyExpressed2018}.
-We therefore evaluated a simple linear normalization based on the sum of these genes, as well as nuclear genes.
-
-An important motivation for the development of \textit{sctransform} was the observation that, even after normalization, the first principal components of various datasets tended to correlate with library size, suggesting an inadequate normalization \cite{hafemeisterSCtransform2019}.
-However, as library size tends to vary across subpopulations, part of this effect can simply reflect biological differences.
-We therefore assessed to what extent the first principal component still retained a correlation with the library size and the number of detected features, removing the confounding covariation with the subpopulations (Figure \ref{fig:figure5}A).
-The simple step of scaling tended to remove much of the correlation with these features and, in the absence of scaling, standard \textit{Seurat} normalization resulted in fairly high correlation with technical covariates.
-\textit{sctransform} led to the lowest correlation but most methods, including normalization based on stable genes, were able to remove most of the correlation when combined with scaling.
-The only exception is \textit{Seurat} normalization with scaling regressing out covariates, which surprisingly increased the correlation in 8 of the 9 datasets.
-
-We further evaluated normalization methods by investigating their impact on the separability of the subpopulations (Figure \ref{fig:figure5}B-C).
-Since clustering accuracy metrics such as the ARI are very strongly influenced by the number of clusters, we complemented it with silhouette width \cite{RousseeuwSil1987} and mutual information (MI).
-We found most methods (including no normalization at all) to perform fairly well in most of the subpopulations.
-An exception was \textit{scVI} (Supplementary Figure 12) which performed poorly on most datasets in terms of both the subpopulation silhouette and clustering accuracy.
-Scaling tended to reduce the average silhouette width of some subpopulations and to increase that of some less distinguishable ones and was generally, but not always, beneficial on the accuracy of the final clustering.
-Regressing out covariates systematically gave poorer performance on all metrics.
-\textit{sctransform} systematically outperformed other methods and, even though it was developed to be applied to data with unique molecular identifiers (UMI), it also performed fairly well with the Smart-seq protocol (Koh and Kumar datasets).
+\begin{figure}
+    \centering
+    \includegraphics[scale=0.65, keepaspectratio]{{main_figures/normalization.pdf}}
+    \caption{\textbf{Evaluation of normalization procedures.} \textbf{A:} Variance in PC1 explained by library size (left) and detection rate (right) after accounting for subpopulation differences.
+\textbf{B:} Average silhouette width per true subpopulation, where higher silhouette width means a higher separability.
+\textbf{C:} Clustering accuracy (\textit{Seurat} clustering), measured by mutual information (MI), Adjusted Rand Index (ARI) and ARI at the true number of cluster.
+'feat_regress', 'mt_regress' and 'feat_mt_regress' respectively stand for the regressing out of the number of features, the proportion of mitochondrial reads, or both during scaling.
+}
+    \label{fig:figure5}
+\end{figure}
 
@@
     \centering
-    \includegraphics[scale=0.55, keepaspectratio]{{main_figures/normalization_files/figure-html/figure3-1.png}}
-    \caption{\textbf{Evaluation of normalization procedures.} `Seurat.feat\_mt\_regress' regresses out the number of features and proportion of mitochondrial reads during scaling; `Seurat.feat\_regress` regresses out the number of features only.
-Residual correlation of the first principal component with library size (left) and detection rate (right), after accounting for biological differences between subpopulations.
-\textbf{B:} Average silhouette width per true subpopulation, where higher silhouette width means a higher separability.
-\textbf{C:} Clustering accuracy (\textit{Seurat} clustering), measured by the ARI at the true number of cluster (left) and by the average MI of the cluster assignment with true subpopulations.}
-    \label{fig:figure5}
-\end{figure}
-
-Finally, we monitored whether, under the same downstream clustering analysis, different normalization methods tended to lead to an over- or under-estimation of the number of clusters.
-Although some methods had a tendency to lead to a higher (e.g., \textit{sctransform}) or lower (e.g., stable genes) number of clusters, the effect was very mild and not entirely systematic (Supplementary Figure 13).
-We also confirmed that, especially in the UMI-based datasets, \textit{sctransform} did in fact successfully stabilize variance across mean counts (Supplementary Figure 14).
-
-\subsection*{Feature selection}
-
-A standard clustering pipeline typically involves a step of highly-variable genes selection, which is complicated by the digital nature and the mean-variance relationship of (sc)RNAseq.
-\textit{Seurat}'s earlier approaches involved the use of dispersion estimates standardized for the mean expression levels, while more recent versions ($\geq$3.0) rely on a different measure of variance, again standardized.
-While adjusting for the mean-variance relationship removes much of the bias towards highly-expressed genes, it is plausible that this relationship may in fact sometimes reflects biological relevance and would be helpful in classifying cell types.
-Another common practice in feature selection is to use those with the highest mean expression.
-Recently, \cite{townesGlmpca2019} instead suggested to use deviance, while \textit{sctransform} provides its own ordering of genes based on transformed variance.
-
-Reasoning that a selection method should ideally select genes whose variability is higher between subpopulations than within, we first assessed to what extent each method selected genes with a high proportion of variance or deviance explained by (real) subpopulation.
-As the proportion of variability in a gene attributable to subpopulations can be measured in various ways, we first compared three approaches: ANOVA on log-normalized count, ANOVA on \textit{sctransform} normalization, and deviance explained.
-The ANOVAs performed on a standard \textit{Seurat} normalization and on \textit{sctransform} data were highly correlated (Supplementary Figure 15A).
-These estimates were also in good agreement with the deviance explained, although lowly-expressed genes could have a high deviance explained without having much of their variance explained by subpopulation (Supplementary Figure 15B-D).
-We therefore compared the proportion of the cumulative variance/deviance explained by the top X genes that could be retrieved by each gene ranking method (Supplementary Figures 16-17).
-We first focused on the first 1000 genes to highlight the differences between methods, although a higher number of selected genes decreased the differences between methods (Supplementary Figures 16-18).
-The standardized measures of variability were systematically worse than their non-standardized counterparts in selecting genes with a high proportion of variance explained by subpopulation.
-Regarding the percentage of deviance explained however, the standardized measures were often superior (Figure \ref{fig:figure6}A and Supplementary Figures 16-17).
-Deviance proved the method of choice to prioritize genes with a high variance explained by subpopulations (with mere expression level proving surprisingly good) but did not perform so well to select genes with a high deviance explained.
-
-\begin{figure}
-    \centering
-    \includegraphics[scale=0.65]{{main_figures/selection_files/figure-html/figure4-1.png}}
+    \includegraphics[scale=0.65]{{main_figures/selection.pdf}}
     \caption{\textbf{Evaluation of feature selection methods.
@@
 
-We next evaluated how the use of different feature selection methods affected the clustering accuracy (Figure \ref{fig:figure6}B).
-To validate the previous assay on the proportion of variance/deviance explained by real populations, we included genes that maximized these two latter measures.
-Interestingly, while these selections were on average the top-ranking methods, they were not systematically best for all datasets.
-The previous observations were reflected in the ARI of the resulting clustering (Figure \ref{fig:figure6}B): non-standardized measures of variability, including mere expression level, tended to outperform more complex metrics.
-In general, we found deviance and unstandardized estimates of variance to provide the best results across datasets and normalization methods.
-Increasing the number of features selected also systematically led to an increase in the accuracy of the clustering, typically plateauing after 4000 features (Supplementary Figure 18).
-
-\subsection*{Dimensionality reduction}
-
-Since the various PCA approaches and implementations were recently benchmarked in a similar context \cite{TsuyuzakiPCA2020}, we focused on widely used approaches which had not yet been compared: \textit{Seurat}'s PCA, \textit{scran}'s \texttt{denoisePCA}, and GLM-PCA \cite{townesGlmpca2019}.
-When relevant, we combined them with \textit{sctransform} normalization.
-Given that \textit{Seurat}'s default PCA weights the cell embeddings by the variance of each component, we also evaluated the impact of this weighting with each method.
-
-The impact of the choice of dimensionality reduction method was far greater than that of normalization or feature selection (Figure \ref{fig:figure7}).
-GLM-PCA tended to increase the average silhouette width of already well-defined subpopulations, but \textit{Seurat}'s PCA procedure however proved superior on all metrics.
-Like GLM-PCA, \textit{scVI}'s Linear Decoder (LD) does not explicitly rely on normalized counts data.
-Its performance was lower than a pipeline with non-weighted \textit{Seurat} PCA on all measures for datasets with more subpopulations (Supplementary Figure 12).
-Overall, weighting the principal components by their variance (as \textit{Seurat} does) had a positive impact on silhouette widths and ARI scores.
-
 \begin{figure}
     \centering
-    \includegraphics[scale=0.40, keepaspectratio]{{main_figures/dr_files/figure-html/dimred-1.png}}
+    \includegraphics[scale=0.50, keepaspectratio]{{main_figures/dr.pdf}}
     \caption{\textbf{Evaluation of common dimensionality reduction methods.
@@
 
-\subsubsection*{Estimating the number of dimensions}
-
-A common step following dimension reduction is the selection of an appropriate number of dimensions to use for downstream analysis.
-Since Euclidean distance decreases as the number of non-discriminating dimensions increases, there is usually a trade-off between selecting enough dimensions to keep most information and excluding smaller dimensions that may represent technical noise or other unwanted sources of variation.
-Overall, increasing the number of dimensions robustly led to a decrease in the number of clusters (Supplementary Figures 13 and 19).
-This tended to affect the accuracy of the clustering (Supplementary Figure 20), although in both cases (number of clusters and ARI) \textit{Seurat}'s resolution parameter had a much stronger impact.
-
-Different approaches have been proposed to select the appropriate number of dimensions, from the visual identification of an 'Elbow' (inflexion point) of the variance explained, to more complex algorithms.
-We evaluated the performance of dimensionality estimators implemented in the \textit{intrinsicDimension} package \cite{johnssonDimensionality2015}, as well as common procedures such as the `elbow` method (inflexion point in the variance explained by each component), some scRNAseq-specific methods such as the JackStraw procedure \cite{ChungJackstraw2015} or \textit{scran}'s \texttt{denoisePCA} \cite{LunScran2016}, and the recent application of Fisher Separability analysis \cite{AlberganteSepar2019}.
-
-We compared the various estimates of dimensionality in their ability to retrieve the intrinsic number of dimensions in a dataset, based on Seurat's weighted PCA space.
-As a first approximation of the true dimensionality, we computed the variance in each principal component that was explained by the subpopulations, which sharply decreased after the first few components in most datasets (Figure \ref{fig:figure8}A).
-Figure \ref{fig:figure8}B shows the difference between the dimension estimates of the above methods and that based on the subpopulations (i.e.
-from Figure \ref{fig:figure8}A).
-Of note, the methods differ widely in view of their computing time (Figure \ref{fig:figure8}B) and we saw no relationship between the accuracy of the estimates and the complexity of the method.
-Reasoning that over-estimating the number of dimensions is less problematic than under-estimating it, we kept the former methods for a full analysis of their impact on clustering (Figure \ref{fig:figure8}C-D), when combined with \textit{sctransform} or the standard \textit{Seurat} normalization.
-Although most methods performed well on the various clustering measures, the global maximum likelihood based on translated Poisson Mixture Model (\textit{maxLikGlobal}, using the 20 nearest neighbors) provided the dimensionality estimate that best separated the subpopulations (Figure \ref{fig:figure8}C) and resulted in the best clustering accuracy (Figure \ref{fig:figure8}D).
-This method systematically selected many more components than were associated with the subpopulations, suggesting that although these additional components appear individually uninformative, in combination they nevertheless contribute to classification.
-An implication is that the commonly-used elbow method most likely underestimates dimensionality.
-
 \begin{figure}
     \centering
-    \includegraphics[scale=0.53, keepaspectratio]{{main_figures/dimensionality_files/figure-html/figure6-1.png}}
+    \includegraphics[scale=0.6, keepaspectratio]{{main_figures/dimensionality.pdf}}
     \caption{\textbf{Estimating dimensionality.
@@
 
-\subsection*{Clustering}
-
-The last step evaluated in our pipeline was clustering.
-Given previous works on the topic \cite{duoClustering2018,freytagComparison2018} and the success of graph-based clustering methods for scRNAseq, we restricted our evaluation to Seurat's method and variations of \textit{scran}'s graph-based clustering using random walks (\textit{walktrap} method) or the optimization of the modularity score (\textit{fast\_greedy}) on various underlying graphs.
-Specifically, we compared rank-based graphs (\textit{scran}'s default) with graphs using the components' values directly, SNN vs KNN graphs, and the exact nearest neighbors to the Annoy approximation.
-Again, the tested methods were combined with \textit{Seurat}'s standard normalization and \textit{sctransform}, otherwise using the parameters found optimal in the previous steps.
-
-Since ARI is dominated by differences in the number of clusters (Supplementary Figures 1-2) and no single metric is perfect, we diversified them (Figure \ref{fig:figure9}).
-MI has the virtue of not decreasing when a true subpopulation is split into two clusters, which is arguably less problematic (and might well reflect unknown biological subgroups), but as a consequence it can be biased towards methods producing higher resolution clustering (Supplementary Figure 2).
-Similarly, precision per true subpopulation is considerably more robust to differences in the number of clusters.
-We also tracked the mean F1 score and the ARI at the true number of clusters.
-
-The MI score and minimum precision, which are largely independent of the estimated number of clusters, were overall higher for the \textit{walktrap} method (Figure \ref{fig:figure9}), while the mean F1 score favored both \textit{scran} methods (\textit{walktrap} and \textit{fast greedy}) over \textit{Seurat}.
-Finally, the ARI score at the true number of clusters, when available, showed similar performances.
-However, because \textit{Seurat}'s resolution parameter had a large impact on the number of clusters identified (Supplementary Figures 1 and 21), \textit{Seurat} could always be coerced into producing the right number of clusters.
-Instead, the number of clusters found by \textit{scran} was considerably less influenced by the available parameters (number of nearest neighbors or steps in the random walk - see Supplementary Figure 21), and as a result \textit{scran}-based clustering sometimes never produced a partitioning with the right number of clusters.
-This observation, along with \textit{scran}'s higher MI score, suggest that \textit{scran} sometimes simply divides a real subpopulation into two clusters (possibly tracking some unknown biological differences) rather than committing misclassification errors.
-Overall, the \textit{walktrap} method appeared superior to the \textit{fast greedy} algorithm and was generally less prone to misclassification than \textit{Seurat} clustering, although the latter offered more control over the resolution.
-A major difference between \textit{walktrap}-based clustering and \textit{Seurat} is the computing time.
-We found that using the Annoy approximation to the nearest neighbors somewhat reduced computing time at no apparent cost to accuracy; however, it nevertheless remained considerably longer than \textit{Seurat}.
-Finally, some poorly distinguishable subpopulations from both the Zhengmix8eq and simMix1 datasets remained very inaccurately classified by all methods and in regards to all metrics.
+\begin{figure}
+    \centering
+    \includegraphics[scale=0.47, keepaspectratio]{{main_figures/clustering.pdf}}
+    \caption{\textbf{Evaluation of clustering methods.} Clustering accuracy (\textbf{A}) and run time (\textbf{B}) of \textit{scran}- and \textit{Seurat}-based clustering tools in combination with \textit{sctransform} and standard \textit{Seurat}'s normalization.}
+    \label{fig:figure9}
+\end{figure}
 
@@
     \centering
-    \includegraphics[scale=0.41, keepaspectratio]{{main_figures/clustering_files/figure-html/clustering-1.png}}
-    \caption{\textbf{Evaluation of clustering methods.} Clustering accuracy (\textbf{A}) and computing time (\textbf{B}) of \textit{scran}- and \textit{Seurat}-based clustering tools in combination with \textit{sctransform} and standard \textit{Seurat}'s normalization.}
-    \label{fig:figure9}
-\end{figure}
-
-\subsection*{Further extensions to the pipeline: imputation/denoising}
-
-The basic pipeline presented here can be extended with additional analysis steps while keeping the same evaluation metrics.
-To demonstrate this, we evaluated various imputation or denoising techniques based on their impact on classification.
-Since preliminary analysis showed that all methods performed equally well or better on normalized data, we applied them after filtering and normalization, but before scaling and reduction.
-Although some of the methods (e.g., \textit{DRImpute\_process} and \textit{alra\_norm}) did improve the separability of some more elusive subpopulations, no method had a systematically positive impact on the average silhouette width across all subpopulations (Figure \ref{fig:figure10}A).
-When restricting ourselves to clustering analyses that yielded the `right' number of clusters, all tested methods improved classification  compared to a scenario with no imputation step (`none' label, Figure \ref{fig:figure10}B).
-However, the situation was not so straightforward with alternative metrics, where some methods (e.g., \textit{enhance}) consistently underperformed.
-10X datasets, which are typically characterized by a lower per-cell coverage and feature detection rate, benefited more from imputation and, in this context, \textit{DrImpute} and \textit{dca} tended to show the best performance.
-On the contrary, imputing on normalized counts originating from Smart-seq technology was instead rather deleterious to the clustering accuracy.
-
-\begin{figure}
-    \centering
-    \includegraphics[width=\textwidth]{{main_figures/imputation_files/figure-html/imputation-1.png}}
+    \includegraphics[scale=0.55, keepaspectratio]{{main_figures/imputation.pdf}}
      \caption{\textbf{Evaluation of imputation/denoising methods} Average silhouette width per subpopulation (\textbf{A}) and clustering accuracy (\textbf{B}) with or without (indicated as \textit{none}) application of a denoising/imputation method.}
@@
 
-\section*{Discussion}
-
-\subsection*{Concrete recommendations}
-
-On the basis of our findings, we can make a number of concrete analysis recommendations, also summarized in Figure \ref{fig:figure11}:
-
-\begin{enumerate}
-   \item Filtering
-   \begin{itemize}
-     \item Doublet detection and removal is advised and can be performed at little computing cost with software such as \textit{scDblFinder} or \textit{scds}.
-     \item Distribution-based cell filtering fails to capture doublets and should use relatively lenient cutoffs (e.g., 5 MADs, or 3 MADs in at least 2 distributions) to exclude poor-quality cells while avoiding bias against some subpopulations.
-     \item Features filtering based on feature type did not appear beneficial.
-   \end{itemize}
-   \item Normalization and scaling
-   \begin{itemize}
-     \item Most normalization methods tested yielded a fair performance, especially when combined with scaling, which tended to have a positive impact on clustering.
-     \item \textit{sctransform} offered the best overall performance in terms of the separability of the subpopulations, as well as removing the effect of library size and detection rate.
-     \item The common practice of regressing out cell covariates, such as the detection rate or proportion of mitochondrial reads nearly always had a negative impact, leading to increased correlation with covariates and decreased clustering accuracy.
-We therefore advise against this practice.
-   \end{itemize}
-   \item Feature selection
-   \begin{itemize}
-     \item Deviance \cite{townesGlmpca2019} offered the best ranking of genes for feature selection.
-     \item Increasing the number of features included tended to lead to better classifications, plateauing from 4000 features in our datasets.
-   \end{itemize}
-   \item Denoising/imputation
-   \begin{itemize}
-       \item Denoising appeared beneficial to the identification of subpopulations in 10x datasets, but not in Smart-seq datasets.
-       \item We found especially \textit{alra} (with prior normalization), \textit{DrImpute} (with prior processing) and \textit{dca} to offer the best performances.
-   \end{itemize}
-   \item PCA
-   \begin{itemize}
-     \item Similarly to previous reports \cite{SunDimRed2019}, we recommend the irlba-based PCA using weighting of the components, as implemented in \textit{Seurat}. 
-     \item We advise against the commonly-used elbow method (because it is too conservative) and the jackstraw method (low performance at great computational costs) for deciding on the components to include, and recommend a method like the \textit{global maximum likelihood based on translated Poisson Mixture Model} method (e.g., implemented in the  \texttt{maxLikGlobalDimEst} function of \texttt{intrinsicDimension}, in our case using the 20 nearest neighbors).
-   \end{itemize}
-   \item Clustering
-   \begin{itemize}
-    \item We found \textit{scran}-based \textit{walktrap} clustering to show good performances, at however increased computational costs which might be prohibitive with large datasets.
-    \item In cases where prior knowledge can guide the choice of a resolution, \textit{Seurat} can be useful in affording manual control of it while, in the absence of such knowledge, \textit{scran}-based \textit{walktrap} clustering provided reasonable estimates.
-   \end{itemize}
-\end{enumerate}
-
 \begin{figure}[h]
     \centering
-    \includegraphics[scale=0.08,keepaspectratio]{main_figures/pipeComp_summary.png}
+    \includegraphics[scale=0.9,keepaspectratio]{main_figures/pipeComp_summary.pdf}
     \caption{\textbf{Recommendations of tools from filtering to clustering.}}
@@
 
-\subsection*{Limitations and open questions}
-
-In this study, we evaluated tools commonly used for the processing of scRNAseq with a focus on droplet-based datasets, namely from the 10x technology.
-Although this platform has been used in almost half of the scRNAseq studies in 2019 \cite{SvenssonDB2019}, other popular technologies such as Drop-seq, InDrops or Smart-seq2/3 were not represented in the present benchmarking.
-Differences between such protocols, even among droplet-based technologies, can have a very large impact on cell capture efficiency, cell numbers and clustering \cite{MereuCellAtlas2019, ZhangDroplet2019, SalomonDroplet2019}.
-Although most top-ranking methods in our comparison performed well on both Smart-seq and 10x datasets that we tested, future benchmarking efforts should strive to include less represented technologies.
-In addition, we did not compare any of the alignment and/or quantification methods used to obtain the count matrix, which was for instance discussed in \cite{viethSystematic2019}.
-Some steps, such as the implementation of the PCA, were also not explored in detail here as they have already been the object of recent and thorough study elsewhere \cite{SunDimRed2019, TsuyuzakiPCA2020}.
-We also considered only methods relying on Euclidean distance, while correlation was recently reported to be superior\cite{kim_impact_2019} and would require further investigation.
-
-Our recommendations are in line with the results of several other benchmarking studies.
-For instance, \textit{Seurat} and \textit{scran} methods were already shown to yield good clustering performance \cite{freytagComparison2018} and \textit{DrImpute} to improve the quality of downstream analyses \cite{ZhangImput2018}.
-A previous systematic comparison \cite{viethSystematic2019} instead found that the use of \textit{SAVER} for denoising data and \textit{scran} for clustering yielded better performance on the ability to recover differentially-expressed genes, whereas we focused on the ability to recover cell populations.
-Another disagreement with the present recommendations come from the study of Hou et al.
-\cite{HouImput2020} that recommends \textit{SAVERX} and discourage the use of \textit{dca} for better clustering, whereas our study found opposite results for these tools.
-Several differences in the design of our study could explain this contrast, such as the use of different metrics (ARI at true number of clusters, mean precision/ recall), of different datasets (Koh, Kumar, simMix simulations) and the use of different filtering and normalization strategies.
-\textit{alra} however yielded good results in both studies and may be the most conservative option.
-The study from Hou et al.
-also highlighted the usefulness of \textit{alra} when followed by differential and trajectory analysis both in UMI and Fluidigm technologies.
-
-Here, we chose to concentrate on what could be learned from datasets with known cell labels (as opposed to labels inferred from the data, as in \cite{MereuCellAtlas2019}).
-In contrast to Tian et al.
-\cite{tianMixology2018}, who used RNA mixtures of known proportions, we chose to rely chiefly on real cells and their representative form of variability.
-Given the limited availability of such well-described datasets however, several aspects of single-cell analysis could not be compared, such as batch effect correction or multi-dataset integration.
-For these aspects of scRNAseq processing that are critical in some experimental designs, we refer the reader to previous evaluations \cite{TranBatch2020, StuartIntegration2019}.
-In addition, a focus on the identification of the subpopulations might fail to reveal methods that are instead best performing for tasks other than clustering, such as differential expression analysis or trajectory inference.
-Several informative benchmarks have already been performed on some of these topics \cite{CrowellMuscat2019, DalMolinDE2017, JaakkolaDE2017, SaelensTraject2019, SonesonDE2018, WangDE2019}.
-Yet, such evaluations could benefit from considering methods not in isolation, but as parts of a connected workflow, as we have done here.
-We believe that the \textit{pipeComp} framework is modular and flexible enough to integrate new steps in the pipeline, as shown by an example with imputation/denoising methods.
-
-We developed \texttt{pipeComp} to address the need of a framework that can simultaneously evaluate the interaction of multiple tools.
-The work of Vieth and colleagues\cite{viethSystematic2019} already offered an important precedent in this respect, evaluating the interaction of various steps in the context of scRNAseq differential expression analysis, but did not offer a platform for doing so.
-Instead, the \textit{CellBench} package was recently proposed to address a similar need \cite{suCellbench}, offering an elegant piping syntax to combine alternative methods at successive steps.
-A key additional feature of \textit{pipeComp} is its ability to perform evaluation in parallel to the pipeline and offering on-the-fly evaluation at any step.
-This avoid the need to store potentially large intermediate data for all possible permutations and thus allowing a combinatorial benchmark.
-In addition, the fact that benchmark functions are stored in the \texttt{PipelineDefinition} makes the benchmark more smoothly portable, making it easy to modify, extend with further methods, or apply to other datasets.
-
-With respect to the methods themselves, we believe there is still space for improvement in some of the steps.
-Concerning cell filtering, we noticed that the current approach based on whole-population characteristic (e.g., MAD cut-off) can be biased against certain subpopulations, suggesting that more refined methods expecting multimodal distributions should be used rather than relying on whole-population characteristics.
-In addition, most common filtering approaches do not harness the relationship between cell QC properties.
-Finally, imputation had a varied impact on the clustering analysis and seemed to be linked to the technology that was used to generate the data.
-Also, the good performance of \textit{DrImpute} is in line with a previous study focused on the preservation of data structure in trajectory inference from scRNAseq \cite{ZhangImput2018} (\textit{alra} was however not included in this previous benchmark).
-Our results also align with the hypothesis of this study on the respective strengths of linear and non-linear models and their use to different types of data, such as the highest performance of non-linear methods in developmental studies.
-
-\section*{Conclusion}
-
-\textit{pipeComp} is a flexible R framework for evaluating methods and parameter combinations in a complex pipeline, computing on-the-fly multilevel evaluation metrics.
-Applying this framework to scRNAseq clustering enabled us to make concrete recommendations on the steps of filtering, normalization, feature selection, denoising, dimensionality reduction and clustering.
-We demonstrate how a diversity of multilevel metrics can be more robust, more sensitive, and more nuanced than simply evaluating the final clustering performance.
-We hope that the \textit{pipeComp} framework can be applied to extend the current field of benchmarking, as well as to apply it to other range of methods.
-
-\section*{Methods}
-
-\subsection*{Code and data availability}
-
-All analyses were performed through the \href{https://github.com/plger/pipeComp}{pipeComp} R package, which implements the pipeline framework described here.
-Clustering and scVI results were produced with pipeComp 0.99.4, while all other results were produced with pipeComp 0.99.2 (`biorxiv` branch of the repository).
-All code to reproduce the simulations and figures is available in the \url{https://github.com/markrobinsonuzh/scRNA\_pipelines\_paper} repository, which also includes the basic datasets with a standardized annotation.
-
-The gene counts for the two mixology datasets were downloaded from the  \href{https://github.com/LuyiTian/CellBench_data/data/sincell_with_class.RData}{CellBench repository}, commit 74fe79e.
-For the other datasets, we used the unfiltered counts from \cite{duoClustering2018}, available on the corresponding repository \url{https://github.com/markrobinsonuzh/scRNAseq\_clustering\_comparison}.
-For simplicity, all starting \textit{SingleCellExperiment} objects with standardized metadata are available on \url{https://doi.org/10.6084/m9.figshare.11787210.v1}.
-
-\subsection*{Software and package versions}
-Analyses were performed in R 3.6.0 (Bioconductor 3.9) and the following packages were installed from GitHub repositories: \textit{Seurat} (version 3.0.0), \textit{sctransform} (0.2.0), \textit{DoubletFinder} (2.0.1), \textit{scDblFinder} (1.1.4), \textit{scds} (1.0.0), \textit{SAVERX} (1.0.0), \textit{scImpute} (0.0.9), \textit{ALRA} (commit 7636de8), \textit{DCA} (0.2.2), \textit{DrImpute} (1.2), \textit{ENHANCE} (R version, commit 1571696), \textit{SAVERX} (1.0.0).
-The code for the glmPCA was obtained from \url{https://github.com/willtownes/scrna2019} (commit 1ddcc30ebb95d083a685f12fe81d35dd1b0cb1b2).
-
-\subsection*{Simulated datasets}
-The simMix1 dataset was based on the human PBMC CITE-seq data deposited under accession code GSE100866 of the Gene Expression Omnibus (GEO) website.
-Both RNA and ADT count data were downloaded from GEO and processed independently using \textit{Seurat}.
-We then considered cells that were in the same cluster both in the RNA-based and ADT-based analyses to be real subpopulations and focused on the 4 most abundant ones.
-We then performed 3 sampling-based simulations with various degrees of separation using \textit{muscat} \cite{CrowellMuscat2019} and merged the three simulations.
-The simMix2 dataset was generated from the mouse brain data published in \cite{CrowellMuscat2019} in a similar fashion.
-The specific code is available on \url{https://github.com/markrobinsonuzh/scRNA\_pipelines\_paper}.
-
-\subsection*{Default pipeline parameters}
-
-Where unspecified, the following default parameters or parameter sets were used:
-\begin{itemize}
-    \item \textit{scDblFinder} was used for doublet identification
-    \item the default filter sets (see below) were applied
-    \item standard \textit{Seurat} normalization was employed
-    \item \textit{Seurat} ($\geq$3.0) variable feature selection was employed, selecting 2000 genes
-    \item standard \textit{Seurat} scaling and PCA were employed
-    \item To study the impact of upstream steps on clustering, various \textit{Seurat} clustering analyses were performed, selecting different numbers of dimensions (5, 10, 15, 20, 30, and 50) and using various resolution parameters (0.005, 0.01, 0.02, 0.05, 0.1, 0.15, 0.2, 0.3, 0.4, 0.5, 0.8, 1, 1.2, 1.5, 2, 4).
-The range of resolution parameters was selected to ensure that the right number of clusters could be obtained in all datasets.
-\end{itemize}
-
-\subsection*{Denoising/imputation}
-Most imputation/denoising methods were run with the default parameters with the following exceptions; \textit{DrImpute} documentation advises to process the data prior to imputation by removing lowly expressed genes and cells expressing less than 2 genes.
-As it is not clear if this step is a hard requirement for the method to perform well, \textit{DrImpute} was run with and without prior processing (\textit{DrImpute\_process} and \textit{DrImpute\_noprocess} labels, respectively).
-The \textit{dca} method only accepts integers counts, while two of the datasets had non-integer quantification of expected counts.
-For these datasets, we rounded up the counts prior to imputation.
-\textit{alra} is designed for normalized data but as we are evaluating normalization downstream to imputation, we used the method on both non-normalized (\textit{alra} label) and normalized counts (\textit{alra\_norm} label).
-\textit{scImpute} requires an estimation of the expected number of clusters with the input data.
-As the estimation of the true number of cluster may not be known by the user, we evaluated the tool using the true number of clusters (\textit{scImpute} label) and using an over/under-estimation of this number (\textit{scImpute\_plus5} and \textit{scImpute\_min5} labels, respectively).
-\textit{ENHANCE} uses a k-nearest neighbor aggregation method and automatically estimates the number of neighbors to merge prior to the imputation.
-With the smallest datasets, this parameters was estimated to be 1, which lead to an early stop of the function.
-In such cases, we manually set this parameter to 2 for the method to work.
-
-\subsection*{Dimensionality estimates}
-
-For the methods that produce local dimensionality estimates, we used the maximum.
-For the elbow method, we implemented an automatic procedure by taking the farthest point from a line drawn between the variance explained by the first and last (i.e.
-50th) components calculated.
-For the \texttt{JackStraw} method, since the \textit{Seurat} documentation advises not to use a p-value threshold (which would typically yield a very large number of dimensions) but rather look for a drop in significance, we applied the same farthest point algorithm on the log10(p-values), which in our hands reproduced manual threshold selection.
-
-\subsection*{Filter sets}
-The \textit{default} set of filters excludes cells that are outliers according to at least two of the following thresholds: log10\_total\_counts $>$2.5 MADs or $<$5 MADs, log10\_total\_features $>$2.5 MADs or $<$5 MADs, pct\_counts\_in\_top\_20\_features $>$ or $<$ 5 MADs, featcount\_dist (distance to expected ratio of log10 counts and features) $>$ or $<$ 5 MADs, pct\_counts\_Mt $>$ 2.5 MADs and $>$ 0.08.
-
-The \textit{stringent} set of filters uses the same thresholds, but excludes a cell if it is an outlier on any single distribution.
-
-The \textit{lenient} set of filters excludes cells that are outliers on at least two distributions by at least 5 MADs, except for pct\_counts\_Mt where the threshold is $>$ 3 MADs and $>$ 0.08.
-
-For cluster-wise filters, clusters were first identified with \textit{scran::quickCluster} and the filters were then applied separately for each cluster.
-
-The `pca' and `pca2' clusters refer to the multivariate outlier detection methods implemented in \textit{scater}, running \textit{runPCA} with \texttt{use\_coldata=TRUE, detect\_outliers=TRUE}.
-`pca` uses all covariates, while `pca2` uses only the log10(counts), log10(features), proportion mitochondrial and proportion in the top 50 features.
-
-\subsection*{Variance and deviance explained}
-
-Unless specified otherwise, the variance in gene expression explained by subpopulations was calculated on the data normalized and transformed through \textit{sctransform}.
-For each gene, we fitted a linear model using the subpopulation as only independent variable ($\sim subpopulation$) and used the R-squared as a measure of the variance explained.
-The same method was used for principal components.
-
-The deviance explained by subpopulations was calculated directly on counts using the \linebreak \texttt{getDevianceExplained} function of \texttt{pipeComp}.
-The function uses \textit{edgeR} to fit two models, a full ($\sim librarySize + population$) and a reduced one ($\sim librarySize$).
-For each gene, the deviance explained is then the difference between the deviance of each models, divided by the deviance of the reduced model.
-In the rare cases where this resulted in a negative deviance explained, it was set to zero.
-
-To estimate the correlation between the principal components and covariates such as library size, we first fitted a linear model on the subpopulations and correlated the residuals of this model with the covariate of interest.
-
-\subsection*{\textit{scVI} normalization and LD}
-
-The evaluation of \textit{scVI} algorithm was performed separate to the other pipeline combinations as the model training on our reference datasets was especially resource-consuming.
-As it was not practically feasible to perform the training step on many combination of parameters as it was done with other tools, we evaluated it on a representative subset of pipelines using \textit{sctransform} normalization and \textit{Seurat}'s PCA dimension reduction method.
-The resulting evaluation is available in the Supplementary Figure 12.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% Backmatter begins here                   %%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{backmatter}
-
-\section*{Competing interests}
-As developers of the \textit{scDblFinder} package, the authors have an interest in it performing well. The authors declare no further competing interests.
-
-\section*{Author's contributions}
-PLG developed the \textit{pipeComp} framework and implemented the evaluation metrics.
-PLG and AS prepared pipelines, methods wrappers and performed the analyses.
-AS performed all denoising/imputation analyses.
-PLG, AS and MDR wrote the manuscript.
-
-\section*{Acknowledgements}
-This work was supported by the Swiss National Science Foundation (grants 310030\_175841, CRSII5\_177208) as well as the Chan Zuckerberg Initiative DAF (grant number 2018-182828), an advised fund of Silicon Valley Community Foundation.
-MDR acknowledges support from the University Research Priority Program Evolution in Action at the University of Zurich.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                  The Bibliography                       %%
-%%                                                         %%
-%%  Bmc_mathpys.bst  will be used to                       %%
-%%  create a .BBL file for submission.                     %%
-%%  After submission of the .TEX file,                     %%
-%%  you will be prompted to submit your .BBL file.         %%
-%%                                                         %%
-%%                                                         %%
-%%  Note that the displayed Bibliography will not          %%
-%%  necessarily be rendered by Latex exactly as specified  %%
-%%  in the online Instructions for Authors.                %%
-%%                                                         %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-% if your bibliography is in bibtex format, use those commands:
-\bibliographystyle{bmc-mathphys} % Style BST file (bmc-mathphys, vancouver, spbasic).
-\bibliography{bmc_article}      % Bibliography file (usually '*.bib' )
-% for author-year bibliography (bmc-mathphys or spbasic)
-% a) write to bib file (bmc-mathphys only)
-% @settings{label, options="nameyear"}
-% b) uncomment next line
-%\nocite{label}
-
-% or include bibliography directly:
-% \begin{thebibliography}
-% \bibitem{b1}
-% \end{thebibliography}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                               %%
-%% Figures                       %%
-%%                               %%
-%% NB: this is for captions and  %%
-%% Titles. All graphics must be  %%
-%% submitted separately and NOT  %%
-%% included in the Tex document  %%
-%%                               %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%
-%% Do not use \listoffigures as most will included as separate files
-
-% \section*{Figures}
-%   \begin{figure}[h!]
-%   \caption{\csentence{Sample figure title.}
-%       A short description of the figure content
-%       should go here.}
-%       \end{figure}
-% 
-% \begin{figure}[h!]
-%   \caption{\csentence{Sample figure title.}
-%       Figure legend text.}
-%       \end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                               %%
-%% Tables                        %%
-%%                               %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%% Use of \listoftables is discouraged.
-%%
-\section*{Tables}
-
-\begin{table}[h!]
-\caption{Overview of the benchmark datasets.}
-\label{tab:table1}
-\begin{tabular}{rlrrl}
-  \hline
-dataset & source & protocol & description \\ 
-  \hline
-Koh & GSE85066 & SMARTer & FACS purified H7 hESC in different differention stages \\ 
-  Kumar & GSE60749 & SMARTer & Mouse ESC cultured in different conditions \\ 
-  Zhengmix4eq & \cite{duoClustering2018} & 10x & Mixtures of FACS purified PBMCs \\ 
-  Zhengmix4uneq & \cite{duoClustering2018} & 10x & Mixtures of FACS purified PBMCs \\ 
-  Zhengmix8eq & \cite{duoClustering2018} & 10x & Mixtures of FACS purified PBMCs \\ 
-  mixology10x3cl & \cite{tianMixology2018} & 10x & Mixture of 3  cancer cell lines from CellBench \\ 
-  mixology10x5cl & \cite{tianMixology2018} & 10x & Mixture of 5 cancer cell lines from CellBench \\ 
-  simMix1 & - & 10x-based & Simulation of 10 human cell subpopulations \\
-  simMix2 & - & 10x-based & Simulation of 9 mouse cell subpopulations \\
-   \hline
-\end{tabular}
-\end{table}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                               %%
-%% Additional Files              %%
-%%                               %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section*{Additional Files}
-  \subsection*{Supplementary figures 1-21}
-    % Additional file descriptions text
-
-\end{backmatter}
 \end{document}

### 2020-03-02 bmc_article.tex

@@
 
+\iffalse
+
 \begin{figure}
@@
     \centering
-    \includegraphics[scale=0.7, keepaspectratio]{main_figures/pipeline_explanation.png}
+    \includegraphics[scale=0.7, keepaspectratio]{main_figures/pipeline_explanation.pdf}
     \caption{\textbf{Overview of the \textit{pipeComp} framework and its application to a scRNAseq clustering pipeline.
@@
 \end{figure}
+
+\fi
 
@@
 
+\iffalse
 \begin{figure}
     \centering
-    \includegraphics[scale=0.65, keepaspectratio]{{main_figures/doublets_files/figure-html/figure1-1.png}}
+    \includegraphics[scale=0.65, keepaspectratio]{{main_figures/doublets.pdf}}
     \caption{\textbf{Identification of doublet cells.
@@
 \end{figure}
+\fi
 
@@
 
+\iffalse
 \begin{figure}
     \centering
-    \includegraphics[scale=0.60, keepaspectratio]{{main_figures/filtering_files/figure-html/figure2-1.png}}
+    \includegraphics[scale=0.60, keepaspectratio]{{main_figures/filtering.pdf}}
     \caption{\textbf{Effect of filtering on cell subpopulation structure and clustering.
@@
 We next investigated the impact of different normalization strategies.
-Beside the standard log-normalization included in \textit{Seurat}, we tested \textit{scran}'s pooling-based normalization \cite{lunPooling2016}, \textit{sctransform}'s variance-stabilizing transformation \cite{hafemeisterSCtransform2019}, and normalization based on stable genes \cite{linStableGenes2018, deekeStablyExpressed2018}.
+Beside the standard log-normalization included in \textit{Seurat}, we tested \textit{scran}'s pooling-based normalization \cite{lunPooling2016}, \textit{sctransform}'s variance-stabilizing transformation \cite{hafemeisterSCtransform2019}, normalization based on stable genes \cite{linStableGenes2018, deekeStablyExpressed2018}, and SCnorm \cite{bacherSCnorm2017} (\textit{scVI} \cite{lopezDeep2018} was included separately on a subset of parameter combinations due to high running times).
 In addition to log-normalization, the standard \textit{Seurat} clustering pipeline performs per-feature unit-variance scaling so that the PCA is not too strongly dominated by highly-expressed features.
 We therefore included versions of the different normalization procedures, with or without a subsequent scaling (\textit{sctransform}'s variance-stabilizing transformation involves an approach analogous to scaling).
-\textit{Seurat}'s scaling function also includes the option to regress out the effect of certain covariates.
+\textit{Seurat}'s scaling function (and \textit{sctransform}) also includes the option to regress out the effect of certain covariates.
 We tested its performance by using the proportion of mitochondrial counts and the number of detected features as covariates.
@@
 However, as library size tends to vary across subpopulations, part of this effect can simply reflect biological differences.
-We therefore assessed to what extent the first principal component still retained a correlation with the library size and the number of detected features, removing the confounding covariation with the subpopulations (Figure \ref{fig:figure5}A).
-The simple step of scaling tended to remove much of the correlation with these features and, in the absence of scaling, standard \textit{Seurat} normalization resulted in fairly high correlation with technical covariates.
-\textit{sctransform} led to the lowest correlation but most methods, including normalization based on stable genes, were able to remove most of the correlation when combined with scaling.
-The only exception is \textit{Seurat} normalization with scaling regressing out covariates, which surprisingly increased the correlation in 8 of the 9 datasets.
+We therefore assessed to what extent the first principal component still retained a correlation with the library size and the number of detected features, removing the confounding covariation with the subpopulations (Figure \ref{fig:figure5}A and Supplementary Figure 12).
+Scaling tended to remove much of the correlation with these features, and while most methods were able to remove most of the effect, the correlation was lowest with \textit{sctransform}.
+Surprisingly, however, regressing out the covariates tended to increase the association with it.
 
@@
 We found most methods (including no normalization at all) to perform fairly well in most of the subpopulations.
-An exception was \textit{scVI} (Supplementary Figure 12), which performed poorly on most datasets in terms of both the subpopulation silhouette and clustering accuracy.
+An exception was \textit{scVI} (Supplementary Figure 13), which performed poorly on most datasets in terms of both the subpopulation silhouette and clustering accuracy.
 For time and resource cost reasons, \textit{scVI} was run on a reduced set of pipelines and compared to \textit{sctransform} (more details in the Methods section).
@@
 
+\iffalse
 \begin{figure}
     \centering
-    \includegraphics[scale=0.55, keepaspectratio]{{main_figures/normalization_files/figure-html/figure3-1.png}}
-    \caption{\textbf{Evaluation of normalization procedures.} `Seurat.feat\_mt\_regress' regresses out the number of features and proportion of mitochondrial reads during scaling; `Seurat.feat\_regress` regresses out the number of features only.
-Residual correlation of the first principal component with library size (left) and detection rate (right), after accounting for biological differences between subpopulations.
+    \includegraphics[scale=0.55, keepaspectratio]{{main_figures/normalization.pdf}}
+    \caption{\textbf{Evaluation of normalization procedures.} \textbf{A:} Variance in PC1 explained by library size (left) and detection rate (right) after accounting for subpopulation differences.
 \textbf{B:} Average silhouette width per true subpopulation, where higher silhouette width means a higher separability.
-\textbf{C:} Clustering accuracy (\textit{Seurat} clustering), measured by the ARI at the true number of cluster (left) and by the average MI of the cluster assignment with true subpopulations.}
+\textbf{C:} Clustering accuracy (\textit{Seurat} clustering), measured by mutual information (MI), Adjusted Rand Index (ARI) and ARI at the true number of cluster.
+'feat_regress', 'mt_regress' and 'feat_mt_regress' respectively stand for the regressing out of the number of features, the proportion of mitochondrial reads, or both during scaling.
+}
     \label{fig:figure5}
 \end{figure}
+\fi
 
 Finally, we monitored whether, under the same downstream clustering analysis, different normalization methods tended to lead to an over- or under-estimation of the number of clusters.
-Although some methods had a tendency to lead to a higher (e.g., \textit{sctransform}) or lower (e.g., stable genes) number of clusters, the effect was very mild and not entirely systematic (Supplementary Figure 13).
-We also confirmed that, especially in the UMI-based datasets, \textit{sctransform} did in fact successfully stabilize variance across mean counts (Supplementary Figure 14).
+Although some methods had a tendency to lead to a higher (e.g., \textit{sctransform}) or lower (e.g., stable genes) number of clusters, the effect was very mild and not entirely systematic (Supplementary Figure 14).
+We also confirmed that, especially in the UMI-based datasets, \textit{sctransform} did in fact successfully stabilize variance across mean counts (Supplementary Figure 15), at little apparent cost on computation time (Supplementary Figure 16).
 
@@
 As the proportion of variability in a gene attributable to subpopulations can be measured in various ways, we first compared three approaches: ANOVA on log-normalized count, ANOVA on \textit{sctransform} normalization, and deviance explained.
-The ANOVAs performed on a standard \textit{Seurat} normalization and on \textit{sctransform} data were highly correlated (Supplementary Figure 15A).
-These estimates were also in good agreement with the deviance explained, although lowly-expressed genes could have a high deviance explained without having much of their variance explained by subpopulation (Supplementary Figure 15B-D).
-We therefore compared the proportion of the cumulative variance/deviance explained by the top X genes that could be retrieved by each gene ranking method (Supplementary Figures 16-17).
-We first focused on the first 1000 genes to highlight the differences between methods, although a higher number of selected genes decreased the differences between methods (Supplementary Figures 16-18).
+The ANOVAs performed on a standard \textit{Seurat} normalization and on \textit{sctransform} data were highly correlated (Supplementary Figure 17A).
+These estimates were also in good agreement with the deviance explained, although lowly-expressed genes could have a high deviance explained without having much of their variance explained by subpopulation (Supplementary Figure 17B-D).
+We therefore compared the proportion of the cumulative variance/deviance explained by the top X genes that could be retrieved by each gene ranking method (Supplementary Figures 18-19).
+We first focused on the first 1000 genes to highlight the differences between methods, although a higher number of selected genes decreased the differences between methods (Supplementary Figures 18-20).
 The standardized measures of variability were systematically worse than their non-standardized counterparts in selecting genes with a high proportion of variance explained by subpopulation.
-Regarding the percentage of deviance explained however, the standardized measures were often superior (Figure \ref{fig:figure6}A and Supplementary Figures 16-17).
+Regarding the percentage of deviance explained however, the standardized measures were often superior (Figure \ref{fig:figure6}A and Supplementary Figures 18-19).
 Deviance proved the method of choice to prioritize genes with a high variance explained by subpopulations (with mere expression level proving surprisingly performant) but did not perform so well to select genes with a high deviance explained.
 
+\iffalse
 \begin{figure}
     \centering
-    \includegraphics[scale=0.65]{{main_figures/selection_files/figure-html/figure4-1.png}}
+    \includegraphics[scale=0.65]{{main_figures/selection.pdf}}
     \caption{\textbf{Evaluation of feature selection methods.
@@
 In general, we found deviance and unstandardized estimates of variance to provide the best results across datasets and normalization methods.
-Increasing the number of features selected also systematically led to an increase in the accuracy of the clustering, typically plateauing after 4000 features (Supplementary Figure 18).
+Increasing the number of features selected also systematically led to an increase in the accuracy of the clustering, typically plateauing after 4000 features (Supplementary Figure 20).
 
@@
 Like GLM-PCA, \textit{scVI}'s Linear Decoder (LD) does not explicitly rely on normalized counts.
-Its performance was lower than a pipeline with non-weighted \textit{Seurat} PCA on all measures for datasets with more subpopulations (Supplementary Figure 12).
+Its performance was lower than a pipeline with non-weighted \textit{Seurat} PCA on all measures for datasets with more subpopulations (Supplementary Figure 13).
 Overall, weighting the principal components by their variance (as \textit{Seurat} does) had a positive impact on silhouette widths and ARI scores.
 
+\iffalse
 \begin{figure}
     \centering
-    \includegraphics[scale=0.40, keepaspectratio]{{main_figures/dr_files/figure-html/dimred-1.png}}
+    \includegraphics[scale=0.40, keepaspectratio]{{main_figures/dr.pdf}}
     \caption{\textbf{Evaluation of common dimensionality reduction methods.
@@
 Since Euclidean distance decreases as the number of non-discriminating dimensions increases, there is usually a trade-off between selecting enough dimensions to keep most information and excluding smaller dimensions that may represent technical noise or other unwanted sources of variation.
-Overall, increasing the number of dimensions robustly led to a decrease in the number of clusters (Supplementary Figures 13 and 19).
-This tended to affect the accuracy of the clustering (Supplementary Figure 20), although in both cases (number of clusters and ARI), \textit{Seurat}'s resolution parameter had a much stronger impact.
+Overall, increasing the number of dimensions robustly led to a decrease in the number of clusters (Supplementary Figures 14 and 21).
+This tended to affect the accuracy of the clustering (Supplementary Figure 22), although in both cases (number of clusters and ARI), \textit{Seurat}'s resolution parameter had a much stronger impact.
 
@@
 Reasoning that over-estimating the number of dimensions is less problematic than under-estimating it, we kept the former methods for a full analysis of their impact on clustering (Figure \ref{fig:figure8}C-D), when combined with \textit{sctransform} or the standard \textit{Seurat} normalization.
-Although most methods performed well on the various clustering measures, the global maximum likelihood based on translated Poisson Mixture Model (\textit{maxLikGlobal}, using the 20 nearest neighbors, although estimates were relatively stable across decent values of \textit{k}; see Supplementary Figure 21) provided the dimensionality estimate that best separated the subpopulations (Figure \ref{fig:figure8}C) and resulted in the best clustering accuracy (Figure \ref{fig:figure8}D).
+Although most methods performed well on the various clustering measures, the global maximum likelihood based on translated Poisson Mixture Model (\textit{maxLikGlobal}, using the 20 nearest neighbors, although estimates were relatively stable across decent values of \textit{k}; see Supplementary Figure 23) provided the dimensionality estimate that best separated the subpopulations (Figure \ref{fig:figure8}C) and resulted in the best clustering accuracy (Figure \ref{fig:figure8}D).
 This method systematically selected many more components than were associated with the subpopulations, suggesting that although these additional components appear individually uninformative, in combination they nevertheless contribute to classification.
@@
 
+\iffalse
 \begin{figure}
     \centering
-    \includegraphics[scale=0.53, keepaspectratio]{{main_figures/dimensionality_files/figure-html/figure6-1.png}}
+    \includegraphics[scale=0.53, keepaspectratio]{{main_figures/dimensionality.pdf}}
     \caption{\textbf{Estimating dimensionality.
@@
 Finally, the ARI score at the true number of clusters, when available, showed similar performances.
-However, because \textit{Seurat}'s resolution parameter had a large impact on the number of clusters identified (Supplementary Figures 1 and 21), \textit{Seurat} could always be coerced into producing the right number of clusters.
-Instead, the number of clusters found by \textit{scran} was considerably less influenced by the available parameters (number of nearest neighbors or steps in the random walk - see Supplementary Figure 22), and as a result \textit{scran}-based clustering sometimes never produced a partitioning with the right number of clusters.
+However, because \textit{Seurat}'s resolution parameter had a large impact on the number of clusters identified (Supplementary Figures 1 and 23), \textit{Seurat} could always be coerced into producing the right number of clusters.
+Instead, the number of clusters found by \textit{scran} was considerably less influenced by the available parameters (number of nearest neighbors or steps in the random walk - see Supplementary Figure 24), and as a result \textit{scran}-based clustering sometimes never produced a partitioning with the right number of clusters.
 This observation, along with \textit{scran}'s higher MI score, suggest that \textit{scran} sometimes simply divides a real subpopulation into two clusters (possibly tracking some unknown biological differences) rather than committing misclassification errors.
@@
 
+\iffalse
 \begin{figure}
     \centering
-    \includegraphics[scale=0.41, keepaspectratio]{{main_figures/clustering_files/figure-html/clustering-1.png}}
+    \includegraphics[scale=0.34, keepaspectratio]{{main_figures/clustering.pdf}}
     \caption{\textbf{Evaluation of clustering methods.} Clustering accuracy (\textbf{A}) and run time (\textbf{B}) of \textit{scran}- and \textit{Seurat}-based clustering tools in combination with \textit{sctransform} and standard \textit{Seurat}'s normalization.}
@@
 
+\iffalse
 \begin{figure}
     \centering
-    \includegraphics[scale=0.49, keepaspectratio]{{main_figures/imputation_files/figure-html/imputation-1.png}}
+    \includegraphics[scale=0.49, keepaspectratio]{{main_figures/imputation.pdf}}
      \caption{\textbf{Evaluation of imputation/denoising methods} Average silhouette width per subpopulation (\textbf{A}) and clustering accuracy (\textbf{B}) with or without (indicated as \textit{none}) application of a denoising/imputation method.}
@@
 
+\iffalse
 \begin{figure}[h]
     \centering
-    \includegraphics[scale=0.08,keepaspectratio]{main_figures/pipeComp_summary.png}
+    \includegraphics[scale=0.9,keepaspectratio]{main_figures/pipeComp_summary.pdf}
     \caption{\textbf{Recommendations of tools from filtering to clustering.}}
@@
 As it was not practically feasible to perform the training step on many combination of parameters as it was done with other tools, we evaluated it on a representative subset of pipelines using \textit{sctransform} normalization and \textit{Seurat}'s PCA dimension reduction method.
-The resulting evaluation is available in the Supplementary Figure 12.
+The resulting evaluation is available in the Supplementary Figure 13.
 
@@
 
-% \section*{Figures}
-%   \begin{figure}[h!]
-%   \caption{\csentence{Sample figure title.}
-%       A short description of the figure content
-%       should go here.}
-%       \end{figure}
-% 
-% \begin{figure}[h!]
-%   \caption{\csentence{Sample figure title.}
-%       Figure legend text.}
-%       \end{figure}
+\section*{Figures}
+
+\begin{figure}[h!]
+    \centering
+    \caption{\textbf{Overview of the benchmark datasets used.}}
+    \label{fig:figure1}
+\end{figure}
+
+\begin{figure}[h!]
+    \centering
+    \caption{\textbf{Overview of the \textit{pipeComp} framework and its application to a scRNAseq clustering pipeline.
+A:} The package is built around a \texttt{PipelineDefinition} S4 class which defines a set of functions to be executed consecutively, as well as optional evaluation functions for each step.
+Each step can accept a number of parameters whose alternative values are provided as a list.
+All (or subsets of) combinations of parameters can then be simultaneously ran and evaluated using the \texttt{runPipeline} function.
+\textbf{B:} Scheme representing the application of \textit{pipeComp} to evaluate a range of methods that are commonly used in scRNAseq studies and some of the metrics monitored at various steps.}
+    \label{fig:figure2}
+\end{figure}
+
+\begin{figure}[h!]
+    \centering
+    \caption{\textbf{Identification of doublet cells.
+A:} Receiver operating characteristic (ROC) curves of the tested doublet detection methods for three datasets with SNP-identified doublets.
+Dots indicate threshold determined by the true number of doublets.
+\textbf{B:} Running time of the different methods (\textit{DoubletFinder} failed on one of the datasets).
+\textbf{C:} Rate of misclassification of the cells identified by \textit{scDblFinder} as doublets (DBL) or singlets (SNG), across a large range of clustering analysis.
+Even in datasets which should not have neotypic doublets, the cells identified as such tend to be misclassified.
+\label{fig:figure3}}
+\end{figure}
+
+\begin{figure}[h!]
+    \centering
+    \caption{\textbf{Effect of filtering on cell subpopulation structure and clustering.
+A:} Filtering on the basis of distance to the whole distribution can lead to strong bias against certain subpopulations.
+The dashed line indicates a threshold of 2.5 median absolute deviations (MADs) from the median of the overall population.
+\textbf{B:} Relationship between the maximum subpopulation exclusion rate and the average clustering accuracy per subpopulation across various filtering strategies.
+Of note, doublet removal appears to be desirable even when, due to the design, there are no heterotypic doublets in the data.
+The PCA methods refer to multivariate outlier detected as implemented in \textit{scater} (see Methods for details).}
+    \label{fig:figure4}
+\end{figure}
+
+\begin{figure}[h!]
+    \centering
+    \caption{\textbf{Evaluation of normalization procedures.} \textbf{A:} Variance in PC1 explained by library size (left) and detection rate (right) after accounting for subpopulation differences.
+\textbf{B:} Average silhouette width per true subpopulation, where higher silhouette width means a higher separability.
+\textbf{C:} Clustering accuracy (\textit{Seurat} clustering), measured by mutual information (MI), Adjusted Rand Index (ARI) and ARI at the true number of cluster.
+'feat_regress', 'mt_regress' and 'feat_mt_regress' respectively stand for the regressing out of the number of features, the proportion of mitochondrial reads, or both during scaling.
+}
+    \label{fig:figure5}
+\end{figure}
+
+\begin{figure}[h!]
+    \centering
+    \caption{\textbf{Evaluation of feature selection methods.
+A:} Ability of different feature ranking methods to capture genes with a high proportion of variance (left) or deviance (right) explained by real subpopulations.
+The asterisk denotes the default Seurat method.
+\textbf{B:} Accuracy of clusterings (at the true number of clusters) when selecting 1000 genes using the given methods.
+Based on standard \textit{Seurat} normalization (left) or \textit{sctransform} (right).
+The  \textit{vst.varExp} and \textit{devianceExplained} methods correspond to the estimates used in \textbf{A} to evaluate the selection methods and were included here only for validation purpose.}
+    \label{fig:figure6}
+\end{figure}
+
+\begin{figure}[h!]
+    \centering
+    \caption{\textbf{Evaluation of common dimensionality reduction methods.
+A:} Average silhouette width per subpopulation resulting from combinations of normalization and dimension reductions.
+\textbf{B:} Clustering accuracy, mean proportion of the variance in the first 5 components explained by real subpopulations (center), and median ARI of the resulting clustering (right).}
+    \label{fig:figure7}
+\end{figure}
+
+\begin{figure}[h!]
+    \centering
+    \caption{\textbf{Estimating dimensionality.
+A:} Estimated dimensionality using the proportion of variance in each component explained by the true subpopulations.
+\textbf{B:} Difference between `real' dimensionality (from \textbf{A}) and dimensionality estimation methods, along with computing time.
+\textbf{C:} Average silhouette width per subpopulation across a selection of methods, combined with \textit{sctransform} or standard \textit{Seurat} normalization.
+\textbf{D:} Clustering accuracy across the normalization/dimensionality estimation methods.}
+    \label{fig:figure8}
+\end{figure}
+
+\begin{figure}[h!]
+    \centering
+    \caption{\textbf{Evaluation of clustering methods.} Clustering accuracy (\textbf{A}) and run time (\textbf{B}) of \textit{scran}- and \textit{Seurat}-based clustering tools in combination with \textit{sctransform} and standard \textit{Seurat}'s normalization.}
+    \label{fig:figure9}
+\end{figure}
+
+\begin{figure}[h!]
+    \centering
+    \caption{\textbf{Evaluation of imputation/denoising methods} Average silhouette width per subpopulation (\textbf{A}) and clustering accuracy (\textbf{B}) with or without (indicated as \textit{none}) application of a denoising/imputation method.}
+    \label{fig:figure10}
+\end{figure}
+
+\begin{figure}[h!]
+    \centering
+    \caption{\textbf{Recommendations of tools from filtering to clustering.}}
+    \label{fig:figure11}
+\end{figure}
 
@@
 \section*{Additional Files}
-  \subsection*{\textbf{Additional file 1:Supplementary figures 1-22}}
+  \subsection*{\textbf{Additional file 1:Supplementary figures 1-24}}
   Additional results of evaluations across datasets among various metrics. (PDF 15 MB)

### 2020-03-02 bmc_article.tex
Updates from Overleaf

@@
 
-\iffalse
-
-\begin{figure}
-    \centering
-    \includegraphics[width=\textwidth,keepaspectratio]{dataset_description}
-    \caption{\textbf{Overview of the benchmark datasets used.}}
-    \label{fig:figure1}
-\end{figure}
-
-\begin{figure}
-    \centering
-    \includegraphics[scale=0.7, keepaspectratio]{main_figures/pipeline_explanation.pdf}
-    \caption{\textbf{Overview of the \textit{pipeComp} framework and its application to a scRNAseq clustering pipeline.
-A:} The package is built around a \texttt{PipelineDefinition} S4 class which defines a set of functions to be executed consecutively, as well as optional evaluation functions for each step.
-Each step can accept a number of parameters whose alternative values are provided as a list.
-All (or subsets of) combinations of parameters can then be simultaneously ran and evaluated using the \texttt{runPipeline} function.
-\textbf{B:} Scheme representing the application of \textit{pipeComp} to evaluate a range of methods that are commonly used in scRNAseq studies and some of the metrics monitored at various steps.}
-    \label{fig:figure2}
-\end{figure}
-
-\fi
-
 \subsection*{Filtering}
@@
 We also found that \textit{scDblFinder} improved the accuracy of the clustering across all benchmark datasets even when, by design, the data contained no heterotypic doublet (Figure \ref{fig:figure4}).
-
-\iffalse
-\begin{figure}
-    \centering
-    \includegraphics[scale=0.65, keepaspectratio]{{main_figures/doublets.pdf}}
-    \caption{\textbf{Identification of doublet cells.
-A:} Receiver operating characteristic (ROC) curves of the tested doublet detection methods for three datasets with SNP-identified doublets.
-Dots indicate threshold determined by the true number of doublets.
-\textbf{B:} Running time of the different methods (\textit{DoubletFinder} failed on one of the datasets).
-\textbf{C:} Rate of misclassification of the cells identified by \textit{scDblFinder} as doublets (DBL) or singlets (SNG), across a large range of clustering analysis.
-Even in datasets which should not have neotypic doublets, the cells identified as such tend to be misclassified.
-\label{fig:figure3}}
-\end{figure}
-\fi
 
@@
 We therefore recommend the use of doublet removal followed by relatively mild filtering, such as is implemented in our `default' filtering (see Methods).
-
-\iffalse
-\begin{figure}
-    \centering
-    \includegraphics[scale=0.60, keepaspectratio]{{main_figures/filtering.pdf}}
-    \caption{\textbf{Effect of filtering on cell subpopulation structure and clustering.
-A:} Filtering on the basis of distance to the whole distribution can lead to strong bias against certain subpopulations.
-The dashed line indicates a threshold of 2.5 median absolute deviations (MADs) from the median of the overall population.
-\textbf{B:} Relationship between the maximum subpopulation exclusion rate and the average clustering accuracy per subpopulation across various filtering strategies.
-Of note, doublet removal appears to be desirable even when, due to the design, there are no heterotypic doublets in the data.
-The PCA methods refer to multivariate outlier detected as implemented in \textit{scater} (see Methods for details).}
-    \label{fig:figure4}
-\end{figure}
-\fi
 
@@
 
-\iffalse
-\begin{figure}
-    \centering
-    \includegraphics[scale=0.55, keepaspectratio]{{main_figures/normalization.pdf}}
-    \caption{\textbf{Evaluation of normalization procedures.} \textbf{A:} Variance in PC1 explained by library size (left) and detection rate (right) after accounting for subpopulation differences.
-\textbf{B:} Average silhouette width per true subpopulation, where higher silhouette width means a higher separability.
-\textbf{C:} Clustering accuracy (\textit{Seurat} clustering), measured by mutual information (MI), Adjusted Rand Index (ARI) and ARI at the true number of cluster.
-'feat_regress', 'mt_regress' and 'feat_mt_regress' respectively stand for the regressing out of the number of features, the proportion of mitochondrial reads, or both during scaling.
-}
-    \label{fig:figure5}
-\end{figure}
-\fi
-
 Finally, we monitored whether, under the same downstream clustering analysis, different normalization methods tended to lead to an over- or under-estimation of the number of clusters.
@@
 
-\iffalse
-\begin{figure}
-    \centering
-    \includegraphics[scale=0.65]{{main_figures/selection.pdf}}
-    \caption{\textbf{Evaluation of feature selection methods.
-A:} Ability of different feature ranking methods to capture genes with a high proportion of variance (left) or deviance (right) explained by real subpopulations.
-The asterisk denotes the default Seurat method.
-\textbf{B:} Accuracy of clusterings (at the true number of clusters) when selecting 1000 genes using the given methods.
-Based on standard \textit{Seurat} normalization (left) or \textit{sctransform} (right).
-The  \textit{vst.varExp} and \textit{devianceExplained} methods correspond to the estimates used in \textbf{A} to evaluate the selection methods and were included here only for validation purpose.}
-    \label{fig:figure6}
-\end{figure}
-\fi
-
 We next evaluated how the use of different feature selection methods affected the clustering accuracy (Figure \ref{fig:figure6}B).
@@
 Overall, weighting the principal components by their variance (as \textit{Seurat} does) had a positive impact on silhouette widths and ARI scores.
-
-\iffalse
-\begin{figure}
-    \centering
-    \includegraphics[scale=0.40, keepaspectratio]{{main_figures/dr.pdf}}
-    \caption{\textbf{Evaluation of common dimensionality reduction methods.
-A:} Average silhouette width per subpopulation resulting from combinations of normalization and dimension reductions.
-\textbf{B:} Clustering accuracy, mean proportion of the variance in the first 5 components explained by real subpopulations (center), and median ARI of the resulting clustering (right).}
-    \label{fig:figure7}
-\end{figure}
-\fi
 
@@
 An implication is that the commonly-used elbow method most likely underestimates dimensionality.
-
-\iffalse
-\begin{figure}
-    \centering
-    \includegraphics[scale=0.53, keepaspectratio]{{main_figures/dimensionality.pdf}}
-    \caption{\textbf{Estimating dimensionality.
-A:} Estimated dimensionality using the proportion of variance in each component explained by the true subpopulations.
-\textbf{B:} Difference between `real' dimensionality (from \textbf{A}) and dimensionality estimation methods, along with computing time.
-\textbf{C:} Average silhouette width per subpopulation across a selection of methods, combined with \textit{sctransform} or standard \textit{Seurat} normalization.
-\textbf{D:} Clustering accuracy across the normalization/dimensionality estimation methods.}
-    \label{fig:figure8}
-\end{figure}
-\fi
 
@@
 
-\iffalse
-\begin{figure}
-    \centering
-    \includegraphics[scale=0.34, keepaspectratio]{{main_figures/clustering.pdf}}
-    \caption{\textbf{Evaluation of clustering methods.} Clustering accuracy (\textbf{A}) and run time (\textbf{B}) of \textit{scran}- and \textit{Seurat}-based clustering tools in combination with \textit{sctransform} and standard \textit{Seurat}'s normalization.}
-    \label{fig:figure9}
-\end{figure}
-\fi
-
 \subsection*{Further extensions to the pipeline: imputation/denoising}
@@
 On the contrary, imputing on normalized counts originating from Smart-seq technology was instead rather deleterious to the clustering accuracy.
-
-\iffalse
-\begin{figure}
-    \centering
-    \includegraphics[scale=0.49, keepaspectratio]{{main_figures/imputation.pdf}}
-     \caption{\textbf{Evaluation of imputation/denoising methods} Average silhouette width per subpopulation (\textbf{A}) and clustering accuracy (\textbf{B}) with or without (indicated as \textit{none}) application of a denoising/imputation method.}
-    \label{fig:figure10}
-\end{figure}
-\fi
 
@@
 
-\iffalse
-\begin{figure}[h]
-    \centering
-    \includegraphics[scale=0.9,keepaspectratio]{main_figures/pipeComp_summary.pdf}
-    \caption{\textbf{Recommendations of tools from filtering to clustering.}}
-    \label{fig:figure11}
-\end{figure}
-\fi
-
 \subsection*{Limitations and open questions}
@@
 For instance, \textit{Seurat} and \textit{scran} methods were already shown to yield good clustering performance \cite{freytagComparison2018} and \textit{DrImpute} to improve the quality of downstream analyses \cite{ZhangImput2018}.
-A previous systematic comparison \cite{viethSystematic2019} instead found that the use of \textit{SAVER} for denoising data and \textit{scran} for clustering yielded better performance on the ability to recover differentially-expressed genes, whereas we focused on the ability to recover cell populations.
+A previous systematic comparison \cite{viethSystematic2019} instead found that the use of \textit{SAVERX} for denoising data and \textit{scran} for clustering yielded better performance on the ability to recover differentially-expressed genes, whereas we focused on the ability to recover cell populations.
 Another disagreement with the present recommendations come from the study of Hou et al.
@@
 
-We developed \texttt{pipeComp} to address the need of a framework that can simultaneously evaluate the interaction of multiple tools.
+We developed \textit{pipeComp} to address the need of a framework that can simultaneously evaluate the interaction of multiple tools.
 The work of Vieth and colleagues\cite{viethSystematic2019} already offered an important precedent in this respect, evaluating the interaction of various steps in the context of scRNAseq differential expression analysis, but did not offer a platform for doing so.
@@
 All analyses were performed through the \href{https://github.com/plger/pipeComp}{pipeComp} R package, which implements the pipeline framework described here.
-scVI results and results of the clustering section were produced with pipeComp 0.99.4, while all other results were produced with pipeComp 0.99.2 (\textit{biorxiv}) branch of the repository).
+scVI results and results of the clustering section were produced with \textit{pipeComp} 0.99.4, while all other results were produced with version 0.99.2 (\textit{biorxiv}) branch of the repository).
 All code to reproduce the simulations and figures is available in the \url{https://github.com/markrobinsonuzh/scRNA\_pipelines\_paper} repository, which also includes the basic datasets with a standardized annotation.
@@
 
-The deviance explained by subpopulations was calculated directly on counts using the \linebreak \texttt{getDevianceExplained} function of \texttt{pipeComp}.
+The deviance explained by subpopulations was calculated directly on counts using the \linebreak \texttt{getDevianceExplained} function of \textit{pipeComp}.
 The function uses \textit{edgeR} to fit two models, a full ($\sim librarySize + population$) and a reduced one ($\sim librarySize$).
@@
 
-\begin{figure}[h!]
+\showfigstrue
+
+\begin{figure}
     \centering
+    \ifshowfigs
+    \includegraphics[width=\textwidth,keepaspectratio]{dataset_description}
+    \fi
     \caption{\textbf{Overview of the benchmark datasets used.}}
@@
 
-\begin{figure}[h!]
+\begin{figure}
     \centering
+    \ifshowfigs
+    \includegraphics[scale=0.75, keepaspectratio]{main_figures/pipeline_explanation.pdf}
+    \fi
     \caption{\textbf{Overview of the \textit{pipeComp} framework and its application to a scRNAseq clustering pipeline.
@@
 
-\begin{figure}[h!]
+\begin{figure}
     \centering
+    \ifshowfigs
+    \includegraphics[scale=0.85, keepaspectratio]{{main_figures/doublets.pdf}}
+    \fi
     \caption{\textbf{Identification of doublet cells.
@@
 
-\begin{figure}[h!]
+\begin{figure}
     \centering
+    \ifshowfigs
+    \includegraphics[scale=0.72, keepaspectratio]{{main_figures/filtering.pdf}}
+    \fi
     \caption{\textbf{Effect of filtering on cell subpopulation structure and clustering.
@@
 
-\begin{figure}[h!]
+\begin{figure}
     \centering
-    \caption{\textbf{Evaluation of normalization procedures.} \textbf{A:} Variance in PC1 explained by library size (left) and detection rate (right) after accounting for subpopulation differences.
+    \ifshowfigs
+    \includegraphics[scale=0.6, keepaspectratio]{{main_figures/normalization.pdf}}
+    \fi
+    \caption{\textbf{Evaluation of normalization procedures.
+A:} Variance in PC1 explained by library size (left) and detection rate (right) after accounting for subpopulation differences.
 \textbf{B:} Average silhouette width per true subpopulation, where higher silhouette width means a higher separability.
 \textbf{C:} Clustering accuracy (\textit{Seurat} clustering), measured by mutual information (MI), Adjusted Rand Index (ARI) and ARI at the true number of cluster.
-'feat_regress', 'mt_regress' and 'feat_mt_regress' respectively stand for the regressing out of the number of features, the proportion of mitochondrial reads, or both during scaling.
-}
+`feat\_regress', `mt\_regress' and `feat\_mt\_regress' respectively stand for the regressing out of the number of features, the proportion of mitochondrial reads, or both during scaling.}
     \label{fig:figure5}
@@
 
-\begin{figure}[h!]
+\begin{figure}
     \centering
+    \ifshowfigs
+    \includegraphics[scale=0.65]{{main_figures/selection.pdf}}
+    \fi
     \caption{\textbf{Evaluation of feature selection methods.
@@
 
-\begin{figure}[h!]
+\begin{figure}
     \centering
+    \ifshowfigs
+    \includegraphics[scale=0.50, keepaspectratio]{{main_figures/dr.pdf}}
+    \fi
     \caption{\textbf{Evaluation of common dimensionality reduction methods.
@@
 
-\begin{figure}[h!]
+\begin{figure}
     \centering
+    \ifshowfigs
+    \includegraphics[scale=0.53, keepaspectratio]{{main_figures/dimensionality.pdf}}
+    \fi
     \caption{\textbf{Estimating dimensionality.
@@
 
-\begin{figure}[h!]
+\begin{figure}
     \centering
+    \ifshowfigs
+    \includegraphics[scale=0.43, keepaspectratio]{{main_figures/clustering.pdf}}
+    \fi
     \caption{\textbf{Evaluation of clustering methods.} Clustering accuracy (\textbf{A}) and run time (\textbf{B}) of \textit{scran}- and \textit{Seurat}-based clustering tools in combination with \textit{sctransform} and standard \textit{Seurat}'s normalization.}
@@
 
-\begin{figure}[h!]
+\begin{figure}
     \centering
+    \ifshowfigs
+    \includegraphics[scale=0.55, keepaspectratio]{{main_figures/imputation.pdf}}
+    \fi
     \caption{\textbf{Evaluation of imputation/denoising methods} Average silhouette width per subpopulation (\textbf{A}) and clustering accuracy (\textbf{B}) with or without (indicated as \textit{none}) application of a denoising/imputation method.}
@@
 
-\begin{figure}[h!]
+\begin{figure}[h]
     \centering
+    \ifshowfigs
+    \includegraphics[scale=0.9,keepaspectratio]{main_figures/pipeComp_summary.pdf}
+    \fi
     \caption{\textbf{Recommendations of tools from filtering to clustering.}}

### 2020-03-02 main_figures.tex
Updates from Overleaf

@@
 
+\showfigstrue
+
 \begin{figure}
     \centering
+    \ifshowfigs
     \includegraphics[width=\textwidth,keepaspectratio]{dataset_description}
+    \fi
     \caption{\textbf{Overview of the benchmark datasets used.}}
@@
     \centering
+    \ifshowfigs
     \includegraphics[scale=0.75, keepaspectratio]{main_figures/pipeline_explanation.pdf}
+    \fi
     \caption{\textbf{Overview of the \textit{pipeComp} framework and its application to a scRNAseq clustering pipeline.
@@
     \centering
+    \ifshowfigs
     \includegraphics[scale=0.85, keepaspectratio]{{main_figures/doublets.pdf}}
+    \fi
     \caption{\textbf{Identification of doublet cells.
@@
     \centering
+    \ifshowfigs
     \includegraphics[scale=0.72, keepaspectratio]{{main_figures/filtering.pdf}}
+    \fi
     \caption{\textbf{Effect of filtering on cell subpopulation structure and clustering.
@@
     \centering
-    \includegraphics[scale=0.65, keepaspectratio]{{main_figures/normalization.pdf}}
-    \caption{\textbf{Evaluation of normalization procedures.} \textbf{A:} Variance in PC1 explained by library size (left) and detection rate (right) after accounting for subpopulation differences.
+    \ifshowfigs
+    \includegraphics[scale=0.6, keepaspectratio]{{main_figures/normalization.pdf}}
+    \fi
+    \caption{\textbf{Evaluation of normalization procedures.
+A:} Variance in PC1 explained by library size (left) and detection rate (right) after accounting for subpopulation differences.
 \textbf{B:} Average silhouette width per true subpopulation, where higher silhouette width means a higher separability.
 \textbf{C:} Clustering accuracy (\textit{Seurat} clustering), measured by mutual information (MI), Adjusted Rand Index (ARI) and ARI at the true number of cluster.
-'feat_regress', 'mt_regress' and 'feat_mt_regress' respectively stand for the regressing out of the number of features, the proportion of mitochondrial reads, or both during scaling.
-}
+`feat\_regress', `mt\_regress' and `feat\_mt\_regress' respectively stand for the regressing out of the number of features, the proportion of mitochondrial reads, or both during scaling.}
     \label{fig:figure5}
@@
     \centering
+    \ifshowfigs
     \includegraphics[scale=0.65]{{main_figures/selection.pdf}}
+    \fi
     \caption{\textbf{Evaluation of feature selection methods.
@@
     \centering
+    \ifshowfigs
     \includegraphics[scale=0.50, keepaspectratio]{{main_figures/dr.pdf}}
+    \fi
     \caption{\textbf{Evaluation of common dimensionality reduction methods.
@@
     \centering
+    \ifshowfigs
     \includegraphics[scale=0.6, keepaspectratio]{{main_figures/dimensionality.pdf}}
+    \fi
     \caption{\textbf{Estimating dimensionality.
@@
     \centering
+    \ifshowfigs
     \includegraphics[scale=0.47, keepaspectratio]{{main_figures/clustering.pdf}}
+    \fi
     \caption{\textbf{Evaluation of clustering methods.} Clustering accuracy (\textbf{A}) and run time (\textbf{B}) of \textit{scran}- and \textit{Seurat}-based clustering tools in combination with \textit{sctransform} and standard \textit{Seurat}'s normalization.}
@@
     \centering
+    \ifshowfigs
     \includegraphics[scale=0.55, keepaspectratio]{{main_figures/imputation.pdf}}
+    \fi
     \caption{\textbf{Evaluation of imputation/denoising methods} Average silhouette width per subpopulation (\textbf{A}) and clustering accuracy (\textbf{B}) with or without (indicated as \textit{none}) application of a denoising/imputation method.}
@@
     \centering
+    \ifshowfigs
     \includegraphics[scale=0.9,keepaspectratio]{main_figures/pipeComp_summary.pdf}
+    \fi
     \caption{\textbf{Recommendations of tools from filtering to clustering.}}

### 2020-03-07 main_figures/clustering.Rmd
updated figures to pdf

@@
 date: "`r format(Sys.time(), '%d %B, %Y')`"
-output:
-  html_document:
-    keep_md: true
+output: html_document
 ---
@@
 
-res <- readRDS("../data/results_clustering2.rds")
+res <- readRDS("../data/results_clustering.rds")
 cl <- res$evaluation$clustering
@@
   x <- gsub("exact ","",x)
-  x <- gsub(" + walktrap","",x,fixed=TRUE)
+  x <- gsub("+walktrap","",x,fixed=TRUE)
 }
@@
 ```{r clustering, fig.width=12, fig.height=8}
+pdf("clustering.pdf", width=15, height=10)
+plot_grid(h2, h3, labels=LETTERS[1:2], ncol=1, rel_heights = c(7,4))
+dev.off()
 plot_grid(h2, h3, labels=LETTERS[1:2], ncol=1, rel_heights = c(7,4))

### 2020-03-07 main_figures/dimensionality.Rmd
updated figures to pdf

@@
   library(matrixStats)
-  library(pipeComp)
 })
@@
 theme_set(theme_cowplot(font_size = 12))
+```
+
+```{r, eval=FALSE}
+# NB: the results were generated using an older version of pipeComp (branch 
+# `biorxiv`), and must be plotted using this version of the package
+suppressPackageStartupMessages({
+  library(pipeComp)
+})
+```
+
+```{r, include=FALSE}
+#devtools::load_all("../../pipeComp_biorxiv/pipeComp/")
 ```
@@
 ```{r figure6, fig.width=9, fig.height=9.5}
+pdf("dimensionality.pdf", width=9, height=10)
+plot_grid( p1, p2, p3, nrow=3, rel_heights=c(0.36,0.32,0.32), labels=c(NA,LETTERS[3:4]))
+dev.off()
 plot_grid( p1, p2, p3, nrow=3, rel_heights=c(0.36,0.32,0.32), labels=c(NA,LETTERS[3:4]))

### 2020-03-07 main_figures/doublets.Rmd
updated figures to pdf

@@
 ---
-title: "Figure 1 doublets"
+title: "Doublets figure"
 author: "Pierre-Luc Germain"
 date: "`r format(Sys.time(), '%d %B, %Y')`"
-output:
-  html_document:
-    keep_md: true
+output: html_document
 
@@
 p1m <- plot_grid(p1, p2, p3, nrow=1)
+pdf("doublets.pdf", width=6, height=7)
+plot_grid(p1m, pt, p4, rel_heights=c(3.5,2,2), labels = LETTERS[1:3], ncol = 1, hjust=0, scale=0.95)
+dev.off()
 plot_grid(p1m, pt, p4, rel_heights=c(3.5,2,2), labels = LETTERS[1:3], ncol = 1, hjust=0)

### 2020-03-07 main_figures/dr.Rmd
updated figures to pdf

@@
   library(matrixStats)
+})
+source("../misc_functions.R")
+```
+
+```{r, eval=FALSE}
+# NB: the results were generated using an older version of pipeComp (branch 
+# `biorxiv`), and must be plotted using this version of the package
+suppressPackageStartupMessages({
   library(pipeComp)
 })
-source("../misc_functions.R")
+```
+
+```{r, include=FALSE}
+#devtools::load_all("../../pipeComp_biorxiv/pipeComp/")
 ```
@@
 
+pdf("dr.pdf", width = 12, height=6)
+plot_grid(p1,p2, nrow=2, labels=LETTERS[1:2], scale=0.95)
+dev.off()
 plot_grid(p1,p2, nrow=2, labels=LETTERS[1:2])

### 2020-03-07 main_figures/filtering.Rmd
updated figures to pdf

@@
 ---
-title: "Figure 2 filtering"
+title: "Filtering figure"
 author: "Pierre-Luc Germain"
 date: "`r format(Sys.time(), '%d %B, %Y')`"
-output:
-  html_document:
-    keep_md: true
-
+output: html_document
 ---
@@
 source("../misc_functions.R")
-data(datasets)
 ```
@@
                      c("nofilter","lenient","default","pca","pca2","stringent","veryStringent"))
-cols <- c("black",plgINS::getQualitativePalette(length(levels(m$filter))-1))
+cols <- c("black", pipeComp::getQualitativePalette(length(levels(m$filter))-1))
 names(cols) <- levels(m$filter)
@@
 
+pdf("filtering.pdf", width = 8, height=8.5)
+plot_grid(p1, p2, nrow=2, rel_heights = c(3,8), labels=LETTERS[1:2])
+dev.off()
+
 plot_grid(p1, p2, nrow=2, rel_heights = c(3,8), labels=LETTERS[1:2])

### 2020-03-07 main_figures/imputation.Rmd
updated figures to pdf

@@
 ```{r imputation, fig.width=10, fig.height=7}
+pdf("imputation.pdf", width = 10, height=7)
+plot_grid( p1, p2, nrow=2, labels=LETTERS[1:2], rel_heights = c(4.5,6))
+dev.off()
+
 plot_grid( p1, p2, nrow=2, labels=LETTERS[1:2], rel_heights = c(4,6))

### 2020-03-07 main_figures/selection.Rmd
updated figures to pdf

@@
 ---
-title: "Figure 4 selection"
+title: "Selection figure"
 author: "Pierre-Luc Germain"
 date: "`r format(Sys.time(), '%d %B, %Y')`"
-output:
-  html_document:
-    keep_md: true
+output: html_document
 ---
@@
 
-```{r figure4, fig.width=7.5, fig.height=6.5}
+```{r}
 res <- readRDS("../data/results_selection.rds")
@@
 p2 <- grid.grabExpr(draw(h1+h2))
+```
+
+```{r figure4, fig.width=7.5, fig.height=6.5}
+pdf("selection.pdf", width = 7.5, height=6.5)
+plot_grid(p1, p2, labels=LETTERS[1:2], nrow=2, rel_heights = c(4,5))
+dev.off()
+
 plot_grid(p1, p2, labels=LETTERS[1:2], nrow=2, rel_heights = c(4,5))

### 2020-03-07 supplementary_figures/supp1b_metrics.Rmd
updated figures to pdf

@@
 dd <- res[!duplicated(res[,c("dataset","true.nbClusts")]),]
+p <- ggplot(res, aes(n_clus, ARI, colour=clustmethod, shape=norm)) + 
+    geom_point() + xlim(c(1,40)) + facet_wrap(~dataset) + xlab("Number of clusters") +
+    geom_vline(aes(xintercept=k), data=dd, linetype="dashed") + theme(legend.position="bottom") + 
+    guides(colour=guide_legend(ncol=1, title = ""), shape=guide_legend(ncol=1, title=""))
 ```
@@
 ```{r, fig.width=7.5, fig.height=7}
-ggplot(res, aes(n_clus, ARI, colour=clustmethod, shape=norm)) + 
-    geom_point() + xlim(c(1,40)) + facet_wrap(~dataset) + xlab("Number of clusters") +
-    geom_vline(aes(xintercept=k), data=dd, linetype="dashed") + theme(legend.position="bottom") + 
-    guides(colour=guide_legend(ncol=1, title = ""), shape=guide_legend(ncol=1, title=""))
+p
 ```

### 2020-03-07 supplementary_figures/supp_figures.Rmd
updated supplementary figures
updated figures to pdf

--- 2019-05-09 supplementary_figures/supp_figures.Rmd
@@
 ---
-title: "Supplementary Figures"
-author: "Pierre-Luc Germain"
-date: "`r format(Sys.time(), '%d %B, %Y')`"
+title: \Large{pipeComp, a general framework for the evaluation of computational pipelines, reveals performant single-cell RNA-seq preprocessing tools}
+subtitle: \vspace{0.2cm}\LARGE{Supplementary Figures}\vspace{0.3cm}
+author:
+- "Pierre-Luc Germain"
+- "Anthony Sonrel"
+- "Mark D. Robinson"
 geometry: margin=1.5cm
@@
   pdf_document:
-    keep_tex: true
     toc: true
@@
 
+```{r child='supp1b_metrics.Rmd'}
+```
+
 ```{r child='supp1_dataset_description.Rmd'}
 ```
+
+```{r child='supp2_filtering.Rmd'}
+```
+
+```{r child='supp3_normalization.Rmd'}
+```
+
+```{r child='supp4_selection.Rmd'}
+```
+
+```{r child='supp5_dr.Rmd'}
+```
+
+```{r child='supp6_clust.Rmd'}
+```

### 2020-03-07 draft.tex

--- 2020-02-27 draft.tex
@@
 Here, we present a flexible R framework for pipeline comparison with multi-level evaluation metrics and apply it to the benchmark of scRNAseq analysis pipelines using datasets with known cell identities.
-We evaluate common steps of such analyses, including filtering, doublet detection (suggesting a new R package, \textit{scDblFinder}), normalization, feature selection, denoising, dimensionality reduction and clustering.
+We evaluate common steps of such analyses, including filtering, doublet detection, normalization, feature selection, denoising, dimensionality reduction and clustering.
 On the basis of these analyses, we make a number of concrete recommendations about analysis choices.
@@
 
-A number of good comparison and benchmark studies have already been performed on various steps related to scRNAseq analysis and can guide the choice of methodology \cite{CobosDeconvolution2020, colePerformance2018, DalMolinDE2017, duoClustering2018, freytagComparison2018, HeiserDimred2019, JaakkolaDE2017, SonesonDE2018, SunDimRed2019, tianMixology2018, TranBatch2020, TsuyuzakiPCA2020, viethSystematic2019, WangDE2019, YipHVGs2018, ZhangImput2018}.
+A number of good comparison and benchmark studies have already been performed on various steps related to scRNAseq processing and analysis and can guide the choice of methodology (\cite{CobosDeconvolution2020, colePerformance2018, DalMolinDE2017, duoClustering2018, freytagComparison2018, GaoProcessing2020, HeiserDimred2019, HouImput2020, JaakkolaDE2017, KrzakClust2019, SonesonDE2018, SunDimRed2019, tianMixology2018, TranBatch2020, TsuyuzakiPCA2020, viethSystematic2019, WangDE2019, YipHVGs2018, ZhangImput2018}).
 However these recommendations need constant updating and often leave open many details of an analysis.
@@
 Finally, we evaluate several recent methods and provide concrete recommendations.
-
-\begin{figure}
-    \centering
-    \includegraphics[width=\textwidth,keepaspectratio]{dataset_description}
-    \caption{\textbf{Overview of the benchmark datasets used.}}
-    \label{fig:figure1}
-\end{figure}
 
@@
 In addition, optional benchmark functions can be set for each step to provide standardized, multi-layered evaluation metrics.
-Given such a \texttt{PipelineDefinition} object, a set of alternative parameters (which might include different subroutines) and benchmark datasets, the \texttt{runPipeline} function then proceeds through all combinations of arguments, avoiding recomputing the same step twice and compiling evaluations (including running time) on the fly.
+Given such a \texttt{PipelineDefinition} object, a set of alternative parameters (which might include different subroutines) and benchmark datasets, the \texttt{runPipeline} function then proceeds through all combinations of arguments, avoiding recomputing the same step twice and compiling evaluations (including run time) on the fly.
 Variations in a given parameter can be evaluated using all metrics from this point downward in the pipeline.
 This is especially important because end-point metrics, such as the adjusted Rand index (ARI) \cite{HubertARI1985} for clustering, are not perfect.
-For example, although the meaning of an ARI score is independent of the number of true subpopulations \cite{steinleyProperties2004}, the number of clusters called is by far the most important determinant of the score: the farther it is from the actual number of subpopulations, the worse the ARI (Supplementary Figure 17).
+For example, although the meaning of an ARI score is independent of the number of true subpopulations \cite{steinleyProperties2004}, the number of clusters called is by far the most important determinant of the score: the farther it is from the actual number of subpopulations, the worse the ARI (Supplementary Figures 1-2).
 In this context, one strategy has been to cluster across various resolutions and only consider the results that have the right number of clusters \cite{duoClustering2018}.
@@
 Clustering results are also very sensitive and might not always capture improvements in earlier steps of the pipeline.
-We therefore favour monitoring several complementary metrics across multiple steps of the process.
-
-\begin{figure}
-    \centering
-    \includegraphics[width=\textwidth,keepaspectratio]{main_figures/pipeline_explanation.png}
-    \caption{\textbf{Overview of the \textit{pipeComp} framework and its application to a scRNAseq clustering pipeline.
-A:} The package is built around a \texttt{PipelineDefinition} S4 class which defines a set of functions to be executed consecutively, as well as optional evaluation functions for each step.
-Each step can accept of number of parameters whose alternative values are provided as a list.
-All (or subsets of) combinations of parameters can then be simultaneously ran and evaluated using the \texttt{runPipeline} function.
-\textbf{B:} Scheme representing the application of \textit{pipeComp} to evaluate a range of methods that are commonly used in scRNAseq studies and some of the metrics monitored at various steps.}
-    \label{fig:figure2}
-\end{figure}
-
-A clustering output is relatively fragile to variations in the pipeline.
-We therefore wanted to capture whether the effect of a given parameter alteration is robust to changes in the rest of the pipeline, or rather specific to a set of other pipeline parameters.
+In addition, we wanted to assess whether the effect of a given parameter alteration is robust to changes in the rest of the pipeline.
+We therefore monitored several complementary metrics across multiple steps of the process.
 We proceeded in a step-wise fashion, first testing a large variety of parameters at the early steps of the pipeline along with only a set of mainstream options downstream, then selecting main alternatives and proceeding to a more detailed benchmark of the next step (Figure \ref{fig:figure2}B).
@@
 In addition to these methods, we reasoned that an approach such as \textit{DoubletFinder} could be simplified by being applied directly on counts and by using a pre-clustering to create neotypic/heterotypic doublets more efficiently.
-We therefore developed a simple and fast Bioconductor package implementing this method for doublet detection, \textit{scDblFinder}, with the added advantage of accounting for uncertainty in the expected doublet rate and using meta-cells from the clusters to even include triplets (see Methods).
-
-\begin{figure}[h]
-    \centering
-    \includegraphics[width=0.9\textwidth]{{main_figures/doublets_files/figure-html/figure1-1.png}}
-    \caption{\textbf{Identification of doublet cells.
-A:} Receiver operating characteristic (ROC) curves of the tested doublet detection methods for three datasets with SNP-identified doublets.
-Dots indicate threshold determined by the true number of doublets.
-\textbf{B:} Running time of the different methods (DoubletFinder failed on one of the datasets).
-\textbf{C:} Rate of misclassification of the cells identified by \textit{scDblFinder} as doublets (DBL) or singlets (SNG), across a large range of clustering analysis.
-Even in datasets which should not have neotypic doublets, the cells identified as such tend to be misclassified.
-\label{fig:figure3}}
-\end{figure}
+We therefore also evaluated a simple and fast Bioconductor package implementing this method for doublet detection, \textit{scDblFinder}, with the added advantage of accounting for uncertainty in the expected doublet rate and using meta-cells from the clusters to even include triplets (see Methods).
 
@@
 This can for instance be performed through the \texttt{isOutlier} function from \textit{scater} that measures, for a given control property, the number of median absolute deviations (MADs) of each cell from the median of all cells.
-Supplementary Figure 1 shows the distributions of some of the typical cell properties commonly used.
+Supplementary Figure 4 shows the distributions of some of the typical cell properties commonly used.
 Of note, these properties tend to be correlated, with some exceptions.
-For example, while a high proportion of mitochondrial reads is often correlated with a high proportion of the counts in the top features, there can also be other reasons for an over-representation of highly-expressed features (Supplementary Figure 3), such as an over-amplification in non-UMI datasets.
-In our experience, 10X datasets also exhibit a very strong correlation between the total counts and the total features even across very different cell types (Supplementary Figure 4).
+For example, while a high proportion of mitochondrial reads is often correlated with a high proportion of the counts in the top features, there can also be other reasons for an over-representation of highly-expressed features (Supplementary Figure 5), such as an over-amplification in non-UMI datasets.
+In our experience, 10X datasets also exhibit a very strong correlation between the total counts and the total features even across very different cell types (Supplementary Figure 6).
 We therefore also measure the ratio between the two and treat cells strongly departing from this trend with suspicion.
 
-Reasoning that the cells we wish to exclude are the cells that would be misclassified, we measured the rate of misclassification of each cell in each dataset across a variety of clustering pipelines, correcting for the median misclassification rate of the subpopulation and then evaluated what properties could be predictive of misclassification (Supplementary Figures 5-7).
+Reasoning that the cells we wish to exclude are the cells that would be misclassified, we measured the rate of misclassification of each cell in each dataset across a variety of clustering pipelines, correcting for the median misclassification rate of the subpopulation and then evaluated what properties could be predictive of misclassification (Supplementary Figures 7-9).
 We could not identify any property or simple combination thereof that would be consistently predictive of misclassification; the only feature that consistently stood out across multiple datasets (the Zheng datasets) was that cells with very high read counts have a higher chance of being misclassified.
-
-\begin{figure}
-    \centering
-    \includegraphics[width=\textwidth]{{main_figures/filtering_files/figure-html/figure2-1.png}}
-    \caption{\textbf{A:} Filtering on the basis of distance to the whole distribution can lead to strong bias against certain subpopulations.
-The dashed line indicates a threshold of 2.5 median absolute deviations (MADs) from the median of the overall population.
-\textbf{B:} Relationship between the maximum subpopulation exclusion rate and the average clustering accuracy per subpopulation across various filtering strategies.
-Of note, doublet removal appears to be desirable even when, due to the design, there are no heterotypic doublets in the data.
-The PCA methods refer to multivariate outlier detected as implemented in \textit{scater} (see Methods for details).}
-    \label{fig:figure4}
-\end{figure}
 
@@
 We next investigated the impact of different normalization strategies.
-Beside the standard log-normalization included in \textit{Seurat}, we tested \textit{scran}'s pooling-based normalization \cite{lunPooling2016}, \textit{sctransform}'s variance-stabilizing transformation \cite{hafemeisterSCtransform2019}, and normalization based on stable genes \cite{linStableGenes2018, deekeStablyExpressed2018}.
+Beside the standard log-normalization included in \textit{Seurat}, we tested \textit{scran}'s pooling-based normalization \cite{lunPooling2016}, \textit{sctransform}'s variance-stabilizing transformation \cite{hafemeisterSCtransform2019}, normalization based on stable genes \cite{linStableGenes2018, deekeStablyExpressed2018}, and SCnorm \cite{bacherSCnorm2017} (\textit{scVI} \cite{lopezDeep2018} was included separately on a subset of parameter combinations due to high running times).
 In addition to log-normalization, the standard \textit{Seurat} clustering pipeline performs per-feature unit-variance scaling so that the PCA is not too strongly dominated by highly-expressed features.
 We therefore included versions of the different normalization procedures, with or without a subsequent scaling (\textit{sctransform}'s variance-stabilizing transformation involves an approach analogous to scaling).
-\textit{Seurat}'s scaling function also includes the option to regress out the effect of certain covariates.
+\textit{Seurat}'s scaling function (and \textit{sctransform}) also includes the option to regress out the effect of certain covariates.
 We tested its performance by using the proportion of mitochondrial counts and the number of detected features as covariates.
-Finally, it has been proposed that the use of stable genes, in particular cytosolic ribosomal genes, can be used to normalize scRNAseq\cite{deekeStablyExpressed2018}.
+Finally, it has been proposed that the use of stable genes, in particular cytosolic ribosomal genes, can be used to normalize scRNAseq \cite{deekeStablyExpressed2018}.
 We therefore evaluated a simple linear normalization based on the sum of these genes, as well as nuclear genes.
@@
 However, as library size tends to vary across subpopulations, part of this effect can simply reflect biological differences.
-We therefore assessed to what extent the first principal component still retained a correlation with the library size and the number of detected features, removing the confounding covariation with the subpopulations (Figure \ref{fig:figure5}A).
-The simple step of scaling tended to remove much of the correlation with these features and, in the absence of scaling, standard \textit{Seurat} normalization resulted in fairly high correlation with technical covariates.
-\textit{sctransform} led to the lowest correlation but most methods, including normalization based on stable genes, were able to remove most of the correlation when combined with scaling.
-The only exception is \textit{Seurat} normalization with scaling regressing out covariates, which surprisingly increased the correlation in 8 of the 9 datasets.
-
-\begin{figure}
-    \centering
-    \includegraphics[width=\textwidth]{{main_figures/normalization_files/figure-html/figure3-1.png}}
-    \caption{\textbf{Evaluation of normalization procedures.} `Seurat.feat\_mt\_regress' regresses out the number of features and proportion of mitochondrial reads during scaling; `Seurat.feat\_regress` regresses out the number of features only.
-Residual correlation of the first principal component with library size (left) and detection rate (right), after accounting for biological differences between subpopulations.
-\textbf{B:} Average silhouette width per true subpopulation, where higher silhouette width means a higher separability.
-\textbf{C:} Clustering accuracy (\textit{Seurat} clustering), measured by the ARI at the true number of cluster (left) and by the average mutual information (MI) of the cluster assignment with true subpopulations.}
-    \label{fig:figure5}
-\end{figure}
+We therefore assessed to what extent the first principal component still retained a correlation with the library size and the number of detected features, removing the confounding covariation with the subpopulations (Figure \ref{fig:figure5}A and Supplementary Figure 12).
+Scaling tended to remove much of the correlation with these features, and while most methods were able to remove most of the effect, the correlation was lowest with \textit{sctransform}.
+Surprisingly, however, regressing out the covariates tended to increase the association with it.
 
 We further evaluated normalization methods by investigating their impact on the separability of the subpopulations (Figure \ref{fig:figure5}B-C).
-Since clustering accuracy metrics such as the ARI are very strongly influenced by the number of clusters, we complemented it with silhouette width \cite{RousseeuwSil1987} and mutual information.
+Since clustering accuracy metrics such as the ARI are very strongly influenced by the number of clusters, we complemented it with silhouette width \cite{RousseeuwSil1987} and mutual information (MI).
 We found most methods (including no normalization at all) to perform fairly well in most of the subpopulations.
+An exception was \textit{scVI} (Supplementary Figure 13), which performed poorly on most datasets in terms of both the subpopulation silhouette and clustering accuracy.
+For time and resource cost reasons, \textit{scVI} was run on a reduced set of pipelines and compared to \textit{sctransform} (more details in the Methods section).
 Scaling tended to reduce the average silhouette width of some subpopulations and to increase that of some less distinguishable ones and was generally, but not always, beneficial on the accuracy of the final clustering.
@@
 Finally, we monitored whether, under the same downstream clustering analysis, different normalization methods tended to lead to an over- or under-estimation of the number of clusters.
-Although some methods had a tendency to lead to a higher (e.g., \textit{sctransform}) or lower (e.g., stable genes) number of clusters, the effect was very mild and not entirely systematic (Supplementary Figure 10).
+Although some methods had a tendency to lead to a higher (e.g., \textit{sctransform}) or lower (e.g., stable genes) number of clusters, the effect was very mild and not entirely systematic (Supplementary Figure 14).
+We also confirmed that, especially in the UMI-based datasets, \textit{sctransform} did in fact successfully stabilize variance across mean counts (Supplementary Figure 15), at little apparent cost on computation time (Supplementary Figure 16).
 
@@
 Another common practice in feature selection is to use those with the highest mean expression.
-Recently, \cite{townesGlmpca2019} instead suggested to use deviance, while \textit{sctransform} provides its own ordering of genes based on transformed variance.
+Recently, it was instead suggested to use deviance \cite{townesGlmpca2019}, while \textit{sctransform} provides its own ordering of genes based on transformed variance.
 
-Reasoning that a selection method should ideally select genes whose variability is higher between subpopulations than within, we first assessed to what extent each method selected genes with a high proportion of variance or deviance explained by (real) subpopulation.
+Reasoning that a selection method should ideally select genes whose variability is higher between subpopulations than within, we first assessed to what extent each method selected genes with a high proportion of variance or deviance explained by the (real) subpopulations.
 As the proportion of variability in a gene attributable to subpopulations can be measured in various ways, we first compared three approaches: ANOVA on log-normalized count, ANOVA on \textit{sctransform} normalization, and deviance explained.
-The ANOVAs performed on a standard \textit{Seurat} normalization and on \textit{sctransform} data were highly correlated (Supplementary Figure 11A).
-These estimates were also in good agreement with the deviance explained, although lowly-expressed genes could have a high deviance explained without having much of their variance explained by subpopulation (Supplementary Figure 11B-D).
-We therefore compared the proportion of the cumulative variance/deviance explained by the top X genes that could be retrieved by each gene ranking method (Supplementary Figures 12-13).
-We first focused on the first 1000 genes to highlight the differences between methods, although a higher number of selected genes decreased the differences between methods (Supplementary Figures 12-14).
+The ANOVAs performed on a standard \textit{Seurat} normalization and on \textit{sctransform} data were highly correlated (Supplementary Figure 17A).
+These estimates were also in good agreement with the deviance explained, although lowly-expressed genes could have a high deviance explained without having much of their variance explained by subpopulation (Supplementary Figure 17B-D).
+We therefore compared the proportion of the cumulative variance/deviance explained by the top X genes that could be retrieved by each gene ranking method (Supplementary Figures 18-19).
+We first focused on the first 1000 genes to highlight the differences between methods, although a higher number of selected genes decreased the differences between methods (Supplementary Figures 18-20).
 The standardized measures of variability were systematically worse than their non-standardized counterparts in selecting genes with a high proportion of variance explained by subpopulation.
-Regarding the percentage of deviance explained however, the standardized measures were often superior (Figure \ref{fig:figure6}A and Supplementary Figures 12-13).
-Deviance proved the method of choice to prioritize genes with a high variance explained by subpopulations (with mere expression level proving surprisingly good) but did not perform so well to select genes with a high deviance explained.
-
-\begin{figure}
-    \centering
-    \includegraphics[scale=0.75]{{main_figures/selection_files/figure-html/figure4-1.png}}
-    \caption{\textbf{Evaluation of feature selection methods.
-A:} Ability of different feature ranking methods to capture genes with a high proportion of variance (left) or deviance (right) explained by real subpopulations.
-The asterix denotes the default Seurat method.
-\textbf{B:} Accuracy of clusterings (at the true number of clusters) when selecting 1000 genes using the given methods.
-Based on standard \textit{Seurat} normalization (left) or \textit{sctransform} (right).
-The  \textit{vst.varExp} and \textit{devianceExplained} methods correspond to the estimates used in \textbf{A} to evaluate the selection methods and were included here only for validation purpose.}
-    \label{fig:figure6}
-\end{figure}
+Regarding the percentage of deviance explained however, the standardized measures were often superior (Figure \ref{fig:figure6}A and Supplementary Figures 18-19).
+Deviance proved the method of choice to prioritize genes with a high variance explained by subpopulations (with mere expression level proving surprisingly performant) but did not perform so well to select genes with a high deviance explained.
 
@@
 In general, we found deviance and unstandardized estimates of variance to provide the best results across datasets and normalization methods.
-Increasing the number of features selected also systematically led to an increase in the accuracy of the clustering, typically plateauing after 4000 features (Supplementary Figure 14).
+Increasing the number of features selected also systematically led to an increase in the accuracy of the clustering, typically plateauing after 4000 features (Supplementary Figure 20).
 
@@
 
-\begin{figure}[h]
-    \centering
-    \includegraphics[width=\textwidth]{{main_figures/dr_files/figure-html/dimred-1.png}}
-    \caption{\textbf{Evaluation of common dimensionality reduction methods.
-A:} Average silhouette width per subpopulation resulting from combinations of normalization and dimension reductions.
-\textbf{B:} Clustering accuracy, mean proportion of the variance in the first 5 components explained by real subpopulations (center), and median ARI of the resulting clustering (right).}
-    \label{fig:figure7}
-\end{figure}
-
-The impact of the choice of dimensionality reduction method was far greater than that of normalization or feature selection (e.g., Supplementary Figure 15).
-GLM-PCA tended to increase the average silhouette width of already well-defined subpopulations, but \textit{Seurat}'s PCA procedure however proved superior on all metrics (Figure \ref{fig:figure7}).
+The impact of the choice of dimensionality reduction method was far greater than that of normalization or feature selection (Figure \ref{fig:figure7}).
+GLM-PCA tended to increase the average silhouette width of already well-defined subpopulations, but \textit{Seurat}'s PCA procedure however proved superior on all metrics.
+Like GLM-PCA, \textit{scVI}'s Linear Decoder (LD) does not explicitly rely on normalized counts.
+Its performance was lower than a pipeline with non-weighted \textit{Seurat} PCA on all measures for datasets with more subpopulations (Supplementary Figure 13).
 Overall, weighting the principal components by their variance (as \textit{Seurat} does) had a positive impact on silhouette widths and ARI scores.
@@
 Since Euclidean distance decreases as the number of non-discriminating dimensions increases, there is usually a trade-off between selecting enough dimensions to keep most information and excluding smaller dimensions that may represent technical noise or other unwanted sources of variation.
-Overall, increasing the number of dimensions robustly led to a decrease in the number of clusters (Supplementary Figures 10 and 15).
-This tended to affect the accuracy of the clustering (Supplementary Figure 16), although in both cases (number of clusters and ARI) \textit{Seurat}'s resolution parameter had a much stronger impact.
+Overall, increasing the number of dimensions robustly led to a decrease in the number of clusters (Supplementary Figures 14 and 21).
+This tended to affect the accuracy of the clustering (Supplementary Figure 22), although in both cases (number of clusters and ARI), \textit{Seurat}'s resolution parameter had a much stronger impact.
 
@@
 We evaluated the performance of dimensionality estimators implemented in the \textit{intrinsicDimension} package \cite{johnssonDimensionality2015}, as well as common procedures such as the `elbow` method (inflexion point in the variance explained by each component), some scRNAseq-specific methods such as the JackStraw procedure \cite{ChungJackstraw2015} or \textit{scran}'s \texttt{denoisePCA} \cite{LunScran2016}, and the recent application of Fisher Separability analysis \cite{AlberganteSepar2019}.
-
-\begin{figure}
-    \centering
-    \includegraphics[width=\textwidth]{{main_figures/dimensionality_files/figure-html/figure6-1.png}}
-    \caption{\textbf{Estimating dimensionality.
-A:} Estimated dimensionality using the proportion of variance in each component explained by the true subpopulations.
-\textbf{B:} Difference between `real' dimensionality (from \textbf{A}) and dimensionality estimation methods, along with computing time.
-\textbf{C:} Average silhouette width per subpopulation across a selection of methods, combined with \textit{sctransform} or standard \textit{Seurat} normalization.
-\textbf{D:} Clustering accuracy across the normalization/dimensionality estimation methods.}
-    \label{fig:figure8}
-\end{figure}
 
@@
 Of note, the methods differ widely in view of their computing time (Figure \ref{fig:figure8}B) and we saw no relationship between the accuracy of the estimates and the complexity of the method.
-Reasoning that over-estimating the number of dimensions is less problematic than under-estimating it, we kept the former methods for a full analysis of their impact on clustering (Figure \ref{fig:figure8}C-D), when combined with sctransform or the standard Seurat normalization.
-Although most methods performed well on the various clustering measures, the global maximum likelihood based on translated Poisson Mixture Model (\textit{maxLikGlobal}) provided the dimensionality estimate that best separated the subpopulations (Figure \ref{fig:figure8}C) and resulted in the best clustering accuracy (Figure \ref{fig:figure8}D).
+Reasoning that over-estimating the number of dimensions is less problematic than under-estimating it, we kept the former methods for a full analysis of their impact on clustering (Figure \ref{fig:figure8}C-D), when combined with \textit{sctransform} or the standard \textit{Seurat} normalization.
+Although most methods performed well on the various clustering measures, the global maximum likelihood based on translated Poisson Mixture Model (\textit{maxLikGlobal}, using the 20 nearest neighbors, although estimates were relatively stable across decent values of \textit{k}; see Supplementary Figure 23) provided the dimensionality estimate that best separated the subpopulations (Figure \ref{fig:figure8}C) and resulted in the best clustering accuracy (Figure \ref{fig:figure8}D).
 This method systematically selected many more components than were associated with the subpopulations, suggesting that although these additional components appear individually uninformative, in combination they nevertheless contribute to classification.
+An implication is that the commonly-used elbow method most likely underestimates dimensionality.
 
@@
 The last step evaluated in our pipeline was clustering.
-Given previous works on the topic \cite{duoClustering2018,freytagComparison2018} and the success of graph-based clustering methods for scRNAseq, we restricted our evaluation to Seurat's method and two \textit{scran} SNN-based clustering approaches, based on random walks (\textit{walktrap} method) or on the optimization of the modularity score (\textit{fast\_greedy}).
+Given previous work on the topic \cite{duoClustering2018,freytagComparison2018} and the success of graph-based clustering methods for scRNAseq, we restricted our evaluation to Seurat's method and variations of \textit{scran}'s graph-based clustering using random walks (\textit{walktrap} method) or the optimization of the modularity score (\textit{fast\_greedy}) on various underlying graphs.
+Specifically, we compared rank-based graphs (\textit{scran}'s default) with graphs using the components' values directly, SNN vs KNN graphs, and the exact nearest neighbors to the Annoy approximation.
 Again, the tested methods were combined with \textit{Seurat}'s standard normalization and \textit{sctransform}, otherwise using the parameters found optimal in the previous steps.
 
-Since ARI is dominated by differences in the number of clusters (Supplementary Figure 17) and no single metric is perfect, we diversified them (Figure \ref{fig:figure9}).
-Mutual information (MI) has the virtue of not decreasing when a true subpopulation is split into two clusters, which is arguably less problematic (and might well reflect unknown biological subgroups), but as a consequence it can be biased towards methods producing higher resolution clustering.
+Since ARI is dominated by differences in the number of clusters (Supplementary Figures 1-2) and no single metric is perfect, we diversified them (Figure \ref{fig:figure9}).
+MI has the virtue of not decreasing when a true subpopulation is split into two clusters, which is arguably less problematic (and might well reflect unknown biological subgroups), but as a consequence it can be biased towards methods producing higher resolution clustering (Supplementary Figure 2).
 Similarly, precision per true subpopulation is considerably more robust to differences in the number of clusters.
@@
 
-\begin{figure}[h]
-    \centering
-    \includegraphics[width=\textwidth]{{main_figures/clustering_files/figure-html/clustering-1.png}}
-    \caption{\textbf{Evaluation of clustering methods.} Clustering accuracy of common clustering tools in combination with \textit{sctransform} and standard Seurat's normalization.}
-    \label{fig:figure9}
-\end{figure}
-
 The MI score and minimum precision, which are largely independent of the estimated number of clusters, were overall higher for the \textit{walktrap} method (Figure \ref{fig:figure9}), while the mean F1 score favored both \textit{scran} methods (\textit{walktrap} and \textit{fast greedy}) over \textit{Seurat}.
 Finally, the ARI score at the true number of clusters, when available, showed similar performances.
-However, because \textit{Seurat}'s resolution parameter had a large impact on the number of clusters identified (Supplementary Figure 18), \textit{Seurat} could always be coerced into producing the right number of clusters.
-Instead, the number of clusters found by \textit{scran} was considerably less influenced by the available parameters (number of nearest neighbors or steps in the random walk - see Supplementary Figure 18), and as a result \textit{scran}-based clustering sometimes never produced a partitioning with the right number of clusters.
+However, because \textit{Seurat}'s resolution parameter had a large impact on the number of clusters identified (Supplementary Figures 1 and 23), \textit{Seurat} could always be coerced into producing the right number of clusters.
+Instead, the number of clusters found by \textit{scran} was considerably less influenced by the available parameters (number of nearest neighbors or steps in the random walk - see Supplementary Figure 24), and as a result \textit{scran}-based clustering sometimes never produced a partitioning with the right number of clusters.
 This observation, along with \textit{scran}'s higher MI score, suggest that \textit{scran} sometimes simply divides a real subpopulation into two clusters (possibly tracking some unknown biological differences) rather than committing misclassification errors.
 Overall, the \textit{walktrap} method appeared superior to the \textit{fast greedy} algorithm and was generally less prone to misclassification than \textit{Seurat} clustering, although the latter offered more control over the resolution.
-Finally, some poorly distinguishable subpopulations from both the Zhengmix8eq and simMix1 datasets remained very inaccurately classified by all methods and in regards to all metrics.
+A major difference between \textit{walktrap}-based clustering and \textit{Seurat} is the computing time (Figure \ref{fig:figure9}B).
+We found that using the Annoy approximation to the nearest neighbors somewhat reduced computing time at no apparent cost to accuracy.
+It nevertheless remained considerably slower than \textit{Seurat}.
+For all methods, some poorly distinguishable subpopulations from both the Zhengmix8eq and simMix1 datasets remained very inaccurately classified in regard to all metrics.
 
 \subsection*{Further extensions to the pipeline: imputation/denoising}
-
-\begin{figure}[h]
-    \centering
-    \includegraphics[width=\textwidth]{{main_figures/imputation_files/figure-html/imputation-1.png}}
-    \caption{\textbf{Imputation/denoising methods} Average silhouette width per subpopulation (\textbf{A}) and clustering accuracy (\textbf{B}) with or without (indicated as \textit{none}) application of a denoising/imputation method.}
-    \label{fig:figure10}
-\end{figure}
 
@@
 \end{enumerate}
-
-\begin{figure}[h]
-    \centering
-    \includegraphics[scale=0.08,keepaspectratio]{main_figures/pipeComp_summary.png}
-    \caption{\textbf{Recommendations of tools from filtering to clustering.}}
-    \label{fig:figure11}
-\end{figure}
 
@@
 
+Our recommendations are in line with the results of several other benchmarking studies.
+For instance, \textit{Seurat} and \textit{scran} methods were already shown to yield good clustering performance \cite{freytagComparison2018} and \textit{DrImpute} to improve the quality of downstream analyses \cite{ZhangImput2018}.
+A previous systematic comparison \cite{viethSystematic2019} instead found that the use of \textit{SAVERX} for denoising data and \textit{scran} for clustering yielded better performance on the ability to recover differentially-expressed genes, whereas we focused on the ability to recover cell populations.
+Another disagreement with the present recommendations come from the study of Hou et al.
+\cite{HouImput2020} that recommends \textit{SAVERX} and discourage the use of \textit{DCA} for better clustering, whereas our study found opposite results for these tools.
+Several differences in the design of our study could explain this contrast, such as the use of different metrics (ARI at true number of clusters, mean precision/ recall), of different datasets (Koh, Kumar, simMix simulations) and the use of different filtering and normalization strategies.
+\textit{ALRA} however yielded good results in both studies and may be the most conservative option.
+The study from Hou et al.
+also highlighted the usefulness of \textit{ALRA} when followed by differential and trajectory analysis both in UMI and Fluidigm technologies.
+
 Here, we chose to concentrate on what could be learned from datasets with known cell labels (as opposed to labels inferred from the data, as in \cite{MereuCellAtlas2019}).
@@
 Finally, imputation had a varied impact on the clustering analysis and seemed to be linked to the technology that was used to generate the data.
-Also, the good performance of \textit{DrImpute} is in line with a previous study focused on the preservation of data structure in trajectory inference from scRNAseq \cite{ZhangImput2018} (\textit{alra} was however not included in this previous benchmark).
+Also, the good performance of \textit{DrImpute} is in line with a previous study focused on the preservation of data structure in trajectory inference from scRNAseq \cite{ZhangImput2018} (\textit{ALRA} was however not included in this previous benchmark).
 Our results also align with the hypothesis of this study on the respective strengths of linear and non-linear models and their use to different types of data, such as the highest performance of non-linear methods in developmental studies.
 
-\section*{Conclusion}
+\section*{Conclusions}
 
 \textit{pipeComp} is a flexible R framework for evaluating methods and parameter combinations in a complex pipeline, computing on-the-fly multilevel evaluation metrics.
-Applying this framework to scRNAseq clustering enabled us to make concrete recommendations on the steps of filtering, normalization, feature selection, denoising, dimensionality reduction and clustering.
+Applying this framework to scRNAseq clustering enabled us to make concrete recommendations on the steps of filtering, normalization, feature selection, denoising, dimensionality reduction and clustering (Figure \ref{fig:figure11}).
 We demonstrate how a diversity of multilevel metrics can be more robust, more sensitive, and more nuanced than simply evaluating the final clustering performance.
-In addition, we provide a new and efficient Bioconductor package for doublet detection, \texttt{scDblFinder}.
-We hope that the \textit{pipeComp} framework can be applied to extend the current field of benchmarking, as well as to apply it to other range of methods.
-
-\newpage
+We hope that the \textit{pipeComp} framework can be applied to extend the current field of benchmarking, as well as to investigate pipelines other domains with multi-step computational analyses.
 
@@
 All analyses were performed through the \href{https://github.com/plger/pipeComp}{pipeComp} R package, which implements the pipeline framework described here.
+scVI results and results of the clustering section were produced with \textit{pipeComp} 0.99.4, while all other results were produced with version 0.99.2 (\textit{biorxiv}) branch of the repository).
 All code to reproduce the simulations and figures is available in the \url{https://github.com/markrobinsonuzh/scRNA\_pipelines\_paper} repository, which also includes the basic datasets with a standardized annotation.
@@
 For the \texttt{JackStraw} method, since the \textit{Seurat} documentation advises not to use a p-value threshold (which would typically yield a very large number of dimensions) but rather look for a drop in significance, we applied the same farthest point algorithm on the log10(p-values), which in our hands reproduced manual threshold selection.
-
-\subsection*{Doublet detection method}
-Our doublet detection method is available at \url{https://github.com/plger/scDblFinder}.
-Briefly, after reducing the data to the most expressed genes, we cluster the cells using the fast-greedy algorithm, favouring overclustering.
-We then create artificial doublets by sampling the two cells specifically from different clusters and sum the counts of each pair of cells.
-We also create meta-cells from each cluster and use them to create additional doublets and triplets.
-We combine them with the real cells, perform PCA and build a KNN graph using \textit{BiocNeighbors}.
-We then calculate, for each cell, the proportion of its neighbors that are artificial doublets, weighted by the distance.
-This ratio serves as a doublet score, which is then thresholded by simultaneously minimizing the error in classifying real vs artificial doublets and the deviation from the distribution of expected doublet rate (accounting for homotypic doublets as done by \textit{DoubletFinder}).
 
@@
 To estimate the correlation between the principal components and covariates such as library size, we first fitted a linear model on the subpopulations and correlated the residuals of this model with the covariate of interest.
+
+\subsection*{\textit{scVI} normalization and LD}
+
+The evaluation of \textit{scVI} algorithm was performed separate to the other pipeline combinations as the model training on our reference datasets was especially resource-consuming.
+As it was not practically feasible to perform the training step on many combination of parameters as it was done with other tools, we evaluated it on a representative subset of pipelines using \textit{sctransform} normalization and \textit{Seurat}'s PCA dimension reduction method.
+The resulting evaluation is available in the Supplementary Figure 13.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                          %%
+%% Backmatter begins here                   %%
+%%                                          %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{backmatter}
+
+\section*{Competing interests}
+As developers of the \textit{scDblFinder} package, the authors have an interest in it performing well. The authors declare no further competing interests.
+
+\section*{Author's contributions}
+PLG developed the \textit{pipeComp} framework, and PLG and MDR designed the benchmark.
+PLG and AS prepared the pipelines, methods wrappers and performed the analyses.
+AS performed all denoising/imputation analyses.
+PLG, AS and MDR wrote the manuscript.
+
+\section*{Acknowledgements}
+This work was supported by the Swiss National Science Foundation (grants 310030\_175841, CRSII5\_177208) as well as the Chan Zuckerberg Initiative DAF (grant number 2018-182828), an advised fund of Silicon Valley Community Foundation.
+MDR acknowledges support from the University Research Priority Program Evolution in Action at the University of Zurich.
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% Backmatter begins here                   %%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\newpage
 
-\begin{backmatter}
+\showfigstrue
 
-\section*{Competing interests}
-As developers of the \textit{scDblFinder} package, the authors have an interest in it performing well. The authors declare no further competing interests.
+\begin{figure}[h!]
+    \centering
+    \ifshowfigs
+    \includegraphics[width=0.95\textwidth,keepaspectratio]{dataset_description}
+    \fi
+    \caption{\textbf{Overview of the benchmark datasets used.}}
+    \label{fig:figure1}
+\end{figure}
 
-% \section*{Author's contributions}
-%    Text for this section \ldots
+\begin{figure}[h!]
+    \centering
+    \ifshowfigs
+    \includegraphics[scale=0.75, keepaspectratio]{main_figures/pipeline_explanation.pdf}
+    \fi
+    \caption{\textbf{Overview of the \textit{pipeComp} framework and its application to a scRNAseq clustering pipeline.
+A:} The package is built around a \texttt{PipelineDefinition} S4 class which defines a set of functions to be executed consecutively, as well as optional evaluation functions for each step.
+Each step can accept a number of parameters whose alternative values are provided as a list.
+All (or subsets of) combinations of parameters can then be simultaneously ran and evaluated using the \texttt{runPipeline} function.
+\textbf{B:} Scheme representing the application of \textit{pipeComp} to evaluate a range of methods that are commonly used in scRNAseq studies and some of the metrics monitored at various steps.}
+    \label{fig:figure2}
+\end{figure}
 
-\section*{Acknowledgements}
-This work was supported by the Swiss National Science Foundation (grants 310030\_175841, CRSII5\_177208) as well as the Chan Zuckerberg Initiative DAF (grant number 2018-182828), an advised fund of Silicon Valley Community Foundation.
-MDR acknowledges support from the University Research Priority Program Evolution in Action at the University of Zurich.
+\begin{figure}[h!]
+    \centering
+    \ifshowfigs
+    \includegraphics[scale=0.85, keepaspectratio]{{main_figures/doublets.pdf}}
+    \fi
+    \caption{\textbf{Identification of doublet cells.
+A:} Receiver operating characteristic (ROC) curves of the tested doublet detection methods for three datasets with SNP-identified doublets.
+Dots indicate threshold determined by the true number of doublets.
+\textbf{B:} Running time of the different methods (\textit{DoubletFinder} failed on one of the datasets).
+\textbf{C:} Rate of misclassification of the cells identified by \textit{scDblFinder} as doublets (DBL) or singlets (SNG), across a large range of clustering analysis.
+Even in datasets which should not have neotypic doublets, the cells identified as such tend to be misclassified.
+\label{fig:figure3}}
+\end{figure}
+
+\begin{figure}[h!]
+    \centering
+    \ifshowfigs
+    \includegraphics[scale=0.72, keepaspectratio]{{main_figures/filtering.pdf}}
+    \fi
+    \caption{\textbf{Effect of filtering on cell subpopulation structure and clustering.
+A:} Filtering on the basis of distance to the whole distribution can lead to strong bias against certain subpopulations.
+The dashed line indicates a threshold of 2.5 median absolute deviations (MADs) from the median of the overall population.
+\textbf{B:} Relationship between the maximum subpopulation exclusion rate and the average clustering accuracy per subpopulation across various filtering strategies.
+Of note, doublet removal appears to be desirable even when, due to the design, there are no heterotypic doublets in the data.
+The PCA methods refer to multivariate outlier detected as implemented in \textit{scater} (see Methods for details).}
+    \label{fig:figure4}
+\end{figure}
+
+\begin{figure}[h!]
+    \centering
+    \ifshowfigs
+    \includegraphics[width=\textwidth, keepaspectratio]{{main_figures/normalization.pdf}}
+    \fi
+    \caption{\textbf{Evaluation of normalization procedures.
+A:} Variance in PC1 explained by library size (left) and detection rate (right) after accounting for subpopulation differences.
+\textbf{B:} Average silhouette width per true subpopulation, where higher silhouette width means a higher separability.
+\textbf{C:} Clustering accuracy (\textit{Seurat} clustering), measured by mutual information (MI), Adjusted Rand Index (ARI) and ARI at the true number of cluster.
+`feat\_regress', `mt\_regress' and `feat\_mt\_regress' respectively stand for the regressing out of the number of features, the proportion of mitochondrial reads, or both during scaling.}
+    \label{fig:figure5}
+\end{figure}
+
+\begin{figure}[h!]
+    \centering
+    \ifshowfigs
+    \includegraphics[scale=0.65]{{main_figures/selection.pdf}}
+    \fi
+    \caption{\textbf{Evaluation of feature selection methods.
+A:} Ability of different feature ranking methods to capture genes with a high proportion of variance (left) or deviance (right) explained by real subpopulations.
+The asterisk denotes the default Seurat method.
+\textbf{B:} Accuracy of clusterings (at the true number of clusters) when selecting 1000 genes using the given methods.
+Based on standard \textit{Seurat} normalization (left) or \textit{sctransform} (right).
+The  \textit{vst.varExp} and \textit{devianceExplained} methods correspond to the estimates used in \textbf{A} to evaluate the selection methods and were included here only for validation purpose.}
+    \label{fig:figure6}
+\end{figure}
+
+\begin{figure}[h!]
+    \centering
+    \ifshowfigs
+        \includegraphics[width=\textwidth, keepaspectratio]{{main_figures/dr.pdf}}
+    \fi
+    \caption{\textbf{Evaluation of common dimensionality reduction methods.
+A:} Average silhouette width per subpopulation resulting from combinations of normalization and dimension reductions.
+\textbf{B:} Clustering accuracy, mean proportion of the variance in the first 5 components explained by real subpopulations (center), and median ARI of the resulting clustering (right).}
+    \label{fig:figure7}
+\end{figure}
+
+\begin{figure}[h!]
+    \centering
+    \ifshowfigs
+    \includegraphics[scale=0.55, keepaspectratio]{{main_figures/dimensionality.pdf}}
+    \fi
+    \caption{\textbf{Estimating dimensionality.
+A:} Estimated dimensionality using the proportion of variance in each component explained by the true subpopulations.
+\textbf{B:} Difference between `real' dimensionality (from \textbf{A}) and dimensionality estimation methods, along with computing time.
+\textbf{C:} Average silhouette width per subpopulation across a selection of methods, combined with \textit{sctransform} or standard \textit{Seurat} normalization.
+\textbf{D:} Clustering accuracy across the normalization/dimensionality estimation methods.}
+    \label{fig:figure8}
+\end{figure}
+
+\begin{figure}[h!]
+    \centering
+    \ifshowfigs
+    \includegraphics[width=\textwidth, keepaspectratio]{{main_figures/clustering.pdf}}
+    \fi
+    \caption{\textbf{Evaluation of clustering methods.} Clustering accuracy (\textbf{A}) and run time (\textbf{B}) of \textit{scran}- and \textit{Seurat}-based clustering tools in combination with \textit{sctransform} and standard \textit{Seurat}'s normalization.}
+    \label{fig:figure9}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \ifshowfigs
+    \includegraphics[scale=0.55, keepaspectratio]{{main_figures/imputation.pdf}}
+    \fi
+    \caption{\textbf{Evaluation of imputation/denoising methods} Average silhouette width per subpopulation (\textbf{A}) and clustering accuracy (\textbf{B}) with or without (indicated as \textit{none}) application of a denoising/imputation method.}
+    \label{fig:figure10}
+\end{figure}
+
+\begin{figure}[h!]
+    \centering
+    \ifshowfigs
+    \includegraphics[scale=0.9,keepaspectratio]{main_figures/pipeComp_summary.pdf}
+    \fi
+    \caption{\textbf{Recommendations of tools from filtering to clustering.}}
+    \label{fig:figure11}
+\end{figure}
 

### 2020-04-20 svadea/datasets/data_prep.Rmd
added SVA-DEA datasets and analysis

---
title: "Dataset preparation"
author: "Pierre-Luc Germain"
date: "4/14/2020"
output: html_document
---

## GSE79636 iPSC lines

```{r}
suppressPackageStartupMessages({
  library(SummarizedExperiment)
  library(iPSCpoweR)
})
data("GSE79636")
se <- SummarizedExperiment( list(counts=GSE79636$dat), colData=GSE79636$annotation)
set.seed(123)
se <- se[,sample(colnames(se), 20)]
se <- se[rowSums(assay(se)>0)>1,]
colData(se) <- droplevels(colData(se))
se$condition <- rep(0:1,each=10)
expressed <- which(rowSums(assay(se)>10)>10)
# add known differential expression
n <- 300
w <- sample(expressed, n)
lfcs <- rep(c(0.25, 0.5, 0.75, 1, 1.25), c(1,2,2,2,1))
lfcs <- sample(c(lfcs, -lfcs), n, replace=TRUE)
assay(se)[w,c(5,12:20)] <- round(assay(se)[w,11:20]*2^lfcs)
rowData(se)$expected.beta <- 0
rowData(se)[w,"expected.beta"] <- lfcs
rowData(se)$isDE <- FALSE
rowData(se)$isDE[w] <- TRUE
saveRDS(se, file="ipsc.rds")
```

## SEQC

```{r}
suppressPackageStartupMessages({
  library(SummarizedExperiment)
  library(seqc)
})
names(smps) <- smps <- c("ILM_refseq_gene_AGR","ILM_refseq_gene_CNL")
ds <- lapply(smps, FUN=function(x){
  e <- get(x)
  RD <<- e[,1:4]
  row.names(RD) <<- row.names(e) <- e$EntrezID
  CD <- as.data.frame(do.call(rbind,strsplit(names(e)[-1:-4], "_")))
  e <- as.matrix(e[,-1:-4])
  colnames(CD) <- c("Mix","Replicate","Lane","Flowcell")
  row.names(CD) <- colnames(e)
  CD$Replicate <- paste0(CD$Mix,CD$Replicate)
  w <- which(CD$Flowcell=="FlowCellA" & CD$Lane %in% c("L01","L02")) # keep only part of the runs
  CD <- CD[w,]
  sapply(split(row.names(CD), CD$Replicate), FUN=function(x){
    rowSums(e[,x,drop=FALSE])
  })
})
e <- do.call(cbind, ds)
lab <- rep(names(ds), sapply(ds,ncol))
mix <- gsub("[0-9]","",colnames(e))
colnames(e) <- paste(lab, colnames(e), sep=".")

se <- SummarizedExperiment(list(counts=e), rowData=RD,
                           colData=data.frame(lab=lab, mix=mix))
se <- se[,which(se$mix %in% c("C","D"))]
se <- se[,c(1:3,9:10,7:8,14:16)]
se$condition <- as.numeric(se$mix!=unique(se$mix)[2])

# expected log2FC between ERCC mixes
ercc <- c(`ERCC-00002` = -1, `ERCC-00003` = -1, `ERCC-00004` = 2, `ERCC-00009` = 0, 
`ERCC-00012` = -0.58, `ERCC-00013` = -1, `ERCC-00014` = -1, `ERCC-00016` = -0.58, 
`ERCC-00017` = 2, `ERCC-00019` = 2, `ERCC-00022` = -1, `ERCC-00024` = -0.58, 
`ERCC-00025` = 0, `ERCC-00028` = 2, `ERCC-00031` = 0, `ERCC-00033` = 2, 
`ERCC-00034` = 0, `ERCC-00035` = 0, `ERCC-00039` = -0.58, `ERCC-00040` = -0.58, 
`ERCC-00041` = -1, `ERCC-00042` = 0, `ERCC-00043` = -1, `ERCC-00044` = -0.58, 
`ERCC-00046` = -1, `ERCC-00048` = -1, `ERCC-00051` = 0, `ERCC-00053` = 0, 
`ERCC-00054` = -0.58, `ERCC-00057` = -0.58, `ERCC-00058` = -0.58, 
`ERCC-00059` = -1, `ERCC-00060` = 0, `ERCC-00061` = -1, `ERCC-00062` = 2, 
`ERCC-00067` = 0, `ERCC-00069` = -1, `ERCC-00071` = -0.58, `ERCC-00073` = 0, 
`ERCC-00074` = -0.58, `ERCC-00075` = 0, `ERCC-00076` = -0.58, 
`ERCC-00077` = -1, `ERCC-00078` = -1, `ERCC-00079` = -1, `ERCC-00081` = -1, 
`ERCC-00083` = 2, `ERCC-00084` = -0.58, `ERCC-00085` = 2, `ERCC-00086` = -1, 
`ERCC-00092` = 2, `ERCC-00095` = 2, `ERCC-00096` = 0, `ERCC-00097` = 2, 
`ERCC-00098` = -0.58, `ERCC-00099` = -0.58, `ERCC-00104` = 0, 
`ERCC-00108` = 2, `ERCC-00109` = 0, `ERCC-00111` = -0.58, `ERCC-00112` = -1, 
`ERCC-00113` = -0.58, `ERCC-00116` = 2, `ERCC-00117` = 0, `ERCC-00120` = -0.58, 
`ERCC-00123` = 2, `ERCC-00126` = 0, `ERCC-00130` = 2, `ERCC-00131` = 2, 
`ERCC-00134` = 2, `ERCC-00136` = 2, `ERCC-00137` = -1, `ERCC-00138` = 0, 
`ERCC-00142` = 0, `ERCC-00143` = -0.58, `ERCC-00144` = 2, `ERCC-00145` = -0.58, 
`ERCC-00147` = 2, `ERCC-00148` = 0, `ERCC-00150` = 0, `ERCC-00154` = 2, 
`ERCC-00156` = 2, `ERCC-00157` = -0.58, `ERCC-00158` = 0, `ERCC-00160` = -1, 
`ERCC-00162` = -0.58, `ERCC-00163` = -1, `ERCC-00164` = -0.58, 
`ERCC-00165` = -1, `ERCC-00168` = -1, `ERCC-00170` = 2, `ERCC-00171` = 0
)
# for a 1:4 vs 3:4 mixes, we expect half the logfcs:
ercc <- ercc/2
rowData(se)$expected.beta <- NA
rowData(se)[names(ercc),"expected.beta"] <- as.numeric(ercc)
rowData(se)$isDE <- NA
rowData(se)[names(ercc),"isDE"] <- ercc != 0
se <- se[rowVars(assay(se))>0,]
saveRDS(se, file="seqc.rds")
```

## Simulation

Loosely based on C. Leek's simulation, see https://github.com/jtleek/svaseq/blob/master/simulateData.Rmd

```{r}
suppressPackageStartupMessages({
  library(polyester)
})
se <- SummarizedExperiment( list(counts=GSE79636$dat), colData=GSE79636$annotation)
# restrict to more homogeneous samples
se <- se[,se$state=="Insulin resistant" & se$race=="White" & se$Sex=="Male"]
se <- se[rowSums(assay(se)>10)>1,]
e <- assay(se)
n <- nrow(e)
params <- get_params(e)
group <- rep(c(0,1),each=8)
batch1 <- rep(c(0, 1, 0),c(6,8,2)) # batch 1 partially correlated to group
batch2 <- c(rnorm(8,0),rnorm(8,1))[rep(1:8,each=2)+0:1*8] # linear batch effect uncorrelated to group
mod = model.matrix(~ 0 + batch1 + batch2 + group)
set.seed(321)
gco <- bco2 <- bco1 <- rep(0,n)
bco1[sample.int(n, floor(n)/3)] <- rnorm(floor(n)/3,sd=1.5)
bco2[sample.int(n, floor(n)/3)] <- rnorm(floor(n)/3,sd=1.3)
lfcs <- c(0.5, 0.75, 1, 1, 1.5, 2)
gco[sample.int(n, 500)] <- sample(c(lfcs,-lfcs), 500, replace=TRUE)
co <- cbind(batch1=bco1, batch2=bco2, group=gco)
dat0 = create_read_numbers(params$mu,params$fit,params$p0,n=nrow(mod),
                           beta=co,mod=mod,seed=321)
row.names(dat0) <- row.names(e)
se <- SummarizedExperiment(list(counts=dat0), rowData=co,
                           colData=data.frame(batch1=factor(batch1), batch2=batch2,
                                              condition=group))
rowData(se)$expected.beta <- gco
rowData(se)$isDE <- gco!=0
se <- se[rowSums(assay(se)>0)>0,]
saveRDS(se, file="simulation.rds")
```

```{r}
homogenizeDEA <- function(x, g=NULL){
  colnames(x)[which(colnames(x) %in% c("FDR","padj","adj.P.Val"))] <- "FDR"
  colnames(x)[which(colnames(x) %in% c("P.Value","pvalue","PValue"))] <- "PValue"
  colnames(x)[which(colnames(x) %in% c("log2FoldChange","logFC"))] <- "logFC"
  if(is.null(g)) g <- row.names(x)
  x <- x[g,c("logFC","PValue","FDR")]
  row.names(x) <- g
  x
}
```

## Session info

```{r}
sessionInfo()
```

### 2020-04-20 svadea/results/results.Rmd
added SVA-DEA datasets and analysis

---
title: "SVA - DEA benchmark"
author: "Pierre-Luc Germain"
date: "4/15/2020"
output: html_document
---

## Benchmark datasets

* **ipsc**: 10 vs 10 random samples from the GSE79636 dataset, which contains heterogeneous samples (background genetic variation, some batch effects). No further technical variation was added, and a foldchange was added on 300 genes.
* **seqc**: 5 vs 5 samples of seqc mixtures C and D respectively, which include two different spike-in mixes. The samples where selected from two batches with technical differences so that there is weak partial correlation with the mixtures (3:2 vs 2:3). Since the true differences between mixtures are not entirely known, the analysis was performed on all genes but only the spike-ins were considered for benchmarking.
* **simulation**: 8 vs 8 samples were simulated using `polyester` and the means/dispersions from the GSE79636 (restricted to a single batch and biological group), with 500 DEGs and two batch effects: 1) a technical batch partially correlated with the group (6:2) and affecting 1/3 of the genes, and 2) a linear vector of technical variation uncorrelated with the groups and affecting 1/3 of the genes.

See `data_prep.Rmd` for the data preparation code.

## Running the pipeline

```{r, eval=FALSE}
suppressPackageStartupMessages({
  library(SummarizedExperiment)
  library(pipeComp)
})
source(system.file("extdata", "dea_wrappers.R", package="pipeComp"))

pd <- dea_pipeline()

ds <- list.files("../data_preparation", pattern="\\.rds$", full.names = TRUE)
names(ds) <- gsub("\\.rds$","",basename(ds))

alternatives <- list(
  filt=c("none","def.filter"),
  minCount=c(10,20),
  sva.method=c("none","sva.svaseq","sva.vstsva","sva.RUVr","sva.RUVs"),
  k=c(1:5,"be","leek"),
  dea.method=c("dea.edgeR", "dea.edgeR.QLF", "dea.DESeq2", "dea.voom")
)

# we get rid of meaningless/redundant combinations:
eg <- buildCombMatrix(alternatives)
eg <- eg[eg$sva.method %in% c("sva.svaseq", "sva.vstsva") | 
           !(eg$k %in% c("be","leek")),]
eg <- eg[eg$filt=="def.filter" | eg$minCount==10,]
eg <- eg[eg$k=="1" | eg$sva.method!="none",]

res <- runPipeline(ds, alternatives, pd, comb = eg, saveEndResults = TRUE, debug=FALSE, nthreads = 3)
```

```{r, include=FALSE}
res <- readRDS("aggregated.rds")
```

## Results

```{r}
suppressPackageStartupMessages({
  library(ggplot2)
  library(cowplot)
})
theme_set(theme_cowplot())
devtools::load_all("../../pipeComp/")
res1b <- res1 <- res
res1$evaluation$dea <- lapply(res1$evaluation$dea, FUN=function(x) x[which(x$k==1),])
res1b$evaluation$dea <- lapply(res1b$evaluation$dea, FUN=function(x){
  x[x$filt=="def.filter" & x$minCount==10,]
})
```

```{r, fig.width=8, fig.height=9.5}
plot_grid(
  grid.grabExpr(draw(
    dea_evalPlot_logFCs(res1, agg.by=c("sva.method","dea.method"), reorder_rows=FALSE, row_split="sva.method") +
    dea_evalPlot_sig(res1, agg.by=c("sva.method","dea.method"), reorder_rows=FALSE)
  )),
  plotElapsed(res, agg.by=c("sva.method","dea.method")),
  nrow=2, labels=LETTERS[1:2]
)
```

```{r, fig.width=11, fig.height=9}
plot_grid(
  dea_evalPlot_curve(res1, agg.by=c("sva.method","dea.method"), colourBy="sva.method", shapeBy="dea.method") + ggtitle("Using a single surrogate variable"),
  dea_evalPlot_curve(res, agg.by=c("sva.method","k"), colourBy="k", shapeBy="sva.method") + ggtitle("Increasing the number of surrogate variables"),
  nrow=2, labels=LETTERS[1:2]
)
```

```{r, fig.width=12, fig.height=6.5}
plot_grid(
  plot_grid(
    dea_evalPlot_curve(res1, agg.by=c("dea.method","filt","minCount"), colourBy = "dea.method", shapeBy="filter") + ggtitle("DEA methods, Averaging across SVA strategies"),
    dea_evalPlot_curve(res1b, agg.by=c("sva.method")) + ggtitle("SVA methods, averaging across DEA methods"),
    nrow=2, labels=LETTERS[1:2]
  ),
  grid.grabExpr(draw(
    dea_evalPlot_sig(res, agg.by=c("sva.method","k"), reorder_rows=FALSE, 
                     show_column_names=TRUE, anno_legend = FALSE, row_split="sva.method")
    )),
  nrow=1, rel_widths=c(5,2), labels=c(NA,"C")
)
```

### 2020-04-20 svadea/run.Rmd
added SVA-DEA datasets and analysis

---
title: "SVA-DEA pipeline run"
author: "Pierre-Luc Germain"
date: "4/14/2020"
output: html_document
---

## Running the pipeline

```{r}
suppressPackageStartupMessages({
  library(SummarizedExperiment)
#  library(pipeComp)
})
devtools::load_all("../pipeComp/")
#source(system.file("extdata", "dea_wrappers.R", package="pipeComp"))
source("../pipeComp/inst/extdata/dea_wrappers.R")

pd <- dea_pipeline()

ds <- list.files("data_preparation", pattern="\\.rds$", full.names = TRUE)
names(ds) <- gsub("\\.rds$","",basename(ds))

none <- function(x, ...)  return(x)

alternatives <- list(
  filt=c("none","def.filter"),
  minCount=c(10,20),
  sva.method=c("none","sva.svaseq","sva.vstsva","sva.RUVr","sva.RUVs"),
  k=c(1:5,"be","leek"),
  dea.method=c("dea.edgeR", "dea.edgeR.QLF", "dea.DESeq2", "dea.voom")
)
eg <- buildCombMatrix(alternatives)
eg <- eg[eg$sva.method %in% c("sva.svaseq", "sva.vstsva") | 
           !(eg$k %in% c("be","leek")),]
eg <- eg[eg$filt=="def.filter" | eg$minCount==10,]
eg <- eg[eg$k=="1" | eg$sva.method!="none",]

res <- runPipeline(ds[2:3], alternatives, pd, output.prefix = "pipelineResults_test/", debug=FALSE, nthreads = 4)
```

## Results

```{r}
dea_evalPlot_logFCs(res, what=c("logFC.pearson","logFC.spearman"), agg.by=c("sva.method","dea.method"), reorder_rows=FALSE)
dea_evalPlot_curve(res, agg.by=c("sva.method","dea.method"), colourBy="sva.method", shapeBy="dea.method")
dea_evalPlot_sig(res, agg.by=c("sva.method","dea.method"), reorder_rows=FALSE)
dea_evalPlot_curve(res, agg.by=c("sva.method")) + ggtitle("Averaging across DEA methods")
dea_evalPlot_curve(res, agg.by=c("dea.method")) + ggtitle("Averaging across SVA strategies")
```

## Session info

```{r}
sessionInfo()
```

### 2020-05-04 main_figures/svadea.Rmd
added sva-dea results

---
title: "SVA-DEA benchmark"
author: "Pierre-Luc Germain"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r}
suppressPackageStartupMessages({
  library(ggplot2)
  library(cowplot)
  library(pipeComp)
  library(grid)
  library(ComplexHeatmap)
  library(circlize)
})
theme_set(theme_cowplot(font_size = 11))
res1 <- res <- readRDS("../svadea/results/aggregated.rds")
res1$evaluation$dea <- lapply(res1$evaluation$dea, FUN=function(x){
  x[which(x$k==1),]
})
```

```{r, fig.width=11.5, fig.height=6.8}
# hts <- dea_evalPlot_sig(res, what=c("logFC.pearson","TPR"), agg.by=c("sva.method","k"), row_split="sva.method", 
#                      anno_legend=FALSE, show_column_names=TRUE, font_factor=0.85, scale="colCenterScale",
#                      show_heatmap_legend=TRUE, heatmap_legend_param=list(direction="horizontal") ) +
#   dea_evalPlot_sig(res, what="FDR", agg.by=c("sva.method","k"), row_split="sva.method", 
#                      anno_legend=FALSE, show_column_names=TRUE, font_factor=0.85, scale="colCenterScale",
#                      col=rev(viridisLite::inferno(100)), value_cols=c("white","black"), 
#                      show_heatmap_legend=TRUE, heatmap_legend_param=list(direction="horizontal") )
# draw(hts, heatmap_legend_side = "bottom", merge_legend=TRUE)

hts <- dea_evalPlot_sig(res, what=c("logFC.pearson","TPR"), agg.by=c("sva.method","k"), row_split="sva.method", 
                     anno_legend=FALSE, show_column_names=TRUE, font_factor=0.85, scale="colCenterScale") +
  dea_evalPlot_sig(res, what="FDR", agg.by=c("sva.method","k"), row_split="sva.method", 
                     anno_legend=FALSE, show_column_names=TRUE, font_factor=0.85, scale="colCenterScale",
                     col=rev(viridisLite::inferno(100)), value_cols=c("white","black"))
lgd <- Legend(col_fun=colorRamp2(1:100, viridisLite::inferno(100)), at=c(1,100), 
              title="MADs", labels=c("worst","best"), direction="horizontal")
hts <- pipeComp:::.renameHrows(hts, f = function(x) gsub("none > 1", "none",x, fixed=TRUE))
#draw(hts, heatmap_legend_side = "bottom", merge_legend=TRUE, annotation_legend_list=lgd)

p <- plot_grid(
  plot_grid(
    dea_evalPlot_curve(res1, agg.by=c("sva.method","dea.method"), colourBy="dea.method", shapeBy="sva.method") + ggtitle("Using a single surrogate variable"),
    dea_evalPlot_curve(res, agg.by=c("sva.method","k"), colourBy="k", shapeBy="sva.method") + ggtitle("Increasing the number of surrogate variables"),
    nrow=2, labels=LETTERS[1:2]
  ),
  grid.grabExpr(draw( hts, heatmap_legend_side = "bottom", merge_legend=TRUE, 
                      annotation_legend_list=lgd)),
  nrow=1, rel_widths=c(5,2.2), labels=c(NA,"C")
)
pdf("svadea.pdf", width=11.5, height=6.8)
p
dev.off()
p
```

### 2020-05-04 supplementary_figures/supp_svadea.Rmd
added sva-dea results

---
title: "Supplementary Figures relative to the SVA-DEA benchmark"
author: "Pierre-Luc Germain"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r, include=FALSE}
if(!exists('FIG_NB')){
  FIG_NB <- 0
  getFigNb <- function(increment=FALSE){
    if(increment) FIG_NB <<- FIG_NB + 1
    FIG_NB
  }
}

suppressPackageStartupMessages({
  library(ggplot2)
  library(cowplot)
  library(pipeComp)
  library(grid)
  library(ComplexHeatmap)
})
theme_set(theme_cowplot(font_size = 12))
res2 <- res1 <- res <- readRDS("../svadea/results/aggregated.rds")
res1$evaluation$dea <- lapply(res1$evaluation$dea, FUN=function(x){
  x[which(x$k==1),]
})
res2$evaluation$dea <- lapply(res2$evaluation$dea, FUN=function(x){
  x[which(x$k==2 | x$sva.method=="none"),]
})
```

# Supplementary Figure `r getFigNb(TRUE)`

```{r, fig.width=8, fig.height=9.5}
plot_grid(
  grid.grabExpr(draw(
    evalHeatmap(res1, what=c("logFC.pearson", "logFC.mad"), agg.by=c("sva.method","dea.method"), row_split="sva.method", font_factor=.9) +
      evalHeatmap(res1, what=c("TPR", "FDR"), agg.by=c("sva.method","dea.method"), row_split="sva.method", font_factor=.9,
                  filter=threshold==0.05)
  )),
  plotElapsed(res, agg.by=c("sva.method","dea.method")),
  nrow=2, labels=LETTERS[1:2]
)
```

### Supplementary Figure `r getFigNb()`
**A:** Accuracy of the estimated logFC (correlation and median absolute deviation from expected logFCs) and of the differential expression analysis (TPR stands for True Positive Rate, and FDR for False Discovery Rate) across the different combinations of SVA and DEA methods (using max 1 dimension). **B:** Running times of the different methods.

\newpage

# Supplementary Figure `r getFigNb(TRUE)`

```{r, fig.width=8.5, fig.height=8.5}
plot_grid(
  dea_evalPlot_curve(res1, agg.by=c("dea.method","filt","minCount"), shapeBy = "dea.method", colourBy = c("filter")) + ggtitle("Using 1 SV"),
  dea_evalPlot_curve(res1, agg.by=c("sva.method","filt","minCount"), shapeBy = "sva.method", colourBy = c("filter")) + ggtitle("Using 1 SV"),
  dea_evalPlot_curve(res2, agg.by=c("sva.method")) + ggtitle("Using 2 SVs"),
  nrow=3, labels=LETTERS[1:3]
)
```

### Supplementary Figure `r getFigNb()`
Accuracy of the differential expression analysis across combinations of: **A:** filters and DEA methods (using max 1 variable), **B:** filters and SVA methods (using max 1 variable), **C:** SVA methods (using 2 variables). In **C**, filled points indicate a nominal FDR below the observed FDR.

\newpage

### 2020-05-04 svadea/results/results.Rmd
added sva-dea results

@@
 
-* **ipsc**: 10 vs 10 random samples from the GSE79636 dataset, which contains heterogeneous samples (background genetic variation, some batch effects).
-No further technical variation was added, and a foldchange was added on 300 genes.
-* **seqc**: 5 vs 5 samples of seqc mixtures C and D respectively, which include two different spike-in mixes.
-The samples where selected from two batches with technical differences so that there is weak partial correlation with the mixtures (3:2 vs 2:3).
-Since the true differences between mixtures are not entirely known, the analysis was performed on all genes but only the spike-ins were considered for benchmarking.
-* **simulation**: 8 vs 8 samples were simulated using `polyester` and the means/dispersions from the GSE79636 (restricted to a single batch and biological group), with 500 DEGs and two batch effects: 1) a technical batch partially correlated with the group (6:2) and affecting 1/3 of the genes, and 2) a linear vector of technical variation uncorrelated with the groups and affecting 1/3 of the genes.
-
 See `data_prep.Rmd` for the data preparation code.
-
-## Running the pipeline
-
-```{r, eval=FALSE}
-suppressPackageStartupMessages({
-  library(SummarizedExperiment)
-  library(pipeComp)
-})
-source(system.file("extdata", "dea_wrappers.R", package="pipeComp"))
-
-pd <- dea_pipeline()
-
-ds <- list.files("../data_preparation", pattern="\\.rds$", full.names = TRUE)
-names(ds) <- gsub("\\.rds$","",basename(ds))
-
-alternatives <- list(
-  filt=c("none","def.filter"),
-  minCount=c(10,20),
-  sva.method=c("none","sva.svaseq","sva.vstsva","sva.RUVr","sva.RUVs"),
-  k=c(1:5,"be","leek"),
-  dea.method=c("dea.edgeR", "dea.edgeR.QLF", "dea.DESeq2", "dea.voom")
-)
-
-# we get rid of meaningless/redundant combinations:
-eg <- buildCombMatrix(alternatives)
-eg <- eg[eg$sva.method %in% c("sva.svaseq", "sva.vstsva") | 
-           !(eg$k %in% c("be","leek")),]
-eg <- eg[eg$filt=="def.filter" | eg$minCount==10,]
-eg <- eg[eg$k=="1" | eg$sva.method!="none",]
-
-res <- runPipeline(ds, alternatives, pd, comb = eg, saveEndResults = TRUE, debug=FALSE, nthreads = 3)
-```
 
@@
   library(cowplot)
+  library(pipeComp)
+  library(grid)
+  library(ComplexHeatmap)
 })
 theme_set(theme_cowplot())
-devtools::load_all("../../pipeComp/")
 res1b <- res1 <- res
@@
   grid.grabExpr(draw(
-    dea_evalPlot_logFCs(res1, agg.by=c("sva.method","dea.method"), reorder_rows=FALSE, row_split="sva.method") +
-    dea_evalPlot_sig(res1, agg.by=c("sva.method","dea.method"), reorder_rows=FALSE)
+    dea_evalPlot_sig(res1, what=c("logFC.pearson", "logFC.mad", "TPR", "FDR"), agg.by=c("sva.method","dea.method"), row_split="sva.method")
   )),
@@
 
-```{r, fig.width=11, fig.height=9}
+```{r, fig.width=12, fig.height=7}
 plot_grid(
-  dea_evalPlot_curve(res1, agg.by=c("sva.method","dea.method"), colourBy="sva.method", shapeBy="dea.method") + ggtitle("Using a single surrogate variable"),
+  plot_grid(
+    dea_evalPlot_curve(res1, agg.by=c("sva.method","dea.method"), colourBy="dea.method", shapeBy="sva.method") + ggtitle("Using a single surrogate variable"),
     dea_evalPlot_curve(res, agg.by=c("sva.method","k"), colourBy="k", shapeBy="sva.method") + ggtitle("Increasing the number of surrogate variables"),
     nrow=2, labels=LETTERS[1:2]
+  ),
+  grid.grabExpr(draw(
+    dea_evalPlot_sig(res, what=c("logFC.pearson", "TPR", "FDR"), agg.by=c("sva.method","k"), row_split="sva.method", 
+                     anno_legend=FALSE, show_column_names=TRUE)
+  )),
+  nrow=1, rel_widths=c(5,2.5), labels=c(NA,"C")
 )
@@
   grid.grabExpr(draw(
-    dea_evalPlot_sig(res, agg.by=c("sva.method","k"), reorder_rows=FALSE, 
-                     show_column_names=TRUE, anno_legend = FALSE, row_split="sva.method")
+    dea_evalPlot_sig(res, what=c("TPR", "FDR"), agg.by=c("sva.method","k"), row_split="sva.method", 
+                     anno_legend=FALSE, show_column_names=TRUE)
     )),

### 2020-05-04 main_figures/clustering.Rmd
updated many figures to improve readability...

@@
   library(ComplexHeatmap)
-  library(pipeComp)
+  library(pipeComp) # version >= 0.99.26
 })
@@
 o <- c(o[grep("seuratvst",o)],o[grep("seuratvst",o,invert=TRUE)])
-h <- scrna_evalPlot_clust(cl, what="MI", agg.by=c("norm","graph","clustmethod"), agg.fn = median, scale=TRUE, anno_legend = FALSE, reorder_rows=o, row_names_max_width = unit(8, "cm"), row_names_gp = gpar(fontsize = 11)) +
-  scrna_evalPlot_clust(res, what="ARI", agg.by=c("norm","graph","clustmethod"), agg.fn = median, scale=TRUE, reorder_rows = o, anno_legend = FALSE) +
-  scrna_evalPlot_clust(res, what="min_pr", agg.by=c("norm","graph","clustmethod"), agg.fn = median, scale=TRUE, reorder_rows = o, anno_legend = FALSE) +
-  scrna_evalPlot_clust(res, what="ARI", atTrueK = TRUE, agg.by=c("norm","graph","clustmethod"), agg.fn = median, reorder_rows = o, anno_legend = FALSE, row_names_max_width = unit(8, "cm"), row_names_gp = gpar(fontsize = 11))
-h <- pipeComp:::.renameHrows(h, renameFn)
+
+h <- renameHrows(
+  evalHeatmap(res, step="clustering", what=c("MI","ARI","min_pr"), row_split = norm, font_factor = 0.85,
+              agg.by=c("norm","graph","clustmethod"), anno_legend=FALSE) +
+  evalHeatmap(res, step="clustering", what="ARI", agg.by=c("norm","graph","clustmethod"), font_factor = 0.85, filter=n_clus==true.nbClusts, title="ARI at\ntrue k", name="ARIatk",anno_legend=FALSE), function(x) sapply(strsplit(x," > "), FUN=function(x) paste(x[-1],collapse=" > ")))
+
 h2 <- grid.grabExpr(draw(h))
@@
 # keep reasonable parameters for a realistic time
-res$evaluation$clustering <- clo[which(clo$steps>2 & clo$k <=20),]
-h3 <- scrna_evalPlot_clust(res, what="elapsed", agg.by=c("graph","clustmethod"),
-                           anno_legend = TRUE, scale=TRUE, value_format = "%.1f" )
-h3 <- h3 + scrna_evalPlot_clust(res, what="nbClust.diff", 
-                                agg.by=c("graph","clustmethod"), agg.fn = median,
+res$evaluation$clustering <- clo[which(clo$k<=20),]
+h3 <- evalHeatmap(res, step="clustering", what="elapsed", agg.by=c("graph","clustmethod"), font_factor = 0.85,
+                           anno_legend = TRUE, scale=TRUE, value_format = "%.1f", reorder_rows=TRUE,
+                  row_names_max_width = unit(8, "cm"))
+h3 <- h3 + evalHeatmap(res, step="clustering", what="nbClust.diff", 
+                       row_names_max_width = unit(8, "cm"),
+                        agg.by=c("graph","clustmethod"), agg.fn = median, font_factor = 0.85,
                         reorder_rows=h3, scale = FALSE, value_format = "%.1f", 
                         col = circlize::colorRamp2(c(-5, 0, 5), c("red", "white", "Darkblue")),
-                                value_cols = c("black","black"), 
-                                row_names_max_width = unit(8, "cm"),
-                                row_names_gp = gpar(fontsize = 12) )
-h3 <- pipeComp:::.renameHrows(h3, renameFn)
+                        value_cols = c("black","black") )
 h3 <- grid.grabExpr(draw(h3))
@@
 
-```{r clustering, fig.width=12, fig.height=8}
-pdf("clustering.pdf", width=15, height=10)
+```{r clustering, fig.width=11, fig.height=8}
+pdf("clustering.pdf", width=11.5, height=8)
 plot_grid(h2, h3, labels=LETTERS[1:2], ncol=1, rel_heights = c(7,4))

### 2020-05-04 main_figures/dimensionality.Rmd
updated many figures to improve readability...

@@
 ```{r, include=FALSE}
-#devtools::load_all("../../pipeComp_biorxiv/pipeComp/")
+devtools::load_all("../../pipeComp_biorxiv/pipeComp/")
+source("../../pipeComp_biorxiv/pipeComp/inst/extdata/scrna_alternatives.R")
 ```
@@
 
+```{r, include=FALSE}
+# the second part of these results was generated with pipeComp >0.99.26
+devtools::load_all("../../pipeComp/")
+```
+
 ```{r, fig.width=8, fig.height=4}
-res1 <- readRDS("../data/results_dimensionality.rds")
-res2 <- readRDS("../data/results_dimensionality2.rds")
-sw1 <- res1$dimensionality$clust.avg.silwidth$selected
-sw2 <- res2$dimensionality$clust.avg.silwidth$selected
-sw2 <- sw2[grep("Local\\.b",row.names(sw2),invert = TRUE),] # redundant
-sw <- rbind(sw1,sw2[setdiff(row.names(sw2),row.names(sw1)),])
-sw <- sw[grep("=doublet",row.names(sw)),]
-n <- sapply(strsplit(row.names(sw),"="),FUN=function(x) rev(x)[1])
-w <- grep("seuratvst",row.names(sw))
-n[w] <- paste0("sctransform+",n[w])
-row.names(sw) <- n
-sw2 <- sw[order(grepl("sctransform",row.names(sw))*1000+rowMeans(sw)+apply(sw,1,probs=0.2,FUN=quantile),decreasing=TRUE),order(colMeans(sw),decreasing=TRUE)]
-h1 <- Heatmap(sw2, cluster_columns = FALSE, cluster_rows = FALSE, 
-        name="silhouette width", col=pipeComp:::.silScale(sw2), show_column_names = FALSE,
-        column_title = "Subpopulation mean silhouette width",
-        bottom_annotation=pipeComp:::.ds_anno(colnames(sw2)), heatmap_legend_param = list(direction="horizontal", legend_width=unit(3, "cm")))
-p2 <- grid.grabExpr(draw(h1, heatmap_legend_side="bottom"))
+res <- readRDS("../data/results_dimensionality.rds")
+H <- renameHrows(
+  scrna_evalPlot_silh( res, step="dimensionality", reorder_rows=TRUE, 
+                       row_split=gsub("norm\\.","",norm),
+                       heatmap_legend_param=list(direction="horizontal")),
+  function(x) gsub("norm\\.","",x))
+p2 <- grid.grabExpr(draw(H, heatmap_legend_side="bottom"))
 
-cl1 <- res1$clustering
-cl2 <- res2$clustering
-res2 <- rbind(cl1, cl2[setdiff(row.names(cl2),row.names(cl1)), colnames(cl1)])
-res2 <- res2[grep("=doublet",row.names(res2)),]
-method <- as.character(parsePipNames(row.names(res2))$dims)
-w <- grep("seuratvst", row.names(res2))
-method[w] <- paste0("sctransform+",method[w])
-row.names(res2) <- paste0("method=",method,";",row.names(res2))
-
-h2 <- scrna_evalPlot_clust(res2, what="NMI", agg.by = c("method"), scale=TRUE, reorder_rows = TRUE, anno_legend = FALSE)
-h2 <- h2 + 
-  scrna_evalPlot_clust(res2, what="min_pr", agg.by = c("method"), reorder_rows=h2, scale=TRUE, anno_legend = FALSE) + 
-scrna_evalPlot_clustAtTrueK(res2, agg.by = c("method"), reorder_rows=h2, scale=TRUE, anno_legend = FALSE)
-p3 <- grid.grabExpr(draw(h2))
+p3 <- renameHrows(
+  evalHeatmap( res, what=c("NMI","min_pr"), agg.by=c("norm","dims"), 
+                   row_split=gsub("norm\\.","",norm), reorder_rows=TRUE, anno_legend=FALSE) + 
+  evalHeatmap(res, what=c("ARI"), agg.by=c("norm","dims"), row_split=gsub("norm\\.","",norm), 
+              reorder_rows=TRUE, filterExpr=n_clus==true.nbClusts, 
+              anno_legend=FALSE, title="ARI at\ntrue k"),
+  function(x) gsub("norm\\.","",x)
+)
+p3 <- grid.grabExpr(draw(p3))
 ```

### 2020-05-04 main_figures/doublets.Rmd
updated many figures to improve readability...

@@
 pw <- function(i, dot.size=4, ...){
-  scores <- res[[i]][,intersect(names(scoresNames),colnames(res[[i]]))]
+  scores <- res[[i]][,intersect(names(scoresNames), colnames(res[[i]]))]
   colnames(scores) <- scoresNames[colnames(scores)]
   plotROCs(scores, factor(res[[i]]$demuxlet_cls, c("SNG","DBL")), dot.size=dot.size) +
-    ylim(0,1) + scale_color_manual(values=cols)
+    ylim(0,1) + scale_color_manual(values=cols) + 
+    facet_wrap(as.formula(paste0('~"', names(res)[i], '"')))
 }
@@
 p1 <- pw(1) + theme(legend.position="none")
-p2 <- pw(2) + theme(legend.position="none")
-p3 <- pw(3) + theme(legend.position=c(0.25,0.25))
+p2 <- pw(2) + theme(legend.position="none") + ylab("")
+p3 <- pw(3) + theme(legend.position=c(0.25,0.25)) + ylab("")
 ```
@@
 mis <- readRDS("../data/misclassification.rds")
-lf <- list.files("../datasets/", full.names = TRUE)
-lf <- lf[grep("simMix",lf,invert=TRUE)]
-names(lf) <- sapply(strsplit(basename(lf),"\\."), FUN=function(x) x[1])
-CDs <- lapply(lf, FUN=function(x) colData(readRDS(x)))
-dbl <- lapply(CDs, FUN=function(x) x[,c("pldf.ratio","pldf.class")])
+#lf <- list.files("../datasets/", full.names = TRUE)
+#lf <- lf[grep("simMix",lf,invert=TRUE)]
+#names(lf) <- sapply(strsplit(basename(lf),"\\."), FUN=function(x) x[1])
+#CDs <- lapply(lf, FUN=function(x) colData(readRDS(x)))
+dbl <- lapply(CDs[grep("simMix", names(CDs), invert=TRUE)], 
+              FUN=function(x) x[,c("pldf.ratio","pldf.class")])
 dbl <- dbl[intersect(names(dbl),names(mis))]
@@
 
-```{r figure1, fig.height=6, fig.width=7}
-p1m <- plot_grid(p1, p2, p3, nrow=1)
-pdf("doublets.pdf", width=6, height=7)
+```{r figure1, fig.height=6, fig.width=8}
+p1m <- plot_grid(p1, p2, p3, nrow=1, rel_widths = c(9,10,10))
+pdf("doublets.pdf", width=8, height=6)
 plot_grid(p1m, pt, p4, rel_heights=c(3.5,2,2), labels = LETTERS[1:3], ncol = 1, hjust=0, scale=0.95)

### 2020-05-04 main_figures/dr.Rmd
updated many figures to improve readability...

--- 2020-02-03 main_figures/dr.Rmd
@@
 
-```{r dimred, fig.width=12, fig.height=6}
-res <- readRDS("../data/results_dimred.rds")
-x <- res$dimreduction$clust.avg.silwidth
-x <- x[grep("norm\\.seurat\\.noscale.*denoise", row.names(x), invert=TRUE),]
-x <- x[order(colSums(apply(x,1,prob=c(0.05,0.5),FUN=quantile))+
-                         rowMeans(x), decreasing=TRUE),order(colMeans(x), decreasing=TRUE)]
-row.names(x) <- gsub("norm\\.","",apply(parsePipNames(row.names(x))[,c("norm","dr")],1,collapse=" + ",FUN=paste))
-row.names(x) <- gsub("seuratvst","sctransform",row.names(x))
-h1 <- Heatmap(x, cluster_columns = FALSE, cluster_rows = FALSE, 
-        name="silhouette width", col=pipeComp:::.silScale(x), show_column_names = FALSE,
-        column_title = "Subpopulation mean silhouette width", row_names_max_width = unit(9,"cm"),
-        bottom_annotation=pipeComp:::.ds_anno(colnames(x)), heatmap_legend_param = list(direction="horizontal", legend_width=unit(3, "cm")))
-p1 <- grid.grabExpr(draw(h1, heatmap_legend_side="bottom"))
+```{r, include=FALSE}
+devtools::load_all("../../pipeComp/")
+```
 
-res$clustering <- res$clustering[grep("norm\\.seurat\\.noscale.*denoise", row.names(res$clustering), invert=TRUE),]
-h <- scrna_evalPlot_clust(res, what="MI", agg.by=c("norm","dr"), scale=TRUE, anno_legend = FALSE)
-h <- h + scrna_evalPlot_clust(res, what="mean_F1", agg.by=c("norm","dr"), scale=TRUE, reorder_rows = h, anno_legend = FALSE) +
-  scrna_evalPlot_clust(res, what="min_pr", agg.by=c("norm","dr"), scale=TRUE, reorder_rows = h, anno_legend = FALSE) +
-  scrna_evalPlot_clustAtTrueK(res, agg.by=c("norm","dr"), reorder_rows = h, anno_legend = FALSE)
-h <- pipeComp:::.renameHrows(h, function(x) gsub("; dr=", " + ", gsub("seuratvst", "sctransform", x)))
+```{r dimred, fig.width=10.5, fig.height=6}
+res <- readRDS("../data/results_dimred2.rds")
+H <- renameHrows(
+  scrna_evalPlot_silh( res, reorder_rows=TRUE, 
+                       heatmap_legend_param=list(direction="horizontal")),
+  function(x) gsub("norm\\.","",x))
+p1 <- grid.grabExpr(draw(H, heatmap_legend_side="bottom"))
 
-p2 <- grid.grabExpr(draw(h))
+p2 <- renameHrows(
+  evalHeatmap( res, what=c("MI","min_pr"), agg.by=c("norm","dr"), reorder_rows=TRUE, anno_legend=FALSE,
+               title=c("MI","Minimum precision")) + 
+  evalHeatmap(res, what=c("ARI"), agg.by=c("norm","dr"), filterExpr=n_clus==true.nbClusts, 
+              anno_legend=FALSE, title="ARI at true k"),
+  function(x) gsub("norm\\.","",x)
+)
+p2 <- grid.grabExpr(draw(p2))
 
+pdf("dr.pdf", width = 10.5, height=6)
+plot_grid(p1,p2, nrow=2, labels=LETTERS[1:2])
+dev.off()
 plot_grid(p1,p2, nrow=2, labels=LETTERS[1:2])
@@
 
-**A:** Average silhouette width per subpopulation according to various normalization and dimension reductions.
-**B:** Clustering accuracy , mean proportion of the variance in the first 5 components explained by real subpopulations (center), and median ARI of the resulting clustering (right).
+Silhouette width (**A**) per subpopulation and clustering accuracy (**B**) according to various normalization and dimension reductions.

### 2020-05-04 main_figures/normalization.Rmd
updated many figures to improve readability...

@@
 
-(Requires `pipeComp` version >=0.99.3)
+(Requires `pipeComp` version >=0.99.26)
 
@@
 ```{r}
-p1 <- scrna_evalPlot_DR(res, "log10_total_counts", agg.by = "norm", anno_legend=FALSE)
-p1 <- p1 + scrna_evalPlot_DR(res, "log10_total_features", agg.by = "norm", anno_legend=FALSE, reorder_rows=p1)
-p1 <- grid.grabExpr(draw(renameNorm(p1)))
+classifyNorms <- function(x){
+  y <- rep("other", length(x))
+  y[grep("sctransform",x)] <- "sctransform"
+  y[grep("seurat",x)] <- "seurat"
+  y[grep("scran",x)] <- "scran"
+  y[grep("stableG",x)] <- "stableG"
+  y
+}
 
-p2 <- grid.grabExpr(draw(
-  renameNorm(scrna_evalPlot_DR(res, what="meanSilWidth", agg.by = "norm", value_format = "", show_heatmap_legend = TRUE))))
+p1 <- evalHeatmap(res, step="dimreduction", what=c("log10_total_counts","log10_total_features"), 
+                  what2="PC1.covar.adjR2", agg.by = "norm", reorder_rows = TRUE, 
+                  row_split = classifyNorms(norm), row_title="Variance in PC1 explained by covariate")
+lgd <- Legend(col_fun=colorRamp2(1:100, viridisLite::inferno(100)), at=c(1,50,100), 
+              title="MADs", labels=c("lowest","median","highest"))
+p1 <- grid.grabExpr(draw(renameNorm(p1), merge_legend=TRUE, annotation_legend_list=lgd))
 
-H <- scrna_evalPlot_clust(res, "MI", agg.by = "norm", scale = TRUE, reorder_rows=TRUE, anno_legend=FALSE)
-H <- H + scrna_evalPlot_clust(res, "ARI", agg.by = "norm", scale = TRUE, reorder_rows = row.names(H@matrix), anno_legend=FALSE) +
-  scrna_evalPlot_clust(res, "ARI", atTrueK = TRUE, agg.by = "norm", scale = TRUE, reorder_rows = row.names(H@matrix), anno_legend=FALSE)
-p3 <- grid.grabExpr(draw(renameNorm(H)))
+p2 <- grid.grabExpr(draw(renameNorm(
+  scrna_evalPlot_silh(res, reorder_rows = TRUE, row_split = classifyNorms(norm),
+                      row_title="Subpopulations silhouette width", anno_legend=FALSE)
+  ), annotation_legend_side="right"))
+
+lgd <- Legend(col_fun=colorRamp2(1:100, viridisLite::inferno(100)), at=c(1,50,100), 
+              title="MADs", labels=c("worst","median","best"), direction="horizontal")
+p3 <- grid.grabExpr(draw(
+  renameNorm(evalHeatmap(res, step="clustering", what=c("MI","ARI"), agg.by = "norm", 
+              reorder_rows=TRUE, anno_legend=FALSE, row_split=classifyNorms(norm), row_title="Clustering accuracy") + 
+    evalHeatmap(res, step="clustering", what="ARI", agg.by = "norm", 
+                filter=n_clus==true.nbClusts, reorder_rows=TRUE, anno_legend=FALSE,
+                row_split=classifyNorms(norm), name="ARI at true k", title="ARI at true k")),
+  heatmap_legend_side = "bottom", merge_legend=TRUE, annotation_legend_list=lgd))
 ```
@@
 pdf("normalization.pdf", width=10, height=11)
-plot_grid(p1,p2,p3, nrow = 3, labels=LETTERS[1:3], rel_heights = c(10,9,10))
+  plot_grid(p1,p2,p3, nrow = 3, labels=LETTERS[1:3], rel_heights = c(9,9,10))
 dev.off()
-plot_grid(p1,p2,p3, nrow = 3, labels=LETTERS[1:3], rel_heights = c(10,9,10))
+plot_grid(p1,p2,p3, nrow = 3, labels=LETTERS[1:3], rel_heights = c(9,9,10))
 ```
@@
 The values are difference in adjusted _R^2_ between a `PC1~subpopulation+covariate` model and `PC1~subpopulation`.
-**B:** Average silhouette width per true subpopulation, where higher silhouette width means a higher separability.
+**B:** Minimum and mean silhouette width per true subpopulation, where higher silhouette width means a higher separability.
 **C:** Clustering accuracy, measured by the mutual information (MI), the adjusted Rand Index (ARI), and the ARI at the true number of cluster.
+The coloring scheme in panels **A** and **B** is based on the per-metric number of median absolute deviations (MADs) from the per-column medians, and mapped to color using a square-root transformation to minimize the effect of outliers.
+The values shown in the cells are the direct metric values (no transformation except rounding).

### 2020-05-04 bmc_article.tex
Updates from Overleaf

@@
 We evaluate common steps of such analyses, including filtering, doublet detection, normalization, feature selection, denoising, dimensionality reduction and clustering.
-On the basis of these analyses, we make a number of concrete recommendations about current analysis choices.
+On the basis of these analyses, we make a number of practical recommendations about current analysis choices.
 The evaluation framework, \textit{pipeComp}, has been implemented so as to easily integrate any other step or tool, allowing extensible benchmarks and easy application to other fields (\url{https://github.com/plger/pipeComp}).
@@
 
-\parttitle{Conclusion} On the basis of our systematic evaluations of analysis pipelines, we make a number of concrete recommendations about current analysis choices and for a comprehensive pipeline.
+\parttitle{Conclusion} On the basis of our systematic evaluations of analysis pipelines, we make a number of practical recommendations about current analysis choices and for a comprehensive pipeline.
 The evaluation framework that we developed, \textit{pipeComp}, has been implemented so as to easily integrate any other step, tool, or evaluation metric allowing extensible benchmarks and easy applications to other fields of reseach in scRNAseq or Bioinformatics (\url{https://github.com/plger/pipeComp}).
@@
 However these recommendations need constant updating and often leave open many details of an analysis.
-Another missing aspect of current benchmarking studies is their limitation to capture all aspects of scRNAseq processing workflow.
+Another missing aspect of current benchmarking studies is their limitation to capture all aspects of a scRNAseq processing workflow.
 Although previous benchmarks already brought valuable recommendations for data processing, some only focused on one aspect of data processing (e.g., \citealp{SunDimRed2019}), did not evaluate how the tool selection affects downstream analysis (e.g., \citealp{TsuyuzakiPCA2020}) or did not tackle all aspects of data processing, such as doublet identification or cell filtering (e.g., \citealp{viethSystematic2019}).
@@
 Therefore, instead of evaluating each step in isolation, we developed a general framework for evaluating nested variations on a pipeline and suggest a multilevel panel of metrics.
-Finally, we evaluate several recent methods and provide concrete recommendations.
+Finally, we evaluate several recent methods and provide practical recommendations.
 
@@
 
-We next investigated the impact of filtering according to various criteria (see Methods).
+We next investigated the impact of filtering according to six methods with varying stringency; three methods based on deviations to MADs with increasing levels of stringency (named \textit{lenient}, \textit{default}, \textit{stringent}, \textit{very stringent}) and two methods based on \textit{scater}'s \textit{runPCA} using all (\textit{pca}) or a restricted number of covariates (\textit{pca2}, see Methods for more details).
 An important risk of excluding cells on the basis of their distance from the whole distribution of cells on some properties (e.g., library size) is that these properties tend to have different distributions across subpopulations.
@@
 We therefore calculated the per-subpopulation precision and recall using the Hungarian algorithm \cite{PapadimitriouHu1998} and monitored their mean F1 score.
-A first observation was that, although the stringent filtering tended to be associated with an increase in accuracy, it could also become deleterious and most of the benefits could be achieved without very stringent filtering and minimizing subpopulation bias (Figure \ref{fig:figure4}B).
-Applying the same filtering criteria on individual clusters of cells (identified through scran's \texttt{quickCluster} method) resulted in nearly no cell being filtered out.
+A first observation was that, although the stringent filtering tended to be associated with an increase in accuracy, it could also become deleterious.
+Most of the benefits could be achieved without very stringent filtering and minimizing subpopulation bias (Figure \ref{fig:figure4}B).
+Applying the same filtering criteria on individual clusters of cells (identified through scran's \texttt{quickCluster} method) rather than on whole dataset resulted in nearly no cells being filtered out.
 This suggests that filtering on the global population tends to discard cells of subpopulations with more extreme properties (e.g.
-high library size), rather than low-quality cells.
-Finally, by changing the distributions, the doublet removal step in conjunction with filtering sometimes resulted in a net decrease in the proportion of excluded cells while retaining or improving accuracy.
-We therefore recommend the use of doublet removal followed by relatively mild filtering, such as is implemented in our `default' filtering (see Methods).
+high library size), rather than low-quality cells, and that more stringent filtering may only remove the cells of these subpopulations.
+On the opposite, lenient or lack of filtering leads to no/ little impact on subpopulations.
+This comes at the cost that lenient filtering performs worse or equally well than other methods among all datasets.
+However, a mild filtering ('default') often leads to higher clustering performance (e.g.
+simMix and Zhengmix4 datasets) or to very little loss (e.g.
+Koh and Zhengmix8eq) and always retain most of the cells in each subpopulation.
+This suggests that a mild ('default') filtering is a very good trade-off between clustering accuracy and subpopulations preservation.
+Finally, the doublet removal step in conjunction with filtering  resulted in a net decrease in the proportion of excluded cells while retaining or improving accuracy among some 10x datasets (mixology, Zheng datasets).
+On the Smart-Seq and simulated datasets however, the removal of doublets had little to no impact on any of the two measures.
+Overall, we therefore recommend the use of doublet removal followed by a `default' filtering (see Methods) or a similar approach.
 
@@
 We next investigated the impact of different normalization strategies.
-Beside the standard log-normalization included in \textit{Seurat}, we tested \textit{scran}'s pooling-based normalization \cite{lunPooling2016}, \textit{sctransform}'s variance-stabilizing transformation \cite{hafemeisterSCtransform2019}, normalization based on stable genes \cite{linStableGenes2018, deekeStablyExpressed2018}, and SCnorm \cite{bacherSCnorm2017} (\textit{scVI} \cite{lopezDeep2018} was included separately on a subset of parameter combinations due to high running times).
+Beside the standard log-normalization included in \textit{Seurat}, we tested \textit{scran}'s pooling-based normalization \cite{lunPooling2016}, \textit{SCTransform}'s variance-stabilizing transformation \cite{hafemeisterSCtransform2019}, normalization based on stable genes \cite{linStableGenes2018, deekeStablyExpressed2018}, and SCnorm \cite{bacherSCnorm2017} (\textit{scVI} \cite{lopezDeep2018} was included separately on a subset of parameter combinations due to high running times).
 In addition to log-normalization, the standard \textit{Seurat} clustering pipeline performs per-feature unit-variance scaling so that the PCA is not too strongly dominated by highly-expressed features.
-We therefore included versions of the different normalization procedures, with or without a subsequent scaling (\textit{sctransform}'s variance-stabilizing transformation involves an approach analogous to scaling).
-\textit{Seurat}'s scaling function (and \textit{sctransform}) also includes the option to regress out the effect of certain covariates.
+We therefore included versions of the different normalization procedures, with or without a subsequent scaling (\textit{SCTransform}'s variance-stabilizing transformation involves an approach analogous to scaling).
+\textit{Seurat}'s scaling function (and \textit{SCTransform}) also includes the option to regress out the effect of certain covariates.
 We tested its performance by using the proportion of mitochondrial counts and the number of detected features as covariates.
@@
 
-An important motivation for the development of \textit{sctransform} was the observation that, even after normalization, the first principal components of various datasets tended to correlate with library size, suggesting an inadequate normalization \cite{hafemeisterSCtransform2019}.
+An important motivation for the development of \textit{SCTransform} was the observation that, even after normalization, the first principal components of various datasets tended to correlate with library size, suggesting an inadequate normalization \cite{hafemeisterSCtransform2019}.
 However, as library size tends to vary across subpopulations, part of this effect can simply reflect biological differences.
 We therefore assessed to what extent the first principal component still retained a correlation with the library size and the number of detected features, removing the confounding covariation with the subpopulations (Figure \ref{fig:figure5}A and Supplementary Figure 12).
-Scaling tended to remove much of the correlation with these features, and while most methods were able to remove most of the effect, the correlation was lowest with \textit{sctransform}.
+Scaling tended to remove much of the correlation with these features, and while most methods were able to remove most of the effect, the correlation was lowest with \textit{SCTransform}.
 Surprisingly, however, regressing out the covariates tended to increase the association with it.
@@
 We found most methods (including no normalization at all) to perform fairly well in most of the subpopulations.
-An exception was \textit{scVI} (Supplementary Figure 13), which performed poorly on most datasets in terms of both the subpopulation silhouette and clustering accuracy.
-For time and resource cost reasons, \textit{scVI} was run on a reduced set of pipelines and compared to \textit{sctransform} (more details in the Methods section).
+In view of the subpopulation silhouette and ARI at true number of clusters, it seems that all methods perform worse on the three Zhengmix datasets, that are the datasets with the highest number of cells and the smaller feature detection rate.
 Scaling tended to reduce the average silhouette width of some subpopulations and to increase that of some less distinguishable ones and was generally, but not always, beneficial on the accuracy of the final clustering.
-Regressing out covariates systematically gave poorer performance on all metrics.
-\textit{sctransform} systematically outperformed other methods and, even though it was developed to be applied to data with unique molecular identifiers (UMI), it also performed fairly well with the Smart-seq protocol (Koh and Kumar datasets).
+Regressing out covariates systematically gave equal  or poorer performance on MI, ARI and ARI at true number of clusters, with a notable negative impact when \textit{Seurat}'s regressing was applied to the Koh and Kumar datasets.
+It is possible that the difference between \textit{Seurat} and \textit{SCTransform}'s regression originates from the use of Pearson's residuals performed during the second round regression of \textit{SCTransform}, which are then independent of sequencing depth.
+\textit{SCTransform} systematically outperformed other methods and, even though it was developed to be applied to data with unique molecular identifiers (UMI), it also performed fairly well with the Smart-seq protocol (Koh and Kumar datasets).\textit{scVI} (Supplementary Figure 13) performed poorly on most datasets in terms of both the subpopulation silhouette and clustering accuracy.
+For time and resource cost reasons, \textit{scVI} was run on a reduced set of pipelines and compared to \textit{SCTransform} (more details in the Methods section).
 
 Finally, we monitored whether, under the same downstream clustering analysis, different normalization methods tended to lead to an over- or under-estimation of the number of clusters.
-Although some methods had a tendency to lead to a higher (e.g., \textit{sctransform}) or lower (e.g., stable genes) number of clusters, the effect was very mild and not entirely systematic (Supplementary Figure 14).
-We also confirmed that, especially in the UMI-based datasets, \textit{sctransform} did in fact successfully stabilize variance across mean counts (Supplementary Figure 15), at little apparent cost on computation time (Supplementary Figure 16).
+Although some methods had a tendency to lead to a higher (e.g., \textit{SCTransform}) or lower (e.g., stable genes) number of clusters, the effect was very mild and not entirely systematic (Supplementary Figure 14).
+We also confirmed that, especially in the UMI-based datasets, \textit{SCTransform} did in fact successfully stabilize variance across mean counts (Supplementary Figure 15), at little apparent cost on computation time (Supplementary Figure 16).
 
@@
 Another common practice in feature selection is to use those with the highest mean expression.
-Recently, it was instead suggested to use deviance \cite{townesGlmpca2019}, while \textit{sctransform} provides its own ordering of genes based on transformed variance.
+Recently, it was instead suggested to use deviance \cite{townesGlmpca2019}, while \textit{SCTransform} provides its own ordering of genes based on transformed variance.
 
 Reasoning that a selection method should ideally select genes whose variability is higher between subpopulations than within, we first assessed to what extent each method selected genes with a high proportion of variance or deviance explained by the (real) subpopulations.
-As the proportion of variability in a gene attributable to subpopulations can be measured in various ways, we first compared three approaches: ANOVA on log-normalized count, ANOVA on \textit{sctransform} normalization, and deviance explained.
-The ANOVAs performed on a standard \textit{Seurat} normalization and on \textit{sctransform} data were highly correlated (Supplementary Figure 17A).
+As the proportion of variability in a gene attributable to subpopulations can be measured in various ways, we first compared three approaches: ANOVA on log-normalized count, ANOVA on \textit{SCTransform} normalization, and deviance explained.
+The ANOVAs performed on a standard \textit{Seurat} normalization and on \textit{SCTransform} data were highly correlated (Pearson correlation >0.9, Supplementary Figure 17A).
 These estimates were also in good agreement with the deviance explained, although lowly-expressed genes could have a high deviance explained without having much of their variance explained by subpopulation (Supplementary Figure 17B-D).
@@
 Regarding the percentage of deviance explained however, the standardized measures were often superior (Figure \ref{fig:figure6}A and Supplementary Figures 18-19).
-Deviance proved the method of choice to prioritize genes with a high variance explained by subpopulations (with mere expression level proving surprisingly performant) but did not perform so well to select genes with a high deviance explained.
+\textit{Deviance} proved the method of choice to prioritize genes with a high variance explained by subpopulations (with mere expression level proving surprisingly performant) but did not perform so well to select genes with a high deviance explained.
 
@@
 Interestingly, while these selections were on average the top-ranking methods, they were not systematically best for all datasets.
-The previous observations were reflected in the ARI of the resulting clustering (Figure \ref{fig:figure6}B): non-standardized measures of variability, including mere expression level, tended to outperform more complex metrics.
+The ARI at true number of clusters was near perfect (>0.94) among the smallest datasets but among bigger datasets (simMix and ZhengMix) the performance of the methods was lower and more variable (Figure \ref{fig:figure6}B).
+Non-standardized measures of variability, including mere expression level, tended to outperform more complex metrics.
 In general, we found deviance and unstandardized estimates of variance to provide the best results across datasets and normalization methods.
@@
 Since the various PCA approaches and implementations were recently benchmarked in a similar context \cite{TsuyuzakiPCA2020}, we focused on widely used approaches that had not yet been compared: \textit{Seurat}'s PCA, \textit{scran}'s \texttt{denoisePCA}, and GLM-PCA \cite{townesGlmpca2019}.
-When relevant, we combined them with \textit{sctransform} normalization.
+When relevant, we combined them with \textit{SCTransform} normalization.
 Given that \textit{Seurat}'s default PCA weights the cell embeddings by the variance of each component, we also evaluated the impact of this weighting with each method.
@@
 The impact of the choice of dimensionality reduction method was far greater than that of normalization or feature selection (Figure \ref{fig:figure7}).
-GLM-PCA tended to increase the average silhouette width of already well-defined subpopulations, but \textit{Seurat}'s PCA procedure however proved superior on all metrics.
+Again, the clustering on the biggest datasets (simMix and Zhengmix) was more challenging for most methods.
+Several of them could not identify the true number of clusters in the simMix1 and Zhengmix8eq datasets, which have the highest number of subpopulations, or lead to very low mean silhouette width and accuracy.
+Overall however, the clustering accuracy was higher for methods that include weighting or scaling.
+Among them, GLM-PCA tended to increase the average silhouette width of already well-defined subpopulations, but \textit{Seurat}'s PCA procedure however proved superior on all metrics.
 Like GLM-PCA, \textit{scVI}'s Linear Decoder (LD) does not explicitly rely on normalized counts.
@@
 
-Different approaches have been proposed to select the appropriate number of dimensions, from the visual identification of an 'Elbow' (inflexion point) of the variance explained, to more complex algorithms.
+Different approaches have been proposed to select the appropriate number of dimensions, from the visual identification of an 'elbow' (inflexion point) of the variance explained, to more complex algorithms.
 We evaluated the performance of dimensionality estimators implemented in the \textit{intrinsicDimension} package \cite{johnssonDimensionality2015}, as well as common procedures such as the `elbow` method (inflexion point in the variance explained by each component), some scRNAseq-specific methods such as the JackStraw procedure \cite{ChungJackstraw2015} or \textit{scran}'s \texttt{denoisePCA} \cite{LunScran2016}, and the recent application of Fisher Separability analysis \cite{AlberganteSepar2019}.
@@
 Of note, the methods differ widely in view of their computing time (Figure \ref{fig:figure8}B) and we saw no relationship between the accuracy of the estimates and the complexity of the method.
-Reasoning that over-estimating the number of dimensions is less problematic than under-estimating it, we kept the former methods for a full analysis of their impact on clustering (Figure \ref{fig:figure8}C-D), when combined with \textit{sctransform} or the standard \textit{Seurat} normalization.
+Reasoning that over-estimating the number of dimensions is less problematic than under-estimating it, we kept the former methods for a full analysis of their impact on clustering (Figure \ref{fig:figure8}C-D), when combined with \textit{SCTransform} or the standard \textit{Seurat} normalization.
 Although most methods performed well on the various clustering measures, the global maximum likelihood based on translated Poisson Mixture Model (\textit{maxLikGlobal}, using the 20 nearest neighbors, although estimates were relatively stable across decent values of \textit{k}; see Supplementary Figure 23) provided the dimensionality estimate that best separated the subpopulations (Figure \ref{fig:figure8}C) and resulted in the best clustering accuracy (Figure \ref{fig:figure8}D).
@@
 Specifically, we compared rank-based graphs (\textit{scran}'s default) with graphs using the components' values directly, SNN vs KNN graphs, and the exact nearest neighbors to the Annoy approximation.
-Again, the tested methods were combined with \textit{Seurat}'s standard normalization and \textit{sctransform}, otherwise using the parameters found optimal in the previous steps.
+Again, the tested methods were combined with \textit{Seurat}'s standard normalization and \textit{SCTransform}, otherwise using the parameters found optimal in the previous steps.
 
@@
 
-\subsection*{Concrete recommendations}
+\subsection*{Practical recommendations}
 
-On the basis of our findings, we can make a number of concrete analysis recommendations, also summarized in Figure \ref{fig:figure11}:
+On the basis of our findings, we can make a number of practical analysis recommendations, also summarized in Figure \ref{fig:figure11}:
 
@@
      \item Most normalization methods tested yielded a fair performance, especially when combined with scaling, which tended to have a positive impact on clustering.
-     \item \textit{sctransform} offered the best overall performance in terms of the separability of the subpopulations, as well as removing the effect of library size and detection rate.
+     \item \textit{SCTransform} offered the best overall performance in terms of the separability of the subpopulations, as well as removing the effect of library size and detection rate.
      \item The common practice of regressing out cell covariates, such as the detection rate or proportion of mitochondrial reads nearly always had a negative impact, leading to increased correlation with covariates and decreased clustering accuracy.
@@
 We also considered only methods relying on Euclidean distance, while correlation was recently reported to be superior\cite{kim_impact_2019} and would require further investigation.
+
+Among all pipelines that were tested, the performance varied greatly among datasets.
+The ZhengMix8eq and simMix1 datasets were the most challenging datasets across all preprocessing steps that we tested.
+The average silhouette width of half of their subpopulations was systematically lower that the other subpopulations/ datasets and the ARI score at true number of clusters was always poor or unable to achieve.
+In comparison to the performance on other datasets, this inconsistency was particularly striking when evaluating the dimensionality estimation dimensionality reduction and the clustering methods, for which half of the evaluated methods could not lead to the true number of clusters.
+It is likely that these two datasets are challenging due to their combined sparsity and complexity (see Figure 1 \ref{fig:figure1}).
+This issue that was already observed for the Zhengmix8eq dataset in a previous benchmark \cite{duoClustering2018}, as it contains deeper levels of annotations for subpopulations of T-cells.
+The Koh dataset harbors as many populations, but originates from more distinct groups (from hESC in different differentiation stages).
+As the Kumar dataset, it harbors a very low subpopulations complexity and most methods applied on them performed well.
 
@@
 \textit{pipeComp} is a flexible R framework for evaluating methods and parameter combinations in a complex pipeline, computing on-the-fly multilevel evaluation metrics.
-Applying this framework to scRNAseq clustering enabled us to make concrete recommendations on the steps of filtering, normalization, feature selection, denoising, dimensionality reduction and clustering (Figure \ref{fig:figure11}).
+Applying this framework to scRNAseq clustering enabled us to make clear recommendations on the steps of filtering, normalization, feature selection, denoising, dimensionality reduction and clustering (Figure \ref{fig:figure11}).
 We demonstrate how a diversity of multilevel metrics can be more robust, more sensitive, and more nuanced than simply evaluating the final clustering performance.
@@
 \subsection*{Software and package versions}
-Analyses were performed in R 3.6.0 (Bioconductor 3.9) and the following packages were installed from GitHub repositories: \textit{Seurat} (version 3.0.0), \textit{sctransform} (0.2.0), \textit{DoubletFinder} (2.0.1), \textit{scDblFinder} (1.1.4), \textit{scds} (1.0.0), \textit{SAVERX} (1.0.0), \textit{scImpute} (0.0.9), \textit{ALRA} (commit 7636de8), \textit{DCA} (0.2.2), \textit{DrImpute} (1.2), \textit{ENHANCE} (R version, commit 1571696), \textit{SAVERX} (1.0.0).
+Analyses were performed in R 3.6.0 (Bioconductor 3.9) and the following packages were installed from GitHub repositories: \textit{Seurat} (version 3.0.0), \textit{SCTransform} (0.2.0), \textit{DoubletFinder} (2.0.1), \textit{scDblFinder} (1.1.4), \textit{scds} (1.0.0), \textit{SAVERX} (1.0.0), \textit{scImpute} (0.0.9), \textit{ALRA} (commit 7636de8), \textit{DCA} (0.2.2), \textit{DrImpute} (1.2), \textit{ENHANCE} (R version, commit 1571696), \textit{SAVERX} (1.0.0).
 The code for the glmPCA was obtained from \url{https://github.com/willtownes/scrna2019} (commit 1ddcc30ebb95d083a685f12fe81d35dd1b0cb1b2).
@@
 
-Unless specified otherwise, the variance in gene expression explained by subpopulations was calculated on the data normalized and transformed through \textit{sctransform}.
+Unless specified otherwise, the variance in gene expression explained by subpopulations was calculated on the data normalized and transformed through \textit{SCTransform}.
 For each gene, we fitted a linear model using the subpopulation as only independent variable ($\sim subpopulation$) and used the R-squared as a measure of the variance explained.
@@
 The evaluation of \textit{scVI} algorithm was performed separate to the other pipeline combinations as the model training on our reference datasets was especially resource-consuming.
-As it was not practically feasible to perform the training step on many combination of parameters as it was done with other tools, we evaluated it on a representative subset of pipelines using \textit{sctransform} normalization and \textit{Seurat}'s PCA dimension reduction method.
+As it was not practically feasible to perform the training step on many combination of parameters as it was done with other tools, we evaluated it on a representative subset of pipelines using \textit{SCTransform} normalization and \textit{Seurat}'s PCA dimension reduction method.
 The resulting evaluation is available in the Supplementary Figure 13.
@@
 \textbf{C:} Clustering accuracy (\textit{Seurat} clustering), measured by mutual information (MI), Adjusted Rand Index (ARI) and ARI at the true number of cluster.
+Grey squares indicate that the true number of clusters could not be reached with the corresponding method and dataset.
 `feat\_regress', `mt\_regress' and `feat\_mt\_regress' respectively stand for the regressing out of the number of features, the proportion of mitochondrial reads, or both during scaling.}
@@
 \textbf{B:} Accuracy of clusterings (at the true number of clusters) when selecting 1000 genes using the given methods.
-Based on standard \textit{Seurat} normalization (left) or \textit{sctransform} (right).
+Based on standard \textit{Seurat} normalization (left) or \textit{SCTransform} (right).
 The  \textit{vst.varExp} and \textit{devianceExplained} methods correspond to the estimates used in \textbf{A} to evaluate the selection methods and were included here only for validation purpose.}
@@
 \textbf{B:} Difference between `real' dimensionality (from \textbf{A}) and dimensionality estimation methods, along with computing time.
-\textbf{C:} Average silhouette width per subpopulation across a selection of methods, combined with \textit{sctransform} or standard \textit{Seurat} normalization.
+\textbf{C:} Average silhouette width per subpopulation across a selection of methods, combined with \textit{SCTransform} or standard \textit{Seurat} normalization.
 \textbf{D:} Clustering accuracy across the normalization/dimensionality estimation methods.}
@@
     \fi
-    \caption{\textbf{Evaluation of clustering methods.} Clustering accuracy (\textbf{A}) and run time (\textbf{B}) of \textit{scran}- and \textit{Seurat}-based clustering tools in combination with \textit{sctransform} and standard \textit{Seurat}'s normalization.}
+    \caption{\textbf{Evaluation of clustering methods.} Clustering accuracy (\textbf{A}) and run time (\textbf{B}) of \textit{scran}- and \textit{Seurat}-based clustering tools in combination with \textit{SCTransform} and standard \textit{Seurat}'s normalization.}
     \label{fig:figure9}

### 2020-05-04 draft.tex
Updates from Overleaf

@@
 
-Different approaches have been proposed to select the appropriate number of dimensions, from the visual identification of an 'Elbow' (inflexion point) of the variance explained, to more complex algorithms.
-We evaluated the performance of dimensionality estimators implemented in the \textit{intrinsicDimension} package \cite{johnssonDimensionality2015}, as well as common procedures such as the `elbow` method (inflexion point in the variance explained by each component), some scRNAseq-specific methods such as the JackStraw procedure \cite{ChungJackstraw2015} or \textit{scran}'s \texttt{denoisePCA} \cite{LunScran2016}, and the recent application of Fisher Separability analysis \cite{AlberganteSepar2019}.
+Different approaches have been proposed to select the appropriate number of dimensions, from the visual identification of an `elbow' (inflexion point) of the variance explained by each component, to more complex algorithms.
+We evaluated the performance of dimensionality estimators implemented in the \textit{intrinsicDimension} package \cite{johnssonDimensionality2015}, as well as common procedures such as the elbow method, some scRNAseq-specific methods such as the JackStraw procedure \cite{ChungJackstraw2015} or \textit{scran}'s \texttt{denoisePCA} \cite{LunScran2016}, and the recent application of Fisher Separability analysis \cite{AlberganteSepar2019}.
 
@@
 
-All analyses were performed through the \href{https://github.com/plger/pipeComp}{pipeComp} R package, which implements the pipeline framework described here.
+All analyses were performed through the \href{https://github.com/plger/pipeComp}{pipeComp} Bioconductor package, which implements the pipeline framework described here.
 scVI results and results of the clustering section were produced with \textit{pipeComp} 0.99.4, while all other results were produced with version 0.99.2 (\textit{biorxiv}) branch of the repository).
@@
   \hline
-dataset & source & protocol & description \\ 
+dataset & source & # cells & protocol & description \\ 
   \hline
-Koh & GSE85066 & SMARTer & FACS purified H7 hESC in different differention stages \\ 
-  Kumar & GSE60749 & SMARTer & Mouse ESC cultured in different conditions \\ 
-  Zhengmix4eq & \cite{duoClustering2018} & 10x & Mixtures of FACS purified PBMCs \\ 
-  Zhengmix4uneq & \cite{duoClustering2018} & 10x & Mixtures of FACS purified PBMCs \\ 
-  Zhengmix8eq & \cite{duoClustering2018} & 10x & Mixtures of FACS purified PBMCs \\ 
-  mixology10x3cl & \cite{tianMixology2018} & 10x & Mixture of 3  cancer cell lines from CellBench \\ 
-  mixology10x5cl & \cite{tianMixology2018} & 10x & Mixture of 5 cancer cell lines from CellBench \\ 
-  simMix1 & - & 10x-based & Simulation of 10 human cell subpopulations \\
-  simMix2 & - & 10x-based & Simulation of 9 mouse cell subpopulations \\
+Koh & GSE85066 & 531 & SMARTer & FACS purified differentiation stages \\ 
+  Kumar & GSE60749 & 246 & SMARTer & Mouse ESC cultured in different conditions \\ 
+  Zhengmix4eq & \cite{duoClustering2018} & 3994 & 10x & Mixtures of FACS purified PBMCs \\ 
+  Zhengmix4uneq & \cite{duoClustering2018} & 6498 & 10x & Mixtures of FACS purified PBMCs \\ 
+  Zhengmix8eq & \cite{duoClustering2018} & 3994 & 10x & Mixtures of FACS purified PBMCs \\ 
+  mixology10x3cl & \cite{tianMixology2018} & 902 & 10x & Mixture of 3  cancer cell lines from CellBench \\ 
+  mixology10x5cl & \cite{tianMixology2018} & 3918 & 10x & Mixture of 5 cancer cell lines from CellBench \\ 
+  simMix1 & - & 2500 & 10x-based & Simulation of 10 human cell subpopulations \\
+  simMix2 & - & 3000 & 10x-based & Simulation of 9 mouse cell subpopulations \\
    \hline

### 2020-05-04 draft.tex
Updates from Overleaf

@@
 
-All analyses were performed through the \href{https://github.com/plger/pipeComp}{pipeComp} Bioconductor package, which implements the pipeline framework described here.
+All analyses were performed through the \href{https://github.com/plger/pipeComp}{pipeComp} R package, which implements the pipeline framework described here.
 scVI results and results of the clustering section were produced with \textit{pipeComp} 0.99.4, while all other results were produced with version 0.99.2 (\textit{biorxiv}) branch of the repository).

### 2020-05-04 main_figures/doublets.Rmd
reduced width

@@
 
-```{r figure1, fig.height=6, fig.width=8}
+```{r figure1, fig.height=6, fig.width=7}
 p1m <- plot_grid(p1, p2, p3, nrow=1, rel_widths = c(9,10,10))
-pdf("doublets.pdf", width=8, height=6)
+pdf("doublets.pdf", width=7, height=6)
 plot_grid(p1m, pt, p4, rel_heights=c(3.5,2,2), labels = LETTERS[1:3], ncol = 1, hjust=0, scale=0.95)

### 2020-05-04 bmc_article.tex

@@
 We evaluate common steps of such analyses, including filtering, doublet detection, normalization, feature selection, denoising, dimensionality reduction and clustering.
-On the basis of these analyses, we make a number of practical recommendations about current analysis choices.
+On the basis of these analyses, we make a number of concrete recommendations about current analysis choices.
 The evaluation framework, \textit{pipeComp}, has been implemented so as to easily integrate any other step or tool, allowing extensible benchmarks and easy application to other fields (\url{https://github.com/plger/pipeComp}).

### 2020-05-04 main_figures/imputation.Rmd
updated imputation figure with new color mapping

@@
 ll2 <- lapply(ll2, FUN=function(x) x[o,])
-ll2s <- lapply(ll2, scale)
+ll2s <- lapply(ll2, pipeComp::colCenterScale)
 

### 2020-05-05 main_figures/filtering.Rmd
reduced dots size and opacity to make overlaps more visible

@@
 
+```{r, include=FALSE}
+# requires old pipeComp version
+devtools::load_all("pipeComp_biorxiv/pipeComp/")
+```
+
 ```{r figure2, fig.width=8, fig.height=8.5}
@@
 
-p2 <- ggplot(m, aes(maxPCout, F1atK, group=method, colour=filter, shape=doubletRemoval)) + geom_point(size=3) +
+p2 <- ggplot(m, aes(maxPCout, F1atK, group=method, colour=filter, shape=doubletRemoval)) + geom_point(size=2.2,alpha=0.8) +
   # geom_line(aes(group=doubletRemoval), colour="lightgrey") + 

### 2020-05-06 main_figures/normalization.Rmd
changed color scale adj R2

@@
   library(pipeComp)
+  library(circlize)
 })
@@
 }
+for(f in rev(colnames(res$evaluation$dimreduction$PC1.covar.adjR2))[1:3]){
+  res$evaluation$dimreduction$PC1.covar.adjR2[[f]] <- 
+    round(res$evaluation$dimreduction$PC1.covar.adjR2[[f]], 3)
+}
 
-p1 <- evalHeatmap(res, step="dimreduction", what=c("log10_total_counts","log10_total_features"), 
-                  what2="PC1.covar.adjR2", agg.by = "norm", reorder_rows = TRUE, 
-                  row_split = classifyNorms(norm), row_title="Variance in PC1 explained by covariate")
-lgd <- Legend(col_fun=colorRamp2(1:100, viridisLite::inferno(100)), at=c(1,50,100), 
-              title="MADs", labels=c("lowest","median","highest"))
-p1 <- grid.grabExpr(draw(renameNorm(p1), merge_legend=TRUE, annotation_legend_list=lgd))
+p1 <- renameNorm(evalHeatmap(res, step="dimreduction", what="log10_total_counts", scale=FALSE, show_heatmap_legend = TRUE, name="adj.R2", what2="PC1.covar.adjR2", agg.by = "norm", reorder_rows = TRUE, col=viridisLite::viridis(100), row_split = classifyNorms(norm), row_title="Variance in PC1 explained by covariate") +
+  evalHeatmap(res, step="dimreduction", what="log10_total_features", scale=FALSE, show_heatmap_legend = FALSE, what2="PC1.covar.adjR2", agg.by = "norm", reorder_rows = TRUE, col=viridisLite::viridis(100)))
+
+p1 <- grid.grabExpr(draw(p1, merge_legend=TRUE))
 
@@
 **C:** Clustering accuracy, measured by the mutual information (MI), the adjusted Rand Index (ARI), and the ARI at the true number of cluster.
-The coloring scheme in panels **A** and **B** is based on the per-metric number of median absolute deviations (MADs) from the per-column medians, and mapped to color using a square-root transformation to minimize the effect of outliers.
+The coloring scheme in panel **C** is based on the per-metric number of median absolute deviations (MADs) from the per-column medians, and mapped to color using a square-root transformation to minimize the effect of outliers.
 The values shown in the cells are the direct metric values (no transformation except rounding).

### 2020-05-06 main_figures/filtering.Rmd
renamed pca filters

@@
 # requires old pipeComp version
-devtools::load_all("pipeComp_biorxiv/pipeComp/")
+devtools::load_all("../../pipeComp_biorxiv/pipeComp/")
 ```
@@
                      c("nofilter","lenient","default","pca","pca2","stringent","veryStringent"))
+levels(m$filter)[4:5] <- c("pca.all","pca.sel")
 cols <- c("black", pipeComp::getQualitativePalette(length(levels(m$filter))-1))

### 2020-05-06 bmc_article.tex

@@
 
-Here, we develop a flexible R framework for pipeline comparison and evaluate the various steps of analysis leading from an initial count matrix to a cluster assignment, which are critical in a wide range of applications.
+Here, we develop a flexible R framework for pipeline comparison and apply it to the evalution of the various steps of analysis leading from an initial count matrix to a cluster assignment, which are critical in a wide range of applications.
 We collected real datasets of known cell composition (Table \ref{tab:table1}) and used a variety of evaluation metrics to investigate in a multilevel fashion the impact of various parameters and variations around a core scRNAseq pipeline.
@@
 In addition, optional benchmark functions can be set for each step to provide standardized, multi-layered evaluation metrics.
-Given such a \texttt{PipelineDefinition} object, a set of alternative parameters (which might include different subroutines) and benchmark datasets, the \texttt{runPipeline} function then proceeds through all combinations of arguments, avoiding recomputing the same step twice and compiling evaluations (including run time) on the fly.
+Given such a \texttt{PipelineDefinition} object, a set of alternative parameters (which might include different subroutines) and benchmark datasets, the \texttt{runPipeline} function then proceeds through all combinations of arguments (or a desired subset), avoiding recomputing the same step twice and compiling evaluations (including run time) on the fly (see Supplementary methods for more detail).
 Variations in a given parameter can be evaluated using all metrics from this point downward in the pipeline.
@@
 
+\subsection*{Application of _pipeComp_ to a different context}
+
+To illustrate the applicability of \textit{pipeComp} to a completely different setting, we benchmarked the impact of surrogate variable analysis (SVA) -based methods on bulk RNAseq differential expression analysis (DEA).
+The rationale behind such approaches is that the identification of factors determining genome-wide variation (i.e.
+surrogate variables) and their inclusion in the generalized linear model for DEA might improve DEA, especially in the presence of sources of variation unrelated to the experimental variables.
+\textit{sva}\citep{leekCapturingHeterogeneityGene2007,leekSvaPackageRemoving2012} and \textit{RUVseq}\citep{rissoNormalizationRNAseq2014a} are two popular packages implementing methods to infer these variables.
+We therefore created a \texttt{PipelineDefinition} with three step: 1) feature filtering, 2) removal of unwanted variation (i.e.
+SVA), and 3) DEA using the most established methods (the package's `pipeComp\_dea' vignette details the creation process).
+We developed 3 benchmark datasets: 1) the \texttt{ipsc} data is a random selection of 10 vs 10 samples from \citet{carcamoOriveAnalysisTranscriptional2017}, which are very heterogeneous, and foldchanges were manually added to 300 genes; 2) the \texttt{seqc} dataset contains 5 vs 5 samples of two mixtures of the SEQC consortium with different ERCC spike-in mixes, and includes a partially correlated batch effect; 3) A **simulation** of 8 vs 8 samples with two technical vectors of variation (see methods for more details).
+As previously reported, we found that more stringent filtering improved DEA (Supplementary Figure 25A) and that the different DEA methods had similar performances (Figure \ref{fig:svadea}A and Supplementary Figure 25A).
+Including a SVA-based step increased sensitivity across all datasets, and particularly so in datasets with technical vectors of variation, and benefited all DEA methods in a similar fashion (Figure \ref{fig:svadea}A,C and Supplementary Figures 25-26).
+With the exception of the \texttt{RUVr} method, which led to a high False Discovery Rate (FDR), all other methods showed comparable gains in sensitivity while maintaining a decent error control (if a little above the nominal FDR), even when increasing the number of surrogate variables used (Figure \ref{fig:svadea}B-C).
+These results therefore suggest that these methods are robust and can be widely applied.
+
 \section*{Discussion}
@@
 
-On the basis of our findings, we can make a number of practical analysis recommendations, also summarized in Figure \ref{fig:figure11}:
+On the basis of our findings, we can make a number of practical analysis recommendations, also summarized in Figure \ref{fig:recommendations}:
 
@@
 \textit{pipeComp} is a flexible R framework for evaluating methods and parameter combinations in a complex pipeline, computing on-the-fly multilevel evaluation metrics.
-Applying this framework to scRNAseq clustering enabled us to make clear recommendations on the steps of filtering, normalization, feature selection, denoising, dimensionality reduction and clustering (Figure \ref{fig:figure11}).
+Applying this framework to scRNAseq clustering enabled us to make clear recommendations on the steps of filtering, normalization, feature selection, denoising, dimensionality reduction and clustering (Figure \ref{fig:recommendations}).
 We demonstrate how a diversity of multilevel metrics can be more robust, more sensitive, and more nuanced than simply evaluating the final clustering performance.
@@
 The \textit{stringent} set of filters uses the same thresholds, but excludes a cell if it is an outlier on any single distribution. 
-
+The \textit{veryStringent} set of filters excludes any cell that is an outlier by at least 2 MADs (lower or higher) on any of: log10\_total\_counts, log10\_total\_features, pct\_counts\_Mt, or pct\_counts\_in\_top\_20\_features.
 The \textit{lenient} set of filters excludes cells that are outliers on at least two distributions by at least 5 MADs, except for pct\_counts\_Mt where the threshold is $>$ 3 MADs and $>$ 0.08.
@@
 The resulting evaluation is available in the Supplementary Figure 13.
+
+\subsection*{SVA-DEA benchmark}
 
@@
     \ifshowfigs
-    \includegraphics[scale=0.75, keepaspectratio]{main_figures/pipeline_explanation.pdf}
+    \includegraphics[width=\textwidth, keepaspectratio]{main_figures/pipeline_explanation.pdf}
     \fi
     \caption{\textbf{Overview of the \textit{pipeComp} framework and its application to a scRNAseq clustering pipeline.
-A:} The package is built around a \texttt{PipelineDefinition} S4 class which defines a set of functions to be executed consecutively, as well as optional evaluation functions for each step.
+A:} The package is built around a \texttt{PipelineDefinition} class which defines a set of functions to be executed consecutively, as well as optional evaluation functions for each step.
 Each step can accept a number of parameters whose alternative values are provided as a list.
-All (or subsets of) combinations of parameters can then be simultaneously ran and evaluated using the \texttt{runPipeline} function.
+All (or subsets of) combinations of parameters can then be simultaneously ran and evaluated using the \texttt{runPipeline} function (see Supplementary Methods for more details).
 \textbf{B:} Scheme representing the application of \textit{pipeComp} to evaluate a range of methods that are commonly used in scRNAseq studies and some of the metrics monitored at various steps.}
@@
     \ifshowfigs
-    \includegraphics[scale=0.85, keepaspectratio]{{main_figures/doublets.pdf}}
+    \includegraphics[width=\textwidth, keepaspectratio]{{main_figures/doublets.pdf}}
     \fi
@@
     \ifshowfigs
-    \includegraphics[scale=0.72, keepaspectratio]{{main_figures/filtering.pdf}}
+    \includegraphics[width=\textwidth, keepaspectratio]{{main_figures/filtering.pdf}}
     \fi
@@
     \ifshowfigs
-    \includegraphics[scale=0.6, keepaspectratio]{{main_figures/normalization.pdf}}
+    \includegraphics[width=\textwidth, keepaspectratio]{{main_figures/normalization.pdf}}
     \fi
@@
     \ifshowfigs
-    \includegraphics[scale=0.65]{{main_figures/selection.pdf}}
+    \includegraphics[width=\textwidth]{{main_figures/selection.pdf}}
     \fi
@@
     \ifshowfigs
-    \includegraphics[scale=0.50, keepaspectratio]{{main_figures/dr.pdf}}
+    \includegraphics[width=\textwidth, keepaspectratio]{{main_figures/dr.pdf}}
     \fi
@@
     \ifshowfigs
-    \includegraphics[scale=0.53, keepaspectratio]{{main_figures/dimensionality.pdf}}
+    \includegraphics[width=\textwidth, keepaspectratio]{{main_figures/dimensionality.pdf}}
     \fi
@@
     \ifshowfigs
-    \includegraphics[scale=0.43, keepaspectratio]{{main_figures/clustering.pdf}}
+    \includegraphics[width=\textwidth, keepaspectratio]{{main_figures/clustering.pdf}}
     \fi
@@
     \ifshowfigs
-    \includegraphics[scale=0.55, keepaspectratio]{{main_figures/imputation.pdf}}
+    \includegraphics[width=\textwidth, keepaspectratio]{{main_figures/imputation.pdf}}
     \fi
@@
     \label{fig:figure10}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \ifshowfigs
+    \includegraphics[width=\textwidth, keepaspectratio]{{main_figures/svadea.pdf}}
+    \fi
+    \caption{\textbf{Evaluation the impact of prior factor analysis on (bulk) RNAseq differential expression analysis using \textit{pipeComp}.
+A:} True positive rates (TPR) and False discovery rates (FDR) of various combinations of SVA-based methods (using a single surrogate variable) and DEA methods.
+\textbf{B:} TPR and FDR of the different SVA-based methods (averaged across the different DEA methods) using increasing numbers of variables (`be' and `leek` refer to the two methods implemented in the \textit{sva} package for estimating the number of surrogate variables).
+In \textbf{A-B}, the x-axis is square-root transformed to improve readability, and three dots for each combination of methods refer to the nominal FDR thresholds of 0.01, 0.05, and 0.1.
+\textbf{C:} Correlation of estimated logFC with expected ones, as well as TPR and FDR (at nominal FDR threshold of 0.05) of the different methods with increasing numbers of variables.}
+    \label{fig:svadea}
 \end{figure}
@@
     \caption{\textbf{Recommendations of tools from filtering to clustering.}}
-    \label{fig:figure11}
+    \label{fig:recommendations}
 \end{figure}
@@
   \hline
-dataset & source & # cells & protocol & description \\ 
+dataset & source & \# cells & protocol & description \\ 
   \hline

### 2020-05-07 supplementary_figures/imputation.Rmd
move some main figures to the supplementary

--- 2020-05-04 main_figures/imputation.Rmd
@@
 
-```{r}
+```{r, include=FALSE}
+if(!exists('FIG_NB')){
+  FIG_NB <- 0
+  getFigNb <- function(increment=FALSE){
+    if(increment) FIG_NB <<- FIG_NB + 1
+    FIG_NB
+  }
+}
+```
+
+# Supplementary Figure `r getFigNb(TRUE)`
+
+```{r, include=FALSE}
 res <- readRDS("../data/results_imputation.rds")
@@
             "ARI at true\nnb clusters"=mycast(ag2, "ARI"),
-            "Mean precision"=mycast(ag, "mean_pr"),
-            "Mean recall"=mycast(ag, "mean_re")
+            "Mean precision"=mycast(ag, "mean_pr")
             )
@@
 p2 <- grid.grabExpr(draw( myh(names(ll2)[1]) + myh(names(ll2)[2]) + 
-                            myh(names(ll2)[3]) + myh(names(ll2)[4]) ))
+                            myh(names(ll2)[3]) ))
 ```
 
-```{r imputation, fig.width=10, fig.height=7}
-pdf("imputation.pdf", width = 10, height=7)
-plot_grid( p1, p2, nrow=2, labels=LETTERS[1:2], rel_heights = c(4.5,6))
-dev.off()
-
+```{r imputation, fig.width=8.5, fig.height=7}
 plot_grid( p1, p2, nrow=2, labels=LETTERS[1:2], rel_heights = c(4,6))
@@
 
-**A:** Average silhouette width per subpopulation across various denoising/imputation methods. **B:** Clustering accuracy across denoising/imputation methods.
+### Supplementary Figure `r getFigNb()`
+**Evaluation of imputation/denoising methods.** Average silhouette width (**A**) and clustering accuracy (**B**) with or without (indicated as _none_) application of a denoising/imputation method.

### 2020-05-07 supplementary_figures/supp2_filtering.Rmd
move some main figures to the supplementary

@@
 
+```{r, fig.height=4, fig.width=7}
+res <- readRDS("../data/filtering_intermediateSummary.rds")
+tmp <- res$filtering$pcCells
+dbl <- 1+grepl("doublet",row.names(tmp))
+row.names(tmp) <- paste0(c("","doublet + ")[dbl],sapply(strsplit(row.names(tmp)," "), FUN=function(x) x[2]))
+cols <- colorRamp2( seq(from=0,to=sqrt(max(tmp)),length.out=100)^2,
+                    colorRampPalette(c("white","red"))(100) )
+draw(Heatmap(tmp, bottom_annotation=pipeComp:::.ds_anno(colnames(tmp)), 
+        show_column_names = FALSE, cluster_columns=FALSE, show_row_dend=FALSE, 
+        row_split=c("No doublet removal","Doublet removal")[dbl], 
+        name="proportion\nexcluded", col=cols, heatmap_legend_param=list(
+          at=c(0,5,10,30,60), labels=paste0(c(0,5,10,30,60),"%"),
+          grid_height=unit(7,"mm")) ),
+     merge_legend=TRUE)
+```
+
+### Supplementary Figure `r getFigNb()`
+
+Proportion of cells filtered out by subpopulation.
+Applying the filters in a cluster-wise fashion (using `scran::quickCluster`) leads to virtually no cell exclusion.
+The color-mapping is square-root transformed.
+
+\newpage
+
+# Supplementary Figure `r getFigNb(TRUE)`
+
 ```{r}

### 2020-05-07 supplementary_figures/supp3_normalization.Rmd
move some main figures to the supplementary

@@
 source("../misc_functions.R")
+
+classifyNorms <- function(x){
+  y <- rep("other", length(x))
+  y[grep("sctransform",x)] <- "sctransform"
+  y[grep("seurat",x)] <- "seurat"
+  y[grep("scran",x)] <- "scran"
+  y[grep("stableG",x)] <- "stableG"
+  y
+}
 ```
@@
 res <- readRDS("../data/results_norm.rds")
-H <- scrna_evalPlot_DR(res, "log10_total_counts", agg.by = "norm", anno_legend=FALSE, covar.type="mean")
-H <- H + scrna_evalPlot_DR(res, "log10_total_features", agg.by = "norm", anno_legend=FALSE, covar.type="mean", reorder_rows = H)
-H
+renameNorm(evalHeatmap(res, step="dimreduction", c("log10_total_counts","log10_total_features"), what2="meanAbsCorr.covariate2", agg.by = "norm", anno_legend=FALSE, row_split=classifyNorms(norm)))
 ```
@@
 silColBreaks <- unique(round(c(x[1],0,x[2]/2,x[2]),1))
-H <- scrna_evalPlot_DR(res, "meanSilWidth", scale=FALSE, agg.by = "method", 
+H <- scrna_evalPlot_silh(res, agg.by = "method", 
                        reorder_rows=TRUE, value_format="", 
@@
 meths <- levels(sil$method)[c(4,2,3,1)]
-H2 <- scrna_evalPlot_clust(res, what="MI", agg.by="method", anno_legend=FALSE, reorder_rows=meths ) + 
-  scrna_evalPlot_clust(res, what="ARI", agg.by="method", anno_legend=FALSE, reorder_rows=meths) + 
-  scrna_evalPlot_clust(res, what="ARI", atTrueK = TRUE, agg.by="method", anno_legend=FALSE, reorder_rows=meths)
+H2 <- evalHeatmap(res, step="clustering", what=c("MI","ARI"), agg.by="method", anno_legend=FALSE, reorder_rows=meths ) + evalHeatmap(res, step="clustering", what="ARI", agg.by="method", anno_legend=FALSE, reorder_rows=meths, filterExpr = n_clus==true.nbClusts, name="ARI at true k", title="ARI at true k" )
+
 ```

### 2020-05-07 supplementary_figures/supp5_dr.Rmd
move some main figures to the supplementary

@@
 ```
+
+# Supplementary Figure `r getFigNb(TRUE)`
+
+```{r dimred, fig.width=10.5, fig.height=6}
+res <- readRDS("../data/results_dimred2.rds")
+H <- renameHrows(
+  scrna_evalPlot_silh( res, reorder_rows=TRUE, 
+                       heatmap_legend_param=list(direction="horizontal")),
+  function(x) gsub("norm\\.","",x))
+p1 <- grid.grabExpr(draw(H, heatmap_legend_side="bottom"))
+
+p2 <- renameHrows(
+  evalHeatmap( res, what=c("MI","min_pr"), agg.by=c("norm","dr"), reorder_rows=TRUE, anno_legend=FALSE,
+               title=c("MI","Minimum precision")) + 
+  evalHeatmap(res, what=c("ARI"), agg.by=c("norm","dr"), filterExpr=n_clus==true.nbClusts, 
+              anno_legend=FALSE, title="ARI at true k"),
+  function(x) gsub("norm\\.","",x)
+)
+p2 <- grid.grabExpr(draw(p2))
+
+plot_grid(p1,p2, nrow=2, labels=LETTERS[1:2])
+```
+
+### Supplementary Figure `r getFigNb()`
+Silhouette width (**A**) per subpopulation and clustering accuracy (**B**) according to various normalization and dimension reductions.
+**Evaluation of common dimensionality reduction methods.
+A:** Minimum (left) and average (right) silhouette width per subpopulation resulting from combinations of normalization and dimension reductions.
+**B:** Clustering accuracy, measured by mutual information (MI), minimum subpopulation precision, and adjusted Rand index (ARI) at the true number of clusters.
+
+\newpage
 

### 2020-05-07 supplementary_figures/supp_figures.Rmd
updated many figures to improve readability...
move some main figures to the supplementary

@@
 ---
-title: \Large{pipeComp, a general framework for the evaluation of computational pipelines, reveals performant single-cell RNA-seq preprocessing tools}
-subtitle: \vspace{0.2cm}\LARGE{Supplementary Figures}\vspace{0.3cm}
+title: \vspace{2cm}\Large{pipeComp, a general framework for the evaluation of computational pipelines, reveals performant single-cell RNA-seq preprocessing tools}
+subtitle: \vspace{2cm}\LARGE{Supplementary Figures}\vspace{2cm}
+date: "`r format(Sys.time(), '%d %B, %Y')`"
 author:
@@
   pdf_document:
-    toc: true
     fig_width: 8
@@
 ```
+
+# Supplementary Figure `r getFigNb(TRUE)`
+
+```{r, fig.align="center", echo=FALSE, out.height="0.9\\textheight"}
+knitr::include_graphics("dataset_description.pdf")
+```
+
+### Supplementary Figure `r getFigNb()`: Overview of the benchmark datasets
 
@@
 ```
+
+```{r child='supp_svadea.Rmd'}
+```
+
+# Supplementary Figure `r getFigNb(TRUE)`
+
+```{r, fig.align="center", echo=FALSE, out.height="0.8\\textheight"}
+knitr::include_graphics("../main_figures/pipeComp_summary.pdf")
+```
+
+### Supplementary Figure `r getFigNb()`: Summary of the recommendations.

### 2020-05-07 supplementary_figures/supp_svadea.Rmd
move some main figures to the supplementary

@@
 
-```{r, fig.width=8, fig.height=9.5}
+```{r, fig.width=8.5, fig.height=8.5}
+plot_grid(
+  dea_evalPlot_curve(res1, agg.by=c("dea.method","filt","minCount"), shapeBy = "dea.method", colourBy = c("filter")) + ggtitle("Using 1 SV"),
+  dea_evalPlot_curve(res1, agg.by=c("sva.method","filt","minCount"), shapeBy = "sva.method", colourBy = c("filter")) + ggtitle("Using 1 SV"),
+  dea_evalPlot_curve(res2, agg.by=c("sva.method")) + ggtitle("Using 2 SVs"),
+  nrow=3, labels=LETTERS[1:3]
+)
+```
+
+### Supplementary Figure `r getFigNb()`
+Accuracy of the differential expression analysis across combinations of: **A:** filters and DEA methods (using max 1 variable), **B:** filters and SVA methods (using max 1 variable), **C:** SVA methods (using 2 variables).
+In **C**, filled points indicate a nominal FDR below the observed FDR.
+
+\newpage
+
+# Supplementary Figure `r getFigNb(TRUE)`
+
+```{r, fig.width=8, fig.height=8.5}
 plot_grid(
@@
 \newpage
-
-# Supplementary Figure `r getFigNb(TRUE)`
-
-```{r, fig.width=8.5, fig.height=8.5}
-plot_grid(
-  dea_evalPlot_curve(res1, agg.by=c("dea.method","filt","minCount"), shapeBy = "dea.method", colourBy = c("filter")) + ggtitle("Using 1 SV"),
-  dea_evalPlot_curve(res1, agg.by=c("sva.method","filt","minCount"), shapeBy = "sva.method", colourBy = c("filter")) + ggtitle("Using 1 SV"),
-  dea_evalPlot_curve(res2, agg.by=c("sva.method")) + ggtitle("Using 2 SVs"),
-  nrow=3, labels=LETTERS[1:3]
-)
-```
-
-### Supplementary Figure `r getFigNb()`
-Accuracy of the differential expression analysis across combinations of: **A:** filters and DEA methods (using max 1 variable), **B:** filters and SVA methods (using max 1 variable), **C:** SVA methods (using 2 variables).
-In **C**, filled points indicate a nominal FDR below the observed FDR.
-
-\newpage

### 2020-05-07 bmc_article.tex

@@
 
-The \textit{pipeComp} package defines a pipeline as, minimally, a list of functions executed consecutively on the output of the previous one (Figure \ref{fig:figure2}A).
+The \textit{pipeComp} package defines a pipeline as, minimally, a list of functions executed consecutively on the output of the previous one (Figure \ref{fig:explanation}A).
 In addition, optional benchmark functions can be set for each step to provide standardized, multi-layered evaluation metrics.
-Given such a \texttt{PipelineDefinition} object, a set of alternative parameters (which might include different subroutines) and benchmark datasets, the \texttt{runPipeline} function then proceeds through all combinations of arguments (or a desired subset), avoiding recomputing the same step twice and compiling evaluations (including run time) on the fly (see Supplementary methods for more detail).
+Given such a \texttt{PipelineDefinition} object, a set of alternative parameters (which might include different subroutines) and benchmark datasets, the \texttt{runPipeline} function then proceeds through all combinations of arguments (or a desired subset), avoiding recomputing the same step twice without the need to save all alternative intermediates, and compiling evaluations (including run time) on the fly (see Supplementary methods for more detail).
 Variations in a given parameter can be evaluated using all metrics from this point downward in the pipeline.
@@
 We therefore monitored several complementary metrics across multiple steps of the process.
-We proceeded in a step-wise fashion, first testing a large variety of parameters at the early steps of the pipeline along with only a set of mainstream options downstream, then selecting main alternatives and proceeding to a more detailed benchmark of the next step (Figure \ref{fig:figure2}B).
+We proceeded in a step-wise fashion, first testing a large variety of parameters at the early steps of the pipeline along with only a set of mainstream options downstream, then selecting main alternatives and proceeding to a more detailed benchmark of the next step (Figure \ref{fig:explanation}B).
 
@@
 
-While most methods accurately identified the doublets in the 3 cell lines dataset (mixology10x3cl), the other two datasets proved more difficult (Figure \ref{fig:figure3}A).
-\textit{scDblFinder} achieved comparable or better accuracy than top alternatives while being the fastest method (Figure \ref{fig:figure3}B).
-Across datasets, cells called as doublets tended to be classified in the wrong cluster more often than other cells (Figure \ref{fig:figure3}C).
-We also found that \textit{scDblFinder} improved the accuracy of the clustering across all benchmark datasets even when, by design, the data contained no heterotypic doublet (Figure \ref{fig:figure4}).
+While most methods accurately identified the doublets in the 3 cell lines dataset (mixology10x3cl), the other two datasets proved more difficult (Figure \ref{fig:doublets}A).
+\textit{scDblFinder} achieved comparable or better accuracy than top alternatives while being the fastest method (Figure \ref{fig:doublets}B).
+Across datasets, cells called as doublets tended to be classified in the wrong cluster more often than other cells (Figure \ref{fig:doublets}C).
+We also found that \textit{scDblFinder} improved the accuracy of the clustering across all benchmark datasets even when, by design, the data contained no heterotypic doublet (Figure \ref{fig:filtering}).
 
@@
 
-We next investigated the impact of filtering according to six methods with varying stringency; three methods based on deviations to MADs with increasing levels of stringency (named \textit{lenient}, \textit{default}, \textit{stringent}, \textit{very stringent}) and two methods based on \textit{scater}'s \textit{runPCA} using all (\textit{pca}) or a restricted number of covariates (\textit{pca2}, see Methods for more details).
+We next investigated the impact of filtering according to six methods with varying stringency; three methods based on deviations to MADs with increasing levels of stringency (named \textit{lenient}, \textit{default}, \textit{stringent}, \textit{very stringent}) and two methods based on \textit{scater}'s \textit{runPCA} using all (\textit{pca.all}) or selected covariates (\textit{pca.sel}, see Methods for more details).
 An important risk of excluding cells on the basis of their distance from the whole distribution of cells on some properties (e.g., library size) is that these properties tend to have different distributions across subpopulations.
-As a result, thresholds in terms of number of MADs from the whole distribution can lead to strong biases against certain subpopulations (Figure \ref{fig:figure4}A).
-We therefore examined the tradeoff between the increased accuracy of filtering and the maximum proportion of cells excluded per subpopulation (Figure \ref{fig:figure4}B and Supplementary Figure 10).
+As a result, thresholds in terms of number of MADs from the whole distribution can lead to strong biases against certain subpopulations (Figure \ref{fig:filtering}A).
+We therefore examined the tradeoff between the increased accuracy of filtering and the maximum proportion of cells excluded per subpopulation (Figure \ref{fig:filtering}B and Supplementary Figure 10).
 Since filtering changes the relative abundance of the different subpopulations, global clustering accuracy metrics such as ARI are not appropriate here.
@@
 A first observation was that, although the stringent filtering tended to be associated with an increase in accuracy, it could also become deleterious.
-Most of the benefits could be achieved without very stringent filtering and minimizing subpopulation bias (Figure \ref{fig:figure4}B).
+Most of the benefits could be achieved without very stringent filtering and minimizing subpopulation bias (Figure \ref{fig:filtering}B).
 Applying the same filtering criteria on individual clusters of cells (identified through scran's \texttt{quickCluster} method) rather than on whole dataset resulted in nearly no cells being filtered out.
-This suggests that filtering on the global population tends to discard cells of subpopulations with more extreme properties (e.g.
-high library size), rather than low-quality cells, and that more stringent filtering may only remove the cells of these subpopulations.
-On the opposite, lenient or lack of filtering leads to no/ little impact on subpopulations.
-This comes at the cost that lenient filtering performs worse or equally well than other methods among all datasets.
-However, a mild filtering ('default') often leads to higher clustering performance (e.g.
-simMix and Zhengmix4 datasets) or to very little loss (e.g.
-Koh and Zhengmix8eq) and always retain most of the cells in each subpopulation.
+This suggests that stringent filtering on the global population tends to discard cells of subpopulations with more extreme properties (e.g.
+high library size), rather than low-quality cells.
+On the opposite, too lenient filtering (or lack thereof) can lead to misclassified cells, as was for instance the case especially with the Zhengmix4 datsets (Figure \ref{fig:filtering}B).
+We found that a relatively mild filtering (`default' - see methods) often leads to high clustering performance (e.g.
+simMix and Zhengmix4 datasets) and always retain most of the cells in each subpopulation (e.g.
+Koh and Zhengmix8eq).
 This suggests that a mild ('default') filtering is a very good trade-off between clustering accuracy and subpopulations preservation.
-Finally, the doublet removal step in conjunction with filtering  resulted in a net decrease in the proportion of excluded cells while retaining or improving accuracy among some 10x datasets (mixology, Zheng datasets).
+Finally, the doublet removal step in conjunction with filtering  resulted in a net decrease in the proportion of excluded cells while retaining or improving accuracy not only in datasets that do include heterotypic doublets (mixology datasets), but also some others in which such doublets are unlikely (e.g.
+Zheng datasets).
 On the Smart-Seq and simulated datasets however, the removal of doublets had little to no impact on any of the two measures.
-Overall, we therefore recommend the use of doublet removal followed by a `default' filtering (see Methods) or a similar approach.
+Overall, we therefore recommend the use of doublet removal followed by our `default' filtering (see Methods) or a similar approach.
 
@@
 However, as library size tends to vary across subpopulations, part of this effect can simply reflect biological differences.
-We therefore assessed to what extent the first principal component still retained a correlation with the library size and the number of detected features, removing the confounding covariation with the subpopulations (Figure \ref{fig:figure5}A and Supplementary Figure 12).
+We therefore assessed to what extent the first principal component still retained a correlation with the library size and the number of detected features, removing the confounding covariation with the subpopulations (Figure \ref{fig:norm}A and Supplementary Figure 12).
 Scaling tended to remove much of the correlation with these features, and while most methods were able to remove most of the effect, the correlation was lowest with \textit{SCTransform}.
@@
 
-We further evaluated normalization methods by investigating their impact on the separability of the subpopulations (Figure \ref{fig:figure5}B-C).
+We further evaluated normalization methods by investigating their impact on the separability of the subpopulations (Figure \ref{fig:norm}B-C).
 Since clustering accuracy metrics such as the ARI are very strongly influenced by the number of clusters, we complemented it with silhouette width \cite{RousseeuwSil1987} and mutual information (MI).
@@
 In view of the subpopulation silhouette and ARI at true number of clusters, it seems that all methods perform worse on the three Zhengmix datasets, that are the datasets with the highest number of cells and the smaller feature detection rate.
-Scaling tended to reduce the average silhouette width of some subpopulations and to increase that of some less distinguishable ones and was generally, but not always, beneficial on the accuracy of the final clustering.
+Scaling (e.g.
+`seurat (no scale)' vs `seurat') tended to reduce the average silhouette width of some subpopulations and to increase that of some less distinguishable ones and was generally, but not always, beneficial on the accuracy of the final clustering.
 Regressing out covariates systematically gave equal  or poorer performance on MI, ARI and ARI at true number of clusters, with a notable negative impact when \textit{Seurat}'s regressing was applied to the Koh and Kumar datasets.
@@
 The standardized measures of variability were systematically worse than their non-standardized counterparts in selecting genes with a high proportion of variance explained by subpopulation.
-Regarding the percentage of deviance explained however, the standardized measures were often superior (Figure \ref{fig:figure6}A and Supplementary Figures 18-19).
+Regarding the percentage of deviance explained however, the standardized measures were often superior (Figure \ref{fig:selection}A and Supplementary Figures 18-19).
 \textit{Deviance} proved the method of choice to prioritize genes with a high variance explained by subpopulations (with mere expression level proving surprisingly performant) but did not perform so well to select genes with a high deviance explained.
 
-We next evaluated how the use of different feature selection methods affected the clustering accuracy (Figure \ref{fig:figure6}B).
+We next evaluated how the use of different feature selection methods affected the clustering accuracy (Figure \ref{fig:selection}B).
 To validate the previous assay on the proportion of variance/deviance explained by real populations, we included genes that maximized these two latter measures.
 Interestingly, while these selections were on average the top-ranking methods, they were not systematically best for all datasets.
-The ARI at true number of clusters was near perfect (>0.94) among the smallest datasets but among bigger datasets (simMix and ZhengMix) the performance of the methods was lower and more variable (Figure \ref{fig:figure6}B).
+The ARI at true number of clusters was near perfect (>0.94) among the smallest datasets but among bigger datasets (simMix and ZhengMix) the performance of the methods was lower and more variable (Figure \ref{fig:selection}B).
 Non-standardized measures of variability, including mere expression level, tended to outperform more complex metrics.
@@
 We compared the various estimates of dimensionality in their ability to retrieve the intrinsic number of dimensions in a dataset, based on Seurat's weighted PCA space.
-As a first approximation of the true dimensionality, we computed the variance in each principal component that was explained by the subpopulations, which sharply decreased after the first few components in most datasets (Figure \ref{fig:figure8}A).
-Figure \ref{fig:figure8}B shows the difference between the dimension estimates of the above methods and that based on the subpopulations (i.e.
-from Figure \ref{fig:figure8}A).
-Of note, the methods differ widely in view of their computing time (Figure \ref{fig:figure8}B) and we saw no relationship between the accuracy of the estimates and the complexity of the method.
-Reasoning that over-estimating the number of dimensions is less problematic than under-estimating it, we kept the former methods for a full analysis of their impact on clustering (Figure \ref{fig:figure8}C-D), when combined with \textit{SCTransform} or the standard \textit{Seurat} normalization.
-Although most methods performed well on the various clustering measures, the global maximum likelihood based on translated Poisson Mixture Model (\textit{maxLikGlobal}, using the 20 nearest neighbors, although estimates were relatively stable across decent values of \textit{k}; see Supplementary Figure 23) provided the dimensionality estimate that best separated the subpopulations (Figure \ref{fig:figure8}C) and resulted in the best clustering accuracy (Figure \ref{fig:figure8}D).
+As a first approximation of the true dimensionality, we computed the variance in each principal component that was explained by the subpopulations, which sharply decreased after the first few components in most datasets (Figure \ref{fig:dim}A).
+Figure \ref{fig:dim}B shows the difference between the dimension estimates of the above methods and that based on the subpopulations (i.e.
+from Figure \ref{fig:dim}A).
+Of note, the methods differ widely in view of their computing time (Figure \ref{fig:dim}B) and we saw no relationship between the accuracy of the estimates and the complexity of the method.
+Reasoning that over-estimating the number of dimensions is less problematic than under-estimating it, we kept the former methods for a full analysis of their impact on clustering (Figure \ref{fig:dim}C-D), when combined with \textit{SCTransform} or the standard \textit{Seurat} normalization.
+Although most methods performed well on the various clustering measures, the global maximum likelihood based on translated Poisson Mixture Model (\textit{maxLikGlobal}, using the 20 nearest neighbors, although estimates were relatively stable across decent values of \textit{k}; see Supplementary Figure 23) provided the dimensionality estimate that best separated the subpopulations (Figure \ref{fig:dim}C) and resulted in the best clustering accuracy (Figure \ref{fig:dim}D).
 This method systematically selected many more components than were associated with the subpopulations, suggesting that although these additional components appear individually uninformative, in combination they nevertheless contribute to classification.
@@
 
-Since ARI is dominated by differences in the number of clusters (Supplementary Figures 1-2) and no single metric is perfect, we diversified them (Figure \ref{fig:figure9}).
+Since ARI is dominated by differences in the number of clusters (Supplementary Figures 1-2) and no single metric is perfect, we diversified them (Figure \ref{fig:clust}).
 MI has the virtue of not decreasing when a true subpopulation is split into two clusters, which is arguably less problematic (and might well reflect unknown biological subgroups), but as a consequence it can be biased towards methods producing higher resolution clustering (Supplementary Figure 2).
@@
 
-The MI score and minimum precision, which are largely independent of the estimated number of clusters, were overall higher for the \textit{walktrap} method (Figure \ref{fig:figure9}), while the mean F1 score favored both \textit{scran} methods (\textit{walktrap} and \textit{fast greedy}) over \textit{Seurat}.
+The MI score and minimum precision, which are largely independent of the estimated number of clusters, were overall higher for the \textit{walktrap} method (Figure \ref{fig:clust}), while the mean F1 score favored both \textit{scran} methods (\textit{walktrap} and \textit{fast greedy}) over \textit{Seurat}.
 Finally, the ARI score at the true number of clusters, when available, showed similar performances.
@@
 Overall, the \textit{walktrap} method appeared superior to the \textit{fast greedy} algorithm and was generally less prone to misclassification than \textit{Seurat} clustering, although the latter offered more control over the resolution.
-A major difference between \textit{walktrap}-based clustering and \textit{Seurat} is the computing time (Figure \ref{fig:figure9}B).
+A major difference between \textit{walktrap}-based clustering and \textit{Seurat} is the computing time (Figure \ref{fig:clust}B).
 We found that using the Annoy approximation to the nearest neighbors somewhat reduced computing time at no apparent cost to accuracy.
@@
 SVA), and 3) DEA using the most established methods (the package's `pipeComp\_dea' vignette details the creation process).
-We developed 3 benchmark datasets: 1) the \texttt{ipsc} data is a random selection of 10 vs 10 samples from \citet{carcamoOriveAnalysisTranscriptional2017}, which are very heterogeneous, and foldchanges were manually added to 300 genes; 2) the \texttt{seqc} dataset contains 5 vs 5 samples of two mixtures of the SEQC consortium with different ERCC spike-in mixes, and includes a partially correlated batch effect; 3) A **simulation** of 8 vs 8 samples with two technical vectors of variation (see methods for more details).
+We developed 3 benchmark datasets: 1) the \texttt{ipsc} data is a random selection of 10 vs 10 samples from \citet{carcamoOriveAnalysisTranscriptional2017}, which are very heterogeneous, and foldchanges were manually added to 300 genes; 2) the \texttt{seqc} dataset contains 5 vs 5 samples of two mixtures of the SEQC consortium with different ERCC spike-in mixes \citep{ComprehensiveAssessment2014}, and includes a partially correlated batch effect; 3) A **simulation** of 8 vs 8 samples with two technical vectors of variation (see methods for more details).
 As previously reported, we found that more stringent filtering improved DEA (Supplementary Figure 25A) and that the different DEA methods had similar performances (Figure \ref{fig:svadea}A and Supplementary Figure 25A).
@@
 
+The full process of creating the SVA-DEA \texttt{PipelineDefinition} is described step by step in the \href{https://github.com/plger/pipeComp/blob/master/vignettes/pipeComp_dea.Rmd}{pipeComp\_dea vignette}.
+
+The three benchmark datasets are available in \url{https://github.com/markrobinsonuzh/scRNA\_pipelines\_paper/blob/master/svadea/datasets/}, along with all code used to generate and prepare them.
+Briefly:
+\begin{itemize}
+    \item For the \textit{ipsc} dataset, we took the quantification of the GSE79636 dataset (fairly heterogeneous) from the \textit{RNAontheBENCH} package \citep{germainRNAontheBENCHComputationalEmpirical2016}, and selected two random sets of 10 samples.
+No further variation was added, and log-foldchanges ranging from 0.25 to 1.25 were added to the counts of 300 random genes.
+All other genes were assumed to be negatives.
+    \item For the \textit{seqc} dataset, we used data from the \textit{seqc} package \citep{ComprehensiveAssessment2014}.
+For both groups (mixtures C and D, which also contain two different spike-in mixes), we selected samples from the \texttt{ILM\_refseq\_gene\_AGR} and \texttt{ILM\_refseq\_gene\_CNL} batches, which have clear technical differences; we selected 5 samples per group with a weak correlation with the batch (3:2 vs 2:3).
+Since the true differences between mixtures are not entirely known, the analysis was performed on all genes but only the spike-ins were considered for benchmarking.
+    \item For the \textit{simulation}, we simulated counts using the \textit{polyester} package \citep{frazeePolyesterSimulating2015} and the means/dispersions from the GSE79636 dataset (restricted to a single batch and biological group), with 500 DEGs (log-foldchanges ranging from 0.5 to 2) and two batch effects: 1) a technical batch partially correlated with the group (6:2) and affecting 1/3 of the genes, and 2) a linear vector of variation uncorrelated with the groups and affecting 1/3 of the genes.
+\end{itemize}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
     \ifshowfigs
-    \includegraphics[width=\textwidth,keepaspectratio]{dataset_description}
-    \fi
-    \caption{\textbf{Overview of the benchmark datasets used.}}
-    \label{fig:figure1}
-\end{figure}
-
-\begin{figure}
-    \centering
-    \ifshowfigs
     \includegraphics[width=\textwidth, keepaspectratio]{main_figures/pipeline_explanation.pdf}
@@
 \textbf{B:} Scheme representing the application of \textit{pipeComp} to evaluate a range of methods that are commonly used in scRNAseq studies and some of the metrics monitored at various steps.}
-    \label{fig:figure2}
+    \label{fig:explanation}
 \end{figure}
@@
 Even in datasets which should not have neotypic doublets, the cells identified as such tend to be misclassified.
-\label{fig:figure3}}
+\label{fig:doublets}}
 \end{figure}
@@
 The PCA methods refer to multivariate outlier detected as implemented in \textit{scater} (see Methods for details).}
-    \label{fig:figure4}
+    \label{fig:filtering}
 \end{figure}
@@
 Grey squares indicate that the true number of clusters could not be reached with the corresponding method and dataset.
-`feat\_regress', `mt\_regress' and `feat\_mt\_regress' respectively stand for the regressing out of the number of features, the proportion of mitochondrial reads, or both during scaling.}
-    \label{fig:figure5}
+`feat\_regress', `mt\_regress' and `feat\_mt\_regress' respectively stand for the regressing out of the number of features, the proportion of mitochondrial reads, or both during scaling.
+Throughout this paper, silhouette plots (panel \textbf{B}) square-root transform values for color mapping.
+Other evaluation metric heatmaps (panels \textbf{A,C}) map colors linearly to the number of (matrix-wise) Median Absolute Deviations from the (column-wise) median, while the printed numbers represent the raw metric values.}
+    \label{fig:norm}
 \end{figure}
@@
 The  \textit{vst.varExp} and \textit{devianceExplained} methods correspond to the estimates used in \textbf{A} to evaluate the selection methods and were included here only for validation purpose.}
-    \label{fig:figure6}
-\end{figure}
-
-\begin{figure}
-    \centering
-    \ifshowfigs
-    \includegraphics[width=\textwidth, keepaspectratio]{{main_figures/dr.pdf}}
-    \fi
-    \caption{\textbf{Evaluation of common dimensionality reduction methods.
-A:} Average silhouette width per subpopulation resulting from combinations of normalization and dimension reductions.
-\textbf{B:} Clustering accuracy, mean proportion of the variance in the first 5 components explained by real subpopulations (center), and median ARI of the resulting clustering (right).}
-    \label{fig:figure7}
+    \label{fig:selection}
 \end{figure}
@@
 \textbf{C:} Average silhouette width per subpopulation across a selection of methods, combined with \textit{SCTransform} or standard \textit{Seurat} normalization.
-\textbf{D:} Clustering accuracy across the normalization/dimensionality estimation methods.}
-    \label{fig:figure8}
+\textbf{D:} Clustering accuracy across the normalization/dimensionality estimation methods.
+The color-mapping scheme is the same as in Figure \ref{fig:norm}}
+    \label{fig:dim}
 \end{figure}
@@
     \fi
-    \caption{\textbf{Evaluation of clustering methods.} Clustering accuracy (\textbf{A}) and run time (\textbf{B}) of \textit{scran}- and \textit{Seurat}-based clustering tools in combination with \textit{SCTransform} and standard \textit{Seurat}'s normalization.}
-    \label{fig:figure9}
-\end{figure}
-
-\begin{figure}
-    \centering
-    \ifshowfigs
-    \includegraphics[width=\textwidth, keepaspectratio]{{main_figures/imputation.pdf}}
-    \fi
-    \caption{\textbf{Evaluation of imputation/denoising methods} Average silhouette width per subpopulation (\textbf{A}) and clustering accuracy (\textbf{B}) with or without (indicated as \textit{none}) application of a denoising/imputation method.}
-    \label{fig:figure10}
+    \caption{\textbf{Evaluation of clustering methods.} Clustering accuracy (\textbf{A}) and run time (\textbf{B}) of \textit{scran}- and \textit{Seurat}-based clustering tools in combination with \textit{SCTransform} and standard \textit{Seurat}'s normalization.
+The color-mapping scheme is the same as in Figure \ref{fig:norm}}
+    \label{fig:clust}
 \end{figure}
@@
 In \textbf{A-B}, the x-axis is square-root transformed to improve readability, and three dots for each combination of methods refer to the nominal FDR thresholds of 0.01, 0.05, and 0.1.
-\textbf{C:} Correlation of estimated logFC with expected ones, as well as TPR and FDR (at nominal FDR threshold of 0.05) of the different methods with increasing numbers of variables.}
+\textbf{C:} Correlation of estimated logFC with expected ones, as well as TPR and FDR (at nominal FDR threshold of 0.05) of the different methods with increasing numbers of variables.
+As in \ref{fig:norm}C, the color mapping tracks the number of (matrix-wise) Median Absolute Deviations from the (column-wise) median, while the printed numbers represent the raw metric values.}
     \label{fig:svadea}
-\end{figure}
-
-\begin{figure}[h]
-    \centering
-    \ifshowfigs
-    \includegraphics[scale=0.9,keepaspectratio]{main_figures/pipeComp_summary.pdf}
-    \fi
-    \caption{\textbf{Recommendations of tools from filtering to clustering.}}
-    \label{fig:recommendations}
 \end{figure}
@@
 \section*{Additional Files}
-  \subsection*{\textbf{Additional file 1:Supplementary figures 1-24}}
+  \subsection*{\textbf{Additional file 1: Supplementary figures}}
   Additional results of evaluations across datasets among various metrics. (PDF 15 MB)
-
+  \subsection*{\textbf{Additional file 2: Supplementary methods}}
+  More detailed description of \textit{pipeComp} and comparison with other tools. (PDF 0.5 MB)
 \end{backmatter}

### 2020-05-07 supplementary_figures/supp3_normalization.Rmd
added missing supp figures

@@
 ggplot(ag, aes(norm, elapsed)) + geom_boxplot() + geom_point(aes(colour=dataset)) + 
-  coord_flip() + scale_colour_manual(values=cols) + ylab("") + xlab("Time (s)")
+  coord_flip() + scale_colour_manual(values=cols) + xlab("") + ylab("Time (s)")
 ```

### 2020-05-07 supplementary_figures/supp5_dr.Rmd
added missing supp figures

@@
 p2 <- renameHrows(
-  evalHeatmap( res, what=c("MI","min_pr"), agg.by=c("norm","dr"), reorder_rows=TRUE, anno_legend=FALSE,
-               title=c("MI","Minimum precision")) + 
-  evalHeatmap(res, what=c("ARI"), agg.by=c("norm","dr"), filterExpr=n_clus==true.nbClusts, 
-              anno_legend=FALSE, title="ARI at true k"),
+  evalHeatmap( res, step="clustering", what=c("MI","min_pr"), agg.by=c("norm","dr"), 
+               reorder_rows=TRUE, anno_legend=FALSE, title=c("MI","Minimum precision")) + 
+  evalHeatmap(res, step="clustering", what=c("ARI"), agg.by=c("norm","dr"), 
+              filterExpr=n_clus==true.nbClusts, anno_legend=FALSE, title="ARI at true k"),
   function(x) gsub("norm\\.","",x)
@@
 ### Supplementary Figure `r getFigNb()`
-Silhouette width (**A**) per subpopulation and clustering accuracy (**B**) according to various normalization and dimension reductions.
 **Evaluation of common dimensionality reduction methods.

### 2020-05-07 bmc_article.tex
Updates from Overleaf

@@
 Although we use some datasets based on other protocols, our focus is especially on droplet-based datasets that do not include exogenous control RNA (i.e.
-spike-ins); see Table \ref{tab:table1} and Figure \ref{fig:figure1} for more details.
+spike-ins); see Table \ref{tab:table1} and Supplementary Figure 1 for more details.
 In addition to previously-used benchmark datasets with true cell labels \cite{duoClustering2018,tianMixology2018}, we simulated two datasets with a hierarchical subpopulation structure based on real 10x human and mouse data using \textit{muscat} \cite{CrowellMuscat2019}.
@@
 
-The \textit{pipeComp} package defines a pipeline as, minimally, a list of functions executed consecutively on the output of the previous one (Figure \ref{fig:explanation}A).
+The \textit{pipeComp} package defines a pipeline as, minimally, a list of functions executed consecutively on the output of the previous one (Figure \ref{fig:explanation}A; see also Supplementary methods).
 In addition, optional benchmark functions can be set for each step to provide standardized, multi-layered evaluation metrics.
@@
 This is especially important because end-point metrics, such as the adjusted Rand index (ARI) \cite{HubertARI1985} for clustering, are not perfect.
-For example, although the meaning of an ARI score is independent of the number of true subpopulations \cite{steinleyProperties2004}, the number of clusters called is by far the most important determinant of the score: the farther it is from the actual number of subpopulations, the worse the ARI (Supplementary Figures 1-2).
+For example, although the meaning of an ARI score is independent of the number of true subpopulations \cite{steinleyProperties2004}, the number of clusters called is by far the most important determinant of the score: the farther it is from the actual number of subpopulations, the worse the ARI (Supplementary Figures 2-3).
 In this context, one strategy has been to cluster across various resolutions and only consider the results that have the right number of clusters \cite{duoClustering2018}.
@@
 While doublets of the same cell type are relatively innocuous for most downstream applications due to their conservation of the relative expression between genes, doublets formed from different cell types or states are likely to be misclassified and could potentially distort downstream analysis.
-In some cases, doublets can be identified through their unusually high number of reads and detected features, but this is not always the case (Supplementary Figure 3).
+In some cases, doublets can be identified through their unusually high number of reads and detected features, but this is not always the case (Supplementary Figure 4).
 A number of methods were developed to identify doublets, in most cases by comparing each cell to artificially-created doublets \cite{mcginnisDoubletfinder2019, LunScran2016, BaisScds2019}.
@@
 This can for instance be performed through the \texttt{isOutlier} function from \textit{scater} that measures, for a given control property, the number of median absolute deviations (MADs) of each cell from the median of all cells.
-Supplementary Figure 4 shows the distributions of some of the typical cell properties commonly used.
+Supplementary Figure 5 shows the distributions of some of the typical cell properties commonly used.
 Of note, these properties tend to be correlated, with some exceptions.
-For example, while a high proportion of mitochondrial reads is often correlated with a high proportion of the counts in the top features, there can also be other reasons for an over-representation of highly-expressed features (Supplementary Figure 5), such as an over-amplification in non-UMI datasets.
-In our experience, 10X datasets also exhibit a very strong correlation between the total counts and the total features even across very different cell types (Supplementary Figure 6).
+For example, while a high proportion of mitochondrial reads is often correlated with a high proportion of the counts in the top features, there can also be other reasons for an over-representation of highly-expressed features (Supplementary Figure 6), such as an over-amplification in non-UMI datasets.
+In our experience, 10X datasets also exhibit a very strong correlation between the total counts and the total features even across very different cell types (Supplementary Figure 7).
 We therefore also measure the ratio between the two and treat cells strongly departing from this trend with suspicion.
 
-Reasoning that the cells we wish to exclude are the cells that would be misclassified, we measured the rate of misclassification of each cell in each dataset across a variety of clustering pipelines, correcting for the median misclassification rate of the subpopulation and then evaluated what properties could be predictive of misclassification (Supplementary Figures 7-9).
+Reasoning that the cells we wish to exclude are the cells that would be misclassified, we measured the rate of misclassification of each cell in each dataset across a variety of clustering pipelines, correcting for the median misclassification rate of the subpopulation and then evaluated what properties could be predictive of misclassification (Supplementary Figures 8-10).
 We could not identify any property or simple combination thereof that would be consistently predictive of misclassification; the only feature that consistently stood out across multiple datasets (the Zheng datasets) was that cells with very high read counts have a higher chance of being misclassified.
@@
 As a result, thresholds in terms of number of MADs from the whole distribution can lead to strong biases against certain subpopulations (Figure \ref{fig:filtering}A).
-We therefore examined the tradeoff between the increased accuracy of filtering and the maximum proportion of cells excluded per subpopulation (Figure \ref{fig:filtering}B and Supplementary Figure 10).
+We therefore examined the tradeoff between the increased accuracy of filtering and the maximum proportion of cells excluded per subpopulation (Figure \ref{fig:filtering}B and Supplementary Figure 11).
 Since filtering changes the relative abundance of the different subpopulations, global clustering accuracy metrics such as ARI are not appropriate here.
@@
 Most of the benefits could be achieved without very stringent filtering and minimizing subpopulation bias (Figure \ref{fig:filtering}B).
-Applying the same filtering criteria on individual clusters of cells (identified through scran's \texttt{quickCluster} method) rather than on whole dataset resulted in nearly no cells being filtered out.
+Applying the same filtering criteria on individual clusters of cells (identified through scran's \texttt{quickCluster} method) rather than on whole dataset resulted in nearly no cells being filtered out (Supplementary Figure 12).
 This suggests that stringent filtering on the global population tends to discard cells of subpopulations with more extreme properties (e.g.
@@
 Mitochondrial reads have been associated with cell degradation and there is evidence that ribosomal genes can influence the clustering output, hiding other biological structure in the analysis \cite{freytagComparison2018}.
-We therefore investigated whether excluding one category of features or the other, or using only protein-coding genes, had an impact on the ability to distinguish subpopulations (Supplementary Figure 11).
+We therefore investigated whether excluding one category of features or the other, or using only protein-coding genes, had an impact on the ability to distinguish subpopulations (Supplementary Figure 13).
 Removal of ribosomal genes robustly reduced the quality of the clustering, suggesting that they represent real biological differences between subpopulations.
@@
 It is possible that the difference between \textit{Seurat} and \textit{SCTransform}'s regression originates from the use of Pearson's residuals performed during the second round regression of \textit{SCTransform}, which are then independent of sequencing depth.
-\textit{SCTransform} systematically outperformed other methods and, even though it was developed to be applied to data with unique molecular identifiers (UMI), it also performed fairly well with the Smart-seq protocol (Koh and Kumar datasets).\textit{scVI} (Supplementary Figure 13) performed poorly on most datasets in terms of both the subpopulation silhouette and clustering accuracy.
+\textit{SCTransform} systematically outperformed other methods and, even though it was developed to be applied to data with unique molecular identifiers (UMI), it also performed fairly well with the Smart-seq protocol (Koh and Kumar datasets).\textit{scVI} (Supplementary Figure 15) performed poorly on most datasets in terms of both the subpopulation silhouette and clustering accuracy.
 For time and resource cost reasons, \textit{scVI} was run on a reduced set of pipelines and compared to \textit{SCTransform} (more details in the Methods section).
@@
 Finally, we monitored whether, under the same downstream clustering analysis, different normalization methods tended to lead to an over- or under-estimation of the number of clusters.
-Although some methods had a tendency to lead to a higher (e.g., \textit{SCTransform}) or lower (e.g., stable genes) number of clusters, the effect was very mild and not entirely systematic (Supplementary Figure 14).
-We also confirmed that, especially in the UMI-based datasets, \textit{SCTransform} did in fact successfully stabilize variance across mean counts (Supplementary Figure 15), at little apparent cost on computation time (Supplementary Figure 16).
+Although some methods had a tendency to lead to a higher (e.g., \textit{SCTransform}) or lower (e.g., stable genes) number of clusters, the effect was very mild and not entirely systematic (Supplementary Figure 16).
+We also confirmed that, especially in the UMI-based datasets, \textit{SCTransform} did in fact successfully stabilize variance across mean counts (Supplementary Figure 17), at little apparent cost on computation time (Supplementary Figure 18).
 
@@
 As the proportion of variability in a gene attributable to subpopulations can be measured in various ways, we first compared three approaches: ANOVA on log-normalized count, ANOVA on \textit{SCTransform} normalization, and deviance explained.
-The ANOVAs performed on a standard \textit{Seurat} normalization and on \textit{SCTransform} data were highly correlated (Pearson correlation >0.9, Supplementary Figure 17A).
-These estimates were also in good agreement with the deviance explained, although lowly-expressed genes could have a high deviance explained without having much of their variance explained by subpopulation (Supplementary Figure 17B-D).
-We therefore compared the proportion of the cumulative variance/deviance explained by the top X genes that could be retrieved by each gene ranking method (Supplementary Figures 18-19).
-We first focused on the first 1000 genes to highlight the differences between methods, although a higher number of selected genes decreased the differences between methods (Supplementary Figures 18-20).
+The ANOVAs performed on a standard \textit{Seurat} normalization and on \textit{SCTransform} data were highly correlated (Pearson correlation >0.9, Supplementary Figure 19A).
+These estimates were also in good agreement with the deviance explained, although lowly-expressed genes could have a high deviance explained without having much of their variance explained by subpopulation (Supplementary Figure 19B-D).
+We therefore compared the proportion of the cumulative variance/deviance explained by the top X genes that could be retrieved by each gene ranking method (Supplementary Figures 20-21).
+We first focused on the first 1000 genes to highlight the differences between methods, although a higher number of selected genes decreased the differences between methods (Supplementary Figures 20-22).
 The standardized measures of variability were systematically worse than their non-standardized counterparts in selecting genes with a high proportion of variance explained by subpopulation.
-Regarding the percentage of deviance explained however, the standardized measures were often superior (Figure \ref{fig:selection}A and Supplementary Figures 18-19).
+Regarding the percentage of deviance explained however, the standardized measures were often superior (Figure \ref{fig:selection}A and Supplementary Figures 20-21).
 \textit{Deviance} proved the method of choice to prioritize genes with a high variance explained by subpopulations (with mere expression level proving surprisingly performant) but did not perform so well to select genes with a high deviance explained.
@@
 In general, we found deviance and unstandardized estimates of variance to provide the best results across datasets and normalization methods.
-Increasing the number of features selected also systematically led to an increase in the accuracy of the clustering, typically plateauing after 4000 features (Supplementary Figure 20).
+Increasing the number of features selected also systematically led to an increase in the accuracy of the clustering, typically plateauing after 4000 features (Supplementary Figure 22).
 
@@
 
-The impact of the choice of dimensionality reduction method was far greater than that of normalization or feature selection (Figure \ref{fig:figure7}).
+The impact of the choice of dimensionality reduction method was far greater than that of normalization or feature selection (Supplementary Figure 23).
 Again, the clustering on the biggest datasets (simMix and Zhengmix) was more challenging for most methods.
@@
 Like GLM-PCA, \textit{scVI}'s Linear Decoder (LD) does not explicitly rely on normalized counts.
-Its performance was lower than a pipeline with non-weighted \textit{Seurat} PCA on all measures for datasets with more subpopulations (Supplementary Figure 13).
+Its performance was lower than a pipeline with non-weighted \textit{Seurat} PCA on all measures for datasets with more subpopulations (Supplementary Figure 15).
 Overall, weighting the principal components by their variance (as \textit{Seurat} does) had a positive impact on silhouette widths and ARI scores.
@@
 Since Euclidean distance decreases as the number of non-discriminating dimensions increases, there is usually a trade-off between selecting enough dimensions to keep most information and excluding smaller dimensions that may represent technical noise or other unwanted sources of variation.
-Overall, increasing the number of dimensions robustly led to a decrease in the number of clusters (Supplementary Figures 14 and 21).
-This tended to affect the accuracy of the clustering (Supplementary Figure 22), although in both cases (number of clusters and ARI), \textit{Seurat}'s resolution parameter had a much stronger impact.
+Overall, increasing the number of dimensions robustly led to a decrease in the number of clusters (Supplementary Figures 16 and 24).
+This tended to affect the accuracy of the clustering (Supplementary Figure 25), although in both cases (number of clusters and ARI), \textit{Seurat}'s resolution parameter had a much stronger impact.
 
@@
 Reasoning that over-estimating the number of dimensions is less problematic than under-estimating it, we kept the former methods for a full analysis of their impact on clustering (Figure \ref{fig:dim}C-D), when combined with \textit{SCTransform} or the standard \textit{Seurat} normalization.
-Although most methods performed well on the various clustering measures, the global maximum likelihood based on translated Poisson Mixture Model (\textit{maxLikGlobal}, using the 20 nearest neighbors, although estimates were relatively stable across decent values of \textit{k}; see Supplementary Figure 23) provided the dimensionality estimate that best separated the subpopulations (Figure \ref{fig:dim}C) and resulted in the best clustering accuracy (Figure \ref{fig:dim}D).
+Although most methods performed well on the various clustering measures, the global maximum likelihood based on translated Poisson Mixture Model (\textit{maxLikGlobal}, using the 20 nearest neighbors, although estimates were relatively stable across decent values of \textit{k}; see Supplementary Figure 26) provided the dimensionality estimate that best separated the subpopulations (Figure \ref{fig:dim}C) and resulted in the best clustering accuracy (Figure \ref{fig:dim}D).
 This method systematically selected many more components than were associated with the subpopulations, suggesting that although these additional components appear individually uninformative, in combination they nevertheless contribute to classification.
@@
 
-Since ARI is dominated by differences in the number of clusters (Supplementary Figures 1-2) and no single metric is perfect, we diversified them (Figure \ref{fig:clust}).
-MI has the virtue of not decreasing when a true subpopulation is split into two clusters, which is arguably less problematic (and might well reflect unknown biological subgroups), but as a consequence it can be biased towards methods producing higher resolution clustering (Supplementary Figure 2).
+Since ARI is dominated by differences in the number of clusters (Supplementary Figures 2-3) and no single metric is perfect, we diversified them (Figure \ref{fig:clust}).
+MI has the virtue of not decreasing when a true subpopulation is split into two clusters, which is arguably less problematic (and might well reflect unknown biological subgroups), but as a consequence it can be biased towards methods producing higher resolution clustering (Supplementary Figure 3).
 Similarly, precision per true subpopulation is considerably more robust to differences in the number of clusters.
@@
 Finally, the ARI score at the true number of clusters, when available, showed similar performances.
-However, because \textit{Seurat}'s resolution parameter had a large impact on the number of clusters identified (Supplementary Figures 1 and 23), \textit{Seurat} could always be coerced into producing the right number of clusters.
-Instead, the number of clusters found by \textit{scran} was considerably less influenced by the available parameters (number of nearest neighbors or steps in the random walk - see Supplementary Figure 24), and as a result \textit{scran}-based clustering sometimes never produced a partitioning with the right number of clusters.
+However, because \textit{Seurat}'s resolution parameter had a large impact on the number of clusters identified (Supplementary Figures 2 and 24), \textit{Seurat} could always be coerced into producing the right number of clusters.
+Instead, the number of clusters found by \textit{scran} was considerably less influenced by the available parameters (number of nearest neighbors or steps in the random walk - see Supplementary Figure 27), and as a result \textit{scran}-based clustering sometimes never produced a partitioning with the right number of clusters.
 This observation, along with \textit{scran}'s higher MI score, suggest that \textit{scran} sometimes simply divides a real subpopulation into two clusters (possibly tracking some unknown biological differences) rather than committing misclassification errors.
@@
 Since preliminary analysis showed that all methods performed equally well or better on normalized data, we applied them after filtering and normalization, but before scaling and reduction.
-Although some of the methods (e.g., \textit{DRImpute\_process} and \textit{alra\_norm}) did improve the separability of some more elusive subpopulations, no method had a systematically positive impact on the average silhouette width across all subpopulations (Figure \ref{fig:figure10}A).
-When restricting ourselves to clustering analyses that yielded the `right' number of clusters, all tested methods improved classification  compared to a scenario with no imputation step (`none' label, Figure \ref{fig:figure10}B).
+Although some of the methods (e.g., \textit{DRImpute\_process} and \textit{alra\_norm}) did improve the separability of some more elusive subpopulations, no method had a systematically positive impact on the average silhouette width across all subpopulations (Supplementary Figure 28A).
+When restricting ourselves to clustering analyses that yielded the `right' number of clusters, all tested methods improved classification  compared to a scenario with no imputation step (`none' label, Supplementary Figure 28B).
 However, the situation was not so straightforward with alternative metrics, where some methods (e.g., \textit{ENHANCE}) consistently underperformed.
@@
 We developed 3 benchmark datasets: 1) the \texttt{ipsc} data is a random selection of 10 vs 10 samples from \citet{carcamoOriveAnalysisTranscriptional2017}, which are very heterogeneous, and foldchanges were manually added to 300 genes; 2) the \texttt{seqc} dataset contains 5 vs 5 samples of two mixtures of the SEQC consortium with different ERCC spike-in mixes \citep{ComprehensiveAssessment2014}, and includes a partially correlated batch effect; 3) A **simulation** of 8 vs 8 samples with two technical vectors of variation (see methods for more details).
-As previously reported, we found that more stringent filtering improved DEA (Supplementary Figure 25A) and that the different DEA methods had similar performances (Figure \ref{fig:svadea}A and Supplementary Figure 25A).
-Including a SVA-based step increased sensitivity across all datasets, and particularly so in datasets with technical vectors of variation, and benefited all DEA methods in a similar fashion (Figure \ref{fig:svadea}A,C and Supplementary Figures 25-26).
+As previously reported, we found that more stringent filtering improved DEA (Supplementary Figure 29A) and that the different DEA methods had similar performances (Figure \ref{fig:svadea}A and Supplementary Figure 29A).
+Including a SVA-based step increased sensitivity across all datasets, and particularly so in datasets with technical vectors of variation, and benefited all DEA methods in a similar fashion (Figure \ref{fig:svadea}A,C and Supplementary Figures 29-30).
 With the exception of the \texttt{RUVr} method, which led to a high False Discovery Rate (FDR), all other methods showed comparable gains in sensitivity while maintaining a decent error control (if a little above the nominal FDR), even when increasing the number of surrogate variables used (Figure \ref{fig:svadea}B-C).
@@
 
-On the basis of our findings, we can make a number of practical analysis recommendations, also summarized in Figure \ref{fig:recommendations}:
+On the basis of our findings, we can make a number of practical analysis recommendations, also summarized in Supplementary Figure 31:
 
@@
 In comparison to the performance on other datasets, this inconsistency was particularly striking when evaluating the dimensionality estimation dimensionality reduction and the clustering methods, for which half of the evaluated methods could not lead to the true number of clusters.
-It is likely that these two datasets are challenging due to their combined sparsity and complexity (see Figure 1 \ref{fig:figure1}).
+It is likely that these two datasets are challenging due to their combined sparsity and complexity (see Supplementary Figure 1).
 This issue that was already observed for the Zhengmix8eq dataset in a previous benchmark \cite{duoClustering2018}, as it contains deeper levels of annotations for subpopulations of T-cells.
@@
 \textit{pipeComp} is a flexible R framework for evaluating methods and parameter combinations in a complex pipeline, computing on-the-fly multilevel evaluation metrics.
-Applying this framework to scRNAseq clustering enabled us to make clear recommendations on the steps of filtering, normalization, feature selection, denoising, dimensionality reduction and clustering (Figure \ref{fig:recommendations}).
+Applying this framework to scRNAseq clustering enabled us to make clear recommendations on the steps of filtering, normalization, feature selection, denoising, dimensionality reduction and clustering (Supplementary Figure 31).
 We demonstrate how a diversity of multilevel metrics can be more robust, more sensitive, and more nuanced than simply evaluating the final clustering performance.

### 2020-05-08 main_figures/dimensionality.Rmd
some figure updates

@@
   library(matrixStats)
+  library(pipeComp) # the first part requires pipeComp >0.99.26
 })
 source("../misc_functions.R")
-source(system.file("extdata/scrna_alternatives.R", package="pipeComp"))
 theme_set(theme_cowplot(font_size = 12))
@@
 
+```{r, fig.width=8, fig.height=4}
+rnfn <- function(x){
+  x <- gsub("norm\\.","",x)
+  x <- gsub("maxLikGlobal$","maxLikGlobal(k=20)",x)
+  gsub("maxLikGlobal10$","maxLikGlobal(k=10)",x)
+}
+res <- readRDS("../data/results_dimensionality.rds")
+H <- renameHrows(
+  scrna_evalPlot_silh( res, step="dimensionality", reorder_rows=TRUE, 
+                       row_split=gsub("norm\\.","",norm),
+                       heatmap_legend_param=list(direction="horizontal")),
+  rnfn)
+p2 <- grid.grabExpr(draw(H, heatmap_legend_side="bottom"))
+
+p3 <- renameHrows(
+  evalHeatmap( res, what=c("NMI","min_pr"), agg.by=c("norm","dims"), 
+                   row_split=gsub("norm\\.","",norm), reorder_rows=TRUE, anno_legend=FALSE) + 
+  evalHeatmap(res, what=c("ARI"), agg.by=c("norm","dims"), row_split=gsub("norm\\.","",norm), 
+              reorder_rows=TRUE, filterExpr=n_clus==true.nbClusts, 
+              anno_legend=FALSE, title="ARI at\ntrue k"),
+  rnfn
+)
+p3 <- grid.grabExpr(draw(p3))
+```
+
 ```{r, eval=FALSE}
-# NB: the results were generated using an older version of pipeComp (branch 
+# NB: the next results were generated using an older version of pipeComp (branch 
 # `biorxiv`), and must be plotted using this version of the package
@@
 
-```{r, include=FALSE}
-# the second part of these results was generated with pipeComp >0.99.26
-devtools::load_all("../../pipeComp/")
-```
-
-```{r, fig.width=8, fig.height=4}
-res <- readRDS("../data/results_dimensionality.rds")
-H <- renameHrows(
-  scrna_evalPlot_silh( res, step="dimensionality", reorder_rows=TRUE, 
-                       row_split=gsub("norm\\.","",norm),
-                       heatmap_legend_param=list(direction="horizontal")),
-  function(x) gsub("norm\\.","",x))
-p2 <- grid.grabExpr(draw(H, heatmap_legend_side="bottom"))
-
-p3 <- renameHrows(
-  evalHeatmap( res, what=c("NMI","min_pr"), agg.by=c("norm","dims"), 
-                   row_split=gsub("norm\\.","",norm), reorder_rows=TRUE, anno_legend=FALSE) + 
-  evalHeatmap(res, what=c("ARI"), agg.by=c("norm","dims"), row_split=gsub("norm\\.","",norm), 
-              reorder_rows=TRUE, filterExpr=n_clus==true.nbClusts, 
-              anno_legend=FALSE, title="ARI at\ntrue k"),
-  function(x) gsub("norm\\.","",x)
-)
-p3 <- grid.grabExpr(draw(p3))
-```
-
 ```{r figure6, fig.width=9, fig.height=9.5}
 pdf("dimensionality.pdf", width=9, height=10)
-plot_grid( p1, p2, p3, nrow=3, rel_heights=c(0.36,0.32,0.32), labels=c(NA,LETTERS[3:4]))
+plot_grid( p1, p2, p3, nrow=3, labels=c(NA,LETTERS[3:4]))
 dev.off()
-plot_grid( p1, p2, p3, nrow=3, rel_heights=c(0.36,0.32,0.32), labels=c(NA,LETTERS[3:4]))
+plot_grid( p1, p2, p3, nrow=3, labels=c(NA,LETTERS[3:4]))
 ```

### 2020-05-08 supplementary_figures/supp2_filtering.Rmd
some figure updates

--- 2020-02-27 supplementary_figures/supp2_filtering.Rmd
@@
   library(pipeComp)
+  library(grid)
 })
@@
 
+```{r, fig.height=8.5, fig.width=7}
+res <- readRDS("../data/filtering_intermediateSummary.rds")
+tmp <- res$filtering$pcCells
+dbl <- 1+grepl("doublet",row.names(tmp))
+row.names(tmp) <- paste0(c("","doublet + ")[dbl],sapply(strsplit(row.names(tmp)," "), FUN=function(x) x[2]))
+cols <- colorRamp2( seq(from=0,to=sqrt(max(tmp)),length.out=100)^2,
+                    colorRampPalette(c("white","red"))(100) )
+H <- Heatmap(tmp, bottom_annotation=pipeComp:::.ds_anno(colnames(tmp)), 
+        show_column_names = FALSE, cluster_columns=FALSE, show_row_dend=FALSE, 
+        row_split=c("No doublet removal","Doublet removal")[dbl], 
+        name="proportion\nexcluded", col=cols, heatmap_legend_param=list(
+          at=c(0,5,10,30,60), labels=paste0(c(0,5,10,30,60),"%"),
+          grid_height=unit(7,"mm")) )
+
+res <- readRDS("../data/overall.rds")
+# get cell numbers
+nb <- aggregate(res$evaluation$doublet$N.before, res$evaluation$doublet[,"dataset",drop=FALSE], FUN=sum)
+row.names(nb) <- nb[,1]
+res <- readRDS("../data/excluded.rds")
+res <- lapply(res, FUN=function(x){
+  names(x) <- gsub("doubletmethod=none;filt=filt.","",names(x))
+  x <- x[names(x)!="doubletmethod=none;filt=none"]
+  names(x) <- c("Doublet removal", "Default filter", "Stringent filter")
+  x
+})
+res2 <- lapply(names(res), FUN=function(x){
+  e <- sapply(res[[x]], FUN=function(i) sapply(res[[x]], FUN=function(j) sum(i %in% j)))
+  e/nb[x,2]*100
+})
+names(res2) <- names(res)
+h2 <- Heatmap(do.call(cbind, res2), column_split = rep(names(res2), each=ncol(res2[[1]])), 
+              cluster_columns = FALSE, cluster_rows = FALSE, col=c("white","red"), 
+              name="% of\ndataset", column_title_gp = gpar(font.size=10), column_title=" ",
+              top_annotation=pipeComp:::.ds_anno(rep(names(res),each=ncol(res2[[1]])),legend = FALSE))
+```
+
+```{r, fig.height=6.5, fig.width=8}
+plot_grid( grid.grabExpr(draw(H,merge_legend=TRUE)), grid.grabExpr(draw(h2)), 
+           nrow=2, rel_heights=c(6,4), labels=LETTERS[1:2])
+```
+
+### Supplementary Figure `r getFigNb()`
+
+**A:** Proportion of cells filtered out by subpopulation.
+Applying the same filters in a cluster-wise fashion (using `scran::quickCluster`, and desginated here with `clustFilter*`) leads to virtually no cell exclusion.
+The color-mapping is square-root transformed to improve the visibility of differences at low proportions.
+**B:** Overlap between cells excluded by doublet removal (`scDblFinder`) and those excluded by MAD-based filters (without doublet removal; the filters are described in the methods), expressed as a proportion of the dataset.
+The cells excluded as doublets do not tend to be excluded by (even stringent) MAD-based filtering.
+
+\newpage
+
+# Supplementary Figure `r getFigNb(TRUE)`
+
 ```{r}

### 2020-05-08 supplementary_figures/supp_overall.Rmd
some figure updates

---
title: "Supplementary Figures relative to feature selection"
author: "Pierre-Luc Germain"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r, include=FALSE}
if(!exists('FIG_NB')){
  FIG_NB <- 0
  getFigNb <- function(increment=FALSE){
    if(increment) FIG_NB <<- FIG_NB + 1
    FIG_NB
  }
}

suppressPackageStartupMessages({
  library(ggplot2)
  library(cowplot)
  library(ComplexHeatmap)
  library(pipeComp)
})
theme_set(theme_cowplot(font_size = 11))
source("../misc_functions.R")
```

# Supplementary Figure `r getFigNb(TRUE)`

```{r, fig.width=7, fig.height=6}
ov <- readRDS("../data/overall.rds")

cl <- ov$evaluation$clustering
cl$norm <- relevel(cl$norm, "norm.seurat")
cl$doubletmethod <- relevel(cl$doubletmethod, "none")
cl$filt <- relevel(cl$filt, "none")
cl$sel <- relevel(cl$sel, "sel.vst")
cl$clustmethod <- relevel(cl$clustmethod, "clust.seurat")
cl$clustmethod <- relevel(cl$clustmethod, "clust.seurat")

ff <- function(x, trueK=FALSE){
  f <- paste0(x,"~dataset+resolution*dataset+doubletmethod+filt+norm+clustmethod+dims")
  if(trueK) cl <- cl[cl$n_clus==cl$true.nbClusts,]
  mod <- lm(as.formula(f), data=cl)
  co <- as.data.frame(coef(summary(mod)))
  co <- co[grep("sctransfrom|scran|filt|elbow|doublet|Annoy", row.names(co)),]
  colnames(co)[2:4] <- c("stderr","t","pvalue")
  co$coef <- gsub(" +$","",row.names(co))
  aov <- as.data.frame(summary(aov(mod))[[1]])[,-1]
  aov <- aov[grep("^Residuals|resolution|dataset", row.names(aov), invert=TRUE),]
  aov$varExpl <- aov[,1]/sum(aov[,1])
  aov <- aov[grep("^Residuals", row.names(aov), invert=TRUE),4:5]
  colnames(aov)[1] <- "pvalue"
  aov$parameter <- gsub(" +$","",row.names(aov))
  list( coefs=co, aov=aov )
}

cos <- lapply(c(MI="MI",ARI="ARI",minPre="min_pr"), FUN=ff)
aov <- dplyr::bind_rows(lapply(cos, FUN=function(x) x$aov), .id="metric")

cos2 <- lapply(c(ARI="ARI"), trueK=TRUE, FUN=ff)
aov2 <- dplyr::bind_rows(lapply(cos2, FUN=function(x) x$aov), .id="metric")
p1 <- ggplot(aov, aes(reorder(parameter, -varExpl), varExpl, group=parameter, fill=metric)) + 
  geom_col() + xlab("") + ylab("Proportion of variance explained") + coord_flip()
p2 <- ggplot(aov2, aes(reorder(parameter, -varExpl), varExpl, group=parameter, fill=metric)) + 
  geom_col() + xlab("") + ylab("Variance explained") + theme(legend.position="none") + 
  coord_flip() + ggtitle("ARI at true k")

co <- dplyr::bind_rows(lapply(cos, FUN=function(x) x$coef), .id="metric")
co2 <- dplyr::bind_rows(lapply(cos2, FUN=function(x) x$coef), .id="metric")
for(f in c("norm","filt","doubletmethod","dims","clustmethod")){
  co$coef <- gsub(paste0("^",f),"",co$coef)
  co2$coef <- gsub(paste0("^",f),"",co2$coef)
}
p3 <- ggplot(co, aes(coef, Estimate, group=coef, fill=metric)) + geom_col() + xlab("") +
  ylab("Estimated coefficient") + geom_hline(yintercept = 0, linetype="dashed") + 
  coord_flip() + theme(legend.position="none")
p4 <- ggplot(co2, aes(coef, Estimate, group=coef, fill=metric)) + geom_col() + xlab("") +
  ylab("Estimated coefficient") + geom_hline(yintercept = 0, linetype="dashed") + 
  coord_flip() + theme(legend.position="none")

# plot_grid( plot_grid(p1,p2,rel_widths=c(4,3),labels=LETTERS[1:2]),
#            plot_grid(p3,p4,rel_widths=c(4,3),labels=LETTERS[3:4]),
#            nrow=2, rel_heights=c(3,4))
plot_grid(p1 + ylab("Proportion of variance in metric explained by the parameter"),p2,nrow=2,labels=LETTERS[1:2])
```

### Supplementary Figure `r getFigNb()`

**A:** Proportion of the variance in clustering accuracy metrics explained by main parameters. **B:** Proportion of the variance in Adjusted Rand Index (ARI) explained by the parameters, among the analyses giving the true number of clusters. In both panels, the proportion is calculated after accounting for differences due to the dataset, the resolution and their interaction.

\newpage

# Supplementary Figure `r getFigNb(TRUE)`

```{r}
rnfn <- function(x){
  x <- gsub("^norm\\.","",x)
  x <- gsub("^doublet\\.","",x)
  x <- gsub("> none > ","> no filter > ",x)
  gsub("^none > ","",x)
}

H1 <- renameHrows( scrna_evalPlot_silh(ov, agg.by = c("doubletmethod","filt","norm"), row_split = norm,
                                       heatmap_legend_param=list(direction="horizontal"),font_factor = 0.8,
                                       row_names_max_width=unit(7, "cm")), rnfn)
H2 <- renameHrows(
     evalHeatmap(ov, step="clustering", what=c("MI","min_pr"), anno_legend=FALSE, title=c("MI","Min precision"), 
                 agg.by=c("doubletmethod","filt","norm"), font_factor = 0.8, 
                 row_split=norm,row_names_max_width=unit(7, "cm")) + 
    evalHeatmap(ov, step="clustering", what=c("ARI"), agg.by=c("doubletmethod","filt","norm"), 
                row_split=norm,filter=n_clus==true.nbClusts, title="ARI at true k",
                anno_legend=FALSE, font_factor = 0.8,row_names_max_width=unit(7, "cm")),
  rnfn)

H3 <- renameHrows( scrna_evalPlot_silh(ov, agg.by = c("doubletmethod","sel","norm"), row_split = norm,
                                       heatmap_legend_param=list(direction="horizontal"),
                                       row_names_max_width=unit(7, "cm")), rnfn)
H4 <- renameHrows(
     evalHeatmap(ov, step="clustering", what=c("MI","min_pr"), anno_legend=FALSE, title=c("MI","Min precision"),
                 agg.by=c("doubletmethod","sel","norm"), row_split=norm,row_names_max_width=unit(7, "cm")) + 
    evalHeatmap(ov, step="clustering", what=c("ARI"), agg.by=c("doubletmethod","sel","norm"), 
                row_split=norm,filter=n_clus==true.nbClusts, title="ARI at true k", 
                anno_legend=FALSE, row_names_max_width=unit(7, "cm")),
  rnfn)
```

```{r, fig.width=9, fig.height=8.5}
plot_grid( grid.grabExpr(draw(H1, heatmap_legend_side="bottom")), 
           grid.grabExpr(draw(H2)), nrow=2, labels=LETTERS[1:2])
```

### Supplementary Figure `r getFigNb()`

Silhouette widths of the real subpopulations (**A**) and clustering accuracy (**B**) using different combinations of methods.

\newpage

# Supplementary Figure `r getFigNb(TRUE)`

```{r, fig.width=9, fig.height=8.5}
plot_grid( grid.grabExpr(draw(H3, heatmap_legend_side="bottom")), 
           grid.grabExpr(draw(H4)), nrow=2, labels=LETTERS[1:2])
```

### Supplementary Figure `r getFigNb()`

Silhouette widths of the real subpopulations (**A**) and clustering accuracy (**B**) using different combinations of methods.

### 2020-05-08 bmc_article.tex

@@
 \begin{table}[h!]
-\caption{Overview of the benchmark datasets.}
+\caption{\textbf{Overview of the benchmark datasets.} `\# features' indicates the number of features detected in at least 10 cells.}
 \label{tab:table1}
@@
   \hline
-dataset & source & \# cells & protocol & description \\ 
+dataset & \# features & \# cells & protocol & description \\ 
   \hline
-Koh & GSE85066 & 531 & SMARTer & FACS purified differentiation stages \\ 
-  Kumar & GSE60749 & 246 & SMARTer & Mouse ESC cultured in different conditions \\ 
-  Zhengmix4eq & \cite{duoClustering2018} & 3994 & 10x & Mixtures of FACS purified PBMCs \\ 
-  Zhengmix4uneq & \cite{duoClustering2018} & 6498 & 10x & Mixtures of FACS purified PBMCs \\ 
-  Zhengmix8eq & \cite{duoClustering2018} & 3994 & 10x & Mixtures of FACS purified PBMCs \\ 
-  mixology10x3cl & \cite{tianMixology2018} & 902 & 10x & Mixture of 3  cancer cell lines from CellBench \\ 
-  mixology10x5cl & \cite{tianMixology2018} & 3918 & 10x & Mixture of 5 cancer cell lines from CellBench \\ 
-  simMix1 & - & 2500 & 10x-based & Simulation of 10 human cell subpopulations \\
-  simMix2 & - & 3000 & 10x-based & Simulation of 9 mouse cell subpopulations \\
+Koh & 33922 & 531 & SMARTer & FACS purified differentiation stages \\ 
+  Kumar & 41930 & 246 & SMARTer & Mouse ESC cultured in different conditions \\ 
+  Zhengmix4eq & 10434 & 3994 & 10x & Mixtures of FACS purified PBMCs \\ 
+  Zhengmix4uneq & 11369 & 6498 & 10x & Mixtures of FACS purified PBMCs \\ 
+  Zhengmix8eq & 10600 & 3994 & 10x & Mixtures of FACS purified PBMCs \\ 
+  mixology10x3cl & 16208 & 902 & 10x & Mixture of 3  cancer cell lines from CellBench \\ 
+  mixology10x5cl & 11786 & 3918 & 10x & Mixture of 5 cancer cell lines from CellBench \\ 
+  simMix1 & 3696 & 2500 & 10x-based & Simulation of 10 human cell subpopulations \\
+  simMix2 & 8893 & 3000 & 10x-based & Simulation of 9 mouse cell subpopulations \\
    \hline
@@
   \subsection*{\textbf{Additional file 1: Supplementary figures}}
-  Additional results of evaluations across datasets among various metrics. (PDF 15 MB)
+  Additional results of evaluations across datasets among various metrics. (PDF 16 MB)
   \subsection*{\textbf{Additional file 2: Supplementary methods}}

### 2020-05-08 data/overall_run.Rmd
added run info for the overall benchmark

---
title: "Overall run"
output: html_document
---

```{r}
suppressPackageStartupMessages({
  library(Seurat)
  library(SingleCellExperiment)
  library(Matrix)
  library(devtools)
  library(scater)
  library(pipeComp)
})
source(system.file("extdata", "scrna_alternatives.R", package="pipeComp"))
ds_files <- list.files("datasets", full.names = T)
names(ds_files) <- gsub(".SCE.rds","",basename(ds_files),fixed=T)
names(ds_files)
```

## Methods

```{r}
clust.scran.knnAnnoy <- function(ds, ...){
  clust.scran(ds, graph.type="knn", neighbor.method="Annoy", ...)
}
none <- function(x, ...) return(x)

pd <- scrna_pipeline(pipeClass="seurat")

alternatives <- list(
  doubletmethod=c("none","doublet.scDblFinder"),
  filt=c("none","filt.default","filt.stringent"),
  norm=c("norm.seurat","norm.sctransform","norm.scran"),
  sel=c("sel.vst","sel.deviance","sel.expr"),
  selnb=c(2000,4000),
  dr=c("seurat.pca"),
  maxdim=50,
  dims=c("elbow", 10, "maxLikGlobal"),
  clustmethod=c("clust.seurat","clust.scran.knnAnnoy"),
  k=c(5,8,10,15,20,30),
  steps=4,
  resolution=c(0.001,0.005,0.01,0.02,0.05,0.1,0.15,0.2,0.3,0.4,0.5,0.8,1,1.2,1.5,2,4),
  min.size=20
)

comb <- buildCombMatrix(alternatives)
comb <- comb[(comb$clustmethod=="clust.seurat" | comb$resolution==0.8),]
```

## Run

```{r}
res <- runPipeline(ds_files, alternatives, pd, output.prefix="pipeComp_overall/", nthreads=18, comb=comb)
saveRDS(res, file="pipeComp_overall/summary.rds")
```

```{r session_info}
sessionInfo()
```

### 2020-05-08 bmc_article.tex

@@
 
-We next investigated the impact of filtering according to six methods with varying stringency; three methods based on deviations to MADs with increasing levels of stringency (named \textit{lenient}, \textit{default}, \textit{stringent}, \textit{very stringent}) and two methods based on \textit{scater}'s \textit{runPCA} using all (\textit{pca.all}) or selected covariates (\textit{pca.sel}, see Methods for more details).
+We next investigated the impact of filtering according to several methods (see Methods for more details): four methods based on deviations to MADs with increasing levels of stringency (named \textit{lenient}, \textit{default}, \textit{stringent}, \textit{very stringent}) and two methods based on \textit{scater}'s \textit{runPCA} using all (\textit{pca.all}) or selected covariates (\textit{pca.sel}).
 An important risk of excluding cells on the basis of their distance from the whole distribution of cells on some properties (e.g., library size) is that these properties tend to have different distributions across subpopulations.
@@
 Most of the benefits could be achieved without very stringent filtering and minimizing subpopulation bias (Figure \ref{fig:filtering}B).
-Applying the same filtering criteria on individual clusters of cells (identified through scran's \texttt{quickCluster} method) rather than on whole dataset resulted in nearly no cells being filtered out (Supplementary Figure 12).
+Applying the same filtering criteria on individual clusters of cells (identified through scran's \texttt{quickCluster} method) rather than on whole dataset resulted in nearly no cells being filtered out (Supplementary Figure 12A).
 This suggests that stringent filtering on the global population tends to discard cells of subpopulations with more extreme properties (e.g.
@@
 On the opposite, too lenient filtering (or lack thereof) can lead to misclassified cells, as was for instance the case especially with the Zhengmix4 datsets (Figure \ref{fig:filtering}B).
-We found that a relatively mild filtering (`default' - see methods) often leads to high clustering performance (e.g.
-simMix and Zhengmix4 datasets) and always retain most of the cells in each subpopulation (e.g.
+We found that a relatively mild filtering (`default' - see methods) provided a good trade-off, often leading to high clustering performance (e.g.
+simMix and Zhengmix4 datasets) while retaining most of the cells in each subpopulation (e.g.
 Koh and Zhengmix8eq).
-This suggests that a mild ('default') filtering is a very good trade-off between clustering accuracy and subpopulations preservation.
-Finally, the doublet removal step in conjunction with filtering  resulted in a net decrease in the proportion of excluded cells while retaining or improving accuracy not only in datasets that do include heterotypic doublets (mixology datasets), but also some others in which such doublets are unlikely (e.g.
-Zheng datasets).
-On the Smart-Seq and simulated datasets however, the removal of doublets had little to no impact on any of the two measures.
+We found very little overlap between the cells excluded by doublet removal and those excluded by MAD-based filtering (Supplementary Figure 12B).
+Of note, doublet removal in conjunction with filtering resulted in a net decrease in the proportion of excluded cells while retaining or improving accuracy not only in datasets that do include heterotypic doublets (mixology datasets), but also some others in which such doublets are unlikely given the experimental design (i.e.
+sorted datasets such as the Zheng datasets).
+Indeed, the datasets where doublet removal did not have an impact (e.g.
+Koh, Kumar, and the simulations) are those in which heterotypic doublets are not expected.
 Overall, we therefore recommend the use of doublet removal followed by our `default' filtering (see Methods) or a similar approach.
@@
 We next investigated the impact of different normalization strategies.
-Beside the standard log-normalization included in \textit{Seurat}, we tested \textit{scran}'s pooling-based normalization \cite{lunPooling2016}, \textit{SCTransform}'s variance-stabilizing transformation \cite{hafemeisterSCtransform2019}, normalization based on stable genes \cite{linStableGenes2018, deekeStablyExpressed2018}, and SCnorm \cite{bacherSCnorm2017} (\textit{scVI} \cite{lopezDeep2018} was included separately on a subset of parameter combinations due to high running times).
+Beside the standard log-normalization included in \textit{Seurat}, we tested \textit{scran}'s pooling-based normalization \cite{lunPooling2016}, \textit{SCTransform}'s variance-stabilizing transformation \cite{hafemeisterSCtransform2019}, normalization based on stable genes \cite{linStableGenes2018, deekeStablyExpressed2018}, and \textit{SCnorm} \cite{bacherSCnorm2017} (\textit{scVI} \cite{lopezDeep2018} was included separately on a subset of parameter combinations due to high running times).
 In addition to log-normalization, the standard \textit{Seurat} clustering pipeline performs per-feature unit-variance scaling so that the PCA is not too strongly dominated by highly-expressed features.
@@
 However, as library size tends to vary across subpopulations, part of this effect can simply reflect biological differences.
-We therefore assessed to what extent the first principal component still retained a correlation with the library size and the number of detected features, removing the confounding covariation with the subpopulations (Figure \ref{fig:norm}A and Supplementary Figure 12).
+We therefore assessed to what extent the first principal component still retained a correlation with the library size and the number of detected features, removing the confounding covariation with the subpopulations (Figure \ref{fig:norm}A and Supplementary Figure 14).
 Scaling tended to remove much of the correlation with these features, and while most methods were able to remove most of the effect, the correlation was lowest with \textit{SCTransform}.
@@
 
-\subsection*{Application of _pipeComp_ to a different context}
+\subsection*{Application of \textif{pipeComp} to a different context}
 
@@
 The rationale behind such approaches is that the identification of factors determining genome-wide variation (i.e.
-surrogate variables) and their inclusion in the generalized linear model for DEA might improve DEA, especially in the presence of sources of variation unrelated to the experimental variables.
-\textit{sva}\citep{leekCapturingHeterogeneityGene2007,leekSvaPackageRemoving2012} and \textit{RUVseq}\citep{rissoNormalizationRNAseq2014a} are two popular packages implementing methods to infer these variables.
+surrogate variables) and their inclusion in the DEA's generalized linear model might improve DEA, especially in the presence of sources of variation unrelated to the experimental variables.
+\textit{sva}\citep{leekCapturingHeterogeneityGene2007,leekSvaPackageRemoving2012} and \textit{RUVseq}\citep{rissoNormalizationRNAseq2014a} are two popular packages each implementing different methods to infer these variables.
 We therefore created a \texttt{PipelineDefinition} with three step: 1) feature filtering, 2) removal of unwanted variation (i.e.
-SVA), and 3) DEA using the most established methods (the package's `pipeComp\_dea' vignette details the creation process).
+SVA), and 3) DEA using the most established methods.
+The package's `pipeComp\_dea' vignette details the creation process.
 We developed 3 benchmark datasets: 1) the \texttt{ipsc} data is a random selection of 10 vs 10 samples from \citet{carcamoOriveAnalysisTranscriptional2017}, which are very heterogeneous, and foldchanges were manually added to 300 genes; 2) the \texttt{seqc} dataset contains 5 vs 5 samples of two mixtures of the SEQC consortium with different ERCC spike-in mixes \citep{ComprehensiveAssessment2014}, and includes a partially correlated batch effect; 3) A **simulation** of 8 vs 8 samples with two technical vectors of variation (see methods for more details).
 As previously reported, we found that more stringent filtering improved DEA (Supplementary Figure 29A) and that the different DEA methods had similar performances (Figure \ref{fig:svadea}A and Supplementary Figure 29A).
-Including a SVA-based step increased sensitivity across all datasets, and particularly so in datasets with technical vectors of variation, and benefited all DEA methods in a similar fashion (Figure \ref{fig:svadea}A,C and Supplementary Figures 29-30).
+Including a SVA-based step increased sensitivity across all datasets, particularly in datasets with technical vectors of variation (\textit{seqc} and \textit{simulation}), and benefited all DEA methods in a similar fashion (Figure \ref{fig:svadea}A,C and Supplementary Figures 29-30).
 With the exception of the \texttt{RUVr} method, which led to a high False Discovery Rate (FDR), all other methods showed comparable gains in sensitivity while maintaining a decent error control (if a little above the nominal FDR), even when increasing the number of surrogate variables used (Figure \ref{fig:svadea}B-C).
+While the FDR did increase in the \textit{ipsc} dataset, we suspect that much of the effect could be due to spurious differences \citep{germainTamingHumanGenetic2017} between the groups which we wrongly assume to be false positives, and the FDR nevertheless remained acceptable with 1-3 variables.
 These results therefore suggest that these methods are robust and can be widely applied.
@@
 Analyses were performed in R 3.6.0 (Bioconductor 3.9) and the following packages were installed from GitHub repositories: \textit{Seurat} (version 3.0.0), \textit{SCTransform} (0.2.0), \textit{DoubletFinder} (2.0.1), \textit{scDblFinder} (1.1.4), \textit{scds} (1.0.0), \textit{SAVERX} (1.0.0), \textit{scImpute} (0.0.9), \textit{ALRA} (commit 7636de8), \textit{DCA} (0.2.2), \textit{DrImpute} (1.2), \textit{ENHANCE} (R version, commit 1571696), \textit{SAVERX} (1.0.0).
-The code for the glmPCA was obtained from \url{https://github.com/willtownes/scrna2019} (commit 1ddcc30ebb95d083a685f12fe81d35dd1b0cb1b2).
+The glmPCA was performed using the \texti{glmpca} package (The code for the glmPCA was obtained from \url{https://github.com/willtownes/scrna2019} (commit 1ddcc30ebb95d083a685f12fe81d35dd1b0cb1b2).
 
@@
 \begin{itemize}
-    \item For the \textit{ipsc} dataset, we took the quantification of the GSE79636 dataset (fairly heterogeneous) from the \textit{RNAontheBENCH} package \citep{germainRNAontheBENCHComputationalEmpirical2016}, and selected two random sets of 10 samples.
+    \item For the \textit{ipsc} dataset, we took the quantification of the GSE79636 dataset (fairly heterogeneous) from the \textit{iPSCpower} package \citep{germainTamingHumanGenetic2017}, and selected two random sets of 10 samples.
 No further variation was added, and log-foldchanges ranging from 0.25 to 1.25 were added to the counts of 300 random genes.
@@
 Throughout this paper, silhouette plots (panel \textbf{B}) square-root transform values for color mapping.
-Other evaluation metric heatmaps (panels \textbf{A,C}) map colors linearly to the number of (matrix-wise) Median Absolute Deviations from the (column-wise) median, while the printed numbers represent the raw metric values.}
+Other evaluation metric heatmaps (e.g.
+panel \textbf{C} here) map colors linearly to the number of (matrix-wise) Median Absolute Deviations from the (column-wise) median.
+Using this mapping, the same \textit{difference} in color has the same meaning across datasets, and the color-scale is not primarily capturing baseline differences between datasets.
+The numbers printed in the cells represent the raw (i.e.
+unscaled) metric values.}
     \label{fig:norm}
@@
     \fi
-    \caption{\textbf{Evaluation the impact of prior factor analysis on (bulk) RNAseq differential expression analysis using \textit{pipeComp}.
-A:} True positive rates (TPR) and False discovery rates (FDR) of various combinations of SVA-based methods (using a single surrogate variable) and DEA methods.
+    \caption{\textbf{Evaluation of the impact of surrogate variable analysis (SVA) on (bulk) RNAseq differential expression analysis using \textit{pipeComp}.} \textit{RUVr} and \textit{RUVs} refer to the two \textit{RUVSeq} functions by the same name; \textit{svaseq} refers to the function of the same name in \textit{sva}, while \textit{vstsva} refers to the original \textit{sva} method applied no variance-stabilized data.
+\textbf{A:} True positive rates (TPR) and False discovery rates (FDR) of various combinations of SVA-based methods (using a single surrogate variable) and DEA methods.
 \textbf{B:} TPR and FDR of the different SVA-based methods (averaged across the different DEA methods) using increasing numbers of variables (`be' and `leek` refer to the two methods implemented in the \textit{sva} package for estimating the number of surrogate variables).

### 2020-05-08 bmc_article.tex

@@
 
-We further evaluated normalization methods by investigating their impact on the separability of the subpopulations (Figure \ref{fig:norm}B-C).
-Since clustering accuracy metrics such as the ARI are very strongly influenced by the number of clusters, we complemented it with silhouette width \cite{RousseeuwSil1987} and mutual information (MI).
-We found most methods (including no normalization at all) to perform fairly well in most of the subpopulations.
-In view of the subpopulation silhouette and ARI at true number of clusters, it seems that all methods perform worse on the three Zhengmix datasets, that are the datasets with the highest number of cells and the smaller feature detection rate.
+We further evaluated normalization methods by investigating their impact on the separability of the subpopulations.
+Since clustering accuracy metrics such as the ARI are very strongly influenced by the number of clusters, we complemented it with silhouette width \cite{RousseeuwSil1987} and mutual information (MI, Figure \ref{fig:norm}B-C).
+We found most methods (including no normalization at all) to perform fairly well in most of the subpopulations, and that the same datasets (e.g.
+Zhengmix8eq and simMix1) and subpopulations were most challenging across for all methods.
+An exception was \textit{scVI}, which performed poorly on most datasets in terms of both the subpopulation silhouette and clustering accuracy (Supplementary Figure 13; \textit{scVI} was run on a reduced set of pipelines).
 Scaling (e.g.
 `seurat (no scale)' vs `seurat') tended to reduce the average silhouette width of some subpopulations and to increase that of some less distinguishable ones and was generally, but not always, beneficial on the accuracy of the final clustering.
-Regressing out covariates systematically gave equal  or poorer performance on MI, ARI and ARI at true number of clusters, with a notable negative impact when \textit{Seurat}'s regressing was applied to the Koh and Kumar datasets.
-It is possible that the difference between \textit{Seurat} and \textit{SCTransform}'s regression originates from the use of Pearson's residuals performed during the second round regression of \textit{SCTransform}, which are then independent of sequencing depth.
-\textit{SCTransform} systematically outperformed other methods and, even though it was developed to be applied to data with unique molecular identifiers (UMI), it also performed fairly well with the Smart-seq protocol (Koh and Kumar datasets).\textit{scVI} (Supplementary Figure 15) performed poorly on most datasets in terms of both the subpopulation silhouette and clustering accuracy.
-For time and resource cost reasons, \textit{scVI} was run on a reduced set of pipelines and compared to \textit{SCTransform} (more details in the Methods section).
+Regressing out covariates systematically gave equal  or poorer performance on MI, ARI and ARI at true number of clusters, with a notable negative impact in the Smart-seq datasets (Koh and Kumar).
+When performed with \textit{SCTransform}, regressing out covariates using tended to have neither a positive or negative effect in droplet-based datsets, possibly owing to its regularization.
+In general, \textit{SCTransform} systematically outperformed other methods and, even though it was developed to be applied to data with unique molecular identifiers (UMI), it also performed fairly well with the Smart-seq protocol (Koh and Kumar datasets).
 
@@
 To validate the previous assay on the proportion of variance/deviance explained by real populations, we included genes that maximized these two latter measures.
-Interestingly, while these selections were on average the top-ranking methods, they were not systematically best for all datasets.
-The ARI at true number of clusters was near perfect (>0.94) among the smallest datasets but among bigger datasets (simMix and ZhengMix) the performance of the methods was lower and more variable (Figure \ref{fig:selection}B).
+Interestingly, while these selections were on average the top-ranking methods, they were not systematically best for all datasets (Figure \ref{fig:selection}B).
 Non-standardized measures of variability, including mere expression level, tended to outperform more complex metrics.
 In general, we found deviance and unstandardized estimates of variance to provide the best results across datasets and normalization methods.
-Increasing the number of features selected also systematically led to an increase in the accuracy of the clustering, typically plateauing after 4000 features (Supplementary Figure 22).
+Increasing the number of features selected also tended to lead to an increase in the accuracy of the clustering, typically plateauing after 4000 features (Supplementary Figure 22).
 
@@
 The impact of the choice of dimensionality reduction method was far greater than that of normalization or feature selection (Supplementary Figure 23).
-Again, the clustering on the biggest datasets (simMix and Zhengmix) was more challenging for most methods.
-Several of them could not identify the true number of clusters in the simMix1 and Zhengmix8eq datasets, which have the highest number of subpopulations, or lead to very low mean silhouette width and accuracy.
-Overall however, the clustering accuracy was higher for methods that include weighting or scaling.
-Among them, GLM-PCA tended to increase the average silhouette width of already well-defined subpopulations, but \textit{Seurat}'s PCA procedure however proved superior on all metrics.
+Overall, weighting the components by the variance their explain tended to increase silhouette width (Supplementary Figure 23A) and, in most datasets, clustering accuracy (Supplementary Figure 23B).
+GLM-PCA tended to increase the average silhouette width of already well-defined subpopulations, but \textit{Seurat}'s PCA procedure however proved superior on all metrics.
 Like GLM-PCA, \textit{scVI}'s Linear Decoder (LD) does not explicitly rely on normalized counts.
-Its performance was lower than a pipeline with non-weighted \textit{Seurat} PCA on all measures for datasets with more subpopulations (Supplementary Figure 15).
-Overall, weighting the principal components by their variance (as \textit{Seurat} does) had a positive impact on silhouette widths and ARI scores.
+Its performance was however the worst both in terms of silhouette width and clustering accuracy (Supplementary Figure 15), and it only performed well in some of the least challenging datasets (mixology10x5cl and simMix2).
 
@@
 
-Different approaches have been proposed to select the appropriate number of dimensions, from the visual identification of an 'elbow' (inflexion point) of the variance explained, to more complex algorithms.
-We evaluated the performance of dimensionality estimators implemented in the \textit{intrinsicDimension} package \cite{johnssonDimensionality2015}, as well as common procedures such as the `elbow` method (inflexion point in the variance explained by each component), some scRNAseq-specific methods such as the JackStraw procedure \cite{ChungJackstraw2015} or \textit{scran}'s \texttt{denoisePCA} \cite{LunScran2016}, and the recent application of Fisher Separability analysis \cite{AlberganteSepar2019}.
+Different approaches have been proposed to select the appropriate number of dimensions, from the visual identification of an inflexion point in the curve of the variance explained by each component (`elbow' method), to more complex algorithms.
+We evaluated the performance of dimensionality estimators implemented in the \textit{intrinsicDimension} package \cite{johnssonDimensionality2015}, as well as common procedures such as the `elbow` method, some scRNAseq-specific methods such as the JackStraw procedure \cite{ChungJackstraw2015} or \textit{scran}'s \texttt{denoisePCA} \cite{LunScran2016}, and the recent application of Fisher Separability analysis \cite{AlberganteSepar2019}.
 
-We compared the various estimates of dimensionality in their ability to retrieve the intrinsic number of dimensions in a dataset, based on Seurat's weighted PCA space.
-As a first approximation of the true dimensionality, we computed the variance in each principal component that was explained by the subpopulations, which sharply decreased after the first few components in most datasets (Figure \ref{fig:dim}A).
-Figure \ref{fig:dim}B shows the difference between the dimension estimates of the above methods and that based on the subpopulations (i.e.
-from Figure \ref{fig:dim}A).
+We compared the various estimates of in their ability to retrieve the intrinsic number of dimensions in a dataset, based on Seurat's weighted PCA space.
+As a first approximation of the true dimensionality, we computed the variance in each principal component (based on Seurat's weighted PCA space) that was explained by the subpopulations, which sharply decreased after the first few components in most datasets (Figure \ref{fig:dim}A).
+These truth-based estimates where then compared to those of the above dimensionality estimation methods (Figure \ref{fig:dim}B).
 Of note, the methods differ widely in view of their computing time (Figure \ref{fig:dim}B) and we saw no relationship between the accuracy of the estimates and the complexity of the method.
-Reasoning that over-estimating the number of dimensions is less problematic than under-estimating it, we kept the former methods for a full analysis of their impact on clustering (Figure \ref{fig:dim}C-D), when combined with \textit{SCTransform} or the standard \textit{Seurat} normalization.
-Although most methods performed well on the various clustering measures, the global maximum likelihood based on translated Poisson Mixture Model (\textit{maxLikGlobal}, using the 20 nearest neighbors, although estimates were relatively stable across decent values of \textit{k}; see Supplementary Figure 26) provided the dimensionality estimate that best separated the subpopulations (Figure \ref{fig:dim}C) and resulted in the best clustering accuracy (Figure \ref{fig:dim}D).
-This method systematically selected many more components than were associated with the subpopulations, suggesting that although these additional components appear individually uninformative, in combination they nevertheless contribute to classification.
-An implication is that the commonly-used elbow method most likely underestimates dimensionality.
+Reasoning that over-estimating the number of dimensions is less problematic than under-estimating it, we kept the former methods for a full analysis of their impact on clustering (Figure \ref{fig:dim}C-D), when combined with \textit{SCTransform} or standard \textit{Seurat} normalization.
+Although most methods performed well on the various clustering measures, the global maximum likelihood based on translated Poisson Mixture Model (\textit{maxLikGlobal}, using 10 or 20 nearest neighbors, although estimates were relatively stable across decent values of \textit{k}; see Supplementary Figure 26) provided the dimensionality estimate that best separated challenging subpopulations (Figure \ref{fig:dim}C) and tended to result in the best clustering accuracy (Figure \ref{fig:dim}D).
+This method systematically selected many more components than were associated with the subpopulations in Figure \ref{fig:dim}A, suggesting that although these additional components appear individually uninformative, in combination they nevertheless contribute to classification.
+An important implication is that the commonly-used elbow method most likely underestimates dimensionality.
 
@@
 
+\subsection*{Overview of interactions between steps}
+
+We selected the most important alternatives for each parameter (see methods) in order to explore their interactions.
+Analysis of variance (after correcting for baseline differences between datasets and the resolution parameter) attributed the largest proportion of variance in the main cluster metrics to the choice of dimensionality, followed by normalization, and very little to filtering (Supplementary Figure 28).
+Because the choice of dimensions impacts the number of clusters, we repeated this analysis of variance among clusterings that yielded the true number of clusters, and observed that although normalization was the most important parameter among these analyses, the choice of dimensionality came closely behind (Supplementary Figure 28B), suggesting that more attention should be paid to this relatively neglected step.
+
+We next investigated whether the benefits of some methods where conditional on choices at other steps.
+Although analysis of variance identified many weak but significant interactions, in general we found that the benefits brought by important methods were robust to differences in other parts of the pipeline (Supplementary Figures 29-30).
+For instance, we confirmed that doublet removal and \textit{SCtransform} normalization were beneficial independently of other steps, and that, again indepdendently of other steps, more stringent filtering did not yield any major improvement over lenient filtering (Supplementary Figure 29).
+Similarly, deviance-based feature selection outcompeted \textit{Seurat}'s default selection method independently of changes at other steps (Supplementary Figure 30).
+
 \subsection*{Further extensions to the pipeline: imputation/denoising}
@@
 Since preliminary analysis showed that all methods performed equally well or better on normalized data, we applied them after filtering and normalization, but before scaling and reduction.
-Although some of the methods (e.g., \textit{DRImpute\_process} and \textit{alra\_norm}) did improve the separability of some more elusive subpopulations, no method had a systematically positive impact on the average silhouette width across all subpopulations (Supplementary Figure 28A).
-When restricting ourselves to clustering analyses that yielded the `right' number of clusters, all tested methods improved classification  compared to a scenario with no imputation step (`none' label, Supplementary Figure 28B).
+Although some of the methods (e.g., \textit{DRImpute\_process} and \textit{alra\_norm}) did improve the separability of some more elusive subpopulations, no method had a systematically positive impact on the average silhouette width across all subpopulations (Supplementary Figure 31A).
+When restricting ourselves to clustering analyses that yielded the `right' number of clusters, all tested methods improved classification  compared to a scenario with no imputation step (`none' label, Supplementary Figure 31B).
 However, the situation was not so straightforward with alternative metrics, where some methods (e.g., \textit{ENHANCE}) consistently underperformed.
@@
 
-To illustrate the applicability of \textit{pipeComp} to a completely different setting, we benchmarked the impact of surrogate variable analysis (SVA) -based methods on bulk RNAseq differential expression analysis (DEA).
+To illustrate the applicability of \textit{pipeComp} to a completely different problem, we benchmarked the impact of surrogate variable analysis (SVA) -based methods on bulk RNAseq differential expression analysis (DEA).
 The rationale behind such approaches is that the identification of factors determining genome-wide variation (i.e.
@@
 The package's `pipeComp\_dea' vignette details the creation process.
-We developed 3 benchmark datasets: 1) the \texttt{ipsc} data is a random selection of 10 vs 10 samples from \citet{carcamoOriveAnalysisTranscriptional2017}, which are very heterogeneous, and foldchanges were manually added to 300 genes; 2) the \texttt{seqc} dataset contains 5 vs 5 samples of two mixtures of the SEQC consortium with different ERCC spike-in mixes \citep{ComprehensiveAssessment2014}, and includes a partially correlated batch effect; 3) A **simulation** of 8 vs 8 samples with two technical vectors of variation (see methods for more details).
-As previously reported, we found that more stringent filtering improved DEA (Supplementary Figure 29A) and that the different DEA methods had similar performances (Figure \ref{fig:svadea}A and Supplementary Figure 29A).
-Including a SVA-based step increased sensitivity across all datasets, particularly in datasets with technical vectors of variation (\textit{seqc} and \textit{simulation}), and benefited all DEA methods in a similar fashion (Figure \ref{fig:svadea}A,C and Supplementary Figures 29-30).
+We developed 3 benchmark datasets: 1) the \texttt{ipsc} data is a random selection of 10 vs 10 samples from \citet{carcamoOriveAnalysisTranscriptional2017}, which are very heterogeneous, and foldchanges were manually added to 300 genes; 2) the \texttt{seqc} dataset contains 5 vs 5 samples of two mixtures of the SEQC consortium with different ERCC spike-in mixes \citep{ComprehensiveAssessment2014}, and includes a partially correlated batch effect; 3) A \texttt{simulation} of 8 vs 8 samples with two technical vectors of variation (see methods for more details).
+As previously reported, we found that more stringent filtering improved DEA (Supplementary Figure 32A) and that the different DEA methods had similar performances (Figure \ref{fig:svadea}A and Supplementary Figure 32A).
+Including a SVA-based step increased sensitivity across all datasets, particularly in datasets with technical vectors of variation (\textit{seqc} and \textit{simulation}), and benefited all DEA methods in a similar fashion (Figure \ref{fig:svadea}A,C and Supplementary Figures 32-33).
 With the exception of the \texttt{RUVr} method, which led to a high False Discovery Rate (FDR), all other methods showed comparable gains in sensitivity while maintaining a decent error control (if a little above the nominal FDR), even when increasing the number of surrogate variables used (Figure \ref{fig:svadea}B-C).
@@
 
-On the basis of our findings, we can make a number of practical analysis recommendations, also summarized in Supplementary Figure 31:
+On the basis of our findings, we can make a number of practical analysis recommendations, also summarized in Supplementary Figure 34:
 
@@
    \begin{itemize}
-       \item Denoising appeared beneficial to the identification of subpopulations in 10x datasets, but not in Smart-seq datasets.
-       \item We found especially \textit{ALRA} (with prior normalization), \textit{DrImpute} (with prior processing) and \textit{DCA} to offer the best performances.
+       \item Denoising appeared beneficial to the identification of some subpopulations in 10x datasets, but not in Smart-seq datasets.
+       \item We found especially \textit{ALRA} (with prior normalization), \textit{DrImpute} (with prior processing) and \textit{DCA} to offer the best performances, although none of these methods was beneficial over all subpopulations.
    \end{itemize}
@@
      \item Similarly to previous reports \cite{SunDimRed2019}, we recommend the irlba-based PCA using weighting of the components, as implemented in \textit{Seurat}. 
-     \item We advise against the commonly-used elbow method (because it is too conservative) and the jackstraw method (low performance at great computational costs) for deciding on the components to include, and recommend a method like the \textit{global maximum likelihood based on translated Poisson Mixture Model} method (e.g., implemented in the  \texttt{maxLikGlobalDimEst} function of \texttt{intrinsicDimension}, in our case using the 20 nearest neighbors).
+     \item We advise against the commonly-used elbow method (because it is too conservative) and the jackstraw method (low performance at great computational costs) for deciding on the components to include, and recommend a method like the \textit{global maximum likelihood based on translated Poisson Mixture Model} method (e.g., implemented in the  \texttt{maxLikGlobalDimEst} function of \texttt{intrinsicDimension}, in our case using the 10 or 20 nearest neighbors).
    \end{itemize}
@@
 The ZhengMix8eq and simMix1 datasets were the most challenging datasets across all preprocessing steps that we tested.
-The average silhouette width of half of their subpopulations was systematically lower that the other subpopulations/ datasets and the ARI score at true number of clusters was always poor or unable to achieve.
-In comparison to the performance on other datasets, this inconsistency was particularly striking when evaluating the dimensionality estimation dimensionality reduction and the clustering methods, for which half of the evaluated methods could not lead to the true number of clusters.
+The average silhouette width of half of their subpopulations was systematically lower that the other subpopulations/ datasets and the ARI score at true number of clusters, if ever achieved, was always poor.
 It is likely that these two datasets are challenging due to their combined sparsity and complexity (see Supplementary Figure 1).
-This issue that was already observed for the Zhengmix8eq dataset in a previous benchmark \cite{duoClustering2018}, as it contains deeper levels of annotations for subpopulations of T-cells.
-The Koh dataset harbors as many populations, but originates from more distinct groups (from hESC in different differentiation stages).
-As the Kumar dataset, it harbors a very low subpopulations complexity and most methods applied on them performed well.
+The challenge of distinguishing subpopulations of T-cells in the Zhengmix8eq dataset was previously discussed \cite{duoClustering2018}.
+In contrast, the Koh dataset harbors as many populations, but originates from more distinct groups (from hESC in different differentiation stages), while the Kumar and mixology datasets harbor a very low subpopulations complexity and most methods applied on them performed well.
 
@@
 Instead, the \textit{CellBench} package was recently proposed to address a similar need \cite{suCellbench}, offering an elegant piping syntax to combine alternative methods at successive steps.
-A key additional feature of \textit{pipeComp} is its ability to perform evaluation in parallel to the pipeline and offering on-the-fly evaluation at any step.
+A key additional feature of \textit{pipeComp} is its ability to perform evaluation in parallel to the pipeline and offering on-the-fly evaluation at multiple steps.
 This avoids the need to store potentially large intermediate data for all possible permutations and thus allowing a combinatorial benchmark.
+This, along with the set-up simplicity (demonstrated here by our separate SVA-DEA benchmark), also distinguishes \textit{pipeComp} from general workflow frameworks such as \textit{drake}\citep{landauDrakePackage2018}; indeed we estimate that the benchmarks discussed here would have taken approximately 16TB of storage had all non-redundant intermediates been saved (see the Supplementary methods for more detailed explanations of the framework and comparison to \textit{CellBench} and \textit{drake}).
 In addition, the fact that benchmark functions are stored in the \texttt{PipelineDefinition} makes the benchmark more smoothly portable, making it easy to modify, extend with further methods, or apply to other datasets.
@@
 \textit{pipeComp} is a flexible R framework for evaluating methods and parameter combinations in a complex pipeline, computing on-the-fly multilevel evaluation metrics.
-Applying this framework to scRNAseq clustering enabled us to make clear recommendations on the steps of filtering, normalization, feature selection, denoising, dimensionality reduction and clustering (Supplementary Figure 31).
+Applying this framework to scRNAseq clustering enabled us to make clear recommendations on the steps of filtering, normalization, feature selection, denoising, dimensionality reduction and clustering (Supplementary Figure 34).
 We demonstrate how a diversity of multilevel metrics can be more robust, more sensitive, and more nuanced than simply evaluating the final clustering performance.
-We hope that the \textit{pipeComp} framework can be applied to extend the current field of benchmarking, as well as to investigate pipelines other domains with multi-step computational analyses.
+We further showed that the \textit{pipeComp} framework can be applied to extend the current field of benchmarking, as well as to investigate pipelines other domains with multi-step computational analyses.
 
@@
 
+\subsection*{Overall run for interactions across steps}
+
+The set of alternatives found for the overall run (Supplementary Figures 28-30) can be found in the \href{https://github.com/markrobinsonuzh/scRNA_pipelines_paper/blob/master/data/overall_run.Rmd}{paper's repository}.
+The analysis of variance was performed with the following model:
+
+\texttt{metric$\sim$ dataset*resolution+doubletmethod+filt+norm+clustmethod+dims}
+
 \subsection*{SVA-DEA benchmark}
@@
 \end{itemize}
+
+For \textit{vstsva}, we first applied the variance-stabilizing transformation from \textit{DESeq2}\citep{loveModeratedEstimation2014} before applying \texttt{sva::sva}.
+\textit{edgeR.QLF} refers to the quasi-likelihood version \citep{lundDetectingDifferential2012} of \textit{edgeR}'s fit.
+The exact wrappers used can be found in the package at \url{https://github.com/plger/pipeComp/blob/master/inst/extdata/dea\_wrappers.R}.
 
@@
 \textbf{D:} Clustering accuracy across the normalization/dimensionality estimation methods.
-The color-mapping scheme is the same as in Figure \ref{fig:norm}}
+The color-mapping scheme is the same as in Figure \ref{fig:norm}C}
     \label{fig:dim}
@@
     \caption{\textbf{Evaluation of clustering methods.} Clustering accuracy (\textbf{A}) and run time (\textbf{B}) of \textit{scran}- and \textit{Seurat}-based clustering tools in combination with \textit{SCTransform} and standard \textit{Seurat}'s normalization.
-The color-mapping scheme is the same as in Figure \ref{fig:norm}}
+The color-mapping scheme in panel \textbf{A} and the left of panel \textbf{B} is the same as in Figure \ref{fig:norm}C.}
     \label{fig:clust}

### 2020-05-12 main_figures/dimensionality.Rmd
updated color mapping

@@
 )
-p3 <- grid.grabExpr(draw(p3))
+p3 <- grid.grabExpr(draw(p3, annotation_legend_list=.scaledLegend(), merge_legend=TRUE, heatmap_legend_side="bottom"))
 ```

### 2020-05-12 main_figures/filtering.Rmd
updated color mapping

--- 2020-03-07 main_figures/filtering.Rmd
@@
   library(SingleCellExperiment)
-  library(pipeComp)
+  #library(pipeComp)
 })
+devtools::load_all("../../pipeComp/")
 theme_set(theme_cowplot(font_size = 12))
@@
 res <- readRDS("../data/results_filtering.rds")
-m <- scrna_evalPlot_filtering(res, returnTable=TRUE)
-m$doubletRemoval <- !grepl("doubletmethod=none;", m$method)
-m$filter <- factor(sapply(strsplit(as.character(m$method),"\\.|="), FUN=function(x) rev(x)[1]),
-                     c("nofilter","lenient","default","pca","pca2","stringent","veryStringent"))
-cols <- c("black", pipeComp::getQualitativePalette(length(levels(m$filter))-1))
-names(cols) <- levels(m$filter)
 
-# ggplot(m, aes(maxPCout, meanF1, group=method, colour=filter, shape=doubletRemoval)) + geom_point(size=3) + facet_wrap(~dataset, scales="free") + scale_color_manual(values=cols) + xlab("Max proportion of subpopulation excluded")
-
-p2 <- ggplot(m, aes(maxPCout, F1atK, group=method, colour=filter, shape=doubletRemoval)) + geom_point(size=3) +
-  # geom_line(aes(group=doubletRemoval), colour="lightgrey") + 
-  facet_wrap(~dataset, scales="free") + scale_color_manual(values=cols) + 
-  xlab("Maximum proportion of cells excluded per subpopulation") + ylab("mean F1 at true number of clusters")
+p2 <- scrna_evalPlot_filtering(res, atNearestK=TRUE, alpha=0.8) + 
+  ylab("Mean F1 at best number of clusters")
 

### 2020-05-12 main_figures/normalization.Rmd
updated color mapping

@@
                 row_split=classifyNorms(norm), name="ARI at true k", title="ARI at true k")),
-  heatmap_legend_side = "bottom", merge_legend=TRUE, annotation_legend_list=lgd))
+  heatmap_legend_side = "bottom", merge_legend=TRUE, annotation_legend_list=.scaledLegend()))
 ```

### 2020-05-12 main_figures/svadea.Rmd
updated color mapping

@@
 })
+source("../misc_functions.R")
 theme_set(theme_cowplot(font_size = 11))
@@
 
-hts <- dea_evalPlot_sig(res, what=c("logFC.pearson","TPR"), agg.by=c("sva.method","k"), row_split="sva.method", 
-                     anno_legend=FALSE, show_column_names=TRUE, font_factor=0.85, scale="colCenterScale") +
-  dea_evalPlot_sig(res, what="FDR", agg.by=c("sva.method","k"), row_split="sva.method", 
+hts <- evalHeatmap( res, step="dea", what="logFC.pearson", agg.by=c("sva.method","k"), 
+                    row_split="sva.method", anno_legend=FALSE, show_column_names=TRUE, 
+                    font_factor=0.85, scale="colCenterScale" ) +
+  evalHeatmap( res, step="dea", what="TPR", agg.by=c("sva.method","k"), 
+                    row_split="sva.method", anno_legend=FALSE, show_column_names=TRUE, 
+                    font_factor=0.85, scale="colCenterScale", filterExpr=threshold==0.05 ) +
+  evalHeatmap(res, step="dea", what="FDR", agg.by=c("sva.method","k"), row_split="sva.method", 
                      anno_legend=FALSE, show_column_names=TRUE, font_factor=0.85, scale="colCenterScale",
-                     col=rev(viridisLite::inferno(100)), value_cols=c("white","black"))
-lgd <- Legend(col_fun=colorRamp2(1:100, viridisLite::inferno(100)), at=c(1,100), 
-              title="MADs", labels=c("worst","best"), direction="horizontal")
+                     col=rev(viridisLite::inferno(100)), value_cols=c("white","black"), filterExpr=threshold==0.05)
 hts <- pipeComp:::.renameHrows(hts, f = function(x) gsub("none > 1", "none",x, fixed=TRUE))
-#draw(hts, heatmap_legend_side = "bottom", merge_legend=TRUE, annotation_legend_list=lgd)
+#draw(hts, heatmap_legend_side = "bottom", merge_legend=TRUE, annotation_legend_list=.scaledLegend())
 
@@
   grid.grabExpr(draw( hts, heatmap_legend_side = "bottom", merge_legend=TRUE, 
-                      annotation_legend_list=lgd)),
+                      annotation_legend_list=.scaledLegend())),
   nrow=1, rel_widths=c(5,2.2), labels=c(NA,"C")

### 2020-05-12 supplementary_figures/supp2_filtering.Rmd
updated supp

@@
 
-```{r filt_more, fig.width=8, fig.height=9}
+```{r filt_more, fig.width=8.5, fig.height=5}
 res <- readRDS("../data/results_filtering.rds")
-m <- old_evalPlot_filtering(res, returnTable=TRUE)
-m <- m[grep("veryStringent", m$method, invert=TRUE),]
+m <- pipeComp:::.mergeFilterOut(res$evaluation[1:2])
+m <- m[grep("veryStringent",m$filt,invert=TRUE),]
+res$evaluation$clustering <- res$evaluation$clustering[grep("veryStringent",res$evaluation$clustering$filt,invert=TRUE),]
+ag <- aggregate(m$pc.lost, m[,c("dataset","doubletmethod","filt")], FUN=max)
+colnames(ag)[4] <- "maxLost"
+ag$medianLost <- aggregate(m$pc.lost, m[,c("dataset","doubletmethod","filt")], FUN=median)$x
 
-# # horizontal version
-# m <- m[grep("veryStringent", m$method, invert=TRUE),]
-# m.f1 <- reshape2::dcast(m, method~dataset, value.var="F1atK")
-# n <- sapply(strsplit(gsub("=",".",m.f1[,1]),"\\."),FUN=function(x) rev(x)[1])
-# n[grep("=doublet", m.f1[,1])] <- paste0("doubletRemoval+",n[grep("=doublet", m.f1[,1])])
-# m.pc <- reshape2::dcast(m, method~dataset, value.var="maxPCout")[,-1]
-# m.f1m <- reshape2::dcast(m, method~dataset, value.var="meanF1")[,-1]
-# m.pcm <- reshape2::dcast(m, method~dataset, value.var="medianPCout")[,-1]
-# m.f1 <- m.f1[,-1]
-# row.names(m.f1m) <- row.names(m.pcm) <- row.names(m.f1) <- row.names(m.pc) <- n
-# 
-# chm(list("mean F1"=m.f1m, "F1 at true \n# clusters"=m.f1, "max % lost per\nsubpopulation"=m.pc, "median % lost per\nsubpopulation"=m.pcm), scale = "column", colors = viridis::inferno(3), value_format = c("%.2f","%.2f","%.0f","%.0f"), scaleFn=list(scale,scale, scale2,scale2), cluster_columns = FALSE,  row_sortFn = function(x, na.rm){ sum(x[1:18],na.rm=TRUE)-median(x[19:36], na.rm=TRUE)/100})
-
-mycast <- function(x, val, form=dataset~method){ 
-  x <- reshape2::dcast(x, form, value.var=val)
-  row.names(x) <- x[,1]
-  as.matrix(x[,-1])
+renameHrows(evalHeatmap(ag, what=c("maxLost","medianLost"), agg.by=c("doubletmethod","filt"), 
+            scale=TRUE, row_split=doubletmethod, reorder_rows=TRUE, value_format="%.f",
+            title = c("max % lost per\nsubpopulation","median % lost per\nsubpopulation"), 
+            font_factor = 0.75, show_column_names = TRUE, anno_legend = FALSE) + 
+  evalHeatmap(res, step="clustering", what=c("mean_F1"), agg.by=c("doubletmethod","filt"), title="mean F1", 
+              font_factor = 0.75, show_column_names=TRUE, anno_legend=FALSE) + 
+  evalHeatmap(res, step="clustering", what=c("mean_F1"), agg.by=c("doubletmethod","filt"), 
+              filterExpr = n_clus==true.nbClusts, font_factor = 0.75, show_column_names=TRUE, 
+              anno_legend=FALSE, name="F1atK", title="mean F1\nat true K"),
+  function(x){
+    x <- gsub("^doublet\\.","",x)
   }
-m.f1 <- mycast(m, "F1atK")
-n <- sapply(strsplit(gsub("=",".",colnames(m.f1)),"\\."),FUN=function(x) rev(x)[1])
-n[grep("=doublet", colnames(m.f1))] <- paste0("doubletRemoval+",n[grep("=doublet", colnames(m.f1))])
-m.pc <- mycast(m, "maxPCout")
-colnames(m.f1) <- colnames(m.pc) <- n
-o <- order(colMeans(t(scale(t(m.f1))), na.rm=TRUE)*100-colMeans(t(scale(t(m.pc))), na.rm=TRUE))
-m.f1 <- m.f1[,o]
-m.pc <- m.pc[,o]
-m.f1m <- mycast(m, "meanF1")[,o]
-m.pcm <- mycast(m, "medianPCout")[,o]
-colnames(m.f1m) <- colnames(m.pcm) <- colnames(m.f1)
-
-names(nn) <- nn <- c("m.f1m","m.f1","m.pc","m.pcm")
-cellfns <- lapply(nn, FUN=function(mname){
-  if(grepl("f1",mname)){
-    sp <- paste0('sprintf("%.1f", 100*',mname,'[i, j])')
-  }else{
-    sp <- paste0('sprintf("%.1f", ',mname,'[i, j])')
-  }
-  eval(parse(text=paste0('function(j, i, x, y, width, height, fill){',
-      'lab <- ', sp, '; grid.text(lab, x, y, gp = gpar(fontsize = 10)) }')))
-})
-
-names(nn) <- c("mean F1", "F1 at true \n# clusters", "max % lost per\nsubpopulation", "median % lost per\nsubpopulation")
-hs <- lapply(seq_along(nn), FUN=function(i){
-  mat <- get(nn[[i]])
-  if(grepl("f1",nn[[i]])){
-    mat <- t(scale(t(mat)))
-  }else{
-    mat <- t(scale(t(sqrt(mat))))
-  }
-  Heatmap(mat, row_title=names(nn)[[i]], col=viridis::inferno(100), cell_fun = cellfns[[i]], show_heatmap_legend = FALSE, cluster_columns = FALSE, cluster_rows = FALSE)
-})
-
-hs[[1]] %v% hs[[2]] %v% hs[[3]] %v% hs[[4]]
+)
 ```
@@
 Mean clustering F1 score per subpopulation, mean F1 at true number of clusters, as well as maximum and median proportion of excluded cells per subpopulation across various filtering strategies.
-Doublet removal generally improves clustering accuracy with very mild exclusion rates, even in datasets that do not have heterotypic doublets.
+Doublet removal generally improves clustering accuracy with relatively mild increases exclusion rates, even in datasets that do not have heterotypic doublets.
 Stringent distribution-based filtering creates large cell type biases.
@@
 
-```{r}
-res <- readRDS("../data/selection_old.rds")
-res <- res[grep("deviance|seurat.variance.standardized",res$sel),]
-res$filt <- as.factor(res$filt)
-levels(res$filt) <- c("coding only", "no mitochondrial", "no ribosomal", "all")
-res2 <- res[which(abs(res$nbClusters-truek[res$dataset])<=1),]
-ll <- lapply(split(res2, res2$norm), FUN=function(x){
-  m2 <- cast2(x, formula=filt~dataset, fun.aggregate=median, value.var="ARI")
-})
-chm(ll, scale="column", scaleTitle="z-score", sameScale=TRUE, value_format="%.2f", row_sortFn=function(x,...){ median(x,...) }, cluster_columns=FALSE)
+```{r, fig.width=8, fig.height=6.5}
+res <- readRDS("../data/results_featexcl.rds")
+rnfn <- function(x){
+  x <- gsub(" > norm.seurat","",x)
+  x <- gsub(" > norm.sctransform"," > sctransform",x)
+  x <- gsub("coding","coding-only",x)
+  x <- gsub("ribo","-ribo",x)
+  gsub("Mt","-Mt",x)
+}
+draw(renameHrows(
+  evalHeatmap(res, "clustering", what="MI", agg.by=c("classes","norm"), 
+              row_split=norm, reorder_rows=TRUE, show_column_names=TRUE,
+              anno_legend=FALSE, title="Mutual information") +
+    evalHeatmap(res, "clustering", what="ARI", agg.by=c("classes","norm"), 
+                filter=n_clus==true.nbClusts, title="ARI at true K",
+                show_column_names=TRUE, anno_legend=FALSE),
+  rnfn), annotation_legend_list=pipeComp:::.scaledLegend(), merge_legend=TRUE, heatmap_legend_side="bottom")
     ```
@@
 
-Impact of restricting the type of features used on the ARI of the clustering.
+Impact of restricting the type of features used on the Mutual Information (MI, left) and Adjusted Rand Index (ARI, right) of the clustering.
+`all` indicates that all features were used, `-Mt` stands for the exclusion of mitochondrial genes, `-ribo` the exclusion of ribosomal genes, and `coding-only` a restriction to protein-coding genes.
+The features were filtered out prior to normalization.
+
 \newpage

### 2020-05-12 supplementary_figures/supp3_normalization.Rmd
updated supp

@@
 res <- readRDS("../data/results_norm.rds")
-renameNorm(evalHeatmap(res, step="dimreduction", c("log10_total_counts","log10_total_features"), what2="meanAbsCorr.covariate2", agg.by = "norm", anno_legend=FALSE, row_split=classifyNorms(norm)))
+renameNorm(evalHeatmap(res, step="dimreduction", c("log10_total_counts","log10_total_features"), what2="meanAbsCorr.covariate2", agg.by="norm", anno_legend=FALSE, scale=FALSE, row_split=classifyNorms(norm)))
 ```

### 2020-05-12 misc_functions.R

@@ -78,7 +78,7 @@ reduceMethods <- function(x, pos=1:2, FUN=mean, sep=" "){
   x[,-1]
 }
 renameNorm <- function(x){
-  pipeComp:::.renameHrows(x, f=function(x){
+  renameHrows(x, f=function(x){
     x <- gsub("^norm\\.", "", x)
     x <- gsub("seuratvst","sctransform",x)
     x <- gsub(".noscale", " (no scaling)", x)
@@ -329,4 +329,39 @@ old_evalPlot_filtering <- function(res, steps=c("doublet","filtering"), returnTa
   ggplot(m, aes(maxPCout, meanF1, group=method, colour=method)) + geom_point(size=3) + 
     facet_wrap(~dataset, scales="free") + xlab("Proportion") + 
     xlab("Max proportion of subpopulation excluded")
+}
+
+.scaledLegend <- function(){
+  ComplexHeatmap::Legend(
+    col_fun=circlize::colorRamp2(1:100, viridisLite::inferno(100)), 
+    at=c(1,50,100), title="MADs", labels=c("worst","median","best"), 
+    direction="horizontal", title_position = "lefttop" )
+}
+
+# legacy...
+.renameHrows <- function(h, f) renameHrows(h,f)
+#' renameHrows
+#'
+#' Applies a function to the row labels of a Heatmap or HeatmapList
+#'
+#' @param h A `Heatmap` or `HeatmapList`
+#' @param f The function to apply on row labels.
+#'
+#' @return the modified `h`
+#' @export
+#'
+#' @examples
+#' data("exampleResults", package="pipeComp")
+#' H <- evalHeatmap(exampleResults, what=c("ARI", "MI"), agg.by="norm")
+#' H
+#' H <- renameHrows(H, function(x) gsub("^norm\\.", "", x))
+#' H
+renameHrows <- function(h, f=identity){
+  if(is(h,"HeatmapList")){
+    h@ht_list <- lapply(h@ht_list, f=f, FUN=.renameHrows)
+    return(h)
+  }
+  h@row_names_param$anno@var_env$value <- 
+    f(h@row_names_param$anno@var_env$value)
+  h
 }

### 2020-05-12 supplementary_figures/supp_framework/supp_framework_explanations.Rmd
supplementary framework information

---
title: "pipeComp - Supplementary Methods"
author: Pierre-Luc Germain, Anthony Sonrel, Mark D. Robinson
output:
  pdf_document:
    toc: true
    number_sections: true
    toc_depth: 2
urlcolor: blue
---

\vfill

Supplementary methods to:

_pipeComp, a general framework for the evaluation of computational pipelines, reveals performant single-cell RNA-seq preprocessing tools_<br/>

\newpage

```{r, include=FALSE}
library(BiocStyle)
```

# Introduction

This document aims at describing in more detail the `r Rpackage("pipeComp")` benchmarking framework. As it represents a snapshot (`r Rpackage("pipeComp")` version `r paste0(packageVersion("pipeComp"), ", ", format(Sys.time(), '%d %B, %Y'))`), the most up-to-date information can be found on the [github repository](https://github.com/plger/pipeComp).

`r Rpackage("pipeComp")` is especially suited to the benchmarking of pipelines that include many steps/parameters, enabling the exploration of combinations of parameters and of the robustness of methods to various changes in other parts of a pipeline. It is also particularly suited to benchmarks across multiple datasets. It is entirely based on _R_/Bioconductor, meaning that methods outside of _R_ need to be called via _R_ wrappers. `r Rpackage("pipeComp")` handles multithreading in a way that minimizes re-computation and duplicated memory usage, and computes evaluation metrics on the fly to avoid saving many potentially large intermediate files, making it well-suited for benchmarks involving large datasets.

<br/><br/><br/>

# _pipeComp_ overview and main inputs

`r Rpackage("pipeComp")` requires three main inputs, through which we will go in order:

* a `PipelineDefinition` object, which specifies the basic pipeline structure and evaluation metrics
* a list of alternative parameter values
* benchmark datasets

## _PipelineDefinition_ objects

```{r figure1, echo = FALSE, out.width="0.75\\linewidth", fig.align="center", fig.cap = "Overview of the pipeComp framework. A: 'PipelineDefinition' objects include general pipeline structures, concentrated around consecutive steps, as well as all evaluation metrics to be computed after any of these steps. B: Given a 'PipelineDefinition', a list of alternative parameters and a list benchmark datasets, the 'runPipeline' function proceeds through combinations arguments, avoiding recomputing the same step twice and computing evaluations on the fly."}
knitr::include_graphics("framework.pdf")
```

`r Rpackage("pipeComp")` is built around the S4 class `PipelineDefinition` which encapsulates a basic abstract workflow (different steps and their parameters) along with pre-defined evaluation metrics to be computed, as depicted in Figure  \ref{fig:figure1}A. When a pipeline is executed, each step is consecutively applied to a starting object. In addition, each step can optionally have evaluation functions, which take as input the output of the main step function, and generate associated evaluation metrics. These metrics are automatically aggregated across analyses and datasets if they have standard formats (e.g. vectors, data.frames or lists thereof), or otherwise custom aggregation functions can be specified.

Each step of a `PipelineDefinition` can have any number of parameters, which can take any scalar values (e.g. character, numeric, logical). Alternative methods for a step can therefore be passed as the name of wrapper functions which need to be prepared in advance. While the `PipelineDefinition` defines the different parameters, the alternatives parameter values are not contained in the object itself, which enables the use of the same `PipelineDefinition` to run several benchmarks. To facilitate this, however, `PipelineDefinition` objects can also hold information about default parameter values, thereby enabling users to run benchmarks without having to define every single parameter.

The packages comes with two different pre-built `PipelineDefinition`s: the single-cell RNAseq clustering pipeline (in two flavours, see `?scrna_pipeline`) and the SVA-DEA pipeline (see `?dea_pipeline`), each with their in-built evaluation functions. For more information about building (or editing) a `PipelineDefinition` object, refer to the examples in the following vignettes:

```{r, eval=FALSE}
vignette("pipeComp", package = "pipeComp")
vignette("pipeComp_dea", package = "pipeComp")
```

## Alternative parameter values

Each pipeline step defined in the `PipelineDefinition` can have any number of parameters, which can take any scalar values (i.e. character, numeric or logical vectors of length 1), as illustrated in Figure \ref{fig:figure1}B. If a `PipelineDefinition` was built in a way that expects it, distinct alternative methods for a step can be passed as the name of wrapper functions, which need to be prepared in advance and to be loaded in the environment in which `runPipeline` (see below) will be run.

For example:

```{r, eval=FALSE, echo=TRUE}
function_A <- function(x) 1+sqrt(x)
function_B <- function(x) log1p(x)
alternatives <- list(
  par1=c("function_A", "function_B")
)
```

## Benchmark datasets

`r Rpackage("pipeComp")` was especially designed to run parallel benchmarks across multiple datasets. There is no a priori restriction on the format of benchmark datasets, so long as the `PipelineDefinition` and wrappers can deal with them. The datasets objects should contain any data that is needed for the pipeline to run, as well as any information that is needed for the evaluation of the results (e.g. their own truth, such as true cell labels for the scRNAseq application, or true differential expression labels for the SVA-DEA application).

<br/><br/><br/>

# The `runPipeline` function.

Once the three main inputs are available, the actual processing can be launched with the `runPipeline` function, e.g.:

```{r, eval=FALSE}
res <- runPipeline( datasets, alternatives, PipelineDefinition,
                    output.prefix="folder/prefix", nthreads=3 )
```

The function's processing is illustrated in Figure \ref{fig:figure1}B. First, a matrix of parameter combinations is created (it can alternatively be manually provided/edited to run only subsets of possible combinations). Multithreading is handled through `r Biocpkg("BiocParallel")`, by first splitting datasets across threads, and then splitting the parameter matrix in a way that minimizes re-computation. In each thread, the parameter combinations are executed in a branching fashion (Figure \ref{fig:figure1}B) so that, if combinations share the same first steps, these are computed only once. Evaluation is computed in the thread immediately after step processing.

<br/><br/><br/>

# Comparison with other (_R_) tools

## _drake_

`r Rpackage("pipeComp")` is _not_ a general pipeline/workflow framework like, for instance, `r CRANpkg("drake")`, which boasts an impressive array of features (object histories and dependencies, caching and recovery, etc.). Instead, `r Rpackage("pipeComp")` was specifically designed for benchmarking: it makes benchmarks easily deployable, modifiable and extendable by encapsulating together abstract pipeline structures and evaluation procedures, and facilitates running many combinations of step parameters through hierarchical branching.

The same branching pattern could in principle be done with `r CRANpkg("drake")`, however in a much more cumbersome fashion. Indeed, planning a static or dynamic `r CRANpkg("drake")` branching on all combinations would result in a target for each (and hence many duplicated computation) unless it was planned for each step separately, requiring considerably more scripting than when implemented in `r Rpackage("pipeComp")`. In addition, it would mean that all intermediate targets need to be stored, which is not practicable for combinatorial explorations with large datasets (for our scRNAseq comparison, we estimate this to be approximately 16 TB).

On the other hand, a `r CRANpkg("drake")`-based workflow has several advantages, in particular the capacity, upon failure of a pipeline (or changes made to a step), to recover and restart from the very next upstream target. We are currently working on an `bptry`-based approach to error handling and selective recomputation, but for the moment users requiring smooth error recovery should rather rely on `r CRANpkg("drake")`.

## _CellBench_

The `r Biocpkg("CellBench")` package was recently proposed for single-cell benchmarking, offering an elegant piping syntax to combine alternative methods at successive steps. Both packages have similar functionalities, although implemented very differently. For instance, `r Biocpkg("CellBench")` uses on-disk caching to avoid recomputing steps, which has the same drawback as `r CRANpkg("drake")` for large datasets / number of combinations. `r Rpackage("pipeComp")` instead keeps computed intermediates in memory no longer than is strictly necessary to avoid recomputation. In addition, because the benchmark is, in `r Biocpkg("CellBench")`, simply a function piped after the pipeline steps, it does not allow the multi-step evaluation central to `r Rpackage("pipeComp")`.

### 2020-05-12 supplementary_figures/supp_framework/supp_framework_explanations.Rmd
fixed typo

@@
 `r Rpackage("pipeComp")` is _not_ a general pipeline/workflow framework like, for instance, `r CRANpkg("drake")`, which boasts an impressive array of features (object histories and dependencies, caching and recovery, etc.).
-Instead, `r Rpackage("pipeComp")` was specifically designed for benchmarking: it makes benchmarks easily deployable, modifiable and extendable by encapsulating together abstract pipeline structures and evaluation procedures, and facilitates running many combinations of step parameters through hierarchical branching.
+Instead, `r Rpackage("pipeComp")` was specifically designed for benchmarking: it makes benchmarks easily deployable, modifiable and extensible by encapsulating together abstract pipeline structures and evaluation procedures, and facilitates running many combinations of step parameters through hierarchical branching.
 

### 2020-05-12 bmc_article.tex

@@
 
-Here, we develop a flexible R framework for pipeline comparison and apply it to the evalution of the various steps of analysis leading from an initial count matrix to a cluster assignment, which are critical in a wide range of applications.
+Here, we develop a flexible R framework for pipeline comparison and apply it to the evaluation of the various steps of analysis leading from an initial count matrix to a cluster assignment, which are critical in a wide range of applications.
 We collected real datasets of known cell composition (Table \ref{tab:table1}) and used a variety of evaluation metrics to investigate in a multilevel fashion the impact of various parameters and variations around a core scRNAseq pipeline.
@@
 In addition, optional benchmark functions can be set for each step to provide standardized, multi-layered evaluation metrics.
-Given such a \texttt{PipelineDefinition} object, a set of alternative parameters (which might include different subroutines) and benchmark datasets, the \texttt{runPipeline} function then proceeds through all combinations of arguments (or a desired subset), avoiding recomputing the same step twice without the need to save all alternative intermediates, and compiling evaluations (including run time) on the fly (see Supplementary methods for more detail).
+Given such a \texttt{PipelineDefinition} object, a set of alternative parameters (which might include different subroutines calling different methods) and benchmark datasets, the \texttt{runPipeline} function then proceeds through all combinations of arguments (or a desired subset), avoiding recomputing the same step twice without the need to save all alternative intermediates, and compiling evaluations (including run time) on the fly (see Supplementary methods for more detail).
 Variations in a given parameter can be evaluated using all metrics from this point downward in the pipeline.
@@
 
-We next investigated the impact of filtering according to several methods (see Methods for more details): four methods based on deviations to MADs with increasing levels of stringency (named \textit{lenient}, \textit{default}, \textit{stringent}, \textit{very stringent}) and two methods based on \textit{scater}'s \textit{runPCA} using all (\textit{pca.all}) or selected covariates (\textit{pca.sel}).
+We next investigated the impact of several filtering methods (see Methods for more details): four methods based on deviations to MADs with increasing levels of stringency (named \textit{lenient}, \textit{default}, \textit{stringent}, \textit{veryStringent}) and two methods based on \textit{scater}'s \textit{runPCA} using all (\textit{pca.all}) or selected covariates (\textit{pca.sel}).
 An important risk of excluding cells on the basis of their distance from the whole distribution of cells on some properties (e.g., library size) is that these properties tend to have different distributions across subpopulations.
@@
 Since filtering changes the relative abundance of the different subpopulations, global clustering accuracy metrics such as ARI are not appropriate here.
-We therefore calculated the per-subpopulation precision and recall using the Hungarian algorithm \cite{PapadimitriouHu1998} and monitored their mean F1 score.
-A first observation was that, although the stringent filtering tended to be associated with an increase in accuracy, it could also become deleterious.
+We therefore calculated the per-subpopulation precision and recall using the Hungarian algorithm \cite{PapadimitriouHu1998} and monitored the mean F1 score across subpopulations.
+A first observation was that, although more stringent filtering tended to be associated with an increase in accuracy, it tended to plateau and could also become deleterious.
 Most of the benefits could be achieved without very stringent filtering and minimizing subpopulation bias (Figure \ref{fig:filtering}B).
@@
 high library size), rather than low-quality cells.
-On the opposite, too lenient filtering (or lack thereof) can lead to misclassified cells, as was for instance the case especially with the Zhengmix4 datsets (Figure \ref{fig:filtering}B).
+On the opposite, too lenient filtering (or lack thereof) can lead to misclassified cells, as was for instance the case especially with the Zhengmix4 datasets (Figure \ref{fig:filtering}B).
 We found that a relatively mild filtering (`default' - see methods) provided a good trade-off, often leading to high clustering performance (e.g.
@@
 Koh and Zhengmix8eq).
+PCA-based selection did not appear superior to feature-wise filters.
 We found very little overlap between the cells excluded by doublet removal and those excluded by MAD-based filtering (Supplementary Figure 12B).
-Of note, doublet removal in conjunction with filtering resulted in a net decrease in the proportion of excluded cells while retaining or improving accuracy not only in datasets that do include heterotypic doublets (mixology datasets), but also some others in which such doublets are unlikely given the experimental design (i.e.
-sorted datasets such as the Zheng datasets).
-Indeed, the datasets where doublet removal did not have an impact (e.g.
+Of note, doublet removal in conjunction with filtering tended improving accuracy not only in datasets that do include heterotypic doublets (mixology datasets), but also some others in which such doublets are unlikely given the experimental design (e.g.
+FACS-sorted datasets).
+Indeed, the datasets where doublet removal did not have a clear positive impact (e.g.
 Koh, Kumar, and the simulations) are those in which heterotypic doublets are not expected.
@@
 Since clustering accuracy metrics such as the ARI are very strongly influenced by the number of clusters, we complemented it with silhouette width \cite{RousseeuwSil1987} and mutual information (MI, Figure \ref{fig:norm}B-C).
-We found most methods (including no normalization at all) to perform fairly well in most of the subpopulations, and that the same datasets (e.g.
-Zhengmix8eq and simMix1) and subpopulations were most challenging across for all methods.
+We found most methods (including no normalization at all) to perform fairly well in most of the subpopulations, and that some datasets (e.g.
+Zhengmix8eq and simMix1) and subpopulations were consistently more challenging across all methods.
 An exception was \textit{scVI}, which performed poorly on most datasets in terms of both the subpopulation silhouette and clustering accuracy (Supplementary Figure 13; \textit{scVI} was run on a reduced set of pipelines).
@@
 `seurat (no scale)' vs `seurat') tended to reduce the average silhouette width of some subpopulations and to increase that of some less distinguishable ones and was generally, but not always, beneficial on the accuracy of the final clustering.
-Regressing out covariates systematically gave equal  or poorer performance on MI, ARI and ARI at true number of clusters, with a notable negative impact in the Smart-seq datasets (Koh and Kumar).
-When performed with \textit{SCTransform}, regressing out covariates using tended to have neither a positive or negative effect in droplet-based datsets, possibly owing to its regularization.
+Regressing out covariates systematically gave equal  or poorer performance on MI, ARI and ARI at true number of clusters, with a notable negative impact among the Smart-seq datasets (Koh and Kumar).
+When performed with \textit{SCTransform}, regressing out covariates using tended to have neither a positive or negative effect in droplet-based datasets, possibly owing to its regularization.
 In general, \textit{SCTransform} systematically outperformed other methods and, even though it was developed to be applied to data with unique molecular identifiers (UMI), it also performed fairly well with the Smart-seq protocol (Koh and Kumar datasets).
@@
 The impact of the choice of dimensionality reduction method was far greater than that of normalization or feature selection (Supplementary Figure 23).
-Overall, weighting the components by the variance their explain tended to increase silhouette width (Supplementary Figure 23A) and, in most datasets, clustering accuracy (Supplementary Figure 23B).
+Overall, weighting the components by their explained variance tended to increase silhouette width (Supplementary Figure 23A) and, in most datasets, clustering accuracy (Supplementary Figure 23B).
 GLM-PCA tended to increase the average silhouette width of already well-defined subpopulations, but \textit{Seurat}'s PCA procedure however proved superior on all metrics.
@@
 
-We selected the most important alternatives for each parameter (see methods) in order to explore their interactions.
+We selected the most important alternatives for each parameter (see Methods) in order to explore their interactions.
 Analysis of variance (after correcting for baseline differences between datasets and the resolution parameter) attributed the largest proportion of variance in the main cluster metrics to the choice of dimensionality, followed by normalization, and very little to filtering (Supplementary Figure 28).
-Because the choice of dimensions impacts the number of clusters, we repeated this analysis of variance among clusterings that yielded the true number of clusters, and observed that although normalization was the most important parameter among these analyses, the choice of dimensionality came closely behind (Supplementary Figure 28B), suggesting that more attention should be paid to this relatively neglected step.
+Because the choice of dimensions impacts the number of clusters, we repeated the analysis of variance among the combinations of methods (up to clustering) that yielded the true number of clusters.
+We observed that although normalization was the most important parameter among these analyses, the choice of dimensionality came closely behind (Supplementary Figure 28B), suggesting that more attention should be paid to this relatively neglected step.
 
-We next investigated whether the benefits of some methods where conditional on choices at other steps.
-Although analysis of variance identified many weak but significant interactions, in general we found that the benefits brought by important methods were robust to differences in other parts of the pipeline (Supplementary Figures 29-30).
-For instance, we confirmed that doublet removal and \textit{SCtransform} normalization were beneficial independently of other steps, and that, again indepdendently of other steps, more stringent filtering did not yield any major improvement over lenient filtering (Supplementary Figure 29).
+We next investigated whether the benefits of some methods where conditional on the choices at other steps.
+Although analysis of variance identified many weak but significant interactions, in general we found that the benefits brought by performant methods were robust to differences in other parts of the pipeline (Supplementary Figures 29-30).
+For instance, we confirmed that doublet removal and \textit{SCtransform} normalization were beneficial independently of other steps, and that, again independendently of other steps, more stringent filtering did not yield any major improvement over lenient filtering (Supplementary Figure 29).
 Similarly, deviance-based feature selection outcompeted \textit{Seurat}'s default selection method independently of changes at other steps (Supplementary Figure 30).
@@
 
-To illustrate the applicability of \textit{pipeComp} to a completely different problem, we benchmarked the impact of surrogate variable analysis (SVA) -based methods on bulk RNAseq differential expression analysis (DEA).
+To illustrate the applicability of \textit{pipeComp} to a completely different problem, we benchmarked the impact of surrogate variable analysis (SVA)-based methods on bulk RNAseq differential expression analysis (DEA).
 The rationale behind such approaches is that the identification of factors determining genome-wide variation (i.e.
@@
 \textit{sva}\citep{leekCapturingHeterogeneityGene2007,leekSvaPackageRemoving2012} and \textit{RUVseq}\citep{rissoNormalizationRNAseq2014a} are two popular packages each implementing different methods to infer these variables.
-We therefore created a \texttt{PipelineDefinition} with three step: 1) feature filtering, 2) removal of unwanted variation (i.e.
+We therefore created a \texttt{PipelineDefinition} with three steps: 1) feature filtering, 2) removal of unwanted variation (i.e.
 SVA), and 3) DEA using the most established methods.
 The package's `pipeComp\_dea' vignette details the creation process.
-We developed 3 benchmark datasets: 1) the \texttt{ipsc} data is a random selection of 10 vs 10 samples from \citet{carcamoOriveAnalysisTranscriptional2017}, which are very heterogeneous, and foldchanges were manually added to 300 genes; 2) the \texttt{seqc} dataset contains 5 vs 5 samples of two mixtures of the SEQC consortium with different ERCC spike-in mixes \citep{ComprehensiveAssessment2014}, and includes a partially correlated batch effect; 3) A \texttt{simulation} of 8 vs 8 samples with two technical vectors of variation (see methods for more details).
+We developed 3 benchmark datasets: 1) the \texttt{ipsc} data is a random selection of 10 vs 10 samples from \citet{carcamoOriveAnalysisTranscriptional2017}, which are very heterogeneous, and foldchanges were manually added to 300 genes; 2) the \texttt{seqc} dataset contains 5 vs 5 samples of two mixtures of the SEQC consortium with different ERCC spike-in mixes \citep{ComprehensiveAssessment2014}, and includes a partially correlated batch effect; 3) A \texttt{simulation} of 8 vs 8 samples with two technical vectors of variation (see Methods for more details).
 As previously reported, we found that more stringent filtering improved DEA (Supplementary Figure 32A) and that the different DEA methods had similar performances (Figure \ref{fig:svadea}A and Supplementary Figure 32A).
@@
 With the exception of the \texttt{RUVr} method, which led to a high False Discovery Rate (FDR), all other methods showed comparable gains in sensitivity while maintaining a decent error control (if a little above the nominal FDR), even when increasing the number of surrogate variables used (Figure \ref{fig:svadea}B-C).
-While the FDR did increase in the \textit{ipsc} dataset, we suspect that much of the effect could be due to spurious differences \citep{germainTamingHumanGenetic2017} between the groups which we wrongly assume to be false positives, and the FDR nevertheless remained acceptable with 1-3 variables.
+While the FDR did increase in the \textit{ipsc} dataset, we suspect that much of the effect could be due to spurious differences \citep{germainTamingHumanGenetic2017} between the groups which we wrongly assume to be false positives.
+Nevertheless, the FDR remained acceptable with 1-3 variables.
 These results therefore suggest that these methods are robust and can be widely applied.
@@
      \item Doublet detection and removal is advised and can be performed at little computing cost with software such as \textit{scDblFinder} or \textit{scds}.
-     \item Distribution-based cell filtering fails to capture doublets and should use relatively lenient cutoffs (e.g., 5 MADs, or 3 MADs in at least 2 distributions) to exclude poor-quality cells while avoiding bias against some subpopulations.
+     \item Distribution-based cell filtering fails to capture doublets and should use relatively lenient cutoffs (e.g., 5 MADs, or 3 MADs in at least 2 distributions - see for instance our 'default' filters in the methods) to exclude poor-quality cells while avoiding bias against some subpopulations.
      \item Features filtering based on feature type did not appear beneficial.
@@
 The challenge of distinguishing subpopulations of T-cells in the Zhengmix8eq dataset was previously discussed \cite{duoClustering2018}.
-In contrast, the Koh dataset harbors as many populations, but originates from more distinct groups (from hESC in different differentiation stages), while the Kumar and mixology datasets harbor a very low subpopulations complexity and most methods applied on them performed well.
+In contrast, while the Koh dataset harbors as many populations, they originate from more distinct groups (from hESC in different differentiation stages), while the Kumar and mixology datasets harbor a very low subpopulations complexity and most methods applied on them performed well.
 
@@
 Analyses were performed in R 3.6.0 (Bioconductor 3.9) and the following packages were installed from GitHub repositories: \textit{Seurat} (version 3.0.0), \textit{SCTransform} (0.2.0), \textit{DoubletFinder} (2.0.1), \textit{scDblFinder} (1.1.4), \textit{scds} (1.0.0), \textit{SAVERX} (1.0.0), \textit{scImpute} (0.0.9), \textit{ALRA} (commit 7636de8), \textit{DCA} (0.2.2), \textit{DrImpute} (1.2), \textit{ENHANCE} (R version, commit 1571696), \textit{SAVERX} (1.0.0).
-The glmPCA was performed using the \texti{glmpca} package (The code for the glmPCA was obtained from \url{https://github.com/willtownes/scrna2019} (commit 1ddcc30ebb95d083a685f12fe81d35dd1b0cb1b2).
+The glmPCA was performed using the \texti{glmpca} package, while the code for deviance calculation was obtained from \url{https://github.com/willtownes/scrna2019} (commit 1ddcc30ebb95d083a685f12fe81d35dd1b0cb1b2).
 
@@
 
-The `pca' and `pca2' clusters refer to the multivariate outlier detection methods implemented in \textit{scater}, running \textit{runPCA} with \texttt{use\_coldata=TRUE, detect\_outliers=TRUE}.
-`pca` uses all covariates, while `pca2` uses only the log10(counts), log10(features), proportion mitochondrial and proportion in the top 50 features.
+The `pca.all' and `pca.sel' clusters refer to the multivariate outlier detection methods implemented in \textit{scater}, running \textit{runPCA} with \texttt{use\_coldata=TRUE, detect\_outliers=TRUE}.
+`pca.all` uses all covariates, while `pca.sel` uses only the log10(counts), log10(features), proportion mitochondrial and proportion in the top 50 features.
 
@@
 Other evaluation metric heatmaps (e.g.
-panel \textbf{C} here) map colors linearly to the number of (matrix-wise) Median Absolute Deviations from the (column-wise) median.
-Using this mapping, the same \textit{difference} in color has the same meaning across datasets, and the color-scale is not primarily capturing baseline differences between datasets.
+panel \textbf{C} here) map colors to the signed square-root of the number of (matrix-wise) Median Absolute Deviations from the (column-wise) median.
+Using this mapping, \textit{differences} in color have the same meaning across datasets, and the color-scale is not primarily capturing outliers or baseline differences between datasets.
 The numbers printed in the cells represent the raw (i.e.
@@
   \hline
-Koh & 33922 & 531 & SMARTer & FACS purified differentiation stages \\ 
-  Kumar & 41930 & 246 & SMARTer & Mouse ESC cultured in different conditions \\ 
+Koh & 33922 & 531 & SMARTer & 9 FACS purified differentiation stages \\ 
+  Kumar & 41930 & 246 & SMARTer & Mouse ESC cultured in 3 different conditions \\ 
   Zhengmix4eq & 10434 & 3994 & 10x & Mixtures of FACS purified PBMCs \\ 

### 2020-05-13 main_figures/dimensionality.Rmd
fixed label colors in heatmaps

@@
 ```{r figure6, fig.width=9, fig.height=9.5}
-pdf("dimensionality.pdf", width=9, height=10)
+pdf("dimensionality.pdf", width=9, height=10.5)
 plot_grid( p1, p2, p3, nrow=3, labels=c(NA,LETTERS[3:4]))

### 2020-05-13 main_figures/normalization.Rmd
fixed label colors in heatmaps

@@
 
-p1 <- renameNorm(evalHeatmap(res, step="dimreduction", what="log10_total_counts", scale=FALSE, show_heatmap_legend = TRUE, name="adj.R2", what2="PC1.covar.adjR2", agg.by = "norm", reorder_rows = TRUE, col=viridisLite::viridis(100), row_split = classifyNorms(norm), row_title="Variance in PC1 explained by covariate") +
-  evalHeatmap(res, step="dimreduction", what="log10_total_features", scale=FALSE, show_heatmap_legend = FALSE, what2="PC1.covar.adjR2", agg.by = "norm", reorder_rows = TRUE, col=viridisLite::viridis(100)))
+p1 <- renameNorm(evalHeatmap(res, step="dimreduction", what="log10_total_counts", scale=FALSE, show_heatmap_legend = TRUE, name="adj.R2", what2="PC1.covar.adjR2", agg.by = "norm", reorder_rows = TRUE, col=viridisLite::viridis(100), row_split = classifyNorms(norm), row_title="Variance in PC1 explained by covariate", value_cols=c("black","black")) +
+  evalHeatmap(res, step="dimreduction", what="log10_total_features", scale=FALSE, show_heatmap_legend = FALSE, what2="PC1.covar.adjR2", agg.by = "norm", reorder_rows = TRUE, col=viridisLite::viridis(100), value_cols=c("black","black")))
 

### 2020-05-13 main_figures/svadea.Rmd
fixed label colors in heatmaps

@@
                     row_split="sva.method", anno_legend=FALSE, show_column_names=TRUE, 
-                    font_factor=0.85, scale="colCenterScale" ) +
+                    font_factor=0.85 ) +
   evalHeatmap( res, step="dea", what="TPR", agg.by=c("sva.method","k"), 
                     row_split="sva.method", anno_legend=FALSE, show_column_names=TRUE, 
-                    font_factor=0.85, scale="colCenterScale", filterExpr=threshold==0.05 ) +
+                    font_factor=0.85, filterExpr=threshold==0.05 ) +
   evalHeatmap(res, step="dea", what="FDR", agg.by=c("sva.method","k"), row_split="sva.method", 
-                     anno_legend=FALSE, show_column_names=TRUE, font_factor=0.85, scale="colCenterScale",
-                     col=rev(viridisLite::inferno(100)), value_cols=c("white","black"), filterExpr=threshold==0.05)
-hts <- pipeComp:::.renameHrows(hts, f = function(x) gsub("none > 1", "none",x, fixed=TRUE))
+                     anno_legend=FALSE, show_column_names=TRUE, font_factor=0.85, 
+                     col=rev(viridisLite::inferno(101)), value_cols=c("white","black"), filterExpr=threshold==0.05)
+hts <- .renameHrows(hts, f = function(x) gsub("none > 1", "none",x, fixed=TRUE))
 #draw(hts, heatmap_legend_side = "bottom", merge_legend=TRUE, annotation_legend_list=.scaledLegend())

### 2020-05-13 supplementary_figures/supp_overall.Rmd
fixed label colors in heatmaps

@@
 **B:** Proportion of the variance in Adjusted Rand Index (ARI) explained by the parameters, among the analyses giving the true number of clusters.
-In both panels, the proportion is calculated after accounting for differences due to the dataset, the resolution and their interaction.
+The ANOVA were performed with the model
+
+`metric ~ dataset*resolution+doubletmethod+filt+norm+clustmethod+dims`
+
+In both panels, the proportion is calculated after substracting the variance attributable to differences due to the dataset, the resolution and their interaction.
+The alternatives included in the analysis (other than the default ones described in the methods) were:
+```
+list( doubletmethod=c("none","doublet.scDblFinder"),
+      filt=c("none","filt.default","filt.stringent"),
+      norm=c("norm.seurat","norm.sctransform","norm.scran"),
+      sel=c("sel.vst","sel.deviance","sel.expr"),
+      selnb=c(2000,4000),
+      dr=c("seurat.pca"),
+      dims=c("elbow", 10, "maxLikGlobal"),
+      clustmethod=c("clust.seurat","clust.scran.knnAnnoy") )
+```
 

### 2020-05-13 bmc_article.tex

@@
 
-The \textit{pipeComp} package defines a pipeline as, minimally, a list of functions executed consecutively on the output of the previous one (Figure \ref{fig:explanation}A; see also Supplementary methods).
+The \textit{pipeComp} package defines a pipeline as, minimally, a list of functions executed consecutively on the output of the previous one {\color{red}(Figure \ref{fig:explanation}A; see also Supplementary methods).
 In addition, optional benchmark functions can be set for each step to provide standardized, multi-layered evaluation metrics.
-Given such a \texttt{PipelineDefinition} object, a set of alternative parameters (which might include different subroutines calling different methods) and benchmark datasets, the \texttt{runPipeline} function then proceeds through all combinations of arguments (or a desired subset), avoiding recomputing the same step twice without the need to save all alternative intermediates, and compiling evaluations (including run time) on the fly (see Supplementary methods for more detail).
-Variations in a given parameter can be evaluated using all metrics from this point downward in the pipeline.
+Given such a \texttt{PipelineDefinition} object, a set of alternative parameters (which might include different subroutines calling different methods) and benchmark datasets, the \texttt{runPipeline} function then proceeds through all combinations of arguments (or a desired subset), avoiding recomputing the same step twice without the need to save all alternative intermediates, and compiling evaluations (including run time) on the fly (see Supplementary methods for more detail).} Variations in a given parameter can be evaluated using all metrics from this point downward in the pipeline.
 This is especially important because end-point metrics, such as the adjusted Rand index (ARI) \cite{HubertARI1985} for clustering, are not perfect.
@@
 
-We next investigated the impact of several filtering methods (see Methods for more details): four methods based on deviations to MADs with increasing levels of stringency (named \textit{lenient}, \textit{default}, \textit{stringent}, \textit{veryStringent}) and two methods based on \textit{scater}'s \textit{runPCA} using all (\textit{pca.all}) or selected covariates (\textit{pca.sel}).
+{\color{red}We next investigated the impact of several filtering methods (see Methods for more details): four methods based on deviations to MADs with increasing levels of stringency (named \textit{lenient}, \textit{default}, \textit{stringent}, \textit{veryStringent}) and two methods based on \textit{scater}'s \textit{runPCA} using all (\textit{pca.all}) or selected covariates (\textit{pca.sel})}.
 An important risk of excluding cells on the basis of their distance from the whole distribution of cells on some properties (e.g., library size) is that these properties tend to have different distributions across subpopulations.
@@
 Since filtering changes the relative abundance of the different subpopulations, global clustering accuracy metrics such as ARI are not appropriate here.
-We therefore calculated the per-subpopulation precision and recall using the Hungarian algorithm \cite{PapadimitriouHu1998} and monitored the mean F1 score across subpopulations.
+We therefore calculated the per-subpopulation precision and recall using the Hungarian algorithm \cite{PapadimitriouHu1998} and monitored the mean F1 score {\color{red}across subpopulations.
 A first observation was that, although more stringent filtering tended to be associated with an increase in accuracy, it tended to plateau and could also become deleterious.
@@
 high library size), rather than low-quality cells.
-On the opposite, too lenient filtering (or lack thereof) can lead to misclassified cells, as was for instance the case especially with the Zhengmix4 datasets (Figure \ref{fig:filtering}B).
+In contrast, too lenient filtering (or lack thereof) can lead to misclassified cells, as was the case for the Zhengmix4 datasets (Figure \ref{fig:filtering}B).
 We found that a relatively mild filtering (`default' - see methods) provided a good trade-off, often leading to high clustering performance (e.g.
@@
 We found very little overlap between the cells excluded by doublet removal and those excluded by MAD-based filtering (Supplementary Figure 12B).
-Of note, doublet removal in conjunction with filtering tended improving accuracy not only in datasets that do include heterotypic doublets (mixology datasets), but also some others in which such doublets are unlikely given the experimental design (e.g.
+Of note, doublet removal in conjunction with filtering tended to improve accuracy not only in datasets that do include heterotypic doublets (mixology datasets), but also some others in which such doublets are unlikely given the experimental design (e.g.
 FACS-sorted datasets).
@@
 Koh, Kumar, and the simulations) are those in which heterotypic doublets are not expected.
-Overall, we therefore recommend the use of doublet removal followed by our `default' filtering (see Methods) or a similar approach.
+Overall, we therefore recommend the use of doublet removal followed by our `default' filtering (see Methods) or a similar approach.}
 
@@
 Since clustering accuracy metrics such as the ARI are very strongly influenced by the number of clusters, we complemented it with silhouette width \cite{RousseeuwSil1987} and mutual information (MI, Figure \ref{fig:norm}B-C).
-We found most methods (including no normalization at all) to perform fairly well in most of the subpopulations, and that some datasets (e.g.
+We found most methods (including no normalization at all) to perform fairly well in most of the subpopulations{\color{red}, and that some datasets (e.g.
 Zhengmix8eq and simMix1) and subpopulations were consistently more challenging across all methods.
@@
 Regressing out covariates systematically gave equal  or poorer performance on MI, ARI and ARI at true number of clusters, with a notable negative impact among the Smart-seq datasets (Koh and Kumar).
-When performed with \textit{SCTransform}, regressing out covariates using tended to have neither a positive or negative effect in droplet-based datasets, possibly owing to its regularization.
-In general, \textit{SCTransform} systematically outperformed other methods and, even though it was developed to be applied to data with unique molecular identifiers (UMI), it also performed fairly well with the Smart-seq protocol (Koh and Kumar datasets).
+When combined with \textit{SCTransform}, regressing out covariates using tended to have neither a positive or negative effect in droplet-based datasets, possibly owing to its regularization.
+In general, \textit{SCTransform} systematically outperformed other methods and, even though it was developed to be applied to data with unique molecular identifiers (UMI), it also performed fairly well with the Smart-seq protocol (Koh and Kumar datasets).}
 
@@
 As the proportion of variability in a gene attributable to subpopulations can be measured in various ways, we first compared three approaches: ANOVA on log-normalized count, ANOVA on \textit{SCTransform} normalization, and deviance explained.
-The ANOVAs performed on a standard \textit{Seurat} normalization and on \textit{SCTransform} data were highly correlated (Pearson correlation >0.9, Supplementary Figure 19A).
+The ANOVAs performed on a standard \textit{Seurat} normalization and on \textit{SCTransform} data were highly correlated ({\color{red}Pearson correlation >0.9,} Supplementary Figure 19A).
 These estimates were also in good agreement with the deviance explained, although lowly-expressed genes could have a high deviance explained without having much of their variance explained by subpopulation (Supplementary Figure 19B-D).
@@
 In general, we found deviance and unstandardized estimates of variance to provide the best results across datasets and normalization methods.
-Increasing the number of features selected also tended to lead to an increase in the accuracy of the clustering, typically plateauing after 4000 features (Supplementary Figure 22).
+Increasing the number of features selected also {\color{red}tended to lead} to an increase in the accuracy of the clustering, typically plateauing after 4000 features (Supplementary Figure 22).
 
@@
 The impact of the choice of dimensionality reduction method was far greater than that of normalization or feature selection (Supplementary Figure 23).
-Overall, weighting the components by their explained variance tended to increase silhouette width (Supplementary Figure 23A) and, in most datasets, clustering accuracy (Supplementary Figure 23B).
+{\color{red}Overall, weighting the components by their explained variance tended to increase silhouette width (Supplementary Figure 23A) and, in most datasets, clustering accuracy (Supplementary} Figure 23B).
 GLM-PCA tended to increase the average silhouette width of already well-defined subpopulations, but \textit{Seurat}'s PCA procedure however proved superior on all metrics.
 Like GLM-PCA, \textit{scVI}'s Linear Decoder (LD) does not explicitly rely on normalized counts.
-Its performance was however the worst both in terms of silhouette width and clustering accuracy (Supplementary Figure 15), and it only performed well in some of the least challenging datasets (mixology10x5cl and simMix2).
+{\color{red}Its performance was however the worst both in terms of silhouette width and clustering accuracy (Supplementary Figure 15), and it only performed well in some of the least challenging datasets (mixology10x5cl and simMix2).}
 
@@
 
-Different approaches have been proposed to select the appropriate number of dimensions, from the visual identification of an inflexion point in the curve of the variance explained by each component (`elbow' method), to more complex algorithms.
-We evaluated the performance of dimensionality estimators implemented in the \textit{intrinsicDimension} package \cite{johnssonDimensionality2015}, as well as common procedures such as the `elbow` method, some scRNAseq-specific methods such as the JackStraw procedure \cite{ChungJackstraw2015} or \textit{scran}'s \texttt{denoisePCA} \cite{LunScran2016}, and the recent application of Fisher Separability analysis \cite{AlberganteSepar2019}.
+Different approaches have been proposed to select the appropriate number of dimensions, from the visual identification of an {\color{red}inflexion point in the curve of the variance explained by each component (`elbow' method), to more complex algorithms.} We evaluated the performance of dimensionality estimators implemented in the \textit{intrinsicDimension} package \cite{johnssonDimensionality2015}, as well as common procedures such as the `elbow` method, some scRNAseq-specific methods such as the JackStraw procedure \cite{ChungJackstraw2015} or \textit{scran}'s \texttt{denoisePCA} \cite{LunScran2016}, and the recent application of Fisher Separability analysis \cite{AlberganteSepar2019}.
 
-We compared the various estimates of in their ability to retrieve the intrinsic number of dimensions in a dataset, based on Seurat's weighted PCA space.
+{\color{red}We compared the various estimates in their ability to retrieve the intrinsic number of dimensions in a dataset, based on Seurat's weighted PCA space.
 As a first approximation of the true dimensionality, we computed the variance in each principal component (based on Seurat's weighted PCA space) that was explained by the subpopulations, which sharply decreased after the first few components in most datasets (Figure \ref{fig:dim}A).
-These truth-based estimates where then compared to those of the above dimensionality estimation methods (Figure \ref{fig:dim}B).
-Of note, the methods differ widely in view of their computing time (Figure \ref{fig:dim}B) and we saw no relationship between the accuracy of the estimates and the complexity of the method.
+These truth-based estimates were then compared to those of the above dimensionality estimation methods (Figure \ref{fig:dim}B).} Of note, the methods differ widely in compute time (Figure \ref{fig:dim}B) and we saw no relationship between the accuracy of the estimates and the complexity of the method.
 Reasoning that over-estimating the number of dimensions is less problematic than under-estimating it, we kept the former methods for a full analysis of their impact on clustering (Figure \ref{fig:dim}C-D), when combined with \textit{SCTransform} or standard \textit{Seurat} normalization.
-Although most methods performed well on the various clustering measures, the global maximum likelihood based on translated Poisson Mixture Model (\textit{maxLikGlobal}, using 10 or 20 nearest neighbors, although estimates were relatively stable across decent values of \textit{k}; see Supplementary Figure 26) provided the dimensionality estimate that best separated challenging subpopulations (Figure \ref{fig:dim}C) and tended to result in the best clustering accuracy (Figure \ref{fig:dim}D).
+Although most methods performed well on the various clustering measures, the global maximum likelihood based on translated Poisson Mixture Model {\color{red}(\textit{maxLikGlobal}, using 10 or 20 nearest neighbors, although estimates were relatively stable across various values of \textit{k}; see Supplementary Figure 26) provided the dimensionality estimate that best separated challenging subpopulations (Figure \ref{fig:dim}C) and tended to result in the best clustering accuracy (Figure \ref{fig:dim}D).
 This method systematically selected many more components than were associated with the subpopulations in Figure \ref{fig:dim}A, suggesting that although these additional components appear individually uninformative, in combination they nevertheless contribute to classification.
-An important implication is that the commonly-used elbow method most likely underestimates dimensionality.
+An important implication is that the commonly-used elbow method most likely underestimates dimensionality.}
 
@@
 
-\subsection*{Overview of interactions between steps}
+{\color{red}\subsection*{Overview of interactions between steps}
 
 We selected the most important alternatives for each parameter (see Methods) in order to explore their interactions.
-Analysis of variance (after correcting for baseline differences between datasets and the resolution parameter) attributed the largest proportion of variance in the main cluster metrics to the choice of dimensionality, followed by normalization, and very little to filtering (Supplementary Figure 28).
-Because the choice of dimensions impacts the number of clusters, we repeated the analysis of variance among the combinations of methods (up to clustering) that yielded the true number of clusters.
+After accounting, in the model, for baseline differences between datasets and the impact of the resolution parameter, ANOVA attributed the largest proportion of variance in the main cluster metrics to the choice of dimensionality reduction, followed by normalization, and very little to filtering (Supplementary Figure 28A).
+Because the choice of dimension reduction impacts the number of clusters, we repeated the analysis of variance in ARI among the combinations of parameters that yielded the true number of clusters (Supplementary Figure 28B).
 We observed that although normalization was the most important parameter among these analyses, the choice of dimensionality came closely behind (Supplementary Figure 28B), suggesting that more attention should be paid to this relatively neglected step.
 
-We next investigated whether the benefits of some methods where conditional on the choices at other steps.
-Although analysis of variance identified many weak but significant interactions, in general we found that the benefits brought by performant methods were robust to differences in other parts of the pipeline (Supplementary Figures 29-30).
+We next investigated whether the benefits of some methods were conditional on the choices at other steps.
+Although analysis of variance identified many weak but significant interactions, we found that the benefits brought by performant methods were robust to differences in other parts of the pipeline (Supplementary Figures 29-30).
 For instance, we confirmed that doublet removal and \textit{SCtransform} normalization were beneficial independently of other steps, and that, again independendently of other steps, more stringent filtering did not yield any major improvement over lenient filtering (Supplementary Figure 29).
-Similarly, deviance-based feature selection outcompeted \textit{Seurat}'s default selection method independently of changes at other steps (Supplementary Figure 30).
+Similarly, deviance-based feature selection outcompeted \textit{Seurat}'s default selection method independently of changes at other steps (Supplementary Figure 30).}
 
@@
 
-\subsection*{Application of \textif{pipeComp} to a different context}
+{\color{red}\subsection*{Application of \textif{pipeComp} to a different context}
 
@@
 surrogate variables) and their inclusion in the DEA's generalized linear model might improve DEA, especially in the presence of sources of variation unrelated to the experimental variables.
-\textit{sva}\citep{leekCapturingHeterogeneityGene2007,leekSvaPackageRemoving2012} and \textit{RUVseq}\citep{rissoNormalizationRNAseq2014a} are two popular packages each implementing different methods to infer these variables.
+\textit{sva} \citep{leekCapturingHeterogeneityGene2007,leekSvaPackageRemoving2012} and \textit{RUVseq} \citep{rissoNormalizationRNAseq2014a} are two popular packages each implementing different methods to infer these variables.
 We therefore created a \texttt{PipelineDefinition} with three steps: 1) feature filtering, 2) removal of unwanted variation (i.e.
-SVA), and 3) DEA using the most established methods.
+SVA), and 3) DEA using the most established methods  \citep{germainRNAontheBENCHComputationalEmpirical2016}.
 The package's `pipeComp\_dea' vignette details the creation process.
-We developed 3 benchmark datasets: 1) the \texttt{ipsc} data is a random selection of 10 vs 10 samples from \citet{carcamoOriveAnalysisTranscriptional2017}, which are very heterogeneous, and foldchanges were manually added to 300 genes; 2) the \texttt{seqc} dataset contains 5 vs 5 samples of two mixtures of the SEQC consortium with different ERCC spike-in mixes \citep{ComprehensiveAssessment2014}, and includes a partially correlated batch effect; 3) A \texttt{simulation} of 8 vs 8 samples with two technical vectors of variation (see Methods for more details).
-As previously reported, we found that more stringent filtering improved DEA (Supplementary Figure 32A) and that the different DEA methods had similar performances (Figure \ref{fig:svadea}A and Supplementary Figure 32A).
+We developed 3 benchmark datasets: 1) the \texttt{ipsc} data is a random selection of 10 vs 10 samples from \citep{carcamoOriveAnalysisTranscriptional2017}, which are very heterogeneous, and foldchanges were manually added to 300 genes; 2) the \texttt{seqc} dataset contains 5 vs 5 samples of two mixtures of the SEQC consortium with different ERCC spike-in mixes \citep{ComprehensiveAssessment2014}, and includes a partially correlated batch effect; 3) A \texttt{simulation} of 8 vs 8 samples with two technical vectors of variation (see Methods for more details).
+As previously reported \citep{ComprehensiveAssessment2014}, we found that more stringent filtering improved DEA (Supplementary Figure 32A) and that the different DEA methods had similar performances (Figure \ref{fig:svadea}A and Supplementary Figure 32A).
 Including a SVA-based step increased sensitivity across all datasets, particularly in datasets with technical vectors of variation (\textit{seqc} and \textit{simulation}), and benefited all DEA methods in a similar fashion (Figure \ref{fig:svadea}A,C and Supplementary Figures 32-33).
@@
 Nevertheless, the FDR remained acceptable with 1-3 variables.
-These results therefore suggest that these methods are robust and can be widely applied.
+These results therefore suggest that these methods are robust and can be widely applied.}
 
@@
 
-\subsection*{Practical recommendations}
+\subsection*{{\color{red}Practical} recommendations}
 
-On the basis of our findings, we can make a number of practical analysis recommendations, also summarized in Supplementary Figure 34:
+On the basis of our findings, we can make a number of \textcolor{red}{practical} analysis recommendations, also summarized in Supplementary Figure 34:
 
@@
      \item Doublet detection and removal is advised and can be performed at little computing cost with software such as \textit{scDblFinder} or \textit{scds}.
-     \item Distribution-based cell filtering fails to capture doublets and should use relatively lenient cutoffs (e.g., 5 MADs, or 3 MADs in at least 2 distributions - see for instance our 'default' filters in the methods) to exclude poor-quality cells while avoiding bias against some subpopulations.
+     \item Distribution-based cell filtering fails to capture doublets and should use relatively lenient cutoffs (e.g., 5 MADs, or 3 MADs in at least 2 distributions \textcolor{ref}{- see for instance our 'default' filters in the methods}) to exclude poor-quality cells while avoiding bias against some subpopulations.
      \item Features filtering based on feature type did not appear beneficial.
@@
        \item Denoising appeared beneficial to the identification of some subpopulations in 10x datasets, but not in Smart-seq datasets.
-       \item We found especially \textit{ALRA} (with prior normalization), \textit{DrImpute} (with prior processing) and \textit{DCA} to offer the best performances, although none of these methods was beneficial over all subpopulations.
+       \item We found especially \textit{ALRA} (with prior normalization), \textit{DrImpute} (with prior processing) and \textit{DCA} to offer the best performances\textcolor{red}{, although none of these methods was beneficial over all subpopulations}.
    \end{itemize}
@@
      \item Similarly to previous reports \cite{SunDimRed2019}, we recommend the irlba-based PCA using weighting of the components, as implemented in \textit{Seurat}. 
-     \item We advise against the commonly-used elbow method (because it is too conservative) and the jackstraw method (low performance at great computational costs) for deciding on the components to include, and recommend a method like the \textit{global maximum likelihood based on translated Poisson Mixture Model} method (e.g., implemented in the  \texttt{maxLikGlobalDimEst} function of \texttt{intrinsicDimension}, in our case using the 10 or 20 nearest neighbors).
+     \item We advise against the commonly-used elbow method (because it is too conservative) and the jackstraw method (low performance at great computational costs) for deciding on the components to include, and recommend a method like the \textit{global maximum likelihood based on translated Poisson Mixture Model} method (e.g., implemented in the  \texttt{maxLikGlobalDimEst} function of \texttt{intrinsicDimension}, in our case using the \textcolor{red}{10 or }20 nearest neighbors).
    \end{itemize}
@@
 
-Among all pipelines that were tested, the performance varied greatly among datasets.
+{\color{red}Among all pipelines tested, performance varied greatly across datasets.
 The ZhengMix8eq and simMix1 datasets were the most challenging datasets across all preprocessing steps that we tested.
-The average silhouette width of half of their subpopulations was systematically lower that the other subpopulations/ datasets and the ARI score at true number of clusters, if ever achieved, was always poor.
+The average silhouette width of half of their subpopulations was systematically lower than the other subpopulations/ datasets and the ARI score at true number of clusters, if ever achieved, was always poor.
 It is likely that these two datasets are challenging due to their combined sparsity and complexity (see Supplementary Figure 1).
 The challenge of distinguishing subpopulations of T-cells in the Zhengmix8eq dataset was previously discussed \cite{duoClustering2018}.
-In contrast, while the Koh dataset harbors as many populations, they originate from more distinct groups (from hESC in different differentiation stages), while the Kumar and mixology datasets harbor a very low subpopulations complexity and most methods applied on them performed well.
+In contrast, while the Koh dataset harbors as many populations, they originate from more distinct groups (from hESC in different differentiation stages), while the Kumar and mixology datasets harbor a very low subpopulation complexity and most methods applied on them performed well.}
 
@@
 Instead, the \textit{CellBench} package was recently proposed to address a similar need \cite{suCellbench}, offering an elegant piping syntax to combine alternative methods at successive steps.
-A key additional feature of \textit{pipeComp} is its ability to perform evaluation in parallel to the pipeline and offering on-the-fly evaluation at multiple steps.
+A key additional feature of \textit{pipeComp} is its ability to perform evaluation in parallel to the pipeline and offering on-the-fly evaluation at {\color{red}multiple steps.
 This avoids the need to store potentially large intermediate data for all possible permutations and thus allowing a combinatorial benchmark.
-This, along with the set-up simplicity (demonstrated here by our separate SVA-DEA benchmark), also distinguishes \textit{pipeComp} from general workflow frameworks such as \textit{drake}\citep{landauDrakePackage2018}; indeed we estimate that the benchmarks discussed here would have taken approximately 16TB of storage had all non-redundant intermediates been saved (see the Supplementary methods for more detailed explanations of the framework and comparison to \textit{CellBench} and \textit{drake}).
-In addition, the fact that benchmark functions are stored in the \texttt{PipelineDefinition} makes the benchmark more smoothly portable, making it easy to modify, extend with further methods, or apply to other datasets.
+This, along with the set-up simplicity (demonstrated here by our separate SVA-DEA benchmark), also distinguishes \textit{pipeComp} from general workflow frameworks such as \textit{drake}\citep{landauDrakePackage2018}; indeed we estimate that the benchmarks discussed here would have taken approximately 16TB of storage had all non-redundant intermediates been saved (see the Supplementary methods for more detailed explanations of the framework and comparison to \textit{CellBench} and \textit{drake}).} In addition, the fact that benchmark functions are stored in the \texttt{PipelineDefinition} makes the benchmark more smoothly portable, making it easy to modify, extend with further methods, or apply to other datasets.
 
@@
 \textit{pipeComp} is a flexible R framework for evaluating methods and parameter combinations in a complex pipeline, computing on-the-fly multilevel evaluation metrics.
-Applying this framework to scRNAseq clustering enabled us to make clear recommendations on the steps of filtering, normalization, feature selection, denoising, dimensionality reduction and clustering (Supplementary Figure 34).
+Applying this framework to scRNAseq clustering enabled us to make {\color{red}clear} recommendations on the steps of filtering, normalization, feature selection, denoising, dimensionality reduction and clustering (Supplementary Figure 34).
 We demonstrate how a diversity of multilevel metrics can be more robust, more sensitive, and more nuanced than simply evaluating the final clustering performance.
-We further showed that the \textit{pipeComp} framework can be applied to extend the current field of benchmarking, as well as to investigate pipelines other domains with multi-step computational analyses.
+We {\color{red}further showed} that the \textit{pipeComp} framework can be applied to extend the current field of benchmarking, as well as to investigate pipelines other domains with multi-step computational analyses.
 
@@
 Analyses were performed in R 3.6.0 (Bioconductor 3.9) and the following packages were installed from GitHub repositories: \textit{Seurat} (version 3.0.0), \textit{SCTransform} (0.2.0), \textit{DoubletFinder} (2.0.1), \textit{scDblFinder} (1.1.4), \textit{scds} (1.0.0), \textit{SAVERX} (1.0.0), \textit{scImpute} (0.0.9), \textit{ALRA} (commit 7636de8), \textit{DCA} (0.2.2), \textit{DrImpute} (1.2), \textit{ENHANCE} (R version, commit 1571696), \textit{SAVERX} (1.0.0).
-The glmPCA was performed using the \texti{glmpca} package, while the code for deviance calculation was obtained from \url{https://github.com/willtownes/scrna2019} (commit 1ddcc30ebb95d083a685f12fe81d35dd1b0cb1b2).
+{\color{red}The glmPCA was performed using the \texti{glmpca} package, while the code for deviance calculation was obtained from \url{https://github.com/willtownes/scrna2019} (commit 1ddcc30ebb95d083a685f12fe81d35dd1b0cb1b2).}
 
@@
 The \textit{stringent} set of filters uses the same thresholds, but excludes a cell if it is an outlier on any single distribution. 
-The \textit{veryStringent} set of filters excludes any cell that is an outlier by at least 2 MADs (lower or higher) on any of: log10\_total\_counts, log10\_total\_features, pct\_counts\_Mt, or pct\_counts\_in\_top\_20\_features.
+{\color{red}The \textit{veryStringent} set of filters excludes any cell that is an outlier by at least 2 MADs (lower or higher) on any of: log10\_total\_counts, log10\_total\_features, pct\_counts\_Mt, or pct\_counts\_in\_top\_20\_features.}
 The \textit{lenient} set of filters excludes cells that are outliers on at least two distributions by at least 5 MADs, except for pct\_counts\_Mt where the threshold is $>$ 3 MADs and $>$ 0.08.
@@
 
-\subsection*{Overall run for interactions across steps}
+{\color{red}\subsection*{Overall run for interactions across steps}
 
@@
 
+Filtering was performed with edgeR's \texttt{filterByExpr} function, changing the \texttt{min.count} argument.
 For \textit{vstsva}, we first applied the variance-stabilizing transformation from \textit{DESeq2}\citep{loveModeratedEstimation2014} before applying \texttt{sva::sva}.
@@
 The exact wrappers used can be found in the package at \url{https://github.com/plger/pipeComp/blob/master/inst/extdata/dea\_wrappers.R}.
-
+}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \textbf{C:} Clustering accuracy (\textit{Seurat} clustering), measured by mutual information (MI), Adjusted Rand Index (ARI) and ARI at the true number of cluster.
-Grey squares indicate that the true number of clusters could not be reached with the corresponding method and dataset.
-`feat\_regress', `mt\_regress' and `feat\_mt\_regress' respectively stand for the regressing out of the number of features, the proportion of mitochondrial reads, or both during scaling.
-Throughout this paper, silhouette plots (panel \textbf{B}) square-root transform values for color mapping.
+{\color{red}Grey squares indicate that the true number of clusters could not be reached with the corresponding method, dataset and parameter set.} `feat\_regress', `mt\_regress' and `feat\_mt\_regress' respectively stand for the regressing out of the number of features, the proportion of mitochondrial reads, or both during scaling.
+{\color{red}Throughout this paper, silhouette plots (panel \textbf{B}) square-root transform values for color mapping.
 Other evaluation metric heatmaps (e.g.
@@
 The numbers printed in the cells represent the raw (i.e.
-unscaled) metric values.}
+unscaled) metric values, and are printed in white or black depending on whether they are above or below the median.}}
     \label{fig:norm}
@@
 \textbf{D:} Clustering accuracy across the normalization/dimensionality estimation methods.
-The color-mapping scheme is the same as in Figure \ref{fig:norm}C}
+{\color{red}The color-mapping scheme is the same as in Figure \ref{fig:norm}C}}
     \label{fig:dim}
@@
     \caption{\textbf{Evaluation of clustering methods.} Clustering accuracy (\textbf{A}) and run time (\textbf{B}) of \textit{scran}- and \textit{Seurat}-based clustering tools in combination with \textit{SCTransform} and standard \textit{Seurat}'s normalization.
-The color-mapping scheme in panel \textbf{A} and the left of panel \textbf{B} is the same as in Figure \ref{fig:norm}C.}
+{\color{red} The color-mapping scheme in panel \textbf{A} and the left of panel \textbf{B} is the same as in Figure \ref{fig:norm}C.}}
     \label{fig:clust}
@@
     \fi
-    \caption{\textbf{Evaluation of the impact of surrogate variable analysis (SVA) on (bulk) RNAseq differential expression analysis using \textit{pipeComp}.} \textit{RUVr} and \textit{RUVs} refer to the two \textit{RUVSeq} functions by the same name; \textit{svaseq} refers to the function of the same name in \textit{sva}, while \textit{vstsva} refers to the original \textit{sva} method applied no variance-stabilized data.
+    \caption{{\color{red}\textbf{Evaluation of the impact of surrogate variable analysis (SVA) on (bulk) RNAseq differential expression analysis using \textit{pipeComp}.} \textit{RUVr} and \textit{RUVs} refer to the two \textit{RUVSeq} functions by the same name; \textit{svaseq} refers to the function of the same name in \textit{sva}, while \textit{vstsva} refers to the original \textit{sva} method applied no variance-stabilized data.
 \textbf{A:} True positive rates (TPR) and False discovery rates (FDR) of various combinations of SVA-based methods (using a single surrogate variable) and DEA methods.
@@
 \textbf{C:} Correlation of estimated logFC with expected ones, as well as TPR and FDR (at nominal FDR threshold of 0.05) of the different methods with increasing numbers of variables.
-As in \ref{fig:norm}C, the color mapping tracks the number of (matrix-wise) Median Absolute Deviations from the (column-wise) median, while the printed numbers represent the raw metric values.}
+As in \ref{fig:norm}C, the color mapping tracks the number of (matrix-wise) Median Absolute Deviations from the (column-wise) median, while the printed numbers represent the raw metric values.}}
     \label{fig:svadea}
@@
 \begin{table}[h!]
-\caption{\textbf{Overview of the benchmark datasets.} `\# features' indicates the number of features detected in at least 10 cells.}
+\caption{\textbf{Overview of the benchmark datasets.} {\color{red}`\# features' indicates the number of features detected in at least 10 cells.}}
 \label{tab:table1}
@@
   \hline
-dataset & \# features & \# cells & protocol & description \\ 
+dataset & {\color{red}\# features} & \# cells & protocol & description \\ 
   \hline
@@
   Additional results of evaluations across datasets among various metrics. (PDF 16 MB)
-  \subsection*{\textbf{Additional file 2: Supplementary methods}}
-  More detailed description of \textit{pipeComp} and comparison with other tools. (PDF 0.5 MB)
+  \subsection*{\textcolor{red}{\textbf{Additional file 2: Supplementary methods}}}
+  \textcolor{red}{More detailed description of \textit{pipeComp} and comparison with other tools. (PDF 0.5 MB)}
 \end{backmatter}

### 2020-05-13 bmc_article.tex
Updates from Overleaf

@@
 \parttitle{Conclusion} On the basis of our systematic evaluations of analysis pipelines, we make a number of practical recommendations about current analysis choices and for a comprehensive pipeline.
-The evaluation framework that we developed, \textit{pipeComp}, has been implemented so as to easily integrate any other step, tool, or evaluation metric allowing extensible benchmarks and easy applications to other fields of reseach in scRNAseq or Bioinformatics (\url{https://github.com/plger/pipeComp}).
+The evaluation framework that we developed, \textit{pipeComp} (\url{https://github.com/plger/pipeComp}), has been implemented so as to easily integrate any other step, tool, or evaluation metric allowing extensible benchmarks and easy applications to other fields of research in Bioinformatics, as we demonstrate through a study of the impact of removal of unwanted variation on differential expression analysis.
 

### 2020-05-13 bmc_article.tex
Updates from Overleaf

@@
 
-\begin{abstract} % abstract
-%\parttitle{Background} %if any
-%The massive growth of single-cell RNA-sequencing (scRNAseq) and methods for its analysis still lacks sufficient and up-to-date benchmarks that would guide analytical choices.
-Moreover, current studies are often focused on isolated steps of the process.
-Here, we present a flexible R framework for pipeline comparison with multi-level evaluation metrics and apply it to the benchmark of scRNAseq analysis pipelines using datasets with known cell identities.
-We evaluate common steps of such analyses, including filtering, doublet detection, normalization, feature selection, denoising, dimensionality reduction and clustering.
-On the basis of these analyses, we make a number of concrete recommendations about current analysis choices.
-The evaluation framework, \textit{pipeComp}, has been implemented so as to easily integrate any other step or tool, allowing extensible benchmarks and easy application to other fields (\url{https://github.com/plger/pipeComp}).
+\begin{abstract}
 
@@
 In contrast, too lenient filtering (or lack thereof) can lead to misclassified cells, as was the case for the Zhengmix4 datasets (Figure \ref{fig:filtering}B).
-We found that a relatively mild filtering (`default' - see methods) provided a good trade-off, often leading to high clustering performance (e.g.
+We found that a relatively mild filtering (\textit{default} - see methods) provided a good trade-off, often leading to high clustering performance (e.g.
 simMix and Zhengmix4 datasets) while retaining most of the cells in each subpopulation (e.g.
@@
 Koh, Kumar, and the simulations) are those in which heterotypic doublets are not expected.
-Overall, we therefore recommend the use of doublet removal followed by our `default' filtering (see Methods) or a similar approach.}
+Overall, we therefore recommend the use of doublet removal followed by our \textit{default} filtering (see Methods) or a similar approach.}
 
@@
 
-{\color{red}\subsection*{Application of \textif{pipeComp} to a different context}
+{\color{red}\subsection*{Application of \textit{pipeComp} to a different context}
 
@@
      \item Doublet detection and removal is advised and can be performed at little computing cost with software such as \textit{scDblFinder} or \textit{scds}.
-     \item Distribution-based cell filtering fails to capture doublets and should use relatively lenient cutoffs (e.g., 5 MADs, or 3 MADs in at least 2 distributions \textcolor{ref}{- see for instance our 'default' filters in the methods}) to exclude poor-quality cells while avoiding bias against some subpopulations.
+     \item Distribution-based cell filtering fails to capture doublets and should use relatively lenient cutoffs (e.g., 5 MADs, or 3 MADs in at least 2 distributions \textcolor{ref}{- see for instance our \textit{default} filters in the methods}) to exclude poor-quality cells while avoiding bias against some subpopulations.
      \item Features filtering based on feature type did not appear beneficial.

### 2020-05-13 bmc_article.tex
Updates from Overleaf

@@
 Since filtering changes the relative abundance of the different subpopulations, global clustering accuracy metrics such as ARI are not appropriate here.
-We therefore calculated the per-subpopulation precision and recall using the Hungarian algorithm \cite{PapadimitriouHu1998} and monitored the mean F1 score {\color{red}across subpopulations.
+We therefore calculated the per-subpopulation precision and recall using the Hungarian algorithm \cite{kuhnHungarianMethod1955} and monitored the mean F1 score {\color{red}across subpopulations.
 A first observation was that, although more stringent filtering tended to be associated with an increase in accuracy, it tended to plateau and could also become deleterious.
@@
      \item Doublet detection and removal is advised and can be performed at little computing cost with software such as \textit{scDblFinder} or \textit{scds}.
-     \item Distribution-based cell filtering fails to capture doublets and should use relatively lenient cutoffs (e.g., 5 MADs, or 3 MADs in at least 2 distributions \textcolor{ref}{- see for instance our \textit{default} filters in the methods}) to exclude poor-quality cells while avoiding bias against some subpopulations.
+     \item Distribution-based cell filtering fails to capture doublets and should use relatively lenient cutoffs (e.g., 5 MADs, or 3 MADs in at least 2 distributions \textcolor{red}{- see for instance our \textit{default} filters in the methods}) to exclude poor-quality cells while avoiding bias against some subpopulations.
      \item Features filtering based on feature type did not appear beneficial.
@@
 Analyses were performed in R 3.6.0 (Bioconductor 3.9) and the following packages were installed from GitHub repositories: \textit{Seurat} (version 3.0.0), \textit{SCTransform} (0.2.0), \textit{DoubletFinder} (2.0.1), \textit{scDblFinder} (1.1.4), \textit{scds} (1.0.0), \textit{SAVERX} (1.0.0), \textit{scImpute} (0.0.9), \textit{ALRA} (commit 7636de8), \textit{DCA} (0.2.2), \textit{DrImpute} (1.2), \textit{ENHANCE} (R version, commit 1571696), \textit{SAVERX} (1.0.0).
-{\color{red}The glmPCA was performed using the \texti{glmpca} package, while the code for deviance calculation was obtained from \url{https://github.com/willtownes/scrna2019} (commit 1ddcc30ebb95d083a685f12fe81d35dd1b0cb1b2).}
+{\color{red}The glmPCA was performed using the \textit{glmpca} package, while the code for deviance calculation was obtained from \url{https://github.com/willtownes/scrna2019} (commit 1ddcc30ebb95d083a685f12fe81d35dd1b0cb1b2).}
 

### 2020-05-25 draft.tex
fixed latex errors

--- 2020-03-07 draft.tex
@@
 
-The massive growth of single-cell RNA-sequencing (scRNAseq) and methods for its analysis still lacks sufficient and up-to-date benchmarks that would guide analytical choices.
-Moreover, current studies are often focused on isolated steps of the process.
-Here, we present a flexible R framework for pipeline comparison with multi-level evaluation metrics and apply it to the benchmark of scRNAseq analysis pipelines using datasets with known cell identities.
-We evaluate common steps of such analyses, including filtering, doublet detection, normalization, feature selection, denoising, dimensionality reduction and clustering.
-On the basis of these analyses, we make a number of concrete recommendations about analysis choices.
-The evaluation framework, \textit{pipeComp}, has been implemented so as to easily integrate any other step or tool, allowing extensible benchmarks and easy application to other fields (\url{https://github.com/plger/pipeComp}).
+The massive growth of single-cell RNA-sequencing (scRNAseq) and the methods for its analysis still lack sufficient and up-to-date benchmarks that could guide analytical choices.
+Numerous benchmark studies already exist and cover most of scRNAseq processing and analytical methods but only a few give advice on a comprehensive pipeline.
+Moreover, current studies often focused on isolated steps of the process and do not address the impact of a tool on both the intermediate and the final steps of the analysis.
+Here, we present a flexible R framework for pipeline comparison with multi-level evaluation metrics.
+We apply it to the benchmark of scRNAseq analysis pipelines using simulated and real datasets with known cell identities, covering common methods of filtering, doublet detection, normalization, feature selection, denoising, dimensionality reduction and clustering.
+We evaluate the choice of these tools with multi-purpose metrics to assess their ability to reveal cell population structure and lead to efficient clustering.
+On the basis of our systematic evaluations of analysis pipelines, we make a number of practical recommendations about current analysis choices and for a comprehensive pipeline.
+The evaluation framework that we developed, \textit{pipeComp} (\url{https://github.com/plger/pipeComp}), has been implemented so as to easily integrate any other step, tool, or evaluation metric allowing extensible benchmarks and easy applications to other fields of research in Bioinformatics, as we demonstrate through a study of the impact of removal of unwanted variation on differential expression analysis.
 
@@
 \section*{Background}
-
 Single-cell RNA-sequencing (scRNAseq) and the set of attached analysis methods are evolving fast, with more than 560 software tools available to the community \cite{ZappiaDB2018}, roughly half of which are dedicated to tasks related to data processing such as clustering, ordering, dimension reduction or normalization.
@@
 
-A number of good comparison and benchmark studies have already been performed on various steps related to scRNAseq processing and analysis and can guide the choice of methodology (\cite{CobosDeconvolution2020, colePerformance2018, DalMolinDE2017, duoClustering2018, freytagComparison2018, GaoProcessing2020, HeiserDimred2019, HouImput2020, JaakkolaDE2017, KrzakClust2019, SonesonDE2018, SunDimRed2019, tianMixology2018, TranBatch2020, TsuyuzakiPCA2020, viethSystematic2019, WangDE2019, YipHVGs2018, ZhangImput2018}).
+A number of good comparison and benchmark studies have already been performed on various steps related to scRNAseq processing and analysis and can guide the choice of methodology (\citealp{CobosDeconvolution2020, colePerformance2018, DalMolinDE2017, duoClustering2018, freytagComparison2018, GaoProcessing2020, HeiserDimred2019, HouImput2020, JaakkolaDE2017, KrzakClust2019, SonesonDE2018, SunDimRed2019, tianMixology2018, TranBatch2020, TsuyuzakiPCA2020, viethSystematic2019, WangDE2019, YipHVGs2018, ZhangImput2018}).
 However these recommendations need constant updating and often leave open many details of an analysis.
-Another missing aspect of current benchmarking studies is their limitation to capture all aspects of scRNAseq processing workflow.
-Although previous benchmarks already brought valuable recommendations for data processing, some only focused on one aspect of data processing (e.g., \cite{SunDimRed2019}), did not evaluate how the tool selection affects downstream analysis (e.g., \cite{TsuyuzakiPCA2020}) or did not tackle all aspects of data processing, such as doublet identification or cell filtering (e.g., \cite{viethSystematic2019}).
+Another missing aspect of current benchmarking studies is their limitation to capture all aspects of a scRNAseq processing workflow.
+Although previous benchmarks already brought valuable recommendations for data processing, some only focused on one aspect of data processing (e.g., \citealp{SunDimRed2019}), did not evaluate how the tool selection affects downstream analysis (e.g., \citealp{TsuyuzakiPCA2020}) or did not tackle all aspects of data processing, such as doublet identification or cell filtering (e.g., \citealp{viethSystematic2019}).
 A thorough evaluation of the tools covering all major processing steps is however urgently needed as previous benchmarking studies highlighted that a combination of tools can have a drastic impact on downstream analysis, such as differential expression analysis and cell-type deconvolution\cite{viethSystematic2019,CobosDeconvolution2020}.
@@
 
-Here, we develop a flexible R framework for pipeline comparison and evaluate the various steps of analysis leading from an initial count matrix to a cluster assignment, which are critical in a wide range of applications.
+Here, we develop a flexible R framework for pipeline comparison and apply it to the evaluation of the various steps of analysis leading from an initial count matrix to a cluster assignment, which are critical in a wide range of applications.
 We collected real datasets of known cell composition (Table \ref{tab:table1}) and used a variety of evaluation metrics to investigate in a multilevel fashion the impact of various parameters and variations around a core scRNAseq pipeline.
 Although we use some datasets based on other protocols, our focus is especially on droplet-based datasets that do not include exogenous control RNA (i.e.
-spike-ins); see Table \ref{tab:table1} and Figure \ref{fig:figure1} for more details.
+spike-ins); see Table \ref{tab:table1} and Supplementary Figure 1 for more details.
 In addition to previously-used benchmark datasets with true cell labels \cite{duoClustering2018,tianMixology2018}, we simulated two datasets with a hierarchical subpopulation structure based on real 10x human and mouse data using \textit{muscat} \cite{CrowellMuscat2019}.
@@
 
-The \textit{pipeComp} package defines a pipeline as, minimally, a list of functions executed consecutively on the output of the previous one (Figure \ref{fig:figure2}A).
+The \textit{pipeComp} package defines a pipeline as, minimally, a list of functions executed consecutively on the output of the previous one {\color{red}(Figure \ref{fig:explanation}A; see also Supplementary methods).
 In addition, optional benchmark functions can be set for each step to provide standardized, multi-layered evaluation metrics.
-Given such a \texttt{PipelineDefinition} object, a set of alternative parameters (which might include different subroutines) and benchmark datasets, the \texttt{runPipeline} function then proceeds through all combinations of arguments, avoiding recomputing the same step twice and compiling evaluations (including run time) on the fly.
-Variations in a given parameter can be evaluated using all metrics from this point downward in the pipeline.
+Given such a \texttt{PipelineDefinition} object, a set of alternative parameters (which might include different subroutines calling different methods) and benchmark datasets, the \texttt{runPipeline} function then proceeds through all combinations of arguments (or a desired subset), avoiding recomputing the same step twice without the need to save all alternative intermediates, and compiling evaluations (including run time) on the fly (see Supplementary methods for more detail).} Variations in a given parameter can be evaluated using all metrics from this point downward in the pipeline.
 This is especially important because end-point metrics, such as the adjusted Rand index (ARI) \cite{HubertARI1985} for clustering, are not perfect.
-For example, although the meaning of an ARI score is independent of the number of true subpopulations \cite{steinleyProperties2004}, the number of clusters called is by far the most important determinant of the score: the farther it is from the actual number of subpopulations, the worse the ARI (Supplementary Figures 1-2).
+For example, although the meaning of an ARI score is independent of the number of true subpopulations \cite{steinleyProperties2004}, the number of clusters called is by far the most important determinant of the score: the farther it is from the actual number of subpopulations, the worse the ARI (Supplementary Figures 2-3).
 In this context, one strategy has been to cluster across various resolutions and only consider the results that have the right number of clusters \cite{duoClustering2018}.
@@
 This can for instance be performed through the \texttt{isOutlier} function from \textit{scater} that measures, for a given control property, the number of median absolute deviations (MADs) of each cell from the median of all cells.
-Supplementary Figure 4 shows the distributions of some of the typical cell properties commonly used.
+Supplementary Figure 5 shows the distributions of some of the typical cell properties commonly used.
 Of note, these properties tend to be correlated, with some exceptions.
-For example, while a high proportion of mitochondrial reads is often correlated with a high proportion of the counts in the top features, there can also be other reasons for an over-representation of highly-expressed features (Supplementary Figure 5), such as an over-amplification in non-UMI datasets.
-In our experience, 10X datasets also exhibit a very strong correlation between the total counts and the total features even across very different cell types (Supplementary Figure 6).
+For example, while a high proportion of mitochondrial reads is often correlated with a high proportion of the counts in the top features, there can also be other reasons for an over-representation of highly-expressed features (Supplementary Figure 6), such as an over-amplification in non-UMI datasets.
+In our experience, 10X datasets also exhibit a very strong correlation between the total counts and the total features even across very different cell types (Supplementary Figure 7).
 We therefore also measure the ratio between the two and treat cells strongly departing from this trend with suspicion.
 
-Reasoning that the cells we wish to exclude are the cells that would be misclassified, we measured the rate of misclassification of each cell in each dataset across a variety of clustering pipelines, correcting for the median misclassification rate of the subpopulation and then evaluated what properties could be predictive of misclassification (Supplementary Figures 7-9).
+Reasoning that the cells we wish to exclude are the cells that would be misclassified, we measured the rate of misclassification of each cell in each dataset across a variety of clustering pipelines, correcting for the median misclassification rate of the subpopulation and then evaluated what properties could be predictive of misclassification (Supplementary Figures 8-10).
 We could not identify any property or simple combination thereof that would be consistently predictive of misclassification; the only feature that consistently stood out across multiple datasets (the Zheng datasets) was that cells with very high read counts have a higher chance of being misclassified.
 
-We next investigated the impact of filtering according to various criteria (see Methods).
+{\color{red}We next investigated the impact of several filtering methods (see Methods for more details): four methods based on deviations to MADs with increasing levels of stringency (named \textit{lenient}, \textit{default}, \textit{stringent}, \textit{veryStringent}) and two methods based on \textit{scater}'s \textit{runPCA} using all (\textit{pca.all}) or selected covariates (\textit{pca.sel})}.
 An important risk of excluding cells on the basis of their distance from the whole distribution of cells on some properties (e.g., library size) is that these properties tend to have different distributions across subpopulations.
-As a result, thresholds in terms of number of MADs from the whole distribution can lead to strong biases against certain subpopulations (Figure \ref{fig:figure4}A).
-We therefore examined the tradeoff between the increased accuracy of filtering and the maximum proportion of cells excluded per subpopulation (Figure \ref{fig:figure4}B and Supplementary Figure 10).
+As a result, thresholds in terms of number of MADs from the whole distribution can lead to strong biases against certain subpopulations (Figure \ref{fig:filtering}A).
+We therefore examined the tradeoff between the increased accuracy of filtering and the maximum proportion of cells excluded per subpopulation (Figure \ref{fig:filtering}B and Supplementary Figure 11).
 Since filtering changes the relative abundance of the different subpopulations, global clustering accuracy metrics such as ARI are not appropriate here.
-We therefore calculated the per-subpopulation precision and recall using the Hungarian algorithm \cite{PapadimitriouHu1998} and monitored their mean F1 score.
-A first observation was that, although the stringent filtering tended to be associated with an increase in accuracy, it could also become deleterious and most of the benefits could be achieved without very stringent filtering and minimizing subpopulation bias (Figure \ref{fig:figure4}B).
-Applying the same filtering criteria on individual clusters of cells (identified through scran's \texttt{quickCluster} method) resulted in nearly no cell being filtered out.
-This suggests that filtering on the global population tends to discard cells of subpopulations with more extreme properties (e.g.
+We therefore calculated the per-subpopulation precision and recall using the Hungarian algorithm \cite{kuhnHungarianMethod1955} and monitored the mean F1 score {\color{red}across subpopulations.
+A first observation was that, although more stringent filtering tended to be associated with an increase in accuracy, it tended to plateau and could also become deleterious.
+Most of the benefits could be achieved without very stringent filtering and minimizing subpopulation bias (Figure \ref{fig:filtering}B).
+Applying the same filtering criteria on individual clusters of cells (identified through scran's \texttt{quickCluster} method) rather than on whole dataset resulted in nearly no cells being filtered out (Supplementary Figure 12A).
+This suggests that stringent filtering on the global population tends to discard cells of subpopulations with more extreme properties (e.g.
 high library size), rather than low-quality cells.
-Finally, by changing the distributions, the doublet removal step in conjunction with filtering sometimes resulted in a net decrease in the proportion of excluded cells while retaining or improving accuracy.
-We therefore recommend the use of doublet removal followed by relatively mild filtering, such as is implemented in our `default' filtering (see Methods).
+In contrast, too lenient filtering (or lack thereof) can lead to misclassified cells, as was the case for the Zhengmix4 datasets (Figure \ref{fig:filtering}B).
+We found that a relatively mild filtering (\textit{default} - see methods) provided a good trade-off, often leading to high clustering performance (e.g.
+simMix and Zhengmix4 datasets) while retaining most of the cells in each subpopulation (e.g.
+Koh and Zhengmix8eq).
+PCA-based selection did not appear superior to feature-wise filters.
+We found very little overlap between the cells excluded by doublet removal and those excluded by MAD-based filtering (Supplementary Figure 12B).
+Of note, doublet removal in conjunction with filtering tended to improve accuracy not only in datasets that do include heterotypic doublets (mixology datasets), but also some others in which such doublets are unlikely given the experimental design (e.g.
+FACS-sorted datasets).
+Indeed, the datasets where doublet removal did not have a clear positive impact (e.g.
+Koh, Kumar, and the simulations) are those in which heterotypic doublets are not expected.
+Overall, we therefore recommend the use of doublet removal followed by our \textit{default} filtering (see Methods) or a similar approach.}
 
@@
 We next investigated the impact of different normalization strategies.
-Beside the standard log-normalization included in \textit{Seurat}, we tested \textit{scran}'s pooling-based normalization \cite{lunPooling2016}, \textit{sctransform}'s variance-stabilizing transformation \cite{hafemeisterSCtransform2019}, normalization based on stable genes \cite{linStableGenes2018, deekeStablyExpressed2018}, and SCnorm \cite{bacherSCnorm2017} (\textit{scVI} \cite{lopezDeep2018} was included separately on a subset of parameter combinations due to high running times).
+Beside the standard log-normalization included in \textit{Seurat}, we tested \textit{scran}'s pooling-based normalization \cite{lunPooling2016}, \textit{sctransform}'s variance-stabilizing transformation \cite{hafemeisterSCtransform2019}, normalization based on stable genes \cite{linStableGenes2018, deekeStablyExpressed2018}, and \textit{SCnorm} \cite{bacherSCnorm2017} (\textit{scVI} \cite{lopezDeep2018} was included separately on a subset of parameter combinations due to high running times).
 In addition to log-normalization, the standard \textit{Seurat} clustering pipeline performs per-feature unit-variance scaling so that the PCA is not too strongly dominated by highly-expressed features.
@@
 However, as library size tends to vary across subpopulations, part of this effect can simply reflect biological differences.
-We therefore assessed to what extent the first principal component still retained a correlation with the library size and the number of detected features, removing the confounding covariation with the subpopulations (Figure \ref{fig:figure5}A and Supplementary Figure 12).
+We therefore assessed to what extent the first principal component still retained a correlation with the library size and the number of detected features, removing the confounding covariation with the subpopulations (Figure \ref{fig:norm}A and Supplementary Figure 14).
 Scaling tended to remove much of the correlation with these features, and while most methods were able to remove most of the effect, the correlation was lowest with \textit{sctransform}.
@@
 
-We further evaluated normalization methods by investigating their impact on the separability of the subpopulations (Figure \ref{fig:figure5}B-C).
-Since clustering accuracy metrics such as the ARI are very strongly influenced by the number of clusters, we complemented it with silhouette width \cite{RousseeuwSil1987} and mutual information (MI).
-We found most methods (including no normalization at all) to perform fairly well in most of the subpopulations.
-An exception was \textit{scVI} (Supplementary Figure 13), which performed poorly on most datasets in terms of both the subpopulation silhouette and clustering accuracy.
-For time and resource cost reasons, \textit{scVI} was run on a reduced set of pipelines and compared to \textit{sctransform} (more details in the Methods section).
-Scaling tended to reduce the average silhouette width of some subpopulations and to increase that of some less distinguishable ones and was generally, but not always, beneficial on the accuracy of the final clustering.
-Regressing out covariates systematically gave poorer performance on all metrics.
-\textit{sctransform} systematically outperformed other methods and, even though it was developed to be applied to data with unique molecular identifiers (UMI), it also performed fairly well with the Smart-seq protocol (Koh and Kumar datasets).
+We further evaluated normalization methods by investigating their impact on the separability of the subpopulations.
+Since clustering accuracy metrics such as the ARI are very strongly influenced by the number of clusters, we complemented it with silhouette width \cite{RousseeuwSil1987} and mutual information (MI, Figure \ref{fig:norm}B-C).
+We found most methods (including no normalization at all) to perform fairly well in most of the subpopulations{\color{red}, and that some datasets (e.g.
+Zhengmix8eq and simMix1) and subpopulations were consistently more challenging across all methods.
+An exception was \textit{scVI}, which performed poorly on most datasets in terms of both the subpopulation silhouette and clustering accuracy (Supplementary Figure 13; \textit{scVI} was run on a reduced set of pipelines).
+Scaling (e.g.
+`seurat (no scale)' vs `seurat') tended to reduce the average silhouette width of some subpopulations and to increase that of some less distinguishable ones and was generally, but not always, beneficial on the accuracy of the final clustering.
+Regressing out covariates systematically gave equal or poorer performance on MI, ARI and ARI at true number of clusters, with a strongest negative impact among the Smart-seq datasets (Koh and Kumar).
+Regressing out covariates using \textit{sctransform} instead, tended to have a much milder effect, possibly owing to its regularization.
+In general, \textit{sctransform} systematically outperformed other methods and, even though it was developed to be applied to data with unique molecular identifiers (UMI), it also performed fairly well with the Smart-seq protocol (Koh and Kumar datasets).}
 
 Finally, we monitored whether, under the same downstream clustering analysis, different normalization methods tended to lead to an over- or under-estimation of the number of clusters.
-Although some methods had a tendency to lead to a higher (e.g., \textit{sctransform}) or lower (e.g., stable genes) number of clusters, the effect was very mild and not entirely systematic (Supplementary Figure 14).
-We also confirmed that, especially in the UMI-based datasets, \textit{sctransform} did in fact successfully stabilize variance across mean counts (Supplementary Figure 15), at little apparent cost on computation time (Supplementary Figure 16).
+Although some methods had a tendency to lead to a higher (e.g., \textit{sctransform}) or lower (e.g., stable genes) number of clusters, the effect was very mild and not entirely systematic (Supplementary Figure 16).
+We also confirmed that, especially in the UMI-based datasets, \textit{sctransform} did in fact successfully stabilize variance across mean counts (Supplementary Figure 17), at little apparent cost on computation time (Supplementary Figure 18).
 
@@
 As the proportion of variability in a gene attributable to subpopulations can be measured in various ways, we first compared three approaches: ANOVA on log-normalized count, ANOVA on \textit{sctransform} normalization, and deviance explained.
-The ANOVAs performed on a standard \textit{Seurat} normalization and on \textit{sctransform} data were highly correlated (Supplementary Figure 17A).
-These estimates were also in good agreement with the deviance explained, although lowly-expressed genes could have a high deviance explained without having much of their variance explained by subpopulation (Supplementary Figure 17B-D).
-We therefore compared the proportion of the cumulative variance/deviance explained by the top X genes that could be retrieved by each gene ranking method (Supplementary Figures 18-19).
-We first focused on the first 1000 genes to highlight the differences between methods, although a higher number of selected genes decreased the differences between methods (Supplementary Figures 18-20).
+The ANOVAs performed on a standard \textit{Seurat} normalization and on \textit{sctransform} data were highly correlated ({\color{red}Pearson correlation >0.9,} Supplementary Figure 19A).
+These estimates were also in good agreement with the deviance explained, although lowly-expressed genes could have a high deviance explained without having much of their variance explained by subpopulation (Supplementary Figure 19B-D).
+We therefore compared the proportion of the cumulative variance/deviance explained by the top X genes that could be retrieved by each gene ranking method (Supplementary Figures 20-21).
+We first focused on the first 1000 genes to highlight the differences between methods, although a higher number of selected genes decreased the differences between methods (Supplementary Figures 20-22).
 The standardized measures of variability were systematically worse than their non-standardized counterparts in selecting genes with a high proportion of variance explained by subpopulation.
-Regarding the percentage of deviance explained however, the standardized measures were often superior (Figure \ref{fig:figure6}A and Supplementary Figures 18-19).
-Deviance proved the method of choice to prioritize genes with a high variance explained by subpopulations (with mere expression level proving surprisingly performant) but did not perform so well to select genes with a high deviance explained.
+Regarding the percentage of deviance explained however, the standardized measures were often superior (Figure \ref{fig:selection}A and Supplementary Figures 20-21).
+\textit{Deviance} proved the method of choice to prioritize genes with a high variance explained by subpopulations (with mere expression level proving surprisingly performant) but did not perform so well to select genes with a high deviance explained.
 
-We next evaluated how the use of different feature selection methods affected the clustering accuracy (Figure \ref{fig:figure6}B).
+We next evaluated how the use of different feature selection methods affected the clustering accuracy (Figure \ref{fig:selection}B).
 To validate the previous assay on the proportion of variance/deviance explained by real populations, we included genes that maximized these two latter measures.
-Interestingly, while these selections were on average the top-ranking methods, they were not systematically best for all datasets.
-The previous observations were reflected in the ARI of the resulting clustering (Figure \ref{fig:figure6}B): non-standardized measures of variability, including mere expression level, tended to outperform more complex metrics.
+Interestingly, while these selections were on average the top-ranking methods, they were not systematically best for all datasets (Figure \ref{fig:selection}B).
+Non-standardized measures of variability, including mere expression level, tended to outperform more complex metrics.
 In general, we found deviance and unstandardized estimates of variance to provide the best results across datasets and normalization methods.
-Increasing the number of features selected also systematically led to an increase in the accuracy of the clustering, typically plateauing after 4000 features (Supplementary Figure 20).
+Increasing the number of features selected also {\color{red}tended to lead} to an increase in the accuracy of the clustering, typically plateauing after 4000 features (Supplementary Figure 22).
 
@@
 
-The impact of the choice of dimensionality reduction method was far greater than that of normalization or feature selection (Figure \ref{fig:figure7}).
+The impact of the choice of dimensionality reduction method was far greater than that of normalization or feature selection (Supplementary Figure 23).
+{\color{red}Overall, weighting the components by their explained variance tended to increase silhouette width (Supplementary Figure 23A) and, in most datasets, clustering accuracy (Supplementary} Figure 23B).
 GLM-PCA tended to increase the average silhouette width of already well-defined subpopulations, but \textit{Seurat}'s PCA procedure however proved superior on all metrics.
 Like GLM-PCA, \textit{scVI}'s Linear Decoder (LD) does not explicitly rely on normalized counts.
-Its performance was lower than a pipeline with non-weighted \textit{Seurat} PCA on all measures for datasets with more subpopulations (Supplementary Figure 13).
-Overall, weighting the principal components by their variance (as \textit{Seurat} does) had a positive impact on silhouette widths and ARI scores.
+{\color{red}Its performance was however the worst both in terms of silhouette width and clustering accuracy (Supplementary Figure 15), and it only performed well in some of the least challenging datasets (mixology10x5cl and simMix2).}
 
@@
 Since Euclidean distance decreases as the number of non-discriminating dimensions increases, there is usually a trade-off between selecting enough dimensions to keep most information and excluding smaller dimensions that may represent technical noise or other unwanted sources of variation.
-Overall, increasing the number of dimensions robustly led to a decrease in the number of clusters (Supplementary Figures 14 and 21).
-This tended to affect the accuracy of the clustering (Supplementary Figure 22), although in both cases (number of clusters and ARI), \textit{Seurat}'s resolution parameter had a much stronger impact.
+Overall, increasing the number of dimensions robustly led to a decrease in the number of clusters (Supplementary Figures 16 and 24).
+This tended to affect the accuracy of the clustering (Supplementary Figure 25), although in both cases (number of clusters and ARI), \textit{Seurat}'s resolution parameter had a much stronger impact.
 
-Different approaches have been proposed to select the appropriate number of dimensions, from the visual identification of an 'Elbow' (inflexion point) of the variance explained, to more complex algorithms.
-We evaluated the performance of dimensionality estimators implemented in the \textit{intrinsicDimension} package \cite{johnssonDimensionality2015}, as well as common procedures such as the `elbow` method (inflexion point in the variance explained by each component), some scRNAseq-specific methods such as the JackStraw procedure \cite{ChungJackstraw2015} or \textit{scran}'s \texttt{denoisePCA} \cite{LunScran2016}, and the recent application of Fisher Separability analysis \cite{AlberganteSepar2019}.
+Different approaches have been proposed to select the appropriate number of dimensions, from the visual identification of an {\color{red}inflexion point in the curve of the variance explained by each component (`elbow' method), to more complex algorithms.} We evaluated the performance of dimensionality estimators implemented in the \textit{intrinsicDimension} package \cite{johnssonDimensionality2015}, as well as common procedures such as the `elbow` method, some scRNAseq-specific methods such as the JackStraw procedure \cite{ChungJackstraw2015} or \textit{scran}'s \texttt{denoisePCA} \cite{LunScran2016}, and the recent application of Fisher Separability analysis \cite{AlberganteSepar2019}.
 
-We compared the various estimates of dimensionality in their ability to retrieve the intrinsic number of dimensions in a dataset, based on Seurat's weighted PCA space.
-As a first approximation of the true dimensionality, we computed the variance in each principal component that was explained by the subpopulations, which sharply decreased after the first few components in most datasets (Figure \ref{fig:figure8}A).
-Figure \ref{fig:figure8}B shows the difference between the dimension estimates of the above methods and that based on the subpopulations (i.e.
-from Figure \ref{fig:figure8}A).
-Of note, the methods differ widely in view of their computing time (Figure \ref{fig:figure8}B) and we saw no relationship between the accuracy of the estimates and the complexity of the method.
-Reasoning that over-estimating the number of dimensions is less problematic than under-estimating it, we kept the former methods for a full analysis of their impact on clustering (Figure \ref{fig:figure8}C-D), when combined with \textit{sctransform} or the standard \textit{Seurat} normalization.
-Although most methods performed well on the various clustering measures, the global maximum likelihood based on translated Poisson Mixture Model (\textit{maxLikGlobal}, using the 20 nearest neighbors, although estimates were relatively stable across decent values of \textit{k}; see Supplementary Figure 23) provided the dimensionality estimate that best separated the subpopulations (Figure \ref{fig:figure8}C) and resulted in the best clustering accuracy (Figure \ref{fig:figure8}D).
-This method systematically selected many more components than were associated with the subpopulations, suggesting that although these additional components appear individually uninformative, in combination they nevertheless contribute to classification.
-An implication is that the commonly-used elbow method most likely underestimates dimensionality.
+{\color{red}We compared the various estimates in their ability to retrieve the intrinsic number of dimensions in a dataset, based on Seurat's weighted PCA space.
+As a first approximation of the true dimensionality, we computed the variance in each principal component (based on Seurat's weighted PCA space) that was explained by the subpopulations, which sharply decreased after the first few components in most datasets (Figure \ref{fig:dim}A).
+These truth-based estimates were then compared to those of the above dimensionality estimation methods (Figure \ref{fig:dim}B).} Of note, the methods differ widely in compute time (Figure \ref{fig:dim}B) and we saw no relationship between the accuracy of the estimates and the complexity of the method.
+Reasoning that over-estimating the number of dimensions is less problematic than under-estimating it, we kept the former methods for a full analysis of their impact on clustering (Figure \ref{fig:dim}C-D), when combined with \textit{sctransform} or standard \textit{Seurat} normalization.
+Although most methods performed well on the various clustering measures, the global maximum likelihood based on translated Poisson Mixture Model {\color{red}(\textit{maxLikGlobal}, using 10 or 20 nearest neighbors, although estimates were relatively stable across various values of \textit{k}; see Supplementary Figure 26) provided the dimensionality estimate that best separated challenging subpopulations (Figure \ref{fig:dim}C) and tended to result in the best clustering accuracy (Figure \ref{fig:dim}D).
+This method systematically selected many more components than were associated with the subpopulations in Figure \ref{fig:dim}A, suggesting that although these additional components appear individually uninformative, in combination they nevertheless contribute to classification.
+An important implication is that the commonly-used elbow method most likely underestimates dimensionality.}
 
@@
 
-Since ARI is dominated by differences in the number of clusters (Supplementary Figures 1-2) and no single metric is perfect, we diversified them (Figure \ref{fig:figure9}).
-MI has the virtue of not decreasing when a true subpopulation is split into two clusters, which is arguably less problematic (and might well reflect unknown biological subgroups), but as a consequence it can be biased towards methods producing higher resolution clustering (Supplementary Figure 2).
+Since ARI is dominated by differences in the number of clusters (Supplementary Figures 2-3) and no single metric is perfect, we diversified them (Figure \ref{fig:clust}).
+MI has the virtue of not decreasing when a true subpopulation is split into two clusters, which is arguably less problematic (and might well reflect unknown biological subgroups), but as a consequence it can be biased towards methods producing higher resolution clustering (Supplementary Figure 3).
 Similarly, precision per true subpopulation is considerably more robust to differences in the number of clusters.
@@
 
-The MI score and minimum precision, which are largely independent of the estimated number of clusters, were overall higher for the \textit{walktrap} method (Figure \ref{fig:figure9}), while the mean F1 score favored both \textit{scran} methods (\textit{walktrap} and \textit{fast greedy}) over \textit{Seurat}.
+The MI score and minimum precision, which are largely independent of the estimated number of clusters, were overall higher for the \textit{walktrap} method (Figure \ref{fig:clust}), while the mean F1 score favored both \textit{scran} methods (\textit{walktrap} and \textit{fast greedy}) over \textit{Seurat}.
 Finally, the ARI score at the true number of clusters, when available, showed similar performances.
-However, because \textit{Seurat}'s resolution parameter had a large impact on the number of clusters identified (Supplementary Figures 1 and 23), \textit{Seurat} could always be coerced into producing the right number of clusters.
-Instead, the number of clusters found by \textit{scran} was considerably less influenced by the available parameters (number of nearest neighbors or steps in the random walk - see Supplementary Figure 24), and as a result \textit{scran}-based clustering sometimes never produced a partitioning with the right number of clusters.
+However, because \textit{Seurat}'s resolution parameter had a large impact on the number of clusters identified (Supplementary Figures 2 and 24), \textit{Seurat} could always be coerced into producing the right number of clusters.
+Instead, the number of clusters found by \textit{scran} was considerably less influenced by the available parameters (number of nearest neighbors or steps in the random walk - see Supplementary Figure 27), and as a result \textit{scran}-based clustering sometimes never produced a partitioning with the right number of clusters.
 This observation, along with \textit{scran}'s higher MI score, suggest that \textit{scran} sometimes simply divides a real subpopulation into two clusters (possibly tracking some unknown biological differences) rather than committing misclassification errors.
 Overall, the \textit{walktrap} method appeared superior to the \textit{fast greedy} algorithm and was generally less prone to misclassification than \textit{Seurat} clustering, although the latter offered more control over the resolution.
-A major difference between \textit{walktrap}-based clustering and \textit{Seurat} is the computing time (Figure \ref{fig:figure9}B).
+A major difference between \textit{walktrap}-based clustering and \textit{Seurat} is the computing time (Figure \ref{fig:clust}B).
 We found that using the Annoy approximation to the nearest neighbors somewhat reduced computing time at no apparent cost to accuracy.
@@
 For all methods, some poorly distinguishable subpopulations from both the Zhengmix8eq and simMix1 datasets remained very inaccurately classified in regard to all metrics.
+
+{\color{red}\subsection*{Overview of interactions between steps}
+
+We selected the most important alternatives for each parameter (see Methods) in order to explore their interactions.
+After accounting, in the model, for baseline differences between datasets and the impact of the resolution parameter, ANOVA attributed the largest proportion of variance in the main cluster metrics to the choice of dimensionality reduction, followed by normalization, and very little to filtering (Supplementary Figure 28A).
+Because the choice of dimension reduction impacts the number of clusters, we repeated the analysis of variance in ARI among the combinations of parameters that yielded the true number of clusters (Supplementary Figure 28B).
+We observed that although normalization was the most important parameter among these analyses, the choice of dimensionality came closely behind (Supplementary Figure 28B), suggesting that more attention should be paid to this relatively neglected step.
+
+We next investigated whether the benefits of some methods were conditional on the choices at other steps.
+Although analysis of variance identified many weak but significant interactions, we found that the benefits brought by performant methods were robust to differences in other parts of the pipeline (Supplementary Figures 29-30).
+For instance, we confirmed that doublet removal and \textit{sctransform} normalization were beneficial independently of other steps, and that, again independendently of other steps, more stringent filtering did not yield any major improvement over lenient filtering (Supplementary Figure 29).
+Similarly, deviance-based feature selection outcompeted \textit{Seurat}'s default selection method independently of changes at other steps (Supplementary Figure 30).}
 
@@
 Since preliminary analysis showed that all methods performed equally well or better on normalized data, we applied them after filtering and normalization, but before scaling and reduction.
-Although some of the methods (e.g., \textit{DRImpute\_process} and \textit{alra\_norm}) did improve the separability of some more elusive subpopulations, no method had a systematically positive impact on the average silhouette width across all subpopulations (Figure \ref{fig:figure10}A).
-When restricting ourselves to clustering analyses that yielded the `right' number of clusters, all tested methods improved classification  compared to a scenario with no imputation step (`none' label, Figure \ref{fig:figure10}B).
+Although some of the methods (e.g., \textit{DRImpute\_process} and \textit{alra\_norm}) did improve the separability of some more elusive subpopulations, no method had a systematically positive impact on the average silhouette width across all subpopulations (Supplementary Figure 31A).
+When restricting ourselves to clustering analyses that yielded the `right' number of clusters, all tested methods improved classification  compared to a scenario with no imputation step (`none' label, Supplementary Figure 31B).
 However, the situation was not so straightforward with alternative metrics, where some methods (e.g., \textit{ENHANCE}) consistently underperformed.
@@
 
+{\color{red}\subsection*{Application of \textit{pipeComp} to a different context}
+
+To illustrate the applicability of \textit{pipeComp} to a completely different problem, we benchmarked the impact of surrogate variable analysis (SVA)-based methods on bulk RNAseq differential expression analysis (DEA).
+The rationale behind such approaches is that the identification of factors determining genome-wide variation (i.e.
+surrogate variables) and their inclusion in the DEA's generalized linear model might improve DEA, especially in the presence of sources of variation unrelated to the experimental variables.
+\textit{sva} \citep{leekCapturingHeterogeneityGene2007,leekSvaPackageRemoving2012} and \textit{RUVseq} \citep{rissoNormalizationRNAseq2014a} are two popular packages each implementing different methods to infer these variables.
+We therefore created a \texttt{PipelineDefinition} with three steps: 1) feature filtering, 2) removal of unwanted variation (i.e.
+SVA), and 3) DEA using the most established methods  \citep{germainRNAontheBENCHComputationalEmpirical2016}.
+The package's `pipeComp\_dea' vignette details the creation process.
+We developed 3 benchmark datasets: 1) the \texttt{ipsc} data is a random selection of 10 vs 10 samples from \citep{carcamoOriveAnalysisTranscriptional2017}, which are very heterogeneous, and foldchanges were manually added to 300 genes; 2) the \texttt{seqc} dataset contains 5 vs 5 samples of two mixtures of the SEQC consortium with different ERCC spike-in mixes \citep{ComprehensiveAssessment2014}, and includes a partially correlated batch effect; 3) A \texttt{simulation} of 8 vs 8 samples with two technical vectors of variation (see Methods for more details).
+As previously reported \citep{ComprehensiveAssessment2014}, we found that more stringent filtering improved DEA (Supplementary Figure 32A) and that the different DEA methods had similar performances (Figure \ref{fig:svadea}A and Supplementary Figure 32A).
+Including a SVA-based step increased sensitivity across all datasets, particularly in datasets with technical vectors of variation (\textit{seqc} and \textit{simulation}), and benefited all DEA methods in a similar fashion (Figure \ref{fig:svadea}A,C and Supplementary Figures 32-33).
+With the exception of the \texttt{RUVr} method, which led to a high False Discovery Rate (FDR), all other methods showed comparable gains in sensitivity while maintaining a decent error control (if a little above the nominal FDR), even when increasing the number of surrogate variables used (Figure \ref{fig:svadea}B-C).
+While the FDR did increase in the \textit{ipsc} dataset, we suspect that much of the effect could be due to spurious differences \citep{germainTamingHumanGenetic2017} between the groups which we wrongly assume to be false positives.
+Nevertheless, the FDR remained acceptable with up to three variables.
+These results therefore suggest that these methods are robust and can be widely applied.}
+
 \section*{Discussion}
 
-\subsection*{Concrete recommendations}
+\subsection*{{\color{red}Practical} recommendations}
 
-On the basis of our findings, we can make a number of concrete analysis recommendations, also summarized in Figure \ref{fig:figure11}:
+On the basis of our findings, we can make a number of \textcolor{red}{practical} analysis recommendations, also summarized in Supplementary Figure 34:
 
@@
      \item Doublet detection and removal is advised and can be performed at little computing cost with software such as \textit{scDblFinder} or \textit{scds}.
-     \item Distribution-based cell filtering fails to capture doublets and should use relatively lenient cutoffs (e.g., 5 MADs, or 3 MADs in at least 2 distributions) to exclude poor-quality cells while avoiding bias against some subpopulations.
+     \item Distribution-based cell filtering fails to capture doublets and should use relatively lenient cutoffs (e.g., 5 MADs, or 3 MADs in at least 2 distributions \textcolor{red}{- see for instance our \textit{default} filters in the methods}) to exclude poor-quality cells while avoiding bias against some subpopulations.
      \item Features filtering based on feature type did not appear beneficial.
@@
    \begin{itemize}
-       \item Denoising appeared beneficial to the identification of subpopulations in 10x datasets, but not in Smart-seq datasets.
-       \item We found especially \textit{ALRA} (with prior normalization), \textit{DrImpute} (with prior processing) and \textit{DCA} to offer the best performances.
+       \item Denoising appeared beneficial to the identification of some subpopulations in 10x datasets, but not in Smart-seq datasets.
+       \item We found especially \textit{ALRA} (with prior normalization), \textit{DrImpute} (with prior processing) and \textit{DCA} to offer the best performances\textcolor{red}{, although none of these methods was beneficial over all subpopulations}.
    \end{itemize}
@@
      \item Similarly to previous reports \cite{SunDimRed2019}, we recommend the irlba-based PCA using weighting of the components, as implemented in \textit{Seurat}. 
-     \item We advise against the commonly-used elbow method (because it is too conservative) and the jackstraw method (low performance at great computational costs) for deciding on the components to include, and recommend a method like the \textit{global maximum likelihood based on translated Poisson Mixture Model} method (e.g., implemented in the  \texttt{maxLikGlobalDimEst} function of \texttt{intrinsicDimension}, in our case using the 20 nearest neighbors).
+     \item We advise against the commonly-used elbow method (because it is too conservative) and the jackstraw method (low performance at great computational costs) for deciding on the components to include, and recommend a method like the \textit{global maximum likelihood based on translated Poisson Mixture Model} method (e.g., implemented in the  \texttt{maxLikGlobalDimEst} function of \texttt{intrinsicDimension}, in our case using the \textcolor{red}{10 or }20 nearest neighbors).
    \end{itemize}
@@
 Some steps, such as the implementation of the PCA, were also not explored in detail here as they have already been the object of recent and thorough study elsewhere \cite{SunDimRed2019, TsuyuzakiPCA2020}.
-We also considered only methods relying on Euclidean distance, while correlation was recently reported to be superior\cite{kim_impact_2019} and would require further investigation.
+We also considered only methods relying on Euclidean distance, while correlation was recently reported to be superior \cite{kim_impact_2019} and would require further investigation.
+
+{\color{red}Among all tested pipelines, performance varied greatly across datasets.
+The ZhengMix8eq and simMix1 datasets were the most challenging datasets across all preprocessing steps that we tested.
+The average silhouette width of half of their subpopulations was systematically lower than the other subpopulations/ datasets and the ARI score at true number of clusters, if ever achieved, was always poor.
+It is likely that these two datasets are challenging due to their combined sparsity and complexity (see Supplementary Figure 1).
+The challenge of distinguishing subpopulations of T-cells in the Zhengmix8eq dataset was previously discussed \cite{duoClustering2018}.
+In contrast, while the Koh dataset harbors as many populations, they originate from more distinct groups (from hESC in different differentiation stages), while the Kumar and mixology datasets harbor a very low subpopulation complexity and most methods applied on them performed well.}
 
@@
 Instead, the \textit{CellBench} package was recently proposed to address a similar need \cite{suCellbench}, offering an elegant piping syntax to combine alternative methods at successive steps.
-A key additional feature of \textit{pipeComp} is its ability to perform evaluation in parallel to the pipeline and offering on-the-fly evaluation at any step.
+A key additional feature of \textit{pipeComp} is its ability to perform evaluation in parallel to the pipeline and offering on-the-fly evaluation at {\color{red}multiple steps.
 This avoids the need to store potentially large intermediate data for all possible permutations and thus allowing a combinatorial benchmark.
-In addition, the fact that benchmark functions are stored in the \texttt{PipelineDefinition} makes the benchmark more smoothly portable, making it easy to modify, extend with further methods, or apply to other datasets.
+This, along with the set-up simplicity (demonstrated here by our separate SVA-DEA benchmark), also distinguishes \textit{pipeComp} from general workflow frameworks such as \textit{drake} \citep{landauDrakePackage2018}; indeed we estimate that the benchmarks discussed here would have taken approximately 16TB of storage had all non-redundant intermediates been saved (see the Supplementary methods for more detailed explanations of the framework and comparison to \textit{CellBench} and \textit{drake}).} In addition, the fact that benchmark functions are stored in the \texttt{PipelineDefinition} makes the benchmark more smoothly portable, making it easy to modify, extend with further methods, or apply to other datasets.
 
@@
 \textit{pipeComp} is a flexible R framework for evaluating methods and parameter combinations in a complex pipeline, computing on-the-fly multilevel evaluation metrics.
-Applying this framework to scRNAseq clustering enabled us to make concrete recommendations on the steps of filtering, normalization, feature selection, denoising, dimensionality reduction and clustering (Figure \ref{fig:figure11}).
+Applying this framework to scRNAseq clustering enabled us to make {\color{red}clear} recommendations on the steps of filtering, normalization, feature selection, denoising, dimensionality reduction and clustering (Supplementary Figure 34).
 We demonstrate how a diversity of multilevel metrics can be more robust, more sensitive, and more nuanced than simply evaluating the final clustering performance.
-We hope that the \textit{pipeComp} framework can be applied to extend the current field of benchmarking, as well as to investigate pipelines other domains with multi-step computational analyses.
+We {\color{red}further showed} that the \textit{pipeComp} framework can be applied to extend the current field of benchmarking, as well as to investigate pipelines other domains with multi-step computational analyses.
 
@@
 Analyses were performed in R 3.6.0 (Bioconductor 3.9) and the following packages were installed from GitHub repositories: \textit{Seurat} (version 3.0.0), \textit{sctransform} (0.2.0), \textit{DoubletFinder} (2.0.1), \textit{scDblFinder} (1.1.4), \textit{scds} (1.0.0), \textit{SAVERX} (1.0.0), \textit{scImpute} (0.0.9), \textit{ALRA} (commit 7636de8), \textit{DCA} (0.2.2), \textit{DrImpute} (1.2), \textit{ENHANCE} (R version, commit 1571696), \textit{SAVERX} (1.0.0).
-The code for the glmPCA was obtained from \url{https://github.com/willtownes/scrna2019} (commit 1ddcc30ebb95d083a685f12fe81d35dd1b0cb1b2).
+{\color{red}The glmPCA was performed using the \textit{glmpca} package, while the code for deviance calculation was obtained from \url{https://github.com/willtownes/scrna2019} (commit 1ddcc30ebb95d083a685f12fe81d35dd1b0cb1b2).}
 
@@
 The \textit{stringent} set of filters uses the same thresholds, but excludes a cell if it is an outlier on any single distribution. 
-
+{\color{red}The \textit{veryStringent} set of filters excludes any cell that is an outlier by at least 2 MADs (lower or higher) on any of: log10\_total\_counts, log10\_total\_features, pct\_counts\_Mt, or pct\_counts\_in\_top\_20\_features.}
 The \textit{lenient} set of filters excludes cells that are outliers on at least two distributions by at least 5 MADs, except for pct\_counts\_Mt where the threshold is $>$ 3 MADs and $>$ 0.08.
@@
 The resulting evaluation is available in the Supplementary Figure 13.
+
+{\color{red}\subsection*{Overall run for interactions across steps}
+
+The set of alternatives found for the overall run (Supplementary Figures 28-30) can be found in the \href{https://github.com/markrobinsonuzh/scRNA_pipelines_paper/blob/master/data/overall_run.Rmd}{paper's repository}.
+The analysis of variance was performed with the following model:
+
+\texttt{metric$\sim$ dataset*resolution+doubletmethod+filt+norm+clustmethod+dims}
+
+\subsection*{SVA-DEA benchmark}
+
+The full process of creating the SVA-DEA \texttt{PipelineDefinition} is described step by step in the \href{https://github.com/plger/pipeComp/blob/master/vignettes/pipeComp_dea.Rmd}{pipeComp\_dea vignette}.
+
+The three benchmark datasets are available in \url{https://github.com/markrobinsonuzh/scRNA\_pipelines\_paper/blob/master/svadea/datasets/}, along with all code used to generate and prepare them.
+Briefly:
+\begin{itemize}
+    \item For the \textit{ipsc} dataset, we took the quantification of the GSE79636 dataset (fairly heterogeneous) from the \textit{iPSCpower} package \citep{germainTamingHumanGenetic2017}, and selected two random sets of 10 samples.
+No further variation was added, and log-foldchanges ranging from 0.25 to 1.25 were added to the counts of 300 random genes.
+All other genes were assumed to be negatives.
+    \item For the \textit{seqc} dataset, we used data from the \textit{seqc} package \citep{ComprehensiveAssessment2014}.
+For both groups (mixtures C and D, which also contain two different spike-in mixes), we selected samples from the \texttt{ILM\_refseq\_gene\_AGR} and \texttt{ILM\_refseq\_gene\_CNL} batches, which have clear technical differences; we selected 5 samples per group with a weak correlation with the batch (3:2 vs 2:3).
+Since the true differences between mixtures are not entirely known, the analysis was performed on all genes but only the spike-ins were considered for benchmarking.
+    \item For the \textit{simulation}, we simulated counts using the \textit{polyester} package \citep{frazeePolyesterSimulating2015} and the means/dispersions from the GSE79636 dataset (restricted to a single batch and biological group), with 500 DEGs (log-foldchanges ranging from 0.5 to 2) and two batch effects: 1) a technical batch partially correlated with the group (6:2) and affecting 1/3 of the genes, and 2) a linear vector of variation uncorrelated with the groups and affecting 1/3 of the genes.
+\end{itemize}
+
+Filtering was performed with edgeR's \texttt{filterByExpr} function, changing the \texttt{min.count} argument.
+For \textit{vstsva}, we first applied the variance-stabilizing transformation from \textit{DESeq2}\citep{loveModeratedEstimation2014} before applying \texttt{sva::sva}.
+\textit{edgeR.QLF} refers to the quasi-likelihood version \citep{lundDetectingDifferential2012} of \textit{edgeR}'s fit.
+The exact wrappers used can be found in the package at \url{https://github.com/plger/pipeComp/blob/master/inst/extdata/dea\_wrappers.R}.
+}
 
@@
 \begin{table}[h!]
-\caption{Overview of the benchmark datasets.}
+\caption{\textbf{Overview of the benchmark datasets.} {\color{red}`\# features' indicates the number of features detected in at least 10 cells.}}
 \label{tab:table1}
@@
   \hline
-dataset & source & protocol & description \\ 
+dataset & {\color{red}\# features} & \# cells & protocol & description \\ 
   \hline
-Koh & GSE85066 & SMARTer & FACS purified H7 hESC in different differention stages \\ 
-  Kumar & GSE60749 & SMARTer & Mouse ESC cultured in different conditions \\ 
-  Zhengmix4eq & \cite{duoClustering2018} & 10x & Mixtures of FACS purified PBMCs \\ 
-  Zhengmix4uneq & \cite{duoClustering2018} & 10x & Mixtures of FACS purified PBMCs \\ 
-  Zhengmix8eq & \cite{duoClustering2018} & 10x & Mixtures of FACS purified PBMCs \\ 
-  mixology10x3cl & \cite{tianMixology2018} & 10x & Mixture of 3  cancer cell lines from CellBench \\ 
-  mixology10x5cl & \cite{tianMixology2018} & 10x & Mixture of 5 cancer cell lines from CellBench \\ 
-  simMix1 & - & 10x-based & Simulation of 10 human cell subpopulations \\
-  simMix2 & - & 10x-based & Simulation of 9 mouse cell subpopulations \\
+Koh & 33922 & 531 & SMARTer & 9 FACS purified differentiation stages \\ 
+  Kumar & 41930 & 246 & SMARTer & Mouse ESC cultured in 3 different conditions \\ 
+  Zhengmix4eq & 10434 & 3994 & 10x & Mixtures of FACS purified PBMCs \\ 
+  Zhengmix4uneq & 11369 & 6498 & 10x & Mixtures of FACS purified PBMCs \\ 
+  Zhengmix8eq & 10600 & 3994 & 10x & Mixtures of FACS purified PBMCs \\ 
+  mixology10x3cl & 16208 & 902 & 10x & Mixture of 3  cancer cell lines from CellBench \\ 
+  mixology10x5cl & 11786 & 3918 & 10x & Mixture of 5 cancer cell lines from CellBench \\ 
+  simMix1 & 3696 & 2500 & 10x-based & Simulation of 10 human cell subpopulations \\
+  simMix2 & 8893 & 3000 & 10x-based & Simulation of 9 mouse cell subpopulations \\
    \hline
@@
 
-\newpage
+\section*{Figures}
 
-\showfigstrue
-
-\begin{figure}[h!]
+\begin{figure}
     \centering
-    \ifshowfigs
-    \includegraphics[width=0.95\textwidth,keepaspectratio]{dataset_description}
-    \fi
-    \caption{\textbf{Overview of the benchmark datasets used.}}
-    \label{fig:figure1}
+    \includegraphics[width=\textwidth, keepaspectratio]{main_figures/pipeline_explanation.pdf}
+    \caption{\textbf{Overview of the \textit{pipeComp} framework and its application to a scRNAseq clustering pipeline.
+A:} The package is built around a \texttt{PipelineDefinition} class which defines a set of functions to be executed consecutively, as well as optional evaluation functions for each step.
+Each step can accept a number of parameters whose alternative values are provided as a list.
+All (or subsets of) combinations of parameters can then be simultaneously ran and evaluated using the \texttt{runPipeline} function (see Supplementary Methods for more details).
+\textbf{B:} Scheme representing the application of \textit{pipeComp} to evaluate a range of methods that are commonly used in scRNAseq studies and some of the metrics monitored at various steps.}
+    \label{fig:explanation}
 \end{figure}
 
-\begin{figure}[h!]
+\begin{figure}
     \centering
-    \ifshowfigs
-    \includegraphics[scale=0.75, keepaspectratio]{main_figures/pipeline_explanation.pdf}
-    \fi
-    \caption{\textbf{Overview of the \textit{pipeComp} framework and its application to a scRNAseq clustering pipeline.
-A:} The package is built around a \texttt{PipelineDefinition} S4 class which defines a set of functions to be executed consecutively, as well as optional evaluation functions for each step.
-Each step can accept a number of parameters whose alternative values are provided as a list.
-All (or subsets of) combinations of parameters can then be simultaneously ran and evaluated using the \texttt{runPipeline} function.
-\textbf{B:} Scheme representing the application of \textit{pipeComp} to evaluate a range of methods that are commonly used in scRNAseq studies and some of the metrics monitored at various steps.}
-    \label{fig:figure2}
-\end{figure}
-
-\begin{figure}[h!]
-    \centering
-    \ifshowfigs
-    \includegraphics[scale=0.85, keepaspectratio]{{main_figures/doublets.pdf}}
-    \fi
+    \includegraphics[width=\textwidth, keepaspectratio]{{main_figures/doublets.pdf}}
     \caption{\textbf{Identification of doublet cells.
@@
 Even in datasets which should not have neotypic doublets, the cells identified as such tend to be misclassified.
-\label{fig:figure3}}
+\label{fig:doublets}}
 \end{figure}
 
-\begin{figure}[h!]
+\begin{figure}
     \centering
-    \ifshowfigs
-    \includegraphics[scale=0.72, keepaspectratio]{{main_figures/filtering.pdf}}
-    \fi
+    \includegraphics[width=\textwidth, keepaspectratio]{{main_figures/filtering.pdf}}
     \caption{\textbf{Effect of filtering on cell subpopulation structure and clustering.
@@
 The PCA methods refer to multivariate outlier detected as implemented in \textit{scater} (see Methods for details).}
-    \label{fig:figure4}
+    \label{fig:filtering}
 \end{figure}
 
-\begin{figure}[h!]
+\begin{figure}
     \centering
-    \ifshowfigs
     \includegraphics[width=\textwidth, keepaspectratio]{{main_figures/normalization.pdf}}
-    \fi
     \caption{\textbf{Evaluation of normalization procedures.
@@
 \textbf{C:} Clustering accuracy (\textit{Seurat} clustering), measured by mutual information (MI), Adjusted Rand Index (ARI) and ARI at the true number of cluster.
-`feat\_regress', `mt\_regress' and `feat\_mt\_regress' respectively stand for the regressing out of the number of features, the proportion of mitochondrial reads, or both during scaling.}
-    \label{fig:figure5}
+{\color{red}Grey squares indicate that the true number of clusters could not be reached with the corresponding method, dataset and parameter set.} `feat\_regress', `mt\_regress' and `feat\_mt\_regress' respectively stand for the regressing out of the number of features, the proportion of mitochondrial reads, or both during scaling.
+{\color{red}Throughout this paper, silhouette plots (panel \textbf{B}) square-root transform values for color mapping.
+Other evaluation metric heatmaps (e.g.
+panel \textbf{C} here) map colors to the signed square-root of the number of (matrix-wise) Median Absolute Deviations from the (column-wise) median.
+Using this mapping, \textit{differences} in color have the same meaning across datasets, and the color-scale is not primarily capturing outliers or baseline differences between datasets.
+The numbers printed in the cells represent the raw (i.e.
+unscaled) metric values, and are printed in white or black depending on whether they are above or below the median.}}
+    \label{fig:norm}
 \end{figure}
 
-\begin{figure}[h!]
+\begin{figure}
     \centering
-    \ifshowfigs
-    \includegraphics[scale=0.65]{{main_figures/selection.pdf}}
-    \fi
+    \includegraphics[width=\textwidth]{{main_figures/selection.pdf}}
     \caption{\textbf{Evaluation of feature selection methods.
@@
 The  \textit{vst.varExp} and \textit{devianceExplained} methods correspond to the estimates used in \textbf{A} to evaluate the selection methods and were included here only for validation purpose.}
-    \label{fig:figure6}
+    \label{fig:selection}
 \end{figure}
 
-\begin{figure}[h!]
+\begin{figure}
     \centering
-    \ifshowfigs
-        \includegraphics[width=\textwidth, keepaspectratio]{{main_figures/dr.pdf}}
-    \fi
-    \caption{\textbf{Evaluation of common dimensionality reduction methods.
-A:} Average silhouette width per subpopulation resulting from combinations of normalization and dimension reductions.
-\textbf{B:} Clustering accuracy, mean proportion of the variance in the first 5 components explained by real subpopulations (center), and median ARI of the resulting clustering (right).}
-    \label{fig:figure7}
-\end{figure}
-
-\begin{figure}[h!]
-    \centering
-    \ifshowfigs
-    \includegraphics[scale=0.55, keepaspectratio]{{main_figures/dimensionality.pdf}}
-    \fi
+    \includegraphics[width=\textwidth, keepaspectratio]{{main_figures/dimensionality.pdf}}
     \caption{\textbf{Estimating dimensionality.
@@
 \textbf{C:} Average silhouette width per subpopulation across a selection of methods, combined with \textit{sctransform} or standard \textit{Seurat} normalization.
-\textbf{D:} Clustering accuracy across the normalization/dimensionality estimation methods.}
-    \label{fig:figure8}
-\end{figure}
-
-\begin{figure}[h!]
-    \centering
-    \ifshowfigs
-    \includegraphics[width=\textwidth, keepaspectratio]{{main_figures/clustering.pdf}}
-    \fi
-    \caption{\textbf{Evaluation of clustering methods.} Clustering accuracy (\textbf{A}) and run time (\textbf{B}) of \textit{scran}- and \textit{Seurat}-based clustering tools in combination with \textit{sctransform} and standard \textit{Seurat}'s normalization.}
-    \label{fig:figure9}
+\textbf{D:} Clustering accuracy across the normalization/dimensionality estimation methods.
+{\color{red}The color-mapping scheme is the same as in Figure \ref{fig:norm}C}}
+    \label{fig:dim}
 \end{figure}
@@
     \centering
-    \ifshowfigs
-    \includegraphics[scale=0.55, keepaspectratio]{{main_figures/imputation.pdf}}
-    \fi
-    \caption{\textbf{Evaluation of imputation/denoising methods} Average silhouette width per subpopulation (\textbf{A}) and clustering accuracy (\textbf{B}) with or without (indicated as \textit{none}) application of a denoising/imputation method.}
-    \label{fig:figure10}
+    \includegraphics[width=\textwidth, keepaspectratio]{{main_figures/clustering.pdf}}
+    \caption{\textbf{Evaluation of clustering methods.} Clustering accuracy (\textbf{A}) and run time (\textbf{B}) of \textit{scran}- and \textit{Seurat}-based clustering tools in combination with \textit{sctransform} and standard \textit{Seurat}'s normalization.
+{\color{red} The color-mapping scheme in panel \textbf{A} and the left of panel \textbf{B} is the same as in Figure \ref{fig:norm}C.}}
+    \label{fig:clust}
 \end{figure}
 
-\begin{figure}[h!]
+\begin{figure}
     \centering
-    \ifshowfigs
-    \includegraphics[scale=0.9,keepaspectratio]{main_figures/pipeComp_summary.pdf}
-    \fi
-    \caption{\textbf{Recommendations of tools from filtering to clustering.}}
-    \label{fig:figure11}
+    \includegraphics[width=\textwidth, keepaspectratio]{{main_figures/svadea.pdf}}
+    \caption{{\color{red}\textbf{Evaluation of the impact of surrogate variable analysis (SVA) on (bulk) RNAseq differential expression analysis using \textit{pipeComp}.} \textit{RUVr} and \textit{RUVs} refer to the two \textit{RUVSeq} functions by the same name; \textit{svaseq} refers to the function of the same name in \textit{sva}, while \textit{vstsva} refers to the original \textit{sva} method applied no variance-stabilized data.
+\textbf{A:} True positive rates (TPR) and False discovery rates (FDR) of various combinations of SVA-based methods (using a single surrogate variable) and DEA methods.
+\textbf{B:} TPR and FDR of the different SVA-based methods (averaged across the different DEA methods) using increasing numbers of variables (`be' and `leek` refer to the two methods implemented in the \textit{sva} package for estimating the number of surrogate variables).
+In \textbf{A-B}, the x-axis is square-root transformed to improve readability, and three dots for each combination of methods refer to the nominal FDR thresholds of 0.01, 0.05, and 0.1.
+\textbf{C:} Correlation of estimated logFC with expected ones, as well as TPR and FDR (at nominal FDR threshold of 0.05) of the different methods with increasing numbers of variables.
+As in \ref{fig:norm}C, the color mapping tracks the number of (matrix-wise) Median Absolute Deviations from the (column-wise) median, while the printed numbers represent the raw metric values.}}
+    \label{fig:svadea}
 \end{figure}

### 2020-05-25 main_figures/selection.Rmd
updated many figures to improve readability...
fixed latex errors

@@
 ```{r, include=FALSE}
-#devtools::load_all("../../pipeComp_biorxiv/pipeComp/")
+devtools::load_all("../../pipeComp_biorxiv/pipeComp/")
 ```
@@
 snorm <- cl[grep("seuratvst",row.names(cl), invert=TRUE),]
-h1 <- scrna_evalPlot_clustAtTrueK(snorm[grep("=1000;",row.names(snorm)),], "ARI", scale = FALSE, title="ARI at true # clusters", anno_legend = FALSE)
+h1 <- scrna_evalPlot_clustAtTrueK(snorm[grep("=1000;",row.names(snorm)),], "ARI", scale = FALSE, title="ARI at true # clusters", anno_legend = FALSE, show_heatmap_legend=TRUE, heatmap_legend_param=list(direction="horizontal", at=5:10/10, title_position="leftcenter"))
 h2 <- scrna_evalPlot_clustAtTrueK(vst[grep("=1000;",row.names(vst)),], "ARI", scale = FALSE, reorder_rows=row.names(h1@matrix), title="sctransform\nARI at true # clusters", name="ARI2", anno_legend = FALSE)
-p2 <- grid.grabExpr(draw(h1+h2))
+p2 <- grid.grabExpr(draw(h1+h2, heatmap_legend_side = "bottom", merge_legend=TRUE))
 ```

### 2020-06-18 main_figures/overall.Rmd
added results-overview figure

---
title: "Overall"
author: "Pierre-Luc Germain"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r}
suppressPackageStartupMessages({
  library(ComplexHeatmap)
  library(S4Vectors)
  library(matrixStats)
  library(pipeComp) # version >= 0.99.26
})
source("../misc_functions.R")
```

```{r}
##########
## Note that a very similar plot can be created much more simply using the
## `scrna_evalPlot_overall` function of pipeComp. Here we do it manually in
## order to customize the positions of legends and such.

ov <- readRDS("../data/overall.rds")

ll <- lapply(c("minSilWidth", "meanSilWidth"), FUN=function(x){
  pipeComp:::.prepRes(ov$evaluation$dimreduction$silhouette$top_10_dims, what=x,
                      agg.by=c("doubletmethod","filt","norm","sel"), 
                      pipDef=metadata(ov)$PipelineDefinition, 
                      returnParams=TRUE, shortNames=TRUE)
})

pp1 <- ll[[1]]$pp
pp1$method <- NULL
ll1 <- lapply(ll, FUN=function(x) x$res)

ll <- lapply(c("ARI", "MI"), FUN=function(x){
  pipeComp:::.prepRes(ov$evaluation$clustering, what=x,
                      agg.by=c("doubletmethod","filt","norm","sel","dims","clustmethod"), 
                      pipDef=metadata(ov)$PipelineDefinition, 
                      returnParams=TRUE, shortNames=TRUE)
})
ll[[3]] <- pipeComp:::.prepRes(ov$evaluation$clustering, what="ARI",
                      agg.by=c("doubletmethod","filt","norm","sel","dims","clustmethod"), 
                      pipDef=metadata(ov)$PipelineDefinition, 
                      returnParams=TRUE, shortNames=TRUE, filt=rlang::expr(true.nbClusts==n_clus))
pp <- ll[[1]]$pp
pp$method <- NULL
ll2 <- lapply(ll, FUN=function(x) colCenterScale(x$res))

pclost <- scrna_evalPlot_filtering(ov, returnTable = TRUE)
pclost <- reshape2::dcast(pclost, doubletmethod+filt~dataset, value.var="max.lost")
row.names(pclost) <- apply(pclost[,1:2],1,collapse=" > ",FUN=paste)
tmp <- as.character(apply(pp[,colnames(pclost)[1:2]],1,collapse=" > ",FUN=paste))
pclost <- pclost[tmp,-1:-2]
row.names(pclost) <- row.names(ll2[[1]])
pclost <- apply(pclost,1,FUN=max)

tmp <- as.character(apply(pp[,colnames(pp1)],1,collapse=" > ",FUN=paste))
ll1 <- lapply(ll1, FUN=function(x){ x <- x[tmp,]; row.names(x) <- row.names(ll2[[1]]); x})
ll <- c(ll1,ll2)
```

```{r}
ll2 <- list( list(mat=ll[[1]], title="min silhouette\nwidth", cluster_columns=TRUE, name="silhouette width"),
             list(mat=ll[[2]], title="mean silhouette\nwidth", cluster_columns=TRUE, show_heatmap_legend=FALSE),
             list(mat=ll[[3]], title="mean ARI", name="ARI (MADs)", cluster_columns=FALSE),
             list(mat=ll[[4]], title="mean MI", name="MI (MADs)", cluster_columns=FALSE),
             list(mat=ll[[5]], title="mean ARI at\ntrue k", name="ARI at true k (MADs)", cluster_columns=FALSE) )
silhscale <- pipeComp:::.silScale(cbind(ll2[[1]]$mat, ll2[[2]]$mat))

ro <- rowMeans(do.call(cbind, lapply(ll[c(2:4)], FUN=function(x){
  x/sd(x,na.rm=TRUE)
  rowMeans(x,na.rm=TRUE)+rowMins(x,na.rm=TRUE)
})))
names(ro) <- row.names(ll[[1]])
h <- hclust(dist(do.call(cbind,ll)))
h2 <- reorder(as.dendrogram(h), ro)
ro2 <- names(sort(ro,decreasing=TRUE))

pp$doubletmethod <- gsub("doublet\\.","",pp$doubletmethod)
pp$filt <- factor(gsub("filt\\.","",pp$filt), c("none","default","stringent"))
pp$sel <- gsub("sel\\.","",pp$sel)
pp$norm <- gsub("norm\\.","",pp$norm)
pp$clustmethod <- gsub("clust\\.","",pp$clustmethod)
colnames(pp) <- c("doubet","filtering","normalization","feature sel","# dimensions","clustering")

cols <- getQualitativePalette(15)
vals <- as.character(unlist(lapply(lapply(pp, FUN=unique), as.character)))
cols <- c("none"="white", "scDblFinder"="black", "stringent"="black", "default"="darkgrey", sctransform="#114477", scran="#117755", seurat="#AA4444", deviance="#777711", vst="#771111", expr="#DD77AA", "maxLikGlobal"="#77AADD", "10"="#AAAA44", elbow="#DD7777", "scran.knnAnnoy"="#44AA88")
cols <- lapply(pp, FUN=function(x) cols[as.character(sort(unique(x)))])
ha <- HeatmapAnnotation(which = "row", "max\n% lost"=anno_barplot(pclost, bar_width = 1, border=FALSE, gp = gpar(fill = "#282828", col="#282828"), width=unit(1.5,"cm")), df=pp, col=cols, show_legend=FALSE)
legends <- lapply(names(cols), FUN=function(f){
  Legend( labels=names(cols[[f]]), title=f, legend_gp=gpar(fill=cols[[f]]) )
})

do.cluster <- TRUE
H <- NULL
for(i in seq_along(ll2)){
  if((do.cluster || is.null(ro)) && i==1){
    hi <- h2
  }else{
    hi <- FALSE
  }
  if(grepl("silhouette", ll2[[i]]$title)){
    col <- silhscale
  }else{
    col <- pipeComp:::.defaultColorMapping(ll2[[i]]$mat, center=TRUE)
  }
  la <- ra <- NULL
  if(i==length(ll2)) ra <- ha
  name <- ifelse(is.null(ll2[[i]]$name),ll2[[i]]$title,ll2[[i]]$name)
  H <- H + Heatmap(ll2[[i]]$mat, name=name, cluster_rows=hi, row_order=ro2, show_row_names=FALSE, show_column_names=FALSE, row_title=ifelse(i==1,"Unsupervised clustering",""), heatmap_legend_param = list(direction="horizontal", byrow=TRUE, nrow=1), column_title=ll2[[i]]$title, width=unit(ifelse(i<=2,3.2,2.7), "cm"), cluster_columns=ll2[[i]]$cluster_columns, show_column_dend=FALSE, bottom_annotation = pipeComp:::.ds_anno(colnames(ll2[[i]]$mat), legend=FALSE), right_annotation = ra, col=col,  column_names_gp = gpar(fontsize=10), column_title_gp = gpar(fontsize=13), show_heatmap_legend=ifelse(is.null(ll2[[i]]$show_heatmap_legend),TRUE,ll2[[i]]$show_heatmap_legend) )
}

top <- ro2[1:15]
ll3 <- lapply(ll2, FUN=function(x){
  x$mat <- x$mat[top,]
  x
})

ha2 <- HeatmapAnnotation(which = "row", "max\n% lost"=anno_barplot(pclost[top], bar_width = 1, border=FALSE, gp = gpar(fill = "#282828", col="#282828"), width=unit(1.5,"cm")), df=pp[top,], col=cols, show_legend=FALSE)

H2 <- NULL
for(i in seq_along(ll3)){
  if(grepl("silhouette", ll2[[i]]$title)){
    col <- silhscale
  }else{
    col <- pipeComp:::.defaultColorMapping(ll2[[i]]$mat, center=TRUE)
  }
  la <- ra <- NULL
  if(i==length(ll2)) ra <- ha2
  name <- ifelse(is.null(ll2[[i]]$name),ll2[[i]]$title,ll2[[i]]$name)
  H2 <- H2 + Heatmap(ll3[[i]]$mat, name=name, cluster_rows=FALSE, show_row_names=FALSE, row_title="Top 15 analyses", show_column_names=FALSE, heatmap_legend_param = list(direction="horizontal"), column_title=ll2[[i]]$title, width=unit(ifelse(i<=2,3.2,2.7), "cm"), cluster_columns=ll2[[i]]$cluster_columns, show_column_dend=FALSE, bottom_annotation = pipeComp:::.ds_anno(colnames(ll2[[i]]$mat), legend=TRUE), right_annotation = ra, col=col,  column_names_gp = gpar(fontsize=10), column_title_gp = gpar(fontsize=13), show_heatmap_legend=ifelse(is.null(ll2[[i]]$show_heatmap_legend),TRUE,ll2[[i]]$show_heatmap_legend) )
}
```

```{r, fig.width=10, fig.height=7}
pdf("merged.clustered.2.pdf", width=10, height=7)
draw(H, annotation_legend_side="right", heatmap_legend_side="bottom", annotation_legend_list=legends)
dev.off()
draw(H, annotation_legend_side="right", heatmap_legend_side="bottom", annotation_legend_list=legends)
```

```{r, fig.width=10, fig.height=3.5}
pdf("merged.clustered.2.pdf", width=10, height=4)
draw(H2, annotation_legend_side="right", heatmap_legend_side="bottom")
dev.off()
draw(H2, annotation_legend_side="right", heatmap_legend_side="bottom")
```

### 2020-06-18 bmc_article.tex
Updates from Overleaf
the actual good version

@@
 %\RequirePackage{natbib}
-\RequirePackage[authoryear]{natbib}% uncomment this for author-year bibliography
+%\RequirePackage[authoryear]{natbib}% uncomment this for author-year bibliography
+\usepackage[square,sort,comma,numbers]{natbib}
 \usepackage{hyperref}
@@
 We next investigated the impact of different normalization strategies.
-Beside the standard log-normalization included in \textit{Seurat}, we tested \textit{scran}'s pooling-based normalization \cite{lunPooling2016}, \textit{SCTransform}'s variance-stabilizing transformation \cite{hafemeisterSCtransform2019}, normalization based on stable genes \cite{linStableGenes2018, deekeStablyExpressed2018}, and \textit{SCnorm} \cite{bacherSCnorm2017} (\textit{scVI} \cite{lopezDeep2018} was included separately on a subset of parameter combinations due to high running times).
+Beside the standard log-normalization included in \textit{Seurat}, we tested \textit{scran}'s pooling-based normalization \cite{lunPooling2016}, \textit{sctransform}'s variance-stabilizing transformation \cite{hafemeisterSCtransform2019}, normalization based on stable genes \cite{linStableGenes2018, deekeStablyExpressed2018}, and \textit{SCnorm} \cite{bacherSCnorm2017} (\textit{scVI} \cite{lopezDeep2018} was included separately on a subset of parameter combinations due to high running times).
 In addition to log-normalization, the standard \textit{Seurat} clustering pipeline performs per-feature unit-variance scaling so that the PCA is not too strongly dominated by highly-expressed features.
-We therefore included versions of the different normalization procedures, with or without a subsequent scaling (\textit{SCTransform}'s variance-stabilizing transformation involves an approach analogous to scaling).
-\textit{Seurat}'s scaling function (and \textit{SCTransform}) also includes the option to regress out the effect of certain covariates.
+We therefore included versions of the different normalization procedures, with or without a subsequent scaling (\textit{sctransform}'s variance-stabilizing transformation involves an approach analogous to scaling).
+\textit{Seurat}'s scaling function (and \textit{sctransform}) also includes the option to regress out the effect of certain covariates.
 We tested its performance by using the proportion of mitochondrial counts and the number of detected features as covariates.
@@
 
-An important motivation for the development of \textit{SCTransform} was the observation that, even after normalization, the first principal components of various datasets tended to correlate with library size, suggesting an inadequate normalization \cite{hafemeisterSCtransform2019}.
+An important motivation for the development of \textit{sctransform} was the observation that, even after normalization, the first principal components of various datasets tended to correlate with library size, suggesting an inadequate normalization \cite{hafemeisterSCtransform2019}.
 However, as library size tends to vary across subpopulations, part of this effect can simply reflect biological differences.
 We therefore assessed to what extent the first principal component still retained a correlation with the library size and the number of detected features, removing the confounding covariation with the subpopulations (Figure \ref{fig:norm}A and Supplementary Figure 14).
-Scaling tended to remove much of the correlation with these features, and while most methods were able to remove most of the effect, the correlation was lowest with \textit{SCTransform}.
+Scaling tended to remove much of the correlation with these features, and while most methods were able to remove most of the effect, the correlation was lowest with \textit{sctransform}.
 Surprisingly, however, regressing out the covariates tended to increase the association with it.
@@
 `seurat (no scale)' vs `seurat') tended to reduce the average silhouette width of some subpopulations and to increase that of some less distinguishable ones and was generally, but not always, beneficial on the accuracy of the final clustering.
-Regressing out covariates systematically gave equal  or poorer performance on MI, ARI and ARI at true number of clusters, with a notable negative impact among the Smart-seq datasets (Koh and Kumar).
-When combined with \textit{SCTransform}, regressing out covariates using tended to have neither a positive or negative effect in droplet-based datasets, possibly owing to its regularization.
-In general, \textit{SCTransform} systematically outperformed other methods and, even though it was developed to be applied to data with unique molecular identifiers (UMI), it also performed fairly well with the Smart-seq protocol (Koh and Kumar datasets).}
+Regressing out covariates systematically gave equal or poorer performance on MI, ARI and ARI at true number of clusters, with a strongest negative impact among the Smart-seq datasets (Koh and Kumar).
+Regressing out covariates using \textit{sctransform} instead, tended to have a much milder effect, possibly owing to its regularization.
+In general, \textit{sctransform} systematically outperformed other methods and, even though it was developed to be applied to data with unique molecular identifiers (UMI), it also performed fairly well with the Smart-seq protocol (Koh and Kumar datasets).}
 
 Finally, we monitored whether, under the same downstream clustering analysis, different normalization methods tended to lead to an over- or under-estimation of the number of clusters.
-Although some methods had a tendency to lead to a higher (e.g., \textit{SCTransform}) or lower (e.g., stable genes) number of clusters, the effect was very mild and not entirely systematic (Supplementary Figure 16).
-We also confirmed that, especially in the UMI-based datasets, \textit{SCTransform} did in fact successfully stabilize variance across mean counts (Supplementary Figure 17), at little apparent cost on computation time (Supplementary Figure 18).
+Although some methods had a tendency to lead to a higher (e.g., \textit{sctransform}) or lower (e.g., stable genes) number of clusters, the effect was very mild and not entirely systematic (Supplementary Figure 16).
+We also confirmed that, especially in the UMI-based datasets, \textit{sctransform} did in fact successfully stabilize variance across mean counts (Supplementary Figure 17), at little apparent cost on computation time (Supplementary Figure 18).
 
@@
 Another common practice in feature selection is to use those with the highest mean expression.
-Recently, it was instead suggested to use deviance \cite{townesGlmpca2019}, while \textit{SCTransform} provides its own ordering of genes based on transformed variance.
+Recently, it was instead suggested to use deviance \cite{townesGlmpca2019}, while \textit{sctransform} provides its own ordering of genes based on transformed variance.
 
 Reasoning that a selection method should ideally select genes whose variability is higher between subpopulations than within, we first assessed to what extent each method selected genes with a high proportion of variance or deviance explained by the (real) subpopulations.
-As the proportion of variability in a gene attributable to subpopulations can be measured in various ways, we first compared three approaches: ANOVA on log-normalized count, ANOVA on \textit{SCTransform} normalization, and deviance explained.
-The ANOVAs performed on a standard \textit{Seurat} normalization and on \textit{SCTransform} data were highly correlated ({\color{red}Pearson correlation >0.9,} Supplementary Figure 19A).
+As the proportion of variability in a gene attributable to subpopulations can be measured in various ways, we first compared three approaches: ANOVA on log-normalized count, ANOVA on \textit{sctransform} normalization, and deviance explained.
+The ANOVAs performed on a standard \textit{Seurat} normalization and on \textit{sctransform} data were highly correlated ({\color{red}Pearson correlation >0.9,} Supplementary Figure 19A).
 These estimates were also in good agreement with the deviance explained, although lowly-expressed genes could have a high deviance explained without having much of their variance explained by subpopulation (Supplementary Figure 19B-D).
@@
 Since the various PCA approaches and implementations were recently benchmarked in a similar context \cite{TsuyuzakiPCA2020}, we focused on widely used approaches that had not yet been compared: \textit{Seurat}'s PCA, \textit{scran}'s \texttt{denoisePCA}, and GLM-PCA \cite{townesGlmpca2019}.
-When relevant, we combined them with \textit{SCTransform} normalization.
+When relevant, we combined them with \textit{sctransform} normalization.
 Given that \textit{Seurat}'s default PCA weights the cell embeddings by the variance of each component, we also evaluated the impact of this weighting with each method.
@@
 These truth-based estimates were then compared to those of the above dimensionality estimation methods (Figure \ref{fig:dim}B).} Of note, the methods differ widely in compute time (Figure \ref{fig:dim}B) and we saw no relationship between the accuracy of the estimates and the complexity of the method.
-Reasoning that over-estimating the number of dimensions is less problematic than under-estimating it, we kept the former methods for a full analysis of their impact on clustering (Figure \ref{fig:dim}C-D), when combined with \textit{SCTransform} or standard \textit{Seurat} normalization.
+Reasoning that over-estimating the number of dimensions is less problematic than under-estimating it, we kept the former methods for a full analysis of their impact on clustering (Figure \ref{fig:dim}C-D), when combined with \textit{sctransform} or standard \textit{Seurat} normalization.
 Although most methods performed well on the various clustering measures, the global maximum likelihood based on translated Poisson Mixture Model {\color{red}(\textit{maxLikGlobal}, using 10 or 20 nearest neighbors, although estimates were relatively stable across various values of \textit{k}; see Supplementary Figure 26) provided the dimensionality estimate that best separated challenging subpopulations (Figure \ref{fig:dim}C) and tended to result in the best clustering accuracy (Figure \ref{fig:dim}D).
@@
 Specifically, we compared rank-based graphs (\textit{scran}'s default) with graphs using the components' values directly, SNN vs KNN graphs, and the exact nearest neighbors to the Annoy approximation.
-Again, the tested methods were combined with \textit{Seurat}'s standard normalization and \textit{SCTransform}, otherwise using the parameters found optimal in the previous steps.
+Again, the tested methods were combined with \textit{Seurat}'s standard normalization and \textit{sctransform}, otherwise using the parameters found optimal in the previous steps.
 
@@
 Although analysis of variance identified many weak but significant interactions, we found that the benefits brought by performant methods were robust to differences in other parts of the pipeline (Supplementary Figures 29-30).
-For instance, we confirmed that doublet removal and \textit{SCtransform} normalization were beneficial independently of other steps, and that, again independendently of other steps, more stringent filtering did not yield any major improvement over lenient filtering (Supplementary Figure 29).
+For instance, we confirmed that doublet removal and \textit{sctransform} normalization were beneficial independently of other steps, and that, again independendently of other steps, more stringent filtering did not yield any major improvement over lenient filtering (Supplementary Figure 29).
 Similarly, deviance-based feature selection outcompeted \textit{Seurat}'s default selection method independently of changes at other steps (Supplementary Figure 30).}
@@
 While the FDR did increase in the \textit{ipsc} dataset, we suspect that much of the effect could be due to spurious differences \citep{germainTamingHumanGenetic2017} between the groups which we wrongly assume to be false positives.
-Nevertheless, the FDR remained acceptable with 1-3 variables.
+Nevertheless, the FDR remained acceptable with up to three variables.
 These results therefore suggest that these methods are robust and can be widely applied.}
@@
      \item Most normalization methods tested yielded a fair performance, especially when combined with scaling, which tended to have a positive impact on clustering.
-     \item \textit{SCTransform} offered the best overall performance in terms of the separability of the subpopulations, as well as removing the effect of library size and detection rate.
+     \item \textit{sctransform} offered the best overall performance in terms of the separability of the subpopulations, as well as removing the effect of library size and detection rate.
      \item The common practice of regressing out cell covariates, such as the detection rate or proportion of mitochondrial reads nearly always had a negative impact, leading to increased correlation with covariates and decreased clustering accuracy.
@@
 Some steps, such as the implementation of the PCA, were also not explored in detail here as they have already been the object of recent and thorough study elsewhere \cite{SunDimRed2019, TsuyuzakiPCA2020}.
-We also considered only methods relying on Euclidean distance, while correlation was recently reported to be superior\cite{kim_impact_2019} and would require further investigation.
+We also considered only methods relying on Euclidean distance, while correlation was recently reported to be superior \cite{kim_impact_2019} and would require further investigation.
 
-{\color{red}Among all pipelines tested, performance varied greatly across datasets.
+{\color{red}Among all tested pipelines, performance varied greatly across datasets.
 The ZhengMix8eq and simMix1 datasets were the most challenging datasets across all preprocessing steps that we tested.
@@
 This avoids the need to store potentially large intermediate data for all possible permutations and thus allowing a combinatorial benchmark.
-This, along with the set-up simplicity (demonstrated here by our separate SVA-DEA benchmark), also distinguishes \textit{pipeComp} from general workflow frameworks such as \textit{drake}\citep{landauDrakePackage2018}; indeed we estimate that the benchmarks discussed here would have taken approximately 16TB of storage had all non-redundant intermediates been saved (see the Supplementary methods for more detailed explanations of the framework and comparison to \textit{CellBench} and \textit{drake}).} In addition, the fact that benchmark functions are stored in the \texttt{PipelineDefinition} makes the benchmark more smoothly portable, making it easy to modify, extend with further methods, or apply to other datasets.
+This, along with the set-up simplicity (demonstrated here by our separate SVA-DEA benchmark), also distinguishes \textit{pipeComp} from general workflow frameworks such as \textit{drake} \citep{landauDrakePackage2018}; indeed we estimate that the benchmarks discussed here would have taken approximately 16TB of storage had all non-redundant intermediates been saved (see the Supplementary methods for more detailed explanations of the framework and comparison to \textit{CellBench} and \textit{drake}).} In addition, the fact that benchmark functions are stored in the \texttt{PipelineDefinition} makes the benchmark more smoothly portable, making it easy to modify, extend with further methods, or apply to other datasets.
 
@@
 \subsection*{Software and package versions}
-Analyses were performed in R 3.6.0 (Bioconductor 3.9) and the following packages were installed from GitHub repositories: \textit{Seurat} (version 3.0.0), \textit{SCTransform} (0.2.0), \textit{DoubletFinder} (2.0.1), \textit{scDblFinder} (1.1.4), \textit{scds} (1.0.0), \textit{SAVERX} (1.0.0), \textit{scImpute} (0.0.9), \textit{ALRA} (commit 7636de8), \textit{DCA} (0.2.2), \textit{DrImpute} (1.2), \textit{ENHANCE} (R version, commit 1571696), \textit{SAVERX} (1.0.0).
+Analyses were performed in R 3.6.0 (Bioconductor 3.9) and the following packages were installed from GitHub repositories: \textit{Seurat} (version 3.0.0), \textit{sctransform} (0.2.0), \textit{DoubletFinder} (2.0.1), \textit{scDblFinder} (1.1.4), \textit{scds} (1.0.0), \textit{SAVERX} (1.0.0), \textit{scImpute} (0.0.9), \textit{ALRA} (commit 7636de8), \textit{DCA} (0.2.2), \textit{DrImpute} (1.2), \textit{ENHANCE} (R version, commit 1571696), \textit{SAVERX} (1.0.0).
 {\color{red}The glmPCA was performed using the \textit{glmpca} package, while the code for deviance calculation was obtained from \url{https://github.com/willtownes/scrna2019} (commit 1ddcc30ebb95d083a685f12fe81d35dd1b0cb1b2).}
@@
 
-Unless specified otherwise, the variance in gene expression explained by subpopulations was calculated on the data normalized and transformed through \textit{SCTransform}.
+Unless specified otherwise, the variance in gene expression explained by subpopulations was calculated on the data normalized and transformed through \textit{sctransform}.
 For each gene, we fitted a linear model using the subpopulation as only independent variable ($\sim subpopulation$) and used the R-squared as a measure of the variance explained.
@@
 The evaluation of \textit{scVI} algorithm was performed separate to the other pipeline combinations as the model training on our reference datasets was especially resource-consuming.
-As it was not practically feasible to perform the training step on many combination of parameters as it was done with other tools, we evaluated it on a representative subset of pipelines using \textit{SCTransform} normalization and \textit{Seurat}'s PCA dimension reduction method.
+As it was not practically feasible to perform the training step on many combination of parameters as it was done with other tools, we evaluated it on a representative subset of pipelines using \textit{sctransform} normalization and \textit{Seurat}'s PCA dimension reduction method.
 The resulting evaluation is available in the Supplementary Figure 13.
@@
 
-\showfigstrue
-
 \begin{figure}
     \centering
-    \ifshowfigs
-    \includegraphics[width=\textwidth, keepaspectratio]{main_figures/pipeline_explanation.pdf}
-    \fi
+    \includegraphics[width=\textwidth, keepaspectratio]{pipeline_explanation.pdf}
     \caption{\textbf{Overview of the \textit{pipeComp} framework and its application to a scRNAseq clustering pipeline.
@@
     \centering
-    \ifshowfigs
-    \includegraphics[width=\textwidth, keepaspectratio]{{main_figures/doublets.pdf}}
-    \fi
+    \includegraphics[width=\textwidth, keepaspectratio]{{doublets.pdf}}
     \caption{\textbf{Identification of doublet cells.
@@
     \centering
-    \ifshowfigs
-    \includegraphics[width=\textwidth, keepaspectratio]{{main_figures/filtering.pdf}}
-    \fi
+    \includegraphics[width=\textwidth, keepaspectratio]{{filtering.pdf}}
     \caption{\textbf{Effect of filtering on cell subpopulation structure and clustering.
@@
     \centering
-    \ifshowfigs
-    \includegraphics[width=\textwidth, keepaspectratio]{{main_figures/normalization.pdf}}
-    \fi
+    \includegraphics[width=\textwidth, keepaspectratio]{{normalization.pdf}}
     \caption{\textbf{Evaluation of normalization procedures.
@@
     \centering
-    \ifshowfigs
-    \includegraphics[width=\textwidth]{{main_figures/selection.pdf}}
-    \fi
+    \includegraphics[width=\textwidth]{{selection.pdf}}
     \caption{\textbf{Evaluation of feature selection methods.
@@
 \textbf{B:} Accuracy of clusterings (at the true number of clusters) when selecting 1000 genes using the given methods.
-Based on standard \textit{Seurat} normalization (left) or \textit{SCTransform} (right).
+Based on standard \textit{Seurat} normalization (left) or \textit{sctransform} (right).
 The  \textit{vst.varExp} and \textit{devianceExplained} methods correspond to the estimates used in \textbf{A} to evaluate the selection methods and were included here only for validation purpose.}
@@
     \centering
-    \ifshowfigs
-    \includegraphics[width=\textwidth, keepaspectratio]{{main_figures/dimensionality.pdf}}
-    \fi
+    \includegraphics[width=\textwidth, keepaspectratio]{{dimensionality.pdf}}
     \caption{\textbf{Estimating dimensionality.
@@
 \textbf{B:} Difference between `real' dimensionality (from \textbf{A}) and dimensionality estimation methods, along with computing time.
-\textbf{C:} Average silhouette width per subpopulation across a selection of methods, combined with \textit{SCTransform} or standard \textit{Seurat} normalization.
+\textbf{C:} Average silhouette width per subpopulation across a selection of methods, combined with \textit{sctransform} or standard \textit{Seurat} normalization.
 \textbf{D:} Clustering accuracy across the normalization/dimensionality estimation methods.
@@
     \centering
-    \ifshowfigs
-    \includegraphics[width=\textwidth, keepaspectratio]{{main_figures/clustering.pdf}}
-    \fi
-    \caption{\textbf{Evaluation of clustering methods.} Clustering accuracy (\textbf{A}) and run time (\textbf{B}) of \textit{scran}- and \textit{Seurat}-based clustering tools in combination with \textit{SCTransform} and standard \textit{Seurat}'s normalization.
+    \includegraphics[width=\textwidth, keepaspectratio]{{clustering.pdf}}
+    \caption{\textbf{Evaluation of clustering methods.} Clustering accuracy (\textbf{A}) and run time (\textbf{B}) of \textit{scran}- and \textit{Seurat}-based clustering tools in combination with \textit{sctransform} and standard \textit{Seurat}'s normalization.
 {\color{red} The color-mapping scheme in panel \textbf{A} and the left of panel \textbf{B} is the same as in Figure \ref{fig:norm}C.}}
@@
     \centering
-    \ifshowfigs
-    \includegraphics[width=\textwidth, keepaspectratio]{{main_figures/svadea.pdf}}
-    \fi
+    \includegraphics[width=\textwidth, keepaspectratio]{{svadea.pdf}}
     \caption{{\color{red}\textbf{Evaluation of the impact of surrogate variable analysis (SVA) on (bulk) RNAseq differential expression analysis using \textit{pipeComp}.} \textit{RUVr} and \textit{RUVs} refer to the two \textit{RUVSeq} functions by the same name; \textit{svaseq} refers to the function of the same name in \textit{sva}, while \textit{vstsva} refers to the original \textit{sva} method applied no variance-stabilized data.

### 2020-06-29 main_figures/dimensionality.Rmd
updated figure on dimensionality

--- 2020-05-08 main_figures/dimensionality.Rmd
@@
   library(matrixStats)
-  library(pipeComp) # the first part requires pipeComp >0.99.26
+  library(pipeComp) # requires pipeComp >0.99.26
 })
@@
                        row_split=gsub("norm\\.","",norm), 
-                       heatmap_legend_param=list(direction="horizontal")),
+                       filter=sel=="sel.vst" & !(dims %in% c("10","30","essLocal.b")),
+                       heatmap_legend_param=list(direction="horizontal"),
+                       row_names_max_width = unit(6.2, "cm")),
   rnfn)
@@
   evalHeatmap( res, what=c("NMI","min_pr"), agg.by=c("norm","dims"), 
-                   row_split=gsub("norm\\.","",norm), reorder_rows=TRUE, anno_legend=FALSE) + 
+               filter=sel=="sel.vst" & !(dims %in% c("10","30","essLocal.b")),
+                   row_split=gsub("norm\\.","",norm), reorder_rows=TRUE, anno_legend=FALSE,
+               row_names_max_width = unit(6.2, "cm")) + 
   evalHeatmap(res, what=c("ARI"), agg.by=c("norm","dims"), row_split=gsub("norm\\.","",norm), 
-              reorder_rows=TRUE, filterExpr=n_clus==true.nbClusts, 
-              anno_legend=FALSE, title="ARI at\ntrue k"),
+              filterExpr=n_clus==true.nbClusts & sel=="sel.vst" & !(dims %in%                                                                      c("10","30","essLocal.b")), row_names_max_width = unit(6.2, "cm"),
+              reorder_rows=TRUE, anno_legend=FALSE, title="ARI at\ntrue k"),
   rnfn
 )
-p3 <- grid.grabExpr(draw(p3))
-```
-
-```{r, eval=FALSE}
-# NB: the next results were generated using an older version of pipeComp (branch 
-# `biorxiv`), and must be plotted using this version of the package
-suppressPackageStartupMessages({
-  library(pipeComp)
-})
-```
-
-```{r, include=FALSE}
-devtools::load_all("../../pipeComp_biorxiv/pipeComp/")
-source("../../pipeComp_biorxiv/pipeComp/inst/extdata/scrna_alternatives.R")
+p3 <- grid.grabExpr(draw(p3, annotation_legend_list=.scaledLegend(), merge_legend=TRUE, heatmap_legend_side="bottom"))
 ```
@@
 o <- lapply(o, FUN=function(x){
-  x$nselected$elbow <- x$nselected$elbow-1
-  x
+  x <- x[setdiff(names(x), c("essLocal.b","pcaLocal.FO"))]
 })
@@
 o2$avgTime <- tmp[as.character(o2$method)]
-p1b <- ggplot(o2, aes(reorder(method, diff, function(x) mean(abs(x))), diff, group=method, fill=Time)) + geom_violin() + coord_flip() + geom_hline(yintercept=0, linetype="dashed") + ylab("Difference to expected dimensionality") + scale_fill_manual(values=tcols) + theme(axis.title.y=element_blank(), legend.position=c(0.7, 0.3))
+p1b <- ggplot(o2, aes(reorder(method, diff, function(x) mean(abs(x))), diff, 
+                      group=method, fill=Time, colour=Time)) + geom_violin() + 
+  coord_flip() + geom_hline(yintercept=0, linetype="dashed") + 
+  ylab("Difference to expected dimensionality") + scale_fill_manual(values=tcols) +
+  scale_color_manual(values=tcols) + 
+  theme(axis.title.y=element_blank(), legend.position=c(0.7, 0.3))
 p1 <- plot_grid(p1a, p1b, rel_widths = c(0.42,0.58), labels=LETTERS[1:2])
@@
 ```{r figure6, fig.width=9, fig.height=9.5}
-pdf("dimensionality.pdf", width=9, height=10)
-plot_grid( p1, p2, p3, nrow=3, labels=c(NA,LETTERS[3:4]))
+pdf("dimensionality.pdf", width=9, height=10.5)
+plot_grid( p1, p2, p3, nrow=3, labels=c(NA,LETTERS[3:4]), rel_heights = c(3,4,4))
 dev.off()
-plot_grid( p1, p2, p3, nrow=3, labels=c(NA,LETTERS[3:4]))
+plot_grid( p1, p2, p3, nrow=3, labels=c(NA,LETTERS[3:4]), rel_heights = c(3,4,4))
 ```

### 2020-06-29 supplementary_figures/supp_overall.Rmd
simplified labels, reduced text size

@@
 
-```{r, fig.width=7, fig.height=6}
+```{r, fig.width=6, fig.height=5}
 ov <- readRDS("../data/overall.rds")
@@
 ff <- function(x, trueK=FALSE){
-  f <- paste0(x,"~dataset+resolution*dataset+doubletmethod+filt+norm+clustmethod+dims")
+  f <- paste0(x,"~dataset+resolution*dataset+doubletmethod+sel+filt+norm+clustmethod+dims")
   if(trueK) cl <- cl[cl$n_clus==cl$true.nbClusts,]
@@
 co2 <- dplyr::bind_rows(lapply(cos2, FUN=function(x) x$coef), .id="metric")
-for(f in c("norm","filt","doubletmethod","dims","clustmethod")){
+for(f in c("norm","filt","doubletmethod","dims","sel","clustmethod")){
   co$coef <- gsub(paste0("^",f),"",co$coef)
@@
 
-`metric ~ dataset*resolution+doubletmethod+filt+norm+clustmethod+dims`
+`metric ~ dataset*resolution+doubletmethod+filt+sel+norm+clustmethod+dims`
 
@@
 rnfn <- function(x){
-  x <- gsub("^norm\\.","",x)
+  x <- gsub("norm\\.","",x)
   x <- gsub("^doublet\\.","",x)
@@
 
-H1 <- renameHrows( scrna_evalPlot_silh(ov, agg.by = c("doubletmethod","filt","norm"), row_split = norm,
+H1 <- renameHrows( scrna_evalPlot_silh(ov, agg.by = c("doubletmethod","filt","norm"), row_split = rnfn(norm),
                                        heatmap_legend_param=list(direction="horizontal"),font_factor = 0.8,
@@
                  agg.by=c("doubletmethod","filt","norm"), font_factor = 0.8, 
-                 row_split=norm,row_names_max_width=unit(7, "cm")) + 
+                 row_split=rnfn(norm),row_names_max_width=unit(7, "cm")) + 
     evalHeatmap(ov, step="clustering", what=c("ARI"), agg.by=c("doubletmethod","filt","norm"), 
-                row_split=norm,filter=n_clus==true.nbClusts, title="ARI at true k",
-                anno_legend=FALSE, font_factor = 0.8,row_names_max_width=unit(7, "cm")),
+                row_split=rnfn(norm),filter=n_clus==true.nbClusts, title="ARI at true k",
+                anno_legend=FALSE, font_factor = 0.8, row_names_max_width=unit(7, "cm")),
   rnfn)
 
-H3 <- renameHrows( scrna_evalPlot_silh(ov, agg.by = c("doubletmethod","sel","norm"), row_split = norm,
+H3 <- renameHrows( scrna_evalPlot_silh(ov, agg.by = c("doubletmethod","sel","norm"), row_split = rnfn(norm),
                                        heatmap_legend_param=list(direction="horizontal"),
+                                       font_factor = 0.8, 
                                        row_names_max_width=unit(7, "cm")), rnfn)
@@
      evalHeatmap(ov, step="clustering", what=c("MI","min_pr"), anno_legend=FALSE, title=c("MI","Min precision"),
-                 agg.by=c("doubletmethod","sel","norm"), row_split=norm,row_names_max_width=unit(7, "cm")) + 
+                 agg.by=c("doubletmethod","sel","norm"), row_split=rnfn(norm), 
+                 font_factor = 0.8, row_names_max_width=unit(7, "cm")) + 
     evalHeatmap(ov, step="clustering", what=c("ARI"), agg.by=c("doubletmethod","sel","norm"), 
-                row_split=norm,filter=n_clus==true.nbClusts, title="ARI at true k", 
-                anno_legend=FALSE, row_names_max_width=unit(7, "cm")),
+                row_split=rnfn(norm),filter=n_clus==true.nbClusts, title="ARI at true k", 
+                anno_legend=FALSE, font_factor = 0.8, row_names_max_width=unit(7, "cm")),
   rnfn)

### 2020-06-29 supplementary_figures/supp3_normalization.Rmd
updated scVI figures

@@
 res <- readRDS("../data/scVI.rds")
-sil <- res$evaluation$dimreduction$silhouette$top_10_dims
-sil$method <- factor(paste(sil$norm, sil$dr))
-levels(sil$method) <- meths <- c("scVI.LD", "scran+seurat.pca", "scVI.LD", "scVI+seurat.pca","scVI.LD","sctransform")
-levels(sil$dataset) <- c("Koh","Kumar","mixology10x3cl","mixology10x5cl","simMix1","simMix2","Zhengmix4eq","Zhengmix4uneq","Zhengmix8eq")
-res$evaluation$dimreduction$silhouette$top_10_dims <- sil
-x <- range(res$evaluation$dimreduction$silhouette$top_10_dims$meanSilWidth)
-silColBreaks <- unique(round(c(x[1],0,x[2]/2,x[2]),1))
-H <- scrna_evalPlot_silh(res, agg.by = "method", 
-                       reorder_rows=TRUE, value_format="", 
-                       show_heatmap_legend=TRUE, anno_legend = TRUE, 
-                       heatmap_legend_param=list(at=silColBreaks))
+res$evaluation$dimreduction$silhouette[[1]] <- res$evaluation$dimreduction$silhouette[[1]][with(res$evaluation$dimreduction$silhouette[[1]], filt=="filt.default" & (!(dr %in% c("scVI.latent","scVI.LD")) | norm=="norm.scVI")),]
+res$evaluation$dimreduction$silhouette[[1]]$norm <- gsub("norm.seuratvst","sctransform",res$evaluation$dimreduction$silhouette[[1]]$norm)
+res$evaluation$clustering <- res$evaluation$clustering[with(res$evaluation$clustering, filt=="filt.default" & (!(dr %in% c("scVI.latent","scVI.LD")) | norm=="norm.scVI")),]
+res$evaluation$clustering$norm <- gsub("norm.seuratvst","sctransform",res$evaluation$clustering$norm)
 
-res$evaluation$clustering$method <- factor(paste(res$evaluation$clustering$norm, res$evaluation$clustering$dr))
+H <- scrna_evalPlot_silh(res)
+
 res$evaluation$clustering$delta.nbClust <- res$evaluation$clustering$n_clus-res$evaluation$clustering$true.nbClusts
-levels(res$evaluation$clustering$method) <- meths
-meths <- levels(sil$method)[c(4,2,3,1)]
-H2 <- evalHeatmap(res, step="clustering", what=c("MI","ARI"), agg.by="method", anno_legend=FALSE, reorder_rows=meths ) + evalHeatmap(res, step="clustering", what="ARI", agg.by="method", anno_legend=FALSE, reorder_rows=meths, filterExpr = n_clus==true.nbClusts, name="ARI at true k", title="ARI at true k" )
-
+H2 <- evalHeatmap(res, step="clustering", what=c("MI"), agg.by=c("norm","dr"), anno_legend=FALSE, width=unit(6.5,"cm")) +
+  evalHeatmap(res, step="clustering", what="ARI", agg.by=c("norm","dr"), anno_legend=FALSE, filterExpr = n_clus==true.nbClusts, name="ARI at true k", title="ARI at true k" ) + 
+  evalHeatmap(res, step="clustering", what=c("delta.nbClust"), value_format = "%.0f", scale=FALSE, col=circlize::colorRamp2(c(-3,0,9),c("red","white","blue")), value_cols = c("black","black"),agg.by=c("norm","dr"), anno_legend=FALSE, show_heatmap_legend = TRUE, heatmap_legend_param=list(direction="horizontal"))
 ```
 
-```{r, fig.width=9, fig.height=4}
-plot_grid(grid.grabExpr(draw(H)), grid.grabExpr(draw(H2)), ncol=1, labels = LETTERS[1:2])
+```{r, fig.width=9, fig.height=5.5}
+plot_grid(grid.grabExpr(draw(H)), grid.grabExpr(draw(H2, heatmap_legend_side="bottom")), ncol=1, labels = LETTERS[1:2], rel_heights=c(3,4))
 ```
@@
 **scVI evaluation.
-A:** Average silhouette width per subpopulation using either sctransform, scran or scVI normalization followed by Seurat PCA, or the scVI linear decoder (LD).
+A:** Average silhouette width per subpopulation using either sctransform, scran or scVI normalization followed by Seurat PCA, or the scVI latent space (latent) or imputed values (LD) of the linear decoder.
 **B:** Clustering accuracy across the same methods followed by Seurat clustering.

### 2020-06-29 bmc_article.tex

@@
 
-The \textit{pipeComp} package defines a pipeline as, minimally, a list of functions executed consecutively on the output of the previous one {\color{red}(Figure \ref{fig:explanation}A; see also Supplementary methods).
+The \textit{pipeComp} package defines a pipeline as, minimally, a list of functions executed consecutively on the output of the previous one (Figure \ref{fig:explanation}A; see also Supplementary methods).
 In addition, optional benchmark functions can be set for each step to provide standardized, multi-layered evaluation metrics.
-Given such a \texttt{PipelineDefinition} object, a set of alternative parameters (which might include different subroutines calling different methods) and benchmark datasets, the \texttt{runPipeline} function then proceeds through all combinations of arguments (or a desired subset), avoiding recomputing the same step twice without the need to save all alternative intermediates, and compiling evaluations (including run time) on the fly (see Supplementary methods for more detail).} Variations in a given parameter can be evaluated using all metrics from this point downward in the pipeline.
+Given such a \texttt{PipelineDefinition} object, a set of alternative parameters (which might include different subroutines calling different methods) and benchmark datasets, the \texttt{runPipeline} function then proceeds through all combinations of arguments (or a desired subset), avoiding recomputing the same step twice without the need to save all alternative intermediates, and compiling evaluations (including run time) on the fly (see Supplementary methods for more detail).
+Variations in a given parameter can be evaluated using all metrics from this point downward in the pipeline.
 This is especially important because end-point metrics, such as the adjusted Rand index (ARI) \cite{HubertARI1985} for clustering, are not perfect.
@@
 
-{\color{red}We next investigated the impact of several filtering methods (see Methods for more details): four methods based on deviations to MADs with increasing levels of stringency (named \textit{lenient}, \textit{default}, \textit{stringent}, \textit{veryStringent}) and two methods based on \textit{scater}'s \textit{runPCA} using all (\textit{pca.all}) or selected covariates (\textit{pca.sel})}.
+We next investigated the impact of several filtering methods (see Methods for more details): four methods based on deviations to MADs with increasing levels of stringency (named \textit{lenient}, \textit{default}, \textit{stringent}, \textit{veryStringent}) and two methods based on \textit{scater}'s \textit{runPCA} using all (\textit{pca.all}) or selected covariates (\textit{pca.sel}).
 An important risk of excluding cells on the basis of their distance from the whole distribution of cells on some properties (e.g., library size) is that these properties tend to have different distributions across subpopulations.
@@
 Since filtering changes the relative abundance of the different subpopulations, global clustering accuracy metrics such as ARI are not appropriate here.
-We therefore calculated the per-subpopulation precision and recall using the Hungarian algorithm \cite{kuhnHungarianMethod1955} and monitored the mean F1 score {\color{red}across subpopulations.
+We therefore calculated the per-subpopulation precision and recall using the Hungarian algorithm \cite{kuhnHungarianMethod1955} and monitored the mean F1 score across subpopulations.
 A first observation was that, although more stringent filtering tended to be associated with an increase in accuracy, it tended to plateau and could also become deleterious.
@@
 Koh, Kumar, and the simulations) are those in which heterotypic doublets are not expected.
-Overall, we therefore recommend the use of doublet removal followed by our \textit{default} filtering (see Methods) or a similar approach.}
+Overall, we therefore recommend the use of doublet removal followed by our \textit{default} filtering (see Methods) or a similar approach.
 
@@
 Since clustering accuracy metrics such as the ARI are very strongly influenced by the number of clusters, we complemented it with silhouette width \cite{RousseeuwSil1987} and mutual information (MI, Figure \ref{fig:norm}B-C).
-We found most methods (including no normalization at all) to perform fairly well in most of the subpopulations{\color{red}, and that some datasets (e.g.
+We found most methods (including no normalization at all) to perform fairly well in most of the subpopulations, and that some datasets (e.g.
 Zhengmix8eq and simMix1) and subpopulations were consistently more challenging across all methods.
@@
 Regressing out covariates using \textit{sctransform} instead, tended to have a much milder effect, possibly owing to its regularization.
-In general, \textit{sctransform} systematically outperformed other methods and, even though it was developed to be applied to data with unique molecular identifiers (UMI), it also performed fairly well with the Smart-seq protocol (Koh and Kumar datasets).}
+In general, \textit{sctransform} systematically outperformed other methods and, even though it was developed to be applied to data with unique molecular identifiers (UMI), it also performed fairly well with the Smart-seq protocol (Koh and Kumar datasets).
 
@@
 As the proportion of variability in a gene attributable to subpopulations can be measured in various ways, we first compared three approaches: ANOVA on log-normalized count, ANOVA on \textit{sctransform} normalization, and deviance explained.
-The ANOVAs performed on a standard \textit{Seurat} normalization and on \textit{sctransform} data were highly correlated ({\color{red}Pearson correlation >0.9,} Supplementary Figure 19A).
+The ANOVAs performed on a standard \textit{Seurat} normalization and on \textit{sctransform} data were highly correlated (Pearson correlation >0.9, Supplementary Figure 19A).
 These estimates were also in good agreement with the deviance explained, although lowly-expressed genes could have a high deviance explained without having much of their variance explained by subpopulation (Supplementary Figure 19B-D).
@@
 In general, we found deviance and unstandardized estimates of variance to provide the best results across datasets and normalization methods.
-Increasing the number of features selected also {\color{red}tended to lead} to an increase in the accuracy of the clustering, typically plateauing after 4000 features (Supplementary Figure 22).
+Increasing the number of features selected also tended to lead to an increase in the accuracy of the clustering, typically plateauing after 4000 features (Supplementary Figure 22).
 
@@
 The impact of the choice of dimensionality reduction method was far greater than that of normalization or feature selection (Supplementary Figure 23).
-{\color{red}Overall, weighting the components by their explained variance tended to increase silhouette width (Supplementary Figure 23A) and, in most datasets, clustering accuracy (Supplementary} Figure 23B).
+Overall, weighting the components by their explained variance tended to increase silhouette width (Supplementary Figure 23A) and, in most datasets, clustering accuracy (Supplementary Figure 23B).
 GLM-PCA tended to increase the average silhouette width of already well-defined subpopulations, but \textit{Seurat}'s PCA procedure however proved superior on all metrics.
 Like GLM-PCA, \textit{scVI}'s Linear Decoder (LD) does not explicitly rely on normalized counts.
-{\color{red}Its performance was however the worst both in terms of silhouette width and clustering accuracy (Supplementary Figure 15), and it only performed well in some of the least challenging datasets (mixology10x5cl and simMix2).}
+Its performance was however the worst both in terms of silhouette width and clustering accuracy (Supplementary Figure 15), and it only performed well in some of the least challenging datasets (mixology10x5cl and simMix2).
 
@@
 
-Different approaches have been proposed to select the appropriate number of dimensions, from the visual identification of an {\color{red}inflexion point in the curve of the variance explained by each component (`elbow' method), to more complex algorithms.} We evaluated the performance of dimensionality estimators implemented in the \textit{intrinsicDimension} package \cite{johnssonDimensionality2015}, as well as common procedures such as the `elbow` method, some scRNAseq-specific methods such as the JackStraw procedure \cite{ChungJackstraw2015} or \textit{scran}'s \texttt{denoisePCA} \cite{LunScran2016}, and the recent application of Fisher Separability analysis \cite{AlberganteSepar2019}.
+Different approaches have been proposed to select the appropriate number of dimensions, from the visual identification of an inflexion point in the curve of the variance explained by each component (`elbow' method), to more complex algorithms.
+We evaluated the performance of dimensionality estimators implemented in the \textit{intrinsicDimension} package \cite{johnssonDimensionality2015}, as well as common procedures such as the `elbow` method, some scRNAseq-specific methods such as the JackStraw procedure \cite{ChungJackstraw2015} or \textit{scran}'s \texttt{denoisePCA} \cite{LunScran2016}, and the recent application of Fisher Separability analysis \cite{AlberganteSepar2019}.
+We also included a variation of the molecular cross-validation (MCVR, modified to be applied on normalized data - see methods) approach originally developed to select an calibrate denoising methods \cite{batsonMolecularCrossValidationSingleCell2019}, but which has also been applied to estimate dimensionality \citep{wagnerMonetOpensource2020}.
 
-{\color{red}We compared the various estimates in their ability to retrieve the intrinsic number of dimensions in a dataset, based on Seurat's weighted PCA space.
+We compared the various estimates in their ability to retrieve the intrinsic number of dimensions in a dataset, based on Seurat's weighted PCA space.
 As a first approximation of the true dimensionality, we computed the variance in each principal component (based on Seurat's weighted PCA space) that was explained by the subpopulations, which sharply decreased after the first few components in most datasets (Figure \ref{fig:dim}A).
-These truth-based estimates were then compared to those of the above dimensionality estimation methods (Figure \ref{fig:dim}B).} Of note, the methods differ widely in compute time (Figure \ref{fig:dim}B) and we saw no relationship between the accuracy of the estimates and the complexity of the method.
+These truth-based estimates were then compared to those of the above dimensionality estimation methods (Figure \ref{fig:dim}B).
+Of note, the methods differ widely in compute time (Figure \ref{fig:dim}B) and we saw no relationship between the accuracy of the estimates and the complexity of the method.
 Reasoning that over-estimating the number of dimensions is less problematic than under-estimating it, we kept the former methods for a full analysis of their impact on clustering (Figure \ref{fig:dim}C-D), when combined with \textit{sctransform} or standard \textit{Seurat} normalization.
-Although most methods performed well on the various clustering measures, the global maximum likelihood based on translated Poisson Mixture Model {\color{red}(\textit{maxLikGlobal}, using 10 or 20 nearest neighbors, although estimates were relatively stable across various values of \textit{k}; see Supplementary Figure 26) provided the dimensionality estimate that best separated challenging subpopulations (Figure \ref{fig:dim}C) and tended to result in the best clustering accuracy (Figure \ref{fig:dim}D).
-This method systematically selected many more components than were associated with the subpopulations in Figure \ref{fig:dim}A, suggesting that although these additional components appear individually uninformative, in combination they nevertheless contribute to classification.
-An important implication is that the commonly-used elbow method most likely underestimates dimensionality.}
+Although most methods performed well on the various clustering measures, the global maximum likelihood based on translated Poisson Mixture Model (\textit{maxLikGlobal}, using 10 or 20 nearest neighbors, although estimates were relatively stable across various values of \textit{k}; see Supplementary Figure 26) provided the dimensionality estimate that best separated challenging subpopulations (Figure \ref{fig:dim}C) and tended to result in the best clustering accuracy (Figure \ref{fig:dim}D).
+MCVR did not perform well for the datasets without UMIs (Koh and Kumar), which violate the assumptions of the method, and while it did increase the separability of some subpopulations (Figure \ref{fig:dim}C), this was at the expense of others.
+We observed a strong negative correlation (\textit{r}=-0.995, p~1.6e-8) between the number of dimensions called by MCVR and the log10 average library size of a dataset, which explained the former considerably better than the number of subpopulations.
+
+Of note, the optimal method in our evaluation (\textit{maxLikGlobal}) systematically selected many more components than were associated with the subpopulations in Figure \ref{fig:dim}A, suggesting that although these additional components appear individually uninformative, in combination they nevertheless contribute to classification.
+An important implication is that the commonly-used elbow method most likely underestimates dimensionality.
 
@@
 
-{\color{red}\subsection*{Overview of interactions between steps}
+\subsection*{Overview of interactions between steps}
 
 We selected the most important alternatives for each parameter (see Methods) in order to explore their interactions.
-After accounting, in the model, for baseline differences between datasets and the impact of the resolution parameter, ANOVA attributed the largest proportion of variance in the main cluster metrics to the choice of dimensionality reduction, followed by normalization, and very little to filtering (Supplementary Figure 28A).
+{\color{red}Unsupervised clustering of the results reveals general patterns in the results as well as the relative importance of the parameters (Figure \ref{fig:overall}), with the clustering method having the most important impact on the results.
+This is however chiefly due to number of clusters called: an ANOVA on all analyses after accounting, in the model, for baseline differences between datasets and the impact of Seurat's resolution parameter, attributed relatively little importance to the clustering method, and instead attributed the largest proportion of variance in the main cluster metrics to the number of dimensions used, followed by normalization (Supplementary Figure 28A).
 Because the choice of dimension reduction impacts the number of clusters, we repeated the analysis of variance in ARI among the combinations of parameters that yielded the true number of clusters (Supplementary Figure 28B).
-We observed that although normalization was the most important parameter among these analyses, the choice of dimensionality came closely behind (Supplementary Figure 28B), suggesting that more attention should be paid to this relatively neglected step.
+We observed that although normalization was the most important parameter among these analyses, the choice of dimensionality (as well as feature selection) came closely behind (Supplementary Figure 28B), suggesting that more attention should be paid to this relatively neglected step.
+
+The greater importance of normalization, dimensionality and feature selection over filtering steps is also visible in the clustering of (Figure \ref{fig:overall}).
+The combination of \textit{scran} normalization with the elbow method for selection the number of dimensions appeared to offer the worst results.
+The top methods (bottom of Figure \ref{fig:overall}) confirm the results of the previous sections, highlighting especially \textit{sctransform}, deviance-based feature selection, and \textit{maxLikGlobal} dimensionality estimates as top methods.
+Interestingly, while the combination of \textit{sctransform} with \textit{Seurat} clustering was particularly performant for most datasets, for the simulated `simMix1` dataset instead standard \textit{Seurat} normalization combined with \textit{scran}-based clustering proved superior.}
 
@@
 For instance, we confirmed that doublet removal and \textit{sctransform} normalization were beneficial independently of other steps, and that, again independendently of other steps, more stringent filtering did not yield any major improvement over lenient filtering (Supplementary Figure 29).
-Similarly, deviance-based feature selection outcompeted \textit{Seurat}'s default selection method independently of changes at other steps (Supplementary Figure 30).}
+Similarly, deviance-based feature selection outcompeted \textit{Seurat}'s default selection method independently of changes at other steps (Supplementary Figure 30).
 
@@
 
-{\color{red}\subsection*{Application of \textit{pipeComp} to a different context}
+\subsection*{Application of \textit{pipeComp} to a different context}
 
@@
 Nevertheless, the FDR remained acceptable with up to three variables.
-These results therefore suggest that these methods are robust and can be widely applied.}
+These results therefore suggest that these methods are robust and can be widely applied.
 
@@
 
-\subsection*{{\color{red}Practical} recommendations}
+\subsection*{Practical recommendations}
 
-On the basis of our findings, we can make a number of \textcolor{red}{practical} analysis recommendations, also summarized in Supplementary Figure 34:
+On the basis of our findings, we can make a number of practical analysis recommendations, also summarized in Supplementary Figure 34:
 
@@
      \item Doublet detection and removal is advised and can be performed at little computing cost with software such as \textit{scDblFinder} or \textit{scds}.
-     \item Distribution-based cell filtering fails to capture doublets and should use relatively lenient cutoffs (e.g., 5 MADs, or 3 MADs in at least 2 distributions \textcolor{red}{- see for instance our \textit{default} filters in the methods}) to exclude poor-quality cells while avoiding bias against some subpopulations.
+     \item Distribution-based cell filtering fails to capture doublets and should use relatively lenient cutoffs (e.g., 5 MADs, or 3 MADs in at least 2 distributions - see for instance our \textit{default} filters in the methods) to exclude poor-quality cells while avoiding bias against some subpopulations.
      \item Features filtering based on feature type did not appear beneficial.
@@
        \item Denoising appeared beneficial to the identification of some subpopulations in 10x datasets, but not in Smart-seq datasets.
-       \item We found especially \textit{ALRA} (with prior normalization), \textit{DrImpute} (with prior processing) and \textit{DCA} to offer the best performances\textcolor{red}{, although none of these methods was beneficial over all subpopulations}.
+       \item We found especially \textit{ALRA} (with prior normalization), \textit{DrImpute} (with prior processing) and \textit{DCA} to offer the best performances, although none of these methods was beneficial over all subpopulations.
    \end{itemize}
@@
      \item Similarly to previous reports \cite{SunDimRed2019}, we recommend the irlba-based PCA using weighting of the components, as implemented in \textit{Seurat}. 
-     \item We advise against the commonly-used elbow method (because it is too conservative) and the jackstraw method (low performance at great computational costs) for deciding on the components to include, and recommend a method like the \textit{global maximum likelihood based on translated Poisson Mixture Model} method (e.g., implemented in the  \texttt{maxLikGlobalDimEst} function of \texttt{intrinsicDimension}, in our case using the \textcolor{red}{10 or }20 nearest neighbors).
+     \item We advise against the commonly-used elbow method (because it is too conservative) and the jackstraw method (low performance at great computational costs) for deciding on the components to include, and recommend a method like the \textit{global maximum likelihood based on translated Poisson Mixture Model} method (e.g., implemented in the  \texttt{maxLikGlobalDimEst} function of \texttt{intrinsicDimension}, in our case using the 10 or 20 nearest neighbors).
    \end{itemize}
@@
 
-{\color{red}Among all tested pipelines, performance varied greatly across datasets.
+Among all tested pipelines, performance varied greatly across datasets.
 The ZhengMix8eq and simMix1 datasets were the most challenging datasets across all preprocessing steps that we tested.
@@
 The challenge of distinguishing subpopulations of T-cells in the Zhengmix8eq dataset was previously discussed \cite{duoClustering2018}.
-In contrast, while the Koh dataset harbors as many populations, they originate from more distinct groups (from hESC in different differentiation stages), while the Kumar and mixology datasets harbor a very low subpopulation complexity and most methods applied on them performed well.}
+In contrast, while the Koh dataset harbors as many populations, they originate from more distinct groups (from hESC in different differentiation stages), while the Kumar and mixology datasets harbor a very low subpopulation complexity and most methods applied on them performed well.
 
@@
 also highlighted the usefulness of \textit{ALRA} when followed by differential and trajectory analysis both in UMI and Fluidigm technologies.
+The performance of \textit{scVI} in our evaluation was lower than in the original publication presenting the method \cite{lopezDeep2018}.
+One of the main strengths of \textit{scVI} is its scalability to bigger datasets as the model can be enhanced by a bigger training set.
+The model was thus reported as less adapted to datasets with a higher number of genes compared to their number of cells.
+As all of our benchmarking datasets fall in this category, it is very likely that the poor performance of \textit{scVI} in our evaluation is a consequence of the poorer training that \textit{scVI} can perform.
+For instance, the original article could train their model on a PBMC dataset of 12'039 cells whereas we used a FACS-sorted version of the same experiment, down-sampled to only 6'498 cells at maximum.
+Another limitation is that the imputation and normalization performed by \textit{scVI} were initially not designed to improve cluster detection.
+On the opposite to the low-dimensional space returned by the model was intended for this purpose and indeed showed a higher ranking compared to other methods, as demonstrated in the original article [NOTE TO MYSELF; adapt to results.
+For instance Hou2020 found that scVI was not among the top performers wrt to clustering based on latent space.
+And Aparicio 2020 that found a low silhouette score using scVI's imputations as an input for clustering].
 
@@
 Instead, the \textit{CellBench} package was recently proposed to address a similar need \cite{suCellbench}, offering an elegant piping syntax to combine alternative methods at successive steps.
-A key additional feature of \textit{pipeComp} is its ability to perform evaluation in parallel to the pipeline and offering on-the-fly evaluation at {\color{red}multiple steps.
+A key additional feature of \textit{pipeComp} is its ability to perform evaluation in parallel to the pipeline and offering on-the-fly evaluation at multiple steps.
 This avoids the need to store potentially large intermediate data for all possible permutations and thus allowing a combinatorial benchmark.
-This, along with the set-up simplicity (demonstrated here by our separate SVA-DEA benchmark), also distinguishes \textit{pipeComp} from general workflow frameworks such as \textit{drake} \citep{landauDrakePackage2018}; indeed we estimate that the benchmarks discussed here would have taken approximately 16TB of storage had all non-redundant intermediates been saved (see the Supplementary methods for more detailed explanations of the framework and comparison to \textit{CellBench} and \textit{drake}).} In addition, the fact that benchmark functions are stored in the \texttt{PipelineDefinition} makes the benchmark more smoothly portable, making it easy to modify, extend with further methods, or apply to other datasets.
+This, along with the set-up simplicity (demonstrated here by our separate SVA-DEA benchmark), also distinguishes \textit{pipeComp} from general workflow frameworks such as \textit{drake} \citep{landauDrakePackage2018}; indeed we estimate that the benchmarks discussed here would have taken approximately 16TB of storage had all non-redundant intermediates been saved (see the Supplementary methods for more detailed explanations of the framework and comparison to \textit{CellBench} and \textit{drake}).
+In addition, the fact that benchmark functions are stored in the \texttt{PipelineDefinition} makes the benchmark more smoothly portable, making it easy to modify, extend with further methods, or apply to other datasets.
 
@@
 \textit{pipeComp} is a flexible R framework for evaluating methods and parameter combinations in a complex pipeline, computing on-the-fly multilevel evaluation metrics.
-Applying this framework to scRNAseq clustering enabled us to make {\color{red}clear} recommendations on the steps of filtering, normalization, feature selection, denoising, dimensionality reduction and clustering (Supplementary Figure 34).
+Applying this framework to scRNAseq clustering enabled us to make
+clear recommendations on the steps of filtering, normalization, feature selection, denoising, dimensionality reduction and clustering (Supplementary Figure 34).
 We demonstrate how a diversity of multilevel metrics can be more robust, more sensitive, and more nuanced than simply evaluating the final clustering performance.
-We {\color{red}further showed} that the \textit{pipeComp} framework can be applied to extend the current field of benchmarking, as well as to investigate pipelines other domains with multi-step computational analyses.
+We further showed that the \textit{pipeComp} framework can be applied to extend the current field of benchmarking, as well as to investigate pipelines other domains with multi-step computational analyses.
 
@@
 Analyses were performed in R 3.6.0 (Bioconductor 3.9) and the following packages were installed from GitHub repositories: \textit{Seurat} (version 3.0.0), \textit{sctransform} (0.2.0), \textit{DoubletFinder} (2.0.1), \textit{scDblFinder} (1.1.4), \textit{scds} (1.0.0), \textit{SAVERX} (1.0.0), \textit{scImpute} (0.0.9), \textit{ALRA} (commit 7636de8), \textit{DCA} (0.2.2), \textit{DrImpute} (1.2), \textit{ENHANCE} (R version, commit 1571696), \textit{SAVERX} (1.0.0).
-{\color{red}The glmPCA was performed using the \textit{glmpca} package, while the code for deviance calculation was obtained from \url{https://github.com/willtownes/scrna2019} (commit 1ddcc30ebb95d083a685f12fe81d35dd1b0cb1b2).}
+The glmPCA was performed using the \textit{glmpca} package, while the code for deviance calculation was obtained from \url{https://github.com/willtownes/scrna2019} (commit 1ddcc30ebb95d083a685f12fe81d35dd1b0cb1b2).
 
@@
 
+For the molecular cross-validation approach, we relied on a R implementation by Matthew Young (\url{https://github.com/constantAmateur/MCVR}), with minor modifications (see wrapper in the \textit{pipeComp} package).
+Briefly, since there is no a priori relationship between the dimensionality of the count data and that of the PCA based on the normalized data, MCVR does not use the original approach based Poisson loss, but the mean squared error of the normalized data, and uses balanced 50-50 splits.
+Because of the 50\% split, dimensionality may be underestimated, and titration did suggest that was the case for some of our datasets.
+
 \subsection*{Filter sets}
@@
 The \textit{stringent} set of filters uses the same thresholds, but excludes a cell if it is an outlier on any single distribution. 
-{\color{red}The \textit{veryStringent} set of filters excludes any cell that is an outlier by at least 2 MADs (lower or higher) on any of: log10\_total\_counts, log10\_total\_features, pct\_counts\_Mt, or pct\_counts\_in\_top\_20\_features.}
+The \textit{veryStringent} set of filters excludes any cell that is an outlier by at least 2 MADs (lower or higher) on any of: log10\_total\_counts, log10\_total\_features, pct\_counts\_Mt, or pct\_counts\_in\_top\_20\_features.
 The \textit{lenient} set of filters excludes cells that are outliers on at least two distributions by at least 5 MADs, except for pct\_counts\_Mt where the threshold is $>$ 3 MADs and $>$ 0.08.
@@
 
-\subsection*{\textit{scVI} normalization and LD}
+\subsection*{\textit{scVI} normalization and LD} [NOTE TO MYSELF; adapt to the new runs]
 
@@
 The resulting evaluation is available in the Supplementary Figure 13.
+
+Using \textit{scVI}, we considered as imputed values the mean of the negative binomial distribution modeling the counts (accessed through the \textit{imputation} method).
+Following \textit{scVI}'s online documentation, we used the \textit{VAE} model, using a train size of 1, a training on 400 epochs and a learning rate of 0.001.
+As a major limitation of \textit{scVI} is the training on dataset with low gene to cell ratio, the features were subsampled using the \textit{subsample_genes} function.
+All other parameters were left unmodified.
+
+The normalized values were computed from the normalized mean of the negative binomial distribution modeling the counts (accessed through the \textit{get_sample_scale} method).
+The same parameters as above were used, among which the early feature selection step that is usually performed latter in our pipeline.
+
+Two models of the \textit{scVI} package computing a latent representation of the data were tested.
+First, we evaluated the mean of the posterior distribution for the latent space, still based on the \textit{VAE} model (accessed through the \textit{get_latent} function).
+All other parameters are the same as above.
+The second model that we evaluated was the linearly decoded VAE of \textit{scVI} (\textit{LDVAE} model).
+It was trained on 250 epochs, a learning rate of 0.001 and computed 50 dimensions.
+For some smaller datasets, the training phase aborted.
+In such cases, we followed the recommendations of the package and reduced the learning rate by a factor 10 until the model could properly converge.
+Given that the other methods that we evaluated return PCs ordered by their variance explained, we applied to the same reordering to the latent space returned by \textit{scVI}.
+For both models.
+we relied on the feature selections results from the pipeline and not on the \textit{subsample_genes} function.
+All other parameters were left unmodified.
 
@@
     \centering
+    \includegraphics[width=\textwidth, keepaspectratio]{{overall.pdf}}
+    \caption{\textbf{Overview of the combination of main alternative parameters.} Unsupervised clustering of the results (above) and top results (below).
+The color-mapping schemes are the same as described in the previous figures.}
+    \label{fig:overall}
+\end{figure}
+
+\begin{figure}
+    \centering
     \includegraphics[width=\textwidth, keepaspectratio]{{svadea.pdf}}

### 2020-07-02 supplementary_figures/supp5_dr.Rmd
added supp figure on dimensionality-nbClusts

@@
 ```{r, fig.width=8, fig.height=8.5}
-library(viridis)
-res <- readRDS("../data/clust_endSummary.rds")
-res2 <- res[which(res$clust=="clust.seurat" & res$k==20 & res$dims>2 & res$resolution > 0.05 & res$resolution <= 1),]
-res2 <- res2[which(res2$dr=="seurat.pca" & res2$dataset!="Kumar"),]
 res3 <- aggregate(res2[,"ARI",drop=FALSE], by=res2[,c("dataset","dims","resolution")], FUN=mean)
@@
 ### Supplementary Figure `r getFigNb()`
-Adjusted Rand Index of clustering depending on the resolution and number of dimensions used. Based on sctransform and seurat PCA.
+
+Adjusted Rand Index of clustering depending on the resolution and number of dimensions used. Based on sctransform
+and seurat PCA.
+
+\newpage
+
+# Supplementary Figure `r getFigNb(TRUE)`
+
+```{r, fig.width=7.5, fig.height=8.5}
+cl <- readRDS("../data/results_dimensionality.rds")$evaluation$clustering
+cl$delta.nbClust <- cl$n_clus-cl$true.nbClusts
+
+cl$dims <- gsub("maxLikGlobal$","maxLikGlobal(k=20)",cl$dims)
+cl$dims <- gsub("maxLikGlobal10$","maxLikGlobal(k=10)",cl$dims)
+cl <- cl[cl$norm=="norm.seurat" & !(cl$dims %in% c("10","30","essLocal.b")),]
+
+ggplot(cl, aes(x=dims, y=delta.nbClust, fill=dims)) + geom_violin() + 
+  facet_wrap(~dataset, scales = "free_x", ncol = 2) + coord_flip() + 
+  stat_summary(size=0.25, fun.data = mean_se, na.rm = TRUE) + 
+  geom_hline(yintercept = 0, linetype="dashed") + xlab("") +
+  theme(legend.position="none", text=element_text(size=10)) + 
+  ylab("Deviation from true number of clusters")
+```
+
+### Supplementary Figure `r getFigNb()`
+
+Deviation from the true number of clusters using different number of principal components (based the indicated dimensionality estimates) of the same Seurat-based PCA.
+Default pipeline parameters were used for the other steps, and the distributions represent the different resolutions of Seurat clustering.
+Across datasets, MCVR and maxLikGlobal appear to depart less from the true number of clusters.
+
+\newpage
 

### 2020-07-02 supplementary_figures/supp_overall.Rmd
added supp figure on dimensionality-nbClusts

@@
 cl$clustmethod <- relevel(cl$clustmethod, "clust.seurat")
+
+# checking for interactions:
+# mod <- lm(ARI~resolution*dataset+doubletmethod*sel*filt*norm*clustmethod*dims, data=cl)
+# co <- as.data.frame(coef(summary(mod)))
+# coi <- co[grep(":",row.names(co)),]
+# coi <- coi[grep("dataset", row.names(coi), invert=TRUE),]
+# colnames(coi) <- c("Estimate","SE","t","p.value")
+# coi <- coi[order(coi$p.value),]
 

### 2020-07-02 bmc_article.tex
Updates from Overleaf

@@
 Reasoning that the cells we wish to exclude are the cells that would be misclassified, we measured the rate of misclassification of each cell in each dataset across a variety of clustering pipelines, correcting for the median misclassification rate of the subpopulation and then evaluated what properties could be predictive of misclassification (Supplementary Figures 8-10).
-We could not identify any property or simple combination thereof that would be consistently predictive of misclassification; the only feature that consistently stood out across multiple datasets (the Zheng datasets) was that cells with very high read counts have a higher chance of being misclassified.
+We could not identify any property or simple combination thereof that would be consistently predictive of misclassification; the only feature that stood out across more than one dataset (the Zheng datasets) was that cells with very high read counts have a higher chance of being misclassified{\color{red}, although such a pattern was not observed in the other datasets.}
 
@@
 We next investigated the impact of different normalization strategies.
-Beside the standard log-normalization included in \textit{Seurat}, we tested \textit{scran}'s pooling-based normalization \cite{lunPooling2016}, \textit{sctransform}'s variance-stabilizing transformation \cite{hafemeisterSCtransform2019}, normalization based on stable genes \cite{linStableGenes2018, deekeStablyExpressed2018}, and \textit{SCnorm} \cite{bacherSCnorm2017} (\textit{scVI} \cite{lopezDeep2018} was included separately on a subset of parameter combinations due to high running times).
+Beside the standard log-normalization included in \textit{Seurat}, we tested \textit{scran}'s pooling-based normalization \cite{lunPooling2016}, \textit{sctransform}'s variance-stabilizing transformation \cite{hafemeisterSCtransform2019}, normalization based on stable genes \cite{linStableGenes2018, deekeStablyExpressed2018}, and \textit{SCnorm} \cite{bacherSCnorm2017} (\textit{scVI}-based normalization \cite{lopezDeep2018} was included separately{\color{red}, as it was not meant for this purpose and follows a slightly different flow}).
 In addition to log-normalization, the standard \textit{Seurat} clustering pipeline performs per-feature unit-variance scaling so that the PCA is not too strongly dominated by highly-expressed features.
@@
 Zhengmix8eq and simMix1) and subpopulations were consistently more challenging across all methods.
-An exception was \textit{scVI}, which performed poorly on most datasets in terms of both the subpopulation silhouette and clustering accuracy (Supplementary Figure 13; \textit{scVI} was run on a reduced set of pipelines).
+{\color{red}An exception was \textit{scVI}-based normalization, which performed similarly to other methods on some datasets (Kumar, simMix2, Zhengmix4eq, Zhengmix4uneq), but very poorly on others, in terms of both the subpopulation silhouette and clustering accuracy (Supplementary Figure 13; of note, these values were not meant to be used for clustering, and the usage of the \textit{scVI} latent space will be discussed below along with dimensionality reductions)}.
 Scaling (e.g.
@@
 
-The impact of the choice of dimensionality reduction method was far greater than that of normalization or feature selection (Supplementary Figure 23).
+The impact of the dimensionality reduction methods tested was far greater than that of normalization or feature selection (Supplementary Figure 23).
 Overall, weighting the components by their explained variance tended to increase silhouette width (Supplementary Figure 23A) and, in most datasets, clustering accuracy (Supplementary Figure 23B).
 GLM-PCA tended to increase the average silhouette width of already well-defined subpopulations, but \textit{Seurat}'s PCA procedure however proved superior on all metrics.
-Like GLM-PCA, \textit{scVI}'s Linear Decoder (LD) does not explicitly rely on normalized counts.
-Its performance was however the worst both in terms of silhouette width and clustering accuracy (Supplementary Figure 15), and it only performed well in some of the least challenging datasets (mixology10x5cl and simMix2).
+{\color{red} Like GLM-PCA, \textit{scVI}'s Linearly Decoded (LD) data \cite{SvenssonLDVAE2020} and latent space do not explicitly rely on normalized counts.
+Their performance was however worse than \textit{Seurat}'s PCA in view of the silhouette width (Supplementary Figure 15).
+In terms of clustering accuracy \textit{scVI}'s latent space performed better that its LD but was nevertheless less accurate than PCA-based alternatives on the datasets with a very low amount of cells (Koh and Kumar) and harboring many subpopulations (simMix1 and Zhengmix8eq).}
 
@@
 We evaluated the performance of dimensionality estimators implemented in the \textit{intrinsicDimension} package \cite{johnssonDimensionality2015}, as well as common procedures such as the `elbow` method, some scRNAseq-specific methods such as the JackStraw procedure \cite{ChungJackstraw2015} or \textit{scran}'s \texttt{denoisePCA} \cite{LunScran2016}, and the recent application of Fisher Separability analysis \cite{AlberganteSepar2019}.
-We also included a variation of the molecular cross-validation (MCVR, modified to be applied on normalized data - see methods) approach originally developed to select an calibrate denoising methods \cite{batsonMolecularCrossValidationSingleCell2019}, but which has also been applied to estimate dimensionality \citep{wagnerMonetOpensource2020}.
+{\color{red} We also included a variation of the molecular cross-validation (MCVR, modified to be applied on normalized data - see methods) approach originally developed to select an calibrate denoising methods \cite{batsonMolecularCrossValidationSingleCell2019}, but which has also been applied to estimate dimensionality \citep{wagnerMonetOpensource2020}.}
 
@@
 Reasoning that over-estimating the number of dimensions is less problematic than under-estimating it, we kept the former methods for a full analysis of their impact on clustering (Figure \ref{fig:dim}C-D), when combined with \textit{sctransform} or standard \textit{Seurat} normalization.
-Although most methods performed well on the various clustering measures, the global maximum likelihood based on translated Poisson Mixture Model (\textit{maxLikGlobal}, using 10 or 20 nearest neighbors, although estimates were relatively stable across various values of \textit{k}; see Supplementary Figure 26) provided the dimensionality estimate that best separated challenging subpopulations (Figure \ref{fig:dim}C) and tended to result in the best clustering accuracy (Figure \ref{fig:dim}D).
-MCVR did not perform well for the datasets without UMIs (Koh and Kumar), which violate the assumptions of the method, and while it did increase the separability of some subpopulations (Figure \ref{fig:dim}C), this was at the expense of others.
-We observed a strong negative correlation (\textit{r}=-0.995, p~1.6e-8) between the number of dimensions called by MCVR and the log10 average library size of a dataset, which explained the former considerably better than the number of subpopulations.
 
-Of note, the optimal method in our evaluation (\textit{maxLikGlobal}) systematically selected many more components than were associated with the subpopulations in Figure \ref{fig:dim}A, suggesting that although these additional components appear individually uninformative, in combination they nevertheless contribute to classification.
+{\color{red}\textit{MCVR} and the global maximum likelihood based on translated Poisson Mixture Model (\textit{maxLikGlobal}) tended to produce clusterings with smaller deviations to the true number of subpopulations (Supplementary Figure 26).
+Although most methods performed well on the various clustering measures, \textit{maxLikGlobal} (using 10 or 20 nearest neighbors, although estimates were relatively stable across various values of \textit{k} -- see Supplementary Figure 27) provided the dimensionality estimate that best separated challenging subpopulations (Figure \ref{fig:dim}C) and tended to result in the best clustering accuracy (Figure \ref{fig:dim}D).
+\texit{MCVR} did not perform well for the datasets without UMIs (Koh and Kumar), which violate the assumptions of the method, and while it did increase the separability of some subpopulations (Figure \ref{fig:dim}C), it was detrimental to the classification of some others.
+We observed a strong negative correlation (\textit{r}=-0.995, p~1.6e-8) between the number of dimensions called by \textit{MCVR} and the log10 average library size of a dataset, which explained the former considerably better than the number of subpopulations.
+
+Of note, the optimal method in our evaluation (\textit{maxLikGlobal})} systematically selected many more components than were associated with the subpopulations in Figure \ref{fig:dim}A, suggesting that although these additional components appear individually uninformative, in combination they nevertheless contribute to classification.
 An important implication is that the commonly-used elbow method most likely underestimates dimensionality.
@@
 However, because \textit{Seurat}'s resolution parameter had a large impact on the number of clusters identified (Supplementary Figures 2 and 24), \textit{Seurat} could always be coerced into producing the right number of clusters.
-Instead, the number of clusters found by \textit{scran} was considerably less influenced by the available parameters (number of nearest neighbors or steps in the random walk - see Supplementary Figure 27), and as a result \textit{scran}-based clustering sometimes never produced a partitioning with the right number of clusters.
+Instead, the number of clusters found by \textit{scran} was considerably less influenced by the available parameters (number of nearest neighbors or steps in the random walk - see Supplementary Figure 28), and as a result \textit{scran}-based clustering sometimes never produced a partitioning with the right number of clusters.
 This observation, along with \textit{scran}'s higher MI score, suggest that \textit{scran} sometimes simply divides a real subpopulation into two clusters (possibly tracking some unknown biological differences) rather than committing misclassification errors.
@@
 We selected the most important alternatives for each parameter (see Methods) in order to explore their interactions.
-{\color{red}Unsupervised clustering of the results reveals general patterns in the results as well as the relative importance of the parameters (Figure \ref{fig:overall}), with the clustering method having the most important impact on the results.
-This is however chiefly due to number of clusters called: an ANOVA on all analyses after accounting, in the model, for baseline differences between datasets and the impact of Seurat's resolution parameter, attributed relatively little importance to the clustering method, and instead attributed the largest proportion of variance in the main cluster metrics to the number of dimensions used, followed by normalization (Supplementary Figure 28A).
-Because the choice of dimension reduction impacts the number of clusters, we repeated the analysis of variance in ARI among the combinations of parameters that yielded the true number of clusters (Supplementary Figure 28B).
-We observed that although normalization was the most important parameter among these analyses, the choice of dimensionality (as well as feature selection) came closely behind (Supplementary Figure 28B), suggesting that more attention should be paid to this relatively neglected step.
+{\color{red}Unsupervised clustering on the evaluation metrics reveals general patterns in the results as well as the relative importance of the parameters (Figure \ref{fig:overall}), with the clustering method having the most important impact on the results.
+This is however chiefly due to number of clusters called: an ANOVA on all analyses after accounting, in the model, for baseline differences between datasets and the impact of \textit{Seurat}'s resolution parameter, attributed relatively little importance to the clustering method, and instead attributed the largest proportion of variance in the main cluster metrics to the number of dimensions used, followed by normalization (Supplementary Figure 29A).
+Because the choice of dimension reduction impacts the number of clusters, we repeated the analysis of variance in ARI among the combinations of parameters that yielded the true number of clusters (Supplementary Figure 29B).
+We observed that although normalization was the most important parameter among these analyses, the choice of dimensionality (as well as feature selection) came closely behind (Supplementary Figure 29B), suggesting that more attention should be paid to this relatively neglected step.
 
 The greater importance of normalization, dimensionality and feature selection over filtering steps is also visible in the clustering of (Figure \ref{fig:overall}).
-The combination of \textit{scran} normalization with the elbow method for selection the number of dimensions appeared to offer the worst results.
 The top methods (bottom of Figure \ref{fig:overall}) confirm the results of the previous sections, highlighting especially \textit{sctransform}, deviance-based feature selection, and \textit{maxLikGlobal} dimensionality estimates as top methods.
@@
 We next investigated whether the benefits of some methods were conditional on the choices at other steps.
-Although analysis of variance identified many weak but significant interactions, we found that the benefits brought by performant methods were robust to differences in other parts of the pipeline (Supplementary Figures 29-30).
-For instance, we confirmed that doublet removal and \textit{sctransform} normalization were beneficial independently of other steps, and that, again independendently of other steps, more stringent filtering did not yield any major improvement over lenient filtering (Supplementary Figure 29).
-Similarly, deviance-based feature selection outcompeted \textit{Seurat}'s default selection method independently of changes at other steps (Supplementary Figure 30).
+Although analysis of variance identified many weak but significant interactions, we found that the benefits brought by performant methods were robust to differences in other parts of the pipeline (Supplementary Figures 30-31).
+For instance, we confirmed that doublet removal and \textit{sctransform} normalization were beneficial independently of other steps, and that, again independently of other steps, more stringent filtering did not yield any major improvement over lenient filtering (Supplementary Figure 30).
+Similarly, deviance-based feature selection outcompeted \textit{Seurat}'s default selection method independently of changes at other steps (Supplementary Figure 31).
+{\color{red} ANOVA did reveal some significant (negative) interactions: the top two-way interaction was between \textit{scran} normalization and the elbow method for selection the number of dimensions, and indeed that combination was involved in all of the 30 lowest-ranking pipelines.
+Furthermore, all significant interactions involved the elbow method, suggesting that some methods were less robust to underestimated dimensionality.}
 
@@
 Since preliminary analysis showed that all methods performed equally well or better on normalized data, we applied them after filtering and normalization, but before scaling and reduction.
-Although some of the methods (e.g., \textit{DRImpute\_process} and \textit{alra\_norm}) did improve the separability of some more elusive subpopulations, no method had a systematically positive impact on the average silhouette width across all subpopulations (Supplementary Figure 31A).
-When restricting ourselves to clustering analyses that yielded the `right' number of clusters, all tested methods improved classification  compared to a scenario with no imputation step (`none' label, Supplementary Figure 31B).
+Although some of the methods (e.g., \textit{DRImpute\_process} and \textit{alra\_norm}) did improve the separability of some more elusive subpopulations, no method had a systematically positive impact on the average silhouette width across all subpopulations (Supplementary Figure 32A).
+When restricting ourselves to clustering analyses that yielded the `right' number of clusters, all tested methods improved classification  compared to a scenario with no imputation step (`none' label, Supplementary Figure 32B).
 However, the situation was not so straightforward with alternative metrics, where some methods (e.g., \textit{ENHANCE}) consistently underperformed.
@@
 We developed 3 benchmark datasets: 1) the \texttt{ipsc} data is a random selection of 10 vs 10 samples from \citep{carcamoOriveAnalysisTranscriptional2017}, which are very heterogeneous, and foldchanges were manually added to 300 genes; 2) the \texttt{seqc} dataset contains 5 vs 5 samples of two mixtures of the SEQC consortium with different ERCC spike-in mixes \citep{ComprehensiveAssessment2014}, and includes a partially correlated batch effect; 3) A \texttt{simulation} of 8 vs 8 samples with two technical vectors of variation (see Methods for more details).
-As previously reported \citep{ComprehensiveAssessment2014}, we found that more stringent filtering improved DEA (Supplementary Figure 32A) and that the different DEA methods had similar performances (Figure \ref{fig:svadea}A and Supplementary Figure 32A).
-Including a SVA-based step increased sensitivity across all datasets, particularly in datasets with technical vectors of variation (\textit{seqc} and \textit{simulation}), and benefited all DEA methods in a similar fashion (Figure \ref{fig:svadea}A,C and Supplementary Figures 32-33).
+As previously reported \citep{ComprehensiveAssessment2014}, we found that more stringent filtering improved DEA (Supplementary Figure 33A) and that the different DEA methods had similar performances (Figure \ref{fig:svadea}A and Supplementary Figure 33A).
+Including a SVA-based step increased sensitivity across all datasets, particularly in datasets with technical vectors of variation (\textit{seqc} and \textit{simulation}), and benefited all DEA methods in a similar fashion (Figure \ref{fig:svadea}A,C and Supplementary Figures 33-34).
 With the exception of the \texttt{RUVr} method, which led to a high False Discovery Rate (FDR), all other methods showed comparable gains in sensitivity while maintaining a decent error control (if a little above the nominal FDR), even when increasing the number of surrogate variables used (Figure \ref{fig:svadea}B-C).
@@
 
-On the basis of our findings, we can make a number of practical analysis recommendations, also summarized in Supplementary Figure 34:
+On the basis of our findings, we can make a number of practical analysis recommendations, also summarized in Supplementary Figure 35:
 
@@
 also highlighted the usefulness of \textit{ALRA} when followed by differential and trajectory analysis both in UMI and Fluidigm technologies.
-The performance of \textit{scVI} in our evaluation was lower than in the original publication presenting the method \cite{lopezDeep2018}.
-One of the main strengths of \textit{scVI} is its scalability to bigger datasets as the model can be enhanced by a bigger training set.
-The model was thus reported as less adapted to datasets with a higher number of genes compared to their number of cells.
-As all of our benchmarking datasets fall in this category, it is very likely that the poor performance of \textit{scVI} in our evaluation is a consequence of the poorer training that \textit{scVI} can perform.
-For instance, the original article could train their model on a PBMC dataset of 12'039 cells whereas we used a FACS-sorted version of the same experiment, down-sampled to only 6'498 cells at maximum.
-Another limitation is that the imputation and normalization performed by \textit{scVI} were initially not designed to improve cluster detection.
-On the opposite to the low-dimensional space returned by the model was intended for this purpose and indeed showed a higher ranking compared to other methods, as demonstrated in the original article [NOTE TO MYSELF; adapt to results.
-For instance Hou2020 found that scVI was not among the top performers wrt to clustering based on latent space.
-And Aparicio 2020 that found a low silhouette score using scVI's imputations as an input for clustering].
+
+{\color{red} The performance of \textit{scVI} in our evaluation was lower than in the original publication presenting the method \cite{lopezDeep2018}.
+Our results are however in line with other studies comparing silhouette score \cite{Aparicio2020} and clustering accuracy \cite{HouImput2020} of \textit{scVI}'s latent space.
+However, these studies, like ours, used datasets with relatively few cells (i.e.
+fewer than genes), for which \textit{scVI} was reported by its authors to be less adapted.
+Further work will therefore be required to assess compare its performance on larger datasets.}
 
@@
 Applying this framework to scRNAseq clustering enabled us to make
-clear recommendations on the steps of filtering, normalization, feature selection, denoising, dimensionality reduction and clustering (Supplementary Figure 34).
+clear recommendations on the steps of filtering, normalization, feature selection, denoising, dimensionality reduction and clustering (Supplementary Figure 35).
 We demonstrate how a diversity of multilevel metrics can be more robust, more sensitive, and more nuanced than simply evaluating the final clustering performance.
@@
 All code to reproduce the simulations and figures is available in the \url{https://github.com/markrobinsonuzh/scRNA\_pipelines\_paper} repository, which also includes the basic datasets with a standardized annotation.
+{\color{red}All wrappers for methods used in the study can be found in the external data of the \textit{pipeComp} package (\url{https://github.com/plger/pipeComp/tree/master/inst/extdata}).}
 
@@
 
-For the molecular cross-validation approach, we relied on a R implementation by Matthew Young (\url{https://github.com/constantAmateur/MCVR}), with minor modifications (see wrapper in the \textit{pipeComp} package).
+{\color{red} For the molecular cross-validation approach, we relied on a R implementation by Matthew Young (\url{https://github.com/constantAmateur/MCVR}), with minor modifications (see wrapper in the \textit{pipeComp} package).
 Briefly, since there is no a priori relationship between the dimensionality of the count data and that of the PCA based on the normalized data, MCVR does not use the original approach based Poisson loss, but the mean squared error of the normalized data, and uses balanced 50-50 splits.
-Because of the 50\% split, dimensionality may be underestimated, and titration did suggest that was the case for some of our datasets.
+Because of the 50\% split, dimensionality may be underestimated, and titration did suggest that was the case for some of our datasets.}
 
@@
 
-\subsection*{\textit{scVI} normalization and LD} [NOTE TO MYSELF; adapt to the new runs]
+\subsection*{\textit{scVI} normalization and LD}
 
-The evaluation of \textit{scVI} algorithm was performed separate to the other pipeline combinations as the model training on our reference datasets was especially resource-consuming.
-As it was not practically feasible to perform the training step on many combination of parameters as it was done with other tools, we evaluated it on a representative subset of pipelines using \textit{sctransform} normalization and \textit{Seurat}'s PCA dimension reduction method.
+The evaluation of \textit{scVI} algorithm was performed separate to the other pipeline combinations as the model training on our reference datasets was especially resource-consuming {\color{red} and follows a different flow from the other pipelines.
+In particular, for datasets with fewer cells than genes, \textit{scVI} requires a feature selection to be performed prior to the normalization and imputation steps; furthermore, its dimensional reduction steps rely directly on raw counts.
+We therefore evaluated it separately and compared it to a representative subset of pipelines using \textit{sctransform} normalization and \textit{Seurat}'s PCA dimension reduction method.
 The resulting evaluation is available in the Supplementary Figure 13.
@@
 Following \textit{scVI}'s online documentation, we used the \textit{VAE} model, using a train size of 1, a training on 400 epochs and a learning rate of 0.001.
-As a major limitation of \textit{scVI} is the training on dataset with low gene to cell ratio, the features were subsampled using the \textit{subsample_genes} function.
+As a major limitation of \textit{scVI} is the training on dataset with low gene to cell ratio, the features were subsampled to 1000 using the \textit{subsample\_genes} function, which resembles the \textit{Seurat} V3 VST gene selection method.
 All other parameters were left unmodified.
+The normalized values were computed from the normalized mean of the negative binomial distribution modeling the counts (accessed through the \textit{get\_sample\_scale} method).
 
-The normalized values were computed from the normalized mean of the negative binomial distribution modeling the counts (accessed through the \textit{get_sample_scale} method).
-The same parameters as above were used, among which the early feature selection step that is usually performed latter in our pipeline.
-
-Two models of the \textit{scVI} package computing a latent representation of the data were tested.
-First, we evaluated the mean of the posterior distribution for the latent space, still based on the \textit{VAE} model (accessed through the \textit{get_latent} function).
-All other parameters are the same as above.
+Two models of the \textit{scVI} package computing a dimension reduction on the data were tested.
+First, we evaluated the mean of the posterior distribution for the latent space, still based on the \textit{VAE} model (accessed through the \textit{get\_latent} function).
+All other parameters were set as described above or left as default.
 The second model that we evaluated was the linearly decoded VAE of \textit{scVI} (\textit{LDVAE} model).
@@
 In such cases, we followed the recommendations of the package and reduced the learning rate by a factor 10 until the model could properly converge.
-Given that the other methods that we evaluated return PCs ordered by their variance explained, we applied to the same reordering to the latent space returned by \textit{scVI}.
+Given that the other methods that we evaluated returned PCs ordered by their variance explained, we applied to the same reordering to the latent space returned by \textit{scVI}.
 For both models.
-we relied on the feature selections results from the pipeline and not on the \textit{subsample_genes} function.
+we relied on the feature selections results from the pipeline and not on the \textit{subsample\_genes} function.
 All other parameters were left unmodified.
+}
 
@@
 
-The set of alternatives found for the overall run (Supplementary Figures 28-30) can be found in the \href{https://github.com/markrobinsonuzh/scRNA_pipelines_paper/blob/master/data/overall_run.Rmd}{paper's repository}.
+The set of alternatives found for the overall run (Supplementary Figures 29-31) can be found in the \href{https://github.com/markrobinsonuzh/scRNA_pipelines_paper/blob/master/data/overall_run.Rmd}{paper's repository}.
 The analysis of variance was performed with the following model:

### 2020-07-02 bmc_article.tex
Updates from Overleaf

@@
 
-{\color{red}\textit{MCVR} and the global maximum likelihood based on translated Poisson Mixture Model (\textit{maxLikGlobal}) tended to produce clusterings with smaller deviations to the true number of subpopulations (Supplementary Figure 26).
+{\color{red} \textit{MCVR} and the global maximum likelihood based on translated Poisson Mixture Model (\textit{maxLikGlobal}) tended to produce clusterings with smaller deviations to the true number of subpopulations (Supplementary Figure 26).
 Although most methods performed well on the various clustering measures, \textit{maxLikGlobal} (using 10 or 20 nearest neighbors, although estimates were relatively stable across various values of \textit{k} -- see Supplementary Figure 27) provided the dimensionality estimate that best separated challenging subpopulations (Figure \ref{fig:dim}C) and tended to result in the best clustering accuracy (Figure \ref{fig:dim}D).
-\texit{MCVR} did not perform well for the datasets without UMIs (Koh and Kumar), which violate the assumptions of the method, and while it did increase the separability of some subpopulations (Figure \ref{fig:dim}C), it was detrimental to the classification of some others.
+\textit{MCVR} did not perform well for the datasets without UMIs (Koh and Kumar), which violate the assumptions of the method, and while it did increase the separability of some subpopulations (Figure \ref{fig:dim}C), it was detrimental to the classification of some others.
 We observed a strong negative correlation (\textit{r}=-0.995, p~1.6e-8) between the number of dimensions called by \textit{MCVR} and the log10 average library size of a dataset, which explained the former considerably better than the number of subpopulations.

### 2020-07-03 bmc_article.tex
Updates from Overleaf

@@
 Their performance was however worse than \textit{Seurat}'s PCA in view of the silhouette width (Supplementary Figure 15).
-In terms of clustering accuracy \textit{scVI}'s latent space performed better that its LD but was nevertheless less accurate than PCA-based alternatives on the datasets with a very low amount of cells (Koh and Kumar) and harboring many subpopulations (simMix1 and Zhengmix8eq).}
+In terms of clustering accuracy, \textit{scVI}'s latent space performed better that its LD but was nevertheless less accurate than PCA-based alternatives on the datasets with a very low amount of cells (Koh and Kumar) and harboring many subpopulations (simMix1 and Zhengmix8eq).}
 
@@
 
-The greater importance of normalization, dimensionality and feature selection over filtering steps is also visible in the clustering of (Figure \ref{fig:overall}).
+The greater importance of normalization, dimensionality and feature selection over filtering steps is also visible in the clustering (Figure \ref{fig:overall}).
 The top methods (bottom of Figure \ref{fig:overall}) confirm the results of the previous sections, highlighting especially \textit{sctransform}, deviance-based feature selection, and \textit{maxLikGlobal} dimensionality estimates as top methods.
-Interestingly, while the combination of \textit{sctransform} with \textit{Seurat} clustering was particularly performant for most datasets, for the simulated `simMix1` dataset instead standard \textit{Seurat} normalization combined with \textit{scran}-based clustering proved superior.}
+Interestingly, while the combination of \textit{sctransform} with \textit{Seurat} clustering was particularly performant for most datasets, for the simulated `simMix1` dataset,  standard \textit{Seurat} normalization combined with \textit{scran}-based clustering proved superior.}
 
@@
 {\color{red} For the molecular cross-validation approach, we relied on a R implementation by Matthew Young (\url{https://github.com/constantAmateur/MCVR}), with minor modifications (see wrapper in the \textit{pipeComp} package).
-Briefly, since there is no a priori relationship between the dimensionality of the count data and that of the PCA based on the normalized data, MCVR does not use the original approach based Poisson loss, but the mean squared error of the normalized data, and uses balanced 50-50 splits.
+Briefly, since there is no \textit{a priori} relationship between the dimensionality of the count data and that of the PCA based on the normalized data, MCVR does not use the original approach based Poisson loss, but the mean squared error of the normalized data, and uses balanced 50-50 splits.
 Because of the 50\% split, dimensionality may be underestimated, and titration did suggest that was the case for some of our datasets.}
@@
 
-The evaluation of \textit{scVI} algorithm was performed separate to the other pipeline combinations as the model training on our reference datasets was especially resource-consuming {\color{red} and follows a different flow from the other pipelines.
+The evaluation of \textit{scVI} algorithm was performed separate to the other pipeline combinations since the model training on our reference datasets was especially resource-consuming {\color{red} and follows a different flow from the other pipelines.
 In particular, for datasets with fewer cells than genes, \textit{scVI} requires a feature selection to be performed prior to the normalization and imputation steps; furthermore, its dimensional reduction steps rely directly on raw counts.
@@
 Following \textit{scVI}'s online documentation, we used the \textit{VAE} model, using a train size of 1, a training on 400 epochs and a learning rate of 0.001.
-As a major limitation of \textit{scVI} is the training on dataset with low gene to cell ratio, the features were subsampled to 1000 using the \textit{subsample\_genes} function, which resembles the \textit{Seurat} V3 VST gene selection method.
+As a major limitation of \textit{scVI} is the training on dataset with low gene to cell ratio, the features were subsampled to 2000 using the \textit{subsample\_genes} function, which resembles the \textit{Seurat} V3 VST gene selection method.
 All other parameters were left unmodified.
@@
 Filtering was performed with edgeR's \texttt{filterByExpr} function, changing the \texttt{min.count} argument.
-For \textit{vstsva}, we first applied the variance-stabilizing transformation from \textit{DESeq2}\citep{loveModeratedEstimation2014} before applying \texttt{sva::sva}.
+For \textit{vstsva}, we first applied the variance-stabilizing transformation from \textit{DESeq2} \citep{loveModeratedEstimation2014} before applying \texttt{sva::sva}.
 \textit{edgeR.QLF} refers to the quasi-likelihood version \citep{lundDetectingDifferential2012} of \textit{edgeR}'s fit.

### 2020-07-03 bmc_article.tex

@@
 Although most methods performed well on the various clustering measures, \textit{maxLikGlobal} (using 10 or 20 nearest neighbors, although estimates were relatively stable across various values of \textit{k} -- see Supplementary Figure 27) provided the dimensionality estimate that best separated challenging subpopulations (Figure \ref{fig:dim}C) and tended to result in the best clustering accuracy (Figure \ref{fig:dim}D).
-\textit{MCVR} did not perform well for the datasets without UMIs (Koh and Kumar), which violate the assumptions of the method, and while it did increase the separability of some subpopulations (Figure \ref{fig:dim}C), it was detrimental to the classification of some others.
+\textit{MCVR} did not perform well for the datasets without UMIs (Koh and Kumar), which violates the assumptions of the method, and while it did increase the separability of some subpopulations (Figure \ref{fig:dim}C), it was detrimental to the classification of some others.
 We observed a strong negative correlation (\textit{r}=-0.995, p~1.6e-8) between the number of dimensions called by \textit{MCVR} and the log10 average library size of a dataset, which explained the former considerably better than the number of subpopulations.
@@
 Our results are however in line with other studies comparing silhouette score \cite{Aparicio2020} and clustering accuracy \cite{HouImput2020} of \textit{scVI}'s latent space.
-However, these studies, like ours, used datasets with relatively few cells (i.e.
+These studies, like ours, used datasets with relatively few cells (i.e.
 fewer than genes), for which \textit{scVI} was reported by its authors to be less adapted.
-Further work will therefore be required to assess compare its performance on larger datasets.}
+We think that further work will be required to assess its performance on larger datasets.}
 
@@
 
-The evaluation of \textit{scVI} algorithm was performed separate to the other pipeline combinations since the model training on our reference datasets was especially resource-consuming {\color{red} and follows a different flow from the other pipelines.
+The evaluation of \textit{scVI} algorithm was performed separate to the other pipeline combinations since the model training on our reference datasets was especially resource-consuming {\color{red} and follows a different workflow from the other pipelines.
 In particular, for datasets with fewer cells than genes, \textit{scVI} requires a feature selection to be performed prior to the normalization and imputation steps; furthermore, its dimensional reduction steps rely directly on raw counts.
@@
 
-{\color{red}\subsection*{Overall run for interactions across steps}
+\subsection*{Overall run for interactions across steps}
 
@@
 The exact wrappers used can be found in the package at \url{https://github.com/plger/pipeComp/blob/master/inst/extdata/dea\_wrappers.R}.
-}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2020-07-03 bmc_article.tex
Updates from Overleaf

@@
 
-\parttitle{Background} The massive growth of single-cell RNA-sequencing (scRNAseq) and the methods for its analysis still lack sufficient and up-to-date benchmarks that could guide analytical choices.
+\parttitle{Background} Despite the massive growth of single-cell RNA-sequencing (scRNAseq) and corresponding computational methods, sufficient and up-to-date benchmarks to guide analytical choices are still lacking.
 Numerous benchmark studies already exist and cover most of scRNAseq processing and analytical methods but only a few give advice on a comprehensive pipeline.
@@
 The evaluation of \textit{scVI} algorithm was performed separate to the other pipeline combinations since the model training on our reference datasets was especially resource-consuming {\color{red} and follows a different workflow from the other pipelines.
-In particular, for datasets with fewer cells than genes, \textit{scVI} requires a feature selection to be performed prior to the normalization and imputation steps; furthermore, its dimensional reduction steps rely directly on raw counts.
+In particular, for datasets with fewer cells than genes, \textit{scVI} requires a feature selection to be performed prior to the normalization step; furthermore, its dimensional reduction steps rely directly on raw counts.
 We therefore evaluated it separately and compared it to a representative subset of pipelines using \textit{sctransform} normalization and \textit{Seurat}'s PCA dimension reduction method.
@@
 
-Using \textit{scVI}, we considered as imputed values the mean of the negative binomial distribution modeling the counts (accessed through the \textit{imputation} method).
+Using \textit{scVI}, we considered as normalized values the normalized mean of the negative binomial distribution modeling the counts (accessed through the \textit{get\_sample\_scale} method).
 Following \textit{scVI}'s online documentation, we used the \textit{VAE} model, using a train size of 1, a training on 400 epochs and a learning rate of 0.001.
@@
 All other parameters were left unmodified.
-The normalized values were computed from the normalized mean of the negative binomial distribution modeling the counts (accessed through the \textit{get\_sample\_scale} method).
+.
 

### 2020-08-04 bmc_article.tex
4overleaf

@@
-%% BioMed_Central_Tex_Template_v1.06
-%%                                      %
-%  bmc_article.tex            ver: 1.06 %
-%                                       %
+\documentclass[11pt]{article}
 
-%%IMPORTANT: do not delete the first line of this template
-%%It must be present to enable the BMC Submission system to
-%%recognise this template!!
+\renewcommand\familydefault{\sfdefault}
+\usepackage[left=3cm,right=3cm,top=2cm,bottom=3cm]{geometry}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                     %%
-%%  LaTeX template for BioMed Central  %%
-%%     journal article submissions     %%
-%%                                     %%
-%%          <8 June 2012>              %%
-%%                                     %%
-%%                                     %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% graphics, colors & math
+\usepackage{graphicx}
+\usepackage[dvipsnames,usenames]{xcolor}
+\usepackage{amsmath,amssymb, amsfonts,nicefrac,textcomp,icomma,bm}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                                                 %%
-%% For instructions on how to fill out this Tex template           %%
-%% document please refer to Readme.html and the instructions for   %%
-%% authors page on the biomed central website                      %%
-%% http://www.biomedcentral.com/info/authors/                      %%
-%%                                                                 %%
-%% Please do not use \input{...} to include other tex files.       %%
-%% Submit your LaTeX manuscript as one .tex document.              %%
-%%                                                                 %%
-%% All additional figures and files should be attached             %%
-%% separately and not embedded in the \TeX\ document itself.       %%
-%%                                                                 %%
-%% BioMed Central currently use the MikTex distribution of         %%
-%% TeX for Windows) of TeX and LaTeX.  This is available from      %%
-%% http://www.miktex.org                                           %%
-%%                                                                 %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% links & figures
+\usepackage{hyperref}
+\usepackage[labelfont=bf,font=small]{caption}
+\captionsetup[figure]{font={stretch=1}}
+\newcommand{\bcaption}[2]{\caption[#1]{\textbf{#1} #2}}
+\usepackage{float}
 
-%%% additional documentclass options:
-%  [doublespacing]
-%  [linenumbers]   - put the line numbers on margins
+% cross-referencing
+\usepackage{xr}
+\makeatletter
+\newcommand*{\addFileDependency}[1]{%
+  \typeout{(#1)}
+  \@addtofilelist{#1}
+  \IfFileExists{#1}{}{\typeout{No file #1.}}
+}
+\makeatother
 
-%%% loading packages, author definitions
+\usepackage{pdfpages}
 
-%\documentclass[twocolumn]{bmcart}% uncomment this for twocolumn layout and comment line below
-\documentclass{bmcart}
+% line spacing
+\usepackage{setspace}
+\renewcommand{\baselinestretch}{1.5}
+\setlength{\parindent}{2em}
 
-%%% Load packages
-%\usepackage{amsthm,amsmath}
-%\RequirePackage{natbib}
-%\RequirePackage[authoryear]{natbib}% uncomment this for author-year bibliography
-\usepackage[square,sort,comma,numbers]{natbib}
-\usepackage{hyperref}
-\usepackage[utf8]{inputenc} %unicode support
-%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
-%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
-\usepackage{graphicx}
-\usepackage{url}
+% line numbering
+\usepackage{etoolbox,lineno}
+\newcommand*{\Section}[1]{\par\nolinenumbers\section*{#1}\linenumbers}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                             %%
-%%  If you wish to display your graphics for   %%
-%%  your own use using includegraphic or       %%
-%%  includegraphics, then comment out the      %%
-%%  following two lines of code.               %%
-%%  NB: These line *must* be included when     %%
-%%  submitting to BMC.                         %%
-%%  All figure files must be submitted as      %%
-%%  separate graphics through the BMC          %%
-%%  submission process, not included in the    %%
-%%  submitted article.                         %%
-%%                                             %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% title layout
+\usepackage[affil-it]{authblk}
+\renewcommand\Authands{ \& }
 
-%\def\includegraphic{}
-%\def\includegraphics{}
+\usepackage{lipsum}
+\makeatletter
+\renewcommand{\maketitle}{\bgroup
+\setlength{\parindent}{0pt}\begin{flushleft}
+\textbf{\Large{\@title}}\\\vspace{1em}\@author
+\end{flushleft}\egroup}
+\makeatother
 
-%%% Put your definitions there:
-\startlocaldefs
-\endlocaldefs
+\author[1,2,3,*]{Pierre-Luc Germain}
+\author[1,2]{Anthony Sonrel}
+\author[1,2,*]{Mark D. Robinson}
 
-%%% Begin ...
+\affil[1]{Department of Molecular Life Sciences, University of Z{\"u}rich, Switzerland}
+\affil[2]{SIB Swiss Institute of Bioinformatics, Z{\"u}rich, Switzerland}
+\affil[3]{D-HEST Institute for Neuroscience, Swiss Federal Institute of Technology (ETH), Z{\"u}rich, Switzerland}
+\affil[*]{Correspondence to Pierre-Luc Germain (pierre-luc.germain@hest.ethz.ch) and Mark D. Robinson (mark.robinson@imls.uzh.ch)}
+
+\date{\today}
+
+% code font-style
+\newsavebox{\foobox}
+\newcommand{\slantbox}[2][0]{\mbox{%
+        \sbox{\foobox}{#2}%
+        \hskip\wd\foobox
+        \pdfsave
+        \pdfsetmatrix{1 0 #1 1}%
+        \llap{\usebox{\foobox}}%
+        \pdfrestore
+}}
+\newcommand\unslant[2][-.25]{\slantbox[#1]{$#2$}}
+\newcommand{\code}[1]{\unslant{\mathit{#1}}}
+
+% citations
+\usepackage[style=nature,citestyle=numeric,sorting=none]{biblatex}
+\renewcommand*{\bibfont}{\footnotesize}
+\addbibresource{refs.bib}
+
+\renewcommand{\cite}[1]{\textcolor{Blue}{$^[$\supercite{#1}$^]$}}
+\newcommand{\citet}[1]{\citeauthor{#1}\textcolor{Blue}{$^[$\supercite{#1}$^]$}}
+
+% check marks
+\usepackage{pifont}
+\newcommand{\cmark}{\textcolor{Green}{\ding{51}}}
+\newcommand{\xmark}{\textcolor{Red}{\ding{55}}}
+
+% tables
+\usepackage{colortbl}
+
+\title{pipeComp, a general framework for the evaluation of computational pipelines, reveals performant single-cell RNA-seq preprocessing tools}
+
 \begin{document}
 
-%%% Start of article front matter
-\begin{frontmatter}
+\thispagestyle{empty}
+\maketitle
 
-\begin{fmbox}
-\dochead{Research}
+\linenumbers
+%\modulolinenumbers[5]
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% Enter the title of your article here     %%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\Section{Abstract}
 
-\title{pipeComp, a general framework for the evaluation of computational pipelines, reveals performant single cell RNA-seq preprocessing tools}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% Enter the authors here                   %%
-%%                                          %%
-%% Specify information, if available,       %%
-%% in the form:                             %%
-%%   <key>={<id1>,<id2>}                    %%
-%%   <key>=                                 %%
-%% Comment or delete the keys which are     %%
-%% not used. Repeat \author command as much %%
-%% as required.                             %%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\author[
-   addressref={aff1,aff2,aff3},          % id's of addresses, e.g. {aff1,aff2}
-   email={pierre-luc.germain@hest.ethz.ch}   % email address
-]{\inits{PLG}\fnm{Pierre-Luc} \snm{Germain}}
-\author[
-   addressref={aff1,aff2},
-    email={anthony.sonrel@uzh.ch}
-]{\fnm{Anthony} \snm{Sonrel}}
-\author[
-   addressref={aff1,aff2},
-   corref={aff1},  % id of corresponding address, if any
-   email={mark.robinson@imls.uzh.ch}
-]{\inits{MR}\fnm{Mark D.} \snm{Robinson}}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% Enter the authors' addresses here        %%
-%%                                          %%
-%% Repeat \address commands as much as      %%
-%% required.                                %%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\address[id=aff1]{
-  \orgname{Department of Molecular Life Sciences, University of Zürich}, % university, etc
-  \street{Winterthurerstrasse 190},
-  \postcode{8057}
-  \city{Zürich},
-  \cny{Switzerland}
-}
-\address[id=aff2]{
-  \orgname{SIB Swiss Institute of Bioinformatics}, % university, etc
-  \city{Zürich},
-  \cny{Switzerland}
-}
-\address[id=aff3]{
-  \orgname{D-HEST Institute for Neurosciences, ETH Zürich},
-  \street{Winterthurerstrasse 190},
-  \postcode{8057}
-  \city{Zürich},
-  \cny{Switzerland}
-}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\end{fmbox}% comment this for two column layout
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% The Abstract begins here                 %%
-%%                                          %%
-%% Please refer to the Instructions for     %%
-%% authors on http://www.biomedcentral.com  %%
-%% and include the section headings         %%
-%% accordingly for your article type.       %%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{abstractbox}
-
-\begin{abstract}
-
-\parttitle{Background} Despite the massive growth of single-cell RNA-sequencing (scRNAseq) and corresponding computational methods, sufficient and up-to-date benchmarks to guide analytical choices are still lacking.
+The massive growth of single-cell RNA-sequencing (scRNAseq) and the methods for its analysis still lack sufficient and up-to-date benchmarks that could guide analytical choices.
 Numerous benchmark studies already exist and cover most of scRNAseq processing and analytical methods but only a few give advice on a comprehensive pipeline.
 Moreover, current studies often focused on isolated steps of the process and do not address the impact of a tool on both the intermediate and the final steps of the analysis.
-
-\parttitle{Results} Here, we present a flexible R framework for pipeline comparison with multi-level evaluation metrics.
+Here, we present a flexible R framework for pipeline comparison with multi-level evaluation metrics.
 We apply it to the benchmark of scRNAseq analysis pipelines using simulated and real datasets with known cell identities, covering common methods of filtering, doublet detection, normalization, feature selection, denoising, dimensionality reduction and clustering.
 We evaluate the choice of these tools with multi-purpose metrics to assess their ability to reveal cell population structure and lead to efficient clustering.
-
-\parttitle{Conclusion} On the basis of our systematic evaluations of analysis pipelines, we make a number of practical recommendations about current analysis choices and for a comprehensive pipeline.
+On the basis of our systematic evaluations of analysis pipelines, we make a number of practical recommendations about current analysis choices and for a comprehensive pipeline.
 The evaluation framework that we developed, \textit{pipeComp} (\url{https://github.com/plger/pipeComp}), has been implemented so as to easily integrate any other step, tool, or evaluation metric allowing extensible benchmarks and easy applications to other fields of research in Bioinformatics, as we demonstrate through a study of the impact of removal of unwanted variation on differential expression analysis.
 
-\end{abstract}
+\newpage
+\pagenumbering{arabic}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% The keywords begin here                  %%
-%%                                          %%
-%% Put each keyword in separate \kwd{}.     %%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{keyword}
-\kwd{single-cell RNA sequencing (scRNAseq)}
-\kwd{pipeline}
-\kwd{clustering}
-\kwd{normalization}
-\kwd{dimensionality reduction}
-\kwd{filtering}
-\kwd{benchmark}
-\end{keyword}
-
-% MSC classifications codes, if any
-%\begin{keyword}[class=AMS]
-%\kwd[Primary ]{}
-%\kwd{}
-%\kwd[; secondary ]{}
-%\end{keyword}
-
-\end{abstractbox}
-%
-%\end{fmbox}% uncomment this for twcolumn layout
-
-\end{frontmatter}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% The Main Body begins here                %%
-%%                                          %%
-%% Please refer to the instructions for     %%
-%% authors on:                              %%
-%% http://www.biomedcentral.com/info/authors%%
-%% and include the section headings         %%
-%% accordingly for your article type.       %%
-%%                                          %%
-%% See the Results and Discussion section   %%
-%% for details on how to create sub-sections%%
-%%                                          %%
-%% use \cite{...} to cite references        %%
-%%  \cite{koon} and                         %%
-%%  \cite{oreg,khar,zvai,xjon,schn,pond}    %%
-%%  \nocite{smith,marg,hunn,advi,koha,mouse}%%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%% start of article main body
-% <put your article body there>
-
-%%%%%%%%%%%%%%%%
-%% Background %%
-%%
 \section*{Background}
-
 Single-cell RNA-sequencing (scRNAseq) and the set of attached analysis methods are evolving fast, with more than 560 software tools available to the community \cite{ZappiaDB2018}, roughly half of which are dedicated to tasks related to data processing such as clustering, ordering, dimension reduction or normalization.
@@
 
-A number of good comparison and benchmark studies have already been performed on various steps related to scRNAseq processing and analysis and can guide the choice of methodology (\citealp{CobosDeconvolution2020, colePerformance2018, DalMolinDE2017, duoClustering2018, freytagComparison2018, GaoProcessing2020, HeiserDimred2019, HouImput2020, JaakkolaDE2017, KrzakClust2019, SonesonDE2018, SunDimRed2019, tianMixology2018, TranBatch2020, TsuyuzakiPCA2020, viethSystematic2019, WangDE2019, YipHVGs2018, ZhangImput2018}).
+A number of good comparison and benchmark studies have already been performed on various steps related to scRNAseq processing and analysis and can guide the choice of methodology \cite{CobosDeconvolution2020, colePerformance2018, DalMolinDE2017, duoClustering2018, freytagComparison2018, GaoProcessing2020, HeiserDimred2019, HouImput2020, JaakkolaDE2017, KrzakClust2019, SonesonDE2018, SunDimRed2019, tianMixology2018, TranBatch2020, TsuyuzakiPCA2020, viethSystematic2019, WangDE2019, YipHVGs2018, ZhangImput2018}.
 However these recommendations need constant updating and often leave open many details of an analysis.
 Another missing aspect of current benchmarking studies is their limitation to capture all aspects of a scRNAseq processing workflow.
-Although previous benchmarks already brought valuable recommendations for data processing, some only focused on one aspect of data processing (e.g., \citealp{SunDimRed2019}), did not evaluate how the tool selection affects downstream analysis (e.g., \citealp{TsuyuzakiPCA2020}) or did not tackle all aspects of data processing, such as doublet identification or cell filtering (e.g., \citealp{viethSystematic2019}).
+Although previous benchmarks already brought valuable recommendations for data processing, some only focused on one aspect of data processing (e.g., \cite{SunDimRed2019}), did not evaluate how the tool selection affects downstream analysis (e.g., \cite{TsuyuzakiPCA2020}) or did not tackle all aspects of data processing, such as doublet identification or cell filtering (e.g., \cite{viethSystematic2019}).
 A thorough evaluation of the tools covering all major processing steps is however urgently needed as previous benchmarking studies highlighted that a combination of tools can have a drastic impact on downstream analysis, such as differential expression analysis and cell-type deconvolution\cite{viethSystematic2019,CobosDeconvolution2020}.
@@
 Reasoning that the cells we wish to exclude are the cells that would be misclassified, we measured the rate of misclassification of each cell in each dataset across a variety of clustering pipelines, correcting for the median misclassification rate of the subpopulation and then evaluated what properties could be predictive of misclassification (Supplementary Figures 8-10).
-We could not identify any property or simple combination thereof that would be consistently predictive of misclassification; the only feature that stood out across more than one dataset (the Zheng datasets) was that cells with very high read counts have a higher chance of being misclassified{\color{red}, although such a pattern was not observed in the other datasets.}
+We could not identify any property or simple combination thereof that would be consistently predictive of misclassification; the only feature that consistently stood out across multiple datasets (the Zheng datasets) was that cells with very high read counts have a higher chance of being misclassified.
 
@@
 We next investigated the impact of different normalization strategies.
-Beside the standard log-normalization included in \textit{Seurat}, we tested \textit{scran}'s pooling-based normalization \cite{lunPooling2016}, \textit{sctransform}'s variance-stabilizing transformation \cite{hafemeisterSCtransform2019}, normalization based on stable genes \cite{linStableGenes2018, deekeStablyExpressed2018}, and \textit{SCnorm} \cite{bacherSCnorm2017} (\textit{scVI}-based normalization \cite{lopezDeep2018} was included separately{\color{red}, as it was not meant for this purpose and follows a slightly different flow}).
+Beside the standard log-normalization included in \textit{Seurat}, we tested \textit{scran}'s pooling-based normalization \cite{lunPooling2016}, \textit{sctransform}'s variance-stabilizing transformation \cite{hafemeisterSCtransform2019}, normalization based on stable genes \cite{linStableGenes2018, deekeStablyExpressed2018}, and \textit{SCnorm} \cite{bacherSCnorm2017} (\textit{scVI} \cite{lopezDeep2018} was included separately on a subset of parameter combinations due to high running times).
 In addition to log-normalization, the standard \textit{Seurat} clustering pipeline performs per-feature unit-variance scaling so that the PCA is not too strongly dominated by highly-expressed features.
@@
 Zhengmix8eq and simMix1) and subpopulations were consistently more challenging across all methods.
-{\color{red}An exception was \textit{scVI}-based normalization, which performed similarly to other methods on some datasets (Kumar, simMix2, Zhengmix4eq, Zhengmix4uneq), but very poorly on others, in terms of both the subpopulation silhouette and clustering accuracy (Supplementary Figure 13; of note, these values were not meant to be used for clustering, and the usage of the \textit{scVI} latent space will be discussed below along with dimensionality reductions)}.
+An exception was \textit{scVI}, which performed poorly on most datasets in terms of both the subpopulation silhouette and clustering accuracy (Supplementary Figure 13; \textit{scVI} was run on a reduced set of pipelines).
 Scaling (e.g.
@@
 
-The impact of the dimensionality reduction methods tested was far greater than that of normalization or feature selection (Supplementary Figure 23).
+The impact of the choice of dimensionality reduction method was far greater than that of normalization or feature selection (Supplementary Figure 23).
 Overall, weighting the components by their explained variance tended to increase silhouette width (Supplementary Figure 23A) and, in most datasets, clustering accuracy (Supplementary Figure 23B).
 GLM-PCA tended to increase the average silhouette width of already well-defined subpopulations, but \textit{Seurat}'s PCA procedure however proved superior on all metrics.
-{\color{red} Like GLM-PCA, \textit{scVI}'s Linearly Decoded (LD) data \cite{SvenssonLDVAE2020} and latent space do not explicitly rely on normalized counts.
-Their performance was however worse than \textit{Seurat}'s PCA in view of the silhouette width (Supplementary Figure 15).
-In terms of clustering accuracy, \textit{scVI}'s latent space performed better that its LD but was nevertheless less accurate than PCA-based alternatives on the datasets with a very low amount of cells (Koh and Kumar) and harboring many subpopulations (simMix1 and Zhengmix8eq).}
+Like GLM-PCA, \textit{scVI}'s Linear Decoder (LD) does not explicitly rely on normalized counts.
+Its performance was however the worst both in terms of silhouette width and clustering accuracy (Supplementary Figure 15), and it only performed well in some of the least challenging datasets (mixology10x5cl and simMix2).
 
@@
 We evaluated the performance of dimensionality estimators implemented in the \textit{intrinsicDimension} package \cite{johnssonDimensionality2015}, as well as common procedures such as the `elbow` method, some scRNAseq-specific methods such as the JackStraw procedure \cite{ChungJackstraw2015} or \textit{scran}'s \texttt{denoisePCA} \cite{LunScran2016}, and the recent application of Fisher Separability analysis \cite{AlberganteSepar2019}.
-{\color{red} We also included a variation of the molecular cross-validation (MCVR, modified to be applied on normalized data - see methods) approach originally developed to select an calibrate denoising methods \cite{batsonMolecularCrossValidationSingleCell2019}, but which has also been applied to estimate dimensionality \citep{wagnerMonetOpensource2020}.}
 
@@
 Reasoning that over-estimating the number of dimensions is less problematic than under-estimating it, we kept the former methods for a full analysis of their impact on clustering (Figure \ref{fig:dim}C-D), when combined with \textit{sctransform} or standard \textit{Seurat} normalization.
-
-{\color{red} \textit{MCVR} and the global maximum likelihood based on translated Poisson Mixture Model (\textit{maxLikGlobal}) tended to produce clusterings with smaller deviations to the true number of subpopulations (Supplementary Figure 26).
-Although most methods performed well on the various clustering measures, \textit{maxLikGlobal} (using 10 or 20 nearest neighbors, although estimates were relatively stable across various values of \textit{k} -- see Supplementary Figure 27) provided the dimensionality estimate that best separated challenging subpopulations (Figure \ref{fig:dim}C) and tended to result in the best clustering accuracy (Figure \ref{fig:dim}D).
-\textit{MCVR} did not perform well for the datasets without UMIs (Koh and Kumar), which violates the assumptions of the method, and while it did increase the separability of some subpopulations (Figure \ref{fig:dim}C), it was detrimental to the classification of some others.
-We observed a strong negative correlation (\textit{r}=-0.995, p~1.6e-8) between the number of dimensions called by \textit{MCVR} and the log10 average library size of a dataset, which explained the former considerably better than the number of subpopulations.
-
-Of note, the optimal method in our evaluation (\textit{maxLikGlobal})} systematically selected many more components than were associated with the subpopulations in Figure \ref{fig:dim}A, suggesting that although these additional components appear individually uninformative, in combination they nevertheless contribute to classification.
+Although most methods performed well on the various clustering measures, the global maximum likelihood based on translated Poisson Mixture Model \textit{maxLikGlobal}, using 10 or 20 nearest neighbors, although estimates were relatively stable across various values of \textit{k}; see Supplementary Figure 26) provided the dimensionality estimate that best separated challenging subpopulations (Figure \ref{fig:dim}C) and tended to result in the best clustering accuracy (Figure \ref{fig:dim}D).
+This method systematically selected many more components than were associated with the subpopulations in Figure \ref{fig:dim}A, suggesting that although these additional components appear individually uninformative, in combination they nevertheless contribute to classification.
 An important implication is that the commonly-used elbow method most likely underestimates dimensionality.
@@
 However, because \textit{Seurat}'s resolution parameter had a large impact on the number of clusters identified (Supplementary Figures 2 and 24), \textit{Seurat} could always be coerced into producing the right number of clusters.
-Instead, the number of clusters found by \textit{scran} was considerably less influenced by the available parameters (number of nearest neighbors or steps in the random walk - see Supplementary Figure 28), and as a result \textit{scran}-based clustering sometimes never produced a partitioning with the right number of clusters.
+Instead, the number of clusters found by \textit{scran} was considerably less influenced by the available parameters (number of nearest neighbors or steps in the random walk - see Supplementary Figure 27), and as a result \textit{scran}-based clustering sometimes never produced a partitioning with the right number of clusters.
 This observation, along with \textit{scran}'s higher MI score, suggest that \textit{scran} sometimes simply divides a real subpopulation into two clusters (possibly tracking some unknown biological differences) rather than committing misclassification errors.
@@
 We selected the most important alternatives for each parameter (see Methods) in order to explore their interactions.
-{\color{red}Unsupervised clustering on the evaluation metrics reveals general patterns in the results as well as the relative importance of the parameters (Figure \ref{fig:overall}), with the clustering method having the most important impact on the results.
-This is however chiefly due to number of clusters called: an ANOVA on all analyses after accounting, in the model, for baseline differences between datasets and the impact of \textit{Seurat}'s resolution parameter, attributed relatively little importance to the clustering method, and instead attributed the largest proportion of variance in the main cluster metrics to the number of dimensions used, followed by normalization (Supplementary Figure 29A).
-Because the choice of dimension reduction impacts the number of clusters, we repeated the analysis of variance in ARI among the combinations of parameters that yielded the true number of clusters (Supplementary Figure 29B).
-We observed that although normalization was the most important parameter among these analyses, the choice of dimensionality (as well as feature selection) came closely behind (Supplementary Figure 29B), suggesting that more attention should be paid to this relatively neglected step.
-
-The greater importance of normalization, dimensionality and feature selection over filtering steps is also visible in the clustering (Figure \ref{fig:overall}).
-The top methods (bottom of Figure \ref{fig:overall}) confirm the results of the previous sections, highlighting especially \textit{sctransform}, deviance-based feature selection, and \textit{maxLikGlobal} dimensionality estimates as top methods.
-Interestingly, while the combination of \textit{sctransform} with \textit{Seurat} clustering was particularly performant for most datasets, for the simulated `simMix1` dataset,  standard \textit{Seurat} normalization combined with \textit{scran}-based clustering proved superior.}
+After accounting, in the model, for baseline differences between datasets and the impact of the resolution parameter, ANOVA attributed the largest proportion of variance in the main cluster metrics to the choice of dimensionality reduction, followed by normalization, and very little to filtering (Supplementary Figure 28A).
+Because the choice of dimension reduction impacts the number of clusters, we repeated the analysis of variance in ARI among the combinations of parameters that yielded the true number of clusters (Supplementary Figure 28B).
+We observed that although normalization was the most important parameter among these analyses, the choice of dimensionality came closely behind (Supplementary Figure 28B), suggesting that more attention should be paid to this relatively neglected step.
 
 We next investigated whether the benefits of some methods were conditional on the choices at other steps.
-Although analysis of variance identified many weak but significant interactions, we found that the benefits brought by performant methods were robust to differences in other parts of the pipeline (Supplementary Figures 30-31).
-For instance, we confirmed that doublet removal and \textit{sctransform} normalization were beneficial independently of other steps, and that, again independently of other steps, more stringent filtering did not yield any major improvement over lenient filtering (Supplementary Figure 30).
-Similarly, deviance-based feature selection outcompeted \textit{Seurat}'s default selection method independently of changes at other steps (Supplementary Figure 31).
-{\color{red} ANOVA did reveal some significant (negative) interactions: the top two-way interaction was between \textit{scran} normalization and the elbow method for selection the number of dimensions, and indeed that combination was involved in all of the 30 lowest-ranking pipelines.
-Furthermore, all significant interactions involved the elbow method, suggesting that some methods were less robust to underestimated dimensionality.}
+Although analysis of variance identified many weak but significant interactions, we found that the benefits brought by performant methods were robust to differences in other parts of the pipeline (Supplementary Figures 29-30).
+For instance, we confirmed that doublet removal and \textit{sctransform} normalization were beneficial independently of other steps, and that, again independendently of other steps, more stringent filtering did not yield any major improvement over lenient filtering (Supplementary Figure 29).
+Similarly, deviance-based feature selection outcompeted \textit{Seurat}'s default selection method independently of changes at other steps (Supplementary Figure 30).
 
@@
 Since preliminary analysis showed that all methods performed equally well or better on normalized data, we applied them after filtering and normalization, but before scaling and reduction.
-Although some of the methods (e.g., \textit{DRImpute\_process} and \textit{alra\_norm}) did improve the separability of some more elusive subpopulations, no method had a systematically positive impact on the average silhouette width across all subpopulations (Supplementary Figure 32A).
-When restricting ourselves to clustering analyses that yielded the `right' number of clusters, all tested methods improved classification  compared to a scenario with no imputation step (`none' label, Supplementary Figure 32B).
+Although some of the methods (e.g., \textit{DRImpute\_process} and \textit{alra\_norm}) did improve the separability of some more elusive subpopulations, no method had a systematically positive impact on the average silhouette width across all subpopulations (Supplementary Figure 31A).
+When restricting ourselves to clustering analyses that yielded the `right' number of clusters, all tested methods improved classification  compared to a scenario with no imputation step (`none' label, Supplementary Figure 31B).
 However, the situation was not so straightforward with alternative metrics, where some methods (e.g., \textit{ENHANCE}) consistently underperformed.
@@
 surrogate variables) and their inclusion in the DEA's generalized linear model might improve DEA, especially in the presence of sources of variation unrelated to the experimental variables.
-\textit{sva} \citep{leekCapturingHeterogeneityGene2007,leekSvaPackageRemoving2012} and \textit{RUVseq} \citep{rissoNormalizationRNAseq2014a} are two popular packages each implementing different methods to infer these variables.
+\textit{sva} \cite{leekCapturingHeterogeneityGene2007,leekSvaPackageRemoving2012} and \textit{RUVseq} \cite{rissoNormalizationRNAseq2014a} are two popular packages each implementing different methods to infer these variables.
 We therefore created a \texttt{PipelineDefinition} with three steps: 1) feature filtering, 2) removal of unwanted variation (i.e.
-SVA), and 3) DEA using the most established methods  \citep{germainRNAontheBENCHComputationalEmpirical2016}.
+SVA), and 3) DEA using the most established methods  \cite{germainRNAontheBENCHComputationalEmpirical2016}.
 The package's `pipeComp\_dea' vignette details the creation process.
-We developed 3 benchmark datasets: 1) the \texttt{ipsc} data is a random selection of 10 vs 10 samples from \citep{carcamoOriveAnalysisTranscriptional2017}, which are very heterogeneous, and foldchanges were manually added to 300 genes; 2) the \texttt{seqc} dataset contains 5 vs 5 samples of two mixtures of the SEQC consortium with different ERCC spike-in mixes \citep{ComprehensiveAssessment2014}, and includes a partially correlated batch effect; 3) A \texttt{simulation} of 8 vs 8 samples with two technical vectors of variation (see Methods for more details).
-As previously reported \citep{ComprehensiveAssessment2014}, we found that more stringent filtering improved DEA (Supplementary Figure 33A) and that the different DEA methods had similar performances (Figure \ref{fig:svadea}A and Supplementary Figure 33A).
-Including a SVA-based step increased sensitivity across all datasets, particularly in datasets with technical vectors of variation (\textit{seqc} and \textit{simulation}), and benefited all DEA methods in a similar fashion (Figure \ref{fig:svadea}A,C and Supplementary Figures 33-34).
+We developed 3 benchmark datasets: 1) the \texttt{ipsc} data is a random selection of 10 vs 10 samples from \cite{carcamoOriveAnalysisTranscriptional2017}, which are very heterogeneous, and foldchanges were manually added to 300 genes; 2) the \texttt{seqc} dataset contains 5 vs 5 samples of two mixtures of the SEQC consortium with different ERCC spike-in mixes \cite{ComprehensiveAssessment2014}, and includes a partially correlated batch effect; 3) A \texttt{simulation} of 8 vs 8 samples with two technical vectors of variation (see Methods for more details).
+As previously reported \cite{ComprehensiveAssessment2014}, we found that more stringent filtering improved DEA (Supplementary Figure 32A) and that the different DEA methods had similar performances (Figure \ref{fig:svadea}A and Supplementary Figure 32A).
+Including a SVA-based step increased sensitivity across all datasets, particularly in datasets with technical vectors of variation (\textit{seqc} and \textit{simulation}), and benefited all DEA methods in a similar fashion (Figure \ref{fig:svadea}A,C and Supplementary Figures 32-33).
 With the exception of the \texttt{RUVr} method, which led to a high False Discovery Rate (FDR), all other methods showed comparable gains in sensitivity while maintaining a decent error control (if a little above the nominal FDR), even when increasing the number of surrogate variables used (Figure \ref{fig:svadea}B-C).
-While the FDR did increase in the \textit{ipsc} dataset, we suspect that much of the effect could be due to spurious differences \citep{germainTamingHumanGenetic2017} between the groups which we wrongly assume to be false positives.
+While the FDR did increase in the \textit{ipsc} dataset, we suspect that much of the effect could be due to spurious differences \cite{germainTamingHumanGenetic2017} between the groups which we wrongly assume to be false positives.
 Nevertheless, the FDR remained acceptable with up to three variables.
@@
 
-On the basis of our findings, we can make a number of practical analysis recommendations, also summarized in Supplementary Figure 35:
+On the basis of our findings, we can make a number of practical analysis recommendations, also summarized in Supplementary Figure 34:
 
@@
 
-{\color{red} The performance of \textit{scVI} in our evaluation was lower than in the original publication presenting the method \cite{lopezDeep2018}.
-Our results are however in line with other studies comparing silhouette score \cite{Aparicio2020} and clustering accuracy \cite{HouImput2020} of \textit{scVI}'s latent space.
-These studies, like ours, used datasets with relatively few cells (i.e.
-fewer than genes), for which \textit{scVI} was reported by its authors to be less adapted.
-We think that further work will be required to assess its performance on larger datasets.}
-
 Here, we chose to concentrate on what could be learned from datasets with known cell labels (as opposed to labels inferred from the data, as in \cite{MereuCellAtlas2019}).
@@
 This avoids the need to store potentially large intermediate data for all possible permutations and thus allowing a combinatorial benchmark.
-This, along with the set-up simplicity (demonstrated here by our separate SVA-DEA benchmark), also distinguishes \textit{pipeComp} from general workflow frameworks such as \textit{drake} \citep{landauDrakePackage2018}; indeed we estimate that the benchmarks discussed here would have taken approximately 16TB of storage had all non-redundant intermediates been saved (see the Supplementary methods for more detailed explanations of the framework and comparison to \textit{CellBench} and \textit{drake}).
+This, along with the set-up simplicity (demonstrated here by our separate SVA-DEA benchmark), also distinguishes \textit{pipeComp} from general workflow frameworks such as \textit{drake} \cite{landauDrakePackage2018}; indeed we estimate that the benchmarks discussed here would have taken approximately 16TB of storage had all non-redundant intermediates been saved (see the Supplementary methods for more detailed explanations of the framework and comparison to \textit{CellBench} and \textit{drake}).
 In addition, the fact that benchmark functions are stored in the \texttt{PipelineDefinition} makes the benchmark more smoothly portable, making it easy to modify, extend with further methods, or apply to other datasets.
@@
 \textit{pipeComp} is a flexible R framework for evaluating methods and parameter combinations in a complex pipeline, computing on-the-fly multilevel evaluation metrics.
-Applying this framework to scRNAseq clustering enabled us to make
-clear recommendations on the steps of filtering, normalization, feature selection, denoising, dimensionality reduction and clustering (Supplementary Figure 35).
+Applying this framework to scRNAseq clustering enabled us to make clear recommendations on the steps of filtering, normalization, feature selection, denoising, dimensionality reduction and clustering (Supplementary Figure 34).
 We demonstrate how a diversity of multilevel metrics can be more robust, more sensitive, and more nuanced than simply evaluating the final clustering performance.
@@
 All code to reproduce the simulations and figures is available in the \url{https://github.com/markrobinsonuzh/scRNA\_pipelines\_paper} repository, which also includes the basic datasets with a standardized annotation.
-{\color{red}All wrappers for methods used in the study can be found in the external data of the \textit{pipeComp} package (\url{https://github.com/plger/pipeComp/tree/master/inst/extdata}).}
 
@@
 
-{\color{red} For the molecular cross-validation approach, we relied on a R implementation by Matthew Young (\url{https://github.com/constantAmateur/MCVR}), with minor modifications (see wrapper in the \textit{pipeComp} package).
-Briefly, since there is no \textit{a priori} relationship between the dimensionality of the count data and that of the PCA based on the normalized data, MCVR does not use the original approach based Poisson loss, but the mean squared error of the normalized data, and uses balanced 50-50 splits.
-Because of the 50\% split, dimensionality may be underestimated, and titration did suggest that was the case for some of our datasets.}
-
 \subsection*{Filter sets}
@@
 
-The evaluation of \textit{scVI} algorithm was performed separate to the other pipeline combinations since the model training on our reference datasets was especially resource-consuming {\color{red} and follows a different workflow from the other pipelines.
-In particular, for datasets with fewer cells than genes, \textit{scVI} requires a feature selection to be performed prior to the normalization step; furthermore, its dimensional reduction steps rely directly on raw counts.
-We therefore evaluated it separately and compared it to a representative subset of pipelines using \textit{sctransform} normalization and \textit{Seurat}'s PCA dimension reduction method.
+The evaluation of \textit{scVI} algorithm was performed separate to the other pipeline combinations as the model training on our reference datasets was especially resource-consuming.
+As it was not practically feasible to perform the training step on many combination of parameters as it was done with other tools, we evaluated it on a representative subset of pipelines using \textit{sctransform} normalization and \textit{Seurat}'s PCA dimension reduction method.
 The resulting evaluation is available in the Supplementary Figure 13.
-
-Using \textit{scVI}, we considered as normalized values the normalized mean of the negative binomial distribution modeling the counts (accessed through the \textit{get\_sample\_scale} method).
-Following \textit{scVI}'s online documentation, we used the \textit{VAE} model, using a train size of 1, a training on 400 epochs and a learning rate of 0.001.
-As a major limitation of \textit{scVI} is the training on dataset with low gene to cell ratio, the features were subsampled to 2000 using the \textit{subsample\_genes} function, which resembles the \textit{Seurat} V3 VST gene selection method.
-All other parameters were left unmodified.
-.
-
-Two models of the \textit{scVI} package computing a dimension reduction on the data were tested.
-First, we evaluated the mean of the posterior distribution for the latent space, still based on the \textit{VAE} model (accessed through the \textit{get\_latent} function).
-All other parameters were set as described above or left as default.
-The second model that we evaluated was the linearly decoded VAE of \textit{scVI} (\textit{LDVAE} model).
-It was trained on 250 epochs, a learning rate of 0.001 and computed 50 dimensions.
-For some smaller datasets, the training phase aborted.
-In such cases, we followed the recommendations of the package and reduced the learning rate by a factor 10 until the model could properly converge.
-Given that the other methods that we evaluated returned PCs ordered by their variance explained, we applied to the same reordering to the latent space returned by \textit{scVI}.
-For both models.
-we relied on the feature selections results from the pipeline and not on the \textit{subsample\_genes} function.
-All other parameters were left unmodified.
-}
 
@@
 
-The set of alternatives found for the overall run (Supplementary Figures 29-31) can be found in the \href{https://github.com/markrobinsonuzh/scRNA_pipelines_paper/blob/master/data/overall_run.Rmd}{paper's repository}.
+The set of alternatives found for the overall run (Supplementary Figures 28-30) can be found in the \href{https://github.com/markrobinsonuzh/scRNA_pipelines_paper/blob/master/data/overall_run.Rmd}{paper's repository}.
 The analysis of variance was performed with the following model:
@@
 \begin{itemize}
-    \item For the \textit{ipsc} dataset, we took the quantification of the GSE79636 dataset (fairly heterogeneous) from the \textit{iPSCpower} package \citep{germainTamingHumanGenetic2017}, and selected two random sets of 10 samples.
+    \item For the \textit{ipsc} dataset, we took the quantification of the GSE79636 dataset (fairly heterogeneous) from the \textit{iPSCpower} package \cite{germainTamingHumanGenetic2017}, and selected two random sets of 10 samples.
 No further variation was added, and log-foldchanges ranging from 0.25 to 1.25 were added to the counts of 300 random genes.
 All other genes were assumed to be negatives.
-    \item For the \textit{seqc} dataset, we used data from the \textit{seqc} package \citep{ComprehensiveAssessment2014}.
+    \item For the \textit{seqc} dataset, we used data from the \textit{seqc} package \cite{ComprehensiveAssessment2014}.
 For both groups (mixtures C and D, which also contain two different spike-in mixes), we selected samples from the \texttt{ILM\_refseq\_gene\_AGR} and \texttt{ILM\_refseq\_gene\_CNL} batches, which have clear technical differences; we selected 5 samples per group with a weak correlation with the batch (3:2 vs 2:3).
 Since the true differences between mixtures are not entirely known, the analysis was performed on all genes but only the spike-ins were considered for benchmarking.
-    \item For the \textit{simulation}, we simulated counts using the \textit{polyester} package \citep{frazeePolyesterSimulating2015} and the means/dispersions from the GSE79636 dataset (restricted to a single batch and biological group), with 500 DEGs (log-foldchanges ranging from 0.5 to 2) and two batch effects: 1) a technical batch partially correlated with the group (6:2) and affecting 1/3 of the genes, and 2) a linear vector of variation uncorrelated with the groups and affecting 1/3 of the genes.
+    \item For the \textit{simulation}, we simulated counts using the \textit{polyester} package \cite{frazeePolyesterSimulating2015} and the means/dispersions from the GSE79636 dataset (restricted to a single batch and biological group), with 500 DEGs (log-foldchanges ranging from 0.5 to 2) and two batch effects: 1) a technical batch partially correlated with the group (6:2) and affecting 1/3 of the genes, and 2) a linear vector of variation uncorrelated with the groups and affecting 1/3 of the genes.
 \end{itemize}
@@
 Filtering was performed with edgeR's \texttt{filterByExpr} function, changing the \texttt{min.count} argument.
-For \textit{vstsva}, we first applied the variance-stabilizing transformation from \textit{DESeq2} \citep{loveModeratedEstimation2014} before applying \texttt{sva::sva}.
-\textit{edgeR.QLF} refers to the quasi-likelihood version \citep{lundDetectingDifferential2012} of \textit{edgeR}'s fit.
+For \textit{vstsva}, we first applied the variance-stabilizing transformation from \textit{DESeq2}\cite{loveModeratedEstimation2014} before applying \texttt{sva::sva}.
+\textit{edgeR.QLF} refers to the quasi-likelihood version \cite{lundDetectingDifferential2012} of \textit{edgeR}'s fit.
 The exact wrappers used can be found in the package at \url{https://github.com/plger/pipeComp/blob/master/inst/extdata/dea\_wrappers.R}.
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                  The Bibliography                       %%
-%%                                                         %%
-%%  Bmc_mathpys.bst  will be used to                       %%
-%%  create a .BBL file for submission.                     %%
-%%  After submission of the .TEX file,                     %%
-%%  you will be prompted to submit your .BBL file.         %%
-%%                                                         %%
-%%                                                         %%
-%%  Note that the displayed Bibliography will not          %%
-%%  necessarily be rendered by Latex exactly as specified  %%
-%%  in the online Instructions for Authors.                %%
-%%                                                         %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section*{Tables}
 
-% if your bibliography is in bibtex format, use those commands:
-\bibliographystyle{bmc-mathphys} % Style BST file (bmc-mathphys, vancouver, spbasic).
-\bibliography{bmc_article}      % Bibliography file (usually '*.bib' )
-% for author-year bibliography (bmc-mathphys or spbasic)
-% a) write to bib file (bmc-mathphys only)
-% @settings{label, options="nameyear"}
-% b) uncomment next line
-%\nocite{label}
+\begin{table}[H]
+\caption{\textbf{Overview of the benchmark datasets.} `\# features' indicates the number of features detected in at least 10 cells.}
+\label{tab:table1}
+\begin{tabular}{rlrrl}
+  \hline
+dataset & \# features & \# cells & protocol & description \\ 
+  \hline
+Koh & 33922 & 531 & SMARTer & 9 FACS purified differentiation stages \\ 
+  Kumar & 41930 & 246 & SMARTer & Mouse ESC cultured in 3 conditions \\ 
+  Zhengmix4eq & 10434 & 3994 & 10x & Mixtures of FACS purified PBMCs \\ 
+  Zhengmix4uneq & 11369 & 6498 & 10x & Mixtures of FACS purified PBMCs \\ 
+  Zhengmix8eq & 10600 & 3994 & 10x & Mixtures of FACS purified PBMCs \\ 
+  mixology10x3cl & 16208 & 902 & 10x & Mixture of 3  cancer lines from CellBench \\ 
+  mixology10x5cl & 11786 & 3918 & 10x & Mixture of 5 cancer lines from CellBench \\ 
+  simMix1 & 3696 & 2500 & 10x-based & Simulation of 10 human subpopulations \\
+  simMix2 & 8893 & 3000 & 10x-based & Simulation of 9 mouse subpopulations \\
+   \hline
+\end{tabular}
+\end{table}
 
-% or include bibliography directly:
-% \begin{thebibliography}
-% \bibitem{b1}
-% \end{thebibliography}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                               %%
-%% Figures                       %%
-%%                               %%
-%% NB: this is for captions and  %%
-%% Titles. All graphics must be  %%
-%% submitted separately and NOT  %%
-%% included in the Tex document  %%
-%%                               %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%
-%% Do not use \listoffigures as most will included as separate files
+\vspace{20mm}
 
@@
     \centering
-    \includegraphics[width=\textwidth, keepaspectratio]{pipeline_explanation.pdf}
+    \includegraphics[width=\textwidth, keepaspectratio]{main_figures/pipeline_explanation.pdf}
     \caption{\textbf{Overview of the \textit{pipeComp} framework and its application to a scRNAseq clustering pipeline.
@@
     \centering
-    \includegraphics[width=\textwidth, keepaspectratio]{{doublets.pdf}}
+    \includegraphics[width=\textwidth, keepaspectratio]{{main_figures/doublets.pdf}}
     \caption{\textbf{Identification of doublet cells.
@@
     \centering
-    \includegraphics[width=\textwidth, keepaspectratio]{{filtering.pdf}}
+    \includegraphics[width=\textwidth, keepaspectratio]{{main_figures/filtering.pdf}}
     \caption{\textbf{Effect of filtering on cell subpopulation structure and clustering.
@@
     \centering
-    \includegraphics[width=\textwidth, keepaspectratio]{{normalization.pdf}}
+    \includegraphics[width=\textwidth, keepaspectratio]{{main_figures/normalization.pdf}}
     \caption{\textbf{Evaluation of normalization procedures.
@@
 \textbf{C:} Clustering accuracy (\textit{Seurat} clustering), measured by mutual information (MI), Adjusted Rand Index (ARI) and ARI at the true number of cluster.
-{\color{red}Grey squares indicate that the true number of clusters could not be reached with the corresponding method, dataset and parameter set.} `feat\_regress', `mt\_regress' and `feat\_mt\_regress' respectively stand for the regressing out of the number of features, the proportion of mitochondrial reads, or both during scaling.
-{\color{red}Throughout this paper, silhouette plots (panel \textbf{B}) square-root transform values for color mapping.
+Grey squares indicate that the true number of clusters could not be reached with the corresponding method, dataset and parameter set.
+`feat\_regress', `mt\_regress' and `feat\_mt\_regress' respectively stand for the regressing out of the number of features, the proportion of mitochondrial reads, or both during scaling.
+Throughout this paper, silhouette plots (panel \textbf{B}) square-root transform values for color mapping.
 Other evaluation metric heatmaps (e.g.
@@
 The numbers printed in the cells represent the raw (i.e.
-unscaled) metric values, and are printed in white or black depending on whether they are above or below the median.}}
+unscaled) metric values, and are printed in white or black depending on whether they are above or below the median.}
     \label{fig:norm}
@@
     \centering
-    \includegraphics[width=\textwidth]{{selection.pdf}}
+    \includegraphics[width=\textwidth]{{main_figures/selection.pdf}}
     \caption{\textbf{Evaluation of feature selection methods.
@@
     \centering
-    \includegraphics[width=\textwidth, keepaspectratio]{{dimensionality.pdf}}
+    \includegraphics[width=\textwidth, keepaspectratio]{{main_figures/dimensionality.pdf}}
     \caption{\textbf{Estimating dimensionality.
@@
 \textbf{D:} Clustering accuracy across the normalization/dimensionality estimation methods.
-{\color{red}The color-mapping scheme is the same as in Figure \ref{fig:norm}C}}
+The color-mapping scheme is the same as in Figure \ref{fig:norm}C}
     \label{fig:dim}
@@
     \centering
-    \includegraphics[width=\textwidth, keepaspectratio]{{clustering.pdf}}
+    \includegraphics[width=\textwidth, keepaspectratio]{{main_figures/clustering.pdf}}
     \caption{\textbf{Evaluation of clustering methods.} Clustering accuracy (\textbf{A}) and run time (\textbf{B}) of \textit{scran}- and \textit{Seurat}-based clustering tools in combination with \textit{sctransform} and standard \textit{Seurat}'s normalization.
-{\color{red} The color-mapping scheme in panel \textbf{A} and the left of panel \textbf{B} is the same as in Figure \ref{fig:norm}C.}}
+The color-mapping scheme in panel \textbf{A} and the left of panel \textbf{B} is the same as in Figure \ref{fig:norm}C.}
     \label{fig:clust}
@@
     \centering
-    \includegraphics[width=\textwidth, keepaspectratio]{{overall.pdf}}
-    \caption{\textbf{Overview of the combination of main alternative parameters.} Unsupervised clustering of the results (above) and top results (below).
-The color-mapping schemes are the same as described in the previous figures.}
-    \label{fig:overall}
-\end{figure}
-
-\begin{figure}
-    \centering
-    \includegraphics[width=\textwidth, keepaspectratio]{{svadea.pdf}}
-    \caption{{\color{red}\textbf{Evaluation of the impact of surrogate variable analysis (SVA) on (bulk) RNAseq differential expression analysis using \textit{pipeComp}.} \textit{RUVr} and \textit{RUVs} refer to the two \textit{RUVSeq} functions by the same name; \textit{svaseq} refers to the function of the same name in \textit{sva}, while \textit{vstsva} refers to the original \textit{sva} method applied no variance-stabilized data.
+    \includegraphics[width=\textwidth, keepaspectratio]{{main_figures/svadea.pdf}}
+    \caption{\textbf{Evaluation of the impact of surrogate variable analysis (SVA) on (bulk) RNAseq differential expression analysis using \textit{pipeComp}.} \textit{RUVr} and \textit{RUVs} refer to the two \textit{RUVSeq} functions by the same name; \textit{svaseq} refers to the function of the same name in \textit{sva}, while \textit{vstsva} refers to the original \textit{sva} method applied no variance-stabilized data.
 \textbf{A:} True positive rates (TPR) and False discovery rates (FDR) of various combinations of SVA-based methods (using a single surrogate variable) and DEA methods.
@@
 \textbf{C:} Correlation of estimated logFC with expected ones, as well as TPR and FDR (at nominal FDR threshold of 0.05) of the different methods with increasing numbers of variables.
-As in \ref{fig:norm}C, the color mapping tracks the number of (matrix-wise) Median Absolute Deviations from the (column-wise) median, while the printed numbers represent the raw metric values.}}
+As in \ref{fig:norm}C, the color mapping tracks the number of (matrix-wise) Median Absolute Deviations from the (column-wise) median, while the printed numbers represent the raw metric values.}
     \label{fig:svadea}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                               %%
-%% Tables                        %%
-%%                               %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\end{backmatter}
 
-%% Use of \listoftables is discouraged.
-%%
-\section*{Tables}
+\clearpage
+\singlespacing
+\nolinenumbers
+\printbibliography
 
-\begin{table}[h!]
-\caption{\textbf{Overview of the benchmark datasets.} {\color{red}`\# features' indicates the number of features detected in at least 10 cells.}}
-\label{tab:table1}
-\begin{tabular}{rlrrl}
-  \hline
-dataset & {\color{red}\# features} & \# cells & protocol & description \\ 
-  \hline
-Koh & 33922 & 531 & SMARTer & 9 FACS purified differentiation stages \\ 
-  Kumar & 41930 & 246 & SMARTer & Mouse ESC cultured in 3 different conditions \\ 
-  Zhengmix4eq & 10434 & 3994 & 10x & Mixtures of FACS purified PBMCs \\ 
-  Zhengmix4uneq & 11369 & 6498 & 10x & Mixtures of FACS purified PBMCs \\ 
-  Zhengmix8eq & 10600 & 3994 & 10x & Mixtures of FACS purified PBMCs \\ 
-  mixology10x3cl & 16208 & 902 & 10x & Mixture of 3  cancer cell lines from CellBench \\ 
-  mixology10x5cl & 11786 & 3918 & 10x & Mixture of 5 cancer cell lines from CellBench \\ 
-  simMix1 & 3696 & 2500 & 10x-based & Simulation of 10 human cell subpopulations \\
-  simMix2 & 8893 & 3000 & 10x-based & Simulation of 9 mouse cell subpopulations \\
-   \hline
-\end{tabular}
-\end{table}
+\newpage
+\pagenumbering{arabic}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                               %%
-%% Additional Files              %%
-%%                               %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% \includepdf[pages=-]{supplementary_figures/supp_figures.pdf}
 
-\section*{Additional Files}
-  \subsection*{\textbf{Additional file 1: Supplementary figures}}
-  Additional results of evaluations across datasets among various metrics. (PDF 16 MB)
-  \subsection*{\textcolor{red}{\textbf{Additional file 2: Supplementary methods}}}
-  \textcolor{red}{More detailed description of \textit{pipeComp} and comparison with other tools. (PDF 0.5 MB)}
-\end{backmatter}
 \end{document}

### 2020-08-04 main_figures/anova.Rmd
added new anova figure

--- 2020-06-29 supplementary_figures/supp_overall.Rmd
@@
 ---
-title: "Supplementary Figures relative to feature selection"
+title: "Overall"
 author: "Pierre-Luc Germain"
@@
 
-```{r, include=FALSE}
-if(!exists('FIG_NB')){
-  FIG_NB <- 0
-  getFigNb <- function(increment=FALSE){
-    if(increment) FIG_NB <<- FIG_NB + 1
-    FIG_NB
-  }
-}
-
+```{r}
 suppressPackageStartupMessages({
@@
 
-# Supplementary Figure `r getFigNb(TRUE)`
-
-```{r, fig.width=6, fig.height=5}
+```{r, fig.width=4, fig.height=4}
 ov <- readRDS("../data/overall.rds")
@@
 cl$sel <- relevel(cl$sel, "sel.vst")
-cl$clustmethod <- relevel(cl$clustmethod, "clust.seurat")
 cl$clustmethod <- relevel(cl$clustmethod, "clust.seurat")
@@
 aov2 <- dplyr::bind_rows(lapply(cos2, FUN=function(x) x$aov), .id="metric")
-p1 <- ggplot(aov, aes(reorder(parameter, -varExpl), varExpl, group=parameter, fill=metric)) + 
-  geom_col() + xlab("") + ylab("Proportion of variance explained") + coord_flip()
-p2 <- ggplot(aov2, aes(reorder(parameter, -varExpl), varExpl, group=parameter, fill=metric)) + 
-  geom_col() + xlab("") + ylab("Variance explained") + theme(legend.position="none") + 
-  coord_flip() + ggtitle("ARI at true k")
-
-co <- dplyr::bind_rows(lapply(cos, FUN=function(x) x$coef), .id="metric")
-co2 <- dplyr::bind_rows(lapply(cos2, FUN=function(x) x$coef), .id="metric")
-for(f in c("norm","filt","doubletmethod","dims","sel","clustmethod")){
-  co$coef <- gsub(paste0("^",f),"",co$coef)
-  co2$coef <- gsub(paste0("^",f),"",co2$coef)
-}
-p3 <- ggplot(co, aes(coef, Estimate, group=coef, fill=metric)) + geom_col() + xlab("") +
-  ylab("Estimated coefficient") + geom_hline(yintercept = 0, linetype="dashed") + 
-  coord_flip() + theme(legend.position="none")
-p4 <- ggplot(co2, aes(coef, Estimate, group=coef, fill=metric)) + geom_col() + xlab("") +
-  ylab("Estimated coefficient") + geom_hline(yintercept = 0, linetype="dashed") + 
-  coord_flip() + theme(legend.position="none")
-
-# plot_grid( plot_grid(p1,p2,rel_widths=c(4,3),labels=LETTERS[1:2]),
-#            plot_grid(p3,p4,rel_widths=c(4,3),labels=LETTERS[3:4]),
-#            nrow=2, rel_heights=c(3,4))
-plot_grid(p1 + ylab("Proportion of variance in metric explained by the parameter"),p2,nrow=2,labels=LETTERS[1:2])
+aov2 <- aov2[aov2$metric=="ARI",]
+aov2$metric <- "ARI at true k"
+p <- ggplot(rbind(aov,aov2), aes(reorder(parameter, -varExpl), varExpl, fill=metric, colour=metric)) + 
+  geom_col(position="dodge") + xlab("") + ylab("Proportion of variance explained") + 
+  coord_flip() + theme(legend.position=c(.62,.85))
+pdf("anova.pdf", width=4, height=4)
+p
+dev.off()
+p
 ```
-
-### Supplementary Figure `r getFigNb()`
-
-**A:** Proportion of the variance in clustering accuracy metrics explained by main parameters.
-**B:** Proportion of the variance in Adjusted Rand Index (ARI) explained by the parameters, among the analyses giving the true number of clusters.
-The ANOVA were performed with the model
-
-`metric ~ dataset*resolution+doubletmethod+filt+sel+norm+clustmethod+dims`
-
-In both panels, the proportion is calculated after substracting the variance attributable to differences due to the dataset, the resolution and their interaction.
-The alternatives included in the analysis (other than the default ones described in the methods) were:
-```
-list( doubletmethod=c("none","doublet.scDblFinder"),
-      filt=c("none","filt.default","filt.stringent"),
-      norm=c("norm.seurat","norm.sctransform","norm.scran"),
-      sel=c("sel.vst","sel.deviance","sel.expr"),
-      selnb=c(2000,4000),
-      dr=c("seurat.pca"),
-      dims=c("elbow", 10, "maxLikGlobal"),
-      clustmethod=c("clust.seurat","clust.scran.knnAnnoy") )
-```
-
-\newpage
-
-# Supplementary Figure `r getFigNb(TRUE)`
-
-```{r}
-rnfn <- function(x){
-  x <- gsub("norm\\.","",x)
-  x <- gsub("^doublet\\.","",x)
-  x <- gsub("> none > ","> no filter > ",x)
-  gsub("^none > ","",x)
-}
-
-H1 <- renameHrows( scrna_evalPlot_silh(ov, agg.by = c("doubletmethod","filt","norm"), row_split = rnfn(norm),
-                                       heatmap_legend_param=list(direction="horizontal"),font_factor = 0.8,
-                                       row_names_max_width=unit(7, "cm")), rnfn)
-H2 <- renameHrows(
-     evalHeatmap(ov, step="clustering", what=c("MI","min_pr"), anno_legend=FALSE, title=c("MI","Min precision"), 
-                 agg.by=c("doubletmethod","filt","norm"), font_factor = 0.8, 
-                 row_split=rnfn(norm),row_names_max_width=unit(7, "cm")) + 
-    evalHeatmap(ov, step="clustering", what=c("ARI"), agg.by=c("doubletmethod","filt","norm"), 
-                row_split=rnfn(norm),filter=n_clus==true.nbClusts, title="ARI at true k",
-                anno_legend=FALSE, font_factor = 0.8, row_names_max_width=unit(7, "cm")),
-  rnfn)
-
-H3 <- renameHrows( scrna_evalPlot_silh(ov, agg.by = c("doubletmethod","sel","norm"), row_split = rnfn(norm),
-                                       heatmap_legend_param=list(direction="horizontal"),
-                                       font_factor = 0.8, 
-                                       row_names_max_width=unit(7, "cm")), rnfn)
-H4 <- renameHrows(
-     evalHeatmap(ov, step="clustering", what=c("MI","min_pr"), anno_legend=FALSE, title=c("MI","Min precision"),
-                 agg.by=c("doubletmethod","sel","norm"), row_split=rnfn(norm), 
-                 font_factor = 0.8, row_names_max_width=unit(7, "cm")) + 
-    evalHeatmap(ov, step="clustering", what=c("ARI"), agg.by=c("doubletmethod","sel","norm"), 
-                row_split=rnfn(norm),filter=n_clus==true.nbClusts, title="ARI at true k", 
-                anno_legend=FALSE, font_factor = 0.8, row_names_max_width=unit(7, "cm")),
-  rnfn)
-```
-
-```{r, fig.width=9, fig.height=8.5}
-plot_grid( grid.grabExpr(draw(H1, heatmap_legend_side="bottom")), 
-           grid.grabExpr(draw(H2)), nrow=2, labels=LETTERS[1:2])
-```
-
-### Supplementary Figure `r getFigNb()`
-
-Silhouette widths of the real subpopulations (**A**) and clustering accuracy (**B**) using different combinations of methods.
-
-\newpage
-
-# Supplementary Figure `r getFigNb(TRUE)`
-
-```{r, fig.width=9, fig.height=8.5}
-plot_grid( grid.grabExpr(draw(H3, heatmap_legend_side="bottom")), 
-           grid.grabExpr(draw(H4)), nrow=2, labels=LETTERS[1:2])
-```
-
-### Supplementary Figure `r getFigNb()`
-
-Silhouette widths of the real subpopulations (**A**) and clustering accuracy (**B**) using different combinations of methods.

### 2020-08-04 bmc_article.tex
Updates from Overleaf

--- 2020-07-03 bmc_article.tex
@@
 
-\parttitle{Background} Despite the massive growth of single-cell RNA-sequencing (scRNAseq) and corresponding computational methods, sufficient and up-to-date benchmarks to guide analytical choices are still lacking.
-Numerous benchmark studies already exist and cover most of scRNAseq processing and analytical methods but only a few give advice on a comprehensive pipeline.
-Moreover, current studies often focused on isolated steps of the process and do not address the impact of a tool on both the intermediate and the final steps of the analysis.
-
-\parttitle{Results} Here, we present a flexible R framework for pipeline comparison with multi-level evaluation metrics.
-We apply it to the benchmark of scRNAseq analysis pipelines using simulated and real datasets with known cell identities, covering common methods of filtering, doublet detection, normalization, feature selection, denoising, dimensionality reduction and clustering.
-We evaluate the choice of these tools with multi-purpose metrics to assess their ability to reveal cell population structure and lead to efficient clustering.
-
-\parttitle{Conclusion} On the basis of our systematic evaluations of analysis pipelines, we make a number of practical recommendations about current analysis choices and for a comprehensive pipeline.
-The evaluation framework that we developed, \textit{pipeComp} (\url{https://github.com/plger/pipeComp}), has been implemented so as to easily integrate any other step, tool, or evaluation metric allowing extensible benchmarks and easy applications to other fields of research in Bioinformatics, as we demonstrate through a study of the impact of removal of unwanted variation on differential expression analysis.
+We present \textit{pipeComp} (\url{https://github.com/plger/pipeComp}), a flexible R framework for pipeline comparison handling interactions between analysis steps and relying on multi-level evaluation metrics.
+We apply it to the benchmark of single-cell RNA-sequencing analysis pipelines using simulated and real datasets with known cell identities, covering common methods of filtering, doublet detection, normalization, feature selection, denoising, dimensionality reduction and clustering.
+\textit{pipeComp} can easily integrate any other step, tool, or evaluation metric, allowing extensible benchmarks and easy applications to other fields, as we demonstrate through a study of the impact of removal of unwanted variation on differential expression analysis.
 
@@
 \kwd{normalization}
-\kwd{dimensionality reduction}
 \kwd{filtering}
@@
 We selected the most important alternatives for each parameter (see Methods) in order to explore their interactions.
-{\color{red}Unsupervised clustering on the evaluation metrics reveals general patterns in the results as well as the relative importance of the parameters (Figure \ref{fig:overall}), with the clustering method having the most important impact on the results.
-This is however chiefly due to number of clusters called: an ANOVA on all analyses after accounting, in the model, for baseline differences between datasets and the impact of \textit{Seurat}'s resolution parameter, attributed relatively little importance to the clustering method, and instead attributed the largest proportion of variance in the main cluster metrics to the number of dimensions used, followed by normalization (Supplementary Figure 29A).
-Because the choice of dimension reduction impacts the number of clusters, we repeated the analysis of variance in ARI among the combinations of parameters that yielded the true number of clusters (Supplementary Figure 29B).
-We observed that although normalization was the most important parameter among these analyses, the choice of dimensionality (as well as feature selection) came closely behind (Supplementary Figure 29B), suggesting that more attention should be paid to this relatively neglected step.
+Unsupervised clustering on the evaluation metrics reveals general patterns in the results as well as the relative importance of the parameters (Supplementary Figure 29), with the clustering method having the most important impact on the results.
+This is however chiefly due to number of clusters called: an ANOVA on all analyses after accounting, in the model, for baseline differences between datasets and the impact of \textit{Seurat}'s resolution parameter, attributed relatively little importance to the clustering method, and instead attributed the largest proportion of variance in the main cluster metrics to the normalization method and the number of dimensions used (Figure \ref{fig:overall}).
+Part of the impact of the estimated dimensionality depended on its effect on the number of clusters detected, and it indeed explained less variance in ARI at the true number of clusters (Figure \ref{fig:overall}).
+Nevertheless, the choice of dimensionality (as well as feature selection) came closely behind normalization, suggesting that more attention should be paid to this relatively neglected step.
 
-The greater importance of normalization, dimensionality and feature selection over filtering steps is also visible in the clustering (Figure \ref{fig:overall}).
-The top methods (bottom of Figure \ref{fig:overall}) confirm the results of the previous sections, highlighting especially \textit{sctransform}, deviance-based feature selection, and \textit{maxLikGlobal} dimensionality estimates as top methods.
-Interestingly, while the combination of \textit{sctransform} with \textit{Seurat} clustering was particularly performant for most datasets, for the simulated `simMix1` dataset,  standard \textit{Seurat} normalization combined with \textit{scran}-based clustering proved superior.}
+The greater importance of normalization, dimensionality and feature selection over filtering steps is also visible in the clustering (Supplementary Figure 29).
+The top methods (bottom of Supplementary Figure 29) confirm the results of the previous sections, highlighting especially \textit{sctransform}, deviance-based feature selection, and \textit{maxLikGlobal} dimensionality estimates as top methods.
+Interestingly, while the combination of \textit{sctransform} with \textit{Seurat} clustering was particularly performant for most datasets, for the simulated `simMix1` dataset,  standard \textit{Seurat} normalization combined with \textit{scran}-based clustering proved superior.
 
 We next investigated whether the benefits of some methods were conditional on the choices at other steps.
-Although analysis of variance identified many weak but significant interactions, we found that the benefits brought by performant methods were robust to differences in other parts of the pipeline (Supplementary Figures 30-31).
-For instance, we confirmed that doublet removal and \textit{sctransform} normalization were beneficial independently of other steps, and that, again independently of other steps, more stringent filtering did not yield any major improvement over lenient filtering (Supplementary Figure 30).
-Similarly, deviance-based feature selection outcompeted \textit{Seurat}'s default selection method independently of changes at other steps (Supplementary Figure 31).
-{\color{red} ANOVA did reveal some significant (negative) interactions: the top two-way interaction was between \textit{scran} normalization and the elbow method for selection the number of dimensions, and indeed that combination was involved in all of the 30 lowest-ranking pipelines.
-Furthermore, all significant interactions involved the elbow method, suggesting that some methods were less robust to underestimated dimensionality.}
+Although analysis of variance identified some significant interactions (Supplementary Figure 30), we found that the benefits brought by performant methods were robust to differences in other parts of the pipeline (e.g.
+Supplementary Figure 31).
+For instance, we confirmed that doublet removal and \textit{sctransform} normalization were beneficial independently of other steps, and that deviance-based feature selection outcompeted \textit{Seurat}'s default selection method independently of changes at other steps (Supplementary Figure 31).
+ANOVA did reveal some significant (negative) interactions (Supplementary Figure 30): the top two-way interaction was between \textit{scran} normalization and the elbow method for selection the number of dimensions, and indeed that combination was involved in all of the 30 lowest-ranking pipelines.
+Furthermore, all significant interactions involved the elbow method, suggesting that some methods were less robust to underestimated dimensionality.
 
@@
 
-\subsection*{Code and data availability}
-
-All analyses were performed through the \href{https://github.com/plger/pipeComp}{pipeComp} R package, which implements the pipeline framework described here.
-scVI results and results of the clustering section were produced with \textit{pipeComp} 0.99.4, while all other results were produced with version 0.99.2 (\textit{biorxiv}) branch of the repository).
-All code to reproduce the simulations and figures is available in the \url{https://github.com/markrobinsonuzh/scRNA\_pipelines\_paper} repository, which also includes the basic datasets with a standardized annotation.
-{\color{red}All wrappers for methods used in the study can be found in the external data of the \textit{pipeComp} package (\url{https://github.com/plger/pipeComp/tree/master/inst/extdata}).}
-
-The gene counts for the two mixology datasets were downloaded from the  \href{https://github.com/LuyiTian/CellBench_data/data/sincell_with_class.RData}{CellBench repository}, commit 74fe79e.
-For the other datasets, we used the unfiltered counts from \cite{duoClustering2018}, available on the corresponding repository \url{https://github.com/markrobinsonuzh/scRNAseq\_clustering\_comparison}.
-For simplicity, all starting \textit{SingleCellExperiment} objects with standardized metadata are available on \url{https://doi.org/10.6084/m9.figshare.11787210.v1}.
-
-\subsection*{Software and package versions}
-Analyses were performed in R 3.6.0 (Bioconductor 3.9) and the following packages were installed from GitHub repositories: \textit{Seurat} (version 3.0.0), \textit{sctransform} (0.2.0), \textit{DoubletFinder} (2.0.1), \textit{scDblFinder} (1.1.4), \textit{scds} (1.0.0), \textit{SAVERX} (1.0.0), \textit{scImpute} (0.0.9), \textit{ALRA} (commit 7636de8), \textit{DCA} (0.2.2), \textit{DrImpute} (1.2), \textit{ENHANCE} (R version, commit 1571696), \textit{SAVERX} (1.0.0).
-The glmPCA was performed using the \textit{glmpca} package, while the code for deviance calculation was obtained from \url{https://github.com/willtownes/scrna2019} (commit 1ddcc30ebb95d083a685f12fe81d35dd1b0cb1b2).
-
 \subsection*{Simulated datasets}
@@
 
-The set of alternatives found for the overall run (Supplementary Figures 29-31) can be found in the \href{https://github.com/markrobinsonuzh/scRNA_pipelines_paper/blob/master/data/overall_run.Rmd}{paper's repository}.
+The set of alternatives found for the overall run (Figure 8 and Supplementary Figures 29-31) can be found in the \href{https://github.com/markrobinsonuzh/scRNA_pipelines_paper/blob/master/data/overall_run.Rmd}{paper's repository}.
 The analysis of variance was performed with the following model:
@@
 
+\subsection*{Software and package versions}
+
+Analyses were performed in R 3.6.0 (Bioconductor 3.9) and the following packages were installed from GitHub repositories: \textit{Seurat} (version 3.0.0), \textit{sctransform} (0.2.0), \textit{DoubletFinder} (2.0.1), \textit{scDblFinder} (1.1.4), \textit{scds} (1.0.0), \textit{SAVERX} (1.0.0), \textit{scImpute} (0.0.9), \textit{ALRA} (commit 7636de8), \textit{DCA} (0.2.2), \textit{DrImpute} (1.2), \textit{ENHANCE} (R version, commit 1571696), \textit{SAVERX} (1.0.0).
+The glmPCA was performed using the \textit{glmpca} package, while the code for deviance calculation was obtained from \url{https://github.com/willtownes/scrna2019} (commit 1ddcc30ebb95d083a685f12fe81d35dd1b0cb1b2).
+
+\section*{Availability of Data and Materials}
+
+All analyses were performed through the \href{https://github.com/plger/pipeComp}{pipeComp} R package, which implements the pipeline framework described here.
+scVI results and results of the clustering section were produced with \textit{pipeComp} 0.99.4, while all other results were produced with version 0.99.2 (\textit{biorxiv}) branch of the repository).
+All code to reproduce the simulations and figures is available in the \url{https://github.com/markrobinsonuzh/scRNA\_pipelines\_paper} repository, which also includes the basic datasets with a standardized annotation.
+All wrappers for methods used in the study can be found in the external data of the \textit{pipeComp} package (\url{https://github.com/plger/pipeComp/tree/master/inst/extdata}).
+
+The gene counts for the two mixology datasets were downloaded from the  \href{https://github.com/LuyiTian/CellBench_data/data/sincell_with_class.RData}{CellBench repository}, commit 74fe79e.
+For the other datasets, we used the unfiltered counts from \cite{duoClustering2018}, available on the corresponding repository \url{https://github.com/markrobinsonuzh/scRNAseq\_clustering\_comparison}.
+For simplicity, all starting \textit{SingleCellExperiment} objects with standardized metadata are available on \url{https://doi.org/10.6084/m9.figshare.11787210.v1}.
+
+All data and code are released under a GNU General Public License.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \begin{backmatter}
+
+\section*{Ethics approval}
+Not applicable.
 
@@
 
-\section*{Acknowledgements}
+\section*{Funding}
+
 This work was supported by the Swiss National Science Foundation (grants 310030\_175841, CRSII5\_177208) as well as the Chan Zuckerberg Initiative DAF (grant number 2018-182828), an advised fund of Silicon Valley Community Foundation.
@@
     \centering
-    \includegraphics[width=\textwidth, keepaspectratio]{{overall.pdf}}
-    \caption{\textbf{Overview of the combination of main alternative parameters.} Unsupervised clustering of the results (above) and top results (below).
-The color-mapping schemes are the same as described in the previous figures.}
+    \includegraphics[width=\textwidth, keepaspectratio]{{anova.pdf}}
+    \caption{\textbf{Variance in clustering metrics explained by the parameters.} See methods for more detail.}
     \label{fig:overall}
@@
 \begin{table}[h!]
-\caption{\textbf{Overview of the benchmark datasets.} {\color{red}`\# features' indicates the number of features detected in at least 10 cells.}}
+\caption{\textbf{Table S1: Overview of the benchmark datasets.} \# features' indicates the number of features detected in at least 10 cells.}
 \label{tab:table1}
@@
 \section*{Additional Files}
-  \subsection*{\textbf{Additional file 1: Supplementary figures}}
-  Additional results of evaluations across datasets among various metrics. (PDF 16 MB)
-  \subsection*{\textcolor{red}{\textbf{Additional file 2: Supplementary methods}}}
-  \textcolor{red}{More detailed description of \textit{pipeComp} and comparison with other tools. (PDF 0.5 MB)}
+  \subsection*{\textbf{Additional file 1}}
+  Supplementary Figures (PDF 16 MB)
+  \subsection*{\textbf{Additional file 2}}
+  More detailed description of \textit{pipeComp} and comparison with other tools. (PDF 0.5 MB)
 \end{backmatter}

### 2020-08-04 bmc_article.tex
changed links to supp figures

@@
 This is especially important because end-point metrics, such as the adjusted Rand index (ARI) \cite{HubertARI1985} for clustering, are not perfect.
-For example, although the meaning of an ARI score is independent of the number of true subpopulations \cite{steinleyProperties2004}, the number of clusters called is by far the most important determinant of the score: the farther it is from the actual number of subpopulations, the worse the ARI (Supplementary Figures 2-3).
+For example, although the meaning of an ARI score is independent of the number of true subpopulations \cite{steinleyProperties2004}, the number of clusters called is by far the most important determinant of the score: the farther it is from the actual number of subpopulations, the worse the ARI (Supplementary Figure 2-3).
 In this context, one strategy has been to cluster across various resolutions and only consider the results that have the right number of clusters \cite{duoClustering2018}.
@@
 
-Reasoning that the cells we wish to exclude are the cells that would be misclassified, we measured the rate of misclassification of each cell in each dataset across a variety of clustering pipelines, correcting for the median misclassification rate of the subpopulation and then evaluated what properties could be predictive of misclassification (Supplementary Figures 8-10).
+Reasoning that the cells we wish to exclude are the cells that would be misclassified, we measured the rate of misclassification of each cell in each dataset across a variety of clustering pipelines, correcting for the median misclassification rate of the subpopulation and then evaluated what properties could be predictive of misclassification (Supplementary Figure 8-10).
 We could not identify any property or simple combination thereof that would be consistently predictive of misclassification; the only feature that stood out across more than one dataset (the Zheng datasets) was that cells with very high read counts have a higher chance of being misclassified{\color{red}, although such a pattern was not observed in the other datasets.}
@@
 These estimates were also in good agreement with the deviance explained, although lowly-expressed genes could have a high deviance explained without having much of their variance explained by subpopulation (Supplementary Figure 19B-D).
-We therefore compared the proportion of the cumulative variance/deviance explained by the top X genes that could be retrieved by each gene ranking method (Supplementary Figures 20-21).
-We first focused on the first 1000 genes to highlight the differences between methods, although a higher number of selected genes decreased the differences between methods (Supplementary Figures 20-22).
+We therefore compared the proportion of the cumulative variance/deviance explained by the top X genes that could be retrieved by each gene ranking method (Supplementary Figure 20-21).
+We first focused on the first 1000 genes to highlight the differences between methods, although a higher number of selected genes decreased the differences between methods (Supplementary Figure 20-22).
 The standardized measures of variability were systematically worse than their non-standardized counterparts in selecting genes with a high proportion of variance explained by subpopulation.
-Regarding the percentage of deviance explained however, the standardized measures were often superior (Figure \ref{fig:selection}A and Supplementary Figures 20-21).
+Regarding the percentage of deviance explained however, the standardized measures were often superior (Figure \ref{fig:selection}A and Supplementary Figure 20-21).
 \textit{Deviance} proved the method of choice to prioritize genes with a high variance explained by subpopulations (with mere expression level proving surprisingly performant) but did not perform so well to select genes with a high deviance explained.
@@
 Since Euclidean distance decreases as the number of non-discriminating dimensions increases, there is usually a trade-off between selecting enough dimensions to keep most information and excluding smaller dimensions that may represent technical noise or other unwanted sources of variation.
-Overall, increasing the number of dimensions robustly led to a decrease in the number of clusters (Supplementary Figures 16 and 24).
+Overall, increasing the number of dimensions robustly led to a decrease in the number of clusters (Supplementary Figure 16 and 24).
 This tended to affect the accuracy of the clustering (Supplementary Figure 25), although in both cases (number of clusters and ARI), \textit{Seurat}'s resolution parameter had a much stronger impact.
@@
 
-Since ARI is dominated by differences in the number of clusters (Supplementary Figures 2-3) and no single metric is perfect, we diversified them (Figure \ref{fig:clust}).
+Since ARI is dominated by differences in the number of clusters (Supplementary Figure 2-3) and no single metric is perfect, we diversified them (Figure \ref{fig:clust}).
 MI has the virtue of not decreasing when a true subpopulation is split into two clusters, which is arguably less problematic (and might well reflect unknown biological subgroups), but as a consequence it can be biased towards methods producing higher resolution clustering (Supplementary Figure 3).
@@
 Finally, the ARI score at the true number of clusters, when available, showed similar performances.
-However, because \textit{Seurat}'s resolution parameter had a large impact on the number of clusters identified (Supplementary Figures 2 and 24), \textit{Seurat} could always be coerced into producing the right number of clusters.
+However, because \textit{Seurat}'s resolution parameter had a large impact on the number of clusters identified (Supplementary Figure 2 and 24), \textit{Seurat} could always be coerced into producing the right number of clusters.
 Instead, the number of clusters found by \textit{scran} was considerably less influenced by the available parameters (number of nearest neighbors or steps in the random walk - see Supplementary Figure 28), and as a result \textit{scran}-based clustering sometimes never produced a partitioning with the right number of clusters.
@@
 As previously reported \citep{ComprehensiveAssessment2014}, we found that more stringent filtering improved DEA (Supplementary Figure 33A) and that the different DEA methods had similar performances (Figure \ref{fig:svadea}A and Supplementary Figure 33A).
-Including a SVA-based step increased sensitivity across all datasets, particularly in datasets with technical vectors of variation (\textit{seqc} and \textit{simulation}), and benefited all DEA methods in a similar fashion (Figure \ref{fig:svadea}A,C and Supplementary Figures 33-34).
+Including a SVA-based step increased sensitivity across all datasets, particularly in datasets with technical vectors of variation (\textit{seqc} and \textit{simulation}), and benefited all DEA methods in a similar fashion (Figure \ref{fig:svadea}A,C and Supplementary Figure 33-34).
 With the exception of the \texttt{RUVr} method, which led to a high False Discovery Rate (FDR), all other methods showed comparable gains in sensitivity while maintaining a decent error control (if a little above the nominal FDR), even when increasing the number of surrogate variables used (Figure \ref{fig:svadea}B-C).
@@
 
-The set of alternatives found for the overall run (Figure 8 and Supplementary Figures 29-31) can be found in the \href{https://github.com/markrobinsonuzh/scRNA_pipelines_paper/blob/master/data/overall_run.Rmd}{paper's repository}.
+The set of alternatives found for the overall run (Figure 8 and Supplementary Figure 29-31) can be found in the \href{https://github.com/markrobinsonuzh/scRNA_pipelines_paper/blob/master/data/overall_run.Rmd}{paper's repository}.
 The analysis of variance was performed with the following model:
@@
   \subsection*{\textbf{Additional file 1}}
-  Supplementary Figures (PDF 16 MB)
+  Supplementary Figure (PDF 16 MB)
   \subsection*{\textbf{Additional file 2}}

### 2020-08-04 bmc_article.tex
changed supp fig links

@@
 Although we use some datasets based on other protocols, our focus is especially on droplet-based datasets that do not include exogenous control RNA (i.e.
-spike-ins); see Table \ref{tab:table1} and Supplementary Figure 1 for more details.
+spike-ins); see Table \ref{tab:table1} and Additional File 1: Fig S1 for more details.
 In addition to previously-used benchmark datasets with true cell labels \cite{duoClustering2018,tianMixology2018}, we simulated two datasets with a hierarchical subpopulation structure based on real 10x human and mouse data using \textit{muscat} \cite{CrowellMuscat2019}.
@@
 This is especially important because end-point metrics, such as the adjusted Rand index (ARI) \cite{HubertARI1985} for clustering, are not perfect.
-For example, although the meaning of an ARI score is independent of the number of true subpopulations \cite{steinleyProperties2004}, the number of clusters called is by far the most important determinant of the score: the farther it is from the actual number of subpopulations, the worse the ARI (Supplementary Figure 2-3).
+For example, although the meaning of an ARI score is independent of the number of true subpopulations \cite{steinleyProperties2004}, the number of clusters called is by far the most important determinant of the score: the farther it is from the actual number of subpopulations, the worse the ARI (Additional File 1: Fig S2-3).
 In this context, one strategy has been to cluster across various resolutions and only consider the results that have the right number of clusters \cite{duoClustering2018}.
@@
 While doublets of the same cell type are relatively innocuous for most downstream applications due to their conservation of the relative expression between genes, doublets formed from different cell types or states are likely to be misclassified and could potentially distort downstream analysis.
-In some cases, doublets can be identified through their unusually high number of reads and detected features, but this is not always the case (Supplementary Figure 4).
+In some cases, doublets can be identified through their unusually high number of reads and detected features, but this is not always the case (Additional File 1: Fig S4).
 A number of methods were developed to identify doublets, in most cases by comparing each cell to artificially-created doublets \cite{mcginnisDoubletfinder2019, LunScran2016, BaisScds2019}.
@@
 This can for instance be performed through the \texttt{isOutlier} function from \textit{scater} that measures, for a given control property, the number of median absolute deviations (MADs) of each cell from the median of all cells.
-Supplementary Figure 5 shows the distributions of some of the typical cell properties commonly used.
+Additional File 1: Fig S5 shows the distributions of some of the typical cell properties commonly used.
 Of note, these properties tend to be correlated, with some exceptions.
-For example, while a high proportion of mitochondrial reads is often correlated with a high proportion of the counts in the top features, there can also be other reasons for an over-representation of highly-expressed features (Supplementary Figure 6), such as an over-amplification in non-UMI datasets.
-In our experience, 10X datasets also exhibit a very strong correlation between the total counts and the total features even across very different cell types (Supplementary Figure 7).
+For example, while a high proportion of mitochondrial reads is often correlated with a high proportion of the counts in the top features, there can also be other reasons for an over-representation of highly-expressed features (Additional File 1: Fig S6), such as an over-amplification in non-UMI datasets.
+In our experience, 10X datasets also exhibit a very strong correlation between the total counts and the total features even across very different cell types (Additional File 1: Fig S7).
 We therefore also measure the ratio between the two and treat cells strongly departing from this trend with suspicion.
 
-Reasoning that the cells we wish to exclude are the cells that would be misclassified, we measured the rate of misclassification of each cell in each dataset across a variety of clustering pipelines, correcting for the median misclassification rate of the subpopulation and then evaluated what properties could be predictive of misclassification (Supplementary Figure 8-10).
+Reasoning that the cells we wish to exclude are the cells that would be misclassified, we measured the rate of misclassification of each cell in each dataset across a variety of clustering pipelines, correcting for the median misclassification rate of the subpopulation and then evaluated what properties could be predictive of misclassification (Additional File 1: Fig S8-10).
 We could not identify any property or simple combination thereof that would be consistently predictive of misclassification; the only feature that stood out across more than one dataset (the Zheng datasets) was that cells with very high read counts have a higher chance of being misclassified{\color{red}, although such a pattern was not observed in the other datasets.}
@@
 As a result, thresholds in terms of number of MADs from the whole distribution can lead to strong biases against certain subpopulations (Figure \ref{fig:filtering}A).
-We therefore examined the tradeoff between the increased accuracy of filtering and the maximum proportion of cells excluded per subpopulation (Figure \ref{fig:filtering}B and Supplementary Figure 11).
+We therefore examined the tradeoff between the increased accuracy of filtering and the maximum proportion of cells excluded per subpopulation (Figure \ref{fig:filtering}B and Additional File 1: Fig S11).
 Since filtering changes the relative abundance of the different subpopulations, global clustering accuracy metrics such as ARI are not appropriate here.
@@
 Most of the benefits could be achieved without very stringent filtering and minimizing subpopulation bias (Figure \ref{fig:filtering}B).
-Applying the same filtering criteria on individual clusters of cells (identified through scran's \texttt{quickCluster} method) rather than on whole dataset resulted in nearly no cells being filtered out (Supplementary Figure 12A).
+Applying the same filtering criteria on individual clusters of cells (identified through scran's \texttt{quickCluster} method) rather than on whole dataset resulted in nearly no cells being filtered out (Additional File 1: Fig S12A).
 This suggests that stringent filtering on the global population tends to discard cells of subpopulations with more extreme properties (e.g.
@@
 PCA-based selection did not appear superior to feature-wise filters.
-We found very little overlap between the cells excluded by doublet removal and those excluded by MAD-based filtering (Supplementary Figure 12B).
+We found very little overlap between the cells excluded by doublet removal and those excluded by MAD-based filtering (Additional File 1: Fig S12B).
 Of note, doublet removal in conjunction with filtering tended to improve accuracy not only in datasets that do include heterotypic doublets (mixology datasets), but also some others in which such doublets are unlikely given the experimental design (e.g.
@@
 Mitochondrial reads have been associated with cell degradation and there is evidence that ribosomal genes can influence the clustering output, hiding other biological structure in the analysis \cite{freytagComparison2018}.
-We therefore investigated whether excluding one category of features or the other, or using only protein-coding genes, had an impact on the ability to distinguish subpopulations (Supplementary Figure 13).
+We therefore investigated whether excluding one category of features or the other, or using only protein-coding genes, had an impact on the ability to distinguish subpopulations (Additional File 1: Fig S13).
 Removal of ribosomal genes robustly reduced the quality of the clustering, suggesting that they represent real biological differences between subpopulations.
@@
 However, as library size tends to vary across subpopulations, part of this effect can simply reflect biological differences.
-We therefore assessed to what extent the first principal component still retained a correlation with the library size and the number of detected features, removing the confounding covariation with the subpopulations (Figure \ref{fig:norm}A and Supplementary Figure 14).
+We therefore assessed to what extent the first principal component still retained a correlation with the library size and the number of detected features, removing the confounding covariation with the subpopulations (Figure \ref{fig:norm}A and Additional File 1: Fig S14).
 Scaling tended to remove much of the correlation with these features, and while most methods were able to remove most of the effect, the correlation was lowest with \textit{sctransform}.
@@
 Zhengmix8eq and simMix1) and subpopulations were consistently more challenging across all methods.
-{\color{red}An exception was \textit{scVI}-based normalization, which performed similarly to other methods on some datasets (Kumar, simMix2, Zhengmix4eq, Zhengmix4uneq), but very poorly on others, in terms of both the subpopulation silhouette and clustering accuracy (Supplementary Figure 13; of note, these values were not meant to be used for clustering, and the usage of the \textit{scVI} latent space will be discussed below along with dimensionality reductions)}.
+{\color{red}An exception was \textit{scVI}-based normalization, which performed similarly to other methods on some datasets (Kumar, simMix2, Zhengmix4eq, Zhengmix4uneq), but very poorly on others, in terms of both the subpopulation silhouette and clustering accuracy (Additional File 1: Fig S13; of note, these values were not meant to be used for clustering, and the usage of the \textit{scVI} latent space will be discussed below along with dimensionality reductions)}.
 Scaling (e.g.
@@
 Finally, we monitored whether, under the same downstream clustering analysis, different normalization methods tended to lead to an over- or under-estimation of the number of clusters.
-Although some methods had a tendency to lead to a higher (e.g., \textit{sctransform}) or lower (e.g., stable genes) number of clusters, the effect was very mild and not entirely systematic (Supplementary Figure 16).
-We also confirmed that, especially in the UMI-based datasets, \textit{sctransform} did in fact successfully stabilize variance across mean counts (Supplementary Figure 17), at little apparent cost on computation time (Supplementary Figure 18).
+Although some methods had a tendency to lead to a higher (e.g., \textit{sctransform}) or lower (e.g., stable genes) number of clusters, the effect was very mild and not entirely systematic (Additional File 1: Fig S16).
+We also confirmed that, especially in the UMI-based datasets, \textit{sctransform} did in fact successfully stabilize variance across mean counts (Additional File 1: Fig S17), at little apparent cost on computation time (Additional File 1: Fig S18).
 
@@
 As the proportion of variability in a gene attributable to subpopulations can be measured in various ways, we first compared three approaches: ANOVA on log-normalized count, ANOVA on \textit{sctransform} normalization, and deviance explained.
-The ANOVAs performed on a standard \textit{Seurat} normalization and on \textit{sctransform} data were highly correlated (Pearson correlation >0.9, Supplementary Figure 19A).
-These estimates were also in good agreement with the deviance explained, although lowly-expressed genes could have a high deviance explained without having much of their variance explained by subpopulation (Supplementary Figure 19B-D).
-We therefore compared the proportion of the cumulative variance/deviance explained by the top X genes that could be retrieved by each gene ranking method (Supplementary Figure 20-21).
-We first focused on the first 1000 genes to highlight the differences between methods, although a higher number of selected genes decreased the differences between methods (Supplementary Figure 20-22).
+The ANOVAs performed on a standard \textit{Seurat} normalization and on \textit{sctransform} data were highly correlated (Pearson correlation >0.9, Additional File 1: Fig S19A).
+These estimates were also in good agreement with the deviance explained, although lowly-expressed genes could have a high deviance explained without having much of their variance explained by subpopulation (Additional File 1: Fig S19B-D).
+We therefore compared the proportion of the cumulative variance/deviance explained by the top X genes that could be retrieved by each gene ranking method (Additional File 1: Fig S20-21).
+We first focused on the first 1000 genes to highlight the differences between methods, although a higher number of selected genes decreased the differences between methods (Additional File 1: Fig S20-22).
 The standardized measures of variability were systematically worse than their non-standardized counterparts in selecting genes with a high proportion of variance explained by subpopulation.
-Regarding the percentage of deviance explained however, the standardized measures were often superior (Figure \ref{fig:selection}A and Supplementary Figure 20-21).
+Regarding the percentage of deviance explained however, the standardized measures were often superior (Figure \ref{fig:selection}A and Additional File 1: Fig S20-21).
 \textit{Deviance} proved the method of choice to prioritize genes with a high variance explained by subpopulations (with mere expression level proving surprisingly performant) but did not perform so well to select genes with a high deviance explained.
@@
 In general, we found deviance and unstandardized estimates of variance to provide the best results across datasets and normalization methods.
-Increasing the number of features selected also tended to lead to an increase in the accuracy of the clustering, typically plateauing after 4000 features (Supplementary Figure 22).
+Increasing the number of features selected also tended to lead to an increase in the accuracy of the clustering, typically plateauing after 4000 features (Additional File 1: Fig S22).
 
@@
 
-The impact of the dimensionality reduction methods tested was far greater than that of normalization or feature selection (Supplementary Figure 23).
-Overall, weighting the components by their explained variance tended to increase silhouette width (Supplementary Figure 23A) and, in most datasets, clustering accuracy (Supplementary Figure 23B).
+The impact of the dimensionality reduction methods tested was far greater than that of normalization or feature selection (Additional File 1: Fig S23).
+Overall, weighting the components by their explained variance tended to increase silhouette width (Additional File 1: Fig S23A) and, in most datasets, clustering accuracy (Additional File 1: Fig S23B).
 GLM-PCA tended to increase the average silhouette width of already well-defined subpopulations, but \textit{Seurat}'s PCA procedure however proved superior on all metrics.
 {\color{red} Like GLM-PCA, \textit{scVI}'s Linearly Decoded (LD) data \cite{SvenssonLDVAE2020} and latent space do not explicitly rely on normalized counts.
-Their performance was however worse than \textit{Seurat}'s PCA in view of the silhouette width (Supplementary Figure 15).
+Their performance was however worse than \textit{Seurat}'s PCA in view of the silhouette width (Additional File 1: Fig S15).
 In terms of clustering accuracy, \textit{scVI}'s latent space performed better that its LD but was nevertheless less accurate than PCA-based alternatives on the datasets with a very low amount of cells (Koh and Kumar) and harboring many subpopulations (simMix1 and Zhengmix8eq).}
@@
 Since Euclidean distance decreases as the number of non-discriminating dimensions increases, there is usually a trade-off between selecting enough dimensions to keep most information and excluding smaller dimensions that may represent technical noise or other unwanted sources of variation.
-Overall, increasing the number of dimensions robustly led to a decrease in the number of clusters (Supplementary Figure 16 and 24).
-This tended to affect the accuracy of the clustering (Supplementary Figure 25), although in both cases (number of clusters and ARI), \textit{Seurat}'s resolution parameter had a much stronger impact.
+Overall, increasing the number of dimensions robustly led to a decrease in the number of clusters (Additional File 1: Fig S16 and 24).
+This tended to affect the accuracy of the clustering (Additional File 1: Fig S25), although in both cases (number of clusters and ARI), \textit{Seurat}'s resolution parameter had a much stronger impact.
 
@@
 
-{\color{red} \textit{MCVR} and the global maximum likelihood based on translated Poisson Mixture Model (\textit{maxLikGlobal}) tended to produce clusterings with smaller deviations to the true number of subpopulations (Supplementary Figure 26).
-Although most methods performed well on the various clustering measures, \textit{maxLikGlobal} (using 10 or 20 nearest neighbors, although estimates were relatively stable across various values of \textit{k} -- see Supplementary Figure 27) provided the dimensionality estimate that best separated challenging subpopulations (Figure \ref{fig:dim}C) and tended to result in the best clustering accuracy (Figure \ref{fig:dim}D).
+{\color{red} \textit{MCVR} and the global maximum likelihood based on translated Poisson Mixture Model (\textit{maxLikGlobal}) tended to produce clusterings with smaller deviations to the true number of subpopulations (Additional File 1: Fig S26).
+Although most methods performed well on the various clustering measures, \textit{maxLikGlobal} (using 10 or 20 nearest neighbors, although estimates were relatively stable across various values of \textit{k} -- see Additional File 1: Fig S27) provided the dimensionality estimate that best separated challenging subpopulations (Figure \ref{fig:dim}C) and tended to result in the best clustering accuracy (Figure \ref{fig:dim}D).
 \textit{MCVR} did not perform well for the datasets without UMIs (Koh and Kumar), which violates the assumptions of the method, and while it did increase the separability of some subpopulations (Figure \ref{fig:dim}C), it was detrimental to the classification of some others.
@@
 
-Since ARI is dominated by differences in the number of clusters (Supplementary Figure 2-3) and no single metric is perfect, we diversified them (Figure \ref{fig:clust}).
-MI has the virtue of not decreasing when a true subpopulation is split into two clusters, which is arguably less problematic (and might well reflect unknown biological subgroups), but as a consequence it can be biased towards methods producing higher resolution clustering (Supplementary Figure 3).
+Since ARI is dominated by differences in the number of clusters (Additional File 1: Fig S2-3) and no single metric is perfect, we diversified them (Figure \ref{fig:clust}).
+MI has the virtue of not decreasing when a true subpopulation is split into two clusters, which is arguably less problematic (and might well reflect unknown biological subgroups), but as a consequence it can be biased towards methods producing higher resolution clustering (Additional File 1: Fig S3).
 Similarly, precision per true subpopulation is considerably more robust to differences in the number of clusters.
@@
 Finally, the ARI score at the true number of clusters, when available, showed similar performances.
-However, because \textit{Seurat}'s resolution parameter had a large impact on the number of clusters identified (Supplementary Figure 2 and 24), \textit{Seurat} could always be coerced into producing the right number of clusters.
-Instead, the number of clusters found by \textit{scran} was considerably less influenced by the available parameters (number of nearest neighbors or steps in the random walk - see Supplementary Figure 28), and as a result \textit{scran}-based clustering sometimes never produced a partitioning with the right number of clusters.
+However, because \textit{Seurat}'s resolution parameter had a large impact on the number of clusters identified (Additional File 1: Fig S2 and 24), \textit{Seurat} could always be coerced into producing the right number of clusters.
+Instead, the number of clusters found by \textit{scran} was considerably less influenced by the available parameters (number of nearest neighbors or steps in the random walk - see Additional File 1: Fig S28), and as a result \textit{scran}-based clustering sometimes never produced a partitioning with the right number of clusters.
 This observation, along with \textit{scran}'s higher MI score, suggest that \textit{scran} sometimes simply divides a real subpopulation into two clusters (possibly tracking some unknown biological differences) rather than committing misclassification errors.
@@
 We selected the most important alternatives for each parameter (see Methods) in order to explore their interactions.
-Unsupervised clustering on the evaluation metrics reveals general patterns in the results as well as the relative importance of the parameters (Supplementary Figure 29), with the clustering method having the most important impact on the results.
+Unsupervised clustering on the evaluation metrics reveals general patterns in the results as well as the relative importance of the parameters (Additional File 1: Fig S29), with the clustering method having the most important impact on the results.
 This is however chiefly due to number of clusters called: an ANOVA on all analyses after accounting, in the model, for baseline differences between datasets and the impact of \textit{Seurat}'s resolution parameter, attributed relatively little importance to the clustering method, and instead attributed the largest proportion of variance in the main cluster metrics to the normalization method and the number of dimensions used (Figure \ref{fig:overall}).
@@
 
-The greater importance of normalization, dimensionality and feature selection over filtering steps is also visible in the clustering (Supplementary Figure 29).
-The top methods (bottom of Supplementary Figure 29) confirm the results of the previous sections, highlighting especially \textit{sctransform}, deviance-based feature selection, and \textit{maxLikGlobal} dimensionality estimates as top methods.
+The greater importance of normalization, dimensionality and feature selection over filtering steps is also visible in the clustering (Additional File 1: Fig S29).
+The top methods (bottom of Additional File 1: Fig S29) confirm the results of the previous sections, highlighting especially \textit{sctransform}, deviance-based feature selection, and \textit{maxLikGlobal} dimensionality estimates as top methods.
 Interestingly, while the combination of \textit{sctransform} with \textit{Seurat} clustering was particularly performant for most datasets, for the simulated `simMix1` dataset,  standard \textit{Seurat} normalization combined with \textit{scran}-based clustering proved superior.
@@
 We next investigated whether the benefits of some methods were conditional on the choices at other steps.
-Although analysis of variance identified some significant interactions (Supplementary Figure 30), we found that the benefits brought by performant methods were robust to differences in other parts of the pipeline (e.g.
-Supplementary Figure 31).
-For instance, we confirmed that doublet removal and \textit{sctransform} normalization were beneficial independently of other steps, and that deviance-based feature selection outcompeted \textit{Seurat}'s default selection method independently of changes at other steps (Supplementary Figure 31).
-ANOVA did reveal some significant (negative) interactions (Supplementary Figure 30): the top two-way interaction was between \textit{scran} normalization and the elbow method for selection the number of dimensions, and indeed that combination was involved in all of the 30 lowest-ranking pipelines.
+Although analysis of variance identified some significant interactions (Additional File 1: Fig S30), we found that the benefits brought by performant methods were robust to differences in other parts of the pipeline (e.g.
+Additional File 1: Fig S31).
+For instance, we confirmed that doublet removal and \textit{sctransform} normalization were beneficial independently of other steps, and that deviance-based feature selection outcompeted \textit{Seurat}'s default selection method independently of changes at other steps (Additional File 1: Fig S31).
+ANOVA did reveal some significant (negative) interactions (Additional File 1: Fig S30): the top two-way interaction was between \textit{scran} normalization and the elbow method for selection the number of dimensions, and indeed that combination was involved in all of the 30 lowest-ranking pipelines.
 Furthermore, all significant interactions involved the elbow method, suggesting that some methods were less robust to underestimated dimensionality.
@@
 Since preliminary analysis showed that all methods performed equally well or better on normalized data, we applied them after filtering and normalization, but before scaling and reduction.
-Although some of the methods (e.g., \textit{DRImpute\_process} and \textit{alra\_norm}) did improve the separability of some more elusive subpopulations, no method had a systematically positive impact on the average silhouette width across all subpopulations (Supplementary Figure 32A).
-When restricting ourselves to clustering analyses that yielded the `right' number of clusters, all tested methods improved classification  compared to a scenario with no imputation step (`none' label, Supplementary Figure 32B).
+Although some of the methods (e.g., \textit{DRImpute\_process} and \textit{alra\_norm}) did improve the separability of some more elusive subpopulations, no method had a systematically positive impact on the average silhouette width across all subpopulations (Additional File 1: Fig S32A).
+When restricting ourselves to clustering analyses that yielded the `right' number of clusters, all tested methods improved classification  compared to a scenario with no imputation step (`none' label, Additional File 1: Fig S32B).
 However, the situation was not so straightforward with alternative metrics, where some methods (e.g., \textit{ENHANCE}) consistently underperformed.
@@
 We developed 3 benchmark datasets: 1) the \texttt{ipsc} data is a random selection of 10 vs 10 samples from \citep{carcamoOriveAnalysisTranscriptional2017}, which are very heterogeneous, and foldchanges were manually added to 300 genes; 2) the \texttt{seqc} dataset contains 5 vs 5 samples of two mixtures of the SEQC consortium with different ERCC spike-in mixes \citep{ComprehensiveAssessment2014}, and includes a partially correlated batch effect; 3) A \texttt{simulation} of 8 vs 8 samples with two technical vectors of variation (see Methods for more details).
-As previously reported \citep{ComprehensiveAssessment2014}, we found that more stringent filtering improved DEA (Supplementary Figure 33A) and that the different DEA methods had similar performances (Figure \ref{fig:svadea}A and Supplementary Figure 33A).
-Including a SVA-based step increased sensitivity across all datasets, particularly in datasets with technical vectors of variation (\textit{seqc} and \textit{simulation}), and benefited all DEA methods in a similar fashion (Figure \ref{fig:svadea}A,C and Supplementary Figure 33-34).
+As previously reported \citep{ComprehensiveAssessment2014}, we found that more stringent filtering improved DEA (Additional File 1: Fig S33A) and that the different DEA methods had similar performances (Figure \ref{fig:svadea}A and Additional File 1: Fig S33A).
+Including a SVA-based step increased sensitivity across all datasets, particularly in datasets with technical vectors of variation (\textit{seqc} and \textit{simulation}), and benefited all DEA methods in a similar fashion (Figure \ref{fig:svadea}A,C and Additional File 1: Fig S33-34).
 With the exception of the \texttt{RUVr} method, which led to a high False Discovery Rate (FDR), all other methods showed comparable gains in sensitivity while maintaining a decent error control (if a little above the nominal FDR), even when increasing the number of surrogate variables used (Figure \ref{fig:svadea}B-C).
@@
 
-On the basis of our findings, we can make a number of practical analysis recommendations, also summarized in Supplementary Figure 35:
+On the basis of our findings, we can make a number of practical analysis recommendations, also summarized in Additional File 1: Fig S35:
 
@@
 The average silhouette width of half of their subpopulations was systematically lower than the other subpopulations/ datasets and the ARI score at true number of clusters, if ever achieved, was always poor.
-It is likely that these two datasets are challenging due to their combined sparsity and complexity (see Supplementary Figure 1).
+It is likely that these two datasets are challenging due to their combined sparsity and complexity (see Additional File 1: Fig S1).
 The challenge of distinguishing subpopulations of T-cells in the Zhengmix8eq dataset was previously discussed \cite{duoClustering2018}.
@@
 Applying this framework to scRNAseq clustering enabled us to make
-clear recommendations on the steps of filtering, normalization, feature selection, denoising, dimensionality reduction and clustering (Supplementary Figure 35).
+clear recommendations on the steps of filtering, normalization, feature selection, denoising, dimensionality reduction and clustering (Additional File 1: Fig S35).
 We demonstrate how a diversity of multilevel metrics can be more robust, more sensitive, and more nuanced than simply evaluating the final clustering performance.
@@
 We therefore evaluated it separately and compared it to a representative subset of pipelines using \textit{sctransform} normalization and \textit{Seurat}'s PCA dimension reduction method.
-The resulting evaluation is available in the Supplementary Figure 13.
+The resulting evaluation is available in the Additional File 1: Fig S13.
 
@@
 
-The set of alternatives found for the overall run (Figure 8 and Supplementary Figure 29-31) can be found in the \href{https://github.com/markrobinsonuzh/scRNA_pipelines_paper/blob/master/data/overall_run.Rmd}{paper's repository}.
+The set of alternatives found for the overall run (Figure 8 and Additional File 1: Fig S29-31) can be found in the \href{https://github.com/markrobinsonuzh/scRNA_pipelines_paper/blob/master/data/overall_run.Rmd}{paper's repository}.
 The analysis of variance was performed with the following model:
@@
   \subsection*{\textbf{Additional file 1}}
-  Supplementary Figure (PDF 16 MB)
+  Additional File 1: Fig S(PDF 16 MB)
   \subsection*{\textbf{Additional file 2}}

### 2020-08-04 bmc_article.tex
Updates from Overleaf

@@
 Here, we develop a flexible R framework for pipeline comparison and apply it to the evaluation of the various steps of analysis leading from an initial count matrix to a cluster assignment, which are critical in a wide range of applications.
-We collected real datasets of known cell composition (Table \ref{tab:table1}) and used a variety of evaluation metrics to investigate in a multilevel fashion the impact of various parameters and variations around a core scRNAseq pipeline.
+We collected real datasets of known cell composition (Table S\ref{tab:table1}) and used a variety of evaluation metrics to investigate in a multilevel fashion the impact of various parameters and variations around a core scRNAseq pipeline.
 Although we use some datasets based on other protocols, our focus is especially on droplet-based datasets that do not include exogenous control RNA (i.e.
-spike-ins); see Table \ref{tab:table1} and Additional File 1: Fig S1 for more details.
+spike-ins); see Table S\ref{tab:table1} and Additional File 1: Fig S1 for more details.
 In addition to previously-used benchmark datasets with true cell labels \cite{duoClustering2018,tianMixology2018}, we simulated two datasets with a hierarchical subpopulation structure based on real 10x human and mouse data using \textit{muscat} \cite{CrowellMuscat2019}.
@@
 
-The \textit{pipeComp} package defines a pipeline as, minimally, a list of functions executed consecutively on the output of the previous one (Figure \ref{fig:explanation}A; see also Supplementary methods).
+The \textit{pipeComp} package defines a pipeline as, minimally, a list of functions executed consecutively on the output of the previous one (Figure \ref{fig:explanation}A; see also Additional File 2).
 In addition, optional benchmark functions can be set for each step to provide standardized, multi-layered evaluation metrics.
-Given such a \texttt{PipelineDefinition} object, a set of alternative parameters (which might include different subroutines calling different methods) and benchmark datasets, the \texttt{runPipeline} function then proceeds through all combinations of arguments (or a desired subset), avoiding recomputing the same step twice without the need to save all alternative intermediates, and compiling evaluations (including run time) on the fly (see Supplementary methods for more detail).
+Given such a \texttt{PipelineDefinition} object, a set of alternative parameters (which might include different subroutines calling different methods) and benchmark datasets, the \texttt{runPipeline} function then proceeds through all combinations of arguments (or a desired subset), avoiding recomputing the same step twice without the need to save all alternative intermediates, and compiling evaluations (including run time) on the fly (see Additional File 2 for more detail).
 Variations in a given parameter can be evaluated using all metrics from this point downward in the pipeline.
@@
 This avoids the need to store potentially large intermediate data for all possible permutations and thus allowing a combinatorial benchmark.
-This, along with the set-up simplicity (demonstrated here by our separate SVA-DEA benchmark), also distinguishes \textit{pipeComp} from general workflow frameworks such as \textit{drake} \citep{landauDrakePackage2018}; indeed we estimate that the benchmarks discussed here would have taken approximately 16TB of storage had all non-redundant intermediates been saved (see the Supplementary methods for more detailed explanations of the framework and comparison to \textit{CellBench} and \textit{drake}).
+This, along with the set-up simplicity (demonstrated here by our separate SVA-DEA benchmark), also distinguishes \textit{pipeComp} from general workflow frameworks such as \textit{drake} \citep{landauDrakePackage2018}; indeed we estimate that the benchmarks discussed here would have taken approximately 16TB of storage had all non-redundant intermediates been saved (see the Additional File 2 for more detailed explanations of the framework and comparison to \textit{CellBench} and \textit{drake}).
 In addition, the fact that benchmark functions are stored in the \texttt{PipelineDefinition} makes the benchmark more smoothly portable, making it easy to modify, extend with further methods, or apply to other datasets.
@@
 Each step can accept a number of parameters whose alternative values are provided as a list.
-All (or subsets of) combinations of parameters can then be simultaneously ran and evaluated using the \texttt{runPipeline} function (see Supplementary Methods for more details).
+All (or subsets of) combinations of parameters can then be simultaneously ran and evaluated using the \texttt{runPipeline} function (see Additional File 2 for more details).
 \textbf{B:} Scheme representing the application of \textit{pipeComp} to evaluate a range of methods that are commonly used in scRNAseq studies and some of the metrics monitored at various steps.}

### 2020-08-04 bmc_article.tex
Updates from Overleaf

@@
 MI has the virtue of not decreasing when a true subpopulation is split into two clusters, which is arguably less problematic (and might well reflect unknown biological subgroups), but as a consequence it can be biased towards methods producing higher resolution clustering (Additional File 1: Fig S3).
-Similarly, precision per true subpopulation is considerably more robust to differences in the number of clusters.
-We also tracked the mean F1 score and the ARI at the true number of clusters.
+Similarly, precision per true subpopulation is considerably more robust to differences in the number of clusters, and we also tracked the ARI at the true number of clusters.
 
-The MI score and minimum precision, which are largely independent of the estimated number of clusters, were overall higher for the \textit{walktrap} method (Figure \ref{fig:clust}), while the mean F1 score favored both \textit{scran} methods (\textit{walktrap} and \textit{fast greedy}) over \textit{Seurat}.
-Finally, the ARI score at the true number of clusters, when available, showed similar performances.
-However, because \textit{Seurat}'s resolution parameter had a large impact on the number of clusters identified (Additional File 1: Fig S2 and 24), \textit{Seurat} could always be coerced into producing the right number of clusters.
+The MI score and minimum precision, which are largely independent of the estimated number of clusters, were overall higher for the \textit{walktrap} method (Figure \ref{fig:clust}A, left) over \textit{Seurat}, although when used in combination with some networks it sometimes yielded a lower ARI or minimum precision.
+The ARI score at the true number of clusters, when available, showed similar performances, especially when using \textit{sctransform}.
+Because \textit{Seurat}'s resolution parameter had a large impact on the number of clusters identified (Additional File 1: Fig S2 and 24), \textit{Seurat} could always be coerced into producing the right number of clusters.
 Instead, the number of clusters found by \textit{scran} was considerably less influenced by the available parameters (number of nearest neighbors or steps in the random walk - see Additional File 1: Fig S28), and as a result \textit{scran}-based clustering sometimes never produced a partitioning with the right number of clusters.
 This observation, along with \textit{scran}'s higher MI score, suggest that \textit{scran} sometimes simply divides a real subpopulation into two clusters (possibly tracking some unknown biological differences) rather than committing misclassification errors.
-Overall, the \textit{walktrap} method appeared superior to the \textit{fast greedy} algorithm and was generally less prone to misclassification than \textit{Seurat} clustering, although the latter offered more control over the resolution.
+Overall, the \textit{walktrap} method appeared superior to the \textit{fast greedy} algorithm and was generally comparable to \textit{Seurat} clustering, although the latter offered more control over the resolution.
 A major difference between \textit{walktrap}-based clustering and \textit{Seurat} is the computing time (Figure \ref{fig:clust}B).
@@
    \begin{itemize}
-    \item We found \textit{scran}-based \textit{walktrap} clustering to show good performances, at however increased computational costs which might be prohibitive with large datasets.
-    \item In cases where prior knowledge can guide the choice of a resolution, \textit{Seurat} can be useful in affording manual control of it while, in the absence of such knowledge, \textit{scran}-based \textit{walktrap} clustering provided reasonable estimates.
+    \item In cases where prior knowledge can guide the choice of a resolution, \textit{Seurat} can be useful in affording manual control of it while, in the absence of such knowledge, \textit{scran}-based \textit{walktrap} clustering provided reasonable estimates, although at greater computational costs.
    \end{itemize}
@@
 All wrappers for methods used in the study can be found in the external data of the \textit{pipeComp} package (\url{https://github.com/plger/pipeComp/tree/master/inst/extdata}).
+In addition, both the package and attending code are archived on \cite{figshare_code}.
 
@@
 For the other datasets, we used the unfiltered counts from \cite{duoClustering2018}, available on the corresponding repository \url{https://github.com/markrobinsonuzh/scRNAseq\_clustering\_comparison}.
-For simplicity, all starting \textit{SingleCellExperiment} objects with standardized metadata are available on \url{https://doi.org/10.6084/m9.figshare.11787210.v1}.
+For simplicity, all starting \textit{SingleCellExperiment} objects with standardized metadata are available on \cite{figshare_sce}.
 
@@
     \centering
-    \includegraphics[width=\textwidth, keepaspectratio]{{anova.pdf}}
+    \includegraphics[width=0.55\textwidth, keepaspectratio]{{anova.pdf}}
     \caption{\textbf{Variance in clustering metrics explained by the parameters.} See methods for more detail.}
@@
   \hline
-dataset & {\color{red}\# features} & \# cells & protocol & description \\ 
+dataset & \# features & \# cells & protocol & description \\ 
   \hline
@@
   \subsection*{\textbf{Additional file 1}}
-  Additional File 1: Fig S(PDF 16 MB)
+  Supplementary Figures (PDF 16 MB)
   \subsection*{\textbf{Additional file 2}}

### 2020-08-04 supplementary_figures/imputation.Rmd
updated supplementary figures, changing figure labels

@@
 
-# Supplementary Figure `r getFigNb(TRUE)`
+# Fig S`r getFigNb(TRUE)`
 
@@
 
-### Supplementary Figure `r getFigNb()`
+### Fig S`r getFigNb()`
 **Evaluation of imputation/denoising methods.** Average silhouette width (**A**) and clustering accuracy (**B**) with or without (indicated as _none_) application of a denoising/imputation method.

### 2020-08-04 supplementary_figures/supp1_dataset_description.Rmd
updated supplementary figures, changing figure labels

@@
 
-### Supplementary Figure `r getFigNb()`
+### Fig S`r getFigNb()`
 The total counts and total features per cell of doublets (red) versus other cells.
@@
 
-### Supplementary Figure `r getFigNb()`
+### Fig S`r getFigNb()`
 Distribution across cells of various control properties in the different datasets. The 
@@
 
-### Supplementary Figure `r getFigNb()`
+### Fig S`r getFigNb()`
 Relationship between selected cell-level QC metrics.
@@
 
-### Supplementary Figure `r getFigNb()`
+### Fig S`r getFigNb()`
 There is a tight relationship, in 10x datasets (i.e.
@@
 
-### Supplementary Figure `r getFigNb()`
+### Fig S`r getFigNb()`
 Relationship between various cellular properties and the frequency of cluster mis-assignment for the mixology10x3cl (A) and mixology10x5cl (B) datasets.
@@
 
-### Supplementary Figure `r getFigNb()`
+### Fig S`r getFigNb()`
 Relationship between various cellular properties and the frequency of cluster mis-assignment for the Zheng equal (A) or unequal (B) mixtures of four cell types.
@@
 
-### Supplementary Figure `r getFigNb()`
+### Fig S`r getFigNb()`
 Relationship between various cellular properties and the frequency of cluster mis-assignment for the Zheng mixture of 8 cell types. See Supplementary Figure `r getFigNb()-2` for more information.

### 2020-08-04 supplementary_figures/supp1b_metrics.Rmd
updated supplementary figures, changing figure labels

@@
 
-### Supplementary Figure `r getFigNb()`
+### Fig S`r getFigNb()`
 The number of clusters called has a much bigger impact on the Adjusted Rand Index (ARI) than differences between methods. The dashed line indicates the true number of clusters.
 
-# Supplementary Figure `r getFigNb(TRUE)`
+# Fig S`r getFigNb(TRUE)`
 
@@
 
-### Supplementary Figure `r getFigNb()`
+### Fig S`r getFigNb()`
 Relationship of various metrics of clustering accuracy between each other and with variations in the number of clusters called (`nbClust.diff` and `nbClust.absDiff`).

### 2020-08-04 supplementary_figures/supp2_filtering.Rmd
updated supplementary figures, changing figure labels

@@
 
-### Supplementary Figure `r getFigNb()`
+### Fig S`r getFigNb()`
 Mean clustering F1 score per subpopulation, mean F1 at true number of clusters, as well as maximum and median proportion of excluded cells per subpopulation across various filtering strategies.
@@
 
-### Supplementary Figure `r getFigNb()`
+### Fig S`r getFigNb()`
 

### 2020-08-04 supplementary_figures/supp3_normalization.Rmd
updated supplementary figures, changing figure labels

@@
 
-### Supplementary Figure `r getFigNb()`
+### Fig S`r getFigNb()`
 Mean per-subpopulation absolute correlation of the first 5 components with library size (left) and the number of detected features (right) across normalization procedures.
@@
 
-### Supplementary Figure `r getFigNb()`
+### Fig S`r getFigNb()`
 Mean difference between the number of detected clusters and the number of real subpopulations, depending on the normalization method, the resolution and the number of dimensions used.
@@
 
-### Supplementary Figure `r getFigNb()`
+### Fig S`r getFigNb()`
 Relationship of the variance with mean count after `sctranform`'s variance stabilizing transformation.
@@
 
-### Supplementary Figure `r getFigNb()`
+### Fig S`r getFigNb()`
 Running time of the normalization methods.

### 2020-08-04 supplementary_figures/supp4_selection.Rmd
updated supplementary figures, changing figure labels

@@
 
-### Supplementary Figure `r getFigNb()`
+### Fig S`r getFigNb()`
 **A:** Comparison of the gene-wise proportion of variance explained by real subpopulations based on Seurat's standard log normalization and on `sctransform` variance-stabilizing transformation.

### 2020-08-04 supplementary_figures/supp5_dr.Rmd
updated supplementary figures, changing figure labels

@@
 
-### Supplementary Figure `r getFigNb()`
+### Fig S`r getFigNb()`
 **Evaluation of common dimensionality reduction methods.
@@
 
-### Supplementary Figure `r getFigNb()`
+### Fig S`r getFigNb()`
 Mean difference between the number of detected clusters and the number of real subpopulations, depending on the resolution and number of dimensions used.
@@
 
-### Supplementary Figure `r getFigNb()`
+### Fig S`r getFigNb()`
 Estimates of dimensionality by the `intrinsicDimension::maxLikGlobalDimEst` method using various 'reasonable' numbers of nearest neighbors (`k` parameter).

### 2020-08-04 supplementary_figures/supp6_clust.Rmd
updated supplementary figures, changing figure labels

@@
 
-### Supplementary Figure `r getFigNb()`
+### Fig S`r getFigNb()`
 Difference between the number of detected clusters and the number of real subpopulations according to different clustering paramters.

### 2020-08-04 supplementary_figures/supp_figures.Rmd
added missing supp figures
some figure updates
updated supplementary figures, changing figure labels

@@
 title: \vspace{2cm}\Large{pipeComp, a general framework for the evaluation of computational pipelines, reveals performant single-cell RNA-seq preprocessing tools}
-subtitle: \vspace{2cm}\LARGE{Supplementary Figures}\vspace{2cm}
+subtitle: \vspace{2cm}\LARGE{Additional File 1 - Supplementary Figures}\vspace{2cm}
 date: "`r format(Sys.time(), '%d %B, %Y')`"
@@
 
-\renewcommand{\thepage}{S\arabic{page}}
-
 ```{r setup, include=FALSE}
@@
 
-### Supplementary Figure `r getFigNb()`: Overview of the benchmark datasets
+### Fig S`r getFigNb()`: Overview of the benchmark datasets
 
@@
 
-```{r child='supp1b_metrics.Rmd'}
+```{r child='supp1b_metrics.Rmd', cache=TRUE}
 ```
 
-```{r child='supp1_dataset_description.Rmd'}
+```{r child='supp1_dataset_description.Rmd', cache=TRUE}
 ```
 
-```{r child='supp2_filtering.Rmd'}
+```{r child='supp2_filtering.Rmd', cache=TRUE}
 ```
 
-```{r child='supp3_normalization.Rmd'}
+```{r child='supp3_normalization.Rmd', cache=TRUE}
 ```
 
-```{r child='supp4_selection.Rmd'}
+```{r child='supp4_selection.Rmd', cache=TRUE}
 ```
 
-```{r child='supp5_dr.Rmd'}
+```{r child='supp5_dr.Rmd', cache=TRUE}
 ```
@@
 
+```{r child='supp_overall.Rmd'}
+```
+
+```{r child='imputation.Rmd'}
+```
+
 ```{r child='supp_svadea.Rmd'}
@@
 
-### Supplementary Figure `r getFigNb()`: Summary of the recommendations.
+### Fig S`r getFigNb()`: Summary of the recommendations.

### 2020-08-04 supplementary_figures/supp_framework/supp_framework_explanations.Rmd
updated supplementary figures, changing figure labels

@@
 ---
-title: "pipeComp - Supplementary Methods"
+title: "pipeComp - Additional file 2: Supplementary Methods"
 author: Pierre-Luc Germain, Anthony Sonrel, Mark D. Robinson
@@
 
-On the other hand, a `r CRANpkg("drake")`-based workflow has several advantages, in particular the capacity, upon failure of a pipeline (or changes made to a step), to recover and restart from the very next upstream target.
-We are currently working on an `bptry`-based approach to error handling and selective recomputation, but for the moment users requiring smooth error recovery should rather rely on `r CRANpkg("drake")`.
+On the other hand, a `r CRANpkg("drake")`-based workflow has several advantages, in particular the capacity, upon failure of a pipeline (or changes made to a step), to smoothly recover and restart from the very next upstream target.
+Starting from version 0.99.45, `r Rpackage("pipeComp")` also supports some error handling, allowing users to use results from those pipelines that ran successfully and selectively recompute (e.g.
+after fixing the error) those that encountered errors (see the package's vignettes for more information).
+However, due to the non-persistent nature of `r Rpackage("pipeComp")`'s storage, for those parameter combinations that failed, all steps will have to be recomputed.
 

### 2020-08-04 supplementary_figures/supp_overall.Rmd
updated supplementary figures, changing figure labels

--- 2020-06-29 supplementary_figures/supp_overall.Rmd
@@
 date: "`r format(Sys.time(), '%d %B, %Y')`"
-output: html_document
+output: pdf_document
 ---
@@
   library(ggplot2)
+  library(ggrepel)
   library(cowplot)
@@
 
-# Supplementary Figure `r getFigNb(TRUE)`
+# Fig S`r getFigNb(TRUE)`
 
-```{r, fig.width=6, fig.height=5}
+```{r, fig.align="center", echo=FALSE, out.height="0.95\\textheight"}
+knitr::include_graphics("../main_figures/overall.pdf")
+```
+
+### Fig S`r getFigNb()`
+
+**Overview of the combination of main alternative parameters.** Unsupervised clustering of the results (above) and top results (below).
+The color-mapping schemes are the same as described in the main figures.
+
+\newpage
+
+# Fig S`r getFigNb(TRUE)`
+
+```{r, fig.width=6, fig.height=4}
 ov <- readRDS("../data/overall.rds")
-
 cl <- ov$evaluation$clustering
@@
 cl$clustmethod <- relevel(cl$clustmethod, "clust.seurat")
-cl$clustmethod <- relevel(cl$clustmethod, "clust.seurat")
 
-ff <- function(x, trueK=FALSE){
-  f <- paste0(x,"~dataset+resolution*dataset+doubletmethod+sel+filt+norm+clustmethod+dims")
-  if(trueK) cl <- cl[cl$n_clus==cl$true.nbClusts,]
-  mod <- lm(as.formula(f), data=cl)
+# checking for interactions:
+mod <- lm(ARI~resolution*dataset+doubletmethod*sel*filt*norm*clustmethod*dims, data=cl)
 co <- as.data.frame(coef(summary(mod)))
-  co <- co[grep("sctransfrom|scran|filt|elbow|doublet|Annoy", row.names(co)),]
-  colnames(co)[2:4] <- c("stderr","t","pvalue")
-  co$coef <- gsub(" +$","",row.names(co))
-  aov <- as.data.frame(summary(aov(mod))[[1]])[,-1]
-  aov <- aov[grep("^Residuals|resolution|dataset", row.names(aov), invert=TRUE),]
-  aov$varExpl <- aov[,1]/sum(aov[,1])
-  aov <- aov[grep("^Residuals", row.names(aov), invert=TRUE),4:5]
-  colnames(aov)[1] <- "pvalue"
-  aov$parameter <- gsub(" +$","",row.names(aov))
-  list( coefs=co, aov=aov )
-}
-
-cos <- lapply(c(MI="MI",ARI="ARI",minPre="min_pr"), FUN=ff)
-aov <- dplyr::bind_rows(lapply(cos, FUN=function(x) x$aov), .id="metric")
-
-cos2 <- lapply(c(ARI="ARI"), trueK=TRUE, FUN=ff)
-aov2 <- dplyr::bind_rows(lapply(cos2, FUN=function(x) x$aov), .id="metric")
-p1 <- ggplot(aov, aes(reorder(parameter, -varExpl), varExpl, group=parameter, fill=metric)) + 
-  geom_col() + xlab("") + ylab("Proportion of variance explained") + coord_flip()
-p2 <- ggplot(aov2, aes(reorder(parameter, -varExpl), varExpl, group=parameter, fill=metric)) + 
-  geom_col() + xlab("") + ylab("Variance explained") + theme(legend.position="none") + 
-  coord_flip() + ggtitle("ARI at true k")
-
-co <- dplyr::bind_rows(lapply(cos, FUN=function(x) x$coef), .id="metric")
-co2 <- dplyr::bind_rows(lapply(cos2, FUN=function(x) x$coef), .id="metric")
-for(f in c("norm","filt","doubletmethod","dims","sel","clustmethod")){
-  co$coef <- gsub(paste0("^",f),"",co$coef)
-  co2$coef <- gsub(paste0("^",f),"",co2$coef)
-}
-p3 <- ggplot(co, aes(coef, Estimate, group=coef, fill=metric)) + geom_col() + xlab("") +
-  ylab("Estimated coefficient") + geom_hline(yintercept = 0, linetype="dashed") + 
-  coord_flip() + theme(legend.position="none")
-p4 <- ggplot(co2, aes(coef, Estimate, group=coef, fill=metric)) + geom_col() + xlab("") +
-  ylab("Estimated coefficient") + geom_hline(yintercept = 0, linetype="dashed") + 
-  coord_flip() + theme(legend.position="none")
-
-# plot_grid( plot_grid(p1,p2,rel_widths=c(4,3),labels=LETTERS[1:2]),
-#            plot_grid(p3,p4,rel_widths=c(4,3),labels=LETTERS[3:4]),
-#            nrow=2, rel_heights=c(3,4))
-plot_grid(p1 + ylab("Proportion of variance in metric explained by the parameter"),p2,nrow=2,labels=LETTERS[1:2])
+coi <- co[grep(":",row.names(co)),]
+coi <- coi[grep("dataset", row.names(coi), invert=TRUE),]
+colnames(coi) <- c("Estimate","SE","t","p.value")
+coi <- coi[order(coi$p.value),]
+coi$FDR <- p.adjust(coi$p.value, method="fdr")
+row.names(coi) <- gsub("selsel","sel",gsub("normnorm","norm",row.names(coi)))
+coi$interaction <- gsub(":"," +\n", row.names(coi))
+coi2 <- coi[coi$FDR < 0.05,]
+ggplot(coi, aes(Estimate, -log10(p.value), label=interaction)) + geom_point() + 
+  geom_point(data=coi2, colour="red") + geom_label_repel(data=coi2) + xlab("ARI coefficient") +
+  ggtitle("Significant interactions between parameter values")
 ```
 
-### Supplementary Figure `r getFigNb()`
+### Fig S`r getFigNb()`
 
-**A:** Proportion of the variance in clustering accuracy metrics explained by main parameters.
-**B:** Proportion of the variance in Adjusted Rand Index (ARI) explained by the parameters, among the analyses giving the true number of clusters.
-The ANOVA were performed with the model
+Estimated coefficient and -log10 significance of all interaction coefficients in the linear model defined by
 
-`metric ~ dataset*resolution+doubletmethod+filt+sel+norm+clustmethod+dims`
+`ARI~dataset*resolution+doubletmethod*sel*filt*norm*clustmethod*dims`
 
-In both panels, the proportion is calculated after substracting the variance attributable to differences due to the dataset, the resolution and their interaction.
-The alternatives included in the analysis (other than the default ones described in the methods) were:
-```
-list( doubletmethod=c("none","doublet.scDblFinder"),
-      filt=c("none","filt.default","filt.stringent"),
-      norm=c("norm.seurat","norm.sctransform","norm.scran"),
-      sel=c("sel.vst","sel.deviance","sel.expr"),
-      selnb=c(2000,4000),
-      dr=c("seurat.pca"),
-      dims=c("elbow", 10, "maxLikGlobal"),
-      clustmethod=c("clust.seurat","clust.scran.knnAnnoy") )
+Those terms significant at a FDR < 0.05 are highlighted. The top interaction terms are:
+
+```{r}
+knitr::kable(coi[1:10,c(1,2,4,5)], digits=3)
 ```
@@
 }
-
-H1 <- renameHrows( scrna_evalPlot_silh(ov, agg.by = c("doubletmethod","filt","norm"), row_split = rnfn(norm),
-                                       heatmap_legend_param=list(direction="horizontal"),font_factor = 0.8,
-                                       row_names_max_width=unit(7, "cm")), rnfn)
-H2 <- renameHrows(
-     evalHeatmap(ov, step="clustering", what=c("MI","min_pr"), anno_legend=FALSE, title=c("MI","Min precision"), 
-                 agg.by=c("doubletmethod","filt","norm"), font_factor = 0.8, 
-                 row_split=rnfn(norm),row_names_max_width=unit(7, "cm")) + 
-    evalHeatmap(ov, step="clustering", what=c("ARI"), agg.by=c("doubletmethod","filt","norm"), 
-                row_split=rnfn(norm),filter=n_clus==true.nbClusts, title="ARI at true k",
-                anno_legend=FALSE, font_factor = 0.8, row_names_max_width=unit(7, "cm")),
-  rnfn)
 
@@
 ```{r, fig.width=9, fig.height=8.5}
-plot_grid( grid.grabExpr(draw(H1, heatmap_legend_side="bottom")), 
-           grid.grabExpr(draw(H2)), nrow=2, labels=LETTERS[1:2])
-```
-
-### Supplementary Figure `r getFigNb()`
-
-Silhouette widths of the real subpopulations (**A**) and clustering accuracy (**B**) using different combinations of methods.
-
-\newpage
-
-# Supplementary Figure `r getFigNb(TRUE)`
-
-```{r, fig.width=9, fig.height=8.5}
 plot_grid( grid.grabExpr(draw(H3, heatmap_legend_side="bottom")), 

### 2020-08-04 supplementary_figures/supp_svadea.Rmd
updated supplementary figures, changing figure labels

@@
 
-### Supplementary Figure `r getFigNb()`
+### Fig S`r getFigNb()`
 Accuracy of the differential expression analysis across combinations of: **A:** filters and DEA methods (using max 1 variable), **B:** filters and SVA methods (using max 1 variable), **C:** SVA methods (using 2 variables).
@@
 
-### Supplementary Figure `r getFigNb()`
+### Fig S`r getFigNb()`
 **A:** Accuracy of the estimated logFC (correlation and median absolute deviation from expected logFCs) and of the differential expression analysis (TPR stands for True Positive Rate, and FDR for False Discovery Rate) across the different combinations of SVA and DEA methods (using max 1 dimension).

### 2020-08-04 main_figures/svadea.Rmd
simplified sva-dea figure

@@
   x[which(x$k==1),]
+})
+res$evaluation$dea <- lapply(res$evaluation$dea, FUN=function(x){
+  levels(x$k) <- c(levels(x$k), "0")
+  x$k <- relevel(x$k, "0")
+  x$k[x$sva.method=="none"] <- 0
+  x
 })
@@
 
+res <- lapply(res$evaluation$dea, FUN=function(x){
+  x[which(x$k %in% as.character(0:5)),]
+})
+
+cols <- c("#00204DFF", "#31446BFF", "#666970FF", "#958F78FF", "#CBBA69FF", "#FFEA46FF")
+names(cols) <- 0:5
+
 p <- plot_grid(
   plot_grid(
-    dea_evalPlot_curve(res1, agg.by=c("sva.method","dea.method"), colourBy="dea.method", shapeBy="sva.method") + ggtitle("Using a single surrogate variable"),
-    dea_evalPlot_curve(res, agg.by=c("sva.method","k"), colourBy="k", shapeBy="sva.method") + ggtitle("Increasing the number of surrogate variables"),
+    dea_evalPlot_curve(res1, agg.by=c("sva.method"), colourBy="sva.method", shapeBy="sva.method") + ggtitle("Using a single surrogate variable"),
+    dea_evalPlot_curve(res, agg.by=c("sva.method","k"), colourBy="k", shapeBy="sva.method") + 
+      scale_colour_manual(values=cols) + ggtitle("Increasing the number of surrogate variables"),
     nrow=2, labels=LETTERS[1:2]

### 2020-08-04 supplementary_figures/supp_svadea.Rmd
updated sva-dea supp

@@
 theme_set(theme_cowplot(font_size = 12))
-res2 <- res1 <- res <- readRDS("../svadea/results/aggregated.rds")
-res1$evaluation$dea <- lapply(res1$evaluation$dea, FUN=function(x){
-  x[which(x$k==1),]
+res12 <- res012 <- res0 <- res1 <- res <- readRDS("../svadea/results/aggregated.rds")
+res0$evaluation$dea <- lapply(res$evaluation$dea, FUN=function(x){
+  x[which(x$sva.method=="none"),]
 })
-res2$evaluation$dea <- lapply(res2$evaluation$dea, FUN=function(x){
-  x[which(x$k==2 | x$sva.method=="none"),]
+res1$evaluation$dea <- lapply(res$evaluation$dea, FUN=function(x){
+  x[which(x$k==1 & x$sva.method!="none"),]
+})
+res012$evaluation$dea <- lapply(res$evaluation$dea, FUN=function(x){
+  x$k[x$sva.method=="none"] <- NA
+  x[which(is.na(x$k) | x$k==2 | x$k==1),]
+})
+res12$evaluation$dea <- lapply(res$evaluation$dea, FUN=function(x){
+  x$k[x$sva.method=="none"] <- NA
+  x[which(x$k==2 | x$k==1),]
 })
@@
 plot_grid(
-  dea_evalPlot_curve(res1, agg.by=c("dea.method","filt","minCount"), shapeBy = "dea.method", colourBy = c("filter")) + ggtitle("Using 1 SV"),
-  dea_evalPlot_curve(res1, agg.by=c("sva.method","filt","minCount"), shapeBy = "sva.method", colourBy = c("filter")) + ggtitle("Using 1 SV"),
-  dea_evalPlot_curve(res2, agg.by=c("sva.method")) + ggtitle("Using 2 SVs"),
+  dea_evalPlot_curve(res0, agg.by=c("dea.method","filt","minCount"), shapeBy = "dea.method", colourBy = c("filter")),
+  dea_evalPlot_curve(res12, agg.by=c("dea.method","filt","minCount"), shapeBy = "dea.method", colourBy = c("filter")),
+  dea_evalPlot_curve(res012, agg.by=c("dea.method","sva.method"), shapeBy="dea.method", colourBy = c("sva.method")),
   nrow=3, labels=LETTERS[1:3]
@@
 ### Fig S`r getFigNb()`
-Accuracy of the differential expression analysis across combinations of: **A:** filters and DEA methods (using max 1 variable), **B:** filters and SVA methods (using max 1 variable), **C:** SVA methods (using 2 variables).
-In **C**, filled points indicate a nominal FDR below the observed FDR.
+Accuracy of the differential expression analysis across combinations of: **A:** filters and DEA methods (without any SVA-step), **B:** filters and DEA methods (average across the different SVA strategies using either 1 or 2 surrogate variables), **C:** DEA and SVA methods (using one or two surrogate variables).
 

### 2020-08-04 main_figures/dimensionality.Rmd
forgotten commit...

@@
 
+res$evaluation$clustering <- res$evaluation$clustering[!(res$evaluation$clustering$dims %in% c("10","30","essLocal.b")),]
+
 p3 <- renameHrows(
-  evalHeatmap( res, what=c("NMI","min_pr"), agg.by=c("norm","dims"), 
-               filter=sel=="sel.vst" & !(dims %in% c("10","30","essLocal.b")),
+  evalHeatmap( res, what=c("MI","ARI"), agg.by=c("norm","dims"), 
                    row_split=gsub("norm\\.","",norm), reorder_rows=TRUE, anno_legend=FALSE,
@@
   evalHeatmap(res, what=c("ARI"), agg.by=c("norm","dims"), row_split=gsub("norm\\.","",norm), 
-              filterExpr=n_clus==true.nbClusts & sel=="sel.vst" & !(dims %in%                                                                      c("10","30","essLocal.b")), row_names_max_width = unit(6.2, "cm"),
+              row_names_max_width = unit(6.2, "cm"),
+              filterExpr = n_clus==true.nbClusts,
               reorder_rows=TRUE, anno_legend=FALSE, title="ARI at\ntrue k"),
-  rnfn
+  function(x){
+    x <- rnfn(sapply(strsplit(x," > "), FUN=function(x) rev(x)[1]))
+  }
 )

### 2020-08-04 bmc_article.tex

@@
 We developed 3 benchmark datasets: 1) the \texttt{ipsc} data is a random selection of 10 vs 10 samples from \citep{carcamoOriveAnalysisTranscriptional2017}, which are very heterogeneous, and foldchanges were manually added to 300 genes; 2) the \texttt{seqc} dataset contains 5 vs 5 samples of two mixtures of the SEQC consortium with different ERCC spike-in mixes \citep{ComprehensiveAssessment2014}, and includes a partially correlated batch effect; 3) A \texttt{simulation} of 8 vs 8 samples with two technical vectors of variation (see Methods for more details).
-As previously reported \citep{ComprehensiveAssessment2014}, we found that more stringent filtering improved DEA (Additional File 1: Fig S33A) and that the different DEA methods had similar performances (Figure \ref{fig:svadea}A and Additional File 1: Fig S33A).
-Including a SVA-based step increased sensitivity across all datasets, particularly in datasets with technical vectors of variation (\textit{seqc} and \textit{simulation}), and benefited all DEA methods in a similar fashion (Figure \ref{fig:svadea}A,C and Additional File 1: Fig S33-34).
+As previously reported \citep{ComprehensiveAssessment2014}, we found that more stringent filtering improved DEA (Additional File 1: Fig S33A-B) and that the different DEA methods had similar performances, especially after using SVA-based methods (Additional File 1: Fig S33B).
+Including a SVA-based step increased sensitivity across all datasets (Figure \ref{fig:svadea}A and Additional File 1: Fig S33B), particularly in datasets with technical vectors of variation (\textit{seqc} and \textit{simulation}), and benefited all DEA methods in a similar fashion (Figure \ref{fig:svadea}C and Additional File 1: Fig S33C-34A).
 With the exception of the \texttt{RUVr} method, which led to a high False Discovery Rate (FDR), all other methods showed comparable gains in sensitivity while maintaining a decent error control (if a little above the nominal FDR), even when increasing the number of surrogate variables used (Figure \ref{fig:svadea}B-C).
@@
 \textbf{B:} Relationship between the maximum subpopulation exclusion rate and the average clustering accuracy per subpopulation across various filtering strategies.
-Of note, doublet removal appears to be desirable even when, due to the design, there are no heterotypic doublets in the data.
+Of note, doublet removal appears to have a neutral or positive impact even when, due to the design, there are no heterotypic doublets in the data.
 The PCA methods refer to multivariate outlier detected as implemented in \textit{scater} (see Methods for details).}
@@
     \caption{{\color{red}\textbf{Evaluation of the impact of surrogate variable analysis (SVA) on (bulk) RNAseq differential expression analysis using \textit{pipeComp}.} \textit{RUVr} and \textit{RUVs} refer to the two \textit{RUVSeq} functions by the same name; \textit{svaseq} refers to the function of the same name in \textit{sva}, while \textit{vstsva} refers to the original \textit{sva} method applied no variance-stabilized data.
-\textbf{A:} True positive rates (TPR) and False discovery rates (FDR) of various combinations of SVA-based methods (using a single surrogate variable) and DEA methods.
-\textbf{B:} TPR and FDR of the different SVA-based methods (averaged across the different DEA methods) using increasing numbers of variables (`be' and `leek` refer to the two methods implemented in the \textit{sva} package for estimating the number of surrogate variables).
+\textbf{A:} True positive rates (TPR) and False discovery rates (FDR) of the SVA-based methods (using a single surrogate variable), averaged across DEA methods.
+\textbf{B:} TPR and FDR of the different SVA-based methods (averaged across the different DEA methods) using increasing numbers of variables.
 In \textbf{A-B}, the x-axis is square-root transformed to improve readability, and three dots for each combination of methods refer to the nominal FDR thresholds of 0.01, 0.05, and 0.1.
 \textbf{C:} Correlation of estimated logFC with expected ones, as well as TPR and FDR (at nominal FDR threshold of 0.05) of the different methods with increasing numbers of variables.
-As in \ref{fig:norm}C, the color mapping tracks the number of (matrix-wise) Median Absolute Deviations from the (column-wise) median, while the printed numbers represent the raw metric values.}}
+As in \ref{fig:norm}C, the color mapping tracks the number of (matrix-wise) Median Absolute Deviations from the (column-wise) median, while the printed numbers represent the raw metric values.
+`be' and `leek` refer to the two methods implemented in the \textit{sva} package for estimating the number of surrogate variables.}}
     \label{fig:svadea}

### 2020-08-06 bmc_article.tex
Updates from Overleaf

@@
 \begin{abstract}
-
-We present \textit{pipeComp} (\url{https://github.com/plger/pipeComp}), a flexible R framework for pipeline comparison handling interactions between analysis steps and relying on multi-level evaluation metrics.
+{\color{red} We present \textit{pipeComp} (\url{https://github.com/plger/pipeComp}), a flexible R framework for pipeline comparison handling interactions between analysis steps and relying on multi-level evaluation metrics.
 We apply it to the benchmark of single-cell RNA-sequencing analysis pipelines using simulated and real datasets with known cell identities, covering common methods of filtering, doublet detection, normalization, feature selection, denoising, dimensionality reduction and clustering.
-\textit{pipeComp} can easily integrate any other step, tool, or evaluation metric, allowing extensible benchmarks and easy applications to other fields, as we demonstrate through a study of the impact of removal of unwanted variation on differential expression analysis.
+\textit{pipeComp} can easily integrate any other step, tool, or evaluation metric, allowing extensible benchmarks and easy applications to other fields, as we demonstrate through a study of the impact of removal of unwanted variation on differential expression analysis.}
 
@@
 Here, we develop a flexible R framework for pipeline comparison and apply it to the evaluation of the various steps of analysis leading from an initial count matrix to a cluster assignment, which are critical in a wide range of applications.
-We collected real datasets of known cell composition (Table S\ref{tab:table1}) and used a variety of evaluation metrics to investigate in a multilevel fashion the impact of various parameters and variations around a core scRNAseq pipeline.
+We collected real datasets of known cell composition ({\color{red}Table S\ref{tab:table1}}) and used a variety of evaluation metrics to investigate in a multilevel fashion the impact of various parameters and variations around a core scRNAseq pipeline.
 Although we use some datasets based on other protocols, our focus is especially on droplet-based datasets that do not include exogenous control RNA (i.e.
-spike-ins); see Table S\ref{tab:table1} and Additional File 1: Fig S1 for more details.
+spike-ins); see {\color{red}Table S\ref{tab:table1}} and Additional File 1: Fig S1 for more details.
 In addition to previously-used benchmark datasets with true cell labels \cite{duoClustering2018,tianMixology2018}, we simulated two datasets with a hierarchical subpopulation structure based on real 10x human and mouse data using \textit{muscat} \cite{CrowellMuscat2019}.
@@
 
-The \textit{pipeComp} package defines a pipeline as, minimally, a list of functions executed consecutively on the output of the previous one (Figure \ref{fig:explanation}A; see also Additional File 2).
+The \textit{pipeComp} package defines a pipeline as, minimally, a list of functions executed consecutively on the output of the previous one (Figure \ref{fig:explanation}A; {\color{red}see also Additional File 2}).
 In addition, optional benchmark functions can be set for each step to provide standardized, multi-layered evaluation metrics.
-Given such a \texttt{PipelineDefinition} object, a set of alternative parameters (which might include different subroutines calling different methods) and benchmark datasets, the \texttt{runPipeline} function then proceeds through all combinations of arguments (or a desired subset), avoiding recomputing the same step twice without the need to save all alternative intermediates, and compiling evaluations (including run time) on the fly (see Additional File 2 for more detail).
+Given such a \texttt{PipelineDefinition} object, a set of alternative parameters (which might include different subroutines calling different methods) and benchmark datasets, the \texttt{runPipeline} function then proceeds through all combinations of arguments (or a desired subset), avoiding recomputing the same step twice without the need to save all alternative intermediates, and compiling evaluations (including run time) on the fly ({\color{red}see Additional File 2 for more details}).
 Variations in a given parameter can be evaluated using all metrics from this point downward in the pipeline.
@@
 Across datasets, cells called as doublets tended to be classified in the wrong cluster more often than other cells (Figure \ref{fig:doublets}C).
-We also found that \textit{scDblFinder} improved the accuracy of the clustering across all benchmark datasets even when, by design, the data contained no heterotypic doublet (Figure \ref{fig:filtering}).
+We also found that {\color{red} \textit{scDblFinder} clearly improved the accuracy of the clustering across datasets that are expected to have heterotypic doublets (mixology datasets), and even in a simulation (simMix1) that should not contain any (Figure \ref{fig:filtering}).
+Across the datasets that, being FACS-sorted, should not contain such doublets, doublet removal tended not to robustly impact clustering accuracy, however in few datasets (ZhengMix4eq and Zhengmix4uneq, and simMix2 with a very small difference) doublet removal performed worse when combined with lenient or no filtering.}
 
@@
 Reasoning that the cells we wish to exclude are the cells that would be misclassified, we measured the rate of misclassification of each cell in each dataset across a variety of clustering pipelines, correcting for the median misclassification rate of the subpopulation and then evaluated what properties could be predictive of misclassification (Additional File 1: Fig S8-10).
-We could not identify any property or simple combination thereof that would be consistently predictive of misclassification; the only feature that stood out across more than one dataset (the Zheng datasets) was that cells with very high read counts have a higher chance of being misclassified{\color{red}, although such a pattern was not observed in the other datasets.}
+We could not identify any property or simple combination thereof that would be consistently predictive of misclassification; the only feature that stood out across more than one dataset (the Zheng datasets) was that cells with very high read counts have a higher chance of being misclassified, although such a pattern was not observed in the other datasets.
 
@@
 We next investigated the impact of different normalization strategies.
-Beside the standard log-normalization included in \textit{Seurat}, we tested \textit{scran}'s pooling-based normalization \cite{lunPooling2016}, \textit{sctransform}'s variance-stabilizing transformation \cite{hafemeisterSCtransform2019}, normalization based on stable genes \cite{linStableGenes2018, deekeStablyExpressed2018}, and \textit{SCnorm} \cite{bacherSCnorm2017} (\textit{scVI}-based normalization \cite{lopezDeep2018} was included separately{\color{red}, as it was not meant for this purpose and follows a slightly different flow}).
+Beside the standard log-normalization included in \textit{Seurat}, we tested \textit{scran}'s pooling-based normalization \cite{lunPooling2016}, \textit{sctransform}'s variance-stabilizing transformation \cite{hafemeisterSCtransform2019}, normalization based on stable genes \cite{linStableGenes2018, deekeStablyExpressed2018}, and \textit{SCnorm} \cite{bacherSCnorm2017} (\textit{scVI}-based normalization \cite{lopezDeep2018} was included separately, as it was not meant for this purpose and follows a slightly different flow).
 In addition to log-normalization, the standard \textit{Seurat} clustering pipeline performs per-feature unit-variance scaling so that the PCA is not too strongly dominated by highly-expressed features.
@@
 Zhengmix8eq and simMix1) and subpopulations were consistently more challenging across all methods.
-{\color{red}An exception was \textit{scVI}-based normalization, which performed similarly to other methods on some datasets (Kumar, simMix2, Zhengmix4eq, Zhengmix4uneq), but very poorly on others, in terms of both the subpopulation silhouette and clustering accuracy (Additional File 1: Fig S13; of note, these values were not meant to be used for clustering, and the usage of the \textit{scVI} latent space will be discussed below along with dimensionality reductions)}.
+An exception was \textit{scVI}-based normalization, which performed similarly to other methods on some datasets (Kumar, simMix2, Zhengmix4eq, Zhengmix4uneq), but very poorly on others, in terms of both the subpopulation silhouette and clustering accuracy (Additional File 1: Fig S13; of note, these values were not meant to be used for clustering, and the usage of the \textit{scVI} latent space will be discussed below along with dimensionality reductions).
 Scaling (e.g.
@@
 GLM-PCA tended to increase the average silhouette width of already well-defined subpopulations, but \textit{Seurat}'s PCA procedure however proved superior on all metrics.
-{\color{red} Like GLM-PCA, \textit{scVI}'s Linearly Decoded (LD) data \cite{SvenssonLDVAE2020} and latent space do not explicitly rely on normalized counts.
+Like GLM-PCA, \textit{scVI}'s Linearly Decoded (LD) data \cite{SvenssonLDVAE2020} and latent space do not explicitly rely on normalized counts.
 Their performance was however worse than \textit{Seurat}'s PCA in view of the silhouette width (Additional File 1: Fig S15).
-In terms of clustering accuracy, \textit{scVI}'s latent space performed better that its LD but was nevertheless less accurate than PCA-based alternatives on the datasets with a very low amount of cells (Koh and Kumar) and harboring many subpopulations (simMix1 and Zhengmix8eq).}
+In terms of clustering accuracy, \textit{scVI}'s latent space performed better that its LD but was nevertheless less accurate than PCA-based alternatives on the datasets with a very low amount of cells (Koh and Kumar) and harboring many subpopulations (simMix1 and Zhengmix8eq).
 
@@
 We evaluated the performance of dimensionality estimators implemented in the \textit{intrinsicDimension} package \cite{johnssonDimensionality2015}, as well as common procedures such as the `elbow` method, some scRNAseq-specific methods such as the JackStraw procedure \cite{ChungJackstraw2015} or \textit{scran}'s \texttt{denoisePCA} \cite{LunScran2016}, and the recent application of Fisher Separability analysis \cite{AlberganteSepar2019}.
-{\color{red} We also included a variation of the molecular cross-validation (MCVR, modified to be applied on normalized data - see methods) approach originally developed to select an calibrate denoising methods \cite{batsonMolecularCrossValidationSingleCell2019}, but which has also been applied to estimate dimensionality \citep{wagnerMonetOpensource2020}.}
+We also included a variation of the molecular cross-validation (MCVR, modified to be applied on normalized data - see methods) approach originally developed to select an calibrate denoising methods \cite{batsonMolecularCrossValidationSingleCell2019}, but which has also been applied to estimate dimensionality \citep{wagnerMonetOpensource2020}.
 
@@
 
-{\color{red} \textit{MCVR} and the global maximum likelihood based on translated Poisson Mixture Model (\textit{maxLikGlobal}) tended to produce clusterings with smaller deviations to the true number of subpopulations (Additional File 1: Fig S26).
+\textit{MCVR} and the global maximum likelihood based on translated Poisson Mixture Model (\textit{maxLikGlobal}) tended to produce clusterings with smaller deviations to the true number of subpopulations (Additional File 1: Fig S26).
 Although most methods performed well on the various clustering measures, \textit{maxLikGlobal} (using 10 or 20 nearest neighbors, although estimates were relatively stable across various values of \textit{k} -- see Additional File 1: Fig S27) provided the dimensionality estimate that best separated challenging subpopulations (Figure \ref{fig:dim}C) and tended to result in the best clustering accuracy (Figure \ref{fig:dim}D).
@@
 
-Of note, the optimal method in our evaluation (\textit{maxLikGlobal})} systematically selected many more components than were associated with the subpopulations in Figure \ref{fig:dim}A, suggesting that although these additional components appear individually uninformative, in combination they nevertheless contribute to classification.
+Of note, the optimal method in our evaluation (\textit{maxLikGlobal}) systematically selected many more components than were associated with the subpopulations in Figure \ref{fig:dim}A, suggesting that although these additional components appear individually uninformative, in combination they nevertheless contribute to classification.
 An important implication is that the commonly-used elbow method most likely underestimates dimensionality.
@@
 
-The MI score and minimum precision, which are largely independent of the estimated number of clusters, were overall higher for the \textit{walktrap} method (Figure \ref{fig:clust}A, left) over \textit{Seurat}, although when used in combination with some networks it sometimes yielded a lower ARI or minimum precision.
-The ARI score at the true number of clusters, when available, showed similar performances, especially when using \textit{sctransform}.
-Because \textit{Seurat}'s resolution parameter had a large impact on the number of clusters identified (Additional File 1: Fig S2 and 24), \textit{Seurat} could always be coerced into producing the right number of clusters.
+{\color{red} The MI score, which is largely independent of the estimated number of clusters, was overall higher for the \textit{walktrap} method (Figure \ref{fig:clust}A, left) over \textit{Seurat} and the \textit{fast-greedy} method.
+Depending on what network it was used with, however, it sometimes yielded a lower ARI or minimum precision.
+The ARI score at the true number of clusters, when available, showed similar performances, especially when using \textit{sctransform}} Because \textit{Seurat}'s resolution parameter had a large impact on the number of clusters identified (Additional File 1: Fig S2 and 24), \textit{Seurat} could always be coerced into producing the right number of clusters.
 Instead, the number of clusters found by \textit{scran} was considerably less influenced by the available parameters (number of nearest neighbors or steps in the random walk - see Additional File 1: Fig S28), and as a result \textit{scran}-based clustering sometimes never produced a partitioning with the right number of clusters.
 This observation, along with \textit{scran}'s higher MI score, suggest that \textit{scran} sometimes simply divides a real subpopulation into two clusters (possibly tracking some unknown biological differences) rather than committing misclassification errors.
-Overall, the \textit{walktrap} method appeared superior to the \textit{fast greedy} algorithm and was generally comparable to \textit{Seurat} clustering, although the latter offered more control over the resolution.
-A major difference between \textit{walktrap}-based clustering and \textit{Seurat} is the computing time (Figure \ref{fig:clust}B).
+{\color{red} Overall, the \textit{walktrap} method appeared superior to the \textit{fast greedy} algorithm and was generally comparable to \textit{Seurat} clustering, although the latter offered more control over the resolution.} A major difference between \textit{walktrap}-based clustering and \textit{Seurat} is the computing time (Figure \ref{fig:clust}B).
 We found that using the Annoy approximation to the nearest neighbors somewhat reduced computing time at no apparent cost to accuracy.
@@
 Unsupervised clustering on the evaluation metrics reveals general patterns in the results as well as the relative importance of the parameters (Additional File 1: Fig S29), with the clustering method having the most important impact on the results.
-This is however chiefly due to number of clusters called: an ANOVA on all analyses after accounting, in the model, for baseline differences between datasets and the impact of \textit{Seurat}'s resolution parameter, attributed relatively little importance to the clustering method, and instead attributed the largest proportion of variance in the main cluster metrics to the normalization method and the number of dimensions used (Figure \ref{fig:overall}).
+This is however chiefly due to number of clusters called: an ANOVA on all analyses after accounting, in the model, for baseline differences between datasets and the impact of \textit{Seurat}'s resolution parameter, attributed relatively little importance to the clustering method, and instead attributed the largest proportion of variance in the main cluster metrics to the {\color{red} normalization method and the number of dimensions used (Figure \ref{fig:overall}).
 Part of the impact of the estimated dimensionality depended on its effect on the number of clusters detected, and it indeed explained less variance in ARI at the true number of clusters (Figure \ref{fig:overall}).
-Nevertheless, the choice of dimensionality (as well as feature selection) came closely behind normalization, suggesting that more attention should be paid to this relatively neglected step.
+Nevertheless, the choice of dimensionality (as well as feature selection) came closely behind normalization, suggesting that more attention should be paid to this relatively neglected step.}
 
-The greater importance of normalization, dimensionality and feature selection over filtering steps is also visible in the clustering (Additional File 1: Fig S29).
-The top methods (bottom of Additional File 1: Fig S29) confirm the results of the previous sections, highlighting especially \textit{sctransform}, deviance-based feature selection, and \textit{maxLikGlobal} dimensionality estimates as top methods.
+The greater importance of normalization, dimensionality and feature selection over filtering steps is also visible in the clustering ({\color{red}Additional File 1: Fig S29}).
+The top methods ({\color{red} bottom of Additional File 1: Fig S29}) confirm the results of the previous sections, highlighting especially \textit{sctransform}, deviance-based feature selection, and \textit{maxLikGlobal} dimensionality estimates as top methods.
 Interestingly, while the combination of \textit{sctransform} with \textit{Seurat} clustering was particularly performant for most datasets, for the simulated `simMix1` dataset,  standard \textit{Seurat} normalization combined with \textit{scran}-based clustering proved superior.
 
-We next investigated whether the benefits of some methods were conditional on the choices at other steps.
+{\color{red}We next investigated whether the benefits of some methods were conditional on the choices at other steps.
 Although analysis of variance identified some significant interactions (Additional File 1: Fig S30), we found that the benefits brought by performant methods were robust to differences in other parts of the pipeline (e.g.
-Additional File 1: Fig S31).
-For instance, we confirmed that doublet removal and \textit{sctransform} normalization were beneficial independently of other steps, and that deviance-based feature selection outcompeted \textit{Seurat}'s default selection method independently of changes at other steps (Additional File 1: Fig S31).
-ANOVA did reveal some significant (negative) interactions (Additional File 1: Fig S30): the top two-way interaction was between \textit{scran} normalization and the elbow method for selection the number of dimensions, and indeed that combination was involved in all of the 30 lowest-ranking pipelines.
+Additional File 1: Fig S31).} For instance, we confirmed that doublet removal and \textit{sctransform} normalization were beneficial independently of other steps, and that deviance-based feature selection outcompeted \textit{Seurat}'s default selection method independently of changes at other steps (Additional File 1: Fig S31).
+ANOVA did reveal some significant (negative) interactions ({\color{red} Additional File 1: Fig S30}): the top two-way interaction was between \textit{scran} normalization and the elbow method for selection the number of dimensions, and indeed that combination was involved in all of the 30 lowest-ranking pipelines.
 Furthermore, all significant interactions involved the elbow method, suggesting that some methods were less robust to underestimated dimensionality.
@@
 We developed 3 benchmark datasets: 1) the \texttt{ipsc} data is a random selection of 10 vs 10 samples from \citep{carcamoOriveAnalysisTranscriptional2017}, which are very heterogeneous, and foldchanges were manually added to 300 genes; 2) the \texttt{seqc} dataset contains 5 vs 5 samples of two mixtures of the SEQC consortium with different ERCC spike-in mixes \citep{ComprehensiveAssessment2014}, and includes a partially correlated batch effect; 3) A \texttt{simulation} of 8 vs 8 samples with two technical vectors of variation (see Methods for more details).
-As previously reported \citep{ComprehensiveAssessment2014}, we found that more stringent filtering improved DEA (Additional File 1: Fig S33A-B) and that the different DEA methods had similar performances, especially after using SVA-based methods (Additional File 1: Fig S33B).
-Including a SVA-based step increased sensitivity across all datasets (Figure \ref{fig:svadea}A and Additional File 1: Fig S33B), particularly in datasets with technical vectors of variation (\textit{seqc} and \textit{simulation}), and benefited all DEA methods in a similar fashion (Figure \ref{fig:svadea}C and Additional File 1: Fig S33C-34A).
+As previously reported \citep{ComprehensiveAssessment2014}, {\color{red} we found that more stringent filtering improved DEA (Additional File 1: Fig S33A-B) and that the different DEA methods had similar performances, especially after using SVA-based methods (Additional File 1: Fig S33B).
+Including a SVA-based step increased sensitivity across all datasets (Figure \ref{fig:svadea}A and Additional File 1: Fig S33B)}, particularly in datasets with technical vectors of variation (\textit{seqc} and \textit{simulation}), and benefited all DEA methods in a similar fashion (Figure \ref{fig:svadea}C and Additional File 1: Fig S33C-34A).
 With the exception of the \texttt{RUVr} method, which led to a high False Discovery Rate (FDR), all other methods showed comparable gains in sensitivity while maintaining a decent error control (if a little above the nominal FDR), even when increasing the number of surrogate variables used (Figure \ref{fig:svadea}B-C).
@@
    \begin{itemize}
-    \item In cases where prior knowledge can guide the choice of a resolution, \textit{Seurat} can be useful in affording manual control of it while, in the absence of such knowledge, \textit{scran}-based \textit{walktrap} clustering provided reasonable estimates, although at greater computational costs.
+    \item In cases where prior knowledge can guide the choice of a resolution, \textit{Seurat} can be useful in affording manual control of it while, in the absence of such knowledge, \textit{scran}-based \textit{walktrap} clustering provided reasonable estimates{\color{red}, although at greater computational costs.}
    \end{itemize}
@@
 
-{\color{red} The performance of \textit{scVI} in our evaluation was lower than in the original publication presenting the method \cite{lopezDeep2018}.
+The performance of \textit{scVI} in our evaluation was lower than in the original publication presenting the method \cite{lopezDeep2018}.
 Our results are however in line with other studies comparing silhouette score \cite{Aparicio2020} and clustering accuracy \cite{HouImput2020} of \textit{scVI}'s latent space.
@@
 fewer than genes), for which \textit{scVI} was reported by its authors to be less adapted.
-We think that further work will be required to assess its performance on larger datasets.}
+We think that further work will be required to assess its performance on larger datasets.
 
@@
 This avoids the need to store potentially large intermediate data for all possible permutations and thus allowing a combinatorial benchmark.
-This, along with the set-up simplicity (demonstrated here by our separate SVA-DEA benchmark), also distinguishes \textit{pipeComp} from general workflow frameworks such as \textit{drake} \citep{landauDrakePackage2018}; indeed we estimate that the benchmarks discussed here would have taken approximately 16TB of storage had all non-redundant intermediates been saved (see the Additional File 2 for more detailed explanations of the framework and comparison to \textit{CellBench} and \textit{drake}).
+This, along with the set-up simplicity (demonstrated here by our separate SVA-DEA benchmark), also distinguishes \textit{pipeComp} from general workflow frameworks such as \textit{drake} \citep{landauDrakePackage2018}; indeed we estimate that the benchmarks discussed here would have taken approximately 16TB of storage had all non-redundant intermediates been saved (see the {\color{red} Additional File 2} for more detailed explanations of the framework and comparison to \textit{CellBench} and \textit{drake}).
 In addition, the fact that benchmark functions are stored in the \texttt{PipelineDefinition} makes the benchmark more smoothly portable, making it easy to modify, extend with further methods, or apply to other datasets.
@@
 
-{\color{red} For the molecular cross-validation approach, we relied on a R implementation by Matthew Young (\url{https://github.com/constantAmateur/MCVR}), with minor modifications (see wrapper in the \textit{pipeComp} package).
+For the molecular cross-validation approach, we relied on a R implementation by Matthew Young (\url{https://github.com/constantAmateur/MCVR}), with minor modifications (see wrapper in the \textit{pipeComp} package).
 Briefly, since there is no \textit{a priori} relationship between the dimensionality of the count data and that of the PCA based on the normalized data, MCVR does not use the original approach based Poisson loss, but the mean squared error of the normalized data, and uses balanced 50-50 splits.
-Because of the 50\% split, dimensionality may be underestimated, and titration did suggest that was the case for some of our datasets.}
+Because of the 50\% split, dimensionality may be underestimated, and titration did suggest that was the case for some of our datasets.
 
@@
 
-The evaluation of \textit{scVI} algorithm was performed separate to the other pipeline combinations since the model training on our reference datasets was especially resource-consuming {\color{red} and follows a different workflow from the other pipelines.
+The evaluation of \textit{scVI} algorithm was performed separate to the other pipeline combinations since the model training on our reference datasets was especially resource-consuming  and follows a different workflow from the other pipelines.
 In particular, for datasets with fewer cells than genes, \textit{scVI} requires a feature selection to be performed prior to the normalization step; furthermore, its dimensional reduction steps rely directly on raw counts.
@@
 All other parameters were left unmodified.
-}
 
@@
 
-The set of alternatives found for the overall run (Figure 8 and Additional File 1: Fig S29-31) can be found in the \href{https://github.com/markrobinsonuzh/scRNA_pipelines_paper/blob/master/data/overall_run.Rmd}{paper's repository}.
+The set of alternatives found for the overall run ({\color{red}Figure 8 and Additional File 1: Fig S29-31}) can be found in the \href{https://github.com/markrobinsonuzh/scRNA_pipelines_paper/blob/master/data/overall_run.Rmd}{paper's repository}.
 The analysis of variance was performed with the following model:
@@
 Each step can accept a number of parameters whose alternative values are provided as a list.
-All (or subsets of) combinations of parameters can then be simultaneously ran and evaluated using the \texttt{runPipeline} function (see Additional File 2 for more details).
+All (or subsets of) combinations of parameters can then be simultaneously ran and evaluated using the \texttt{runPipeline} function ({\color{red}see Additional File 2 for more details}).
 \textbf{B:} Scheme representing the application of \textit{pipeComp} to evaluate a range of methods that are commonly used in scRNAseq studies and some of the metrics monitored at various steps.}
@@
 The dashed line indicates a threshold of 2.5 median absolute deviations (MADs) from the median of the overall population.
-\textbf{B:} Relationship between the maximum subpopulation exclusion rate and the average clustering accuracy per subpopulation across various filtering strategies.
-Of note, doublet removal appears to have a neutral or positive impact even when, due to the design, there are no heterotypic doublets in the data.
+\textbf{B:} Relationship between the maximum subpopulation exclusion rate and the average clustering accuracy per subpopulation across various filtering strategies.{\color{red} The datasets for which the presence of doublets could be confirmed with SNP genotypes are labeled in bold.
+Of note, doublet removal appears to have a neutral or positive impact }even when, due to the design, there are no heterotypic doublets in the data.
 The PCA methods refer to multivariate outlier detected as implemented in \textit{scater} (see Methods for details).}
@@
 \textbf{C:} Clustering accuracy (\textit{Seurat} clustering), measured by mutual information (MI), Adjusted Rand Index (ARI) and ARI at the true number of cluster.
-{\color{red}Grey squares indicate that the true number of clusters could not be reached with the corresponding method, dataset and parameter set.} `feat\_regress', `mt\_regress' and `feat\_mt\_regress' respectively stand for the regressing out of the number of features, the proportion of mitochondrial reads, or both during scaling.
-{\color{red}Throughout this paper, silhouette plots (panel \textbf{B}) square-root transform values for color mapping.
+Grey squares indicate that the true number of clusters could not be reached with the corresponding method, dataset and parameter set.} `feat\_regress', `mt\_regress' and `feat\_mt\_regress' respectively stand for the regressing out of the number of features, the proportion of mitochondrial reads, or both during scaling.
+Throughout this paper, silhouette plots (panel \textbf{B}) square-root transform values for color mapping.
 Other evaluation metric heatmaps (e.g.
@@
 The numbers printed in the cells represent the raw (i.e.
-unscaled) metric values, and are printed in white or black depending on whether they are above or below the median.}}
+unscaled) metric values, and are printed in white or black depending on whether they are above or below the median.
     \label{fig:norm}
@@
     \caption{\textbf{Evaluation of clustering methods.} Clustering accuracy (\textbf{A}) and run time (\textbf{B}) of \textit{scran}- and \textit{Seurat}-based clustering tools in combination with \textit{sctransform} and standard \textit{Seurat}'s normalization.
-{\color{red} The color-mapping scheme in panel \textbf{A} and the left of panel \textbf{B} is the same as in Figure \ref{fig:norm}C.}}
+The color-mapping scheme in panel \textbf{A} and the left of panel \textbf{B} is the same as in Figure \ref{fig:norm}C.}
     \label{fig:clust}
@@
     \includegraphics[width=0.55\textwidth, keepaspectratio]{{anova.pdf}}
-    \caption{\textbf{Variance in clustering metrics explained by the parameters.} See methods for more detail.}
+    \caption{{\color{red}\textbf{Variance in clustering metrics explained by the parameters.} Proportion of variance explained by the parameters of the main analysis steps.
+The proportions are calculated from an ANOVA using the model;\\
+    \texttt{metric$\sim$ dataset*resolution+doubletmethod+filt+norm+clustmethod+dims}
+    \\ and substracting the variance attributable to differences due to datasets, resolutions and their interactions. }}
     \label{fig:overall}
@@
     \includegraphics[width=\textwidth, keepaspectratio]{{svadea.pdf}}
-    \caption{{\color{red}\textbf{Evaluation of the impact of surrogate variable analysis (SVA) on (bulk) RNAseq differential expression analysis using \textit{pipeComp}.} \textit{RUVr} and \textit{RUVs} refer to the two \textit{RUVSeq} functions by the same name; \textit{svaseq} refers to the function of the same name in \textit{sva}, while \textit{vstsva} refers to the original \textit{sva} method applied no variance-stabilized data.
-\textbf{A:} True positive rates (TPR) and False discovery rates (FDR) of the SVA-based methods (using a single surrogate variable), averaged across DEA methods.
-\textbf{B:} TPR and FDR of the different SVA-based methods (averaged across the different DEA methods) using increasing numbers of variables.
+    \caption{\textbf{Evaluation of the impact of surrogate variable analysis (SVA) on (bulk) RNAseq differential expression analysis using \textit{pipeComp}.} \textit{RUVr} and \textit{RUVs} refer to the two \textit{RUVSeq} functions by the same name; \textit{svaseq} refers to the function of the same name in \textit{sva}, while \textit{vstsva} refers to the original \textit{sva} method applied no variance-stabilized data.
+\textbf{A:} True positive rates (TPR) and False discovery rates (FDR) of {\color{red} the SVA-based methods (using a single surrogate variable), averaged across DEA methods.} \textbf{B:} TPR and FDR of the different SVA-based methods (averaged across the different DEA methods) using increasing numbers of variables.
 In \textbf{A-B}, the x-axis is square-root transformed to improve readability, and three dots for each combination of methods refer to the nominal FDR thresholds of 0.01, 0.05, and 0.1.
@@
 As in \ref{fig:norm}C, the color mapping tracks the number of (matrix-wise) Median Absolute Deviations from the (column-wise) median, while the printed numbers represent the raw metric values.
-`be' and `leek` refer to the two methods implemented in the \textit{sva} package for estimating the number of surrogate variables.}}
+{\color{red} \textit{be} and \textit{leek} refer to the two methods implemented in the \textit{sva} package for estimating the number of surrogate variables.}}
     \label{fig:svadea}
@@
 \begin{table}[h!]
-\caption{\textbf{Table S1: Overview of the benchmark datasets.} \# features' indicates the number of features detected in at least 10 cells.}
+\caption{{\color{red} \textbf{Table S1:} Overview of the benchmark datasets.} \# features' indicates the number of features detected in at least 10 cells.}
 \label{tab:table1}

### 2020-08-06 bmc_article.tex
Updates from Overleaf

@@
 We also found that {\color{red} \textit{scDblFinder} clearly improved the accuracy of the clustering across datasets that are expected to have heterotypic doublets (mixology datasets), and even in a simulation (simMix1) that should not contain any (Figure \ref{fig:filtering}).
-Across the datasets that, being FACS-sorted, should not contain such doublets, doublet removal tended not to robustly impact clustering accuracy, however in few datasets (ZhengMix4eq and Zhengmix4uneq, and simMix2 with a very small difference) doublet removal performed worse when combined with lenient or no filtering.}
+Across the datasets that, being FACS-sorted, should not contain such doublets, doublet removal tended not to robustly impact clustering accuracy.
+However in few datasets (ZhengMix4eq and Zhengmix4uneq, and simMix2 with a very small difference) doublet removal performed worse when combined with lenient or no filtering.}
 
@@
     \includegraphics[width=0.55\textwidth, keepaspectratio]{{anova.pdf}}
-    \caption{{\color{red}\textbf{Variance in clustering metrics explained by the parameters.} Proportion of variance explained by the parameters of the main analysis steps.
-The proportions are calculated from an ANOVA using the model;\\
-    \texttt{metric$\sim$ dataset*resolution+doubletmethod+filt+norm+clustmethod+dims}
-    \\ and substracting the variance attributable to differences due to datasets, resolutions and their interactions. }}
+    \caption{\color{red}\textbf{Variance in clustering metrics explained by the parameters.} Proportion of variance explained by the parameters of the main analysis steps.
+The proportions are calculated from an ANOVA using the model;
+    \texttt{metric $\sim$ dataset*resolution+doubletmethod+filt+norm+clustmethod+dims}
+    and substracting the variance attributable to differences due to datasets, resolutions and their interactions. }
     \label{fig:overall}

### 2020-08-06 bmc_article.tex
Updates from Overleaf

@@
 Here, we develop a flexible R framework for pipeline comparison and apply it to the evaluation of the various steps of analysis leading from an initial count matrix to a cluster assignment, which are critical in a wide range of applications.
-We collected real datasets of known cell composition ({\color{red}Table S\ref{tab:table1}}) and used a variety of evaluation metrics to investigate in a multilevel fashion the impact of various parameters and variations around a core scRNAseq pipeline.
+We collected real datasets of known cell composition ({\color{red}Table \ref{tab:table1}}) and used a variety of evaluation metrics to investigate in a multilevel fashion the impact of various parameters and variations around a core scRNAseq pipeline.
 Although we use some datasets based on other protocols, our focus is especially on droplet-based datasets that do not include exogenous control RNA (i.e.
-spike-ins); see {\color{red}Table S\ref{tab:table1}} and Additional File 1: Fig S1 for more details.
+spike-ins); see {\color{red}Table \ref{tab:table1}} and Additional File 1: Fig S1 for more details.
 In addition to previously-used benchmark datasets with true cell labels \cite{duoClustering2018,tianMixology2018}, we simulated two datasets with a hierarchical subpopulation structure based on real 10x human and mouse data using \textit{muscat} \cite{CrowellMuscat2019}.
@@
 \textbf{C:} Clustering accuracy (\textit{Seurat} clustering), measured by mutual information (MI), Adjusted Rand Index (ARI) and ARI at the true number of cluster.
-Grey squares indicate that the true number of clusters could not be reached with the corresponding method, dataset and parameter set.} `feat\_regress', `mt\_regress' and `feat\_mt\_regress' respectively stand for the regressing out of the number of features, the proportion of mitochondrial reads, or both during scaling.
+Grey squares indicate that the true number of clusters could not be reached with the corresponding method, dataset and parameter set.
+`feat\_regress', `mt\_regress' and `feat\_mt\_regress' respectively stand for the regressing out of the number of features, the proportion of mitochondrial reads, or both during scaling.
 Throughout this paper, silhouette plots (panel \textbf{B}) square-root transform values for color mapping.
@@
 The numbers printed in the cells represent the raw (i.e.
-unscaled) metric values, and are printed in white or black depending on whether they are above or below the median.
+unscaled) metric values, and are printed in white or black depending on whether they are above or below the median.}
     \label{fig:norm}
@@
     \caption{\color{red}\textbf{Variance in clustering metrics explained by the parameters.} Proportion of variance explained by the parameters of the main analysis steps.
-The proportions are calculated from an ANOVA using the model;
-    \texttt{metric $\sim$ dataset*resolution+doubletmethod+filt+norm+clustmethod+dims}
+The proportions are calculated from an ANOVA using the model \hspace{\textwidth}
+    \texttt{metric $\sim$  dataset*resolution+doubletmethod+filt+norm+clustmethod+dims} \hspace{\textwidth}
     and substracting the variance attributable to differences due to datasets, resolutions and their interactions. }
@@
 \begin{table}[h!]
-\caption{{\color{red} \textbf{Table S1:} Overview of the benchmark datasets.} \# features' indicates the number of features detected in at least 10 cells.}
+\caption{Overview of the benchmark datasets. \# features' indicates the number of features detected in at least 10 cells.}
 \label{tab:table1}

### 2022-06-27 datasets.Rmd
added code to generate the dataset description

---
title: "datasets overview"
author: "Pierre-Luc Germain"
date: "10/23/2019"
output: html_document
---

```{r}
suppressPackageStartupMessages({
  library(ggplot2)
  library(cowplot)
  library(pipeComp)
})
theme_set(theme_cowplot())
ds <- list.files("datasets/", full.names=TRUE)
names(ds) <- gsub("\\.SCE\\.rds","",basename(ds))

sces <- lapply(ds, FUN=function(x){
  library(scran)
  library(scater)
  sce <- readRDS(x)
  sce <- logNormCounts(sce)
  var.stats <- modelGeneVar(sce)
  sce <- denoisePCA(sce, technical=var.stats)
  reducedDim(sce, "tSNE") <- Rtsne::Rtsne(reducedDim(sce))$Y
  reducedDim(sce, "umap") <- uwot::umap(reducedDim(sce))
  sce
})
```

```{r, fig.width=8.5, fig.height=12}
describeDatasets(sces, pt.size = 0.4, alpha=0.7)
```
